### 2022-08-12 writing/manuscript.Rmd
initial push

---
title: "How do resource abundance and stochasticity affect animals' spatial needs?"
subtitle: "A review with quantitative simulations and empirical modeling"
author: "Mezzini S. and Noonan M. J."
date: "`r Sys.Date()`"
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{setspace}\doublespacing # double-space text
    - \usepackage{indentfirst} # indent first line of each paragraph
    - \usepackage[small]{titlesec} # smaller font for headings
    - \usepackage{wrapfig} # to wrap text around figures
    - \usepackage{caption}
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-20pt}

    # - \usepackage{gensymb}
    # - \usepackage{float}
    # - \usepackage{afterpage}
    # - \usepackage{ragged2e}
subparagraph: yes # needed for \usepackage[small]{titlesec}
# classoption:
# - twocolumn
output:
  bookdown::pdf_document2:
    number_sections: true
    toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
```

```{r wrap-figure-function, include = FALSE}
# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures

# this chunk is needed to wrap text around or beside figures

defOut <- knitr::knit_hooks$get('plot')  # save the default plot hook

knitr::knit_hooks$set(plot = function(x, options) {  # set the new plot hook ...

  x <- defOut(x, options)  # first apply the default hook

  # if wrapfigure options are given, create the new opening string for the wrapfigure
  # environment using custom figure alignment and text width
  if(!is.null(options$wrapfigure)) {

    # options must be numeric, they cannot be characters
    wf <- sprintf('\\begin{wrapfigure}{%s}{%g\\textwidth}',
                  options$wrapfigure[[1]], # figure alignment
                  options$wrapfigure[[2]]) # caption width relative to text width
    x  <- gsub(pattern = '\\begin{figure}', # and replace the default one with it
               replacement = wf,
               x = x,
               fixed = TRUE) # pattern is a string, so evaluate as is
    x  <- gsub(pattern = '{figure}', # also replace the environment ending
               replacement = '{wrapfigure}',
               x = x,
               fixed = TRUE)
  }
  return(x)
}
)

# chuck options used to wrap figures:
## fig.align = "right":          align figures right
## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width
```

<!-- place figures before the paragraph they are referred to so wrapfigure places them beside the paragraph -->

<!-- add section explaining what the paper details and what the objectives are? -->

@ciuti and @berger-tal: human presence in nature for other reasons than hunting
"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
habitat productivity hypothesis: Harestad and Bunnel, 1979

Add @mueller_how_2011 for a study that accounted for variance in NDVI for animal movement

The habitat productivity hypothesis proposed by @harestad_home_1979 is supported, but it ignores many other important causes of movement [@seigle-ferrand_systematic_2021].

The ability to move allows animals to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977]. Animals may move for a variety of reasons, including searching for resources (e.g., food, nutrients, water, heat, a new den or nesting spot), a mate, or a new group. Movement also allows animals to escape predation or dangerous competition as well as defend resources and territories [@grant_whether_1993]. @nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model

\begin{equation} \label{nathan-movement}
\vec u_{t+1} = F(\Omega, \Phi, \vec r_t, \vec w_t, \vec u_t),
\end{equation}

such that the animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$). This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible). The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters, doing so with sufficient frequency and accuracy would not be simple. Rather than measuring changes in the various resources and needs directly, we can quantify animals' spatial needs as a response to average resource abundance and the variability around it, instead. This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it. We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI; see @pettorelli_normalized_2011].

# Effects of resource abundance on animals' spatial needs

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and average $\\mathbb V(R)$. (d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and average $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk. Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$). Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$). Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.]. Let $R$ indicate the amount of *resources* in a particular patch or area. For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity. However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.].

Consider the simplistic scenario where $\mu$ changes over time but $\sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a). We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time: $\mathbb E(R) = \mu(t)$) while the variance is constant, i.e., $\mathbb V(R) = \sigma^2$ (figure \@ref(fig:multipanel-constant-var)b). Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d). However, if $\mathbb E(R) = \mu(t)$ changes repetitively and regularly over time (e.g. peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019]. However, since $\mathbb E(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.

This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize. Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance. Examples of temporally homogeneous habitats include regions which throughout the year remain highly productive but fairly predictable (e.g., equatorial rain forests) or have very low productivity and variability (e.g., deserts without strongly distinct seasons). When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case. Most habitats with seasonal changes in climate will require models to account for changes in both $\mathbb E(R)$ and $\mathbb V(R)$, since periods of high productivity will tend to be more variable than times of "dormancy" or lower productivity.

The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where the average resource abundance remains constant or changes over time. Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$, constant $\mathbb V(R)$), and how the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range. In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$, constant $\mathbb V(R)$ ), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$. However, estimates of spatial requirements based on estimated changes in $\mathbb E(R)$ should be interpreted carefully, since model error and unpredictable changes in $\mathbb E(R)$ may increase animals' spatial needs unpredictably. Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs. This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$, constant $\mathbb V(R)$), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible. In cases where $\mathbb E(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often. Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$, constant $\mathbb V(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.

# Effects of environmental variance on animals' spatial needs

Although the effect of resource abundance on animals' spatial needs is often recognized and accounted for, the variance in resource abundance is generally ignored or assumed to remain constant over time. However, resource abundance often varies due to multiple causes. Some causes are repetitive, predictable, or even well-known patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches), while others are infrequent, unpredictable, or poorly-understood events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018]. In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity. We refer to predictable or expected spatiotemporal changes as environmental **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation. For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable. In contrast, the quantity and quality of berries produced during a given year may be stochastic. Thus, $\mathbb V(R)$ is affected by both sources of variation, and changes in $\mathbb V(R)$ may predictable or not.

The effects of $\mathbb V(R)$ on animals' spatial requirements likely depend on the animals' needs. @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous) required exponentially less space than foxes that lived in more homogeneous regions. @nilsen_can_2005 found that the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality, while the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased. @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity. ***ADD MORE EMPIRICAL EXAMPLES?*** However, the effects of environmental variance on animals' spatial needs remains largely undestudied and unaccounted for.

# Empirical examples

```{r tapir-example, fig.cap="Seven-day home range size of a tapir from Medici $et~al.$ (2022) as mean and variance in resource abundance change over the tracking period. (a.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir during the period of reference. (b.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir during the period of reference. (c.) Estimated home range size of the tapir during each reference period, based on 50% (bold) and 95% (thin) utilization distributions.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/tapir-example.png')
```

Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [@lucherini_habitat_1996; @relyea_home_2000]. While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance. The size of a home range, is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [@nilsen_can_2005], since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004].

There are many factors which may affect animal's spatial requirements and how they use their HR [*sensu* @nathan_movement_2008]. In particular, resource abundance and environmental productivity are often hypothesized to be inversely proportional to HR size (or some function of it), such that regions with higher abundance correspond to smaller spatial needs, since animals do not have to range over extensive areas to meet the energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @southwood_habitat_1977], and maintaining a thermoregulatory balance. The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear, since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004]. @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region. However, the effect of many other factors on HR size is likely depends on the species, location, and possibly also on the individual animal. For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; capuchin monkeys, @torrez-herrera_monkeys_2020]. Similarly, predation may force animals to move more frequently to escape predators, or it may prevent them from venturing too far from the safety of their core HR too often [@suraci_beyond_2022]. Patch quality, size, fragmentation, and heterogeneity may cause animals to explore more patches if some are of low value, too small, too disconnected, or too variable [@fahrig_is_2019], but high diversity may also decrease HR size if animals require heterogeneous habitats [@fox_niche_1981; @lucherini_habitat_1996]. Similarly, patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006]. However, not all animals take advantage of linear features or higher predictability in human-altered habitats. @noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed. Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat. Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.

Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022

Simulation refs: @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021

<!--
Notes:

notation:
- $\mu$ mean movement process
- $\epsilon$ movement noise, exploratory movements
- $\tau_p$ positional autocorrelation, mean time reversion to the mean (origin), range crossing time
- $\tau_v$ velocity autocorrelation

protective/defensive adaptations peaks at intermediate levels of competitor density/spatial resource density/spatiotemporal clumping. At low competition, there's no need for defense and the cost (e.g., energetic) and risk (e.g., wounds) outweigh the benefit, while at high values the cost and risk become too high and they outweigh the benefit [@grant_whether_1993].

See *The effects of phenological mismatches on demography*.

@rickbeil_plasticity_2019:
- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
- Elk departed earlier from highly-productive areas? This goes against the idea that animals move as a consequence of need. But maybe the elk simply left when they had enough stored energy to face the migration. Maybe elk migration is triggered when they have sufficient energy (if the weather is still not too harsh).
- Elk left early when exposed to hunting more
-->

# Simulating spatial needs as a function of $\mathbb E(R)$ and $\mathbb V(R)$

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance). (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks. (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$. (c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$. (d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

<!-- here! -->

In areas where animals are not guaranteed that the resources they find during one visit will be there the next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness. Patches with low or high $p$ will be most predictable, since successes can be expected to be very rare (if $p \approx 0$) or very common (if $p \approx 1$). In contrast, patches will be most stochastic when the probabilities of success and failure are approximately the same (i.e., $p \approx 1 - p \implies p \approx 0.5$, see figure \@ref(fig:multipanel-changing-var)b). In stochastic habitats, $\mathbb E(U)$ will depend on $\mathbb E(S)$ as well as $\mathbb E(R)$, since $S$ is no longer constant (figure \@ref(fig:multipanel-changing-var)c). Now, expected usable resources become

\begin{equation} \label{mean-u-definition}
\mathbb E(U) = \mathbb E(R \, S),
\end{equation}

or

\begin{equation} \label{mean-u-definition-indep}
\mathbb E(U) = \mathbb E(R) \, \mathbb E(S)
\end{equation}

if $R$ and $S$ are independent. This model can be applied to all mobile animals, including herbivores, carnivores, and omnivores. In the case of herbivores, $p$ may indicate the chance of finding good forage, which may be absent if regeneration times are long, if competitors have already exhausted the resource, or following a fire. For carnivores, $p$ may indicate the chance of feeding on prey, which may depend on encountering and killing some first [@suraci_beyond_2022]. However, since $R$ is not limited to energetic resources, $p$ may also indicate the chance of finding water or also a suitable location for a den or nest.

While resource availability is often considered in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], an environment's heterogeneity (i.e., diversity), stochasticity (i.e., unpredictability), and how the two change over time are rarely accounted for. In addition, environmental stochasticity, including extreme events, can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness. Therefore, I expect animals living in unpredictable environments to require more space than those in stable environments. Although this hypothesis is supported by a few recent studies [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], many of them are limited in their analytical depth and geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited. There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity. These stresses are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018]. Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018]. As the impacts of habitat loss and climate change will worsen in the future [@ipcc_global_2018; @hansen_high-resolution_2013], it is imperative that we better understand spatial requirements of taxa to protect wildlife existence and biodiversity. Environmental safeguarding is also essential for Reconciliation with Indigenous People in Canada [@truth_and_reconciliation_commission_of_canada_honouring_2015].

# Interaction effects of resource availability and stochasticity

@nilsen_can_2005

In the previous section, we mentioned that a model which does not account for environmental variance may be acceptable when $\sqrt{\mathbb V(R)} / \mathbb E(R)$ is low, as in the case of a habitat with high $R$ and $p \approx 1$. However, since maximum resource abundance is constrained by environmental stochasticity [@chevin_adaptation_2010] and resources can be depleted or rendered inaccessible by other individuals [@grant_whether_1993; @jetz_scaling_2004], $\mathbb E(R)$ does not provide a sufficiently complete picture of resource availability in an environment. Thus, the effect of $\mathbb V(R)$ should be included when possible, even when $p \approx 1$ or $p \approx 0$. <!--Spatiotemporally constant and homogeneous regions will allow animals to have a small and dependable home range, while stochastic regions will force animals to have higher spatial needs and possibly store resources for periods of scarcity. Similarly, heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one. Conversely, if some patches are sufficient and highly valuable, animals may adopt strong defensive behaviors towards the patch [@grant_whether_1993]. Thus, some measure of $\mathbb V(R)$ is needed to understand how stochasticity may affect spatial requirements.-->

Let $H$ indicate the size of an animal's HR.^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$. Statistically, we can imagine $H$ as having a probability distribution with support over the interval from zero (not included) to infinity (also not included), which we can indicate with the notation $H \in (0, \infty)$.] As explained above, $H$ will be higher in regions with lower $\mathbb E(U)$ or higher $\mathbb V(U)$. Figure \@ref(fig:5-5-sims) presents various scenarios with different trends in $\mathbb E(U)$ and $\mathbb V(U)$. Although some of these scenarios may seem overly simplistic, they can help us understand the complex interaction effects of $\mathbb E(U)$ and $\mathbb V(U)$. In regions where $U$ has constant mean and variance, $H$ remains approximately constant, although small oscillations are possible, since $\mathbb V(U) \ne 0$ and thus the environment is still stochastic. When $\mathbb E(U)$ changes over time but $\mathbb V(U)$ remains constant, it is easy to see that $H$ decreases as $\mathbb E(U)$ increases, with changes in $\mathbb E(U)$ having larger effects when $\mathbb E(U)$ is low and smaller effects when $\mathbb E(U)$ is already large. This is best visualized in the scenario in which $\mathbb E(U)$ is increasing linearly while $\mathbb V(U)$ is constant, since decrease in core HR ($H_{50\%}$) and 95% HR ($H_{95\%}$) is decreases over time. Additionally, as $\mathbb E(U)$ increases, the difference between $H_{50\%}$ and $H_{95\%}$ also decreases. Thus, $H_{95\%}$ is more sensitive to changes in $\mathbb E(U)$ than $H_{50\%}$. $H_{95\%}$ is also more sensitive to changes in $\mathbb V(U)$, since changes in $\mathbb V(U)$ cause greater oscillations in $H_{95\%}$ than in $H_{50\%}$.

```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time. The thicker line indicates the animal's core home range (0.5 quantile), while the thin line indicates the 0.95 utilization quantile. Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase linearly as $\\mathbb V(R)$ increases. Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks each day starting from the point (0, 0) until they reach satiety, at which point they returned to (0, 0) over the same amount of time. The animal's spatial variance parameter was then calculated using an OUF model via the $ctmm$ package", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\mathbb E(U)$ and $\mathbb V(U)$, and their impacts on animal space use, are useful examples that can be thought of as simplified scenarios. $\mathbb E(U)$ and $\mathbb V(U)$ may be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently available that changes in $\mathbb E(U)$ and $\mathbb V(U)$ remain undetected. Although it is impossible for $\mathbb E(U)$ and $\mathbb V(U)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease). Additionally, these examples are important because they demonstrate the relationships between $H$, $\mathbb E(U)$, and $\mathbb V(U)$ in a (relatively) simple scenario. Cyclical oscillations in $\mathbb E(U)$ and $\mathbb V(U)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\mathbb E(U)$ and $\mathbb V(U)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity. Finally, erratic changes in $\mathbb E(U)$ and $\mathbb V(U)$ may occur in environments where changes are very sudden, such as fire-prone or floor-prone areas, or habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation). However, if highly stochastic or erratic changes occur frequently, animals are most likely to perceive them as a smooth transition rather a series of small, sudden, changes. Estimating the true trend would often require an excessively high cognitive capacity and an equally unlikely abundance of information. Additionally, although changes in $\mathbb E(U)$ are not due to $\mathbb V(U)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020]. <!-- https://doi.org/10.1080/00207549508930162 ?-->

# Empirical examples

## Sliding-window sampling

<!-- create a pdf for supplementary information on how to create the 5x5 panel and run methods w real data, as well as considerations and potential issues about modeling means and variances -->

```{r tapir-hr, fig.cap="Movement of a tapir during a monitoring period of approximately 3 years (1997-07-10 to 2000-06-08) used in the work by Medici $et~al.$ (2022). Different subsamples may result in different estimates of home range size and centroid (a, b; each span 100 days), so neither subsample is necessarily representative of the space the animal used over the entire monitoring period (c).", out.width = '0.4\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.4)}
knitr::include_graphics('../figures/hr-and-lifespan.png')
```

In the case of animals whose spatial needs change over the their lifetimes (figure \@ref(fig:tapir-hr)), one may define HR as a function of time (with distinct HRs for each period, e.g., salmon spawning) or as a continuous function of age (so the HR changes smoothly). In either case, the accuracy of the estimate parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019]. Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.

spatial needs can change as requirements from the animal change or as $R$ changes.

<!-- see https://doi.org/10.1186/s40462-019-0187-z regarding HR, energetics, and movement along paths -->

* tapirs
* tapir w sliding window

<!--

# Movement types

## Range residency

## Seasonal migration

## Nomadism

predation is more stochastic than herbibory; predators need to match more cycles, depend on less individuals -> more variable food

@teitelbaum_beyond_2019:
- **Nomadic movement**: with irregular timing and/or direction, producing both within-year and between-year variability in location and movement patterns
- **Nomadic movement** is not the same as **migration**, because migration is predictable and cyclical while nomadism is random in time and space
- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"

## Edge dynamics

-->

<!--

# Simulations

Include:

* effects of animal mass,
* resistance to/ease of movement (e.g., wolves and clear-cut gas lines)
* "state-space model" with 2 spaces: within a patch (slow-moving) and between patches (faster movement)

See `R` packages for meta-community simulations

-->

# Future work

The work presented here provides a general framework for viewing animal movement as a function of mean and variance in resource abundance. The models used here do not account for animals' traits or behaviors (e.g., energetic needs, food preferences, ability to move, competition) as well as any environmental barriers to movement, such as lakes, rivers, and anthropogenic environments. Future models can account for more stuff, but modeling movement often already accounts for most of this.

How does this impact the field, future papers?

\newpage <!-- place references on a separate page -->

## The temporal scale of stochastic events (DELETE THIS SECTION?)

The timescale over which an event is (un)predictable will also play a key role in determining how species respond, or not, to stochasticity. Adverse events which occur frequently during an organism's generation time or lifespan (e.g., at least twice, on average) may be considered as threats worth preparing for. Stochastic and destructive events which happen frequently can become part of an organism's habitat and life history, so organisms may either adapt to their occurrence and even learn to predict them [@foley_severe_2008], or they may avoid the risk altogether by moving to a new habitat [@southwood_habitat_1977]. If habitats are (or become) unfavorable, animals may move to a new habitat, migrate seasonally [@geremia_migrating_2019], or simply prefer nomadism over range residency [e.g., @nandintsetseg_variability_2019]. It is unclear when animals switch from range residency to nomadism (or vice-versa), but understanding the connection between the two types of movement is important for quantifying the effect of spatiotemporal stochasticity on animal's spatial needs. From a quantitative perspective, the switch is related to an animal's positional autocorrelation over time (also known as home range crossing time, here indicated as $\tau_p$). Animals without a constant home home range (i.e., without a constant centroid) will spend more time away from the overall mean position, so the time required to crossing their entire range time will be large (i.e., on the order of the animal's lifespan).

Adverse events which are too infrequent (e.g., $p \lesssim 0.1$) may not be perceived as a reoccurring threat, so organisms may be unable to adapt appropriately. Instead, highly infrequent events are more likely to perceived as an oddity rather than something worth preparing for. In contrast, highly stochastic events (e.g., $p \approx 0.5$ or if $p$ changes unpredictably) may be perceived as a threat, but organisms may be unable to predict their occurrence, since it would require refined cognitive abilities. For a species or population to adapt to an event, the event must thus occur with sufficient frequency and for a sufficiently long period of time.

```{r stochasticity-timescale-figure, fig.cap="Fictitious changes in resource abundance mean (a) and variance (b) over the course of a year. The true trends are indicated by the black line, while the colored lines indicate the estimates perceived based on different levels of adaptability. The dashed vertical line indicates the hypothetical current date, such that any data to the right of the line is unknown to the animals, so estimates on the right of the dashed line are extrapolations. The data were simulated using Gaussian noise to ensure mean-variance independence and modeled using Generalized Additive Models with an identity link function to allow linear responses.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
knitr::include_graphics('../figures/stochasticity-timescale.png')
```

What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability. Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977]. Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g. elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004], which allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015]. However, the short generation time and high fertility of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can allow them to develop traits that increase the chances of survival following an extreme event. Additionally, the effects of size and lifespan on sensitivity are likely nonlinear and correlated, since smaller animals tend to have shorter lives (and vice-versa), and small animals also tend to have lower metabolic rates, which often limit their movement speed, home ranges, and how the animals interact with their ecosystem [@brown_toward_2004]. Thus, processes will have stronger impacts on smaller animals than on bigger ones: The grazing pattern of a bison (*Bison bison*) drastically alters the habitat of most crawling insects, but other grazing mammals would not pay attention to changes in grass length at the same spatial or temporal resolution. Similarly, the timing and quantity of yearly snowfall would be a somewhat predictable and expected event for most adult moose (*Alces alces*), but it may be a shock for many adult zooplankton or a new-born wolf (*Canis lupus*). What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal. <!-- add empirical examples -->

Figure \@ref(fig:stochasticity-timescale-figure) illustrates the perceived resource availability mean and variance by animals with four different levels of adaptability and how they may predict the mean and variance to change in the future (after the vertical grey line). In the extreme (and likely fictitious) case where an animal does not adapt to the environment (dark blue), it cannot react to or predict changes in mean resource abundance, and thus the perceived variance is inflated whenever the resource abundance does not match the overall mean. Environments appear highly unpredictable to such animals. Animals which can only perceive simple, linear changes in mean resource abundance (orange) perform better as long as resources continue to change linearly in the same direction (i.e. increasing or decreasing). However, they may be surprised when the trend's direction changes, as indicated by the rapid increase in variance in the top rows of figure \@ref(fig:stochasticity-timescale-figure)b. Animals which are able to adapt smoothly (light blue) can predict changes in mean without a significant bias in estimated variance, as long as the process is not highly stochastic (unsurprisingly, see the bottom rows of figure \@ref(fig:stochasticity-timescale-figure)). Finally, animals which adapt rapidly to changing environments may be most able to take advantage of frequent environmental changes, but they may be unable to produce reasonable predictions based on memory [*sensu* @fagan_spatial_2013; @abrahms_memory_2019]. Instead, they depend on constant information and only predict on the most recent information (as indicated by the deviations from the data when predicting into the future).

<!-- The recent shift to more quantitative statistical methods in environmental biology has pushed scientists to begin thinking more critically about data analysis. However, many still operate using methods and frameworks which were necessary with small datasets, but likely become problematic with an abundance of data. For instance, the widespread frequentist approach of hypothesis testing with a significance level of $\alpha = 0.05$ may often be misleading, because the significance of an effect depends on its standard error, which is a function of its sample size, since larger samples will offer a more precise and accurate measurement of the true mean. Similarly, the question of how much data is needed to detect a significant trend [@gray_how_2018] may seem important to some, but I suggest we pay less attention to whether an event or trend is statistically significant over a period of time based on the dichotomy of significance based on an arbitrary significance level. Since statistical significance does not only depend on effect size but also sample size (including measurement frequency), high-frequency data is likely to produce statistically significant effects, even in the event of a false positive. Instead, we should focus primarily on whether a change has an important impact on the fitness of the animal(s) of interest [@zuur_mixed_2009]. -->

An animal's ability to alter its behavior (including its movement) in response to environmental conditions is essential in stochastic or changing landscapes. Whether the changes be due to highly variable but (potentially) predictable changes such as the seasons and the weather, or whether they be due to more stochastic events [e.g., natural events such as fires or floods, but also anthropogenic events like oil spills, see @matkin_ongoing_2008], an animal's ability to adapt increases its odds of survival. Generally, temporal variation is more likely to promote plasticity over diversity, since adaptability will likely offer better odds of survival than temporally static diversity in a population or species [@bell_spatial_1993]. @rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on variable environmental events and cues such as available forage biomass, hunting pressure, snow fall, and snow melt. Birds have also shown to change their their migration as the climate changes, including large-scale, trans-Saharian migrants [@jonzen_rapid_2006]. And while genetic diversity and polymorphisms also increase the odds of survival for a species [@cavedon_genomic_2022], this project will focus strictly on animal behavior.

In an environment that changes over time, organisms which depend on mutable cycles such as changes in temperature, precipitation, and resource availability are more likely to respond to environmental changes than organisms which depend purely on deterministic cycles (e.g., photoperiod). While this thesis focuses on the movement and spatial use of animals, the adaptability (or lack thereof) of non-animal organisms is also crucial. The ability of most animals to rely on visual cues and move accordingly greatly increases their adaptability and plasticity, particularly for those animals that are able to move large distances over short periods of time (e.g., flying birds and large vertebrates). Although this project will focus on vertebrate animals, such changes are important to consider because the ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.

## The spatial scale of stochastic events (DELETE THIS SECTION?)

To measure the spatial stochasticity of an ecosystem, it is necessary to first define extent of the ecosystem of interest. A lake may seem like a simple and self-contained space to define, but it is often not as simple as deciding the lake's boundaries using a map or satellite pictures. Should one include the shores of the lake? And if so, which region do we consider the lake's shores? Are they the area of land which are currently covered by water, including the sections hit daily by waves, or should we also include the parts of land which are currently not wet but are covered during seasons with more rain? One could base their definition on the species of interest, since coastal areas are not frequented by large fish, but they would likely include them in the ecosystem if they are studying benthic littoral species. Thus, how one measures stochasticity should also depend on which organisms are of most interest.

Since species in a heterogeneous landscape can modify their behavior and move to more favorable locations, spatial variation will favor movement and behavioral adaptations faster than temporal variation will [@bell_spatial_1993].

Transition regions are often have the highest heterogeneity, but, similarly to temporal stochasticity, measures and perceptions of spatial heterogeneity depend on the scale at which they are being considered. The repetitive and cyclical motion of waves may not concern most vertebrates that live along the coast, but it is much more turbulent and chaotic for smaller organisms, such as crustaceans or diatoms. Therefore, the scale at which we consider spatial stochasticity should depend on the size of the organisms of interest as well as the size of their habitat or home range.

Although @bell_spatial_1993 couldn't find a scale at which variance stopped increasing, it seems reasonable to assume that variance cannot increase infinitely. The variance of a finite system (e.g., a lake or a well-defined habitat), should reach a finite maximum somewhere between our smallest scale of interest and the scale of the entire system, even if the maximum is reached when the entire system is considered. This does not imply that a somewhat larger system that contains the first one (e.g. the lake and its coasts) will have the same variance. Variance between lake and coast is high because they are different types of habitats, and the area where they meet (the coast) is potentially even more variable because the mixture of the two creates a third habitat for littoral benthic animals which is subject to high stochasticity due to waves and changing water levels.

The same idea would likely also apply for larger-scale systems, such as continents or a planet. And while stochasticity of a system may also vary over time, it is not reasonable to expect it to increase monotonically through time because that would imply it has been increasing monotonically until the current moment. Still, stochasticity does vary over time, and its estimation depends on the data available, so it may be useful to view it as a random variable with an estimated, conditional posterior distribution through a Bayesian framework [@bolstad_introduction_2017].

<!--
## Frontiers  Hierarchical, Memory-Based Movement Models for Translocated Elk (Cervus canadensis)  Ecology and Evolution

Refs to check:

* humans, nonhuman primates and other large-brained vertebrates make movement decisions based on spatial representations of their environments (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B50)
* 2010). These representations may allow animals to move directly to important sites in their environment that lie outside of their perceptual range (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B40, https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B35)
* Attribute memory, i.e., non-spatial memory, e.g., food type and abundance (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B13)
* Bison remember valuable information about the location and quality of meadows (spatial and attribute memory) and use this information to revisit profitable locations (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B29)
* Roe deer an spatial memory for foraging in absence of territorility (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B41)
* memory in capuchin monkeys (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B8)
* memory in bison (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B18)
* memory in sheep (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B18)
* memory in woodland caribou (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B1)
* memory in chimpanzees (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B20)
* memory constitutes an important mechanism for home range emergence (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B6, Börger et al., 2008; Van Moorter et al., 2009)
-->

<!--
## Urban landscapes and habitat structure

Human activity often fundamentally alters the structure of environments by spatially homogenizing them and imposing temporally heterogeneous and potentially stochastic cycles (e.g., mining, clear-cutting, road traffic, and weekly work schedules). The perception of spatial heterogeneity and cycles in anthropogenic areas varies between species: scavenger animals such as ravens (*Corvus corax*) and raccoons (*Procyon lotor*) often learn to recognize areas that can provide food, such as restaurants or large disposal bins, but other animals may not be able to distinguish between different urban areas. Certain animals have learned to depend on and take advantage of human activity, such as crows who place nuts strategically so that cars will crack them open for them to eat [@nihei_when_2002].
-->

\newpage

### 2022-08-25 writing/manuscript.Rmd
wrote conclusion and added hypothesis figures

@@
 ---
-title: "How do resource abundance and stochasticity affect animals' spatial needs?"
-subtitle: "A review with quantitative simulations and empirical modeling"
-author: "Mezzini S. and Noonan M. J."
-date: "`r Sys.Date()`"
+title: \large How resource abundance and stochasticity affect animals' spatial needs
+author:
+  - name: Stefano Mezzini
+    email: stefano.mezzini@ubc.ca
+    institute: ubco
+  - name: Emilia Patrícia Medici
+    institute: [tapir, iucn, escas]
+  - name: Michael J. Noonan
+    email: michael.noonan@ubc.ca
+    institute: ubco
+institute:
+  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
 bibliography: 'env-var-review.bib'
@@
 header-includes:
-    - \usepackage{setspace}\doublespacing # double-space text
-    - \usepackage{indentfirst} # indent first line of each paragraph
-    - \usepackage[small]{titlesec} # smaller font for headings
-    - \usepackage{wrapfig} # to wrap text around figures
-    - \usepackage{caption}
+    - \usepackage{setspace}\doublespacing # for double-spaced text
+    - \usepackage{indentfirst} # for indenting first line of each paragraph
+    - \usepackage[small]{titlesec} # for smaller font for headings
+    - \usepackage{wrapfig} # for wrapping text around figures
+    - \usepackage{caption} # for more customization of captions
     # smaller caption text, less space above and below caption
-    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-20pt}
-
-    # - \usepackage{gensymb}
-    # - \usepackage{float}
-    # - \usepackage{afterpage}
-    # - \usepackage{ragged2e}
-subparagraph: yes # needed for \usepackage[small]{titlesec}
-# classoption:
-# - twocolumn
+    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
+    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
+    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
+    # - \usepackage{float} # for better floating objects interface; adds more floats 
+    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
+subparagraph: true # needed for \usepackage[small]{titlesec}
+linenumbers: true
 output:
   bookdown::pdf_document2:
-    number_sections: true
+    pandoc_args:
+      - '--lua-filter=scholarly-metadata.lua'
+      - '--lua-filter=author-info-blocks.lua'
+    keep_tex: true
+    number_sections: false
     toc: false
 ---
+
+<!--
+TO DO:
+- write conclusion paragraph
+- write discussion (2-3 pages)
+    - add headers for sections
+- trim down and edit intro and other previous sections
+- move empirical examples within each relevant section (i.e., abundance examples for abundance section)
+- simple figure for demonstrating idea of nonlinear relationship between R and H, with animal moving in empty space?
+    - explain the idea in the caption, straight line can't work because can't have negative space, and there's need for saturation
+    - line increasing as R decreases, 3 options: death, more increase (nomadism), sub-optimal saturation
+- make same figure for V(R), but focus on temporal variance in time:
+    - use grey or low alpha for curves with high uncertainty
+- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
+- drop figure 1? (Fictitious example of variation)
+- move 5x5 figure to interactive effects section
+- add median NDVI behind tracking data of tapir, buffalo, and ??? figure
+-->
+
+\newpage <!-- start main body on a new page -->
 
@@
 ## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width
+
+#' place figures before the paragraph they are referred to so `wrapfigure` places them
+#' beside the paragraph
 ```
 
-<!-- place figures before the paragraph they are referred to so wrapfigure places them beside the paragraph -->
-
-<!-- add section explaining what the paper details and what the objectives are? -->
-
+<!--
 @ciuti and @berger-tal: human presence in nature for other reasons than hunting
@@
 habitat productivity hypothesis: Harestad and Bunnel, 1979
+-->
 
-Add @mueller_how_2011 for a study that accounted for variance in NDVI for animal movement
+# Abstract
 
-The habitat productivity hypothesis proposed by @harestad_home_1979 is supported, but it ignores many other important causes of movement [@seigle-ferrand_systematic_2021].
+Animal's spatial needs are often thought to be tighly linked to the amount of resources ($R$) within their habitats.
+Although this hypothesis has widespread support, most studies only account for the mean $R$, while the variance in $R$ is rarely accounted for.
+Failing to account for such variation is likely to result in an underestimation on animals' spatial needs, which may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict.
+In this paper, we provide a review of animal movement studies that estimate animals' spatial needs as a function of $R$, and we demonstrate the importance of accounting for variance in $R$.
+Additionally, we provide a unifying framework and hypotheses for the effect of mean and variance in $R$ on animals' home range sizes and movement behaviors.
+In the first part of this paper, we use simulated movement data in a variety of example habitats to demonstrate the importance of accounting for the variance in $R$.
+Next, we use the Normalized Difference Vegetation Index to estimate mean and variance in $R$ within the habitats of a tapir from the Pantanal (Brazil) and an African buffalo from South Africa to demostrate how this framework can be applied to real-world tracking data.
+Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to real-world tracking data.
 
-The ability to move allows animals to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977].
-Animals may move for a variety of reasons, including searching for resources (e.g., food, nutrients, water, heat, a new den or nesting spot), a mate, or a new group.
-Movement also allows animals to escape predation or dangerous competition as well as defend resources and territories [@grant_whether_1993].
+\newpage
+
+# Introduction
+
+The ability to move allows motile organisms to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977].
+Animals may move for a variety of reasons, including searching for resources [@refs] or mates [@refs] while avoiding predation [@refs] or dangerous competition [@refs], as well as defending resources and territories [@grant_whether_1993; @refs].
 @nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
@@
 This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it.
-We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI; see @pettorelli_normalized_2011].
+We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI, see @pettorelli_normalized_2011].
 
 # Effects of resource abundance on animals' spatial needs
+
+Figure 2 of @bista_effects_2022 for example of study that assumes linear relationships (w LM) for home range, when there can't be one.
+
+```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of mean resource abundance ($\\mathbb E(R)$).
+At low values of $\\mathbb E(R)$, range-resident animals (bottom dashed line) may exhibit nomadic behavior, such that While $H$ decreases as $\\mathbb E(R)$ increases, it is not possible for $H$ to continuously decrease linearly because animals will require a minimum (nonnegative) amount of space to survive.
+At low values of $\\mathbb E(R)$, $H$", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
+```
+
+Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [figure \@ref(fig:), @relyea_home_2000].
+While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance.
+The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005].
+Animals living in large groups will require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004].
+
+The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk.
+Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$).
+Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$).
+Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.].
+Let $R$ indicate the amount of *resources* in a particular patch or area.
+For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity.
+However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.].
+Next, let $H$ indicate the size of an animal's home range (HR)^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$.].
 
@@
 ```
-
-The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk.
-Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$).
-Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$).
-Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.].
-Let $R$ indicate the amount of *resources* in a particular patch or area.
-For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity.
-However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.].
 
@@
 
-The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where the average resource abundance remains constant or changes over time.
-Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$, constant $\mathbb V(R)$), and how the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range.
-In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$, constant $\mathbb V(R)$ ), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$.
+The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where $\mathbb E(R)$ remains constant or changes over time but $\mathbb V(R)$ remains constant.
+(Although some noise in $H$ are still present since $\mathbb V(R) \ne 0$ and thus the $R$ is not constant).
+Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), with larger changes when $\mathbb E(R)$ is low, since the multiplicative changes in $R$ are larger.
+Additionally, the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range $H_{50\%}$.
+In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$.
 However, estimates of spatial requirements based on estimated changes in $\mathbb E(R)$ should be interpreted carefully, since model error and unpredictable changes in $\mathbb E(R)$ may increase animals' spatial needs unpredictably.
 Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$, constant $\mathbb V(R)$), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible.
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible.
 In cases where $\mathbb E(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often.
-Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$, constant $\mathbb V(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
+Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
+
+```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
+The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile.
+Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase approximately linearly as $\\mathbb V(R)$ increases.
+Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
+The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
+knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+```
 
@@
 
-Although the effect of resource abundance on animals' spatial needs is often recognized and accounted for, the variance in resource abundance is generally ignored or assumed to remain constant over time.
-However, resource abundance often varies due to multiple causes.
-Some causes are repetitive, predictable, or even well-known patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches), while others are infrequent, unpredictable, or poorly-understood events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
-In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity.
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
+(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
+(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$.
+(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
+```
+
+Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], the variance in $R$ has received far less attention.
+However, $R$ can vary strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
+Environmental stochasticity, including extreme events, can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
+A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
+There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+
+@seigle-ferrand_systematic_2021: Landscape heterogeneity influenced the sizes of home ranges in about 96% of studies based on individuals monitored by VHF but in only about 75% for individuals monitored by GPS, whatever the number of studied metrics.
+
+When animals are not guaranteed that the resources they find during one visit will be there the next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness.
+Since $\mathbb V(R)$ is not constant over time, we can now define it as a function of time using the notation $\mathbb V(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
+In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
+
+In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity in $R$.
 We refer to predictable or expected spatiotemporal changes as environmental **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation.
 For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable.
-In contrast, the quantity and quality of berries produced during a given year may be stochastic.
-Thus, $\mathbb V(R)$ is affected by both sources of variation, and changes in $\mathbb V(R)$ may predictable or not.
+In contrast, whether it is a good or bad year for berries is stochastic.
+<!--Thus, changes in $\mathbb V(R)$ may predictable or not, and $\mathbb V(R)$ is affected by both sources of variation.-->
+
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\mathbb E(H)$) illustrates how an animal's home range may vary as $\mathbb V(R)$ changes over time while $\mathbb E(R)$ remains constant.
+Both $\mathbb E(H)$ and $\mathbb V(H)$ increase with $\mathbb V(R)$, although $H_{95\%}$ is more sensitive to changes in $\mathbb V(R)$ than $H_{50\%}$.
+Additionally, the effect of $\mathbb V(R)$ on $H$ depends on $\mathbb E(R)$, since the effect of $\mathbb V(R)$ on $H$ is stronger when $\mathbb E(R)$ is low (see the bottom row with erratic $\mathbb V(R)$).
+
+Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\mathbb E(R)$ and $\mathbb V(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
+$\mathbb E(R)$ and $\mathbb V(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\mathbb E(R)$ and $\mathbb V(R)$ remain undetected.
+Although it is impossible for $\mathbb E(R)$ and $\mathbb V(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease).
+Additionally, these examples are important because they demonstrate the relationships between $H$, $\mathbb E(R)$, and $\mathbb V(R)$ in (relatively) simple scenarios.
+Cyclical oscillations in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\mathbb E(R)$ and $\mathbb V(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
+Finally, erratic changes in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
+However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\mathbb E(R)$ or increases in $\mathbb V(R)$ rather a series of small, sudden, changes.
+Estimating the true trend in $\mathbb E(R)$ and $\mathbb V(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\mathbb E(R)$ are not due to $\mathbb V(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\mathbb E(R)$ and $\mathbb V(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.
+<!-- https://doi.org/10.1080/00207549508930162 ?-->
 
 The effects of $\mathbb V(R)$ on animals' spatial requirements likely depend on the animals' needs.
-@lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous) required exponentially less space than foxes that lived in more homogeneous regions.
-@nilsen_can_2005 found that the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality, while the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased.
 @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
-***ADD MORE EMPIRICAL EXAMPLES?*** However, the effects of environmental variance on animals' spatial needs remains largely undestudied and unaccounted for.
+Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic.
+In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions.
+Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality.
+The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.
 
-# Empirical examples
+What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability.
+Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004], which allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high fertility of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can allow them to develop traits that increase the chances of survival following an extreme event.
+Additionally, the effects of size and lifespan on sensitivity are likely nonlinear and correlated, since smaller animals tend to have shorter lives (and vice-versa), and small animals also tend to have lower metabolic rates, which often limit their movement speed, home ranges, and how the animals interact with their ecosystem [@brown_toward_2004].
+Thus, processes will have stronger impacts on smaller animals than on bigger ones: The grazing pattern of a bison (*Bison bison*) drastically alters the habitat of most crawling insects, but other grazing mammals would not pay attention to changes in grass length at the same spatial or temporal resolution.
+Similarly, the timing and quantity of yearly snowfall would be a somewhat predictable and expected event for most adult moose (*Alces alces*), but it may be a shock for many adult zooplankton or a new-born wolf (*Canis lupus*).
+What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal.
+Therefore, the mean and variance in $R$ should be modeled while accounting for the behavior of the species of interest.
 
-```{r tapir-example, fig.cap="Seven-day home range size of a tapir from Medici $et~al.$ (2022) as mean and variance in resource abundance change over the tracking period.
-(a.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir during the period of reference.
-(b.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir during the period of reference.
-(c.) Estimated home range size of the tapir during each reference period, based on 50% (bold) and 95% (thin) utilization distributions.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/tapir-example.png')
-```
+An unresponsive animal may perceive frequent changes in $R$ as high $\mathbb V(R)$, while a responsive animal may perceive them as changes in $\mathbb E(R)$.
+Thus, $R$ should be modeled using very flexible and wiggly models when animals are highly responsive, while less flexible models are more appropriate for more unresponsive or sedentary species.
+Generalized Additive Models [GAMs, see @wood_generalized_2017] are sufficiently flexible and adaptable to allow one to model $R$ while reflecting various levels of adaptability and responsiveness.
+(See @simpson_modelling_2018 for an overview of how to select appropriate bases sizes and smoothness parameters for GAMs.) The animal's movement (e.g., degree of range residency, range crossing time, average speed, directional persistence) should also inform the complexity of any spatial smooths included in the models.
+Models with complex spatial terms are appropriate for animals that recognize the location of high-quality patches, while smoother terms are best for animals that may perceive differences in $R$ as spatial variance, instead.
 
-Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [@lucherini_habitat_1996; @relyea_home_2000].
-While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance.
-The size of a home range, is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [@nilsen_can_2005], since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004].
+**it may be worth chatting about the NDVI models for the tapir and the buffalo later this month**
+
+# Discussion
 
@@
 In particular, resource abundance and environmental productivity are often hypothesized to be inversely proportional to HR size (or some function of it), such that regions with higher abundance correspond to smaller spatial needs, since animals do not have to range over extensive areas to meet the energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @southwood_habitat_1977], and maintaining a thermoregulatory balance.
-The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear, since larger home ranges can result in higher rates of competition and are harder to defend [@grant_whether_1993; @jetz_scaling_2004].
 @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
-However, the effect of many other factors on HR size is likely depends on the species, location, and possibly also on the individual animal.
+However, the effect of many other factors on HR size likely depends on the species, location, and possibly also on the individual animal.
 For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; capuchin monkeys, @torrez-herrera_monkeys_2020].
@@
 
+* failing to account for $\mathbb V(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
+* find examples of extreme events that caused population collapses or mass die-offs
+* increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness
+* animal mortality following fires, how little we know of mortality following fires: @jolly_animal_2022
+
+## Moving window examples
+
+```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$, a.-d.) and an African buffalo ($Syncerus~caffer$, e.-h.) as mean and variance in resource abundance change over the tracking period.
+(a., e.) Tracking data of the two animals.
+(b., f.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir during the period of reference.
+(c., g.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir during the period of reference.
+(d., h.) Estimated home range size of the tapir during each reference period, based on 50\\% (bold) and 95\\% (thin) utilization distributions.
+The tapir movement data corresponds to the individual named \"Anna\" from the Pantal sample of Medici $et~al.$ (2022), while the buffalo data corresponds to the buffalo named \"Cilla\" in the data of Getz $et~al.$ (2007) available on through the $\\tt{Movebank}$ package (Cross $et~al.$ 2016).
+The buffalo tracking data are also available in the $\\tt{buffalo}$ dataset in the $\\tt{ctmm}$ package.", out.width = '100%'}
+knitr::include_graphics('../figures/moving-window-examples.png')
+```
+
+\newpage
+
+<!--
+in stochastic environments:
+- higher spatial needs and possibly store resources for periods of scarcity, which are expected to get longer with climate change
+- heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one
+-->
+
 <!--
@@
 
-protective/defensive adaptations peaks at intermediate levels of competitor density/spatial resource density/spatiotemporal clumping.
-At low competition, there's no need for defense and the cost (e.g., energetic) and risk (e.g., wounds) outweigh the benefit, while at high values the cost and risk become too high and they outweigh the benefit [@grant_whether_1993].
-
-See *The effects of phenological mismatches on demography*.
-
 @rickbeil_plasticity_2019:
@@
 Maybe elk migration is triggered when they have sufficient energy (if the weather is still not too harsh).
-- Elk left early when exposed to hunting more
+- Elk left early when exposed to more hunting
 -->
-
-# Simulating spatial needs as a function of $\mathbb E(R)$ and $\mathbb V(R)$
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$.
-(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
 
@@
 
-In areas where animals are not guaranteed that the resources they find during one visit will be there the next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness.
-Patches with low or high $p$ will be most predictable, since successes can be expected to be very rare (if $p \approx 0$) or very common (if $p \approx 1$).
-In contrast, patches will be most stochastic when the probabilities of success and failure are approximately the same (i.e., $p \approx 1 - p \implies p \approx 0.5$, see figure \@ref(fig:multipanel-changing-var)b).
-In stochastic habitats, $\mathbb E(U)$ will depend on $\mathbb E(S)$ as well as $\mathbb E(R)$, since $S$ is no longer constant (figure \@ref(fig:multipanel-changing-var)c).
-Now, expected usable resources become
+# Interactive effects of mean and variance in resource abundance
 
-\begin{equation} \label{mean-u-definition}
-\mathbb E(U) = \mathbb E(R \, S),
-\end{equation}
+Although the effects of environmental variance on animals' spatial needs remain largely undestudied and unaccounted for, available evidence suggests that the effect of $\mathbb V(R)$ depends on the species and landscapes studied, so it would be inappropriate to develop a single hypothesis for how $\mathbb V(R)$ affects animals' spatial needs.
 
-or
+[*Needs for understanding effects of variance*] are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
 
-\begin{equation} \label{mean-u-definition-indep}
-\mathbb E(U) = \mathbb E(R) \, \mathbb E(S)
-\end{equation}
+* variance has less of an effect when variance is high
 
-if $R$ and $S$ are independent.
-This model can be applied to all mobile animals, including herbivores, carnivores, and omnivores.
-In the case of herbivores, $p$ may indicate the chance of finding good forage, which may be absent if regeneration times are long, if competitors have already exhausted the resource, or following a fire.
-For carnivores, $p$ may indicate the chance of feeding on prey, which may depend on encountering and killing some first [@suraci_beyond_2022].
-However, since $R$ is not limited to energetic resources, $p$ may also indicate the chance of finding water or also a suitable location for a den or nest.
-
-While resource availability is often considered in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], an environment's heterogeneity (i.e., diversity), stochasticity (i.e., unpredictability), and how the two change over time are rarely accounted for.
-In addition, environmental stochasticity, including extreme events, can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness.
-Therefore, I expect animals living in unpredictable environments to require more space than those in stable environments.
-Although this hypothesis is supported by a few recent studies [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], many of them are limited in their analytical depth and geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
-These stresses are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
-As the impacts of habitat loss and climate change will worsen in the future [@ipcc_global_2018; @hansen_high-resolution_2013], it is imperative that we better understand spatial requirements of taxa to protect wildlife existence and biodiversity.
-Environmental safeguarding is also essential for Reconciliation with Indigenous People in Canada [@truth_and_reconciliation_commission_of_canada_honouring_2015].
-
-# Interaction effects of resource availability and stochasticity
-
-@nilsen_can_2005
-
-In the previous section, we mentioned that a model which does not account for environmental variance may be acceptable when $\sqrt{\mathbb V(R)} / \mathbb E(R)$ is low, as in the case of a habitat with high $R$ and $p \approx 1$.
-However, since maximum resource abundance is constrained by environmental stochasticity [@chevin_adaptation_2010] and resources can be depleted or rendered inaccessible by other individuals [@grant_whether_1993; @jetz_scaling_2004], $\mathbb E(R)$ does not provide a sufficiently complete picture of resource availability in an environment.
-Thus, the effect of $\mathbb V(R)$ should be included when possible, even when $p \approx 1$ or $p \approx 0$.
-<!--Spatiotemporally constant and homogeneous regions will allow animals to have a small and dependable home range, while stochastic regions will force animals to have higher spatial needs and possibly store resources for periods of scarcity.
-Similarly, heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one.
-Conversely, if some patches are sufficient and highly valuable, animals may adopt strong defensive behaviors towards the patch [@grant_whether_1993].
-Thus, some measure of $\mathbb V(R)$ is needed to understand how stochasticity may affect spatial requirements.-->
-
-Let $H$ indicate the size of an animal's HR.^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$.
-Statistically, we can imagine $H$ as having a probability distribution with support over the interval from zero (not included) to infinity (also not included), which we can indicate with the notation $H \in (0, \infty)$.] As explained above, $H$ will be higher in regions with lower $\mathbb E(U)$ or higher $\mathbb V(U)$.
-Figure \@ref(fig:5-5-sims) presents various scenarios with different trends in $\mathbb E(U)$ and $\mathbb V(U)$.
-Although some of these scenarios may seem overly simplistic, they can help us understand the complex interaction effects of $\mathbb E(U)$ and $\mathbb V(U)$.
-In regions where $U$ has constant mean and variance, $H$ remains approximately constant, although small oscillations are possible, since $\mathbb V(U) \ne 0$ and thus the environment is still stochastic.
-When $\mathbb E(U)$ changes over time but $\mathbb V(U)$ remains constant, it is easy to see that $H$ decreases as $\mathbb E(U)$ increases, with changes in $\mathbb E(U)$ having larger effects when $\mathbb E(U)$ is low and smaller effects when $\mathbb E(U)$ is already large.
-This is best visualized in the scenario in which $\mathbb E(U)$ is increasing linearly while $\mathbb V(U)$ is constant, since decrease in core HR ($H_{50\%}$) and 95% HR ($H_{95\%}$) is decreases over time.
-Additionally, as $\mathbb E(U)$ increases, the difference between $H_{50\%}$ and $H_{95\%}$ also decreases.
-Thus, $H_{95\%}$ is more sensitive to changes in $\mathbb E(U)$ than $H_{50\%}$.
-$H_{95\%}$ is also more sensitive to changes in $\mathbb V(U)$, since changes in $\mathbb V(U)$ cause greater oscillations in $H_{95\%}$ than in $H_{50\%}$.
-
-```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
-The thicker line indicates the animal's core home range (0.5 quantile), while the thin line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase linearly as $\\mathbb V(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks each day starting from the point (0, 0) until they reach satiety, at which point they returned to (0, 0) over the same amount of time.
-The animal's spatial variance parameter was then calculated using an OUF model via the $ctmm$ package", out.width='100%'}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
-```
-
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\mathbb E(U)$ and $\mathbb V(U)$, and their impacts on animal space use, are useful examples that can be thought of as simplified scenarios.
-$\mathbb E(U)$ and $\mathbb V(U)$ may be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently available that changes in $\mathbb E(U)$ and $\mathbb V(U)$ remain undetected.
-Although it is impossible for $\mathbb E(U)$ and $\mathbb V(U)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease).
-Additionally, these examples are important because they demonstrate the relationships between $H$, $\mathbb E(U)$, and $\mathbb V(U)$ in a (relatively) simple scenario.
-Cyclical oscillations in $\mathbb E(U)$ and $\mathbb V(U)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\mathbb E(U)$ and $\mathbb V(U)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
-Finally, erratic changes in $\mathbb E(U)$ and $\mathbb V(U)$ may occur in environments where changes are very sudden, such as fire-prone or floor-prone areas, or habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
-However, if highly stochastic or erratic changes occur frequently, animals are most likely to perceive them as a smooth transition rather a series of small, sudden, changes.
-Estimating the true trend would often require an excessively high cognitive capacity and an equally unlikely abundance of information.
-Additionally, although changes in $\mathbb E(U)$ are not due to $\mathbb V(U)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020].
-<!-- https://doi.org/10.1080/00207549508930162 ?-->
-
-# Empirical examples
-
-## Sliding-window sampling
+\newpage
 
@@
 
-```{r tapir-hr, fig.cap="Movement of a tapir during a monitoring period of approximately 3 years (1997-07-10 to 2000-06-08) used in the work by Medici $et~al.$ (2022).
-Different subsamples may result in different estimates of home range size and centroid (a, b; each span 100 days), so neither subsample is necessarily representative of the space the animal used over the entire monitoring period (c).", out.width = '0.4\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.4)}
-knitr::include_graphics('../figures/hr-and-lifespan.png')
-```
-
-In the case of animals whose spatial needs change over the their lifetimes (figure \@ref(fig:tapir-hr)), one may define HR as a function of time (with distinct HRs for each period, e.g., salmon spawning) or as a continuous function of age (so the HR changes smoothly).
-In either case, the accuracy of the estimate parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
-Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
-
-spatial needs can change as requirements from the animal change or as $R$ changes.
-
 <!-- see https://doi.org/10.1186/s40462-019-0187-z regarding HR, energetics, and movement along paths -->
-
-* tapirs
-* tapir w sliding window
 
@@
 # Movement types
+* Range residency
+* Seasonal migration
+* Nomadism
 
-## Range residency
-
-## Seasonal migration
-
-## Nomadism
-
-predation is more stochastic than herbibory; predators need to match more cycles, depend on less individuals -> more variable food
+predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move -> higher effect of V(R)
 
@@
 
-<!--
+# Conclusion
 
-# Simulations
+The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance.
+We provide realistic and flexible hypotheses of the effects of $\mathbb E(R)$ and $\mathbb V(R)$ on animals' spatial needs and movement behavior.
+We demonstrate that animals' spatial needs respond nonlinearly to both $\mathbb E(R)$ and $\mathbb V(R)$, and we demonstrate the importance of accounting for $\mathbb V(R)$ besides $\mathbb E(R)$.
+We also provide a framework for modeling animal behavior and movement continuously rather than imposing arbitrary thresholds between range-resident and nomadic animals.
+(notes on seasons and "seasonal behavior")
 
-Include:
+\newpage
 
-* effects of animal mass,
-* resistance to/ease of movement (e.g., wolves and clear-cut gas lines)
-* "state-space model" with 2 spaces: within a patch (slow-moving) and between patches (faster movement)
+# MISC
 
-See `R` packages for meta-community simulations
-
--->
-
-# Future work
-
-The work presented here provides a general framework for viewing animal movement as a function of mean and variance in resource abundance.
-The models used here do not account for animals' traits or behaviors (e.g., energetic needs, food preferences, ability to move, competition) as well as any environmental barriers to movement, such as lakes, rivers, and anthropogenic environments.
-Future models can account for more stuff, but modeling movement often already accounts for most of this.
-
-How does this impact the field, future papers?
-
-\newpage <!-- place references on a separate page -->
+In the case of animals whose spatial needs change over the their lifetimes (figure \@ref(fig:tapir-hr)), one may define HR as a function of time (with distinct HRs for each period, e.g., salmon spawning) or as a continuous function of age (so the HR changes smoothly).
+In either case, the accuracy of the estimate parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
+Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
 
@@
 
+```{r tapir-hr, fig.cap="Movement of a tapir during a monitoring period of approximately 3 years (1997-07-10 to 2000-06-08) used in the work by Medici $et~al.$ (2022).
+Different subsamples may result in different estimates of home range size and centroid (a, b; each span 100 days), so neither subsample is necessarily representative of the space the animal used over the entire monitoring period (c).", out.width = '0.4\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.4)}
+knitr::include_graphics('../figures/hr-and-lifespan.png')
+```
+
 Adverse events which are too infrequent (e.g., $p \lesssim 0.1$) may not be perceived as a reoccurring threat, so organisms may be unable to adapt appropriately.
@@
 
-```{r stochasticity-timescale-figure, fig.cap="Fictitious changes in resource abundance mean (a) and variance (b) over the course of a year.
-The true trends are indicated by the black line, while the colored lines indicate the estimates perceived based on different levels of adaptability.
-The dashed vertical line indicates the hypothetical current date, such that any data to the right of the line is unknown to the animals, so estimates on the right of the dashed line are extrapolations.
-The data were simulated using Gaussian noise to ensure mean-variance independence and modeled using Generalized Additive Models with an identity link function to allow linear responses.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/stochasticity-timescale.png')
-```
-
-What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability.
+<!-- What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability.
 Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977].
@@
 What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal.
-<!-- add empirical examples -->
+**add empirical examples** -->
 
@@
 
-\newpage
+\newpage <!-- place references on a separate page -->
+
+# References

### 2022-10-04 writing/manuscript.Rmd
mass push to transfer to laptop

@@
       - '--lua-filter=author-info-blocks.lua'
-    keep_tex: true
+    toc: false
     number_sections: false
-    toc: false
 ---
@@
   - add headers for sections
-- trim down and edit intro and other previous sections
-- move empirical examples within each relevant section (i.e., abundance examples for abundance section)
-- simple figure for demonstrating idea of nonlinear relationship between R and H, with animal moving in empty space?
-    - explain the idea in the caption, straight line can't work because can't have negative space, and there's need for saturation
-    - line increasing as R decreases, 3 options: death, more increase (nomadism), sub-optimal saturation
-- make same figure for V(R), but focus on temporal variance in time:
-    - use grey or low alpha for curves with high uncertainty
 - focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
-- drop figure 1? (Fictitious example of variation)
-- move 5x5 figure to interactive effects section
-- add median NDVI behind tracking data of tapir, buffalo, and ??? figure
 -->
@@
 \newpage <!-- start main body on a new page -->
+
+\tableofcontents
+
+\newpage
 
@@
 ```{r wrap-figure-function, include = FALSE}
+# this chunk is needed to wrap text around or beside figures
 # code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures
-
-# this chunk is needed to wrap text around or beside figures
 
@@
 "When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
-habitat productivity hypothesis: Harestad and Bunnel, 1979
+habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
 -->
@@
 
-Animal's spatial needs are often thought to be tighly linked to the amount of resources ($R$) within their habitats.
-Although this hypothesis has widespread support, most studies only account for the mean $R$, while the variance in $R$ is rarely accounted for.
-Failing to account for such variation is likely to result in an underestimation on animals' spatial needs, which may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict.
-In this paper, we provide a review of animal movement studies that estimate animals' spatial needs as a function of $R$, and we demonstrate the importance of accounting for variance in $R$.
-Additionally, we provide a unifying framework and hypotheses for the effect of mean and variance in $R$ on animals' home range sizes and movement behaviors.
-In the first part of this paper, we use simulated movement data in a variety of example habitats to demonstrate the importance of accounting for the variance in $R$.
-Next, we use the Normalized Difference Vegetation Index to estimate mean and variance in $R$ within the habitats of a tapir from the Pantanal (Brazil) and an African buffalo from South Africa to demostrate how this framework can be applied to real-world tracking data.
-Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to real-world tracking data.
+\noindent Since animals need to maintain a positive energetic balance to grow and reproduce, their spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
+Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance in around the mean are still largely unknown.
+This is not a trivial oversight.
+Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs.
+In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
+We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resource abundance on animals' space use.
+Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use.
+Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
 
@@
 
-The ability to move allows motile organisms to respond rapidly and continuously to heterogeneous and changing environments and conditions [@southwood_habitat_1977].
-Animals may move for a variety of reasons, including searching for resources [@refs] or mates [@refs] while avoiding predation [@refs] or dangerous competition [@refs], as well as defending resources and territories [@grant_whether_1993; @refs].
-@nathan_movement_2008 conceptualized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
+<!-- You need a lot more depth here.
+Open with a paragraph on movement generally that leads to the idea that responses to resource are something worth focusing on.
+Follow that up with a paragraph on the history of exploring movement/space use and resources, leading to the concept that, to date, the relationship between abundance and stochasticity have not been fully explored.
+Then finish with a paragraph detailing the aims of the current paper.
+-->
+
+\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce.
+Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018].
+Thus, many organisms will change their behaviours in response to changes in local resource abundance.
+While some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
+However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+
+While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century.
+@burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range.
+Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce.
+Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance.
+More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022].
+Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with larger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005].
+Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago various scientists, inlcuding @lindstedt_home_1986 and @reiss_scaling_1988.
+Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
+
+In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
+First, we set the theoretical background necessary for the successive chapters.
+Concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology.
+Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis of how animals respond to changes in resource abundance.
+Afterwards, we present a hypothesis for how animals respond to resource unpredictability.
+Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability.
+Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado.
+Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.
+
+\newpage
+
+# Concepts and definitions
+
+## Resources as a random variable
+
+In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$).
+Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
+Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify).
+Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.
+
+### Probability distributions
+
+In probability and statistics, random variables are defined by specifying the distribution the variable follows.
+Since the variable is random, it can take multiple values (minimum 2), each with different probabilities.
+Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.
+The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.
+The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.
+
+There are many distributions we can assign to $R$, depending on how we quantify it.
+If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $r \in (0, \infty)$.
+(We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because both 0 and infinity are not included in the support set.) In this case, we can let $R$ follow a Gamma distribution.
+If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution for $R$.
+Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
+The next two sections introduce two parameters that are often used to visualize and define distributions: the expectation and the variance of a random variable.
+
+### Expected resource abundance, $\mathbb E(R)$
+
+The expectation of a random variable (i.e.
+its mean or average, indicated here as ) is the value one can *expect* the random variable to take.
+$\mathbb E(R) = \mu$ to indicate the expectation of the random variable $R$.
+The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\mathbb E(R)$ is still useful for estimating the average long-term resource abundance.
+When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\mathbb E(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\mathbb E(R) = \mu(t)$.
+
+The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022].
+This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
+Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero).
+The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.
+
+### Variance in resource abundance, $\mathbb V(R)$
+
+A random variable's variance is a measure of its unpredictability or variability.
+We use  $\mathbb V(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$.
+For instance, while one may consider $R$ to vary within years, we define $\mathbb V(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others.
+Additionally, we distinguish between two different sources of variation.
+We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation.
+For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley.
+In contrast, whether next year will be a good or bad year for berries is stochastic.
+
+### Simulating resource abundance
+
+In this paper, we can simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
+Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples.
+However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean.
+As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.
+
+### Estimating resource abundance using NDVI
+
+In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011].
+However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation
+
+\begin{equation} \label{ndvi-to-r}
+R = \frac{\nu + 1}{2},
+\end{equation}
+
+where $\nu$ indicates the NDVI value.
+Note that only **linear transformations** (i.e.
+only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906].
+In this case, we can define $R$ as following a Beta distribution using the notation $B (\mu(t), \sigma^2(t))$.
+Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.
+
+## Animal movement as a stochastic process
+
+### Drivers of animal movement
+
+@nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
 
@@
 
-such that the animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$).
+such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$).
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters, doing so with sufficient frequency and accuracy would not be simple.
-Rather than measuring changes in the various resources and needs directly, we can quantify animals' spatial needs as a response to average resource abundance and the variability around it, instead.
-This paper will illustrate how animals' spatial needs depend on average resource abundance and the variance around it.
-We present this idea through a quantitative theoretical framework and empirical examples where resource abundance is estimated using the Normalized Vegetation Index [NDVI, see @pettorelli_normalized_2011].
+The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels, @refs), doing so with sufficient frequency and accuracy would not be simple.
+However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly) <!-- since accounting for a proxy of resource abundance will also account for the effects of many other drivers that depend strongly on resource abundance (e.g., competition, predation)-->.
+
+### Visualizing movement as a stochastic process
+
+To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day.
+If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another.
+However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
+Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity.
+$\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
+Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters.
+Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].
+
+If sampling frequency is too low to provide precise measures of autocorrelation, it will be inappropriate to estimate an animal's average speed, directional persistence, and range crossing time, but the data can still provide reasonable estimates of the animal's spatial needs and usage, and model selection procedures with criteria such as AIC can help the analyst choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
+Estimates of average or instantaneous speed require the highest sampling frequency [$\Delta t \lessapprox \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
+However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior, it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the sampling duration is sufficiently long.
+
+### Quantifying spatial needs
+
+The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
+Thus, here we quantify animal's spatial needs using their positional variance (indicated as $\mathbb V(\text{position})$), instead.
+***explain the concept of positional variance*** [@refs].
+
+Although range-resident animals will generally have smaller $\mathbb V(\text{position})$ than migratory and nomadic animals, $\mathbb V(\text{position})$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\mathbb V(\text{position})$ determines the scale and size of an animal's spatial needs but not its behavior.
+Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement.
+Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
+In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.
+
+### Advantages of modeling animal movement as a stochastic process
+
+Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
+
+Depending on sampling frequency, statistical methods that rely on straight-line displacement are prone to under-estimating the distance traveled by an animal and its speed at coarse sampling frequencies, while they over-estimate them at fine frequencies [@noonan_scale-insensitive_2019].
+In addition, most commonly-used methods for home range estimation assume tracking data to be independent of one another and sampled at consistent and discrete intervals, which is most often not the case.
+Consequently, animals' spatial needs are often under-estimated [@noonan_comprehensive_2019].
+
+### Unassigned
+
+@teitelbaum_beyond_2019:
+
+- **Nomadic movement**: movement with irregular timing and/or direction that produces both within- and between-year variability in location and movement patterns
+- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north Atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"
+
+\newpage
 
@@
 
-Figure 2 of @bista_effects_2022 for example of study that assumes linear relationships (w LM) for home range, when there can't be one.
-
-```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of mean resource abundance ($\\mathbb E(R)$).
-At low values of $\\mathbb E(R)$, range-resident animals (bottom dashed line) may exhibit nomadic behavior, such that While $H$ decreases as $\\mathbb E(R)$ increases, it is not possible for $H$ to continuously decrease linearly because animals will require a minimum (nonnegative) amount of space to survive.
-At low values of $\\mathbb E(R)$, $H$", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of mean resource abundance ($\\mathbb E(R)$).
+We expect low values of $\\mathbb E(R)$ to result in high $\\mathbb V(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\mathbb E(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
+Note that the relationship between $\\mathbb E(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for high values of $\\mathbb E(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 
-Environmental productivity is tightly linked to the amount of space animals need to cover to obtain the resources they needed to survive and reproduce [figure \@ref(fig:), @relyea_home_2000].
-While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction, such as energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @refs], and maintaining a thermoregulatory balance.
-The size of a home range is hypothesized to be proportional to resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005].
-Animals living in large groups will require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004].
+<!-- You need to open with some background on metabolic requirements.
+Individuals have fixed needs that they must meet in order to survive, grow, and reproduce.
+If these are not met, they will die.
+Then transition to the idea that, because of that, they move around in order to secure resources.
+Therefore, when environments are productive, they don't have to move as far to meet their energetic requirements.
+-->
 
-The favorableness of a patch or habitat often depends on a variety of factors, including resource abundance, competitive pressure, and predation risk.
-Since many of these factors' affect resource abundance unpredictably, it is useful to visualize it as a random variable^[In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$, $H$).
-Known values, such as realizations (i.e., known observations) of random variables, are indicated with lower-case letters (e.g., $r$, $h$).
-Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.].
-Let $R$ indicate the amount of *resources* in a particular patch or area.
-For simplicity, we can assume $R$ can take any value between zero and infinity, and we can let $R$ follow a Gamma distribution with mean $\mu$ and variance $\sigma^2$ (which we can write as $R \sim \Gamma(\mu, \sigma^2)$)^[Although Gamma distributions are more often parameterized using shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ for simplicity.
-However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean since as the mean approaches zero the variance also does: $\mu = 0 \iff k \theta = 0 \iff k \theta^2 = 0 \iff \sigma^2 = 0$.].
-Next, let $H$ indicate the size of an animal's home range (HR)^[We can consider specific utilization quantiles of the HR, such as the core HR, $H_{50\%}$, or the 95% HR, $H_{95\%}$, but for simplicity I will refer to the entirety of the HR with $H$.].
+* add refs to generalize beyond terrestrial mammals
+
+\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
+All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref].
+Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death.
+Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)].
+Low values of $\mathbb E(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
+As $\mathbb E(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
+
+Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs.
+From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$).
+As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.
+
+# Effects of environmental variance on animals' spatial needs
+
+```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of resource unpredictability ($\\mathbb V(R)$).
+We expect low values of $\\mathbb V(R)$ to result in low $\\mathbb V(\\text{position})$ as animals are able to depend on somewhat predictable resources.
+As $\\mathbb V(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
+Note that the relationship between $\\mathbb V(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for low values of $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
+```
+
+\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\mathbb V(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\mathbb V(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
+Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
+A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
+There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+
+Simlarly to the effect of $\mathbb E(R)$, we expect $\mathbb V(R)$ to have a nonlinear effect on $\mathbb V(\text{position})$.
+When $\mathbb V(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\mathbb V(R)$ to have a noticeable effect on $\mathbb V(\text{position})$.
+However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).
+
+The effects of $\mathbb V(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently.
+Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\mathbb V(\text{position})$ may not detect such differences.
+@rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
+Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic.
+In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions.
+Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality.
+The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.
+
+* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\mathbb V(R)$
+
+# Interactive effects of mean and variance in resource abundance
 
@@
 
-Consider the simplistic scenario where $\mu$ changes over time but $\sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time: $\mathbb E(R) = \mu(t)$) while the variance is constant, i.e., $\mathbb V(R) = \sigma^2$ (figure \@ref(fig:multipanel-constant-var)b).
+\noindent Consider the simplistic scenario where $\mathbb E(R) = \mu$ changes over time but $\mathbb V(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
 Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
@@
 Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance.
-Examples of temporally homogeneous habitats include regions which throughout the year remain highly productive but fairly predictable (e.g., equatorial rain forests) or have very low productivity and variability (e.g., deserts without strongly distinct seasons).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case.
-Most habitats with seasonal changes in climate will require models to account for changes in both $\mathbb E(R)$ and $\mathbb V(R)$, since periods of high productivity will tend to be more variable than times of "dormancy" or lower productivity.
+Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
+When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs].
+Most often, $\mathbb E(R)$ will vary over time, and animals' behaviors will change in response.
+Additionally, when data availability is sufficiently high to produce precise estimates of $\mathbb V(R)$, the effect of $\mathbb V(R)$ on animals' $\mathbb V(\text{position})$ should be estimated before assuming the effect is negligible.
 
-The top row of figure \@ref(fig:5-5-sims) (constant $\mathbb V(R)$) illustrates the simulated spatial requirements of an animal living in habitats where $\mathbb E(R)$ remains constant or changes over time but $\mathbb V(R)$ remains constant.
-(Although some noise in $H$ are still present since $\mathbb V(R) \ne 0$ and thus the $R$ is not constant).
-Note how the home range size, $H$, decreases nonlinearly as $\mathbb E(R)$ increases linearly (linear $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), with larger changes when $\mathbb E(R)$ is low, since the multiplicative changes in $R$ are larger.
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
+(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
+(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$.
+(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
+```
+
+When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness.
+Since $\mathbb V(R)$ is not constant over time, we can now define it as a function of time using the notation $\mathbb V(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
+In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
+
+*Needs for understanding effects of variance* are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+
+<!--
+in stochastic environments:
+- higher spatial needs and possibly store resources for periods of scarcity, which are expected to get longer with climate change
+- heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one
+-->
+
+Most habitats with seasonal changes in climate will require models to account for changes in both $\mathbb E(R)$ and $\mathbb V(R)$, since periods of high productivity can often be more variable than times of "dormancy" or lower productivity, and the variability in $R$ often changes both within and between years.
+In such cases, failing to account for $\mathbb V(R)$ can result in serious biases in the estimation of animals' spatial needs.
+
+\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\mathbb E(R)$ and $\mathbb V(R)$.
+The top row (constant $\mathbb V(R)$) shows how  remains constant.
+Some noise in $H$ is still present since $\mathbb V(R) \ne 0$ and thus $R$ is not constant).
+Note the nonlinear decrease in $H$ when $\mathbb E(R)$ increases at a constant slope (linear $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), with larger changes when $\mathbb E(R)$ is low, since the multiplicative changes in $R$ are larger.
 Additionally, the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range $H_{50\%}$.
@@
 
-# Effects of environmental variance on animals' spatial needs
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$.
-(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
-
-Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], the variance in $R$ has received far less attention.
-However, $R$ can vary strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
-Environmental stochasticity, including extreme events, can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
-A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
-
-@seigle-ferrand_systematic_2021: Landscape heterogeneity influenced the sizes of home ranges in about 96% of studies based on individuals monitored by VHF but in only about 75% for individuals monitored by GPS, whatever the number of studied metrics.
-
-When animals are not guaranteed that the resources they find during one visit will be there the next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness.
-Since $\mathbb V(R)$ is not constant over time, we can now define it as a function of time using the notation $\mathbb V(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
-In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
-
-In this paper, we separate $\mathbb V(R)$ into heterogeneity and stochasticity in $R$.
-We refer to predictable or expected spatiotemporal changes as environmental **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation.
-For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable.
-In contrast, whether it is a good or bad year for berries is stochastic.
-<!--Thus, changes in $\mathbb V(R)$ may predictable or not, and $\mathbb V(R)$ is affected by both sources of variation.-->
-
 The leftmost column of figure \@ref(fig:5-5-sims) (constant $\mathbb E(H)$) illustrates how an animal's home range may vary as $\mathbb V(R)$ changes over time while $\mathbb E(R)$ remains constant.
@@
 
-The effects of $\mathbb V(R)$ on animals' spatial requirements likely depend on the animals' needs.
-@rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
-Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic.
-In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions.
-Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality.
-The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.
-
-What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability.
-Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004], which allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high fertility of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can allow them to develop traits that increase the chances of survival following an extreme event.
-Additionally, the effects of size and lifespan on sensitivity are likely nonlinear and correlated, since smaller animals tend to have shorter lives (and vice-versa), and small animals also tend to have lower metabolic rates, which often limit their movement speed, home ranges, and how the animals interact with their ecosystem [@brown_toward_2004].
-Thus, processes will have stronger impacts on smaller animals than on bigger ones: The grazing pattern of a bison (*Bison bison*) drastically alters the habitat of most crawling insects, but other grazing mammals would not pay attention to changes in grass length at the same spatial or temporal resolution.
-Similarly, the timing and quantity of yearly snowfall would be a somewhat predictable and expected event for most adult moose (*Alces alces*), but it may be a shock for many adult zooplankton or a new-born wolf (*Canis lupus*).
-What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal.
-Therefore, the mean and variance in $R$ should be modeled while accounting for the behavior of the species of interest.
-
-An unresponsive animal may perceive frequent changes in $R$ as high $\mathbb V(R)$, while a responsive animal may perceive them as changes in $\mathbb E(R)$.
-Thus, $R$ should be modeled using very flexible and wiggly models when animals are highly responsive, while less flexible models are more appropriate for more unresponsive or sedentary species.
-Generalized Additive Models [GAMs, see @wood_generalized_2017] are sufficiently flexible and adaptable to allow one to model $R$ while reflecting various levels of adaptability and responsiveness.
-(See @simpson_modelling_2018 for an overview of how to select appropriate bases sizes and smoothness parameters for GAMs.) The animal's movement (e.g., degree of range residency, range crossing time, average speed, directional persistence) should also inform the complexity of any spatial smooths included in the models.
-Models with complex spatial terms are appropriate for animals that recognize the location of high-quality patches, while smoother terms are best for animals that may perceive differences in $R$ as spatial variance, instead.
-
-**it may be worth chatting about the NDVI models for the tapir and the buffalo later this month**
-
 # Discussion
 
-There are many factors which may affect animal's spatial requirements and how they use their HR [*sensu* @nathan_movement_2008].
-In particular, resource abundance and environmental productivity are often hypothesized to be inversely proportional to HR size (or some function of it), such that regions with higher abundance correspond to smaller spatial needs, since animals do not have to range over extensive areas to meet the energetic needs [e.g., food, water, see @baldwin_nutritional_1984], habitat needs [e.g., dens, trees, tall grass, breeding grounds, protection from predators and competitors, see @southwood_habitat_1977], and maintaining a thermoregulatory balance.
-@nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
-However, the effect of many other factors on HR size likely depends on the species, location, and possibly also on the individual animal.
+\noindent @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
+However, the effect of many other factors on HR size likely depends on the species, diet type [@harestad_home_1979], location, and possibly also on the individual animal.
 For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; capuchin monkeys, @torrez-herrera_monkeys_2020].
@@
 
-```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$, a.-d.) and an African buffalo ($Syncerus~caffer$, e.-h.) as mean and variance in resource abundance change over the tracking period.
-(a., e.) Tracking data of the two animals.
-(b., f.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir during the period of reference.
-(c., g.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir during the period of reference.
-(d., h.) Estimated home range size of the tapir during each reference period, based on 50\\% (bold) and 95\\% (thin) utilization distributions.
-The tapir movement data corresponds to the individual named \"Anna\" from the Pantal sample of Medici $et~al.$ (2022), while the buffalo data corresponds to the buffalo named \"Cilla\" in the data of Getz $et~al.$ (2007) available on through the $\\tt{Movebank}$ package (Cross $et~al.$ 2016).
-The buffalo tracking data are also available in the $\\tt{buffalo}$ dataset in the $\\tt{ctmm}$ package.", out.width = '100%'}
-knitr::include_graphics('../figures/moving-window-examples.png')
+```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
+(a.) GPS tracking data of the tapir.
+(b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir.
+(c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir.
+(d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles.
+(e.) Effect of resource abundance on home range size.
+(f.) Effect of resource unpredictability on home range size.
+The effects in panels e and f were estimated using generalized linear models with Gamma cinditional distributions.
+The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
+knitr::include_graphics('../figures/tapir-example.png')
 ```
@@
 \newpage
-
-<!--
-in stochastic environments:
-- higher spatial needs and possibly store resources for periods of scarcity, which are expected to get longer with climate change
-- heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one
--->
 
@@
 - there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
-- Elk departed earlier from highly-productive areas?
+- ***Elk departed earlier from highly-productive areas?
 This goes against the idea that animals move as a consequence of need.
@@
 Maybe elk migration is triggered when they have sufficient energy (if the weather is still not too harsh).
-- Elk left early when exposed to more hunting
 -->
-
-<!-- here! -->
-
-# Interactive effects of mean and variance in resource abundance
-
-Although the effects of environmental variance on animals' spatial needs remain largely undestudied and unaccounted for, available evidence suggests that the effect of $\mathbb V(R)$ depends on the species and landscapes studied, so it would be inappropriate to develop a single hypothesis for how $\mathbb V(R)$ affects animals' spatial needs.
-
-[*Needs for understanding effects of variance*] are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
-
-* variance has less of an effect when variance is high
 
@@
 
-<!--
-
-# Movement types
-* Range residency
-* Seasonal migration
-* Nomadism
-
-predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move -> higher effect of V(R)
-
-@teitelbaum_beyond_2019:
-- **Nomadic movement**: with irregular timing and/or direction, producing both within-year and between-year variability in location and movement patterns
-- **Nomadic movement** is not the same as **migration**, because migration is predictable and cyclical while nomadism is random in time and space
-- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"
-
-## Edge dynamics
-
--->
-
 # Conclusion
 
-The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance.
+\noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance.
 We provide realistic and flexible hypotheses of the effects of $\mathbb E(R)$ and $\mathbb V(R)$ on animals' spatial needs and movement behavior.
@@
 
-In the case of animals whose spatial needs change over the their lifetimes (figure \@ref(fig:tapir-hr)), one may define HR as a function of time (with distinct HRs for each period, e.g., salmon spawning) or as a continuous function of age (so the HR changes smoothly).
-In either case, the accuracy of the estimate parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
+In the case of animals whose spatial needs change over the their lifetimes [@refs], one may define HR as a function of time with distinct HRs for each period [e.g., salmon spawning: @ref] or as a continuous function of age [so the HR changes smoothly: @ref].
+In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
 Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
 
-## The temporal scale of stochastic events (DELETE THIS SECTION?)
+The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
 
-The timescale over which an event is (un)predictable will also play a key role in determining how species respond, or not, to stochasticity.
-Adverse events which occur frequently during an organism's generation time or lifespan (e.g., at least twice, on average) may be considered as threats worth preparing for.
-Stochastic and destructive events which happen frequently can become part of an organism's habitat and life history, so organisms may either adapt to their occurrence and even learn to predict them [@foley_severe_2008], or they may avoid the risk altogether by moving to a new habitat [@southwood_habitat_1977].
-If habitats are (or become) unfavorable, animals may move to a new habitat, migrate seasonally [@geremia_migrating_2019], or simply prefer nomadism over range residency [e.g., @nandintsetseg_variability_2019].
-It is unclear when animals switch from range residency to nomadism (or vice-versa), but understanding the connection between the two types of movement is important for quantifying the effect of spatiotemporal stochasticity on animal's spatial needs.
-From a quantitative perspective, the switch is related to an animal's positional autocorrelation over time (also known as home range crossing time, here indicated as $\tau_p$).
-Animals without a constant home home range (i.e., without a constant centroid) will spend more time away from the overall mean position, so the time required to crossing their entire range time will be large (i.e., on the order of the animal's lifespan).
+## The spatiotemporal scale of stochastic events: memory and adaptability
 
-```{r tapir-hr, fig.cap="Movement of a tapir during a monitoring period of approximately 3 years (1997-07-10 to 2000-06-08) used in the work by Medici $et~al.$ (2022).
-Different subsamples may result in different estimates of home range size and centroid (a, b; each span 100 days), so neither subsample is necessarily representative of the space the animal used over the entire monitoring period (c).", out.width = '0.4\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.4)}
-knitr::include_graphics('../figures/hr-and-lifespan.png')
-```
+The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism (i.e.
+approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
+Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
 
-Adverse events which are too infrequent (e.g., $p \lesssim 0.1$) may not be perceived as a reoccurring threat, so organisms may be unable to adapt appropriately.
-Instead, highly infrequent events are more likely to perceived as an oddity rather than something worth preparing for.
-In contrast, highly stochastic events (e.g., $p \approx 0.5$ or if $p$ changes unpredictably) may be perceived as a threat, but organisms may be unable to predict their occurrence, since it would require refined cognitive abilities.
-For a species or population to adapt to an event, the event must thus occur with sufficient frequency and for a sufficiently long period of time.
+Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
+@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck].
+@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
 
-<!-- What an animal perceives as a stochastic process depends on the stochasticity of the process relative to the animal's size, current age (or average lifespan), and adaptability.
-Generally, small, short-lived, or young organisms will tend to be more sensitive to small-scale (spatial) or short-term (temporal) changes [@southwood_habitat_1977].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004], which allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high fertility of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can allow them to develop traits that increase the chances of survival following an extreme event.
-Additionally, the effects of size and lifespan on sensitivity are likely nonlinear and correlated, since smaller animals tend to have shorter lives (and vice-versa), and small animals also tend to have lower metabolic rates, which often limit their movement speed, home ranges, and how the animals interact with their ecosystem [@brown_toward_2004].
-Thus, processes will have stronger impacts on smaller animals than on bigger ones: The grazing pattern of a bison (*Bison bison*) drastically alters the habitat of most crawling insects, but other grazing mammals would not pay attention to changes in grass length at the same spatial or temporal resolution.
-Similarly, the timing and quantity of yearly snowfall would be a somewhat predictable and expected event for most adult moose (*Alces alces*), but it may be a shock for many adult zooplankton or a new-born wolf (*Canis lupus*).
-What one animal may perceive as a single stochastic event (or cycle) may be considered as a series of short and highly stochastic events by another animal.
-**add empirical examples** -->
+\newpage
 
-Figure \@ref(fig:stochasticity-timescale-figure) illustrates the perceived resource availability mean and variance by animals with four different levels of adaptability and how they may predict the mean and variance to change in the future (after the vertical grey line).
-In the extreme (and likely fictitious) case where an animal does not adapt to the environment (dark blue), it cannot react to or predict changes in mean resource abundance, and thus the perceived variance is inflated whenever the resource abundance does not match the overall mean.
-Environments appear highly unpredictable to such animals.
-Animals which can only perceive simple, linear changes in mean resource abundance (orange) perform better as long as resources continue to change linearly in the same direction (i.e.
-increasing or decreasing).
-However, they may be surprised when the trend's direction changes, as indicated by the rapid increase in variance in the top rows of figure \@ref(fig:stochasticity-timescale-figure)b.
-Animals which are able to adapt smoothly (light blue) can predict changes in mean without a significant bias in estimated variance, as long as the process is not highly stochastic (unsurprisingly, see the bottom rows of figure \@ref(fig:stochasticity-timescale-figure)).
-Finally, animals which adapt rapidly to changing environments may be most able to take advantage of frequent environmental changes, but they may be unable to produce reasonable predictions based on memory [*sensu* @fagan_spatial_2013; @abrahms_memory_2019].
-Instead, they depend on constant information and only predict on the most recent information (as indicated by the deviations from the data when predicting into the future).
+# Extra notes
 
-<!-- The recent shift to more quantitative statistical methods in environmental biology has pushed scientists to begin thinking more critically about data analysis.
-However, many still operate using methods and frameworks which were necessary with small datasets, but likely become problematic with an abundance of data.
-For instance, the widespread frequentist approach of hypothesis testing with a significance level of $\alpha = 0.05$ may often be misleading, because the significance of an effect depends on its standard error, which is a function of its sample size, since larger samples will offer a more precise and accurate measurement of the true mean.
-Similarly, the question of how much data is needed to detect a significant trend [@gray_how_2018] may seem important to some, but I suggest we pay less attention to whether an event or trend is statistically significant over a period of time based on the dichotomy of significance based on an arbitrary significance level.
-Since statistical significance does not only depend on effect size but also sample size (including measurement frequency), high-frequency data is likely to produce statistically significant effects, even in the event of a false positive.
-Instead, we should focus primarily on whether a change has an important impact on the fitness of the animal(s) of interest [@zuur_mixed_2009].
--->
+@teitelbaum_how_2015 found that resource abundance (measured as NDVI) had a strong negative effect on migration distance, while the variance in NDVI only had a small positive effect on **???**.
 
-An animal's ability to alter its behavior (including its movement) in response to environmental conditions is essential in stochastic or changing landscapes.
-Whether the changes be due to highly variable but (potentially) predictable changes such as the seasons and the weather, or whether they be due to more stochastic events [e.g., natural events such as fires or floods, but also anthropogenic events like oil spills, see @matkin_ongoing_2008], an animal's ability to adapt increases its odds of survival.
-Generally, temporal variation is more likely to promote plasticity over diversity, since adaptability will likely offer better odds of survival than temporally static diversity in a population or species [@bell_spatial_1993].
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on variable environmental events and cues such as available forage biomass, hunting pressure, snow fall, and snow melt.
-Birds have also shown to change their their migration as the climate changes, including large-scale, trans-Saharian migrants [@jonzen_rapid_2006].
-And while genetic diversity and polymorphisms also increase the odds of survival for a species [@cavedon_genomic_2022], this project will focus strictly on animal behavior.
-
-In an environment that changes over time, organisms which depend on mutable cycles such as changes in temperature, precipitation, and resource availability are more likely to respond to environmental changes than organisms which depend purely on deterministic cycles (e.g., photoperiod).
-While this thesis focuses on the movement and spatial use of animals, the adaptability (or lack thereof) of non-animal organisms is also crucial.
-The ability of most animals to rely on visual cues and move accordingly greatly increases their adaptability and plasticity, particularly for those animals that are able to move large distances over short periods of time (e.g., flying birds and large vertebrates).
-Although this project will focus on vertebrate animals, such changes are important to consider because the ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
-
-## The spatial scale of stochastic events (DELETE THIS SECTION?)
-
-To measure the spatial stochasticity of an ecosystem, it is necessary to first define extent of the ecosystem of interest.
-A lake may seem like a simple and self-contained space to define, but it is often not as simple as deciding the lake's boundaries using a map or satellite pictures.
-Should one include the shores of the lake?
-And if so, which region do we consider the lake's shores?
-Are they the area of land which are currently covered by water, including the sections hit daily by waves, or should we also include the parts of land which are currently not wet but are covered during seasons with more rain?
-One could base their definition on the species of interest, since coastal areas are not frequented by large fish, but they would likely include them in the ecosystem if they are studying benthic littoral species.
-Thus, how one measures stochasticity should also depend on which organisms are of most interest.
-
-Since species in a heterogeneous landscape can modify their behavior and move to more favorable locations, spatial variation will favor movement and behavioral adaptations faster than temporal variation will [@bell_spatial_1993].
-
-Transition regions are often have the highest heterogeneity, but, similarly to temporal stochasticity, measures and perceptions of spatial heterogeneity depend on the scale at which they are being considered.
-The repetitive and cyclical motion of waves may not concern most vertebrates that live along the coast, but it is much more turbulent and chaotic for smaller organisms, such as crustaceans or diatoms.
-Therefore, the scale at which we consider spatial stochasticity should depend on the size of the organisms of interest as well as the size of their habitat or home range.
-
-Although @bell_spatial_1993 couldn't find a scale at which variance stopped increasing, it seems reasonable to assume that variance cannot increase infinitely.
-The variance of a finite system (e.g., a lake or a well-defined habitat), should reach a finite maximum somewhere between our smallest scale of interest and the scale of the entire system, even if the maximum is reached when the entire system is considered.
-This does not imply that a somewhat larger system that contains the first one (e.g.
-the lake and its coasts) will have the same variance.
-Variance between lake and coast is high because they are different types of habitats, and the area where they meet (the coast) is potentially even more variable because the mixture of the two creates a third habitat for littoral benthic animals which is subject to high stochasticity due to waves and changing water levels.
-
-The same idea would likely also apply for larger-scale systems, such as continents or a planet.
-And while stochasticity of a system may also vary over time, it is not reasonable to expect it to increase monotonically through time because that would imply it has been increasing monotonically until the current moment.
-Still, stochasticity does vary over time, and its estimation depends on the data available, so it may be useful to view it as a random variable with an estimated, conditional posterior distribution through a Bayesian framework [@bolstad_introduction_2017].
+high-resource areas and explaining 23% of the variation in migration distances.
 
@@
 * Attribute memory, i.e., non-spatial memory, e.g., food type and abundance (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B13)
-* Bison remember valuable information about the location and quality of meadows (spatial and attribute memory) and use this information to revisit profitable locations (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B29)
 * Roe deer an spatial memory for foraging in absence of territorility (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B41)
@@
 
-<!--
-## Urban landscapes and habitat structure
-
-Human activity often fundamentally alters the structure of environments by spatially homogenizing them and imposing temporally heterogeneous and potentially stochastic cycles (e.g., mining, clear-cutting, road traffic, and weekly work schedules).
-The perception of spatial heterogeneity and cycles in anthropogenic areas varies between species: scavenger animals such as ravens (*Corvus corax*) and raccoons (*Procyon lotor*) often learn to recognize areas that can provide food, such as restaurants or large disposal bins, but other animals may not be able to distinguish between different urban areas.
-Certain animals have learned to depend on and take advantage of human activity, such as crows who place nuts strategically so that cars will crack them open for them to eat [@nihei_when_2002].
--->
-
 \newpage <!-- place references on a separate page -->

### 2022-10-26 writing/appendix-1-simulations.Rmd
started the appendix on simulations

---
title: "Appendix 1: Simulations"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
fontsize: 12pt
indent: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
    toc: false
---

<!-- set default code chunk options -->

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE)
```

<!-- add custom commands for formulae -->

\newcommand{\e}{\mathbb E}
\newcommand{\var}{\mathbb V}
\newcommand{\vpos}{\mathbb V(\text{position})}

<!-- have table of contents on pages of its own -->

\newpage

\tableofcontents

# To do {-}

* change all references to scripts to the correct file locations
* give an overview of how the sims work at the beginning of the pdf
* explain the checks in `delta-t-sensitivity.R`
* reduce scaling of figure 2

\newpage

# Overview {-}

This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript. For ease of reference, we also include the figures here:

```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time. The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile. Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase approximately linearly as $\\mathbb V(R)$ increases. Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time. The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
```

\newpage

```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements. The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile. Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
knitr::include_graphics('../figures/simulation-regression-plots.png')
```

## explain how the sims work

To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document. Instead, the R Markdown document ($\tt{writing/appendix-1-simulations.Rmd}$) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file. Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes and as a general guide. To replicate the simulations, we suggest using the `R` scripts, instead.

# Simulating the movement tracks

To reduce sampling variance between simulations, we use the same set of simulated tracks for each time point in each panel. In the $\tt{analysis/simulations/tracks.R}$ script, we generate $2^{10} = 1024$ tracks to check how many tracks are necessary to obtain stable home range estimates in the best- and worst-case scenarios. Most intermediate and diagnostic checks included in the `R` scripts are not included in this document for the sake of brevity and simplicity.

```{r load-packages, eval=TRUE, comment=FALSE}
# NOTE: change your working directory to be "env-var-review/writing", or
# modify all file paths as needed
library('ctmm')   # for generating movement models and movement modeling
library('raster') # for working with raster data
library('dplyr')  # for data wrangling
library('purrr')  # for functional programming
source('../functions/label_visits.R') # decides when animal encounters food

DELTA_T <- 60 # sampling interval in seconds
SAMPLES <- seq(0, 60 * 60 * 12, by = DELTA_T) # 12h every DELTA_T seconds

# projected raster of resources
PROJECTION <- '+proj=aeqd +lon_0=0 +lat_0=0 +datum=WGS84'
HABITAT <- matrix(data = 1, nrow = 500, ncol = 500) %>%
  raster(xmx = 1e3, xmn = -1e3, ymx = 1e3, ymn = -1e3, crs = PROJECTION)

# infinitely diffusive movement model
model <- ctmm(tau = c(Inf, 1e3), sigma = 0.1, mu = c(0, 0))
# summary(model) # FAILS TO KNIT BECAUSE OF TAU CHARACTER # check the parameters of the model
```

```{r simulate-tracks}
N_DAYS <- 2^10 # number of "days" (i.e., tracks with different seeds)

# extracts tracks from a ctmm movement model for given sample times
get_tracks <- function(day, times = SAMPLES) {
  simulate(model, # ctmm movement model
           t = times, # sampling times in seconds
           seed = day, # for a consistent track each day
           complete = TRUE, # add lat, long, and timestamp to telemetry
           crs = PROJECTION) # CRS projection string
}

# generate simulated tracks (will be truncated at satiety later)
tracks <- tibble(day = 1:N_DAYS, # a simulation for each day
                 tel = map(.x = day, # set a seed for consistent results
                           .f = get_tracks)) # function to generate tracks

# find patch visits and calories consumed from the tracks
tracks <-
  transmute(tracks, # drop tel column
            day, # keep day column
            track = map(.x = tel, # add a column of full tracks
                        .f = \(x) {
                          label_visits(.tel = x, .habitat = HABITAT)
                        }))

# make a single, large tibble (will need lots of RAM)
tracks <- tidyr::unnest(tracks, track)
```

After generating the tracks, we performed the following tests to ensure the tracks were appropriate and results would be stable:

## Checking whether adding return trips is necessary

This check was performed with the code in the $\tt{analysis/figures/simulations/return-sensitivity.R}$ script. Adding return trips to $\langle0, 0\rangle$ after an animal reached satiety doubled computational times without appreciable improvements on the home range estimates (including the 95% CIs).

## Checking whether the sampling interval is sufficiently small

This check was performed with the code in the $\tt{analysis/figures/simulations/delta-t-sensitivity.R}$ script. ***explain what I did***

```{r dt-sensitivity, fig.cap="Diagnostic tests used to decide an appropriate sampling interval. (a.) Histograms of the number of encounters as a function of the interval between envounters, with a binwith of 30 seconds. Although some encounters occur with less than 30 seconds between them, most of them occur at least 60 seconds apart. (b.) Number of encounters with food detected as a function of sampling interval. The colored lines indicate the estimated relationship based on a Generalized Additive Model fit using the $\\tt{geom\\_smooth}$ function from the $\\tt{ggplot2}$ package. Although the number of encounters detected decreases as sampling interval doubles, the loss at 30 seconds is negligible. (c.) Beginning of the track generated with seed \"1\" (purple line) with the location of each sample for different sampling intervals. Red dots indicate samples where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell, and thus encountered food. While the number of feeding events detected decreases as the sampling interval increases, most of the feeding events lost at $\\Delta t = 30$s occured because the animal remained almost adjacent to the borders between cells. (d.) The three tracks used in these tests over the raster used to determine when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
knitr::include_graphics('../figures/thinning-examples.png')
```

## Checking how many tracks were necessary

This check was performed with the code in the $\tt{analysis/simulations/hr-simulation-extreme-scenarios.R}$ script. ***explain, add figures***

# Main scripts (to be run in the following order)

2. $\tt{analysis/simulations/hr-mean-variance-simulations-days.R}$
3. $\tt{analysis/simulations/hr-mean-variance-simulations-days-summarized.R}$
4. $\tt{analysis/simulations/hr-mean-variance-simulations-modeling.R}$
5. $\tt{analysis/simulations/hr-mean-variance-simulations-hrs.R}$

### 2022-11-02 writing/appendix-2-empirical-modeling.Rmd
started appendix 2

---
title: "Appendix 2: Empirical modeling"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emilia Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
bibliography: 'env-var-review.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \usepackage{indentfirst} # for indenting first line of each paragraph
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    # smaller caption text, less space above and below caption
    - \captionsetup[figure]{font=scriptsize}
subparagraph: true # needed for \usepackage[small]{titlesec}
linenumbers: true
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
---

<!-- set default code chunk options -->

```{r, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, fig.align = 'center')
```

<!-- add custom commands for formulae -->

\newcommand{\e}{\mathbb E}
\newcommand{\var}{\mathbb V}
\newcommand{\vpos}{\mathbb V(\text{position})}

<!-- have table of contents on pages of its own -->

\newpage

\tableofcontents

# To do {-}

* add scripts to the correct folders and check
* add "for more info, see the tapir project" at the beginning
* add refs for each package used

\newpage

This appendix illustrates the steps necessary to reproduce the tapir movement analysis and figures in the main manuscript. For ease of reference, we also include the figure here:

```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance. (a.) GPS tracking data of the tapir. (b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir. (c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir. (d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles. (e.) Effect of resource abundance on home range size. (f.) Effect of resource unpredictability on home range size. The effects in panels e and f were estimated using generalized linear models with Gamma conditional distributions. The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
knitr::include_graphics('../figures/tapir-example.png')
```

Estimating the effects of resource abundance and unpredictability on the tapir's spatial needs requires us to first estimate the changes in the tapir's spatial needs (section \@ref(movement)) and the changes in resource abundance and variance (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use to recreate the final figure (section \@ref(final-fig)).

To minimize the computational costs of creating this appendix, we load the necessary objects through hidden `R` chunks rather than running all the code in the final pdf. Still, those interested in replicating the analysis can do so by using the code in the pdf document or the related `R` markdown (`Rmd`) document. Listed below are all the packages and source scripts required to run the analyses in this document:

```{r, eval=TRUE, comment=FALSE}
# attach all necessary packages
library('raster')    # to import and save rasters
library('dplyr')     # for data wrangling
library('purrr')     # for functional programming
library('tidyr')     # for data wrangling
library('ggplot2')   # for fancy plots
library('cowplot')   # for fancy multi-panel plots
library('ctmm')      # for movement modeling
library('mgcv')      # for empirical Bayesian modeling
library('lubridate') # for smoother date wrangling
library('sf')        # for spatial features
library('MODIStsp')  # for downloading NDVI rasters

# resolve conflicts between packages
select <- dplyr::select

# source all necessary scripts
source('../analysis/figures/default-figure-styling.R') # NDVI color palette
source('../earthdata-login-info.R') # personal login info for EarthData
source('../analysis/figures/default-figure-styling.R') # for color palettes
source('../functions/window_hr.R') # function to calculate HRs
```

# Modeling the tapir's movement over time {#movement}

The script `analysis/tapir/tapirs-moving-window.R` estimates the seven-day spatial use of various tapirs from the Brazilian Cerrado. Here, we simplified the code so that it only estimates the spatial use of the tapir in the manuscript, Anna:

```{r}
# import tapir data from https://github.com/StefanoMezzini/tapirs
anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
  filter(name.short == 'ANNA')
anna_tel <- anna$data[[1]] # telemetry data

# projection for the region in the Brazilian Cerrado
ctmm::projection(anna_tel) <- '+proj=utm +zone=22 +datum=NAD83 +units=m'

# calculate the 7-day home range estimate
anna_mw <-
  window_hr(
    anna_tel,
    window = 7 %#% 'day', # 1 week of data for sufficient sample size
    dt = 1 %#% 'day', # move window over by a single day each time
    fig_path = '../figures',
    rds_path = '../models'
  )
```

```{r, echo=FALSE, eval=TRUE}
anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
  filter(name.short == 'ANNA')
anna_tel <- anna$data[[1]] # telemetry data

# projection for the region in the Brazilian Cerrado
ctmm::projection(anna_tel) <- '+proj=utm +zone=22 +datum=NAD83 +units=m'

anna_mw <- readRDS('../models/CE_31_ANNA-window-7-days-dt-1-days.rds')
```

The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days that starts with the first seven days of data (2017-06-24 to 2017-07-01) and then repeats the analysis for the second seven-day set (2017-06-25 to 2017-07-02) until it reaches the last set of days (2018-09-23 to 2018-09-30). It then saves an exploratory figure (figure \@ref(fig:mw)) to the `figures` folder and the list of movement models to the `models` folder.

```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function. Panel $\\mathbf{a.}$ shows the tapir's movement data, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals."}
knitr::include_graphics('../figures/CE_31_ANNA-window-7-days-dt-1-days.png')
```

# Modeling $\e(R)$ and $\var(R)$ over time {#ndvi}

To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli-ref]. We downloaded the data using the `MODIStsp` `R` package with the following code:

```{r}
anna_ud <- anna$akde[[1]]

bbox <-
  SpatialPolygonsDataFrame.UD(anna_ud,
                              level.UD = 0.9995, # utilization quantile
                              level = 0) %>% # no CIs
  st_as_sf() %>%
  st_transform(crs = '+proj=longlat') %>%
  st_bbox()

# download NDVI rasters (if needed, create all necessary folders first)
MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
         out_folder = '../data/ndvi-rasters/tapir-anna',
         selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)',
         prod_version = '061', # 2022 raster version
         bandsel = 'NDVI', # NDVI layer only
         sensor = 'Terra', # only terrestrial values, ignore water
         user = USERNAME, # Earthdata username for urs.earthdata.nasa.gov
         password = PASSWORD, # your Earthdata password
         start_date = format(min(anna_tel$timestamp) - 16, '%Y.%m.%d'),
         end_date = format(max(anna_tel$timestamp) + 16, '%Y.%m.%d'),
         spatmeth = 'bbox', # use a bounding box for the extent
         bbox = bbox, # spatial file for raster extent
         out_projsel = 'User Defined', # use specified projection
         output_proj = '+proj=longlat', # download unprojected raster
         resampling = 'bilinear', # raster resampling method for new proj
         delete_hdf = TRUE, # delete HDF files after download is complete
         scale_val = TRUE, # convert from integers to floats within [-1, 1]
         out_format = 'GTiff', # output format
         verbose = TRUE) # print processing messages

# save NDVI data as an rds file of a tibble
anna_ndvi <-
  list.files(
    path = '../data/ndvi-rasters/tapir-anna/VI_16Days_250m_v61/NDVI/',
    pattern = '.tif', full.names = TRUE) %>%
  stack() %>% # import all rasters as a single stack
  rasterToPoints() %>% # convert to a matrix of points
  data.frame() %>% # convert to a data frame
  pivot_longer(-c(x, y)) %>% # change to long format (x, y, name, value)
  transmute(long = x, # rename x column
            lat = y, # rename y column
            date = substr(name, # change name to a date
                          start = nchar('MOD13Q1_NDVI_x'),
                          stop = nchar(name)) %>%
              as.Date(format = '%Y_%j'), # format is year_julian date
            ndvi = value, # rename value column
            dec_date = decimal_date(date))
```

```{r, echo=FALSE, eval=TRUE}
# import NDVI data
anna_ndvi <-
  readRDS('../data/tapir-anna-ndvi-data.rds') %>%
  mutate(dec_date = decimal_date(date))
```

We removed the raster for 2017-12-19 because the values were unusually low for the region. We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood (which is common for the Cerrado):

<!-- using echo = TRUE and eval = TRUE because this needs to be both visible and evaluated -->

```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.width=5, out.width='100%'}
anna_ndvi %>%
  filter(date >= as.Date('2017-08-29'), date <= as.Date('2018-04-07')) %>%
  ggplot() +
  facet_wrap(~ date, nrow = 3) + # a raster for each date
  coord_equal() + # keep the scaling of x and y equal
  geom_tile(aes(long, lat, fill = ndvi)) +
  scale_x_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, breaks = NULL, expand = c(0, 0)) +
  scale_fill_gradientn('NDVI', colours = ndvi_pal, limits = c(-1, 1)) +
  theme(legend.position = 'top')

anna_ndvi <- filter(anna_ndvi, date != '2017-12-19') # remove bad values
```

Ideally, one should model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1. The beta family would be appropriate after applying the linear transformation $$y^* = \frac{y + 1}{2},$$ where $y$ is the original NDVI value (between -1 and 1) while $y^*$ is the NDVI value scaled between 0 and 1. However, the `mgcv` package [@ref] does not allow one to model changes in variance in NDVI using a location-scale model with a beta family. Although the `brms` package allows one to fit beta location-scale models [@ref], the computational costs of using a fully Bayesian approach can be prohibitive for large datasets. Therefore, we decided to use a location-scale Gaussian model via the `mgcv` package (`family = gaulss()` in the code chunk below). Given the large amounts of data, the predictions are sufficiently far from -1 and 1 that any bias appears to be negligible.

```{r}
m_ndvi <-
  gam(list(
    # mean predictor
    ndvi ~
      s(long, lat, bs = 'ds', k = 20) + # mean over space (k = 20 * 20)
      s(dec_date, bs = 'tp', k = 20),   # high k to account for adaptions
    # precision (1/standard deviation) predictor
    ~
      s(long, lat, bs = 'ds', k = 10) + # precision over space
      s(dec_date, bs = 'tp', k = 10)),  # precision over time
    family = gaulss(b = 0.0001), # minimum standard deviation of 0.0001
    data = anna_ndvi,
    method = 'REML') # performs better than GCV (the default)

# plot smooths (predictions are on the link scales!)
plot(m_ndvi, pages = 1, scheme = 1, n = 250)
```

```{r, echo=FALSE, eval=TRUE}
m_ndvi <- readRDS('../models/CE_31_ANNA-mgcv-ndvi-gulss.rds')
plot(m_ndvi, pages = 1, scheme = 1, n = 250)
```

# Modeling the effects of $\e(R)$ and $\var(R)$ on spatial needs {#final-fig}

We start by simulating a movement track that passes near all recorded locations using the movement model estimated from the telemetry data. Note that this step can take hours to run, so we suggest skipping it if one is only interested in replicating the most important portions of the analysis (since the track is only used for panel $\mathbf{a.}$ in the figure).

```{r}
# simulate the movement track with an interval of 100 seconds
anna_ouf <- anna$model[[1]]
track <-
  predict(anna_ouf, data = anna_tel, dt = 100, complete = TRUE) %>%
  data.frame() %>%
  select(timestamp, longitude, latitude) %>%
  rename(long = longitude, lat = latitude) %>%
  mutate(dec_date = decimal_date(timestamp))
```

```{r, eval=TRUE, echo=FALSE}
track <- readRDS('../models/CE_31_ANNA-track-dt-100.rds')
```

To create panels $\mathbf{b.}$-$\mathbf{f.}$, we use the NDVI Gaussian location-scale model to estimate the mean and variance in NDVI for each known location of the tapir.

```{r, eval=TRUE}
# function to make predictions from the NDVI model based on telemetry data
ndvi_preds <- function(.data) {
  .data <- .data %>%
    data.frame() %>% # convert telemetry data to a data.frame
    mutate(year = year(timestamp),
           doy = yday(timestamp))

  predict.gam(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE)%>%
    data.frame() %>% # from list to data frame
    tibble() %>% # from data frame to tibble
    transmute(mu = X1, # rename mean column
              sigma2 = (1/X2)^2) %>% # convert precision to variance
    return()
}

anna_tel <- data.frame(anna_tel) %>%
  tibble() %>%
  rename(long = longitude, lat = latitude) %>%
  mutate(dec_date = decimal_date(timestamp))
anna_tel <- bind_cols(anna_tel, ndvi_preds(anna_tel))
```

We also apply the function to the subset of the telemetry data belonging to each of the seven-day periods.

```{r, eval=TRUE}
anna_mw <-
  mutate(
    anna_mw,
    # estimate mean and variance in NDVI
    preds = map(dataset, \(.d)
                filter(anna_tel, timestamp %in% .d$timestamp)),
    # extract the column of mean NDVI and take the average
    mu = map_dbl(preds, \(.d) mean(.d$mu)),
    # extract the column of variance in NDVI and take the average
    sigma2 = map_dbl(preds, \(.d) mean(.d$sigma2))) %>%
  # only keep necessary coliumns
  select(t_center, mu, sigma2, hr_lwr_50, hr_est_50, hr_upr_50, hr_lwr_95,
         hr_est_95, hr_upr_95) %>%
  # change to long format (one column for both utilization quantiles)
  pivot_longer(c(hr_lwr_50, hr_est_50, hr_upr_50, hr_lwr_95, hr_est_95,
                 hr_upr_95), names_to = c('.value', 'quantile'),
               names_pattern = '(.+)_(.+)') %>%
  mutate(t_center = as.POSIXct(t_center, origin = '1970-01-01'),
         quantile = paste0(quantile, '%'),
         quantile = factor(quantile))
anna_mw
```

We now have all the necessary objects to create the final figure:

```{r}
# min and max of the tracking dates
date_labs <- range(anna_tel$timestamp) %>% as.Date()

p_track <-
  ggplot() +
  coord_equal() +
  geom_path(aes(long, lat), track, alpha = 0.3) +
  geom_point(aes(long, lat, color = dec_date), anna_tel) +
  scale_color_viridis_c(NULL, breaks = range(anna_tel$dec_date),
                        labels = date_labs) +
  labs(x = NULL, y = NULL)

p_mu <-
  ggplot(anna_mw, aes(t_center, mu)) +
  geom_line(color = pal[1]) +
  labs(x = NULL, y = '\U1D707(t)')

p_sigma2 <-
  ggplot(anna_mw, aes(t_center, sigma2)) +
  geom_line(color = pal[2]) +
  labs(x = NULL, y = '\U1D70E\U00B2(t)')

p_hr <-
  ggplot(anna_mw) +
  geom_line(aes(t_center, hr_est, group = quantile, size = quantile),
            color = pal[3], show.legend = FALSE) +
  geom_ribbon(aes(t_center, ymin = `50%`, ymax = `95%`),
              anna_mw %>%
                select(t_center, hr_est, quantile) %>%
                pivot_wider(values_from = hr_est, names_from = quantile),
              fill = pal[3], color = 'transparent', alpha = 0.2,
              inherit.aes = FALSE) +
  labs(x = NULL, y = expression(Home~range~size~(km^2)~'   ')) +
  scale_size_manual(values = c(1, 0.5))
```

To ensure all panels are properly aligned, we convert each plot to a grid object (a "grob") and then align each grob.

```{r}
# convert to grid graphical objects (grobs)
grobs <- map(list(p_track, p_mu, p_sigma2, p_hr), as_grob)

# align left margins of all plots
aligned_widths <- align_margin(map(grobs, function(x) {x$widths}), 'first')

# Setting the dimensions of plots to the aligned dimensions
for (i in seq_along(grobs)) {
  grobs[[i]]$widths <- aligned_widths[[i]]
}
# Draw aligned plots
p_left <- plot_grid(plotlist = grobs, ncol = 1,
                    labels = paste0(letters, '.'))
```

## there should be a model for anna instead of `geom_smooth()!

```{r}
reg_mu <-
  ggplot(anna_mw) +
  coord_cartesian(ylim = c(0, 13)) +
  geom_point(aes(mu, hr_est, group = quantile, shape = quantile), alpha = 0.3,
             color = pal[3]) +
  geom_smooth(aes(mu, hr_est, group = quantile, size = quantile),
              color = pal[1], se = FALSE, method = 'gam',
              formula = y ~ x, method.args = list(family = "Gamma")) +
  scale_x_continuous('Resource abundance') +
  scale_y_continuous(expression(Home~range~size~(km^2)), expand = c(0, 0)) +
  scale_size_manual(values = c(1, 0.5)) +
  theme(legend.position = 'none')

# variance
reg_s2 <-
  ggplot(anna_mw) +
  coord_cartesian(ylim = c(0, 13)) +
  geom_point(aes(sigma2, hr_est, group = quantile, shape = quantile),
             alpha = 0.3, color = pal[3]) +
  geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
              color = pal[2], se = FALSE, method = 'gam',
              formula = y ~ x, method.args = list(family = "Gamma")) +
  scale_x_continuous('Resource unpredictability') +
  scale_y_continuous(expression(Home~range~size~(km^2)), expand = c(0, 0)) +
  scale_size_manual('Quantile', values = c(1, 0.5)) +
  scale_shape('Quantile') +
  theme(legend.position = 'none')

leg <-
  get_legend(
    ggplot(anna_mw) +
      geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
                  se = FALSE, method = 'gam', color = 'black',
                  formula = y ~ x, method.args = list(family = "Gamma")) +
      geom_point(aes(sigma2, hr_est, group = quantile, shape = quantile),
                 color = pal[3]) +
      scale_size_manual('Utilization quantile', values = c(1, 0.5)) +
      scale_shape('Utilization quantile') +
      theme(legend.position = 'top')
  )

# add a legend on the top 
p_regs <- plot_grid(leg, reg_mu, reg_s2, ncol = 1,
                    rel_heights = c(0.1, 1, 1), labels = c('', 'e.', 'f.'))

# group all the plots together
plot_grid(p_left, p_regs)
```

```{r, echo=FALSE, eval=TRUE, out.width='100%'}
knitr::include_graphics('../figures/tapir-example.png')
```

### 2022-11-02 analysis/figures/default-figure-styling.R

@@ -1,0 +1,14 @@
+library('ggplot2') # for fancy figures
+library('stringi') # for working with strings
+
+# set default ggplot theme
+theme_set(theme_bw() + theme(text = element_text(face = 'bold')))
+
+# custom color-blind palette
+pal <- c('#ff8c00', '#4477AA', '#009900', '#66CCEE',
+         '#CCBB44', '#EE6677', '#AA3377', '#BBBBBB')
+
+# custom NDVI color palette
+create_ndvi_pal <- colorRampPalette(c('darkblue', 'dodgerblue', '#744700', '#d9bb94',
+                                      'darkgreen'))
+ndvi_pal <- create_ndvi_pal(100)

### 2022-11-02 functions/window_hr.R

@@ -1,0 +1,149 @@
+# function that loops through a moving window of X days of tracking data
+# computes estimated HR size and CIs for each window
+# plots ML point estimate of HR size and 95% CIs
+# written by Stefano Mezzini based on code by Mark Bidwell and Chris Fleming
+
+library('ctmm')  # for continuous movement modeling
+library('dplyr') # for data wrangling
+library('purrr') # for functional programming
+
+window_hr <- function(tel, window, dt, projection, fig_path = NULL,
+                      rds_path = NULL, cores = 1) {
+  
+  # moving window created at beginning of t; slides forward as far as possible:
+  # |---|.... --> .|---|... --> ..|---|.. --> ...|---|. --> ....|---|
+  
+  times <- seq(min(tel$t), max(tel$t) - window, by = dt)
+  N <- length(times)
+  
+  # to extract home ranges later
+  extract_hr <- function(a, par, l.ud) {
+    # convert HR to km^2
+    summary(a, units = FALSE, level.UD = l.ud)$CI['area (square meters)', ][par] / 1e6
+  }
+  
+  out <-
+    tibble(
+      # add start and end times
+      t_start = times, # left bound
+      t_end = t_start + window, # left bound
+      # subset times within window (t to t + win; can't use filter() bc telmtry)
+      dataset = map2(t_start, t_end,
+                     \(t_1, t_2) tel[tel$t >= t_1 & tel$t <= t_2, ]),
+      models =
+        imap(dataset,
+             \(d, i) {
+               
+               if(nrow(d) > 1) {
+                 cat('Analyzing dataset ', i, ' of ', N, '.\n', sep = '')
+                 tibble(
+                   # find initial guesses for models
+                   guess = ctmm.guess(data = d, interactive = FALSE) %>% list(),
+                   # select best model based on subset of tel
+                   model = ctmm.select(data = d, CTMM = guess[[1]],
+                                       cores = cores) %>%
+                     list(),
+                   # estimate autocorrelated kernel density estimate
+                   akde = akde(data = d, CTMM = model[[1]]) %>% list(),
+                   # find home range estimate
+                   hr_est_50 = extract_hr(a = akde[[1]], par='est', l.ud=0.50),
+                   hr_lwr_50 = extract_hr(a = akde[[1]], par='low', l.ud=0.50),
[…]

### 2022-11-03 writing/appendix-2-empirical-modeling.Rmd
corrected typo, fixed YAML header

@@
 ---
-title: "Appendix 2: Empirical modeling"
+title: \large How resource abundance and stochasticity affect animals' spatial needs
+subtitle: "Appendix 2: Empirical modeling"
 author:
@@
 subparagraph: true # needed for \usepackage[small]{titlesec}
-linenumbers: true
 output:
@@
 ```{r, echo=FALSE, eval=TRUE}
-m_ndvi <- readRDS('../models/CE_31_ANNA-mgcv-ndvi-gulss.rds')
+m_ndvi <- readRDS('../models/CE_31_ANNA-mgcv-ndvi-gaulss.rds')
 plot(m_ndvi, pages = 1, scheme = 1, n = 250)

### 2022-11-03 writing/appendix-1-simulations.Rmd
fixed commenting issue, changed w Kim's comments

@@
 ---
-title: "Appendix 1: Simulations"
+title: \large How resource abundance and stochasticity affect animals' spatial needs
+subtitle: "Appendix 1: Simulations"
 author:
@@
     institute: ubco
+  - name: Emilia Patrícia Medici
+    institute: [tapir, iucn, escas]
   - name: Michael J. Noonan
@@
   - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
 fontsize: 12pt
-indent: true
+header-includes:
+    - \usepackage{indentfirst} # for indenting first line of each paragraph
+    - \usepackage[small]{titlesec} # for smaller font for headings
+    - \usepackage{caption} # for more customization of captions
+    # smaller caption text, less space above and below caption
+    - \captionsetup[figure]{font=scriptsize}
+subparagraph: true # needed for \usepackage[small]{titlesec}
 output:
@@
 ```{r, echo=FALSE}
-knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE)
+knitr::opts_chunk$set(eval = FALSE,    # do not run R code
+                      echo = TRUE,     # print R code
+                      message = FALSE, # hide messages
+                      comment = '')    # don't add anything before outputs
 ```
@@
 * change all references to scripts to the correct file locations
-* give an overview of how the sims work at the beginning of the pdf
+* give an overview of how the sims work at the beginning of the pdf (including defining "encounters" with resources)
 * explain the checks in `delta-t-sensitivity.R`
 * reduce scaling of figure 2
+* fix captions (incomplete, maybe incorrect def of sims, HR not V(pos))
 
@@
 
-# Overview {-}
+# Overview
 
-This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript. For ease of reference, we also include the figures here:
+This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript.
+For ease of reference, we also include the figures here (Figures \@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
+To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
+Instead, the R Markdown document (`writing/appendix-1-simulations.Rmd`) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file.
+Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes and as a general guide.
+We suggest using the `R` scripts to replicate the simulations, instead.
 
@@
 The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase approximately linearly as $\\mathbb V(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
-The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
+Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases.
+Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 100 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
+The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
 knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+```
+
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
+The relationships were estimated using a Generalized Linear Model with a Gamma family.
+The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile.
+Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+knitr::include_graphics('../figures/simulation-regression-plots.png')
 ```
@@
 
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile.
-Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-plots.png')
+# Simulating the movement tracks
+
+To reduce sampling variance between simulations, we use the same set of simulated tracks for each time point in each panel (figure \@ref(fig:sims-example)).
+In the `analysis/simulations/tracks.R` script, we generate $2^{10} = 1024$ tracks to check how many tracks are necessary to obtain stable home range estimates in the best- and worst-case scenarios.
+Most intermediate and diagnostic checks included in the `R` scripts are not included in this document for the sake of brevity and simplicity.
+
+```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' spatial needs were simulated.
+(a.) Animal tracks were simulated using an infinitely diffusive movement model starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
+(b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$.
+(c.) Each time the animal collected more resources, its satiety (purple) increased.
+Once the animal collected sufficient resources, the animal stopped moving (i.e., the track was truncated).
+(d.) The process was repeated 100 times (13 tracks pictured in this panel).
+(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% and 50\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
+knitr::include_graphics('../figures/simulation-example.png')
 ```
 
-## explain how the sims work
-
-To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
-Instead, the R Markdown document ($\tt{writing/appendix-1-simulations.Rmd}$) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file.
-Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes and as a general guide.
-To replicate the simulations, we suggest using the `R` scripts, instead.
-
-# Simulating the movement tracks
-
-To reduce sampling variance between simulations, we use the same set of simulated tracks for each time point in each panel.
-In the $\tt{analysis/simulations/tracks.R}$ script, we generate $2^{10} = 1024$ tracks to check how many tracks are necessary to obtain stable home range estimates in the best- and worst-case scenarios.
-Most intermediate and diagnostic checks included in the `R` scripts are not included in this document for the sake of brevity and simplicity.
-
-```{r load-packages, eval=TRUE, comment=FALSE}
+```{r simulate-tracks, eval=TRUE}
 # NOTE: change your working directory to be "env-var-review/writing", or
@@
 model <- ctmm(tau = c(Inf, 1e3), sigma = 0.1, mu = c(0, 0))
-# summary(model) # FAILS TO KNIT BECAUSE OF TAU CHARACTER # check the parameters of the model
-```
 
-```{r simulate-tracks}
 N_DAYS <- 2^10 # number of "days" (i.e., tracks with different seeds)
@@
                            .f = get_tracks)) # function to generate tracks
+tracks
 
@@
 tracks <- tidyr::unnest(tracks, track)
+tracks
 ```
 
-After generating the tracks, we performed the following tests to ensure the tracks were appropriate and results would be stable:
+After generating the tracks, we performed the following tests to ensure the number and length of the tracks were large enough for results to be stable.
+For the sake of conciseness, the code for each of the checks is not presented in this appendix, but it is available in the `R` scripts referenced in each section.
 
@@
 
-This check was performed with the code in the $\tt{analysis/figures/simulations/return-sensitivity.R}$ script.
+Script: `analysis/figures/simulations/return-sensitivity.R`
+
 Adding return trips to $\langle0, 0\rangle$ after an animal reached satiety doubled computational times without appreciable improvements on the home range estimates (including the 95% CIs).
@@
 
-This check was performed with the code in the $\tt{analysis/figures/simulations/delta-t-sensitivity.R}$ script. ***explain what I did***
+Script: `analysis/figures/simulations/delta-t-sensitivity.R`
 
-```{r dt-sensitivity, fig.cap="Diagnostic tests used to decide an appropriate sampling interval.
-(a.) Histograms of the number of encounters as a function of the interval between envounters, with a binwith of 30 seconds.
+Using three tracks generated using three arbitrary seeds (1, 2, and 3), we explored the effects of sampling interval on the number of encounters with food (i.e., movements to new cells) detected.
+From each of the four checks, we created an exploratory plot that we present in figure \@ref(fig:dt-sensitivity).
+
+**Exploratory plot \@ref(fig:dt-sensitivity)a.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
+Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.
+
+**Exploratory plot \@ref(fig:dt-sensitivity)b.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$
+
+**Exploratory plot \@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks while avoiding excessive amounts of data and an inflated amount of encounters with resources while an animal was near cell boundaries.
+
+**Exploratory plot \@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
+All three tracks have a reasonable amount of tortuous movement and directed movement.
+
+```{r dt-sensitivity, fig.cap="Exploratory plots used to decide an appropriate sampling interval.
+(a.) Histograms of the number of encounters as a function of the interval between encounters, with a binwith of 30 seconds.
 Although some encounters occur with less than 30 seconds between them, most of them occur at least 60 seconds apart.
@@
 Red dots indicate samples where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell, and thus encountered food.
-While the number of feeding events detected decreases as the sampling interval increases, most of the feeding events lost at $\\Delta t = 30$s occured because the animal remained almost adjacent to the borders between cells.
+While the number of encounters detected decreases as the sampling interval increases, most of the encounters lost at $\\Delta t = 30$s occured because the animal remained almost adjacent to the borders between cells.
 (d.) The three tracks used in these tests over the raster used to determine when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
@@
 
+\newpage
+
 ## Checking how many tracks were necessary
 
-This check was performed with the code in the $\tt{analysis/simulations/hr-simulation-extreme-scenarios.R}$ script. ***explain, add figures***
+Script: `analysis/hr-simulation-extreme-scenarios.R`
+
+```{r}
+library('ctmm')    # for continuous-time movement modeling
+library('dplyr')   # for data wrangling (e.g., %>%)
+library('tidyr')   # for data wrangling (e.g., nested tibbles)
+library('ggplot2') # for fancy plots
+source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
+source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/simulations/movement-model.R') # for consistency across scripts
+source('functions/get_hr.R') # for extracting gaussian home range
+theme_set(theme_bw())
+set.seed(1) # for consistent results
+
+tels <- readRDS('simulations/tracks.rds') # list of telemetry tracks
+tracks <- readRDS('simulations/labelled-tracks.rds') # tibble of tracks
+MAX_T <- max(tracks$t) # maximum amount of exploration time
+
+WORST <- filter(d55, mu == min(mu)) %>% # lowest mean resources
+  filter(sigma2 == max(sigma2)) %>% # with highest variance
+  slice(1) # take the first row only
+BEST <- filter(d55, mu == max(mu)) %>% # highest mean resources
+  filter(sigma2 == min(sigma2)) %>% # with lowest variance
+  slice(1) # take the first row only
+
+days <-
+  # modify WORST and BEST to follow the syntax used in 'offspring-simulations.R'
+  transmute(bind_rows(WORST, BEST),
+            animal,
+            mu,
+            sigma2,
+            d = list(tracks),
+            scenario = c('Worst case', 'Best case')) %>%
+  unnest(d) %>% # unnest the datasets so we have a single, large tibble
+  select(-timestamp) %>%
+  # generate the food for each row from a gamma distribution
+  mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
+         # if the animal visits a new cell, it finds food, otherwise it doesn't
+         food = if_else(new_cell, food, 0)) %>%
+  # end the movement once the animal has reached satiety
+  group_by(day, animal, scenario) %>%
+  # calculate the total number of visits, total calories, and if animal is full
+  mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
+         full = satiety >= REQUIRED) %>% # did the animal reach the its needs?
+  filter(cumsum(full) <= 1) %>% # full only once
+  ungroup()
+
+if(FALSE) {
+  # check if the ends of each day are correct and make sense
+  days_end <-
+    days %>%
+    group_by(scenario, day) %>%
+    filter(full, ! duplicated(full)) %>% # take the 1st row where animal is full
+    rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
+    # remove unneded columns (also avoids duplicated colnames with tracks)
+    dplyr::select(-c(x, y, vx, vy, longitude, latitude, food))
+
+  # check max fraction of time used (should be < 1)
+  max(days_end$t_expl) / MAX_T
+
+  # are animals are full only once/day? (should be == 1)
+  sum(days_end$full) / (max(days$day) * 2)
+
+  # plot of satiety over time by animal
+  ggplot(days, aes(t, satiety, group = day)) +
+    facet_wrap(~ scenario) +
+    geom_line(alpha = 0.05) +
+    geom_point(aes(t_expl), days_end, alpha = 0.1) +
+    geom_hline(yintercept = REQUIRED, color = 'red') +
+    geom_vline(xintercept = MAX_T, color = 'blue')
+
+  # check distribution of animals
+  ggplot(days_end, aes(scenario, t_expl)) +
+    geom_hline(yintercept = MAX_T, color = 'red') +
+    geom_violin(fill = 'forestgreen', alpha = 0.3) +
+    geom_boxplot(fill = NA) +
+    labs(x = '', y = 'Exploration time')
+
+  # check home ranges of animals
+  ggplot(days) +
+    facet_grid(. ~ scenario) +
+    coord_equal() +
+    geom_hex(aes(longitude, latitude)) +
+    scale_fill_distiller('Count', type = 'seq', na.value = 'transparent') +
+    theme(legend.position = 'top')
+}
+
+# single estimates that eventually converge to the asymptote ----
+days_summarized <-
+  days %>%
+  # find how long it took to reach satiety
+  group_by(scenario, day) %>%
+  nest(tel_day = -c(scenario, day)) %>%
+  mutate(t_expl = map_dbl(tel_day, \(d) max(d$t))) %>%
+  # add days sequentially
+  group_by(scenario) %>%
+  mutate(t_start = lag(2 * t_expl), # add the return time before next "day"
+         t_start = if_else(is.na(t_start), 0, t_start), # start at 0, not NA
+         t_start = cumsum(t_start), # make start times comsecutive
+         tel_day = map2(day, t_expl,
+                        \(i, te) tels$tel[[i]] %>% # extract tel for the day
+                          data.frame() %>% # for filtering
+                          filter(t <= te))) %>% # end tracks at satiety
+  unnest(tel_day) %>% # make one big dataset
+  mutate(t = t + t_start, # make times consecutive
+         individual.local.identifier = scenario, # ctmm identifier
+         timestamp = as.POSIXct(t, origin = '2000-01-01')) %>% # use new times 
+  ungroup() # remove grouping by scenario
+
+if(FALSE) {
+  # check times are adding up correctly
+  # best case should require less time
+  # blue and red lines should have same length (within the pair)
+  # black lines should be horizontal
+  days_summarized %>%
+    filter(day <= 10) %>%
+    ggplot(aes(day, timestamp)) +
+    facet_wrap(~ scenario, scales = 'free_y') +
+    geom_line() +
+    geom_line(aes(group = day), color = 'blue', lwd = 30) +
+    geom_line(aes(day, timestamp + t_expl, group = day), color = 'red', lwd =30)
+
+  days_summarized %>%
+    filter(day <= 10) %>%
+    ggplot() +
+    facet_wrap(~ scenario) +
+    coord_equal() +
+    geom_path(aes(x, y, group = day), alpha = 0.5) +
+    geom_point(aes(0, 0)) +
+    geom_point(aes(x, y), filter(days_summarized, day <= 10, t == 0),
+               color = 'red')
+}
+
+# estimate saturation curve of home range size over number of days
+saturation_days <-
+  expand_grid(n_days = (2^seq(1, log2(1e3), by = 0.2)) %>% round() %>% unique(),
+              case = unique(days_summarized$scenario)) %>%
+  mutate(data = map2(n_days, case,
+                     \(.n, .case) filter(days_summarized,
+                                         day <= .n,
+                                         scenario == .case)),
+         tel = map(data, as.telemetry), # convert data to telemetry for modeling
+         theta = map(tel, \(x) ctmm.guess(data = x, interactive = FALSE)),
+         m = map(1:n(), \(i) {
+           cat('Fitting model', i, '\n')
+           ctmm.fit(tel[[i]], theta[[i]])
+           }), # fit movement model
+         sigma = map_dbl(m, \(.m) ctmm:::area.covm(.m$sigma)), # var(position)
+         hr = get_hr(.sigma = sigma, quantile = 0.95)) # Gaussian home range
+
+saveRDS(saturation_days, 'simulations/hr-saturation-days.rds')
+
+saturation_days %>%
+  select(case, n_days, sigma, hr) %>%
+  readr::write_csv('simulations/hr-saturation-days.csv')
+
+p_hr_days <-
+  ggplot(saturation_days, aes(n_days, hr)) +
+  facet_wrap(~ case, nrow = 1) +
+  geom_vline(xintercept = 100, color = 'darkorange') +
+  geom_smooth(method = 'gam', color = 'black',
+              formula = y ~ s(x, bs = 'cs', k = 10),
+              method.args = list(family = Gamma(link = 'log'))) +
+  geom_point(alpha = 0.3) +
+  scale_x_continuous(expression(Number~of~days~sampled~(log[2]~scale)),
+                     trans = 'log2', breaks = c(2, 16, 128, 1024),
+                     limits = c(2, 1100)) +
+  scale_y_log10(expression(Estimated~home~range~(log[10]~scale))); p_hr_days
+```
+
+```{r, eval=TRUE, echo=FALSE, fig.cap="Caption."}
+knitr::include_graphics('../figures/hr-over-days.png')
+```
 

### 2022-11-05 writing/appendix-2-empirical-modeling.Rmd
completed all sections, need to check scripts

@@
 fontsize: 12pt
-indent: true
 header-includes:
@@
 ```{r, echo=FALSE}
-knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, fig.align = 'center')
+knitr::opts_chunk$set(eval = FALSE,         # do not run R code
+                      echo = TRUE,          # print R code
+                      message = FALSE,      # hide messages
+                      comment = '',         # don't add anything before outputs
+                      fig.align = 'center', # change figure alignment
+                      cache = TRUE)         # cache all chunk outputs
 ```
@@
 * add scripts to the correct folders and check
-* add "for more info, see the tapir project" at the beginning
-* add refs for each package used
 
@@
 
-This appendix illustrates the steps necessary to reproduce the tapir movement analysis and figures in the main manuscript. For ease of reference, we also include the figure here:
+This appendix illustrates the steps necessary to reproduce the tapir movement analysis and figures in the main manuscript.
+The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at https://github.com/StefanoMezzini/tapirs.
+For ease of reference, we also include the figure below (figure \@ref(fig:mw-examples)).
 
-```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
+```{r mw-example, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
 (a.) GPS tracking data of the tapir.
@@
 Still, those interested in replicating the analysis can do so by using the code in the pdf document or the related `R` markdown (`Rmd`) document.
-Listed below are all the packages and source scripts required to run the analyses in this document:
+Listed below are all the packages and source scripts required to run the analyses in this document.
+For spatial data, we use the `MODIStsp` package [version 2.0.9, @busetto_modistsp_2016] to download the NDVI rasters, the `raster` package [version 3.6-3, @hijmans_raster_2022] to work with the NDVI rasters, and the `sf` package [version 1.0-8, @pebesma_simple_2018].
+We use the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling and the `lubridate` package [version 1.8.0, @grolemund_dates_2011] for converting calendar dates to decimal dates.
+Finally, we used the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] and the `mgcv` package [version 1.8-41, @wood_generalized_2017] for modeling and the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
 
 ```{r, eval=TRUE, comment=FALSE}
+# NOTE: change your working directory to be "env-var-review/writing", or
+# modify all file paths as needed
+
 # attach all necessary packages
@@
 ```
+
+\newpage
 
@@
 
+\newpage
+
 # Modeling $\e(R)$ and $\var(R)$ over time {#ndvi}
 
-To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli-ref].
+To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli_using_2005; @pettorelli_normalized_2011].
 We downloaded the data using the `MODIStsp` `R` package with the following code:
@@
 
-Ideally, one should model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1.
+Next, we estimate the mean and variance in NDVI using a Generalized Additive Model for location and scale (GAMLS) [@wood_generalized_2017; @simpson_modelling_2018; @stasinopoulos_generalized_2007].
+Ideally, we would model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1.
 The beta family would be appropriate after applying the linear transformation $$y^* = \frac{y + 1}{2},$$ where $y$ is the original NDVI value (between -1 and 1) while $y^*$ is the NDVI value scaled between 0 and 1.
-However, the `mgcv` package [@ref] does not allow one to model changes in variance in NDVI using a location-scale model with a beta family.
-Although the `brms` package allows one to fit beta location-scale models [@ref], the computational costs of using a fully Bayesian approach can be prohibitive for large datasets.
+However, the `mgcv` package [@wood_generalized_2017] does not allow one to model changes in variance in NDVI using a location-scale model with a beta family.
+Although the `brms` package allows one to fit beta location-scale models [@burkner_brms_2017; @burkner_advanced_2018], the computational costs of using a fully Bayesian approach can be prohibitive for large datasets.
 Therefore, we decided to use a location-scale Gaussian model via the `mgcv` package (`family = gaulss()` in the code chunk below).
@@
 
-# plot smooths (predictions are on the link scales!)
-plot(m_ndvi, pages = 1, scheme = 1, n = 250)
+# plot smooths (note: predictions are on the link scales!)
+plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
 ```
@@
 m_ndvi <- readRDS('../models/CE_31_ANNA-mgcv-ndvi-gaulss.rds')
-plot(m_ndvi, pages = 1, scheme = 1, n = 250)
+plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
 ```
+
+Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term.
+While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial.
+Excessively smooth terms for the mean can inflate the variance (figure \@ref(fig:undersmooth)), while excessively wiggly smooth terms for the mean can cause the variance to be under-estimated (figure \@ref(fig:oversmooth)).
+Ultimately, each of the `k` values should be decided while in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
+If one is unsure where to start from, keeping the `k` for the scale smooth terms below half the `k` for the mean smooth terms is a good starting point.
+Additionally, note that computation time depends strongly on the complexity of the two-dimensional spatial terms, since changes in `k` alter the size of the term quadratically (since the total size is $\tt k \times \tt k = \tt k^2$).
+
+```{r}
+gam(list(
+  # mean predictor
+  ndvi ~
+    s(long, lat, bs = 'ds', k = 10) + # default k (too small)
+    s(dec_date, bs = 'tp', k = 10),
+  # precision (1/standard deviation) predictor
+  ~
+    s(long, lat, bs = 'ds', k = 20) + # excessively high k
+    s(dec_date, bs = 'tp', k = 20)),
+  family = gaulss(b = 0.0001),
+  data = anna_ndvi,
+  method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+```{r}
+gam(list(
+  # mean predictor
+  ndvi ~
+    s(long, lat, bs = 'ds', k = 50) + # k too large
+    s(dec_date, bs = 'tp', k = 29), # k too large
+  # precision (1/standard deviation) predictor
+  ~
+    s(long, lat, bs = 'ds', k = 5) + # k too low
+    s(dec_date, bs = 'tp', k = 5)),  # k too low
+  family = gaulss(b = 0.0001),
+  data = anna_ndvi,
+  method = 'REML') %>% # performs better than GCV (the default)
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+<!-- chunks to create the plots -->
+
+```{r undersmooth, fig.cap="Estimated mean and variance in NDVI with excessively small basis size $\\tt{k}$ for the mean smooth terms and excessively large for the variance smooth terms.", echo=FALSE, eval=TRUE}
+gam(list(
+  # mean predictor
+  ndvi ~
+    s(long, lat, bs = 'ds', k = 10) + # default k (too small)
+    s(dec_date, bs = 'tp', k = 10),
+  # precision (1/standard deviation) predictor
+  ~
+    s(long, lat, bs = 'ds', k = 20) + # excessively high k
+    s(dec_date, bs = 'tp', k = 20)),
+  family = gaulss(b = 0.0001), # minimum standard deviation of 0.0001
+  data = anna_ndvi,
+  method = 'REML',
+  control = gam.control(nthreads = 4)) %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+```{r oversmooth, fig.cap="Estimated mean and variance in NDVI with excessively large basis size $\\tt{k}$ for the smooth term of time ($\tt{dec\\_date}$) for the mean in NDVI and excessively low for the variance temporal smooth term.", echo=FALSE, eval=TRUE}
+gam(list(
+  # mean predictor
+  ndvi ~
+    s(long, lat, bs = 'ds', k = 50) +
+    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)), # k too large
+  # precision (1/standard deviation) predictor
+  ~
+    s(long, lat, bs = 'ds', k = 5) + # k too low
+    s(dec_date, bs = 'tp', k = 5)),  # k too low
+  family = gaulss(b = 0.0001),
+  data = anna_ndvi,
+  method = 'REML',
+  control = gam.control(nthreads = 4)) %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+\newpage
 
@@
   ggplot(anna_mw) +
-  geom_line(aes(t_center, hr_est, group = quantile, size = quantile),
+  geom_line(aes(t_center, hr_est, group = quantile, lwd = quantile),
             color = pal[3], show.legend = FALSE) +
@@
 }
-# Draw aligned plots
+# create a panel of the plots
 p_left <- plot_grid(plotlist = grobs, ncol = 1,
@@
 
-## there should be a model for anna instead of `geom_smooth()!
+Finally, we estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use using the `geom_smooth()` function from the `ggplot2` package.
+Although one would ideally model the relationships with a formal model, we use `geom_smooth()` because the sampling of the estiamted means and variance are both *auto*correlated (figure \@ref(fig:)) and correlated, so the risk for biased estimates is high (figure \@ref(fig:mu-s2-rel)).
+
+```{r mu-s2-rel, eval=TRUE, echo=FALSE, fig.cap="Scatterplot of estimated mean NDVI and variance in NDVI.
+Both variables are higly temporally autocorrelated, and there is also strong correlation between the two variables."}
+knitr::include_graphics('H:/GitHub/env-var-review/figures/mean-variance-relationship.png',
+                        rel_path = TRUE)
+```
 
 ```{r}
+# mean
 reg_mu <-
@@
              color = pal[3]) +
-  geom_smooth(aes(mu, hr_est, group = quantile, size = quantile),
+  geom_smooth(aes(mu, hr_est, group = quantile, lwd = quantile),
               color = pal[1], se = FALSE, method = 'gam',
@@
              alpha = 0.3, color = pal[3]) +
-  geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
+  geom_smooth(aes(sigma2, hr_est, group = quantile, lwd = quantile),
               color = pal[2], se = FALSE, method = 'gam',
@@
     ggplot(anna_mw) +
-      geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
+      geom_smooth(aes(sigma2, hr_est, group = quantile, lwd = quantile),
                   se = FALSE, method = 'gam', color = 'black',
@@
 ```
+
+\clearpage
+
+# References {-}

### 2022-11-05 functions/get_hr.R

@@ -1,0 +1,5 @@
+#' function to get home range estimate starting from positional variance (`sigma`)
+#' **NOTE:** only works when AKDE is isotropic and Gaussian!
+#' @param .sigma positional variance estimated by a CTMM model
+#' @param quantile quantile within the interval (0, 1) for which HR should be calculated
+get_hr <- function(.sigma, quantile) -2 * log(1 - quantile) * pi * .sigma

### 2022-11-05 functions/mean-variance-trends.R

@@ -1,0 +1,119 @@
+library('dplyr') # for data wrangling (e.g., %>%) and tibble()
+
+# mean functions
+# mean should be far from zero because a small mean with large variance implies the scale
+# parameter is very large, which results in a heavy right tail and a large range in the
+# simulated values.
+# This issue would be clearer when parameterizing in terms of shape and scale, but the
+# parameterization is often harder to plot and visualize.
+min_mu <- 100
+max_mu <- 200
+
+#' animals reach their `REQUIRED` amount of energy after eating many times: 
+#' `REQUIRED` should be high enough that all animals are forced to eat multiple times
+#' `REQUIRED` should *not* be so high that HR estimates become too variable
+#' the variance in HR depends on the variance in total food, i.e. sum of variances
+REQUIRED <- max_mu * 10
+
+mean_constant <- function(generations) {
+  rep(mean(c(min_mu, max_mu)), length(generations))
+}
+
+mean_linear <- function(generations) {
+  b <- (max_mu - min_mu) / (200 - 1)
+  
+  generations * b + min_mu - b
+}
+
+mean_cyclical <- function(generations) {
+  sin(generations / 10) * (max_mu - min_mu) / 2 + mean(c(min_mu, max_mu))
+}
+
+mean_drifting <- function(generations, seed = 8) {
+  set.seed(seed = seed)
+  mu <- cumsum(rnorm(length(generations)))
+  mu <- mu - min(mu) # set minimum to 0
+  mu <- mu / max(mu) # shrink the range to [0, 1]
+  mu <- mu * (max_mu - min_mu) # scale to the range of mu
+  mu <- mu + min_mu # shift to the right intercept
+  return(mu)
+}
+
+mean_erratic <- function(generations) {
+  case_when(generations < quantile(generations, 0.25)~mean(c(min_mu, max_mu)),
+            generations < quantile(generations, 2/3) ~ min_mu,
+            TRUE ~ max_mu)
+}
+
+# variance functions
+min_s2 <- 50^2
+max_s2 <- 500^2
+
+variance_constant <- function(generations) {
+  rep(mean(c(min_s2, max_s2)), length(generations))
+}
+
+variance_linear <- function(generations) {
+  b <- (max_s2 - min_s2) / (200 - 1)
+  
+  generations * b + min_s2 - b
+}
+
[…]

### 2022-11-05 functions/rgamma2.R

@@ -1,0 +1,8 @@
+# re-parameterize rgamma() as a function of mean and variance
+rgamma2 <- function(mu, sigma2, N = n()) {
+  # mean = k * theta
+  # sigma^2 = k * theta^2
+  rgamma(n = N,
+         shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
+         scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
+}

### 2022-11-05 writing/appendix-1-simulations.Rmd
added pkg refs, switched to absolute fig paths

@@
   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+bibliography: 'env-var-review.bib'
+csl: 'the-american-naturalist.csl'
 fontsize: 12pt
@@
 
-```{r, echo=FALSE}
+```{r, echo=TRUE}
 knitr::opts_chunk$set(eval = FALSE,    # do not run R code
@@
                       message = FALSE, # hide messages
-                      comment = '')    # don't add anything before outputs
+                      comment = '',    # don't add anything before outputs
+                      cache = TRUE)    # cache all chunk outputs
+knitr::opts_knit$set(root.dir = '..')
 ```
@@
 The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+knitr::include_graphics('H:/GitHub/env-var-review/figures/mean-variance-5-by-5-hr-sims.png',
+                        rel_path = FALSE)
 ```
@@
 Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-plots.png')
+knitr::include_graphics('H:/GitHub/env-var-review/figures/simulation-regression-plots.png',
+                        rel_path = FALSE)
 ```
@@
 Most intermediate and diagnostic checks included in the `R` scripts are not included in this document for the sake of brevity and simplicity.
+In this script, we use the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for movement modeling, the `raster` package [version 3.6-3, @hijmans_raster_2022] to work with the simulated resource rasters, the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling, and the the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
 
@@
 (e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% and 50\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
-knitr::include_graphics('../figures/simulation-example.png')
+knitr::include_graphics('H:/GitHub/env-var-review/figures/simulation-example.png',
+                        rel_path = FALSE)
 ```
@@
 ```{r simulate-tracks, eval=TRUE}
-# NOTE: change your working directory to be "env-var-review/writing", or
-# modify all file paths as needed
+# NOTE: change your working directory to "env-var-review" or modify file paths
+
+# attach all necessary packages
 library('ctmm')   # for generating movement models and movement modeling
@@
 library('purrr')  # for functional programming
-source('../functions/label_visits.R') # decides when animal encounters food
+library('tidyr')   # for data wrangling (e.g., nested tibbles)
+library('ggplot2') # for fancy plots
+
+source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
+source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/movement-model.R') # for consistency across scripts
+source('functions/get_hr.R') # for extracting gaussian home range
+source('functions/label_visits.R') # decides when animal encounters food
 
@@
 (d.) The three tracks used in these tests over the raster used to determine when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
-knitr::include_graphics('../figures/thinning-examples.png')
+knitr::include_graphics('H:/GitHub/env-var-review/figures/thinning-examples.png',
+                        rel_path = FALSE)
 ```
@@
 ```{r}
-library('ctmm')    # for continuous-time movement modeling
-library('dplyr')   # for data wrangling (e.g., %>%)
-library('tidyr')   # for data wrangling (e.g., nested tibbles)
-library('ggplot2') # for fancy plots
-source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
-source('analysis/simulations/movement-model.R') # for consistency across scripts
-source('functions/get_hr.R') # for extracting gaussian home range
-theme_set(theme_bw())
 set.seed(1) # for consistent results
@@
 
-```{r, eval=TRUE, echo=FALSE, fig.cap="Caption."}
-knitr::include_graphics('../figures/hr-over-days.png')
+```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated spatial needs as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
+In both cases, 100 days are sufficient to produce stable estimates of spatial needs."}
+knitr::include_graphics('H:/GitHub/env-var-review/figures/hr-over-days.png',
+                        rel_path = FALSE)
 ```
@@
 5. $\tt{analysis/simulations/hr-mean-variance-simulations-hrs.R}$
+
+\clearpage
+
+# References {-}

### 2022-11-05 analysis/figures/delta-t-sensitivity.R
updated dt figure

@@ -1,0 +1,195 @@
+library('raster')  # for working with rasters
+library('ctmm')    # for generating movement models
+library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
+library('tidyr')   # for data wrangling (unnest, etc.)
+library('purrr')   # for functional programming (map_*(), etc.)
+library('ggplot2') # for fancy plots
+library('cowplot') # for multi-panel fancy plots
+source('analysis/simulations/movement-model.R') # for consistency between scripts
+theme_set(theme_bw() + # change default ggplot theme
+            theme(legend.position = 'none', strip.background = element_blank(),
+                  strip.text = element_blank()))
+
+# simulate some movement
+get_tracks <- function(.seed) {
+  simulate(model, # ctmm movement model
+           t = 0:(max(SAMPLES)), # 12 hours of data, sampled every second
+           seed = .seed, # set a seed for consistent results
+           complete = TRUE, # prevents missing projection warning
+           crs = PROJECTION) # projection CRS string
+}
+tels <- tibble(seed = 1:3, tel = map(seed, get_tracks)) %>%
+  mutate(seed = factor(seed))
+
+# plot intervals between encounters
+intervals <-
+  transmute(tels,
+            seed,
+            ints = map(tel, \(.tel) {
+              .tel %>%
+                data.frame() %>%
+                mutate(cell_id = cellFromXY(object = HABITAT,
+                                            xy = SpatialPoints.telemetry(.tel)),
+                       new_cell = c(1, diff(cell_id)) != 0) %>%
+                filter(new_cell) %>%
+                mutate(interval = c(NA, diff(t)))
+            })) %>%
+  unnest(ints)
+
+# check the main quantiles
+quantile(intervals$interval, na.rm = TRUE)
+
+# histogram of number of events as by the time interval between events
+p_hist <-
+  ggplot(intervals, aes(interval)) +
+  facet_wrap(~ seed, ncol = 1) +
+  geom_histogram(aes(fill = seed), binwidth = 30, center = 15, lwd = 0.1,
+                 color = 'black', na.rm = TRUE) +
+  scale_fill_viridis_d('Seed', ) +
[…]

### 2022-11-08 writing/manuscript.Rmd
added some refs

@@
 
-\newpage <!-- start main body on a new page -->
+<!-- add custom commands for formulae -->
+
+\newcommand{\e}{\mathbb E}
+\newcommand{\var}{\mathbb V}
+\newcommand{\vpos}{\mathbb V(\text{position})}
+
+<!-- have table of contents on pages of its own -->
+
+\newpage
 
@@
 
+<!-- start main body on a new page -->
 \newpage
@@
 Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018].
-Thus, many organisms will change their behaviours in response to changes in local resource abundance.
-While some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Thus, many organisms will change their behaviors in response to changes in local resource abundance.
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 
@@
 More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022].
-Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with larger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005].
-Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago various scientists, inlcuding @lindstedt_home_1986 and @reiss_scaling_1988.
-Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
+Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005].
+Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988.
+Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
 
@@
 
-### Expected resource abundance, $\mathbb E(R)$
+### Expected resource abundance, $\e(R)$
 
@@
 its mean or average, indicated here as ) is the value one can *expect* the random variable to take.
-$\mathbb E(R) = \mu$ to indicate the expectation of the random variable $R$.
-The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\mathbb E(R)$ is still useful for estimating the average long-term resource abundance.
-When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\mathbb E(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\mathbb E(R) = \mu(t)$.
+$\e(R) = \mu$ to indicate the expectation of the random variable $R$.
+The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance.
+When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.
 
@@
 
-### Variance in resource abundance, $\mathbb V(R)$
+### Variance in resource abundance, $\var(R)$
 
 A random variable's variance is a measure of its unpredictability or variability.
-We use  $\mathbb V(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$.
-For instance, while one may consider $R$ to vary within years, we define $\mathbb V(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others.
+We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$.
+For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others.
 Additionally, we distinguish between two different sources of variation.
@@
 $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
-Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters.
+Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters.
 Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].
@@
 The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Thus, here we quantify animal's spatial needs using their positional variance (indicated as $\mathbb V(\text{position})$), instead.
-***explain the concept of positional variance*** [@refs].
+Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the variance in position of animals that are not range-resident [@refs].
 
-Although range-resident animals will generally have smaller $\mathbb V(\text{position})$ than migratory and nomadic animals, $\mathbb V(\text{position})$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\mathbb V(\text{position})$ determines the scale and size of an animal's spatial needs but not its behavior.
+Although range-resident animals will generally have smaller $\vpos$ than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior.
 Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement.
@@
 Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
+Animals may temporarily change their behavior in response to many causes, including predators [@ref], competitors [@ref], weather [@ref], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
 
@@
 ### Unassigned
+
+@boratynski_energetic_2020: animals' hr is limited by BMI and maximum metabolic rate, which can be accounted for by accounting for body mass [@noonan_effects_2020]
+
+@arechavala-lopez_common_2019: octopuses (*Octopus vulgaris*) congregate in human-altered coastal habitats
+
+@mitsuhashi_home_2018: urban raccoon dogs (*Nyctereutes procyonoides*) have a smaller HR than rural ones
+
+@cronk_home_2021: yellow mongoose (*Cynictis penicillata*) in urban regions have smaller HRs
 
@@
 
-```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of mean resource abundance ($\\mathbb E(R)$).
-We expect low values of $\\mathbb E(R)$ to result in high $\\mathbb V(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\mathbb E(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
-Note that the relationship between $\\mathbb E(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for high values of $\\mathbb E(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of mean resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high $\\var(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
+Note that the relationship between $\\e(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)].
-Low values of $\mathbb E(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
-As $\mathbb E(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
+Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
+As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
 
@@
 
-```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\mathbb V(\\text{position})$) of an animal as a function of resource unpredictability ($\\mathbb V(R)$).
-We expect low values of $\\mathbb V(R)$ to result in low $\\mathbb V(\\text{position})$ as animals are able to depend on somewhat predictable resources.
-As $\\mathbb V(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\mathbb V(R)$ and $\\mathbb V(\\text{position})$ cannot be linear because it would require $\\mathbb V(\\text{position})$ to be negative for low values of $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of resource unpredictability ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low $\\var(\\text{position})$ as animals are able to depend on somewhat predictable resources.
+As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
+Note that the relationship between $\\var(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\mathbb V(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\mathbb V(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
+Although it generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
+
+\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
 Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
@@
 
-Simlarly to the effect of $\mathbb E(R)$, we expect $\mathbb V(R)$ to have a nonlinear effect on $\mathbb V(\text{position})$.
-When $\mathbb V(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\mathbb V(R)$ to have a noticeable effect on $\mathbb V(\text{position})$.
+Simlarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on $\vpos$.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect on $\vpos$.
 However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly.
@@
 
-The effects of $\mathbb V(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently.
-Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\mathbb V(\text{position})$ may not detect such differences.
+The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently.
+Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\vpos$ may not detect such differences.
 @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
@@
 Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality.
-The work by @nilsen_can_2005 also demonstrates that the effect of $\mathbb E(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\mathbb V(R)$, so it is important to account for interaction effects between the two.
+The work by @nilsen_can_2005 also demonstrates that the effect of $\e(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\var(R)$, so it is important to account for interaction effects between the two.
 
-* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\mathbb V(R)$
+* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\var(R)$
 
@@
 (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and average $\\mathbb V(R)$.
-(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and average $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+(c.) Simulated animal movement at a time of lowest $\\e(R)$ and average $\\var(R)$.
+(d.) Simulated animal movement at a time of peak $\\e(R)$ and average $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
@@
 
-\noindent Consider the simplistic scenario where $\mathbb E(R) = \mu$ changes over time but $\mathbb V(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
+\noindent Consider the simplistic scenario where $\e(R) = \mu$ changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
 We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
 Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
-However, if $\mathbb E(R) = \mu(t)$ changes repetitively and regularly over time (e.g.
+However, if $\e(R) = \mu(t)$ changes repetitively and regularly over time (e.g.
 peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019].
-However, since $\mathbb E(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.
+However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.
 
@@
 When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs].
-Most often, $\mathbb E(R)$ will vary over time, and animals' behaviors will change in response.
-Additionally, when data availability is sufficiently high to produce precise estimates of $\mathbb V(R)$, the effect of $\mathbb V(R)$ on animals' $\mathbb V(\text{position})$ should be estimated before assuming the effect is negligible.
+Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
+Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible.
 
@@
 (b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\mathbb E(R)$ and $\\mathbb V(R)$.
-(d.) Simulated animal movement at a time of peak $\\mathbb E(R)$ and $\\mathbb V(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+(c.) Simulated animal movement at a time of lowest $\\e(R)$ and $\\var(R)$.
+(d.) Simulated animal movement at a time of peak $\\e(R)$ and $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
@@
 
-When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\mathbb V(R)$ will have an appreciable effect on the location's favourableness.
-Since $\mathbb V(R)$ is not constant over time, we can now define it as a function of time using the notation $\mathbb V(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
+When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\var(R)$ will have an appreciable effect on the location's favourableness.
+Since $\var(R)$ is not constant over time, we can now define it as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
 In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
@@
 Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+
+# Interactive effects of mean and variance in resource abundance
 
@@
 
-Most habitats with seasonal changes in climate will require models to account for changes in both $\mathbb E(R)$ and $\mathbb V(R)$, since periods of high productivity can often be more variable than times of "dormancy" or lower productivity, and the variability in $R$ often changes both within and between years.
-In such cases, failing to account for $\mathbb V(R)$ can result in serious biases in the estimation of animals' spatial needs.
+Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of "dormancy" or lower productivity, and the variability in $R$ often changes both within and between years.
+In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs.
 
-\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\mathbb E(R)$ and $\mathbb V(R)$.
-The top row (constant $\mathbb V(R)$) shows how  remains constant.
-Some noise in $H$ is still present since $\mathbb V(R) \ne 0$ and thus $R$ is not constant).
-Note the nonlinear decrease in $H$ when $\mathbb E(R)$ increases at a constant slope (linear $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), with larger changes when $\mathbb E(R)$ is low, since the multiplicative changes in $R$ are larger.
-Additionally, the 95% home range is more sensitive to changes in $\mathbb E(R)$ than the core home range $H_{50\%}$.
-In regions where the average resource abundance changes over time (cyclical $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mathbb E(R)$.
-However, estimates of spatial requirements based on estimated changes in $\mathbb E(R)$ should be interpreted carefully, since model error and unpredictable changes in $\mathbb E(R)$ may increase animals' spatial needs unpredictably.
+\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
+The top row (constant $\var(R)$) shows how  remains constant.
+Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant).
+Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$<!--, constant $\var(R)$-->), with larger changes when $\e(R)$ is low, since the multiplicative changes in $R$ are larger.
+Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$.
+In regions where the average resource abundance changes over time (cyclical $\e(R)$<!--, constant $\var(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$.
+However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably.
 Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), since accurate long-term estimates of $\mathbb E(R)$ may be hard to produce, if not impossible.
-In cases where $\mathbb E(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often.
-Finally, in ecosystems where $\mathbb E(R)$ changes erratically and suddenly (erratic $\mathbb E(R)$<!--, constant $\mathbb V(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$<!--, constant $\var(R)$-->), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
+In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often.
+Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$<!--, constant $\var(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
 
@@
 The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\mathbb E(R)$ increases, and they increase approximately linearly as $\\mathbb V(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\mathbb V(R)$ is higher, and changes in $\\mathbb V(R)$ have greater impacts when $\\mathbb E(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
+Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases.
+Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
 The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\mathbb E(H)$) illustrates how an animal's home range may vary as $\mathbb V(R)$ changes over time while $\mathbb E(R)$ remains constant.
-Both $\mathbb E(H)$ and $\mathbb V(H)$ increase with $\mathbb V(R)$, although $H_{95\%}$ is more sensitive to changes in $\mathbb V(R)$ than $H_{50\%}$.
-Additionally, the effect of $\mathbb V(R)$ on $H$ depends on $\mathbb E(R)$, since the effect of $\mathbb V(R)$ on $H$ is stronger when $\mathbb E(R)$ is low (see the bottom row with erratic $\mathbb V(R)$).
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(H)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant.
+Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$.
+Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).
 
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\mathbb E(R)$ and $\mathbb V(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
-$\mathbb E(R)$ and $\mathbb V(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\mathbb E(R)$ and $\mathbb V(R)$ remain undetected.
-Although it is impossible for $\mathbb E(R)$ and $\mathbb V(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease).
-Additionally, these examples are important because they demonstrate the relationships between $H$, $\mathbb E(R)$, and $\mathbb V(R)$ in (relatively) simple scenarios.
-Cyclical oscillations in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\mathbb E(R)$ and $\mathbb V(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
-Finally, erratic changes in $\mathbb E(R)$ and $\mathbb V(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
-However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\mathbb E(R)$ or increases in $\mathbb V(R)$ rather a series of small, sudden, changes.
-Estimating the true trend in $\mathbb E(R)$ and $\mathbb V(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\mathbb E(R)$ are not due to $\mathbb V(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\mathbb E(R)$ and $\mathbb V(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.
+Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
+$\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected.
+Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease).
+Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios.
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
+Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
+However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes.
+Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.
 <!-- https://doi.org/10.1080/00207549508930162 ?-->
+
+[*Needs for understanding effects of variance*] are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
 
@@
 
-\noindent @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
+\noindent
+
+@nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
 However, the effect of many other factors on HR size likely depends on the species, diet type [@harestad_home_1979], location, and possibly also on the individual animal.
@@
 
-* failing to account for $\mathbb V(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
+* failing to account for $\var(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
 * find examples of extreme events that caused population collapses or mass die-offs
@@
 
+## Applying these methods
+
+* The hypotheses we present here allow researchers to combine findings and generalize
+* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
+* can present results building on previous work rather than listing them as *de novo* findings
+* additionally, can test for common behaviors between members of the same population or species
+* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(R)$ or $\var(R)$
+* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior
+
 \newpage
@@
 
-notation:
-- $\mu$ mean movement process
-- $\epsilon$ movement noise, exploratory movements
-- $\tau_p$ positional autocorrelation, mean time reversion to the mean (origin), range crossing time
-- $\tau_v$ velocity autocorrelation
-
 @rickbeil_plasticity_2019:
@@
 - there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
-- ***Elk departed earlier from highly-productive areas?
-This goes against the idea that animals move as a consequence of need.
-But maybe the elk simply left when they had enough stored energy to face the migration.
-Maybe elk migration is triggered when they have sufficient energy (if the weather is still not too harsh).
 -->
@@
 
-<!-- see https://doi.org/10.1186/s40462-019-0187-z regarding HR, energetics, and movement along paths -->
-
 # Conclusion
@@
 \noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance.
-We provide realistic and flexible hypotheses of the effects of $\mathbb E(R)$ and $\mathbb V(R)$ on animals' spatial needs and movement behavior.
-We demonstrate that animals' spatial needs respond nonlinearly to both $\mathbb E(R)$ and $\mathbb V(R)$, and we demonstrate the importance of accounting for $\mathbb V(R)$ besides $\mathbb E(R)$.
+We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
+We demonstrate that animals' spatial needs respond nonlinearly to both $\e(R)$ and $\var(R)$, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
 We also provide a framework for modeling animal behavior and movement continuously rather than imposing arbitrary thresholds between range-resident and nomadic animals.
@@
 @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck].
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck] or fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
 @merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
@@
 # Extra notes
+
+@polansky_elucidating_2015 as example of memory and changes in movement type (exploration vs stationary)
 

### 2022-11-08 writing/manuscript.Rmd
added refs

@@
 \noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce.
-Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019], or even death [@foley_severe_2008; @berger_climate_2018].
+Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], or even death [@foley_severe_2008; @berger_climate_2018].
 Thus, many organisms will change their behaviors in response to changes in local resource abundance.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 
@@
 More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022].
-Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., @herfindal_prey_2005; @nilsen_can_2005].
+Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020].
 Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988.
@@
 @cronk_home_2021: yellow mongoose (*Cynictis penicillata*) in urban regions have smaller HRs
+
+@yu_artificial_2022: rat snakes move to artificial structures for better prey availability and better thermoregulation
+
+Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
 
@@
 (f.) Effect of resource unpredictability on home range size.
-The effects in panels e and f were estimated using generalized linear models with Gamma cinditional distributions.
+The effects in panels e and f were estimated using generalized linear models with Gamma conditional distributions.
 The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}

### 2022-11-10 writing/manuscript.Rmd
cleaned some refs

@@
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels, @refs), doing so with sufficient frequency and accuracy would not be simple.
+The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
 However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly) <!-- since accounting for a proxy of resource abundance will also account for the effects of many other drivers that depend strongly on resource abundance (e.g., competition, predation)-->.
@@
 The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the variance in position of animals that are not range-resident [@refs].
+Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident.
 
@@
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [e.g., @nilsen_can_2005; @ex; see figure \@ref(fig:mu-hyp)].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
 Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
@@
 
+Animal's spatial needs are often estimated using all available tracking data.
+Doing so maximizes the (effective) sample size, but it also assumes that during the tracking period the animal's movement behavior did not change and that its home range did not shift either.
+Although these assumptions may be valid for relatively short tracking periods (and often necessary for reasonable estimates of spatial needs), few animals will behave consistently during the tracking period and use the same space every day.
+***Most animals (if not all) use their space selectively.*** *add examples with humans moving to different regions?*
+
+In the case of animals whose spatial needs change over their lifetimes, one may define HR as a function of time with distinct HRs for each distinct period [e.g., salmon: @ref; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017] or as a continuous function so the HR changes smoothly over time [e.g., elephants: @polansky_elucidating_2015; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021].
+In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
+Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
+
+The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
+
 ```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
@@
 * using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior
+
+## The spatiotemporal scale of stochastic events: memory and adaptability
+
+The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism (i.e.
+approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
+Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
+
+Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
+@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck] or fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
+@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
 
@@
 (notes on seasons and "seasonal behavior")
-
-\newpage
-
-# MISC
-
-In the case of animals whose spatial needs change over the their lifetimes [@refs], one may define HR as a function of time with distinct HRs for each period [e.g., salmon spawning: @ref] or as a continuous function of age [so the HR changes smoothly: @ref].
-In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
-Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
-
-The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
-
-## The spatiotemporal scale of stochastic events: memory and adaptability
-
-The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism (i.e.
-approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
-Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
-
-Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
-@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck] or fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
-@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
 

### 2022-11-14 writing/manuscript.Rmd
resolved conflicts

@@
 
-<!--
-TO DO:
-- write conclusion paragraph
-- write discussion (2-3 pages)
-  - add headers for sections
-- focus on V(R) as temporal variance rather than spatial one. people have already focused on patchiness, but temporal variation is more risky
--->
+<!-- define custom math symbols -->
 
-<!-- add custom commands for formulae -->
+\newcommand*\e{\mathbb E} <!-- expectation (blackboard bold) E -->
 
-\newcommand{\e}{\mathbb E}
-\newcommand{\var}{\mathbb V}
-\newcommand{\vpos}{\mathbb V(\text{position})}
+\newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->
 
-<!-- have table of contents on pages of its own -->
+\newcommand*\vpos{\var (\text{position})} <!-- positional variance -->
 
+<!-- place table of contents on a new page -->
 \newpage
@@
 ```{r setup, include=FALSE}
+# set chunk defaults: don't print code, images 75% of page width and centered
 knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
@@
 <!--
-@ciuti and @berger-tal: human presence in nature for other reasons than hunting
+@ciuti and @berger-tal: human presence in nature for reasons other than hunting
 "When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
@@
 
-<!-- You need a lot more depth here.
-Open with a paragraph on movement generally that leads to the idea that responses to resource are something worth focusing on.
-Follow that up with a paragraph on the history of exploring movement/space use and resources, leading to the concept that, to date, the relationship between abundance and stochasticity have not been fully explored.
-Then finish with a paragraph detailing the aims of the current paper.
--->
-
 \noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce.
@@
 Thus, many organisms will change their behaviors in response to changes in local resource abundance.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 
@@
 Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
+**but animals don't only respond to changes in mean**
+
+**paragraph detailing effects of variance on movement and how little work has been done**
 
@@
 
-The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022].
+The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022].
 This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
@@
 only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906].
-In this case, we can define $R$ as following a Beta distribution using the notation $B (\mu(t), \sigma^2(t))$.
+In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$ (We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is $\mu = \frac{\alpha}{\alpha+\beta}$ while the variance is $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$).
 Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.
@@
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec r_t, \text{and }\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
-However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly) <!-- since accounting for a proxy of resource abundance will also account for the effects of many other drivers that depend strongly on resource abundance (e.g., competition, predation)-->.
+The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
+However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly).
 
@@
 
-# Effects of resource abundance on animals' spatial needs
+# Effects of $\e(R)$
 
-```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of mean resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high $\\var(\\text{position})$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of mean resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high $\\vpos$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
 As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
-Note that the relationship between $\\e(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+Note that the relationship between $\\e(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 \noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
-All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @ref].
+All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref].
 Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death.
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
 Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
@@
 
-# Effects of environmental variance on animals' spatial needs
+# Effects of $\var(R)$
 
-```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\var(\\text{position})$) of an animal as a function of resource unpredictability ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low $\\var(\\text{position})$ as animals are able to depend on somewhat predictable resources.
+```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of resource unpredictability ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low $\\vpos$ as animals are able to depend on somewhat predictable resources.
 As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and $\\var(\\text{position})$ cannot be linear because it would require $\\var(\\text{position})$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+Note that the relationship between $\\var(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-Although it generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
+Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
 
@@
 
-# Interactive effects of mean and variance in resource abundance
+# Interactive effects of $\e(R)$ and $\var(R)$
 
@@
 Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible.
+The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].
 
@@
 
-*Needs for understanding effects of variance* are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years.
+In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].
+In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [@male_deer_growing_antlers].
+In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one.
 
@@
 \noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
-The top row (constant $\var(R)$) shows how  remains constant.
+The top row (constant $\var(R)$) shows how $\vpos$ remains constant.
 Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant).
-Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$<!--, constant $\var(R)$-->), with larger changes when $\e(R)$ is low, since the multiplicative changes in $R$ are larger.
-Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$.
-In regions where the average resource abundance changes over time (cyclical $\e(R)$<!--, constant $\var(R)$-->), animals should have sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$.
+Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low.
+This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g.
+from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
+Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$.
+In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures).
 However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably.
-Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the expected spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$<!--, constant $\var(R)$-->), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
-In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly and often.
-Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$<!--, constant $\var(R)$-->), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
+Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs.
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
+In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time.
+Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
 
@@
 Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
-The animal's spatial variance parameter was then calculated using an OUF model via the $\\tt{ctmm}$ package", out.width='100%'}
+The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
 knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(H)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant.
-Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$.
+Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the core home range ($H_{50\%}$).
 Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).
@@
 Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.
-<!-- https://doi.org/10.1080/00207549508930162 ?-->
 
-[*Needs for understanding effects of variance*] are compounded by climate change, which exposes species to increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Furthermore, anthropogenic structures reduce the habitat available to terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
+The relationships were estimated using a Generalized Linear Model with a Gamma family.
+The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile.
+Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+knitr::include_graphics('../figures/simulation-regression-plots.png')
+```
 
@@
 
-## Moving window examples
+## Applying this framework
+
+* The hypotheses we present here allow researchers to combine findings and generalize
+* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
+* can present results building on previous work rather than listing them as *de novo* findings
+* additionally, can test for common behaviors between members of the same population or species
+* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$
+* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior
+
+### Modeling $R$
+
+* location-scale models
+* `mgcv`, `brms`
+* choosing the best family
+
+### Modeling spatial needs
+
+* importance of AKDE
+* Introduce moving window idea
+
+Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017].
+However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...
+
+### Modeling the relationship
+
+* Gamma distribution for HR
+* possibly location-scale if enough data?
+
+## *Tapirus terrestris* in the Brazilian Pantanal: an empirical example
 
@@
 (f.) Effect of resource unpredictability on home range size.
-The effects in panels e and f were estimated using generalized linear models with Gamma conditional distributions.
+The effects in panels e and f were estimated using independent generalized linear models with Gamma conditional distributions.
 The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
@@
 
+# MISC
+
+In the case of animals whose spatial needs change over the their lifetimes [@refs], one may define HR as a function of time with distinct HRs for each period [e.g., salmon spawning: @ref] or as a continuous function of age [so the HR changes smoothly: @ref].
+In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
+Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
+
+The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
+
+## The spatiotemporal scale of stochastic events: memory and adaptability
+
+The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism (i.e.
+approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
+Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
+
+Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
+@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck].
+@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
+
+\newpage
+
 # Extra notes

### 2022-11-15 writing/manuscript.Rmd
added some refs

@@
 Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020].
-Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists, including @lindstedt_home_1986 and @reiss_scaling_1988.
+Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982].
 Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
@@
 
-The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022].
+The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
 This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
@@
 @jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck] or fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
-@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
+*changes in bird migration, check for changes in mammals*
+
+* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
 

### 2022-11-17 writing/manuscript.Rmd
resolve conflict

--- 2022-11-14 writing/manuscript.Rmd
@@
 
+***Not sure if we need this or if we can tie it in somewhere:***
+
+Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017].
+However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...
+
 ### Advantages of modeling animal movement as a stochastic process
@@
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not monotonic nor linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
 Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
@@
 Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case [@refs; but see @refs].
+When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case.
 Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
@@
 
-<!--
-in stochastic environments:
-- higher spatial needs and possibly store resources for periods of scarcity, which are expected to get longer with climate change
-- heterogeneous environments, such as habitats with a high variety of patches, may force animals to move between patches if they cannot depend on a single one
--->
+Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
+The top row (constant $\var(R)$) shows how $\vpos$ changes as a function of $\e(R)$ while $\var(R)$ remains constant.
+(Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant).
+Note the nonlinear decrease in $H$ as $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(H)$ is low.
+This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g.
+from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
+Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$, since (not sure about this) larger quantiles (e.g., 95% > 50%) are more sensitive to changes in a distribution's variance (in this case $\propto \vpos$).
 
@@
 ```
+
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant.
+Overall, $\vpos$ increases with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the 50% core home range ($H_{50\%}$).
+***here*** <!--Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).-->
 
@@
 
-```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
+```{r mw-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
 (a.) GPS tracking data of the tapir.
@@
 (f.) Effect of resource unpredictability on home range size.
-The effects in panels e and f were estimated using independent generalized linear models with Gamma conditional distributions.
+The effects in panels e and f were estimated using two generalized linear models with Gamma families of distributions.
 The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}

### 2022-11-17 writing/manuscript.Rmd
cleaned sections; added refs

@@
 
+\singlespacing
+
 \tableofcontents
+
+\doublespacing
 
@@
 Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020].
-Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982].
+Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982].
 Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
-**but animals don't only respond to changes in mean**
+And while abundant work has been done on estimating animal's response to changes in resource abundance, there is little to no information on how many species respond to unpredictable changes in the resources they depend on.
 
-**paragraph detailing effects of variance on movement and how little work has been done**
+A habitat's long-term average resource abundance is not sufficient to assess the habitat's quality.
+A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources.
+In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted.
+Although both habitats may have the same long-term resource abundance, the differences in resource predictability result in substantially different habitats.
+Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].
 
@@
 
-In this paper, we can simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
+In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
 Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples.
@@
 Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident.
+***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***
 
@@
 Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
-Animals may temporarily change their behavior in response to many causes, including predators [@ref], competitors [@ref], weather [@ref], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
+Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
 
@@
 ### Unassigned
+
+@pretorius_movement_2020: flamingos -- migrating or nomadic?
 
@@
 
-<!-- You need to open with some background on metabolic requirements.
-Individuals have fixed needs that they must meet in order to survive, grow, and reproduce.
-If these are not met, they will die.
-Then transition to the idea that, because of that, they move around in order to secure resources.
-Therefore, when environments are productive, they don't have to move as far to meet their energetic requirements.
--->
-
-* add refs to generalize beyond terrestrial mammals
-
 \noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
-All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@baldwin_nutritional_1984; @reich_body_2001; @ref].
+All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
 Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism [@ref] and movement costs [@ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @ex].
-Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan \@ref(fig:mu-hyp).
+Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
+Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_20129; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)).
 As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
@@
 
-Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
+\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns as well as infrequent and unpredictable events.
+Examples of predictable changes in $\var(R)$ include changes that occur between seasons, such as changes in temperature in continental summers (e.g., between 15 and 30 $^\circ$C) and winter temperatures (e.g., from -40 to -10$^\circ$C within the span of a few days), changes in precipitation frequency and unpredictability between the dry season (predictable lack of rain) and the wet season (frequent but sudden and unpredictable rains), or changes in fruit and berry abundance with the arrival of spring and summer (since the timing between plants can depend on many unaccounted factors).
+Although each of these events is unpredictable, the change in unpredictability can be predicted and expected.
+In contrast, other sources of stochasticity can be much less predictable, such as fires, floods, droughts, the arrival of new competitors, and changes in climate [@ipcc_global_2018; @noonan_situ_2018].
 
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns (such as daily or seasonal changes in temperature and precipitation, or the location of different patches) as well as infrequent and unpredictable events [such as forest fires, the arrival of new competitors, human activity, fires, floods, droughts, and other consequences of climate change, see @ipcc_global_2018; @noonan_situ_2018].
 Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
@@
 
-Simlarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on $\vpos$.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect on $\vpos$.
-However, as resources become additionally unpredictable, we expect spatial needs to increase superlinearly.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors, including nomadism, dispersal, and migration (figure \@ref(fig:s2-hyp)).
+Similarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)).
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, torpor, and hibernation).
 
-The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs, so the degree to which $R$ varies will likely affects each species differently.
-Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with a larger $\vpos$ may not detect such differences.
+The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs [@boratynski_energetic_2020; @noonan_effects_2020], so the degree to which $R$ varies will likely affects each species differently.
+Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not detect such differences.
 @rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
@@
 In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions.
-Similarly, @nilsen_can_2005 found that the home ranges of wolverines (*Gulo gulo*) and Canadian lynx (*Lynx canadensis*) decreased with seasonality, while the home ranges of bobcats (*Lynx rufus*) and coyotes (*Canis latrans*) increased with seasonality.
-The work by @nilsen_can_2005 also demonstrates that the effect of $\e(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\var(R)$, so it is important to account for interaction effects between the two.
-
-* predation is more stochastic than herbibory; predators need to match more cycles, depend on fewer food sources that are able to move $\rightarrow$ higher effect of $\var(R)$
+Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality.
+A subsequent analysis by @nilsen_can_2005 found that various carnivores respond to resource variability and seasonality (albeit in contrasting ways) using the Canadian lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*).
+Additionally, @nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$, so it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two.
+It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants.
+Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels.
 
@@
 
-\noindent Consider the simplistic scenario where $\e(R) = \mu$ changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that the resource abundance follows a Gamma distribution with a mean that changes over time (i.e., the expected value is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
 Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
-However, if $\e(R) = \mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019].
+However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
+peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019].
 However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.
@@
 Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is rarely the case.
+When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
 Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
-Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' $\vpos$ should be estimated before assuming the effect is negligible.
+Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' spatial needs should be estimated before assuming the effect is negligible.
 The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
 Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
-Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable [@ref] and pose high mortality risks [@noonan_roads_2021; @refs].
+Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
 
@@
 In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].
-In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [@male_deer_growing_antlers].
-In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one.
-
-# Interactive effects of mean and variance in resource abundance
+In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as during pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
+In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one [@lucherini_habitat_1996].
 
 Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
-The top row (constant $\var(R)$) shows how $\vpos$ changes as a function of $\e(R)$ while $\var(R)$ remains constant.
-(Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant).
-Note the nonlinear decrease in $H$ as $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(H)$ is low.
-This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g.
-from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
-Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$, since (not sure about this) larger quantiles (e.g., 95% > 50%) are more sensitive to changes in a distribution's variance (in this case $\propto \vpos$).
-
-Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of "dormancy" or lower productivity, and the variability in $R$ often changes both within and between years.
-In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs.
-
-\noindent Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
 The top row (constant $\var(R)$) shows how $\vpos$ remains constant.
@@
 Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios.
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
 Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
@@
 
+This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
+The hypotheses and methods we present allow researchers to build upon previous findings and datasets rather than stating their results as *de novo* findings.
+Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref].
+Individuals and populations that were once said to have markedly different behaviors and spatial needs may instead have a single common behavior but with different levels of $\e(R)$ and $\var(R)$ [@examples].
+
+In this discussion section, we make considerations about the effects of resource abundance and stochasticity on various adaptations, including movement behaviors, energy storage, and spatiotemporal memory.
+We then illustrate how one may apply this framework to their own data and we also provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
+Finally, we demonstrate how to use this framework to create products that can inform conservation-based decisions, including predicting animals' spatial needs and assessing the quality of potential (and current) conservation areas.
+
+#### Move the following paragraph to previous sections
+
 @nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
 However, the effect of many other factors on HR size likely depends on the species, diet type [@harestad_home_1979], location, and possibly also on the individual animal.
-For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; capuchin monkeys, @torrez-herrera_monkeys_2020].
+For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; Capuchin monkeys, @torrez-herrera_monkeys_2020].
 Similarly, predation may force animals to move more frequently to escape predators, or it may prevent them from venturing too far from the safety of their core HR too often [@suraci_beyond_2022].
@@
 
+## The spatiotemporal scale of stochastic events: memory and adaptability
+
+@ranc_memory_2022: Roe deer create HR based on memory
+
+The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism (i.e.
+approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events [@examples].
+Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@logares_black_2012; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; **mule deer?**: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015; **species?**: @fagan_spatial_2013].
+
+Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
+@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
+*changes in bird migration, check for changes in mammals*
+
+* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
+
+The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism (i.e.
+approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
+Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
+Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
+
+Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
+@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@abrahms_memory_2019].
+@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
+
+<!--
+Notes:
+
+@rickbeil_plasticity_2019:
+- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
+- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
+- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
+-->
+
 ## Applying this framework
@@
 * choosing the best family
+* choosing the right complexity of each term (maybe that's why @nilsen_can_2005 found contrasting results)
 
@@
 
-### Modeling the relationship
+AKDE is good for range-resident animals
+
+In the case of animals whose spatial needs change over their lifetimes [@refs], one may define a HR as a function of time with distinct HRs for each period [salmon: @ref; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017] or as a continuous function of time.
+In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
+Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
+However, when an animal's spatial needs change over time, models which assume range residency are not appropriate.
+Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results.
+Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].
+
+### Modeling the relationships
 
@@
 
-## *Tapirus terrestris* in the Brazilian Pantanal: an empirical example
+## Lowland tapir in the Brazilian Cerrado: an empirical example
 
@@
 
-In the case of animals whose spatial needs change over their lifetimes, one may define HR as a function of time with distinct HRs for each distinct period [e.g., salmon: @ref; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017] or as a continuous function so the HR changes smoothly over time [e.g., elephants: @polansky_elucidating_2015; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021].
-In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
-Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
-
-The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
+The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its food source's or associate's ability to move or adapt, too.
 
@@
 
-## Applying these methods
+## Using this framework in conservation-based decisions
 
-* The hypotheses we present here allow researchers to combine findings and generalize
-* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
-* can present results building on previous work rather than listing them as *de novo* findings
-* additionally, can test for common behaviors between members of the same population or species
-* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(R)$ or $\var(R)$
-* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior
-
-## The spatiotemporal scale of stochastic events: memory and adaptability
-
-The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism (i.e.
-approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
-Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
-
-Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
-@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
-*changes in bird migration, check for changes in mammals*
-
-* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
-
-\newpage
-
-<!--
-Notes:
-
-@rickbeil_plasticity_2019:
-- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
-- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
-- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
--->
+* maps of resource abundance and resource unpredictability
+* predict animal home range sizes for different levels of $\e(R)$ and $\var(R)$; pick stable areas (flatter parts of the curves)
 
@@
 
-In the case of animals whose spatial needs change over the their lifetimes [@refs], one may define HR as a function of time with distinct HRs for each period [e.g., salmon spawning: @ref] or as a continuous function of age [so the HR changes smoothly: @ref].
-In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
-Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
-
 The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
-
-## The spatiotemporal scale of stochastic events: memory and adaptability
-
-The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism (i.e.
-approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
-Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
-
-Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
-@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@whales_getting_stuck].
-@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
 
@@
 
+\newpage
+
 <!--
@@
 * Attribute memory, i.e., non-spatial memory, e.g., food type and abundance (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B13)
-* Roe deer an spatial memory for foraging in absence of territorility (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B41)
 * memory in capuchin monkeys (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B8)
@@
 
+\clearpage
+
 \newpage <!-- place references on a separate page -->

### 2022-11-18 writing/manuscript.Rmd
cleaned text and added refs

@@
 "When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
-habitat productivity hypothesis: Harestad and Bunnel, 1979 [@harestad_home_1979]
 -->
@@
 The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
-Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_20129; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)).
+Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)).
 As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
@@
 Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs.
+While competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
+
+It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose: @singh_migration_2012; eagles: @wheat_migrate_2017; @poessel_interpreting_2022; flamingos: @pretorius_movement_2020].
 From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$).
 As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.
+
+```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance).
+(a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
+(c.) Simulated animal movement at a time of lowest $\\e(R)$ and average $\\var(R)$.
+(d.) Simulated animal movement at a time of peak $\\e(R)$ and average $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
+```
+
+\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
+However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
+peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019] although it will not be possible to predict where high-$R$ locations are, since $\e(R)$ does not vary predictably over space.
+
+This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
+Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance.
+Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
+When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
+Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
+Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' spatial needs should be estimated before assuming the effect is negligible.
+The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
 
@@
 Although each of these events is unpredictable, the change in unpredictability can be predicted and expected.
-In contrast, other sources of stochasticity can be much less predictable, such as fires, floods, droughts, the arrival of new competitors, and changes in climate [@ipcc_global_2018; @noonan_situ_2018].
+In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors, fires, floods, droughts, and changes in climate [@jolly_animal_2022; @ipcc_global_2018; @noonan_situ_2018].
 
@@
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, torpor, and hibernation).
+
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
+(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
+(c.) Simulated animal movement at a time of lowest $\\e(R)$ and $\\var(R)$.
+(d.) Simulated animal movement at a time of peak $\\e(R)$ and $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
+```
+
+When the (un)predictability in $R$ chances over time (figure \@ref(fig:multipanel-changing-var)a), changes in $\var(R)$ may have an appreciable effect on the location's favourableness.
+In this scenario, we can define $\var(R)$ as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
+When both the expected $R$ and its unpredictability change over time, accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
+
+Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years.
+In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs.
+These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
+Thus, failing to account for changes in $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].
 
@@
 It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants.
-Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels.
+Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels [@harestad_home_1979; @noonan_effects_2020].
 
@@
 
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\e(R)$ and average $\\var(R)$.
-(d.) Simulated animal movement at a time of peak $\\e(R)$ and average $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
-```
-
-\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
-Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
-However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019].
-However, since $\e(R)$ does not vary predictably over space, it will not be possible to predict the location of locations with high or low $R$.
-
-This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
-Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance.
-Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
-Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
-Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' spatial needs should be estimated before assuming the effect is negligible.
-The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
-Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\e(R)$ and $\\var(R)$.
-(d.) Simulated animal movement at a time of peak $\\e(R)$ and $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
-
-When animals are not guaranteed that the resources they find during one visit will be there next time (figure \@ref(fig:multipanel-changing-var)a), $\var(R)$ will have an appreciable effect on the location's favourableness.
-Since $\var(R)$ is not constant over time, we can now define it as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
-In this scenario, both the expected $R$ and its predictability change over time, so accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
-
-Most habitats with seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ often changes both within and between years.
-In such cases, failing to account for $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].
-In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or in periods of increased energetic needs, such as during pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
-In heterogeneous environments, such as habitats with a high diversity of clustered resources, animals may have to move between locations if they cannot depend on a single one [@lucherini_habitat_1996].
+In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable.
+Thus, we expect $\var(R)$ to have a stronger effect on resource abundance when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
 
 Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
-The top row (constant $\var(R)$) shows how $\vpos$ remains constant.
-Some noise in $H$ is still present since $\var(R) \ne 0$ and thus $R$ is not constant).
-Note the nonlinear decrease in $H$ when $\e(R)$ increases at a constant slope (linear $\e(R)$), with larger changes when $\e(R)$ is low.
-This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g.
-from 1 to 2) can correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
-Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ since larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the $\vpos$.
-In regions where the average resource abundance changes over time (cyclical $\e(R)$; e.g., seasonal changes), animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable changes in $\e(R)$ may increase animals' spatial needs unpredictably.
-Thus, it is best to avoid changing the space available to animals too close to when resources are expected to be low and include a "buffer" area so the available area is larger than the estimated spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
-In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time.
-Finally, in ecosystems where $\e(R)$ changes erratically and suddenly (erratic $\e(R)$), such as regions prone to large fires or sudden floods, as well as regions with high mining or logging activity, the space available to animals should be sufficient to satisfy needs during the periods of greatest scarcity while also allowing animals to adapt their behavior between periods.
+The top row (constant $\var(R)$) shows how simulated spatial needs vary for different trends in $\e(R)$ while $var(R)$ remains constant.
+As $\e(R)$ increases at a constant slope (linear $\e(R)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home range size approaches the minimum size required by the animal.
+This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g., from 1 to 2) correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
+Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ because larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the home range.
 
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates how an animal's home range may vary as $\var(R)$ changes over time while $\e(R)$ remains constant.
-Overall, $\vpos$ increases with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the 50% core home range ($H_{50\%}$).
-***here*** <!--Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).-->
+In regions where the average resource abundance changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures).
+However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable decreases in $\e(R)$ (such as following fires) may increase animals' spatial needs unpredictably.
+Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
+In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\e(R)$).
 
-Both $\e(H)$ and $\var(H)$ increase with $\var(R)$, although the 95% home range ($H_{95\%}$) is more sensitive to changes in $\var(R)$ than the core home range ($H_{50\%}$).
-Additionally, the effect of $\var(R)$ on $H$ depends on $\e(R)$, since the effect of $\var(R)$ on $H$ is stronger when $\e(R)$ is low (see the bottom row with erratic $\var(R)$).
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated home range while $\e(R)$ remains constant.
+Overall, both quantiles increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$, as with changes in $\e(R)$.
+Additionally, note how both the estimated home range size and the noise in the estimate increase with $\var(R)$ (e.g., constant $\e(R)$, linear $\var(R)$).
+This is because as the habitat becomes more stochastic, the amount of space required to reach satiety also becomes increasingly unpredictable.
+
+The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
+Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both change monotonically, especially when the trends are linear.
+When both $\e(R)$ and $\var(R)$ increase linearly, the change in $H_{95\%}$ is small an not monotonic, while there is no visible change in $H_{50\%}$.
+However, when $\e(R)$ changes nonlinearly, it is easy to see how the increase in $\var(R)$ amplifies the effects of $\e(R)$.
+Similarly, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
+Thus, the estimated spatial needs are largest when $\e(R)$ is low and $\var(R)$ is high and they are smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$, and @lai_movement_2017].
 
 Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
-$\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ remain undetected.
-Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly continuously, such increases may be possible for short periods of time (followed by periods of no change or decrease).
-Additionally, these examples are important because they demonstrate the relationships between $H$, $\e(R)$, and $\var(R)$ in (relatively) simple scenarios.
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments with an abundance of competitors, threats, and stochasticity, such as a habitat with a high degree of human alteration and activity.
+$\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ are relatively small and remain undetected.
+Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease).
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
 Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
-However, if highly stochastic or erratic changes occur frequently, animals may perceive them as smooth changes in $\e(R)$ or increases in $\var(R)$ rather a series of small, sudden, changes.
-Estimating the true trend in $\e(R)$ and $\var(R)$ would often require an excessively high cognitive capacity and an equally unlikely abundance of information, since although changes in $\e(R)$ are not due to $\var(R)$, but often distinguishing between the two is not easy [@steixner-kumar_strategies_2020], and $\e(R)$ and $\var(R)$ are not independent for $R \sim \Gamma(\mu, \sigma^2)$.
+However, if changes are stochastic but sufficiently small and frequent, animals may perceive them as few smooth changes rather a series of small, sudden, changes.
+Estimating trends in $\e(R)$ and $\var(R)$ requires a high cognitive capacity, since distinguishing between $\var (R)$ and changes in $\e(R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent (as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$).
+
+Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the effects of $\e(R)$ and $\var(R)$ are as we hypothesized (figure \@ref(fig:5-5-reg)).
+In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.
 
 ```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Linear Model with a Gamma family.
-The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile.
-Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+The relationships were estimated using a Generalized Linear Model with a Gamma family of distributions that accounted for the effects of both $\\e(R)$ and $\\var(R)$ as well as differences between the two quantiles.
+The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile.
+Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlinear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
 knitr::include_graphics('../figures/simulation-regression-plots.png')
@@
 
-#### Move the following paragraph to previous sections
-
-@nilsen_can_2005 found that the home ranges of brown bears (*Ursus arctos*), leopards (*Panthera pardus*) decreased nonlinearly with the fraction of photosinthetically active radiation absorbed by plants in a given region.
-However, the effect of many other factors on HR size likely depends on the species, diet type [@harestad_home_1979], location, and possibly also on the individual animal.
-For instance, competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], but strong and consistent competition paired with territorial defense may also prevent them from doing so [e.g., wolves, @rich_anthropogenic_2012; feral cats, @bengsen_feral_2016; Capuchin monkeys, @torrez-herrera_monkeys_2020].
-Similarly, predation may force animals to move more frequently to escape predators, or it may prevent them from venturing too far from the safety of their core HR too often [@suraci_beyond_2022].
-Patch quality, size, fragmentation, and heterogeneity may cause animals to explore more patches if some are of low value, too small, too disconnected, or too variable [@fahrig_is_2019], but high diversity may also decrease HR size if animals require heterogeneous habitats [@fox_niche_1981; @lucherini_habitat_1996].
-Similarly, patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006].
-However, not all animals take advantage of linear features or higher predictability in human-altered habitats.
-@noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed.
-Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat.
-Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.
-
 Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022
 
-Simulation refs: @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
+Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
 
@@
 * increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness
-* animal mortality following fires, how little we know of mortality following fires: @jolly_animal_2022
 
@@
 Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
-
-Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
-@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to become trapped in sub-optimal locations [@abrahms_memory_2019].
-@merkle_spatial_2019 suggest that Bison (*Bison bison*) rely memory much more than on tracking spring green-up or autumn snow depth.
 
@@
 
-In the case of animals whose spatial needs change over their lifetimes [@refs], one may define a HR as a function of time with distinct HRs for each period [salmon: @ref; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017] or as a continuous function of time.
+In the case of animals whose spatial needs change over their lifetimes [arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017], one may define a HR as a function of time with distinct HRs for each period or as a continuous function of time.
 In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
@@
 
-\newpage
+competition can promote niche partitioning [@fox_niche_1981]
 
-# Extra notes
+Patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006].
 
-@polansky_elucidating_2015 as example of memory and changes in movement type (exploration vs stationary)
-
-@teitelbaum_how_2015 found that resource abundance (measured as NDVI) had a strong negative effect on migration distance, while the variance in NDVI only had a small positive effect on **???**.
-
-high-resource areas and explaining 23% of the variation in migration distances.
+However, not all animals take advantage of linear features or higher predictability in human-altered habitats.
+@noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed.
+Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat.
+Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.
 

### 2022-11-18 writing/manuscript.Rmd
end of week push

@@
   bookdown::pdf_document2:
+    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
     pandoc_args:
@@
 
-***Not sure if we need this or if we can tie it in somewhere:***
-
-Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017].
-However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...
-
 ### Advantages of modeling animal movement as a stochastic process
@@
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear [see figure \@ref(fig:mu-hyp) and @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear, as illustrated in figure \@ref(fig:mu-hyp) [@nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
 Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)).
@@
 
-\noindent Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
+Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
 We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
@@
 Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
+
+* @berger_climate_2018 fitness decreases with resources scarcity and lower temperatures
 
@@
 
-Environmental variability, including extreme events [@logares_black_2012], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness and increases their spatial needs.
+Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs].
 A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
@@
 These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
-Thus, failing to account for changes in $\var(R)$ can result in serious biases in the estimation of animals' spatial needs [@examples].
+Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@examples].
+
+Recent increases in the frequency and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018].
+When possible, increasing the frequency of extreme events to reduce their frequency (such as with controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
 
@@
 Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease).
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
 Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
@@
 
-Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the effects of $\e(R)$ and $\var(R)$ are as we hypothesized (figure \@ref(fig:5-5-reg)).
+Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the both $\e(R)$ and $\var(R)$ affect spatial needs as we hypothesized (figure \@ref(fig:5-5-reg)).
 In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.
@@
 This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to build upon previous findings and datasets rather than stating their results as *de novo* findings.
+The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
 Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref].
-Individuals and populations that were once said to have markedly different behaviors and spatial needs may instead have a single common behavior but with different levels of $\e(R)$ and $\var(R)$ [@examples].
+Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].
 
-In this discussion section, we make considerations about the effects of resource abundance and stochasticity on various adaptations, including movement behaviors, energy storage, and spatiotemporal memory.
+In this discussion section, we make considerations about the effects of $\e(R)$ and $\var(R)$ on various animal adaptations, including movement behaviors, energy storage, and spatiotemporal memory.
 We then illustrate how one may apply this framework to their own data and we also provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
@@
 
-Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022
+## Animal adaptations to changes in $\e(R)$ and $\var(R)$
 
-Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
-
-* failing to account for $\var(R)$ may lead to decreases in population fitness and size as well as increase the risk of population collapse or human-wildlife conflict
-* find examples of extreme events that caused population collapses or mass die-offs
-* increasing stability of ecosystems and populations is a good way of increasing chances of long-term survival and fitness
-
-## The spatiotemporal scale of stochastic events: memory and adaptability
-
-@ranc_memory_2022: Roe deer create HR based on memory
+@liow_lower_2009: *Lower Extinction Risk in Sleep-or-Hide Mammals*
 
 The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism (i.e.
-approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
+Events which occur at the scale of the organism [i.e.
+approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
+
+Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g.
 elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events [@examples].
-Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@logares_black_2012; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; **mule deer?**: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015; **species?**: @fagan_spatial_2013].
+However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples].
+Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].
 
 Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues and anticipate periods and areas of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-@rickbeil_plasticity_2019 showed that the yearly migration of elk (*Cervus canadensis*) depends on cues such as available forage biomass, hunting pressure, snow fall, and snow melt, while @falcon-cortes_hierarchical_2021 propose that elk move following nonrandom patterns based on memory that can last longer than 11 months.
-@jonzen_rapid_2006 demonstrated that large-scale, trans-Saharian migrant birds adapt their migration timing based on changes in the beginning of spring.
+An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
+For example, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their movement by reducing the amount of time spent exploring.
+Similarly, the ability of trans-Saharian migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006].
 However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
-*changes in bird migration, check for changes in mammals*
-
-* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
 
@@
 Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
+
+* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
 
@@
 
-* The hypotheses we present here allow researchers to combine findings and generalize
-* the analytical methods we use provide an example for how researchers can test these hypotheses using their own data
-* can present results building on previous work rather than listing them as *de novo* findings
-* additionally, can test for common behaviors between members of the same population or species
-* ecologists often conclude that populations from different regions behave differently [@examples], but their differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$
-* using the methods presented here, it is possible to test whether populations and individuals do behave differently while also quantifying the differences in behavior
+The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses.
+We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for.
+In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings.
+Additionally, the analyses can be expanded to test for commonalities and differences in the behaviors between members of the same population or species, as well as between multiple distinct populations or species.
+
+Ecologists often conclude that populations from different regions behave differently [@examples], but the differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$, instead, which would cause animals to behave differently even if they depend on the same functional responses.
+
+We start by providing examples and considerations on how one may model $R$ to estimate $\e(R)$ and $\var(R)$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animal's spatial needs.
+Next we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.
 
 ### Modeling $R$
+
+Location-scale models are a class of statistical models that allow one to model the mean and variance of the response rather than only modeling the mean.
 
@@
 
+Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values.
+Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.
+
 * Gamma distribution for HR
@@
 
-## Lowland tapir in the Brazilian Cerrado: an empirical example
+### An empirical example: Lowland tapir in the Brazilian Cerrado
 
@@
 
-## Using this framework in conservation-based decisions
+## Using this framework in conservation-related decisions
 
@@
 # MISC
+
+Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022
+
+Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
 

### 2022-11-20 writing/manuscript.Rmd
added to modeling $R$ and spatial needs

@@
 
-In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
 Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$).
@@
 
-In probability and statistics, random variables are defined by specifying the distribution the variable follows.
+\noindent In probability and statistics, random variables are defined by specifying the distribution the variable follows.
 Since the variable is random, it can take multiple values (minimum 2), each with different probabilities.
@@
 
-The expectation of a random variable (i.e.
+\noindent The expectation of a random variable (i.e.
 its mean or average, indicated here as ) is the value one can *expect* the random variable to take.
@@
 
-A random variable's variance is a measure of its unpredictability or variability.
+\noindent A random variable's variance is a measure of its unpredictability or variability.
 We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$.
@@
 
-In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
+\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
 Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples.
@@
 
-In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011].
+\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011].
 However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation
@@
 
-@nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
+\noindent @nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
 
@@
 
-To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day.
+\noindent To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day.
 If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another.
@@
 
-The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
+\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
 Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident.
@@
 
-Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
+\noindent Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
 Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
@@
 
-Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs].
+Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017; @yao_emergence_2022], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs].
 A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
@@
 These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
-Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@examples].
+Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
 
-Recent increases in the frequency and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018].
-When possible, increasing the frequency of extreme events to reduce their frequency (such as with controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
+Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018].
+Managing the intensity of and frequency of extreme events (e.g., controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
 
@@
 
-In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+\noindent In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable.
@@
 
-\noindent
-
-This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
+\noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
 The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
@@
 
-The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+\noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
 Events which occur at the scale of the organism [i.e.
 approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
-
 Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g.
@@
 However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples].
-Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008; @examples], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].
+Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].
+
+* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
 
@@
 However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
-
-The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism (i.e.
-approximately one or more times per lifetime and within the animal's spatial range) are more likely to be perceived as a threat or boon worth preparing for [e.g., @foley_severe_2008; @refs].
-Smaller organisms (e.g., mice) are more likely to be be severely impacted by a stochastic event than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] can promote traits that increase survival following frequent extreme events [@examples].
-Frequent events may result in changes in animal behavior and memory [@foley_severe_2008; @geremia_migrating_2019; @examples], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through natural selection and evolution [@logares_black_2012; @examples].
-
-* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
 
@@
 
-The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses.
+\noindent The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses.
 We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for.
@@
 
-Location-scale models are a class of statistical models that allow one to model the mean and variance of the response rather than only modeling the mean.
+\noindent Location-scale models [@stasinopoulos_generalized_2007, @rigby_generalized_2005] are a class of statistical models that us to estimate changes in $\e(R)$ and $\var(R)$ while allowing the mean-variance relationship to vary.
+`mgcv` [@wood_generalized_2017] is a commonly-used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017] for location and scale (GAMLSs).
+Currently, the `mgcv` package allows one to fit location-scale models with a variety distribution families, including Gaussian, scaled t, Gamma, Tweedie, Gumbel families.
 
-* location-scale models
-* `mgcv`, `brms`
-* choosing the best family
-* choosing the right complexity of each term (maybe that's why @nilsen_can_2005 found contrasting results)
+The Gaussian family of distributions is the most flexible, since the mean and standard deviation parameters are assumed to be independent, and the response can be either positive or negative.
+The scaled t family is similar to the Gaussian family, but it is best for high-variance data with heavy tails.
+However, the two distributions' flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data, count data, proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
+
+The Gamma location-scale family is best for strictly positive responses, including elemental compositions (e.g., carbon to nitrogen ratio), total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
+The Tweedie family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey abundance.
+Lastly, the Gumbel distribution is suitable for modeling the extremes of a distribution, such as maximum or minimum daily temperatures.
+
+If one is interested in using other families of distributions, including zero-inflated and hurdle models, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of the family's parameters using a fully Bayesian approach [@burkner_advanced_2018].
+In this paper, we use the `mgcv` package because although `brms` provides more flexibility and a greater choice of families, its computational costs can be prohibitive for large datasets, particularly for location-scale models.
+
+Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
+Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and not monotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatio-temporal terms should be chosen carefully, particularly the terms for the mean (i.e., location).
+Excessively wiggly mean terms will cause the variance to be under-estimated and under-fit, and an excessively smooth mean terms will inflate the variance term, so an improperly defined model can lead to inaccurate estimates.
+Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt should be sufficient.
+Additionally, REstricted Marginal Likelihood [a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
+We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
+Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies.
 
@@
 
-* importance of AKDE
-* Introduce moving window idea
+\noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017].
+Since all models that can be currently fit via the `ctmm` package [@fleming_ctmm_2021] assume the animal to be range-resident and move with a single, consistent behavior, we suggest estimating an animal's spatial needs using a **sliding window** approach.
 
-Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; bald eagles: @wheat_migrate_2017].
-However, the models we use in this paper assume an animal's large-scale behavior and movement to remain consistent over time...
+**here**
 
-AKDE is good for range-resident animals
+Let $w$ indicate a fixed window width.
 
-In the case of animals whose spatial needs change over their lifetimes [arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017], one may define a HR as a function of time with distinct HRs for each period or as a continuous function of time.
 In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
@@
 
-Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values.
+\noindent Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values.
 Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.
@@
 
-Animal's spatial needs are often estimated using all available tracking data.
+\noindent Animal's spatial needs are often estimated using all available tracking data.
 Doing so maximizes the (effective) sample size, but it also assumes that during the tracking period the animal's movement behavior did not change and that its home range did not shift either.
@@
 ***Most animals (if not all) use their space selectively.*** *add examples with humans moving to different regions?*
-
-The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its food source's or associate's ability to move or adapt, too.
 
@@
 
+Additional info is in appendix 2
+
+we use NDVI [@ref] as a proxy for resource abundance.
+
 ## Using this framework in conservation-related decisions
+
+\noindent text
 
@@
 # MISC
+
+The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its food source's or associate's ability to move or adapt, too.
 

### 2022-11-21 writing/manuscript.Rmd
added to the discussion and some refs

@@
 
-Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, blizzards, fires, disease outbreaks, and oil spills are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by causing mass die-offs that drastically reduce the size and genetic diversity of a population [e.g., @matkin_ongoing_2008; @murray_pathogens_2006; @widmer_effects_2004; @reed_frequency_2003; @hollings_trophic_2014; @ameca_y_juarez_natural_2012; @berger_climate_2018].
-Managing the intensity of and frequency of extreme events (e.g., controlled burns) can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
+Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, hurricanes [@widmer_effects_2004]; severe storms [@berger_climate_2018], fires [@ruthrof_how_2016], new diseases [@murray_pathogens_2006; @hollings_trophic_2014; @hoyt_ecology_2021], and anthropogenic die-offs [@matkin_ongoing_2008] are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by drastically reducing the size and genetic diversity of a population [@reed_frequency_2003; @ameca_y_juarez_natural_2012].
+Managing the intensity of and frequency of extreme events [e.g., @aponte_forest_2016] can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
 
@@
 The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
-Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response [@ref].
+Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response using hierarchical models [@pedersen_hierarchical_2019].
 Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].
@@
 
-@liow_lower_2009: *Lower Extinction Risk in Sleep-or-Hide Mammals*
-
 \noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
@@
 Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [@foley_severe_2008; @polansky_elucidating_2015].
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [elephants: @foley_severe_2008; @polansky_elucidating_2015].
 However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples].
-Frequent events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], including the developing of memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; elk: @falcon-cortes_hierarchical_2021; blue whales: @abrahms_memory_2019; elephants: @polansky_elucidating_2015].
+Frequent and repetitive events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008].
 
-* range shifts can help species overcome changing conditions, but they can also damage species that are being encroached upon, such as species that either fail to adapt (i.e., do not shift, too) or species that **cannot** shift
+Events occur which multiple times at relatively predictable intervals during an animal's lifespan and over multiple generations, such as the alternating of seasons, allow it to anticipate times of scarcity and prepare for them by, for instance, increasing its fat storage [marmots: @armitage_energetics_2003; bears: @nespolo_why_2022] or caching resources [@post_caching_2006].
+Of course, an animal with a life span shorter than a year will be unable to perceive the cyclical alternating of the seasons, and an animal with a life span of a few weeks or days may not even experience substantial changes in weather and temperature other than lower temperatures at night than during the day.
+When an animal's expected lifespan is at least one or more years, preparing for predictable times of abundance and scarcity becomes an advantage (and often a necessity).
+In such cases, animals can increase their chances of survival via physiological and behavioral adaptations, including torpor [marmots: @inouye_climate_2000; ground squirrels: @goldberg_hibernation_2021]; and memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; caribou: @cavedon_genomic_2022; elk: @falcon-cortes_hierarchical_2021; elephants: @polansky_elucidating_2015], which is required for migration [butterflies: @stefanescu_timing_2021; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018].
+However, relying on environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
 
@@
 An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-For example, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their movement by reducing the amount of time spent exploring.
-Similarly, the ability of trans-Saharian migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006].
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008] or become trapped in sub-optimal locations if long-term conditions and cycles change unexpectedly [e.g., blue whales: @abrahms_memory_2019; bison: @merkle_spatial_2019].
+For example, yellow-bellied marmots (**) []
+
+Likewise, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time spent exploring.
+Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006].
+However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
+
+Recent changes in climate during the last few decades have caused some species to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@examples].
+While range shifts can help species overcome changing conditions, they can still result in phenological mismatches [@lameris_arctic_2018], and they also add pressure to the species that already live in the habitat, such as species that either fail to adapt (i.e., do not also shift) or species that are unable shift.
+Specialists and obligate symbionts are less likely to shift since their range is determined by their food sources' or associates' ability to move and adapt, too.
+Consequently, the arrival of new species may exacerbate the stressors such animals are already facing.
 
@@
 - Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
-- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006; Merkle et al., 2016)
-- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006; Lameris et al., 2018) and highlighting a potential measure of resilience in the face of climate change
+- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006)
+- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006) and highlighting a potential measure of resilience in the face of climate change
 -->
@@
 
-The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its food source's or associate's ability to move or adapt, too.
-
 Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022

### 2023-01-05 writing/manuscript.Rmd
read up to section on modeling HR

@@
     - \usepackage{indentfirst} # for indenting first line of each paragraph
-    - \usepackage[small]{titlesec} # for smaller font for headings
+    # - \usepackage[small]{titlesec} # for smaller font for headings
     - \usepackage{wrapfig} # for wrapping text around figures
@@
     # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
-    # - \usepackage{float} # for better floating objects interface; adds more floats 
-    # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
+    # - \usepackage{float} # for better floating objects interface; adds more floats     # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 
-\noindent Since animals need to maintain a positive energetic balance to grow and reproduce, their spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
-Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance in around the mean are still largely unknown.
+\noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
+Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
 This is not a trivial oversight.
@@
 In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
-We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resource abundance on animals' space use.
+We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use.
 Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use.
@@
 
-\noindent The amount of resources an organism is able to access is a strong determinant of its probability to survive and reproduce.
-Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], or even death [@foley_severe_2008; @berger_climate_2018].
-Thus, many organisms will change their behaviors in response to changes in local resource abundance.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [@refs], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction.
+Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018].
+Thus, many organisms will adapt their behaviors in response to changes in local resource abundance.
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], and reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @ref; mammals: @ref], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 
@@
 Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce.
-Two years later, @harestad_home_1979 proposed that the simplest hypothesis for the relationship between an animal's home range size is inversely proportional to resource abundance.
+Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home range size is
+
+\begin{equation} \label{harestad-eq}
+H = C / R,
+\end{equation}
+
+\noindent where $H$ is the animal's **home range size**, $C$ is the animal's resource consumption, and $R$ is the abundance of (usable) resources.
 More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022].
 Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020].
-Additionally, an animal's spatial requirements depend strongly on its metabolic rate and diet, as demonstrated over three decades ago by various scientists [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982].
+Additionally, an animal's spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020].
 Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
-And while abundant work has been done on estimating animal's response to changes in resource abundance, there is little to no information on how many species respond to unpredictable changes in the resources they depend on.
+And while abundant work has been done on estimating animals' responses to changes in resource abundance, there is little to no information on they respond to unpredictable changes in the resources they depend on.
 
-A habitat's long-term average resource abundance is not sufficient to assess the habitat's quality.
+In this paper, we refer to a location's average amount of available and accessible resources as **resource abundance**, while use the phrase **resource unpredictability** to indicate the variability in resource abundance due to *unpredictable* causes.
+On its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality.
 A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources.
 In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted.
-Although both habitats may have the same long-term resource abundance, the differences in resource predictability result in substantially different habitats.
+Although both habitats may have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different habitats.
 Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].
@@
 In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
-First, we set the theoretical background necessary for the successive chapters.
-Concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology.
-Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis of how animals respond to changes in resource abundance.
-Afterwards, we present a hypothesis for how animals respond to resource unpredictability.
+First, we set the theoretical background necessary for the successive sections, and concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology.
+Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource unpredictability on animals spatial needs while suggesting a second unifying hypothesis.
 Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability.
@@
 
-\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
-Known values, such as realizations (i.e., known observations or instances) of random variables, are indicated with lower-case letters (e.g., $r$).
+\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
 Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
@@
 
-\noindent In probability and statistics, random variables are defined by specifying the distribution the variable follows.
+\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable follows.
 Since the variable is random, it can take multiple values (minimum 2), each with different probabilities.
 Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.
+
 The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.
-The support of the normal and T distributions are all real numbers, while the Bernoulli distribution only has support over values 0 and 1.
+The support of the normal and T distributions are all real numbers (i.e., from $-\infty$ to $\infty$), while the Bernoulli distribution only has support over values 0 and 1.
 
@@
 
-\noindent The expectation of a random variable (i.e.
-its mean or average, indicated here as ) is the value one can *expect* the random variable to take.
+\noindent The **expectation** of a random variable (i.e.
+its mean or average, indicated here with the symbol $\e$) is the value one can *expect* the random variable to take.
 $\e(R) = \mu$ to indicate the expectation of the random variable $R$.
@@
 When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.
+We indicate the estimated average resource abundance by adding a caret on the symbol: $\hat \mu(t)$.
 
@@
 
-\noindent A random variable's variance is a measure of its unpredictability or variability.
-We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$.
+\noindent A random variable's **variance** is a measure of its unpredictability or variability.
+We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$, and $\widehat{\sigma^2}(t)$ to indicate its estimate conditional on time.
 For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others.
@@
 Note that only **linear transformations** (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution's parameters [@denny_fallacy_2017; @jensen_sur_1906].
-In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$ (We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is $\mu = \frac{\alpha}{\alpha+\beta}$ while the variance is $\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}$).
-Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.
+only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906].
+In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
+
+\begin{equation} \label{beta-mean}
+\mu = \frac{\alpha}{\alpha + \beta}
+\end{equation}
+
+\noindent while the variance is
+
+\begin{equation} \label{beta-var}
+\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
+\end{equation}
+
+\noindent Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.
 
@@
 
-\noindent @nathan_movement_2008 generalized animal movement from location $\vec u_t$ to location $\vec u_{t+1}$ using the model
+\noindent @nathan_movement_2008 generalized animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
 
 \begin{equation} \label{nathan-movement}
-\vec u_{t+1} = F(\Omega, \Phi, \vec r_t, \vec w_t, \vec u_t),
+\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
 \end{equation}
 
-such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec u_t$), as well as how each of the parameters may change with the current environmental factors ($\vec r_t$).
+\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec r_t,$ and $\vec w_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
-However, we can focus on animals' behavior and spatial needs as a response to resource abundance and unpredictability alone (rather than measuring changes in each of an animal's resources, needs, and behavioral states directly).
+The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
+Thus, we can instead focus on animals' behavior and spatial needs as a response to overall resource abundance and unpredictability.
 
@@
 
-\noindent To understand how movement can be viewed as a stochastic process, imagine an animal that starts at its den and moves to collect food and other resources, and it returns to its den once it has collected enough resources for the day.
+\noindent To understand how movement can be viewed as a **stochastic process**, imagine an animal that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
 If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another.
-However, we can use the animal's positional data to estimate the amount of time it spends exploring before returning "home" (or the time required for the position to "decay" back to the average position), which is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
+However, we can use the animal's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
+This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
 Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity.
 $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
-Thus, by modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for a variety properties of tracking data which are problematic for models that assume errors to be independent while also leveraging such properties to estimate important movement parameters.
-Accounting for autocorrelation in movement data (which is present in most modern datasets) avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as path length and average (or instantaneous) speed [@noonan_scale-insensitive_2019].
 
-If sampling frequency is too low to provide precise measures of autocorrelation, it will be inappropriate to estimate an animal's average speed, directional persistence, and range crossing time, but the data can still provide reasonable estimates of the animal's spatial needs and usage, and model selection procedures with criteria such as AIC can help the analyst choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
-Estimates of average or instantaneous speed require the highest sampling frequency [$\Delta t \lessapprox \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
-However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior, it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the sampling duration is sufficiently long.
+By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data.
+While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].
+
+Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
+However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the total **sampling duration**, $T$, is sufficiently long.
+Model selection procedures with criteria such as AIC [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
+
+Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less *why* it moved.
+Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
 
@@
 \noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance of animals that are not range-resident.
-***but we can use a sliding window approach if animals are temporarily range resident or to estimate continuous changes in short-term spatial needs***
+Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.
+In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs using a sliding window approach (see the discussion section).
 
-Although range-resident animals will generally have smaller $\vpos$ than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior.
-Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different kinds of movement.
+**I don't think we need this paragraph:** Although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior.
+Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
 Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it is unlikely to return to locations it visited in the past.
-
-### Advantages of modeling animal movement as a stochastic process
-
-\noindent Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the causes and decisions that determine the movement allows us to focus more on the animal's movement and behavior and less on recognizing each cause of movement.
-Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
-
-Depending on sampling frequency, statistical methods that rely on straight-line displacement are prone to under-estimating the distance traveled by an animal and its speed at coarse sampling frequencies, while they over-estimate them at fine frequencies [@noonan_scale-insensitive_2019].
-In addition, most commonly-used methods for home range estimation assume tracking data to be independent of one another and sampled at consistent and discrete intervals, which is most often not the case.
-Consequently, animals' spatial needs are often under-estimated [@noonan_comprehensive_2019].
-
-### Unassigned
-
-@pretorius_movement_2020: flamingos -- migrating or nomadic?
-
-@boratynski_energetic_2020: animals' hr is limited by BMI and maximum metabolic rate, which can be accounted for by accounting for body mass [@noonan_effects_2020]
-
-@arechavala-lopez_common_2019: octopuses (*Octopus vulgaris*) congregate in human-altered coastal habitats
-
-@mitsuhashi_home_2018: urban raccoon dogs (*Nyctereutes procyonoides*) have a smaller HR than rural ones
-
-@cronk_home_2021: yellow mongoose (*Cynictis penicillata*) in urban regions have smaller HRs
-
-@yu_artificial_2022: rat snakes move to artificial structures for better prey availability and better thermoregulation
-
-Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
-
-@teitelbaum_beyond_2019:
-
-- **Nomadic movement**: movement with irregular timing and/or direction that produces both within- and between-year variability in location and movement patterns
-- "Data from recent developments in animal tracking technology (e.g., longer-lasting and lighter GPS devices) show that species previously considered migrants or range residents can show irregular, often long-distance, movement patterns (flexible foraging movements of leatherback turtles across the north Atlantic ocean, To migrate, stay put, or wander Varied movement strategies in bald eagles, Big data analyses reveal patterns and drivers of the movements of southern elephant seals)"
+In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have a position it returns to regularly.
 
@@
 
-# Effects of $\e(R)$
+# Effects of resource abundance, $\e(R)$
 
@@
 
+```{r, eval=FALSE, echo=FALSE}
+# to help visualize the consequences of different means and variances
+dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
+    # mean = k * theta
+    # sigma^2 = k * theta^2
+    dgamma(x = x,
+           shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
+           scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
+}
+x <- seq(0, 10, length.out = 250)
+layout(matrix(1:4, ncol = 2, byrow = TRUE))
+plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
+layout(1)
+```
+
 \noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
 All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources is likely to result in lower fitness, changes in physiology or behavior, physiological damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@harvey_primate_1981; @ref].
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear, as illustrated in figure \@ref(fig:mu-hyp) [@nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
-Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, and nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan (figure \@ref(fig:mu-hyp)).
-As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assumes the effect to be linear].
+Failing to acquire sufficient resources is likely to result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
+Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
 
-Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-While competition may push animals to explore other areas and expand their HR [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
+The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear, as illustrated in figure \@ref(fig:mu-hyp) [also see: @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
+Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan.
+
+As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assume linear relationships].
+The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
+In contrast, at higher values of $\e(R)$, the same additive changes (e.g.
+from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50).
+
+Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+While competition may push animals to explore other areas and expand their home range [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
 
@@
 ```
-
 Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
@@
 The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
-Additionally, anthropogenic structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
+
+Anthropogenic changes, including human-built structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
 Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
+Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
 
-* @berger_climate_2018 fitness decreases with resources scarcity and lower temperatures
-
-# Effects of $\var(R)$
+# Effects of resource unpredictability, $\var(R)$
 
@@
 Although each of these events is unpredictable, the change in unpredictability can be predicted and expected.
-In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors, fires, floods, droughts, and changes in climate [@jolly_animal_2022; @ipcc_global_2018; @noonan_situ_2018].
+In contrast, other sources of stochasticity can be much less predictable, such as new competitors [@ref], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], droughts [@foley_severe_2008; @haig_assessment_2013], and changes in climate [@ipcc_global_2018; @noonan_situ_2018; @berger_climate_2018; @grant_evolution_2017; @yao_emergence_2022].
 
-Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017; @yao_emergence_2022], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs [@refs].
+Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017; @yao_emergence_2022], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
 A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-There thus remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+
+```{r, eval=FALSE, echo=FALSE}
+# the median decreases as V(R) increases, even if E(R) is constant
+source('functions/qgamma2.R')
+plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
+```
 
@@
 When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)).
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, torpor, and hibernation).
+This is because as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly.
+Additionally, stochastic environments tend to be less productive [@@chevin_adaptation_2010], and the added movement required to search for food increases animals' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, food caching, torpor, and hibernation).
 
@@
 
-When the (un)predictability in $R$ chances over time (figure \@ref(fig:multipanel-changing-var)a), changes in $\var(R)$ may have an appreciable effect on the location's favourableness.
+When the unpredictability in $R$ chances over time (figure \@ref(fig:multipanel-changing-var)a), changes in $\var(R)$ may have an appreciable effect on the location's favourableness.
 In this scenario, we can define $\var(R)$ as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
-When both the expected $R$ and its unpredictability change over time, accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
+When both $\e(R)$ and $\var(R)$ change over time, accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
 
-Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, since periods of high productivity can often be more variable than times of dormancy or lower productivity, and the variability in $R$ can often change both within and between years.
-In highly variable and stochastic environments, animals will likely have higher spatial needs, particularly if they are required to collect and store resources for periods of scarcity or increased energetic needs.
-These include reproductive periods, whether this be due to pregnancy [@ref], to feed offspring [@ref], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
-Thus, failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in population fitness and size while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
+Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, as they often change within and between years, and periods of high productivity can often be more variable than times of lower productivity or dormancy.
+During times of resource unpredictability, animals will likely have higher spatial needs, particularly if they need to collect and store resources for periods scarcity or increased energetic needs.
+These include higher energetic needs during reproductive periods, whether this be due to pregnancy [@schmidt_interplay_2020], to feed offspring [@saraux_plasticity_2011], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
+Failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
 
-Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes, hurricanes [@widmer_effects_2004]; severe storms [@berger_climate_2018], fires [@ruthrof_how_2016], new diseases [@murray_pathogens_2006; @hollings_trophic_2014; @hoyt_ecology_2021], and anthropogenic die-offs [@matkin_ongoing_2008] are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by drastically reducing the size and genetic diversity of a population [@reed_frequency_2003; @ameca_y_juarez_natural_2012].
-Managing the intensity of and frequency of extreme events [e.g., @aponte_forest_2016] can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
+The effects of $\var(R)$ on animals' spatial needs likely vary between species, since energetic and nutritional needs vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
+Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
+@rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation (i.e., standard deviation over mean) in C:N, which suggests that the hares expanded their home range in response to both nitrogen abundance and nitrogen variability.
+Similarly, @mueller_how_2011 demonstrated that ungulates in more variable ecosystems move more and are more likely to be nomadic.
+In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more more heterogeneous habitats required exponentially less space than foxes that lived in more homogeneous regions.
+Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality.
+A subsequent analysis by @nilsen_can_2005 with the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*), found that carnivores respond differently to resource variability and seasonality, although it is worth disentangling the effects of heterogeneity (i.e., changes in $\e(R)$) and and stochasticity ($\var(R)$).
 
-The effects of $\var(R)$ on animals' spatial requirements most likely vary between species, since different species have different energetic and nutritional needs [@boratynski_energetic_2020; @noonan_effects_2020], so the degree to which $R$ varies will likely affects each species differently.
-Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not detect such differences.
-@rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions (UDs) of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation for C:N, which suggests that the hares expanded their home range in response to both resource scarcity and the variance in resource scarcity.
-Similarly, @mueller_how_2011 demonstrated that ungulate species in more variable ecosystems move more than species in more stable landscapes and that they are more likely to be nomadic.
-In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more habitats per hectare (i.e., more heterogeneous habitats) required exponentially less space than foxes that lived in more homogeneous regions.
-Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality.
-A subsequent analysis by @nilsen_can_2005 found that various carnivores respond to resource variability and seasonality (albeit in contrasting ways) using the Canadian lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*).
-Additionally, @nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$, so it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two.
-It is worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants.
-Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels [@harestad_home_1979; @noonan_effects_2020].
+@nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$.
+Therefore, it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two.
+It is also worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants.
+Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels, since animals from higher trophic levels will be affected by the cascading effects of $\e(R)$ and $\var(R)$ on each of the lower trophic levels [@harestad_home_1979; @noonan_effects_2020].
 
@@
 \noindent In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
-A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ is likely to have less of an effect if resources are always abundant, albeit unpredictable.
-Thus, we expect $\var(R)$ to have a stronger effect on resource abundance when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
+A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ will have little effect if resources are always sufficiently abundant.
+Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
 
-Figure \@ref(fig:5-5-sims) illustrates the simulated average spatial requirements of an animal as a function of $\e(R)$ and $\var(R)$.
-The top row (constant $\var(R)$) shows how simulated spatial needs vary for different trends in $\e(R)$ while $var(R)$ remains constant.
+Figure \@ref(fig:5-5-sims) illustrates simulated spatial requirements as a function of $\e(R)$ and $\var(R)$.
+The top row (constant $\var(R)$) shows how simulated spatial needs vary for different trends in $\e(R)$ while $\var(R)$ remains constant.
 As $\e(R)$ increases at a constant slope (linear $\e(R)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home range size approaches the minimum size required by the animal.
-This is because when $\e(R)$ is low, small additive changes in $\e(R)$ (e.g., from 1 to 2) correspond to large multiplicative changes (i.e., 2 is twice as big as 1).
-Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range $H_{50\%}$ because larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the home range.
+As described in the section on $\e(R)$, the nonlinear decrease is because changes in $\e(R)$ have a larger effect when $\e(R)$ is low.
+Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range simply because larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the distribution.
+Additionally, when $\e(R)$ is lower, $\var(H)$ is higher (particularly for the 95% quantile), since resource-poor regions force animals to move more, which results in a greater uncertainty in a longer and thus also more unpredictable search.
 
-```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
-The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
-The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
+```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource abundance ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
+The lower green line indicates the animal's core home range (50\\% quantile), while the top line indicates the 0.95 utilization quantile.
+Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase nonlinearly as $\\var(R)$ increases.
+Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they returned to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
+The animal's home range was then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
+", out.width='100%'}
 knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
@@
 In regions where the average resource abundance changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unpredictable decreases in $\e(R)$ (such as following fires) may increase animals' spatial needs unpredictably.
+However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
 Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated home range while $\e(R)$ remains constant.
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
 Overall, both quantiles increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$, as with changes in $\e(R)$.
-Additionally, note how both the estimated home range size and the noise in the estimate increase with $\var(R)$ (e.g., constant $\e(R)$, linear $\var(R)$).
-This is because as the habitat becomes more stochastic, the amount of space required to reach satiety also becomes increasingly unpredictable.
+Additionally, note how both $H$ and $\var(H)$ increase with $\var(R)$ (e.g., constant $\e(R)$, linear $\var(R)$).
+This is because, similarly to resource-poor areas, more stochastic times require the animal to move for longer periods of time, which results in a greater uncertainty in how much time and space it will require to reach satiety.
 
 The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
-Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both change monotonically, especially when the trends are linear.
-When both $\e(R)$ and $\var(R)$ increase linearly, the change in $H_{95\%}$ is small an not monotonic, while there is no visible change in $H_{50\%}$.
-However, when $\e(R)$ changes nonlinearly, it is easy to see how the increase in $\var(R)$ amplifies the effects of $\e(R)$.
+Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change monotonically, especially when the trends are linear (i.e., linear $\e(R)$ and $\var(R)$).
+When both $\e(R)$ and $\var(R)$ increase linearly, the change in $H_{95\%}$ is small an not monotonic, while there is no clearly visible change in $H_{50\%}$.
+However, when $\e(R)$ changes nonlinearly (e.g., cyclical $\e(R)$ with linear $\var(R)$), it is easy to see how the increase in $\var(R)$ amplifies the effects of $\e(R)$ (and vice-versa).
 Similarly, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
-Thus, the estimated spatial needs are largest when $\e(R)$ is low and $\var(R)$ is high and they are smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$, and @lai_movement_2017].
+Thus, the estimated spatial needs are largest when $\e(R)$ is low and $\var(R)$ is high, and they are smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
 
@@
 Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease).
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
-Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut for mining purposes with a subsequent artificial re-forestation).
-However, if changes are stochastic but sufficiently small and frequent, animals may perceive them as few smooth changes rather a series of small, sudden, changes.
-Estimating trends in $\e(R)$ and $\var(R)$ requires a high cognitive capacity, since distinguishing between $\var (R)$ and changes in $\e(R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent (as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$).
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
+Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
+However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
+Estimating trends in $\e(R)$ and $\var(R)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
 
-Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, the both $\e(R)$ and $\var(R)$ affect spatial needs as we hypothesized (figure \@ref(fig:5-5-reg)).
+Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, both $\e(R)$ and $\var(R)$ affect spatial needs as we hypothesized (figure \@ref(fig:5-5-reg)).
 In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.
@@
 \noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects resource abundance and stochasticity have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
+The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
 Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response using hierarchical models [@pedersen_hierarchical_2019].
-Individuals and populations that were once thought to have markedly different behaviors and spatial needs may instead have a single common functional response but different levels of $\e(R)$ and $\var(R)$ [@examples].
+Individuals and populations that were thought to have different behaviors and spatial needs may instead have a single common behavior as they respond to $\e(R)$ and $\var(R)$ following the same functional response [@examples].
 
 In this discussion section, we make considerations about the effects of $\e(R)$ and $\var(R)$ on various animal adaptations, including movement behaviors, energy storage, and spatiotemporal memory.
-We then illustrate how one may apply this framework to their own data and we also provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
+We then illustrate how one may apply this framework to their own data by providing an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
 Finally, we demonstrate how to use this framework to create products that can inform conservation-based decisions, including predicting animals' spatial needs and assessing the quality of potential (and current) conservation areas.
@@
 Events which occur at the scale of the organism [i.e.
-approximately one or more times per lifetime and within the animal's spatial range; see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
+approximately one or more times per lifetime and within the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
 Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to develop or memory about the frequency and severity of such events [elephants: @foley_severe_2008; @polansky_elucidating_2015].
-However, the short generation time and high reproductive rate of smaller *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates and genetic diversity through the production of numerous offspring [@examples].
-Frequent and repetitive events may result in changes in animal behavior and memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021], while events that occur over larger spatiotemporal scales are more likely to promote slower changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008].
+elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to adapt their behavior to such events, including the development of memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; @polansky_elucidating_2015; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; blue whales: @abrahms_memory_2019].
 
-Events occur which multiple times at relatively predictable intervals during an animal's lifespan and over multiple generations, such as the alternating of seasons, allow it to anticipate times of scarcity and prepare for them by, for instance, increasing its fat storage [marmots: @armitage_energetics_2003; bears: @nespolo_why_2022] or caching resources [@post_caching_2006].
-Of course, an animal with a life span shorter than a year will be unable to perceive the cyclical alternating of the seasons, and an animal with a life span of a few weeks or days may not even experience substantial changes in weather and temperature other than lower temperatures at night than during the day.
-When an animal's expected lifespan is at least one or more years, preparing for predictable times of abundance and scarcity becomes an advantage (and often a necessity).
-In such cases, animals can increase their chances of survival via physiological and behavioral adaptations, including torpor [marmots: @inouye_climate_2000; ground squirrels: @goldberg_hibernation_2021]; and memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; caribou: @cavedon_genomic_2022; elk: @falcon-cortes_hierarchical_2021; elephants: @polansky_elucidating_2015], which is required for migration [butterflies: @stefanescu_timing_2021; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018].
+The short generation time and high reproductive rate of smaller, *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates through higher population size, dispersal, and genetic diversity [@examples].
+Events that occur over large spatiotemporal scales can promote slow changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which are numerically more likely if the starting population is large with abundant offspring [@examples; but see @leung_smaller_2022].
+
+Events which occur multiple times at relatively predictable intervals during an animal's lifespan and over multiple generations (such as the alternating of seasons) allow the animal to anticipate times of scarcity and prepare for them by, for instance, increasing its fat storage [marmots: @armitage_energetics_2003; bears: @nespolo_why_2022], caching resources [@post_caching_2006], or migrating [@geremia_migrating_2019; @mueller_social_2013].
+Of course, an animal with a lifespan shorter than a year will be unable to perceive the cyclical alternating of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather and temperature other than lower temperatures at night than during the day.
+
+When an animal's expected lifespan is longer than a year, preparing for predictable times of abundance and scarcity becomes an advantage (and often a necessity).
+In such cases, animals can increase their chances of survival via physiological and behavioral adaptations, including adapting the timing of reproduction [@ref] and torpor [marmots: @inouye_climate_2000; ground squirrels: @goldberg_hibernation_2021], as well as memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; caribou: @cavedon_genomic_2022; elk: @falcon-cortes_hierarchical_2021; elephants: @polansky_elucidating_2015], which is required for migration [butterflies: @stefanescu_timing_2021; whooping cranes: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018].
 However, relying on environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
 
-Adaptability and responsiveness to change are strong determinants to whether an animal survives in a changing environment [@riotte-lambert_environmental_2020].
+An animal's fitness and chances of survival in a changing environment depend strongly on its adaptability and responsiveness to change [@riotte-lambert_environmental_2020].
 An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-For example, yellow-bellied marmots (**) []
-
-Likewise, ungulates' ability to time their yearly migrations on cues such as available forage biomass, hunting pressure, snow fall, and snow melt allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time spent exploring.
+For example, ungulates' ability to time their yearly migrations on available forage biomass, snow fall, snow melt, and hunting pressure allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time and energy spent exploring.
 Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006].
-However, relying strongly on short-term cues rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
+However, relying strongly on short-term changes rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
 
 Recent changes in climate during the last few decades have caused some species to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@examples].
-While range shifts can help species overcome changing conditions, they can still result in phenological mismatches [@lameris_arctic_2018], and they also add pressure to the species that already live in the habitat, such as species that either fail to adapt (i.e., do not also shift) or species that are unable shift.
-Specialists and obligate symbionts are less likely to shift since their range is determined by their food sources' or associates' ability to move and adapt, too.
+While range shifts can help species overcome changing conditions, they can still result in phenological mismatches [@lameris_arctic_2018], and they also add pressure to the species that already live in the habitat.
+Additionally, specialists and obligate symbionts are less likely to shift their range since it is determined by their food sources' and associates' ability to move and adapt, too.
 Consequently, the arrival of new species may exacerbate the stressors such animals are already facing.
@@
 In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings.
-Additionally, the analyses can be expanded to test for commonalities and differences in the behaviors between members of the same population or species, as well as between multiple distinct populations or species.
+Additionally, the analyses can be expanded to test whether members of the same population or species behave according to common functional responses to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
+In the following sections, we illustrate how one can apply this framework to empirical data.
 
-Ecologists often conclude that populations from different regions behave differently [@examples], but the differences in behavior may only be due to differences in $\e(r)$ or $\var(R)$, instead, which would cause animals to behave differently even if they depend on the same functional responses.
-
-We start by providing examples and considerations on how one may model $R$ to estimate $\e(R)$ and $\var(R)$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animal's spatial needs.
+We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
 Next we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
@@
 
-\noindent Location-scale models [@stasinopoulos_generalized_2007, @rigby_generalized_2005] are a class of statistical models that us to estimate changes in $\e(R)$ and $\var(R)$ while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly-used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017] for location and scale (GAMLSs).
-Currently, the `mgcv` package allows one to fit location-scale models with a variety distribution families, including Gaussian, scaled t, Gamma, Tweedie, Gumbel families.
+\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
+its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
+`mgcv` [@wood_generalized_2017] is a commonly-used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017], including location-scale GLMs and GAMs.
+Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), scaled t, gamma, Tweedie, and Gumbel.
 
@@
 The scaled t family is similar to the Gaussian family, but it is best for high-variance data with heavy tails.
-However, the two distributions' flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data, count data, proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
+However, the two distributions' flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
+area), count data, proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
 
-The Gamma location-scale family is best for strictly positive responses, including elemental compositions (e.g., carbon to nitrogen ratio), total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
+The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
 The Tweedie family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey abundance.
@@
 Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and not monotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatio-temporal terms should be chosen carefully, particularly the terms for the mean (i.e., location).
-Excessively wiggly mean terms will cause the variance to be under-estimated and under-fit, and an excessively smooth mean terms will inflate the variance term, so an improperly defined model can lead to inaccurate estimates.
+Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and nonmonotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatio-temporal terms should be chosen carefully, particularly for mean term.
+An excessively wiggly $\hat \mu(t)$ will cause $\widehat{\sigma^2}(t)$ to be under-estimated and under-fit, and an excessively smooth $\hat \mu(t)$ will inflate $\widehat{\sigma^2}(t)$.
+Thus, models should be fit with care to avoid inaccurate estimates.
 Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt should be sufficient.
-Additionally, REstricted Marginal Likelihood [a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
+Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
 We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies.
+Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size).
 
@@
 \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017].
-Since all models that can be currently fit via the `ctmm` package [@fleming_ctmm_2021] assume the animal to be range-resident and move with a single, consistent behavior, we suggest estimating an animal's spatial needs using a **sliding window** approach.
+However, most commonly-used movement models assume animals' behaviors to remain consistent during the entirety of the tracking period [but see state-space models, e.g., @polansky_elucidating_2015].
 
-**here**
+To quantify an animal's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
+The width of the window, $w$, should be a fixed value greater than two or three times the animal's expected range crossing time [@ref].
+If the animal's spatial use is relatively consistent over time, one may fit a movement model to the whole dataset, and then choose $w$ based on the model's $\tau_p$.
+Alternatively, if the animal's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of 2 or 3 range crossings (i.e., $w \approx 2.5 \,\tau_p$).
+Modeling animal movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
+In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022].
 
-Let $w$ indicate a fixed window width.
+### Modeling effects of $\e(R)$ and $\var(R)$
 
-In either case, the accuracy of the estimated parameters will depend strongly on the length of the observation period as well as the measurement frequency [@noonan_comprehensive_2019].
-Data from a portion of an animal's life may be sufficient if it is representative the animal's movement or if inference is limited to the period(s) for which data is available.
-However, when an animal's spatial needs change over time, models which assume range residency are not appropriate.
-Instead, a sliding window approach that estimates the short-term (e.g., weekly or monthly) spatial needs may produce sensible results.
-Modeling animal movement as an autocorrelated stochastic process is particularly useful in this approach, since it accounts the uncertainty and limited amount of information in short periods of tracking data [by accounting for difference in sample size and effective sample size -- see @noonan_comprehensive_2019; @noonan_scale-insensitive_2019].
+***CONTINUE READING HERE***
 
-### Modeling the relationships
+***CHANGE MODELS FOR SIMS REG TO LOCATION-SCALE GAMMA***
 
-\noindent Any linear model, including ANOVA-like models, would assume spatial needs can span all (complex) numbers, including negative values.
-Since spatial needs are strictly greater than zero, models that allow non-linear relationships while forcing the spatial needs to be strictly greater than zero.
-
-* Gamma distribution for HR
-* possibly location-scale if enough data?
+\noindent As discussed in the section on modeling $R$, the first step to modeling $H$ is to choose an appropriate distribution.
+Since $H$ is strictly positive and continuous, a Gamma distribution seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017].
+The Tweedie distribution would also be appropriate if the mean-variance relationship of the Gamma distribution does not fit the data well.
+The decrease in $\var(H)$ as $\e(R)$ increases and $\var(R)$ decreases in figure \@ref(fig:5-5-reg) suggests that a location-scale model is likely required.
+If there is not sufficient data to fit a location scale model, a simple Gamma model may be sufficient, but we suggest selecting the best distribution using an information criterion such as the Akaike Information Criterion [AIC, see @akaike_new_1974].
+While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest using a GLM rather than a GAM (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic.
+Additionally, restricting the model to a GLM reduces the risk of issues that may arise due to limited sample size and temporal autocorrelation in $\e(R)$ and $\var(R)$.
+If one interested in fitting a GAM but imposing shape restrictions, such as monotonicity, we sugget using the `scam` package [@pya_scam_2022].
 
@@
 
-\noindent Animal's spatial needs are often estimated using all available tracking data.
-Doing so maximizes the (effective) sample size, but it also assumes that during the tracking period the animal's movement behavior did not change and that its home range did not shift either.
-Although these assumptions may be valid for relatively short tracking periods (and often necessary for reasonable estimates of spatial needs), few animals will behave consistently during the tracking period and use the same space every day.
-***Most animals (if not all) use their space selectively.*** *add examples with humans moving to different regions?*
+***CHANGE MODELS FOR TAPIR REG TO LOCATION-SCALE GAMMA***
 
-```{r mw-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
+***ENSURE NDVI MODEL IS BASED ON DATA WITHOUT THE DEC 2017 RASTER***
+
+In this section, we use real-world tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a location-scale Gaussian GAM.
+Although a beta model would be more appropriate, the `mgcv` package does not currently support location-scale beta models.
+While the `brms` model supports beta location-scale models, the dataset is sufficiently large and with data sufficiently away from the boundaries (i.e., -1 and 1) that we expect the estimates to be reasonably accurate for an illustrative example.
+Additional information regarding the modeling can be found in Appendix 2.
+
+```{r tapir-mw, fig.cap="Seven-day home range size of a tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
 (a.) GPS tracking data of the tapir.
@@
 
-Additional info is in appendix 2
+Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
+Panel **a** provides an overview of the tracking data, while panels **b** and **c** show the changes in mean and variance in NDVI, respectively.
+Panel **d** shows the changes in the tapir's spatial use over time.
+Note how the home range size is larger during periods of lower average NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018), despite an increase in the variance in NDVI.
+In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times as much space.
+Finally, panels **e** and **f** suggest that the tapir's spatial needs change based on the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and found in the simulated responses (figure \@ref(fig:5-5-reg)).
 
-we use NDVI [@ref] as a proxy for resource abundance.
+## Applications for conservation
 
-## Using this framework in conservation-related decisions
+\noindent The designation of high-quality habitats is a central concern of many conservation initiatives.
+Using this framework, one can quantify the quality of a habitat based on its resource abundance and stochasticity while also estimating the amount of space animals would need in the habitat.
+Researchers and conservation organizations can use the methods presented in this paper to assess the quality of current protected areas and predict whether or not the areas will be sufficiently large or appropriate in future decades.
+Additionally, the methods can also be used to evaluate new potential conservation areas.
+In light of Canada's pledge to conserve 30% of its landmass and waters by 2030 [@ref], the framework presented here provide useful metrics for maximizing the quality and stability of protected areas while estimating their quality in the decades to come.
 
-\noindent text
+Accounting for changes in resource abundance and resource stochasticity (e,g, between seasons or "good" and "bad" years) allows researchers and conservation managers can ensure the quality of protected areas is sufficiently high to sustain biodiversity for decades after the designation of the protected area.
+Protecting areas with high resource abundance and low resource unpredictability will decrease animal's spatial needs while also increasing the resiliency in home range size to small changes in resource abundance or stochasticity, since changes in spatial needs are more drastic when $\e(R)$ is low or $\var(R)$ is high (figure \@ref(fig:5-5-reg), panels e and f).
 
-* maps of resource abundance and resource unpredictability
-* predict animal home range sizes for different levels of $\e(R)$ and $\var(R)$; pick stable areas (flatter parts of the curves)
+<!-- create a pdf for supplementary information on how to create the 5x5 panel and run methods w real data, as well as considerations and potential issues about modeling means and variances -->
 
 \newpage
-
-<!-- create a pdf for supplementary information on how to create the 5x5 panel and run methods w real data, as well as considerations and potential issues about modeling means and variances -->
 
@@
 We demonstrate that animals' spatial needs respond nonlinearly to both $\e(R)$ and $\var(R)$, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a framework for modeling animal behavior and movement continuously rather than imposing arbitrary thresholds between range-resident and nomadic animals.
-(notes on seasons and "seasonal behavior")
+We also provide way of modeling animal movement continuously rather than imposing arbitrary thresholds between range-resident and nomadic animals.
+The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals will need.
 
@@
 
-Other empirical refs to possibly add: @lindstedt_seasonality_1985, @jonzen_rapid_2006, @wolkovich_warming_2012, @falcon-cortes_hierarchical_2021, @nathan_big-data_2022
+Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes [@emerton_predicting_2022], hurricanes [@widmer_effects_2004]; severe storms [@berger_climate_2018], fires [@ruthrof_how_2016], new diseases [@murray_pathogens_2006; @hollings_trophic_2014; @hoyt_ecology_2021], and anthropogenic die-offs [@matkin_ongoing_2008] are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by drastically reducing the size and genetic diversity of a population [@reed_frequency_2003; @ameca_y_juarez_natural_2012].
+Managing the intensity of and frequency of extreme events [e.g., @aponte_forest_2016] can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
+
+@mueller_social_2013: Whooping Cranes migrate better when there are elders
+
+Other empirical refs to possibly add: @wolkovich_warming_2012, @nathan_big-data_2022
 
 Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
-
-The ability of a specialist or obligate symbiont to shift its home range or adapt will likely depend strongly on its associate's ability to move or adapt, too.
-
-competition can promote niche partitioning [@fox_niche_1981]
 
@@
 * humans, nonhuman primates and other large-brained vertebrates make movement decisions based on spatial representations of their environments (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B50)
-* 2010).
-These representations may allow animals to move directly to important sites in their environment that lie outside of their perceptual range (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B40, https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B35)
+* These representations may allow animals to move directly to important sites in their environment that lie outside of their perceptual range (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B40, https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B35)
 * Attribute memory, i.e., non-spatial memory, e.g., food type and abundance (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B13)
@@
 
-\clearpage
+\clearpage <!-- to prevent figures from floating to the references -->
 

### 2023-01-06 writing/manuscript.Rmd
finished reading

@@
 
-***CONTINUE READING HERE***
-
-***CHANGE MODELS FOR SIMS REG TO LOCATION-SCALE GAMMA***
-
 \noindent As discussed in the section on modeling $R$, the first step to modeling $H$ is to choose an appropriate distribution.
@@
 The Tweedie distribution would also be appropriate if the mean-variance relationship of the Gamma distribution does not fit the data well.
-The decrease in $\var(H)$ as $\e(R)$ increases and $\var(R)$ decreases in figure \@ref(fig:5-5-reg) suggests that a location-scale model is likely required.
-If there is not sufficient data to fit a location scale model, a simple Gamma model may be sufficient, but we suggest selecting the best distribution using an information criterion such as the Akaike Information Criterion [AIC, see @akaike_new_1974].
+
+The decrease in $\var(H)$ as $\e(R)$ increases and as $\var(R)$ decreases in figure \@ref(fig:5-5-reg) suggests that a location-scale model is likely required.
+If there is not sufficient data to fit a location scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best distribution using an information criterion such as the Akaike Information Criterion [AIC, see @akaike_new_1974].
 While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest using a GLM rather than a GAM (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic.
@@
 
-***CHANGE MODELS FOR TAPIR REG TO LOCATION-SCALE GAMMA***
-
-***ENSURE NDVI MODEL IS BASED ON DATA WITHOUT THE DEC 2017 RASTER***
-
-In this section, we use real-world tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
 We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a location-scale Gaussian GAM.
@@
 (c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir.
-(d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles.
+(d.) Estimated home range size during each seven-day period, based on 50\\% and 95\\% utilization quantiles.
 (e.) Effect of resource abundance on home range size.
@@
 The effects in panels e and f were estimated using two generalized linear models with Gamma families of distributions.
+Although a single model would be ideal, the small sample size and collinearity between $\\e(R)$ and $\\var(R)$ prevent us from estimating accurate effects of $\\e(R)$ and $\\var(R)$ within the same model (see Appendix 2).
 The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
@@
 Note how the home range size is larger during periods of lower average NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018), despite an increase in the variance in NDVI.
-In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times as much space.
-Finally, panels **e** and **f** suggest that the tapir's spatial needs change based on the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and found in the simulated responses (figure \@ref(fig:5-5-reg)).
+In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable.
+Finally, panels **e** and **f** show that the tapir's spatial needs change based on the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
 
@@
 \noindent The designation of high-quality habitats is a central concern of many conservation initiatives.
-Using this framework, one can quantify the quality of a habitat based on its resource abundance and stochasticity while also estimating the amount of space animals would need in the habitat.
-Researchers and conservation organizations can use the methods presented in this paper to assess the quality of current protected areas and predict whether or not the areas will be sufficiently large or appropriate in future decades.
-Additionally, the methods can also be used to evaluate new potential conservation areas.
-In light of Canada's pledge to conserve 30% of its landmass and waters by 2030 [@ref], the framework presented here provide useful metrics for maximizing the quality and stability of protected areas while estimating their quality in the decades to come.
+This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity while also estimating the amount of space animals would need in the habitat.
+With the methods presented in this paper, one can assess the quality of current protected areas and predict whether or not the areas will be sufficiently large or appropriate in future decades while also evaluating new potential conservation areas.
+In light of Canada's pledge to conserve 30% of its landmass and waters by 2030 [@ref], the framework and methods presented here provide useful metrics for maximizing the quality and stability of protected areas while estimating their quality in the decades to come under various climate change scenarios [@ref].
 
-Accounting for changes in resource abundance and resource stochasticity (e,g, between seasons or "good" and "bad" years) allows researchers and conservation managers can ensure the quality of protected areas is sufficiently high to sustain biodiversity for decades after the designation of the protected area.
-Protecting areas with high resource abundance and low resource unpredictability will decrease animal's spatial needs while also increasing the resiliency in home range size to small changes in resource abundance or stochasticity, since changes in spatial needs are more drastic when $\e(R)$ is low or $\var(R)$ is high (figure \@ref(fig:5-5-reg), panels e and f).
+Accounting for changes in resource abundance and resource stochasticity (e,g, between seasons or "good" and "bad" years) allows researchers and conservation managers can ensure the quality of protected areas is sufficiently high to sustain biodiversity long term.
+Protecting areas with high resource abundance and low resource unpredictability will minimize animal's spatial needs while also increasing the stability in resiliency in spatial needs, since changes in spatial needs are more drastic when $\e(R)$ is low or $\var(R)$ is high (figure \@ref(fig:5-5-reg), panels e and f).
+
+***need to add a few more ideas?***
 
@@
 We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
-We demonstrate that animals' spatial needs respond nonlinearly to both $\e(R)$ and $\var(R)$, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide way of modeling animal movement continuously rather than imposing arbitrary thresholds between range-resident and nomadic animals.
-The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals will need.
+We demonstrate that animals' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
+We also provide a way of modeling animal movement continuously rather than imposing arbitrary thresholds between different types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism).
+The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currenly need and will need in the future under different climate change scenarios.
 

### 2023-02-15 writing/conference-abstracts/2023-grs-abstract.Rmd
wrote 2023 GRS abstract

---
title: "How resource abundance and stochasticity affect animals’ spatial needs"
date: "`r Sys.Date()`"
author:
  - name: Stefano Mezzini
    email: stefano.mezzini@ubc.ca
    institute: ubco
  - name: Emília Patrícia Medici
    institute: [tapir, iucn, escas]
  - name: Michael J. Noonan
    email: michael.noonan@ubc.ca
    institute: ubco
institute:
  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
output:
  bookdown::pdf_document2:
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
    toc: false
    number_sections: false
---

# Abstract

Animals’ spatial needs are thought to be tightly linked to resource abundance within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats. Although this hypothesis has widespread empirical support, existing studies have focused primarily on animals’ responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown. This is not a trivial oversight. Animals adjust to variable environmental conditions, so failing to consider the effect of resource unpredictability can result in a limited understanding of animals’ spatial needs. In this study, we leverage the currently available information to provide a unifying framework and hypotheses for the effects of mean and variance in resources on animals’ space use. Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use. Finally, we use real-world GPS tracking data to show how this framework can be applied to better understand the movement ecology of free-ranging animals. We present our results with a fully transparent approach that allows researchers to apply this framework to their own data and support area-based conservation efforts.

### 2023-05-19 writing/conference-abstracts/2023-grs-abstract.Rmd
added more data contributors

@@
     institute: ubco
-  - name: Emília Patrícia Medici
+  - name: Adam T. Ford
+    institute: ubco
+  - name: Jacob R. Goheen
+    institute: wyom
+  - name: E. Patrícia Medici
     institute: [tapir, iucn, escas]
@@
   - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
+  - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
@@
     pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
+      - '--lua-filter=../scholarly-metadata.lua'
+      - '--lua-filter=../author-info-blocks.lua'
     toc: false

### 2023-05-20 writing/appendix-1-simulations.Rmd
added authors to the appendices

@@
     institute: ubco
-  - name: Emilia Patrícia Medici
+  - name: Adam T. Ford
+    institute: ubco
+  - name: Jacob R. Goheen
+    institute: wyom
+  - name: E. Patrícia Medici
     institute: [tapir, iucn, escas]
@@
   - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
-  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+  - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
 bibliography: 'env-var-review.bib'
@@
 
-```{r, echo=TRUE}
+```{r, echo=FALSE}
 knitr::opts_chunk$set(eval = FALSE,    # do not run R code
@@
 
-* change all references to scripts to the correct file locations
-* give an overview of how the sims work at the beginning of the pdf (including defining "encounters" with resources)
-* explain the checks in `delta-t-sensitivity.R`
-* reduce scaling of figure 2
-* fix captions (incomplete, maybe incorrect def of sims, HR not V(pos))
+* fix captions (incomplete, HR not V(pos)); take captions from manuscript
 
@@
 This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript.
-For ease of reference, we also include the figures here (Figures \@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
+For ease of reference, we also include the figures here (figures \@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
 To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
@@
 
-```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource availability ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
-The bottom line indicates the animal's core home range (0.5 quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase approximately linearly as $\\var(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 100 tracks at each time point starting from the point $\\langle 0, 0\\rangle$ until they reach satiety, at which point they returned to $\\langle 0, 0\\rangle$ over the same amount of time.
-The animal's spatial variance parameter was then calculated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package", out.width='100%'}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/mean-variance-5-by-5-hr-sims.png',
+```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource abundance ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
+The lower green line indicates the animal's core home range (50\\% quantile), while the top line indicates the 0.95 utilization quantile.
+Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase nonlinearly as $\\var(R)$ increases.
+Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they returned to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
+The animal's home range was then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
+", out.width='100%'}
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/mean-variance-5-by-5-hr-sims.png',
                         rel_path = FALSE)
@@
 ```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Linear Model with a Gamma family.
-The thicker bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the thinner top line indicates the relationship with the 0.95 utilization quantile.
-Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlineear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/simulation-regression-plots.png',
+The relationships were estimated using a Generalized Linear Model with a Gamma family of distributions that accounted for the effects of both $\\e(R)$ and $\\var(R)$ as well as differences between the two quantiles.
+The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile.
+Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlinear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-regression-plots.png',
                         rel_path = FALSE)
@@
 ```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' spatial needs were simulated.
-(a.) Animal tracks were simulated using an infinitely diffusive movement model starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
+(a.) Animal tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU; model, an infinitely diffusive and continous-velocity movement mode), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
 (b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$.
@@
 Once the animal collected sufficient resources, the animal stopped moving (i.e., the track was truncated).
-(d.) The process was repeated 100 times (13 tracks pictured in this panel).
+(d.) The process was repeated $2^{10}=1024$ times (13 tracks pictured in this panel).
 (e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% and 50\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/simulation-example.png',
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-example.png',
                         rel_path = FALSE)
@@
 
-```{r simulate-tracks, eval=TRUE}
-# NOTE: change your working directory to "env-var-review" or modify file paths
+```{r simulate-tracks, eval=TRUE, warning=FALSE}
+# NOTE: change working directory to "hr-resource-stoch" or modify file paths
+setwd('H:/GitHub/hr-resource-stoch')
 
@@
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/figures/mean-variance-trends-panel-data.R') # mu and sigma2
 source('analysis/movement-model.R') # for consistency across scripts
@@
 DELTA_T <- 60 # sampling interval in seconds
-SAMPLES <- seq(0, 60 * 60 * 12, by = DELTA_T) # 12h every DELTA_T seconds
+SAMPLES <- seq(0, 60 * 60 * 12, by = DELTA_T) # 12 hours by DELTA_T seconds
 
@@
 
-# make a single, large tibble (will need lots of RAM)
+# make a single, large tibble
 tracks <- tidyr::unnest(tracks, track)
@@
 (d.) The three tracks used in these tests over the raster used to determine when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/thinning-examples.png',
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/thinning-examples.png',
                         rel_path = FALSE)
@@
 Script: `analysis/hr-simulation-extreme-scenarios.R`
+
+In this script, we check how many tracks are necessary to produce stable and accurate estimates of the spatial needs.
+We do so by estimating the home ranges of animals in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
 
@@
 days <-
-  # modify WORST and BEST to follow the syntax used in 'offspring-simulations.R'
+  # modify WORST and BEST to follow the syntax in 'offspring-simulations.R'
   transmute(bind_rows(WORST, BEST),
@@
   mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
-         # if the animal visits a new cell, it finds food, otherwise it doesn't
+         # the animal finds food if it visits a new cell, otherwise not
          food = if_else(new_cell, food, 0)) %>%
@@
   group_by(day, animal, scenario) %>%
-  # calculate the total number of visits, total calories, and if animal is full
-  mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
-         full = satiety >= REQUIRED) %>% # did the animal reach the its needs?
+  # calculate the total visits, total calories, and if animal is full
+  mutate(satiety = cumsum(food), # for diagnostics if animals don't get full
+         full = satiety >= REQUIRED) %>% # did the animal reach its needs?
   filter(cumsum(full) <= 1) %>% # full only once
@@
     group_by(scenario, day) %>%
-    filter(full, ! duplicated(full)) %>% # take the 1st row where animal is full
+    filter(full, ! duplicated(full)) %>% # take 1st row where animal is full
     rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
@@
     geom_line(aes(group = day), color = 'blue', lwd = 30) +
-    geom_line(aes(day, timestamp + t_expl, group = day), color = 'red', lwd =30)
+    geom_line(aes(day, timestamp + t_expl, group = day), color = 'red',
+              lwd = 30)
 
@@
 saturation_days <-
-  expand_grid(n_days = (2^seq(1, log2(1e3), by = 0.2)) %>% round() %>% unique(),
+  expand_grid(n_days = (2^seq(1, log2(1e3), by = 0.2)) %>%
+                round() %>%
+                unique(),
               case = unique(days_summarized$scenario)) %>%
@@
                                          scenario == .case)),
-         tel = map(data, as.telemetry), # convert data to telemetry for modeling
+         tel = map(data, as.telemetry), # convert to telemetry for modeling
          theta = map(tel, \(x) ctmm.guess(data = x, interactive = FALSE)),
@@
            }), # fit movement model
-         sigma = map_dbl(m, \(.m) ctmm:::area.covm(.m$sigma)), # var(position)
+         sigma = map_dbl(m, \(.m) ctmm:::area.covm(.m$sigma)), # var(pos)
          hr = get_hr(.sigma = sigma, quantile = 0.95)) # Gaussian home range
@@
 
-p_hr_days <-
 ggplot(saturation_days, aes(n_days, hr)) +
@@
                      limits = c(2, 1100)) +
-  scale_y_log10(expression(Estimated~home~range~(log[10]~scale))); p_hr_days
+  scale_y_log10(expression(Estimated~home~range~(log[10]~scale)))
 ```
@@
 In both cases, 100 days are sufficient to produce stable estimates of spatial needs."}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/hr-over-days.png',
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/hr-over-days.png',
                         rel_path = FALSE)
@@
 
-2. $\tt{analysis/simulations/hr-mean-variance-simulations-days.R}$
-3. $\tt{analysis/simulations/hr-mean-variance-simulations-days-summarized.R}$
-4. $\tt{analysis/simulations/hr-mean-variance-simulations-modeling.R}$
-5. $\tt{analysis/simulations/hr-mean-variance-simulations-hrs.R}$
+1. $\tt{analysis/simulations/hr-mean-variance-simulations-days.R}$
+2. $\tt{analysis/simulations/hr-mean-variance-simulations-days-summarized.R}$
+3. $\tt{analysis/simulations/hr-mean-variance-simulations-modeling.R}$
+4. $\tt{analysis/simulations/hr-mean-variance-simulations-hrs.R}$
+6. $\tt{analysis/simulations/modeling-R-and-hr.R}$
 

### 2023-05-20 writing/appendix-2-empirical-modeling.Rmd
added authors to the appendices

@@
 ```{r, eval=TRUE, comment=FALSE}
-# NOTE: change your working directory to be "env-var-review/writing", or
+# NOTE: to replicate the analyses, you need to change your working directory
+#       to "hr-resource-stoch/writing", or
 # modify all file paths as needed
@@
 Both variables are higly temporally autocorrelated, and there is also strong correlation between the two variables."}
-knitr::include_graphics('H:/GitHub/env-var-review/figures/mean-variance-relationship.png',
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/mean-variance-relationship.png',
                         rel_path = TRUE)

### 2023-06-08 analysis/aepyceros-melampus-1757/aepyceros-melampus-1757.R
added comment on spatiotemporal interaction

@@ -59,6 +59,7 @@ if(file.exists('models/aepyceros-melampus/1757-ndvi-gaulss.rds')) {
   m_ndvi <-
     gam(list(
       # mean predictor
+      # adding an interaction of long, lat, dec_date doesn't improve much
       ndvi ~
         s(long, lat, bs = 'ds', k = 20) + # mean over space
         s(dec_date, bs = 'tp', k = 20), # need high k to account for animal adapting

### 2023-06-08 writing/manuscript.Rmd
added authors; minor edits

@@
 
-\doublespacing
-
 <!-- start main body on a new page -->
 \newpage
+
+\doublespacing
 
@@
 Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
-And while abundant work has been done on estimating animals' responses to changes in resource abundance, there is little to no information on they respond to unpredictable changes in the resources they depend on.
+While much work has been done on estimating animals' responses to resource abundance, there is little to no information on they respond to unpredictable changes in resources.
 
 In this paper, we refer to a location's average amount of available and accessible resources as **resource abundance**, while use the phrase **resource unpredictability** to indicate the variability in resource abundance due to *unpredictable* causes.
-On its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality.
-A grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources.
+We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality.
+For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources.
 In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted.
-Although both habitats may have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different habitats.
-Despite this, little work has been done on estimating the effects of resource unpredictability on animals' spatial needs [but see @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].
+Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies being selected for.
+Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-In this study, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
+In the following sections, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections, and concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology.
@@
 
-\newpage
+# Concepts and definitions
 
-# Concepts and definitions
+In this section, we presents the foundational concepts and terms used in this manuscript.
+We introduce each concept without assuming a strong statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.
 
@@
 
-\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable follows.
-Since the variable is random, it can take multiple values (minimum 2), each with different probabilities.
+\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable is assumed to follow.
+Since the variable is random, it can take multiple values, each with different probabilities.
 Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.
@@
 There are many distributions we can assign to $R$, depending on how we quantify it.
-If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $r \in (0, \infty)$.
-(We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because both 0 and infinity are not included in the support set.) In this case, we can let $R$ follow a Gamma distribution.
-If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution for $R$.
+If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$.^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] In this case, we can let $R$ follow a Gamma or log-normal distribution.
+If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
 Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
-The next two sections introduce two parameters that are often used to visualize and define distributions: the expectation and the variance of a random variable.
+The next two sections introduce two parameters that are often used to visualize, compare, and define distributions: the expectation and the variance of a random variable.
 
@@
 
-\noindent The **expectation** of a random variable (i.e.
+\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average.
+The **expectation** of a random variable (i.e.
 its mean or average, indicated here with the symbol $\e$) is the value one can *expect* the random variable to take.
-$\e(R) = \mu$ to indicate the expectation of the random variable $R$.
+Thus, we use $\e(R) = \mu$ to indicate the expectation of the random variable $R$.
 The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance.
@@
 
-The hypothesis that spatial needs decrease with resource abundance is common and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+The hypothesis that spatial needs decrease with resource abundance, $R$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
 This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
 Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero).
-The relationship between resource abundance and animals' spatial needs is explored in more depth in a following section.
+As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assume linear relationships].
+The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
+In contrast, at higher values of $\e(R)$, the same additive changes (e.g.
+from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50).
 
@@
 In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have a position it returns to regularly.
-
-\newpage
 
@@
 
-\newpage
-
 # Conclusion
@@
 We also provide a way of modeling animal movement continuously rather than imposing arbitrary thresholds between different types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism).
-The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currenly need and will need in the future under different climate change scenarios.
+The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currently need and will need in the future under different climate change scenarios.
 
@@
 Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.
-
-\newpage
 

### 2023-07-05 writing/manuscript.Rmd
completed manuscript

@@
     email: stefano.mezzini@ubc.ca
-    institute: ubco
-  - name: Adam T. Ford
-    institute: ubco
-  - name: Jacob R. Goheen
-    institute: wyom
+    institute: [braes, biol]
+  - name: Simon Wood
+    institute: uofe
+  # - name: Adam T. Ford
+  #   institute: biol
+  #   institute: braes
+  # - name: Jacob R. Goheen
+  #   institute: wyom
   - name: E. Patrícia Medici
@@
     email: michael.noonan@ubc.ca
-    institute: ubco
+    institute: [braes, biol, cmps]
 institute:
-  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
-  - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
+  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
@@
   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-bibliography: 'env-var-review.bib'
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+bibliography: 'hr-resource-stoch.bib'
 csl: 'the-american-naturalist.csl'
@@
     - \usepackage{indentfirst} # for indenting first line of each paragraph
-    # - \usepackage[small]{titlesec} # for smaller font for headings
+    - \usepackage[small]{titlesec} # for smaller font for headings
     - \usepackage{wrapfig} # for wrapping text around figures
     - \usepackage{caption} # for more customization of captions
-    # smaller caption text, less space above and below caption
     - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
     - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
-    # - \usepackage{gensymb} # for \degree, \celsius, \perthousand, \micro and \ohm
-    # - \usepackage{float} # for better floating objects interface; adds more floats     # - \usepackage{ragged2e} # for \Centering, \RaggedLeft, and \RaggedRight
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
     number_sections: false
+editor_options: 
+  chunk_output_type: console
 ---
@@
 \newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->
-
-\newcommand*\vpos{\var (\text{position})} <!-- positional variance -->
 
@@
 
-<!--
-@ciuti and @berger-tal: human presence in nature for reasons other than hunting
-"When habitats ensuring different functions (Dunning et al., 1992; Camp et al., 2013) are far from each other..."
--->
-
 # Abstract
@@
 \noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
-Although this hypothesis has widespread empirical support, existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
+This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown.
 This is not a trivial oversight.
-Animals adjust to variable environmental conditions, so failing to account for the effect of resource unpredictability can result in a limited understanding of animals' spatial needs.
+Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs.
 In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
@@
 Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use.
-Finally, we use real-world tracking data on a lowland tapir (\textit{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado <!--an impala (*Aepyceros melampus*), and multiple moose (*Alces alces*)--> to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
 Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
@@
 Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018].
-Thus, many organisms will adapt their behaviors in response to changes in local resource abundance.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], and reducing energetic costs by reducing body heat [@schmidt_interplay_2020] or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @ref; mammals: @ref], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 
-While there are many other causes of animal movement, the relationship between animal movement and resource abundance has been of interest to biologists for over half a century.
+The relationship between animal movement and resource abundance has been of interest to biologists for over half a century.
 @burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range.
-Three decades after, @southwood_habitat_1977 suggested change in resource abundance is a strong determinant of how animals decide when and where to live and reproduce.
-Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home range size is
+Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how animals decide where to live and when to reproduce.
+Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home-range size is
 
-\begin{equation} \label{harestad-eq}
+\begin{equation}
 H = C / R,
+(\#eq:harestad-eq)
 \end{equation}
 
-\noindent where $H$ is the animal's **home range size**, $C$ is the animal's resource consumption, and $R$ is the abundance of (usable) resources.
-More recently, various researches reported that home range size decreases linearly with resource abundance [desert mule deer: @relyea_home_2000; red panda: @bista_effect_2022].
-Others demonstrated that spatial needs decrease nonlinearly as habitat productivity increases, with stronger decreases when resources are scarce [e.g., Canadian lynx: @herfindal_prey_2005; carnivores: @nilsen_can_2005; tigers: @simcharoen_female_2014; ferruginous hawks: @watson_ferruginous_2020].
-Additionally, an animal's spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020].
-Thus, there are many drivers of animal movement and many causes of variety between behaviors, and we currently lack a unifying framework for quantifying the effects of resource abundance and unpredictability on animals' spatial needs.
-While much work has been done on estimating animals' responses to resource abundance, there is little to no information on they respond to unpredictable changes in resources.
+\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** an animal can access.
+However, a species' spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020].
 
-In this paper, we refer to a location's average amount of available and accessible resources as **resource abundance**, while use the phrase **resource unpredictability** to indicate the variability in resource abundance due to *unpredictable* causes.
+While many researchers have demonstrated that animals adapt heir home ranges in response to resources abundance, few studies build upon others' previous work.
+Instead, results are reported as independent, *de-novo* findings.
+Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs.
+In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little to no information on how they respond to unpredictable changes in resources.
+
+In this paper, we refer to a location's average amount of resources as **resource abundance**, while use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality.
-For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with rare, sudden, and strong pulses of resources.
-In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are exhausted.
+For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
+In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
 Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies being selected for.
@@
 
-In the following sections, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
-First, we set the theoretical background necessary for the successive sections, and concepts and notation are introduced without assuming any previous knowledge of statistical notation or movement ecology.
-Next, we provide a review of the effects of resource abundance on animals spatial needs while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource unpredictability on animals spatial needs while suggesting a second unifying hypothesis.
-Subsequently, we test the two hypotheses using quantitative simulated animal responses to changes in resource abundance and unpredictability.
-Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir living in the Brazilian Cerrado.
+Here, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
+First, we set the theoretical background necessary for the successive sections by introducing concepts and notations without assuming any previous knowledge of statistics or movement ecology.
+Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource *stochasticity* on animals spatial needs while suggesting a second unifying hypothesis.
+Subsequently, we test the two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
+Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
 Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.
@@
 
-In this section, we presents the foundational concepts and terms used in this manuscript.
-We introduce each concept without assuming a strong statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.
+<!--\noindent In this section, we presents the foundational concepts and terms used in this manuscript.
+We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.-->
 
@@
 Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Since resource abundance depends on a variety of factors, including climate, weather, competitive pressure, and individual- and species-specific energetics, it is often unpredictable (and hard to quantify).
-Thus, we can let the random variable $R$ indicate the *resource* abundance at a given point in space and time.
+Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate, weather, competitive pressure, and individual- and species-specific energetics.
+Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
 
@@
 
-\noindent In probability and statistics, random variables are defined by specifying the **distribution** the variable is assumed to follow.
-Since the variable is random, it can take multiple values, each with different probabilities.
-Commonly known distributions include the normal (i.e., Gaussian) distribution, the Student's T distribution, and the Bernoulli distribution.
-
+\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli).
+Since the variable is random, it can take multiple possible values, each with different probabilities.
 The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.
-The support of the normal and T distributions are all real numbers (i.e., from $-\infty$ to $\infty$), while the Bernoulli distribution only has support over values 0 and 1.
 
 There are many distributions we can assign to $R$, depending on how we quantify it.
-If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow any distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$.^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] In this case, we can let $R$ follow a Gamma or log-normal distribution.
-If $R$ was a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
+If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution.
+If $R$ is a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
 Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
-The next two sections introduce two parameters that are often used to visualize, compare, and define distributions: the expectation and the variance of a random variable.
 
@@
 \noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average.
-The **expectation** of a random variable (i.e.
-its mean or average, indicated here with the symbol $\e$) is the value one can *expect* the random variable to take.
-Thus, we use $\e(R) = \mu$ to indicate the expectation of the random variable $R$.
-The mean may not be within the distribution's support (e.g., a non-integer number of prey per day), but $\e(R)$ is still useful for estimating the average long-term resource abundance.
-When the mean changes over time, as in the case changes in $R$ in highly seasonal regions, we explicitly indicate that $\e(R) = \mu$ changes over time by writing the expectation of $R$ as a function of time, $t$: $\e(R) = \mu(t)$.
-We indicate the estimated average resource abundance by adding a caret on the symbol: $\hat \mu(t)$.
+The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term.
+Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$.
+When the mean changes over time, such as in strogly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\mu(t)$.
+We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
 
-The hypothesis that spatial needs decrease with resource abundance, $R$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because it can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between spatial needs and resource abundance must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime and the minimum amount of space it requires to survive (which is necessarily greater than zero).
-As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assume linear relationships].
-The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
+The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
+Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive (which is necessarily greater than zero).
+As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+<!--The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
 In contrast, at higher values of $\e(R)$, the same additive changes (e.g.
-from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50).
+from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50), so spatial needs decrease more slowly.-->
 
@@
 
-\noindent A random variable's **variance** is a measure of its unpredictability or variability.
-We use  $\var(R) = \sigma^2$ to indicate the variance in $R$, after accounting for changes in $\mu(t)$, and $\widehat{\sigma^2}(t)$ to indicate its estimate conditional on time.
-For instance, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is lower during some seasons and higher in others.
-Additionally, we distinguish between two different sources of variation.
-We refer to known differences or expected spatiotemporal changes as **heterogeneity**, while we use the term **stochasticity** for unpredictable and unexpected variation.
-For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but predictable once an animal visits the valley.
-In contrast, whether next year will be a good or bad year for berries is stochastic.
+\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$.
+A random variable's **variance** is a measure of its unpredictability (i.e., variability).
+We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$ and $\widehat{\sigma^2}(t)$ to indicate its estimate over time.
+Consequently, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
+Additionally, we distinguish between two different sources of variation: We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
+In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$.
+For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley.
+In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.
 
@@
 
-\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$ (which we write as $R \sim \Gamma(\mu(t), \sigma^2(t))$).
-Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\mu = k \theta$ and $\sigma^2 = k \theta^2$ to facilitate visualizing the examples.
-However, note that $\mu$ and $\sigma^2$ are not independent because the variance depends strongly on the mean.
-As the mean approaches zero (from the positive side, which we write as $\mu \rightarrow0^+$) the variance also does: $\mu \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \sigma^2 = 0^+$.
+\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
+Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
+However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa): As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
+This assumption also holds biologically: In the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
 
@@
 \noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011].
-However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the linear transformation
+However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the **linear transformation**
 
-\begin{equation} \label{ndvi-to-r}
-R = \frac{\nu + 1}{2},
+\begin{equation}
+\nu_{[0, 1]} = \frac{\nu + 1}{2},
+(\#eq:ndvi-to-01)
 \end{equation}
 
-where $\nu$ indicates the NDVI value.
-Note that only **linear transformations** (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906].
-In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+\noindent where $\nu$ indicates the NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
+<!--Note that only **linear transformations** (i.e.
+only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
 We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
 
-\begin{equation} \label{beta-mean}
-\mu = \frac{\alpha}{\alpha + \beta}
+\begin{equation}
+\e(R) = \frac{\alpha}{\alpha + \beta}
+(\#eq:beta-mean)
 \end{equation}
@@
 
-\begin{equation} \label{beta-var}
-\sigma^2 = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
+\begin{equation}
+\var(R) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
+(\#eq:beta-var)
 \end{equation}
 
-\noindent Failing to model or simulate resource abundance appropriately can fail to produce robust, sensible, and accurate results.
+\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):
 
+\begin{equation}
+\nu = 2 \nu_{[0, 1]} - 1.
+(\#eq:01-to-ndvi)
+\end{equation}
+
+\noindent Consequently, we have
+
+\begin{equation}
+\e(\nu|t) = 2 \mu(t) - 1.
+(\#eq:mu-to-e-ndvi)
+\end{equation}
+
+\noindent and
+
+\begin{equation}
+\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
+(\#eq:sigma2-to-v-ndvi)
+\end{equation}
+
+\noindent Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
+Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
+
+<!--
 ## Animal movement as a stochastic process
@@
 
-\noindent @nathan_movement_2008 generalized animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
+\noindent @nathan_movement_2008 suggested we can generalize animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
 
-\begin{equation} \label{nathan-movement}
+\begin{equation}
 \vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
+(\#eq:nathan-movement)
 \end{equation}
@@
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., body size, memory, cortisol levels), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on animals' behavior and spatial needs as a response to overall resource abundance and unpredictability.
+The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
+Thus, we can instead focus on animals' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
 
@@
 This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
-Similarly, we can define $\tau_v$ to be amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity.
+Similarly, we can define $\tau_v$ to be the amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
 $\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
@@
 
-Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the dataset must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
-However, since the time required for an animal to cross the area it uses is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home range size, provided that the total **sampling duration**, $T$, is sufficiently long.
-Model selection procedures with criteria such as AIC [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
+Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
+However, since the time required for an animal to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
+Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
 
-Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less *why* it moved.
+Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less on *why* it moved.
 Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
+Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in animal movement while focusing on the animals' movement behaviors rather than the causes.
 
@@
 \noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 50% and 95% home range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.
-In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs using a sliding window approach (see the discussion section).
-
-**I don't think we need this paragraph:** Although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's spatial needs but not its behavior.
+Although here we quantify animal's spatial needs using 50% and 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Note, however, that although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's movement but not its behavior.
 Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
 Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have a position it returns to regularly.
+In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
+-->
 
@@
 
-```{r mu-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of mean resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high $\\vpos$ as animals are forced to explore greater areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
 As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
-Note that the relationship between $\\e(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 ```{r, eval=FALSE, echo=FALSE}
-# to help visualize the consequences of different means and variances
+# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
 dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
@@
 All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources is likely to result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
+Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
 
-The size of an animal's spatial needs is often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs increase when resources are low, but the relationship is likely not linear, as illustrated in figure \@ref(fig:mu-hyp) [also see: @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022].
-Low values of $\e(R)$ force animals to have larger home ranges [@relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales rather than an animal's lifespan.
+An animal's spatial needs are often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs are high when resources are scarce and low when resources are abundant, but, as noted above, the relationship is necessarily nonlinear, as illustrated in figure \@ref(fig:mu-hyp) [also see:  @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+Low values of $\e(R)$ force animals to have larger home ranges [ @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
 
-As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [see @relyea_home_2000; @bista_effect_2022, although the latter uses models which assume linear relationships].
-The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
-In contrast, at higher values of $\e(R)$, the same additive changes (e.g.
-from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50).
+Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+While competition may push animals to explore other areas and expand their home range [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
+It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose: @singh_migration_2012; eagles: @wheat_migrate_2017; @poessel_interpreting_2022; flamingos: @pretorius_movement_2020].
 
-Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], and larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-While competition may push animals to explore other areas and expand their home range [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
-
-It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between the types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose: @singh_migration_2012; eagles: @wheat_migrate_2017; @poessel_interpreting_2022; flamingos: @pretorius_movement_2020].
-From a quantitative perspective, the switch is related to an animal's home range crossing time (or positional autocorrelation, $\tau_p$).
-As the amount of space an animal uses over its lifespan, the amount of time required to cross the area ($\tau_p$) will also increase, so animals with $\tau_p$ on the order of their expected lifespan will necessarily be nomadic.
-
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with constant variance (after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, raster fill) varies over time and space, the overall amount of variance (dot color) remains constant.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\e(R)$ and average $\\var(R)$.
-(d.) Simulated animal movement at a time of peak $\\e(R)$ and average $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
+(a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
 ```
-Consider the simplistic scenario where the average in resource abundance, $\e(R) = \mu$, changes over time but $\var(R) = \sigma^2$ is constant over time and space (and non-zero, see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
-Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat (figures \@ref(fig:multipanel-constant-var)c-d).
+
+Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
 However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and decreases in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017; @abrahms_memory_2019] although it will not be possible to predict where high-$R$ locations are, since $\e(R)$ does not vary predictably over space.
+peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
 
 This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
-Such a model may be appropriate in regions with little to no changes in variability and where unpredictable events occur at a similar frequency over time, or when data availability is too low to produce appreciable measures of variance.
+Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
 Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
@@
 Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
-Additionally, when data availability is sufficiently high to produce precise estimates of $\var(R)$, the effect of $\var(R)$ on animals' spatial needs should be estimated before assuming the effect is negligible.
-The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018].
+The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018; @yao_emergence_2022].
 
-Anthropogenic changes, including human-built structures and urban areas reduce the habitat available to many terrestrial species [@wilson_habitat_2016], who struggle to move in fragmented [@fahrig_non-optimal_2007], human-dominated landscapes [@tucker_moving_2018].
-Although some species rely on anthropogenic areas and structures for predicable and dependable sources of resources [wild boar: @ikeda_home_2022; elephants: @benitez_precipitation_2022; steppe rat snake: @yu_artificial_2022; yellow mongoose: @cronk_home_2021; common octopus: @arechavala-lopez_common_2019; raccoon dog: @mitsuhashi_home_2018; coyote: @peron_periodic_2017], anthropogenic environments can be highly unpredictable and pose high mortality risks [giant anteaters: @noonan_roads_2021; killer whales: @matkin_ongoing_2008].
-Although it is generally acknowledged that animals will modify their behavior in response to human activity [@rogala_human_2011] as well as changing conditions [@walther_ecological_2002; @chevin_adaptation_2010; @sih_evolution_2011], a fine-scale understanding of behavioral responses is lacking [@sih_evolution_2011; @beever_behavioral_2017].
+# Effects of resource stochasticity, $\var(R)$
 
-# Effects of resource unpredictability, $\var(R)$
+\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as infrequent and unpredictable events.
+For example, berries are predictably scarce out of the growing season, but their abundance more stochastic when in spring and summer since the timing between plants can depend on many unpredictable factors, and an animal may explore a location only after competitors ate all the berries.
+Thus, while a bear cannot to expect to find berries during growing season, it can expect the abundance in berries to be more stochastic.
+In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@noonan_situ_2018; @berger_climate_2018; @grant_evolution_2017; @yao_emergence_2022; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
 
-```{r s2-hyp, fig.cap="Hypothesized spatial needs (indicated as positional variance, $\\vpos$) of an animal as a function of resource unpredictability ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low $\\vpos$ as animals are able to depend on somewhat predictable resources.
+```{r s2-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource stochasticity ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low spatial needs as animals are able to depend on somewhat predictable resources.
 As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and $\\vpos$ cannot be linear because it would require $\\vpos$ to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], $\var(R)$ has received far less attention [but see @lucherini_habitat_1996; @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can change strongly due to many important causes, including repetitive and predictable patterns as well as infrequent and unpredictable events.
-Examples of predictable changes in $\var(R)$ include changes that occur between seasons, such as changes in temperature in continental summers (e.g., between 15 and 30 $^\circ$C) and winter temperatures (e.g., from -40 to -10$^\circ$C within the span of a few days), changes in precipitation frequency and unpredictability between the dry season (predictable lack of rain) and the wet season (frequent but sudden and unpredictable rains), or changes in fruit and berry abundance with the arrival of spring and summer (since the timing between plants can depend on many unaccounted factors).
-Although each of these events is unpredictable, the change in unpredictability can be predicted and expected.
-In contrast, other sources of stochasticity can be much less predictable, such as new competitors [@ref], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], droughts [@foley_severe_2008; @haig_assessment_2013], and changes in climate [@ipcc_global_2018; @noonan_situ_2018; @berger_climate_2018; @grant_evolution_2017; @yao_emergence_2022].
-
-Environmental variability, including extreme events [@logares_black_2012; @anderson_black-swan_2017; @yao_emergence_2022], can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
-A few recent studies support this hypothesis [@morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
+Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
+While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
 Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
@@
 Similarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)).
-This is because as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly.
-Additionally, stochastic environments tend to be less productive [@@chevin_adaptation_2010], and the added movement required to search for food increases animals' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage, food caching, torpor, and hibernation).
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
+If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
 
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resource abundance in a heterogeneous environment with changing variance (even after accounting for changes in mean abundance).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, variance (dot color) is lowest at the beginning and end of the observational period and highest when $R$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with mean $\\mu$ and variance $\\sigma^2$.
-(c.) Simulated animal movement at a time of lowest $\\e(R)$ and $\\var(R)$.
-(d.) Simulated animal movement at a time of peak $\\e(R)$ and $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
+(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
@@
 
-When the unpredictability in $R$ chances over time (figure \@ref(fig:multipanel-changing-var)a), changes in $\var(R)$ may have an appreciable effect on the location's favourableness.
-In this scenario, we can define $\var(R)$ as a function of time using the notation $\var(R) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)b).
-When both $\e(R)$ and $\var(R)$ change over time, accurately and precisely predicting $R$ becomes more complex (figure \@ref(fig:multipanel-changing-var)c-d).
+Like the scenario illustrated in figure \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
+We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)).
+When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
+Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
+Biologically, this is important because some animals may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
+An animal's perception of changes in $R$ will depend strongly on the animal's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
+The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
 
-Most habitats with appreciable seasonal changes in climate will require models to account for changes in both $\e(R)$ and $\var(R)$, as they often change within and between years, and periods of high productivity can often be more variable than times of lower productivity or dormancy.
-During times of resource unpredictability, animals will likely have higher spatial needs, particularly if they need to collect and store resources for periods scarcity or increased energetic needs.
-These include higher energetic needs during reproductive periods, whether this be due to pregnancy [@schmidt_interplay_2020], to feed offspring [@saraux_plasticity_2011], or because the animal's needs change over time [e.g., male deer growing antlers: @french_nutrient_1956; @smolko_soil_2022].
-Failing to account for changes in $\var(R)$ can result in serious underestimations of animals' spatial needs, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
-
-The effects of $\var(R)$ on animals' spatial needs likely vary between species, since energetic and nutritional needs vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
-@rizzuto_forage_2021 found that the 50%, 75%, and 90% utilization distributions of snowshoe hares (*Lepus americanus*) increased with average C:N in lowbush blueberry as well as the coefficient of variation (i.e., standard deviation over mean) in C:N, which suggests that the hares expanded their home range in response to both nitrogen abundance and nitrogen variability.
-Similarly, @mueller_how_2011 demonstrated that ungulates in more variable ecosystems move more and are more likely to be nomadic.
-In contrast, @lucherini_habitat_1996 found that red foxes (*Vulpes vulpes*) living in areas with more more heterogeneous habitats required exponentially less space than foxes that lived in more homogeneous regions.
-Similarly, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality.
-A subsequent analysis by @nilsen_can_2005 with the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*), found that carnivores respond differently to resource variability and seasonality, although it is worth disentangling the effects of heterogeneity (i.e., changes in $\e(R)$) and and stochasticity ($\var(R)$).
-
-@nilsen_can_2005 also demonstrate that the effect of $\var(R)$ on the home range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on $\e(R)$.
-Therefore, it is not only important to account for both $e(R)$ and $\var(R)$ but also interaction effects between the two.
-It is also worth noting that carnivores may be more susceptible to $\var(R)$ since animal prey's ability to move makes their abundance and location less predictable than those of plants.
-Since, the effects of $e(R)$ and $\var(R)$ may also affect prey abundance, it is important to account for both $e(R)$ and $\var(R)$ as well as any interactive effects between the two and how the effects may vary between animals with different diets or from different trophic levels, since animals from higher trophic levels will be affected by the cascading effects of $\e(R)$ and $\var(R)$ on each of the lower trophic levels [@harestad_home_1979; @noonan_effects_2020].
+```{r stoch-timescale, eval = FALSE, fig.cap="Conceptualizations of perceptions of $\\e(R)$ (a.) and $\\var(R)$ (b.) over time for animals with no plasticity in their movement (blue), seasonal memory but no plasticity (orange), or high plasticity but no memory (green).
+The true trends in $\\e(R)$ and $\\var(R)$ are indicated in black in panels (a.) and (b.), respictively.
+Estimates in the white area indicate the animal's estimated $\\mu(t)$ and $\\sigma^2(t)$ based on past experience, while estimates to the grey shaded region indicate the animals' predictions of $\\mu(t)$ and $\\sigma^2(t)$ in the future.
+The four facets show how each behavior performs in environments where $\\e(R)$ is either predictable (top row) or stochastic (bottom row), and $\\var(R)$ is either predictable (left column) or stochastic (right column).
+Animals that do not adapt to changes in $R$ (blue) are unable to predict $\\e(R)$ accurately, and they have high estimates of $\\var(R)$ when $\\mu(t)$ is far from the long-term, temporally static average.
+Memory of seasonal trends in $\\mu(t)$ and $\\var(R)$ (orange) allows animals to perform well in predictable environments, but their lack of adaptability causes them to fair poorly in stochastic habitats, where they perceive $\\var(R)$ to be high when $\\e(R)$ is far from the expected $\\mu(t)$.
+Finally, the ability to adapt rapidly to change (green) allows animals to track changes in $\\e(R)$ and $\\var(R)$ closely, but the lack of memory prevents accurate predictions in predictable environments.
+Since such animals tend to perceive changes in $R$ as changes in $\\mu(t)$, they tend to have lower estimates of $\\sigma^2(t)$ than other animals."}
+# set eval to false: include in a separate paper?
+knitr::include_graphics('../figures/stochasticity-timescale.png')
+```
 
@@
 
-\noindent In habitats where both $\e(R)$ and $\var(R)$ change substantially over time, models should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
-A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ will have little effect if resources are always sufficiently abundant.
+\noindent In systems where both $\e(R)$ and $\var(R)$ change substantially over time, animals may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
+Consequently, models estimating the effects of $R$ on animal's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
 Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
 
-Figure \@ref(fig:5-5-sims) illustrates simulated spatial requirements as a function of $\e(R)$ and $\var(R)$.
-The top row (constant $\var(R)$) shows how simulated spatial needs vary for different trends in $\e(R)$ while $\var(R)$ remains constant.
-As $\e(R)$ increases at a constant slope (linear $\e(R)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home range size approaches the minimum size required by the animal.
-As described in the section on $\e(R)$, the nonlinear decrease is because changes in $\e(R)$ have a larger effect when $\e(R)$ is low.
-Additionally, the 95% home range is more sensitive to changes in $\e(R)$ than the core home range simply because larger quantiles (e.g., 95% > 50%) are more sensitive to changes in the distribution.
-Additionally, when $\e(R)$ is lower, $\var(H)$ is higher (particularly for the 95% quantile), since resource-poor regions force animals to move more, which results in a greater uncertainty in a longer and thus also more unpredictable search.
+<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
+However, the effects of $\var(R)$ on animals' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
+Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
+-->
 
-```{r 5-5-sims, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource abundance ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
-The lower green line indicates the animal's core home range (50\\% quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase nonlinearly as $\\var(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they returned to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's home range was then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
-", out.width='100%'}
+@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\mu(t)}{\sqrt{\sigma^2(t)}}$) in C:N.
+However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
+Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality).
+Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
+A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons<!-- (\@ref(fig:stoch-timescale))-->.
+Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
+
+@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on seasonality (and vice-versa): Animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
+Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
+This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but the they may not respond if $\mu(t)$ is sufficiently high.
+However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
+
+## Simulating animal responses to $\e(R)$ and $\var(R)$
+
+To illustrate how $\e(R)$ and $\var(R)$ affect animals' spatial needs as well as the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations in figure \@ref(fig:5-5-sims) (for methodological details, see Appendix 1).
+The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the animal.
+As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
+
+```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
+Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
+The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
+$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
+Additional information is provided in Appendix 1.", out.width='100%'}
 knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
@@
 
-In regions where the average resource abundance changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\e(R)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\e(R)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
+In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
+However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
 Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\e(R)$), since accurate long-term estimates of $\e(R)$ may be hard to produce, if not impossible.
-In cases where $\e(R)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\e(R)$).
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
+In cases where $\mu(t)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
 
 The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
-Overall, both quantiles increase with $\var(R)$, although $H_{95\%}$ is more sensitive to changes in $\var(R)$ than $H_{50\%}$, as with changes in $\e(R)$.
-Additionally, note how both $H$ and $\var(H)$ increase with $\var(R)$ (e.g., constant $\e(R)$, linear $\var(R)$).
-This is because, similarly to resource-poor areas, more stochastic times require the animal to move for longer periods of time, which results in a greater uncertainty in how much time and space it will require to reach satiety.
+Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
+This is because, similarly to resource-poor periods, times of greater stochasticity require the animal to move over larger areas for longer periods of time.
+This results in a greater uncertainty in how much time and space the animal will require to reach satiety, or indeed whether an animal living in highly stochastic environments can even reach satiety within a finite amount of time.
 
 The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
-Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change monotonically, especially when the trends are linear (i.e., linear $\e(R)$ and $\var(R)$).
-When both $\e(R)$ and $\var(R)$ increase linearly, the change in $H_{95\%}$ is small an not monotonic, while there is no clearly visible change in $H_{50\%}$.
-However, when $\e(R)$ changes nonlinearly (e.g., cyclical $\e(R)$ with linear $\var(R)$), it is easy to see how the increase in $\var(R)$ amplifies the effects of $\e(R)$ (and vice-versa).
-Similarly, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
-Thus, the estimated spatial needs are largest when $\e(R)$ is low and $\var(R)$ is high, and they are smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
+Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
+When both $\e(R)$ and $\var(R)$ increase linearly, $H$ initially increases since the effect of $\var(R)$ is stronger, but then decreases as the effect of $\e(R)$ begins to dominate.
+However, when $\mu(t)$ and $\sigma^2(t)$ follow fairly different trends (e.g., cyclical $\e(R)$ and linear $\var(R)$), it is easy to see how an increase in $\var(R)$ amplifies the effects of $\e(R)$.
+In contrast, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
+Thus, $H$ is largest when $\e(R)$ is low and $\var(R)$ is high, and it is smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
 
 Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
-$\e(R)$ and $\var(R)$ can be assumed to be (approximately) constant in highly homogeneous environments, or environments where resources are sufficiently abundant that changes in $\e(R)$ and $\var(R)$ are relatively small and remain undetected.
-Although it is impossible for $\e(R)$ and $\var(R)$ to increase linearly forever, such increases may be possible for short periods of time (followed by periods of no change or decrease).
+$\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
+Although it is impossible for $\mu(t)$ and $\sigma^2(t)$ to increase linearly forever, these examples are useful to demonstrate that linear changes in $\mu(t)$ and $\sigma^2(t)$ affect $H$ nonlinearly.
 Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
@@
 However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
-Estimating trends in $\e(R)$ and $\var(R)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
 
-Although the trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ do not seem easy to disentangle, both $\e(R)$ and $\var(R)$ affect spatial needs as we hypothesized (figure \@ref(fig:5-5-reg)).
-In the following discussion section, we detail how this framework can be applied to empirical data and used for conservation decision-making.
+Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either E(R) or V(R), rather than time.
+Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
+We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.-->
+
+We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
+Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
+Secondly, the simulated animal's range residency with a consistent (albeit stochastic) movement process within a homogeneous environment.
+Since real-world landscapes are heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event such as a fire or a flood.
+Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
+Additionally, the simulations also ignore the how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
+While these are definitely important effects that should be accounted for [@jetz_scaling_2004], we believed that doing so was beyond the scope of this paper.
 
 ```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Linear Model with a Gamma family of distributions that accounted for the effects of both $\\e(R)$ and $\\var(R)$ as well as differences between the two quantiles.
-The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile.
-Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlinear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions.
+The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
+Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the approximately nonlinear increase in $H$ as $\\var(R)$ increases.
+Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
 knitr::include_graphics('../figures/simulation-regression-plots.png')
 ```
-
-# Discussion
-
-\noindent This paper presents a new unifying framework for estimating the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
-Instead, one can test whether members of the same population or species behave similarly, and whether all members of the group have a common functional response using hierarchical models [@pedersen_hierarchical_2019].
-Individuals and populations that were thought to have different behaviors and spatial needs may instead have a single common behavior as they respond to $\e(R)$ and $\var(R)$ following the same functional response [@examples].
-
-In this discussion section, we make considerations about the effects of $\e(R)$ and $\var(R)$ on various animal adaptations, including movement behaviors, energy storage, and spatiotemporal memory.
-We then illustrate how one may apply this framework to their own data by providing an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
-Finally, we demonstrate how to use this framework to create products that can inform conservation-based decisions, including predicting animals' spatial needs and assessing the quality of potential (and current) conservation areas.
-
-## Animal adaptations to changes in $\e(R)$ and $\var(R)$
-
-\noindent The spatiotemporal scales over which an event occurs are a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism [i.e.
-approximately one or more times per lifetime and within the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019].
-Smaller and short-lived organisms (e.g., mice) are more likely to be be severely impacted by stochastic events than larger ones (e.g.
-elephants), since larger organisms can have bigger energy reserves [@lindstedt_seasonality_1985], can move longer distances over short periods of time [@hirt_general_2017], and tend to have longer lifespans, generation times, and developmental periods [@brown_toward_2004] that allow them to adapt their behavior to such events, including the development of memory [framework: @riotte-lambert_environmental_2020; elephants: @foley_severe_2008; @polansky_elucidating_2015; bison and mule deer: @geremia_migrating_2019; elk: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; blue whales: @abrahms_memory_2019].
-
-The short generation time and high reproductive rate of smaller, *r-selected* species [@pianka_r-_1970; @brown_toward_2004] may also promote traits that increase survival following frequent extreme events by increasing population survival rates through higher population size, dispersal, and genetic diversity [@examples].
-Events that occur over large spatiotemporal scales can promote slow changes through adaptation, natural selection, and evolution [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which are numerically more likely if the starting population is large with abundant offspring [@examples; but see @leung_smaller_2022].
-
-Events which occur multiple times at relatively predictable intervals during an animal's lifespan and over multiple generations (such as the alternating of seasons) allow the animal to anticipate times of scarcity and prepare for them by, for instance, increasing its fat storage [marmots: @armitage_energetics_2003; bears: @nespolo_why_2022], caching resources [@post_caching_2006], or migrating [@geremia_migrating_2019; @mueller_social_2013].
-Of course, an animal with a lifespan shorter than a year will be unable to perceive the cyclical alternating of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather and temperature other than lower temperatures at night than during the day.
-
-When an animal's expected lifespan is longer than a year, preparing for predictable times of abundance and scarcity becomes an advantage (and often a necessity).
-In such cases, animals can increase their chances of survival via physiological and behavioral adaptations, including adapting the timing of reproduction [@ref] and torpor [marmots: @inouye_climate_2000; ground squirrels: @goldberg_hibernation_2021], as well as memory [framework: @riotte-lambert_environmental_2020; @fagan_spatial_2013; roe deer: @ranc_memory_2022; caribou: @cavedon_genomic_2022; elk: @falcon-cortes_hierarchical_2021; elephants: @polansky_elucidating_2015], which is required for migration [butterflies: @stefanescu_timing_2021; whooping cranes: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018].
-However, relying on environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
-
-An animal's fitness and chances of survival in a changing environment depend strongly on its adaptability and responsiveness to change [@riotte-lambert_environmental_2020].
-An animal's ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on the animal's chance of survival and reproduction.
-For example, ungulates' ability to time their yearly migrations on available forage biomass, snow fall, snow melt, and hunting pressure allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time and energy spent exploring.
-Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences $\e(R)$ between years [@jonzen_rapid_2006].
-However, relying strongly on short-term changes rather than depending on memory can also cause animals to fail to find crucial resources [@polansky_elucidating_2015; @foley_severe_2008].
-
-Recent changes in climate during the last few decades have caused some species to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@examples].
-While range shifts can help species overcome changing conditions, they can still result in phenological mismatches [@lameris_arctic_2018], and they also add pressure to the species that already live in the habitat.
-Additionally, specialists and obligate symbionts are less likely to shift their range since it is determined by their food sources' and associates' ability to move and adapt, too.
-Consequently, the arrival of new species may exacerbate the stressors such animals are already facing.
-
-<!--
-Notes:
-
-@rickbeil_plasticity_2019:
-- Many species use seasonal migration to track emerging resources (Fryxell, Greever, & Sinclair, 1988), reduce predation risk (Hebblewhite & Merrill, 2009), or avoid disease and parasites (Altizer, Bartel, & Han, 2011).
-- Migrations to track resources are often timed to coincide with peak plant and insect emergence (Aikens et al., 2017; Jonzén et al., 2006)
-- there is growing evidence that some [bird] populations are adjusting their behavior, suggesting that migration, specifically migration timing, may be more plastic than previously expected (Jonzén, Hedenstrom, & Lundberg, 2007; Jonzén et al., 2006) and highlighting a potential measure of resilience in the face of climate change
--->
 
@@
 We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
-Next we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
+Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
 Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.
@@
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly-used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017], including location-scale GLMs and GAMs.
+`mgcv` [@wood_generalized_2017] is a commonly used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017], including location-scale GLMs and GAMs.
 Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), scaled t, gamma, Tweedie, and Gumbel.
 
-The Gaussian family of distributions is the most flexible, since the mean and standard deviation parameters are assumed to be independent, and the response can be either positive or negative.
-The scaled t family is similar to the Gaussian family, but it is best for high-variance data with heavy tails.
-However, the two distributions' flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-area), count data, proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
+The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
+However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
+available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
 
 The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
-The Tweedie family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey abundance.
-Lastly, the Gumbel distribution is suitable for modeling the extremes of a distribution, such as maximum or minimum daily temperatures.
-
-If one is interested in using other families of distributions, including zero-inflated and hurdle models, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of the family's parameters using a fully Bayesian approach [@burkner_advanced_2018].
-In this paper, we use the `mgcv` package because although `brms` provides more flexibility and a greater choice of families, its computational costs can be prohibitive for large datasets, particularly for location-scale models.
+The Tweedie location-scale family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density.
+In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family.
+This family is not available in `mgcv` at the time of publication, but it should be available soon.
+If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
 
 Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and nonmonotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatio-temporal terms should be chosen carefully, particularly for mean term.
-An excessively wiggly $\hat \mu(t)$ will cause $\widehat{\sigma^2}(t)$ to be under-estimated and under-fit, and an excessively smooth $\hat \mu(t)$ will inflate $\widehat{\sigma^2}(t)$.
-Thus, models should be fit with care to avoid inaccurate estimates.
-Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt should be sufficient.
+Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the location terms.
+An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
+Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
 Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
@@
 Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size).
+Appendix 2 contains additional information on how NDVI was modeled for the empirical example, along with the continuous time models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
 
-### Modeling spatial needs
+<!-- ### Modeling spatial needs -->
 
-\noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017; @wheat_migrate_2017].
-However, most commonly-used movement models assume animals' behaviors to remain consistent during the entirety of the tracking period [but see state-space models, e.g., @polansky_elucidating_2015].
+<!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
+However, most commonly used movement models assume animals' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
+-->
 
-To quantify an animal's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
-The width of the window, $w$, should be a fixed value greater than two or three times the animal's expected range crossing time [@ref].
-If the animal's spatial use is relatively consistent over time, one may fit a movement model to the whole dataset, and then choose $w$ based on the model's $\tau_p$.
-Alternatively, if the animal's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of 2 or 3 range crossings (i.e., $w \approx 2.5 \,\tau_p$).
-Modeling animal movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
+<!-- To quantify an animal's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
+The width of the window, $w$, should be a fixed value greater than two or three times the animal's expected range crossing time [@fleming_overcoming_2019].
+If the animal's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
+Alternatively, if the animal's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
+-->
+
+<!-- Modeling animal movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
 In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022].
+-->
 
-### Modeling effects of $\e(R)$ and $\var(R)$
+### Modeling the effects of $\e(R)$ and $\var(R)$
 
-\noindent As discussed in the section on modeling $R$, the first step to modeling $H$ is to choose an appropriate distribution.
+\noindent As discussed in the precious section, the first step to modeling $H$ is to choose an appropriate distribution.
 Since $H$ is strictly positive and continuous, a Gamma distribution seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017].
-The Tweedie distribution would also be appropriate if the mean-variance relationship of the Gamma distribution does not fit the data well.
+The Tweedie distribution may also be appropriate, if the mean-variance relationship of the Gamma distribution results in a poor fit.
 
-The decrease in $\var(H)$ as $\e(R)$ increases and as $\var(R)$ decreases in figure \@ref(fig:5-5-reg) suggests that a location-scale model is likely required.
-If there is not sufficient data to fit a location scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best distribution using an information criterion such as the Akaike Information Criterion [AIC, see @akaike_new_1974].
-While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest using a GLM rather than a GAM (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic.
-Additionally, restricting the model to a GLM reduces the risk of issues that may arise due to limited sample size and temporal autocorrelation in $\e(R)$ and $\var(R)$.
-If one interested in fitting a GAM but imposing shape restrictions, such as monotonicity, we sugget using the `scam` package [@pya_scam_2022].
+As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
+If there is not sufficient data to fit a location scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
+While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest one use a GLM, instead, (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic.
+Restricting the model to a GLM reduces the risk of issues that may arise due to limited sample size and temporal autocorrelation in $\e(R)$ and $\var(R)$.
+If one is interested in fitting a GAM but imposing shape constraints, such as monotonicity, we suggest fitting a shape-constrained GAM via the `scam` package [as we do here: see the section below, Appendix 2, and @pya_scam_2022].
 
@@
 
-In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
-We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a location-scale Gaussian GAM.
-Although a beta model would be more appropriate, the `mgcv` package does not currently support location-scale beta models.
-While the `brms` model supports beta location-scale models, the dataset is sufficiently large and with data sufficiently away from the boundaries (i.e., -1 and 1) that we expect the estimates to be reasonably accurate for an illustrative example.
+\noindent In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a Beta location-scale GAM, after scaling NDVI to [0, 1] using equation \@ref(eq:ndvi-to-01).
 Additional information regarding the modeling can be found in Appendix 2.
 
-```{r tapir-mw, fig.cap="Seven-day home range size of a tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
-(a.) GPS tracking data of the tapir.
-(b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir.
-(c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir.
-(d.) Estimated home range size during each seven-day period, based on 50\\% and 95\\% utilization quantiles.
-(e.) Effect of resource abundance on home range size.
-(f.) Effect of resource unpredictability on home range size.
-The effects in panels e and f were estimated using two generalized linear models with Gamma families of distributions.
-Although a single model would be ideal, the small sample size and collinearity between $\\e(R)$ and $\\var(R)$ prevent us from estimating accurate effects of $\\e(R)$ and $\\var(R)$ within the same model (see Appendix 2).
+```{r tapir-mw, fig.cap="Seven-day home-range size of a tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
+(a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
+(b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir.
+(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
+(d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
+The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
+(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
+Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
+See Appendix 2 for additional information.
 The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
@@
 Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
-Panel **a** provides an overview of the tracking data, while panels **b** and **c** show the changes in mean and variance in NDVI, respectively.
-Panel **d** shows the changes in the tapir's spatial use over time.
-Note how the home range size is larger during periods of lower average NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018), despite an increase in the variance in NDVI.
+Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period.
+Panel **c** shows the changes in the tapir's spatial use over time.
+Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018).
 In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable.
-Finally, panels **e** and **f** show that the tapir's spatial needs change based on the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
+Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
+Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
+
+# Discussion
+
+\noindent This paper presents a new unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
+The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
+Instead, one can test whether individuals follow these hypotheses, if members of the same population or species behave similarly, and if all members of the group have a common functional response [using hierarchical models, see @pedersen_hierarchical_2019].
+Individuals and populations that were thought to have different behaviors and spatial needs due to differences in behavior may instead be simply responding to different conditions along the same functional response [e.g., @singh_migration_2012].
+
+## Animal adaptations to changes in $\e(R)$ and $\var(R)$
+
+\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+Events which occur at the scale of the organism [i.e.
+approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants, *Loxodonta sp.*: @foley_severe_2008; muskoxen, *Ovibos moschatus*: @berger_climate_2018; bison, *Bison bison*, and mule deer, *Odocoileus hemionus*: @geremia_migrating_2019].
+Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
+However, adapting to changes in environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
+
+In contrast, individuals with short lifespans are less likely to experience stochastic events and memory about them.
+However, the short generation time and high reproductive rate of *r-selected* species [*sensu* @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events.
+Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021].
+Events that occur over large spatiotemporal scales can thus promote slow changes through adaptation, natural selection, and evolution of the species rather than the individual [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which are numerically more likely if the starting population is large and genetically diverse, as in the case of r-selected species [but see @leung_smaller_2022].
+Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than lower temperatures at night than during the day and some weather stochasticity.
+Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].
+
+A population's fitness and chances of survival in a changing environment depend strongly on its adaptability and responsiveness to change [@riotte-lambert_environmental_2020], and its ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on its members' chances of survival and reproduction.
+For example, ungulates' ability to time their yearly migrations on available forage biomass, snow fall, snow melt, and hunting pressure allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time and energy spent exploring.
+Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences in $\mu(t)$ between years [@jonzen_rapid_2006].
+However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources [cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
+However, when anomalies occur repeatedly over time, animals may begin to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
+Range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021].
+However, they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [@wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range since it is determined by their food sources' and associates' ability to move and adapt, too.
 
@@
 
-\noindent The designation of high-quality habitats is a central concern of many conservation initiatives.
-This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity while also estimating the amount of space animals would need in the habitat.
-With the methods presented in this paper, one can assess the quality of current protected areas and predict whether or not the areas will be sufficiently large or appropriate in future decades while also evaluating new potential conservation areas.
-In light of Canada's pledge to conserve 30% of its landmass and waters by 2030 [@ref], the framework and methods presented here provide useful metrics for maximizing the quality and stability of protected areas while estimating their quality in the decades to come under various climate change scenarios [@ref].
+\noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on animal behavior.
+In these times of great change and climatic uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021], it is increasingly important to recognize how environmental stochasticity affects animal behavior and shapes ecosystems.
+Accounting for changes in both resource abundance and resource stochasticity is crucial when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
+Protecting areas with high resource abundance and low resource stochasticity will reduce animal's spatial requirements.
 
-Accounting for changes in resource abundance and resource stochasticity (e,g, between seasons or "good" and "bad" years) allows researchers and conservation managers can ensure the quality of protected areas is sufficiently high to sustain biodiversity long term.
-Protecting areas with high resource abundance and low resource unpredictability will minimize animal's spatial needs while also increasing the stability in resiliency in spatial needs, since changes in spatial needs are more drastic when $\e(R)$ is low or $\var(R)$ is high (figure \@ref(fig:5-5-reg), panels e and f).
+The designation of high-quality habitats is a central concern of many conservation initiatives.
+This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat.
+With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas.
+In light of the UN's objctive to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
 
-***need to add a few more ideas?***
+Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
+Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' spatial needs, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
+This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in spatial needs.
 
-<!-- create a pdf for supplementary information on how to create the 5x5 panel and run methods w real data, as well as considerations and potential issues about modeling means and variances -->
+It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels.
+While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity.
+In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants.
+Consequently, the abundance of primary producers may correlate with habitat productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019].
+Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay.
+Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
 
@@
 
-\noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance.
+\noindent The work presented here provides a unifying framework for viewing animal movement as function of resource abundance and stochasticity.
 We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
 We demonstrate that animals' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a way of modeling animal movement continuously rather than imposing arbitrary thresholds between different types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism).
+We also provide a moving window approach that allows one to model animal movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism.
 The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currently need and will need in the future under different climate change scenarios.
@@
 \newpage
-
-# MISC
-
-Extreme events [@logares_black_2012; @anderson_black-swan_2017] such as heat domes [@emerton_predicting_2022], hurricanes [@widmer_effects_2004]; severe storms [@berger_climate_2018], fires [@ruthrof_how_2016], new diseases [@murray_pathogens_2006; @hollings_trophic_2014; @hoyt_ecology_2021], and anthropogenic die-offs [@matkin_ongoing_2008] are important events which are hard to predict (i.e., highly stochastic) and can cause great population instability by drastically reducing the size and genetic diversity of a population [@reed_frequency_2003; @ameca_y_juarez_natural_2012].
-Managing the intensity of and frequency of extreme events [e.g., @aponte_forest_2016] can increase the stability of a habitat while promoting habitat diversity and adaptations to the event [@staver_tree_2011; @tingley_pyrodiversity_2016; @nimmo_animal_2019].
-
-@mueller_social_2013: Whooping Cranes migrate better when there are elders
-
-Other empirical refs to possibly add: @wolkovich_warming_2012, @nathan_big-data_2022
-
-Simulation refs: @cain_random_1985, @blackwell_heterogeneity_2007, @quaglietta_simulating_2019, @tucker_humaninduced_2021
-
-Patch connectivity and ease of movement may widen HRs by decreasing the energetic cost of movement and favoring exploration [@dickie_resource_2022], or they may shrink HRs by decreasing the energetic cost of movement while increasing encounter rates with resources [@martinez-garcia_how_2020; @bartumeus_influence_2008; @visser_plankton_2006].
-
-However, not all animals take advantage of linear features or higher predictability in human-altered habitats.
-@noonan_roads_2021 found that giant anteaters (*Myrmecophaga tridactyla*) in Brazil did not use roads to reduce movement costs or increase movement speed.
-Instead, roads increased anteater mortality because the animals were attracted to the high-quality foraging found on roadside habitat.
-Similarly, @medici_movement_2022 found that the movement of tapirs (*Tapirus terrestris*) was unaffected by anthropogenic activity or habitat type.
-
-<!--
-## Frontiers  Hierarchical, Memory-Based Movement Models for Translocated Elk (Cervus canadensis)  Ecology and Evolution
-
-Refs to check:
-
-* humans, nonhuman primates and other large-brained vertebrates make movement decisions based on spatial representations of their environments (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B50)
-* These representations may allow animals to move directly to important sites in their environment that lie outside of their perceptual range (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B40, https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B35)
-* Attribute memory, i.e., non-spatial memory, e.g., food type and abundance (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B13)
-* memory in capuchin monkeys (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B8)
-* memory in bison (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B18)
-* memory in sheep (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B18)
-* memory in woodland caribou (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B1)
-* memory in chimpanzees (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B20)
-* memory constitutes an important mechanism for home range emergence (https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/#B6, Börger et al., 2008; Van Moorter et al., 2009)
--->
 

### 2023-07-05 analysis/figures/barga-italy-2023/hr-mean-variance-simulations-barga-2023.R

@@ -1,6 +1,6 @@
 library('ggplot2') # for fancy plots
 library('cowplot') # for multi-panel fancy plots
-source('analysis/figures/mean-variance-trends-panel-data.R') # tibble of mu, var
+source('analysis/mean-variance-trends-panel-data.R') # tibble of mu, var
 
 # for common theme and color palette
 source('analysis/figures/barga-italy-2023/default-figure-styling-barga-2023.R')
@@ -51,48 +51,3 @@ p_full
 ggsave('figures/2023-GRS-GRC-barga-italy/mean-variance-5-by-5-hr-sims.png',
        p_full, width = 72, height = 37, units = 'cm', dpi = 600,
        bg = 'white')
-
-# estimate fraction of animals not satisfied if only mean is accounted for ----
-days_suff <-
-  left_join(days,
-            filter(days, variance == 'constant') %>%
-              select(mean, animal, hr_50, hr_95) %>%
-              rename(hr_50_x = hr_50, hr_95_x = hr_95),
-            by = c('mean', 'animal')) %>%
-  # see if the animal has enough space based on the 95% quantile
-  mutate(enough_50 = hr_50 <= hr_95_x,
-         enough_95 = hr_95 <= hr_95_x,
-         additional = if_else(enough_95, hr_95, hr_95_x))
-
-p_sim_suff <-
-  ggplot(days_suff) +
-  facet_grid(variance ~ mean, drop = FALSE) + # facet by trends in mu and var
-  # area between 50% and 95% HRs
-  geom_ribbon(aes(animal, ymin = hr_50_x, ymax = hr_95_x), fill = pal[3],
-              alpha = 0.2) +
-  # area missing to satisfy needs
-  geom_ribbon(aes(animal, ymin = hr_95, ymax = additional), fill = pal[7],
-              alpha = 0.2) +
-  # add lines for 50% and 95% HRs
-  geom_line(aes(animal, hr_50, color = enough_50, group = 1)) +
-  geom_line(aes(animal, hr_95, color = enough_95, group = 1)) +
-  scale_x_continuous('Time', breaks = NULL) +
-  scale_y_continuous(hr_lab, breaks = NULL, limits = c(0, NA)) +
-  scale_color_manual('Space is sufficient', values = pal[c(7, 3)],
-                     labels = c('No', 'Yes'), aesthetics = c('color', 'fill')) +
-  # remove facet names since they appear in the marginal panels
[…]

### 2023-07-05 analysis/figures/default-figure-styling.R

@@ -4,7 +4,7 @@ library('stringi') # for working with strings
 # set default ggplot theme
 # cannot bold text because unicode characters fail
 theme_set(theme_bw() + theme(legend.position = 'none',
-                             text = element_text(size = 22)))
+                             text = element_text(size = 15)))
 
 # custom color-blind palette
 pal <- c('#ff8c00', '#4477AA', '#009900', '#66CCEE',

### 2023-07-05 analysis/figures/delta-t-sensitivity.R

@@ -191,5 +191,5 @@ panels <-
                       labels = c('a.', 'b.', 'c.', 'd.'), ncol = 2),
             ncol = 1, rel_heights = c(1, 10))
 
-ggsave('figures/simulations/thinning-examples.png', plot = panels,
+ggsave('figures/thinning-examples.png', plot = panels,
        width = 7, height = 5, scale = 1.2, bg = 'white')

### 2023-07-05 analysis/figures/habitat-examples.R

@@ -1,0 +1,194 @@
+# plot(1, xlab = '\U1D53C') # run this before attaching ragg to plot unicode
+# if notation doesn't show, see https://github.com/r-lib/ragg/issues/118
+library('dplyr')     # for data wrangling
+library('tidyr')     # for data wrangling
+library('purrr')     # for functional programming
+library('ctmm')      # for movement modeling
+library('lubridate') # for working with data more smoothly
+library('ggplot2')   # for fancy plots
+library('latex2exp') # for LaTeX in plots
+library('cowplot')   # for fancy plots in grids
+library('ragg')      # needed for custom alpha with coord_cartesian
+# see https://github.com/tidyverse/ggplot2/issues/4029
+
+source('functions/rgamma2.R') # function to generate rgamma() from mean and variance
+source('functions/qgamma2.R') # function to generate qgamma() from mean and variance
+source('analysis/figures/default-figure-styling.R')
+theme_set(theme_get() + theme(legend.position = 'top', panel.grid = element_blank()))
+
+e_r <- '\U1D53C(\U1D445)'
+var_r <- '\U1D54D(\U1D445)'
+
+# resource abundance palette
+LOW <- '#744700'
+MID <- '#d9bb94'
+HIGH <- 'darkgreen'
+
+# variable definition ----
+# functions for trends in mean and variance
+mean_fcn <- function(.t) {
+  sinpi(.t / 4.5 - 0.75) * 20 + 35
+}
+
+var_fcn <- function(.t) {
+  (sinpi(.t / 4.5 - 0.75) * 15 + 17.5)
+}
+
+params <- tibble(t = seq(1, 10, length.out = 250),
+                 mu = mean_fcn(t),
+                 sigma2 = var_fcn(t),
+                 sigma2_constant = mean(sigma2))
+
+lim_food <- c(0,
+              qgamma2(p = 0.01, mu = max(params$mu), max(params$sigma2)) %>%
+                ceiling())
+
+lim_sigma2 <- c(floor(min(params$sigma2)), ceiling(max(params$sigma2)))
+
+# changing mean, constant variance ----
+# samples
+d_1 <- expand_grid(t = 1:10,
+                   patch = letters[t]) %>%
+  mutate(mu = mean_fcn(t),
+         sigma2 = mean(var_fcn(t)), # constant values
[…]

### 2023-07-05 analysis/figures/home-range-hypotheses.R

@@ -9,7 +9,7 @@ source('analysis/figures/default-figure-styling.R') # for common theme and color
 theme_set(theme_get() + theme(legend.position = 'none',
                               text = element_text(size = 22)))
 
-hr_lab <- 'Positional variance' # label for y axis
+hr_lab <- 'Spatial needs' # label for y axis
 
 # hypothesis for change in H over E(R) ----
 d <-

### 2023-07-05 analysis/figures/hr-mean-variance-simulations.R

@@ -1,0 +1,94 @@
+library('ggplot2') # for fancy plots
+library('cowplot') # for multi-panel fancy plots
+source('analysis/figures/default-figure-styling.R') # defaults for figures
+source('analysis/mean-variance-trends-panel-data.R') # tibble of mu, var
+
+# import simulation data
+days <- readRDS('simulations/days-hrs.rds')
+
+# main simulation figure of change in home range (5x5 plot) ----
+e_r <- 'Resource abundance, \U1D53C(\U1D445|t) = \U1D707(\U1D461)'
+v_r <- 'Resource stochasticity, \U1D54D(\U1D445|t) = \U1D70E\U00B2(\U1D461)'
+hr_lab <- 'Home range size, \U1D43B' # label for y axis
+p_sim <-
+  ggplot(days) +
+  facet_grid(variance ~ mean, drop = FALSE) + # facet by trends in mu and var
+  # area between 50% and 95% HRs
+  # geom_ribbon(aes(animal, ymin = hr_50, ymax = hr_95), fill = pal[3],
+  #             alpha = 0.2) +
+  # add lines for 50% and 95% HRs
+  # geom_line(aes(animal, hr_50), color = pal[3]) +
+  geom_line(aes(animal, hr_95), color = pal[3], linewidth = 1) +
+  scale_x_continuous('Time', breaks = NULL) +
+  scale_y_continuous(hr_lab, breaks = NULL, limits = c(0, NA)) +
+  # remove facet names since they appear in the marginal panels
+  theme(strip.background = element_blank(), strip.text = element_blank())
+
+# trends in the mean
+p_mean <-
+  ggplot(d55, aes(animal, mu)) +
+  facet_grid(. ~ mean) +
+  geom_line(color = pal[1], linewidth = 1) +
+  scale_x_continuous(e_r, breaks = NULL, position = 'top') +
+  scale_y_continuous(hr_lab, breaks = NULL) +
+  theme(strip.background = element_blank(),
+        axis.title.y = element_text(colour = 'transparent'))
+
+# trends in the variance
+p_variance <-
+  ggplot(d55, aes(animal, sigma2)) +
+  facet_grid(variance ~ ., switch = 'y') +
+  geom_line(color = pal[2], linewidth = 1) +
+  scale_x_continuous('Time', breaks = NULL) +
+  scale_y_continuous(v_r, breaks = NULL) +
+  theme(strip.background = element_blank(),
+        axis.title.x = element_text(color = 'transparent'))
+
[…]

### 2023-07-05 analysis/figures/mean-variance-trends-panel-data.R

@@ -1,21 +1,0 @@
-#' created a separate script for `d55` since it's used in multiple scripts
-
-library('dplyr') # for data wrangling
-library('tidyr') # for data wrangling (expand_grid())
-library('purrr') # for functional programming (e.g., map_*())
-source('functions/mean-variance-trends.R')  # functions to generate means and variances
-
-N <- 200 # number of animals
-types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic') # functions types
-
-d55 <-
-  expand_grid(mean = types,
-              variance = types) %>%
-  # assign trends in means and variances depending on the type
-  mutate(t = map2(mean, variance,
-                  \(.x, .y) tibble(animal = 1:N,
-                                   mu = get(paste0('mean_', .x))(animal),
-                                   sigma2 = get(paste0('variance_', .y))(animal)))) %>%
-  unnest(t) %>%
-  mutate(mean = factor(mean, levels = types),
-         variance = factor(variance, levels = types))

### 2023-07-05 analysis/figures/modeling-R-and-hr.R

@@ -1,0 +1,136 @@
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('mgcv')    # for empirical Bayes GAMs
+library('ggplot2') # for fancy plots
+library('cowplot') # for fancy multi-panel plots
+source('analysis/figures/default-figure-styling.R') # for color palette
+theme_set(theme_bw() + theme(legend.position = 'none'))
+
+e_r <- 'Resource abundance, \U1D53C(\U1D445)'
+v_r <- 'Resource stochasticity, \U1D54D(\U1D445)'
+
+sims <- readRDS('simulations/days-hrs.rds') %>%
+  tibble() %>%
+  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr') %>%
+  mutate(quantile = factor(quantile))
+
+# sampling is not uniform, but not highly problematic
+# removing the data with constant mu or sigma2 has no effect on the model
+ggplot(sims, aes(mu, sigma2)) +
+  geom_point(alpha = 0.3)
+
+m_1 <- gam(
+  list(
+    # predictor for the mean
+    hr ~ quantile + mu + sigma2,
+    # predictor for the scale (variance = scale * mean)
+    ~ quantile + mu + sigma2),
+  family = gammals(),
+  data = sims)
+
+m_2 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2),
+    ~ quantile + s(mu) + s(sigma2)),
+  family = gammals(),
+  data = sims)
+
+m_3 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5),
+    ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5)),
+  family = gammals(),
+  data = sims)
+
+AIC(m_1, m_2, m_3) # smooth model with interactions is best
+BIC(m_1, m_2, m_3) # smooth model with interactions is best
+
+m <- m_3
+
+plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+
+# estimate predictions ----
+# effect of mu
+newd_mu <- expand_grid(mu = seq(min(sims$mu), max(sims$mu),
+                                length.out = 400),
+                       sigma2 = mean(sims$sigma2),
+                       quantile = unique(sims$quantile))
+
+preds_mu <-
+  bind_cols(
+    newd_mu,
[…]

### 2023-07-05 analysis/hr-simulation-extreme-scenarios.R

@@ -3,7 +3,7 @@ library('dplyr')   # for data wrangling (e.g., %>%)
 library('tidyr')   # for data wrangling (e.g., nested tibbles)
 library('ggplot2') # for fancy plots
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/mean-variance-trends-panel-data.R') # means & variances
 source('analysis/simulations/movement-model.R') # for consistency across scripts
 source('functions/get_hr.R') # for extracting gaussian home range
 theme_set(theme_bw())
@@ -164,5 +164,5 @@ p_hr_days <-
                      limits = c(2, 1100)) +
   scale_y_log10(expression(Estimated~home~range~(log[10]~scale))); p_hr_days
 
-ggsave('figures/simulations/hr-over-days.png', plot = p_hr_days,
+ggsave('figures/hr-over-days.png', plot = p_hr_days,
        width = 6, height = 3, dpi = 'print')

### 2023-07-05 analysis/mean-variance-trends-panel-data.R

@@ -1,0 +1,21 @@
+#' created a separate script for `d55` since it's used in multiple scripts
+
+library('dplyr') # for data wrangling
+library('tidyr') # for data wrangling (expand_grid())
+library('purrr') # for functional programming (e.g., map_*())
+source('functions/mean-variance-trends.R')  # functions to generate means and variances
+
+N <- 200 # number of animals
+types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic') # functions types
+
+d55 <-
+  expand_grid(mean = types,
+              variance = types) %>%
+  # assign trends in means and variances depending on the type
+  mutate(t = map2(mean, variance,
+                  \(.x, .y) tibble(animal = 1:N,
+                                   mu = get(paste0('mean_', .x))(animal),
+                                   sigma2 = get(paste0('variance_', .y))(animal)))) %>%
+  unnest(t) %>%
+  mutate(mean = factor(mean, levels = types),
+         variance = factor(variance, levels = types))

### 2023-07-05 analysis/tapirus-terrestris/anna-betals.R

@@ -1,0 +1,51 @@
+library('ctmm')      # for tracking data and movement modeling
+library('dplyr')     # for data wrangling
+library('ggplot2')   # for fancy plots
+library('mgcv')      # for GAMs
+library('lubridate') # for smoother date wrangling
+
+source('../ndvi-stochasticity/functions/betals.r') # beta location-scale fam 
+source('analysis/figures/default-figure-styling.R') # for NDVI color palette
+
+# import 95% home range estimates with no CIs (optional)
+ud <- readRDS('../tapirs/models/tapirs-akdes.rds') %>%
+  filter(name.short == 'ANNA') %>%
+  pull(akde) %>%
+  purrr::map_dfr(\(x) {
+    SpatialPolygonsDataFrame.UD(x, level.UD = 0.95, level = 0) %>%
+      sf::st_as_sf() %>%
+      sf::st_transform(crs = '+proj=longlat')
+    })
+
+# import NDVI data
+d <- readRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
+  mutate(dec_date = decimal_date(date)) %>%
+  # remove unusual raster (likely due to a flood)
+  filter(date != '2017-12-19')
+
+ggplot(d, aes(long, lat, fill = ndvi)) +
+  facet_wrap(~ round(dec_date, 2)) +
+  geom_raster() +
+  geom_sf(dat = ud, inherit.aes = FALSE, alpha = 0.3) +
+  scale_fill_gradientn('NDVI', colours = ndvi_pal, limits = c(-1, 1))
+
+m_ndvi <-
+  gam(list(
+    # mean predictor
+    ndvi ~ # not scaling because range is in (0, 1)
+      s(long, lat, bs = 'ds', k = 20) + # mean over space
+      s(dec_date, bs = 'tp', k = 20), # need high k to account for animal adapting
+    # precision (1/standard deviation) predictor
+    ~
+      s(long, lat, bs = 'ds', k = 10) +
+      s(dec_date, bs = 'tp', k = 10)),
+    family = betals(),
+    data = d,
+    method = 'REML',
+    control = gam.control(nthreads = 8, trace = TRUE))
+
+saveRDS(m_ndvi, file = 'models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
+
+if(FALSE) {
+  plot(m_ndvi, pages = 1, scheme = 3) # plot smooths
+}

### 2023-07-05 analysis/tapirus-terrestris/anna-ndvi-data-to-local.R

@@ -1,0 +1,76 @@
+# see https://rspatialdata.github.io/vegetation.html for example code and more info
+#' install `rgeoboundaries` with `remotes::install_github('wmgeolab/rgeoboundaries')`
+#' install `MODIStsp` with `remotes::install_github('ropensci/MODIStsp')`
+#' `MODIStsp` version >= 2.0.7 from github fixes login issue due to 
+library('dplyr')    # for data wrangling
+library('sf')       # for spatial features
+library('MODIStsp') # for downloading NDVI rasters
+library('purrr')  # for functional programming
+library('tidyr')  # for data wrangling
+library('raster') # to import and save rasters
+source('earthdata-login-info.R') # import personal login credentials for EarthData
+
+# use a bounding box from tracking data ----
+library('ctmm') # for movement data and models
+tapir <- filter(readRDS('../tapirs/models/tapirs-final.rds'), name.short == 'ANNA')
+tel <- tapir$data[[1]]
+ud <- tapir$akde[[1]]
+
+bbox <-
+  SpatialPolygonsDataFrame.UD(ud, level.UD = 0.9995, level = 0) %>%
+  st_as_sf() %>%
+  st_transform(crs = '+proj=longlat') %>%
+  st_bbox()
+
+# download NDVI
+MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
+         out_folder = 'data/ndvi-rasters/tapir-anna', # '<folder>/VI_16Days_250m_v6/NDVI'
+         selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)', # can't specify Terra here
+         prod_version = '061', # 2022 raster version
+         bandsel = 'NDVI', # Normalized Difference Vegetation Index layer only
+         sensor = 'Terra', # only terrestrial values, ignore main bodies of water
+         user = USERNAME, # your Earthdata username (for urs.earthdata.nasa.gov/home)
+         password = PASSWORD, # your Earthdata password
+         start_date = format(min(tel$timestamp) - 16, '%Y.%m.%d'), # 1 raster before min
+         end_date = format(max(tel$timestamp) + 16, '%Y.%m.%d'), # 1 raster after end
+         spatmeth = 'bbox', # use a bounding box for the extent
+         bbox = bbox, # spatial file for raster extent
+         out_projsel = 'User Defined', # use specified projection instead of default
[…]

### 2023-07-05 analysis/tapirus-terrestris/moving-window-figure-anna-betals.R

@@ -1,0 +1,184 @@
+library('dplyr')     # for data wrangling
+library('purrr')     # for functional programming
+library('tidyr')     # for data wrangling
+library('ctmm')      # for movement modeling
+library('mgcv')      # for empirical Bayesian modeling
+library('scam')      # for shape-constrained splines
+library('lubridate') # for smoother date wrangling
+library('ggplot2')   # for fancy figures
+library('cowplot')   # for fancy multi-panel plots
+
+source('analysis/figures/default-figure-styling.R') # for theme & palettes
+
+# axis lables
+e_r <- 'Resource abundance, \U1D707(t)'
+v_r <- 'Resource stochasticity, \U1D70E\U00B2(t)'
+hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
+
+# import location-scale model for predictions of mean and variance in NDVI
+m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
+
+# function to make predictions from the model
+ndvi_preds <- function(.data) {
+  .data <- .data %>%
+    data.frame() %>% # convert telemetry data to a data.frame
+    mutate(year = year(timestamp),
+           doy = yday(timestamp))
+  
+  predict.gam(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE) %>%
+    data.frame() %>%
+    tibble() %>%
+    transmute(mu = X1,
+              sigma2 = X1 * (1 - X1) * X2) %>% # didn't scale NDVI to [0, 1]
+    return()
+}
+
+# import tapir data (https://doi.org/10.1186/s40462-022-00313-w)
+tapir <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+  filter(name.short == 'ANNA')
+tel <- tapir$data[[1]]
+
+tel <- data.frame(tel) %>%
+  rename(long = longitude, lat = latitude) %>%
+  mutate(dec_date = decimal_date(timestamp))
+tel <- bind_cols(tel, ndvi_preds(tel))
+
+tapir <-
+  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+  # mutate(ess = map_dbl(model, \(.m) summary(.m)$DOF['area']),
+  #        weight = ess / mean(ess)) %>%
[…]

### 2023-07-05 analysis/tapirus-terrestris/tapirs-moving-window.R

@@ -1,0 +1,86 @@
+library('dplyr')   # for data wrangling
+library('purrr')   # for functional programming
+library('tidyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+library('cowplot') # for fancy multi-panel plots
+library('ctmm')    # for movement modeling
+source('functions/window_hr.R') # function to calculate HRs and create figures
+theme_set(theme_bw()) # change default theme
+
+# import tapir data from https://github.com/StefanoMezzini/tapirs
+tapirs <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+  select(region, name, name.short, data) %>%
+  filter(region != 'atlantica') # Atlantic Forest uses VHF => irregular sampling
+
+# check ranges of tracking data
+tapirs <-
+  tapirs %>%
+  mutate(start = map_chr(data, \(d) min(d$timestamp) %>% as.character()),
+         end = map_chr(data, \(d) max(d$timestamp) %>% as.character()),
+         days = map_dbl(data, \(d) diff(range(d$timestamp))),
+         clicks = map_dbl(data, nrow),
+         daily_clicks = clicks / days)%>%
+  arrange(days, clicks, desc(daily_clicks))
+
+if(FALSE) {
+  View(tapirs)
+  
+  # plot the tapir data
+  tapirs %>%
+    filter(region == 'cerrado') %>%
+    mutate(data = map(data, data.frame)) %>% # convert to df for unnesting
+    unnest(data) %>%
+    slice(seq(1, n(), by = 10)) %>%
+    ggplot() +
+    facet_wrap(~ name, scales = 'free') +
+    geom_path(aes(longitude, latitude), alpha = 0.05) +
+    geom_point(aes(longitude, latitude, color = timestamp), alpha = 0.1) +
+    scale_color_viridis_c('Date') +
+    theme(legend.position = c(x = 0.7, y = 0.1),
+          legend.direction = 'horizontal',
+          legend.key.width = unit(0.5, 'in'))
+}
+
+# find some good examples ----
+# Cerrado ----
+ce_proj <- '+proj=utm +zone=22 +datum=NAD83 +units=m'
+
+cerrado <- filter(tapirs, region == 'cerrado')
+select(cerrado, name, days, clicks, daily_clicks)
+
+cerrado %>%
+  filter(name.short %in% # filter to only a few good examples
+           c('ANNA', 'DONALINA', 'KURUKA', 'LOU', 'ZEFA', 'ZACA', 'TITI', 'ZIGGY')) %>%
[…]

### 2023-07-06 writing/manuscript.Rmd
updated manuscript with comments from 2023-07-05

@@
 
-\noindent Animals' spatial needs are thought to be tightly linked to the amount of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
+\noindent The amount of space animals use is thought to be tightly linked to the availability of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
 This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown.
 This is not a trivial oversight.
-Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs.
+Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs, challenging ecological theory and applied conservation alike.
 In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
@@
 \noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** an animal can access.
-However, a species' spatial requirements depend strongly on its metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @boratynski_energetic_2020; @noonan_effects_2020].
-
-While many researchers have demonstrated that animals adapt heir home ranges in response to resources abundance, few studies build upon others' previous work.
+Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
+While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
 Instead, results are reported as independent, *de-novo* findings.
@@
 In this paper, we refer to a location's average amount of resources as **resource abundance**, while use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
-We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality.
+We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use.
 For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
-In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food, while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies being selected for.
+In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
+Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies and life histories being selected for.
 Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-Here, we illustrate how animals' spatial needs depend on the abundance and unpredictability of resources.
-First, we set the theoretical background necessary for the successive sections by introducing concepts and notations without assuming any previous knowledge of statistics or movement ecology.
+Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources.
+First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
 Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis.
 Afterwards, we provide a review of the effects of resource *stochasticity* on animals spatial needs while suggesting a second unifying hypothesis.
-Subsequently, we test the two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
+Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
 Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
@@
 Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate, weather, competitive pressure, and individual- and species-specific energetics.
+Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differents in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
 Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
+Quantifying resources as a numerical random variable (as opposed to using *ad hoc* qualitative descriptions) provides us with the capacity to leverage techniques from probability theory and statistics.
 
@@
 There are many distributions we can assign to $R$, depending on how we quantify it.
-If $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution.
+For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution.
 If $R$ is a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
@@
 We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
-
-The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive (which is necessarily greater than zero).
-As $\e(R)$ increases, we expect animals' spatial requirements to decrease nonlinearly until they reach the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-<!--The spatial requirements initially decrease rapidly since small additive increases (e.g., from 1 to 2) are proportionally large (i.e., 2 is twice 1).
-In contrast, at higher values of $\e(R)$, the same additive changes (e.g.
-from 50 to 51) are proportionally small (i.e., 51 is only 2% larger than 50), so spatial needs decrease more slowly.-->
-
+s
 ### Variance in resource abundance, $\var(R)$
@@
 We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$ and $\widehat{\sigma^2}(t)$ to indicate its estimate over time.
-Consequently, while one may consider $R$ to vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
+Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
 Additionally, we distinguish between two different sources of variation: We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
@@
 This assumption also holds biologically: In the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
+
+# Effects of resource abundance, $\e(R)$
+
+```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
+Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
+```
+
+```{r, eval=FALSE, echo=FALSE}
+# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
+dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
+    # mean = k * theta
+    # sigma^2 = k * theta^2
+    dgamma(x = x,
+           shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
+           scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
+}
+x <- seq(0, 10, length.out = 250)
+layout(matrix(1:4, ncol = 2, byrow = TRUE))
+plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
+plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
+     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
+layout(1)
+```
+
+\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
+All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
+Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
+Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
+The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
+Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+
+When $\e(R)$ is high, we expect animals' spatial requirements be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
+Sill, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
+For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
+
+It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
+Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
+Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).
+
+```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
+(a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
+```
+
+Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
+However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
+peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
+
+This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
+Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
+Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
+When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
+Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
+
+# Effects of resource stochasticity, $\var(R)$
+
+\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
+For example, the abundance of berries is predictably scarce outside the growing season, while it is higher and unpredictable during spring and summer, since a bear may explore a location only after competitors ate all the berries or during a year of unusually low productivity.
+In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
+
+```{r s2-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource stochasticity ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low spatial needs as animals are able to depend on somewhat predictable resources.
+As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
+Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
+```
+
+Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
+While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
+Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+
+```{r, eval=FALSE, echo=FALSE}
+# the median decreases as V(R) increases, even if E(R) is constant
+source('functions/qgamma2.R')
+plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
+```
+
+Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
+If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
+
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
+(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
+```
+
+Like the scenario illustrated in figure \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
+We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)).
+When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
+Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
+Biologically, this is important because some animals may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
+An animal's perception of changes in $R$ will depend strongly on the animal's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
+The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
+Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
+
+```{r stoch-timescale, eval = FALSE, fig.cap="Conceptualizations of perceptions of $\\e(R)$ (a.) and $\\var(R)$ (b.) over time for animals with no plasticity in their movement (blue), seasonal memory but no plasticity (orange), or high plasticity but no memory (green).
+The true trends in $\\e(R)$ and $\\var(R)$ are indicated in black in panels (a.) and (b.), respictively.
+Estimates in the white area indicate the animal's estimated $\\mu(t)$ and $\\sigma^2(t)$ based on past experience, while estimates to the grey shaded region indicate the animals' predictions of $\\mu(t)$ and $\\sigma^2(t)$ in the future.
+The four facets show how each behavior performs in environments where $\\e(R)$ is either predictable (top row) or stochastic (bottom row), and $\\var(R)$ is either predictable (left column) or stochastic (right column).
+Animals that do not adapt to changes in $R$ (blue) are unable to predict $\\e(R)$ accurately, and they have high estimates of $\\var(R)$ when $\\mu(t)$ is far from the long-term, temporally static average.
+Memory of seasonal trends in $\\mu(t)$ and $\\var(R)$ (orange) allows animals to perform well in predictable environments, but their lack of adaptability causes them to fair poorly in stochastic habitats, where they perceive $\\var(R)$ to be high when $\\e(R)$ is far from the expected $\\mu(t)$.
+Finally, the ability to adapt rapidly to change (green) allows animals to track changes in $\\e(R)$ and $\\var(R)$ closely, but the lack of memory prevents accurate predictions in predictable environments.
+Since such animals tend to perceive changes in $R$ as changes in $\\mu(t)$, they tend to have lower estimates of $\\sigma^2(t)$ than other animals."}
+# set eval to false: include in a separate paper?
+knitr::include_graphics('../figures/stochasticity-timescale.png')
+```
+
+# Interactive effects of $\e(R)$ and $\var(R)$
+
+\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' spatial needs, but we presented the two parameters as independent drivers of movement.
+However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, animals may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
+Consequently, models estimating the effects of $R$ on animal's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
+Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
+
+<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
+However, the effects of $\var(R)$ on animals' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
+Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
+-->
+
+@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\mu(t)}{\sqrt{\sigma^2(t)}}$) in C:N.
+However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
+Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality).
+Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
+A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
+Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on animals' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
+Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
+The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@ipcc_global_2018; @noonan_situ_2018; @yao_emergence_2022].
+
+@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa): Animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
+Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
+This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but the they may not respond if $\mu(t)$ is sufficiently high.
+However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
+Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
+Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.
+
+## Simulating animal responses to $\e(R)$ and $\var(R)$
+
+To illustrate how $\e(R)$ and $\var(R)$ affect animals' spatial needs as well as the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations in figure \@ref(fig:5-5-sims) (for methodological details, see Appendix 1).
+The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the animal.
+As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
+
+```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
+Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
+The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
+$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
+Additional information is provided in Appendix 1.", out.width='100%'}
+knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+```
+
+In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
+However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
+Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
+This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
+In cases where $\mu(t)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
+
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
+Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
+This is because, similarly to resource-poor periods, times of greater stochasticity require the animal to move over larger areas for longer periods of time.
+This results in a greater uncertainty in how much time and space the animal will require to reach satiety, or indeed whether an animal living in highly stochastic environments can even reach satiety within a finite amount of time.
+
+The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
+Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
+When both $\e(R)$ and $\var(R)$ increase linearly, $H$ initially increases since the effect of $\var(R)$ is stronger, but then decreases as the effect of $\e(R)$ begins to dominate.
+However, when $\mu(t)$ and $\sigma^2(t)$ follow fairly different trends (e.g., cyclical $\e(R)$ and linear $\var(R)$), it is easy to see how an increase in $\var(R)$ amplifies the effects of $\e(R)$.
+In contrast, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
+Thus, $H$ is largest when $\e(R)$ is low and $\var(R)$ is high, and it is smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
+
+Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
+$\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
+Although it is impossible for $\mu(t)$ and $\sigma^2(t)$ to increase linearly forever, these examples are useful to demonstrate that linear changes in $\mu(t)$ and $\sigma^2(t)$ affect $H$ nonlinearly.
+Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
+Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
+However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
+
+Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either E(R) or V(R), rather than time.
+Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
+We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
+Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
+Secondly, the animal's movement was simulated to be range resident within an homogeneous environment.
+Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood.
+Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
+Additionally, the simulations also ignore the how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
+While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.
+
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
+The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions.
+The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
+Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the approximately nonlinear increase in $H$ as $\\var(R)$ increases.
+Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
+knitr::include_graphics('../figures/simulation-regression-plots.png')
+```
+
+# Applying the framework: The effects of E(R) and V(R) on a lowland tapir
+
+\noindent In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a Beta location-scale GAM, after scaling NDVI to [0, 1].
+Appendix 2 contains additional information on how NDVI was modeled for the empirical example, along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
+
+The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functional responses.
+We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
+In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings.
+Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models with individual-level intercepts and slopes [see models "GS" and "GI" in @pedersen_hierarchical_2019].
+Such models would allow one to test whether members of the same population or species behave according to common functional responses to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
+In the following sections, we illustrate how one can apply this framework to empirical data.
+
+We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
+Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.
+
+### Modeling $R$
+
+\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
+its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
+`mgcv` [@wood_generalized_2017] is a commonly used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
+Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie families.
+
+The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
+However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
+available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and NDVI [@pettorelli_normalized_2011].
+
+The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
+The Tweedie location-scale family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density.
+In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
+If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
+
+Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
+Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the location terms.
+An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
+Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
+Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
+We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
+Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size due to the spatiotemporal autocorrelation).
 
@@
 
-# Effects of resource abundance, $\e(R)$
-
-```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
-Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
-```
-
-```{r, eval=FALSE, echo=FALSE}
-# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
-dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
-    # mean = k * theta
-    # sigma^2 = k * theta^2
-    dgamma(x = x,
-           shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
-           scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
-}
-x <- seq(0, 10, length.out = 250)
-layout(matrix(1:4, ncol = 2, byrow = TRUE))
-plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
-layout(1)
-```
-
-\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
-All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
-
-An animal's spatial needs are often hypothesized to be depend on resource abundance [@burt_territoriality_1943], such that spatial needs are high when resources are scarce and low when resources are abundant, but, as noted above, the relationship is necessarily nonlinear, as illustrated in figure \@ref(fig:mu-hyp) [also see:  @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
-Low values of $\e(R)$ force animals to have larger home ranges [ @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], although large-scale changes in behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
-
-Animals living in large groups will often require larger home ranges because they will have to share resources with conspecifics [@prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-While competition may push animals to explore other areas and expand their home range [@jetz_scaling_2004], strong and consistent competition paired with territorial defense may also prevent them from doing so [wolves: @rich_anthropogenic_2012; feral cats: @bengsen_feral_2016; Capuchin monkeys: @torrez-herrera_monkeys_2020].
-It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose: @singh_migration_2012; eagles: @wheat_migrate_2017; @poessel_interpreting_2022; flamingos: @pretorius_movement_2020].
-
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
-(a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
-```
-
-Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
-Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
-However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
-
-This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
-Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
-Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
-Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
-The need to account for $\var(R)$ is compounded by recent changes in climate, which exposes species to novel situations and increasingly common stochastic events [@ipcc_global_2018; @noonan_situ_2018; @yao_emergence_2022].
-
-# Effects of resource stochasticity, $\var(R)$
-
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as infrequent and unpredictable events.
-For example, berries are predictably scarce out of the growing season, but their abundance more stochastic when in spring and summer since the timing between plants can depend on many unpredictable factors, and an animal may explore a location only after competitors ate all the berries.
-Thus, while a bear cannot to expect to find berries during growing season, it can expect the abundance in berries to be more stochastic.
-In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@noonan_situ_2018; @berger_climate_2018; @grant_evolution_2017; @yao_emergence_2022; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
-
-```{r s2-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource stochasticity ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low spatial needs as animals are able to depend on somewhat predictable resources.
-As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
-```
-
-Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
-While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
-
-```{r, eval=FALSE, echo=FALSE}
-# the median decreases as V(R) increases, even if E(R) is constant
-source('functions/qgamma2.R')
-plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
-```
-
-Similarly to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
-
-Like the scenario illustrated in figure \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
-We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)).
-When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
-Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
-Biologically, this is important because some animals may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
-An animal's perception of changes in $R$ will depend strongly on the animal's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
-
-```{r stoch-timescale, eval = FALSE, fig.cap="Conceptualizations of perceptions of $\\e(R)$ (a.) and $\\var(R)$ (b.) over time for animals with no plasticity in their movement (blue), seasonal memory but no plasticity (orange), or high plasticity but no memory (green).
-The true trends in $\\e(R)$ and $\\var(R)$ are indicated in black in panels (a.) and (b.), respictively.
-Estimates in the white area indicate the animal's estimated $\\mu(t)$ and $\\sigma^2(t)$ based on past experience, while estimates to the grey shaded region indicate the animals' predictions of $\\mu(t)$ and $\\sigma^2(t)$ in the future.
-The four facets show how each behavior performs in environments where $\\e(R)$ is either predictable (top row) or stochastic (bottom row), and $\\var(R)$ is either predictable (left column) or stochastic (right column).
-Animals that do not adapt to changes in $R$ (blue) are unable to predict $\\e(R)$ accurately, and they have high estimates of $\\var(R)$ when $\\mu(t)$ is far from the long-term, temporally static average.
-Memory of seasonal trends in $\\mu(t)$ and $\\var(R)$ (orange) allows animals to perform well in predictable environments, but their lack of adaptability causes them to fair poorly in stochastic habitats, where they perceive $\\var(R)$ to be high when $\\e(R)$ is far from the expected $\\mu(t)$.
-Finally, the ability to adapt rapidly to change (green) allows animals to track changes in $\\e(R)$ and $\\var(R)$ closely, but the lack of memory prevents accurate predictions in predictable environments.
-Since such animals tend to perceive changes in $R$ as changes in $\\mu(t)$, they tend to have lower estimates of $\\sigma^2(t)$ than other animals."}
-# set eval to false: include in a separate paper?
-knitr::include_graphics('../figures/stochasticity-timescale.png')
-```
-
-# Interactive effects of $\e(R)$ and $\var(R)$
-
-\noindent In systems where both $\e(R)$ and $\var(R)$ change substantially over time, animals may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on animal's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
-A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
-Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
-
-<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
-However, the effects of $\var(R)$ on animals' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
--->
-
-@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\mu(t)}{\sqrt{\sigma^2(t)}}$) in C:N.
-However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality).
-Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
-A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons<!-- (\@ref(fig:stoch-timescale))-->.
-Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
-
-@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves (*Canis lupus*) and fishers (*Martes pennanti*) depends on seasonality (and vice-versa): Animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
-Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but the they may not respond if $\mu(t)$ is sufficiently high.
-However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
-
-## Simulating animal responses to $\e(R)$ and $\var(R)$
-
-To illustrate how $\e(R)$ and $\var(R)$ affect animals' spatial needs as well as the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations in figure \@ref(fig:5-5-sims) (for methodological details, see Appendix 1).
-The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the animal.
-As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
-
-```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
-Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
-$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
-Additional information is provided in Appendix 1.", out.width='100%'}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
-```
-
-In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
-Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
-In cases where $\mu(t)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
-
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
-Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
-This is because, similarly to resource-poor periods, times of greater stochasticity require the animal to move over larger areas for longer periods of time.
-This results in a greater uncertainty in how much time and space the animal will require to reach satiety, or indeed whether an animal living in highly stochastic environments can even reach satiety within a finite amount of time.
-
-The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
-Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
-When both $\e(R)$ and $\var(R)$ increase linearly, $H$ initially increases since the effect of $\var(R)$ is stronger, but then decreases as the effect of $\e(R)$ begins to dominate.
-However, when $\mu(t)$ and $\sigma^2(t)$ follow fairly different trends (e.g., cyclical $\e(R)$ and linear $\var(R)$), it is easy to see how an increase in $\var(R)$ amplifies the effects of $\e(R)$.
-In contrast, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
-Thus, $H$ is largest when $\e(R)$ is low and $\var(R)$ is high, and it is smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
-
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
-$\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
-Although it is impossible for $\mu(t)$ and $\sigma^2(t)$ to increase linearly forever, these examples are useful to demonstrate that linear changes in $\mu(t)$ and $\sigma^2(t)$ affect $H$ nonlinearly.
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
-Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
-However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
-
-Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either E(R) or V(R), rather than time.
-Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
-We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.-->
-
-We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
-Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
-Secondly, the simulated animal's range residency with a consistent (albeit stochastic) movement process within a homogeneous environment.
-Since real-world landscapes are heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event such as a fire or a flood.
-Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
-Additionally, the simulations also ignore the how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
-While these are definitely important effects that should be accounted for [@jetz_scaling_2004], we believed that doing so was beyond the scope of this paper.
-
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions.
-The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
-Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the approximately nonlinear increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-plots.png')
-```
-
-## Applying this framework
-
-\noindent The hypotheses we present here allow researchers to combine findings and generalize results to a single, common set of functional responses.
-We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all animals, once differences in metabolic needs and diet are accounted for.
-In this subsection, we illustrate how researchers can test these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings.
-Additionally, the analyses can be expanded to test whether members of the same population or species behave according to common functional responses to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
-In the following sections, we illustrate how one can apply this framework to empirical data.
-
-We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
-Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.
-
-### Modeling $R$
-
-\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
-its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and generalized additive models [GAMs, see @wood_generalized_2017], including location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), scaled t, gamma, Tweedie, and Gumbel.
-
-The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
-However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), NDVI [@pettorelli_normalized_2011], and extremes (such as daily maximum or minimum temperatures).
-
-The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
-The Tweedie location-scale family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density.
-In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family.
-This family is not available in `mgcv` at the time of publication, but it should be available soon.
-If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
-
-Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the location terms.
-An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
-Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
-Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
-We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size).
-Appendix 2 contains additional information on how NDVI was modeled for the empirical example, along with the continuous time models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
-
 <!-- ### Modeling spatial needs -->
@@
 
-\noindent As discussed in the precious section, the first step to modeling $H$ is to choose an appropriate distribution.
+\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate distribution.
 Since $H$ is strictly positive and continuous, a Gamma distribution seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017].
@@
 
-### An empirical example: Lowland tapir in the Brazilian Cerrado
-
-\noindent In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
-We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a Beta location-scale GAM, after scaling NDVI to [0, 1] using equation \@ref(eq:ndvi-to-01).
-Additional information regarding the modeling can be found in Appendix 2.
+### Empirical results
 
@@
 
-Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
+\noindent Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
 Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period.
-Panel **c** shows the changes in the tapir's spatial use over time.
+Panel **c** shows the changes in the tapir's space use over time.
 Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018).
@@
 Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
+Thus, while the time series shown in panel (c.) may appear complicated, the tapir's response can be well described by two simple functional responses.
 
@@
 
-\noindent This paper presents a new unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
+\noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
 The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
 Instead, one can test whether individuals follow these hypotheses, if members of the same population or species behave similarly, and if all members of the group have a common functional response [using hierarchical models, see @pedersen_hierarchical_2019].
-Individuals and populations that were thought to have different behaviors and spatial needs due to differences in behavior may instead be simply responding to different conditions along the same functional response [e.g., @singh_migration_2012].
+Individuals and populations that were thought to have different behaviors and spatial needs due to differences in behavior may instead be simply responding to different local conditions along the same functional response [e.g., @singh_migration_2012].
 
@@
 Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-However, adapting to changes in environmental cues rather than on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
+However, adapting to changes in environmental cues rather than relying on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
 
@@
 Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences in $\mu(t)$ between years [@jonzen_rapid_2006].
-However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources [cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
-However, when anomalies occur repeatedly over time, animals may begin to shift their ranges [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
-Range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021].
-However, they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [@wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range since it is determined by their food sources' and associates' ability to move and adapt, too.
+However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources during unpredictable times [whooping cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
+Instead, if anomalies occur repeatedly over time, animals may adapt to a higher $\sigma^2(t)$ or shift their ranges altogether [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
+Still, while range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021], they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [e.g., @wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range, since it is determined by their food sources' and associates' ability to move and adapt, too.
 
@@
 In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants.
-Consequently, the abundance of primary producers may correlate with habitat productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019].
+Consequently, the abundance of primary consumers may correlate with productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019].
 Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay.
 Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
+However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less sotchastic resources.
 

### 2023-07-25 writing/manuscript.Rmd
updated based on Sarah's comments

@@
 Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by reducing body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020], but movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012], or moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
+However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
 
@@
 Since the variable is random, it can take multiple possible values, each with different probabilities.
-The set or range of values which have non-zero probabilities in a distribution are referred to as the distribution's **support**.
+The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.
 
@@
 For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution.
-If $R$ is a discrete variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
+If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
 Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
+However, since there is no commonly used distribution with that support, the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).
 
@@
 Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$.
-When the mean changes over time, such as in strogly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\mu(t)$.
+When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$.
 We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
-s
+
 ### Variance in resource abundance, $\var(R)$
@@
 A random variable's **variance** is a measure of its unpredictability (i.e., variability).
-We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$ and $\widehat{\sigma^2}(t)$ to indicate its estimate over time.
+We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$).
 Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
-Additionally, we distinguish between two different sources of variation: We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
+Additionally, we distinguish between two different sources of variation.
+We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
 In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$.
@@
 \noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
-Although Gamma distributions are more often parameterized using parameters shape and scale ($k, \theta$) or shape and rate ($\alpha, \beta = 1/\theta$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
-However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa): As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
-This assumption also holds biologically: In the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
+Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
+However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa).
+As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
+This assumption also holds biologically; in the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
 
@@
 It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Sill, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
+Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
 For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
@@
 (a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
+The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
@@
 Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
 Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
@@
 Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
@@
 (a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
+The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
@@
 
-@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\mu(t)}{\sqrt{\sigma^2(t)}}$) in C:N.
+@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
 However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
@@
 Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
-The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@ipcc_global_2018; @noonan_situ_2018; @yao_emergence_2022].
+The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
 
-@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa): Animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
+@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
 Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but the they may not respond if $\mu(t)$ is sufficiently high.
+This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
 However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
@@
 Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
+The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022).
 $H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
@@
 
-Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either E(R) or V(R), rather than time.
+Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
 Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
@@
 Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
-Additionally, the simulations also ignore the how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
+Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
 While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.
@@
 The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
-Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the approximately nonlinear increase in $H$ as $\\var(R)$ increases.
+Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
 Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
@@
 
-# Applying the framework: The effects of E(R) and V(R) on a lowland tapir
+# Applying the framework: The effects of $\e(R)$ and $\var(R)$ on a lowland tapir
 
@@
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used `R` package that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
+`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
 Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie families.
@@
 approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants, *Loxodonta sp.*: @foley_severe_2008; muskoxen, *Ovibos moschatus*: @berger_climate_2018; bison, *Bison bison*, and mule deer, *Odocoileus hemionus*: @geremia_migrating_2019].
-Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
+Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
 However, adapting to changes in environmental cues rather than relying on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].

### 2023-07-27 writing/manuscript.Rmd
updated based on Sarah's comments (2)

@@
     - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
+    -  \usepackage{hanging} # for hanging indents in references
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on animals spatial needs while suggesting a second unifying hypothesis.
+Afterwards, we provide a review of the effects of resource *stochasticity* on animals' spatial needs while suggesting a second unifying hypothesis.
 Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
@@
 Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
-Quantifying resources as a numerical random variable (as opposed to using *ad hoc* qualitative descriptions) provides us with the capacity to leverage techniques from probability theory and statistics.
+Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
 
@@
 
-\noindent In this section, we use empirical tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
-We estimate $\e(R)$ and $\var(R)$ by modeling NDVI [@pettorelli_normalized_2011] using a Beta location-scale GAM, after scaling NDVI to [0, 1].
-Appendix 2 contains additional information on how NDVI was modeled for the empirical example, along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
+\noindent In this section, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011].
+Appendix 2 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
 
-The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functional responses.
+The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
 We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
-In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as *de novo* findings.
-Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models with individual-level intercepts and slopes [see models "GS" and "GI" in @pedersen_hierarchical_2019].
-Such models would allow one to test whether members of the same population or species behave according to common functional responses to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
+In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as new findings.
+Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models (HGAMs) with individual-level intercepts and smooth terms [see models "GS" and "GI" in @pedersen_hierarchical_2019].
+Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
+Individuals and populations that were once thought to have different spatial needs due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
 In the following sections, we illustrate how one can apply this framework to empirical data.
@@
 We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
-Next, we provide an empirical example using tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado and NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals spatial needs under different scenarios.
+Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals' spatial needs under different scenarios.
 
@@
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMS, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie families.
+`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
+Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).
 
@@
 However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and NDVI [@pettorelli_normalized_2011].
+available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].
 
-The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass (e.g., kg of carbon per m$^2$), or energetic intake (e.g., kcal/day).
-The Tweedie location-scale family is similar to the Gamma family, but it allows allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density.
+The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->.
+The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009].
 In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
@@
 
-Irrespectively of whether one uses a frequentist or Bayesian approach, it is important to model the mean and variance terms carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the location terms.
+Modeling the mean and variance terms of $R$ should be done carefully.
+Since trends in both $\e(R)$ and $\var(R)$ can be spatiotemporally nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms.
 An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
 Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
-Additionally, using REstricted Marginal Likelihood [REML, a form of penalized maximum likelihood optimization, see @wood_fast_2011] and setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms should provide reasonably accurate results.
+Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
 We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model if $R$ is measured at high frequencies and with a large sample size (but low effective sample size due to the spatiotemporal autocorrelation).
+Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model.
+@simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.
 
@@
 
-\noindent In this paper's empirical example, we estimate resource abundance using NDVI [for more information, see the relevant section later in the paper and @pettorelli_normalized_2011].
-However, since there is no commonly used distribution with a support over the interval $[-1, 1]$ we scale NDVI to the interval $[0, 1]$ to use a Beta distribution by applying the **linear transformation**
+\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the **linear transformation**
 
@@
 
-\noindent where $\nu$ indicates the NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
+\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
 <!--Note that only **linear transformations** (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a Beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
-We use this parameterization here for ease of explanation, but note that Beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
+only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
 
@@
 
-\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate distribution.
-Since $H$ is strictly positive and continuous, a Gamma distribution seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017].
-The Tweedie distribution may also be appropriate, if the mean-variance relationship of the Gamma distribution results in a poor fit.
+\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate family of distributions.
+Since $H$ is strictly positive and continuous, a Gamma family seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017], but the Tweedie family would also be appropriate if the variance does not scale with the squared mean.
 
 As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
-If there is not sufficient data to fit a location scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
-While a GAM would allow the effects of $\e(R)$ and $\var(R)$ on $H$ to be quite flexible, we suggest one use a GLM, instead, (unlike when modeling $R$) because we expect the relationship to be fairly simple and monotonic.
-Restricting the model to a GLM reduces the risk of issues that may arise due to limited sample size and temporal autocorrelation in $\e(R)$ and $\var(R)$.
-If one is interested in fitting a GAM but imposing shape constraints, such as monotonicity, we suggest fitting a shape-constrained GAM via the `scam` package [as we do here: see the section below, Appendix 2, and @pya_scam_2022].
+If there is not sufficient data to fit a location-scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
+Due to potential correlation between $\e(R)$ and $\var(R)$ in a small and autocorrelated sample, the relationship between $R$ and $H$ should be modeled carefully and with parsimony.
+To appropriately constrain the shape of the effects of $\e(R)$ and $\var(R)$, we suggest using a GLM or a shape-constrained GAM (as long as the wiggliness of each term is chosen carefully).
+Here, we model the effects of $\e(R)$ and $\var(R)$ on $H$ using a shape-constrained GAM with via the `scam` package [@pya_scam_2022].
+Additional information on the model can be found in Appendix 2.
 
@@
 
-```{r tapir-mw, fig.cap="Seven-day home-range size of a tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
+```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
 (a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
@@
 See Appendix 2 for additional information.
-The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
 knitr::include_graphics('../figures/tapir-example.png')
@@
 Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018).
-In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable.
+In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (January 2018).
 Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
 Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
-Thus, while the time series shown in panel (c.) may appear complicated, the tapir's response can be well described by two simple functional responses.
+Thus, while the time series shown in panel **c** may appear complicated, the tapir's response can be well described by two simple functions.
 
@@
 
-\noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as *de novo* findings.
-Instead, one can test whether individuals follow these hypotheses, if members of the same population or species behave similarly, and if all members of the group have a common functional response [using hierarchical models, see @pedersen_hierarchical_2019].
-Individuals and populations that were thought to have different behaviors and spatial needs due to differences in behavior may instead be simply responding to different local conditions along the same functional response [e.g., @singh_migration_2012].
+<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
+The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as new findings.
+-->
 
@@
 \noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events which occur at the scale of the organism [i.e.
-approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [e.g., elephants, *Loxodonta sp.*: @foley_severe_2008; muskoxen, *Ovibos moschatus*: @berger_climate_2018; bison, *Bison bison*, and mule deer, *Odocoileus hemionus*: @geremia_migrating_2019].
-Consequently, organisms with long generations (e.g., elephants) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus respond to and prepare for stochastic events by adjusting the timing of reproduction [framewokrk: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hybernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006], moving long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017] including migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019], and developing more complex cognitive abilities over longer lifespans, generation times, and developmental periods [@brown_toward_2004], including the development of memory of such events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-However, adapting to changes in environmental cues rather than relying on memory alone [ungulates: @severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; birds: @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
+Events that occur at the scale of the organism [i.e.
+approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019].
+Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them.
+Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006].
+Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
+<!-- 
+with species names:
+Animals may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006].
+Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
+However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
 
-In contrast, individuals with short lifespans are less likely to experience stochastic events and memory about them.
-However, the short generation time and high reproductive rate of *r-selected* species [*sensu* @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events.
+In contrast, individuals with short lifespans are less likely to experience stochastic events and develop memory about them.
+However, the short generation time and high reproductive rate of r-selected species [as opposed to k-selected species, see @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events.
 Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021].
-Events that occur over large spatiotemporal scales can thus promote slow changes through adaptation, natural selection, and evolution of the species rather than the individual [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which are numerically more likely if the starting population is large and genetically diverse, as in the case of r-selected species [but see @leung_smaller_2022].
+Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022].
 Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than lower temperatures at night than during the day and some weather stochasticity.
@@
 Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay.
-Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
+Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
 However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less sotchastic resources.
@@
 
-\noindent The work presented here provides a unifying framework for viewing animal movement as function of resource abundance and stochasticity.
+\noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance and stochasticity.
 We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
@@
 # References
+
+\hangparas{1em}{1} <!-- indent all lines but the first -->

### 2023-08-06 writing/manuscript.Rmd
updated based on Lauren's comments; fixed fig pos

@@
 
-In this paper, we refer to a location's average amount of resources as **resource abundance**, while use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
+In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use.
@@
 
-When $\e(R)$ is high, we expect animals' spatial requirements be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+When $\e(R)$ is high, we expect animals' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
 However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
@@
 However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
-For example, the abundance of berries is predictably scarce outside the growing season, while it is higher and unpredictable during spring and summer, since a bear may explore a location only after competitors ate all the berries or during a year of unusually low productivity.
+For example, since berries are scarce outside the growing season, the variance in berries will also be low.
+But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location only after competitors ate all the berries.
 In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
@@
 
-Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
-
 ```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
@@
 ```
+
+Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
+If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
 
@@
 Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
-
-```{r stoch-timescale, eval = FALSE, fig.cap="Conceptualizations of perceptions of $\\e(R)$ (a.) and $\\var(R)$ (b.) over time for animals with no plasticity in their movement (blue), seasonal memory but no plasticity (orange), or high plasticity but no memory (green).
-The true trends in $\\e(R)$ and $\\var(R)$ are indicated in black in panels (a.) and (b.), respictively.
-Estimates in the white area indicate the animal's estimated $\\mu(t)$ and $\\sigma^2(t)$ based on past experience, while estimates to the grey shaded region indicate the animals' predictions of $\\mu(t)$ and $\\sigma^2(t)$ in the future.
-The four facets show how each behavior performs in environments where $\\e(R)$ is either predictable (top row) or stochastic (bottom row), and $\\var(R)$ is either predictable (left column) or stochastic (right column).
-Animals that do not adapt to changes in $R$ (blue) are unable to predict $\\e(R)$ accurately, and they have high estimates of $\\var(R)$ when $\\mu(t)$ is far from the long-term, temporally static average.
-Memory of seasonal trends in $\\mu(t)$ and $\\var(R)$ (orange) allows animals to perform well in predictable environments, but their lack of adaptability causes them to fair poorly in stochastic habitats, where they perceive $\\var(R)$ to be high when $\\e(R)$ is far from the expected $\\mu(t)$.
-Finally, the ability to adapt rapidly to change (green) allows animals to track changes in $\\e(R)$ and $\\var(R)$ closely, but the lack of memory prevents accurate predictions in predictable environments.
-Since such animals tend to perceive changes in $R$ as changes in $\\mu(t)$, they tend to have lower estimates of $\\sigma^2(t)$ than other animals."}
-# set eval to false: include in a separate paper?
-knitr::include_graphics('../figures/stochasticity-timescale.png')
-```
 
@@
 `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with a variety families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).
+Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).
 
@@
 Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022].
-Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than lower temperatures at night than during the day and some weather stochasticity.
+Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity.
 Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].

### 2023-08-06 writing/manuscript.Rmd
addressed Aimee's comments (1); corrected typos

@@
 \noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction.
-Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018].
+Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018].
 Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.
 Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, movement represents one of the most readily available traits that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
+However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
 
@@
 
-\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** an animal can access.
+\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** the animal can access.
 Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
 While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
-Instead, results are reported as independent, *de-novo* findings.
+Instead, results are reported as independent, novel findings.
 Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs.
-In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little to no information on how they respond to unpredictable changes in resources.
+In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
 
@@
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use.
-For instance, a grassland with relatively low but constant forage availability will clearly require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
-In the first scenario, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the second scenario may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement strategies and life histories being selected for.
+For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
+In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
+Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for.
 Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
@@
 Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Resource abundance is often unpredictable (and hard to quantify), since it depends on a variety of factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differents in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
+Resource abundance is often unpredictable (and difficult to quantify), since it depends on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
 Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
@@
 
-\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Beroulli).
+\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Bernoulli).
 Since the variable is random, it can take multiple possible values, each with different probabilities.
@@
 
-There are many distributions we can assign to $R$, depending on how we quantify it.
-For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support over all real numbers between 0 and infinity, which we can indicate with $R \in (0, \infty)$,^[We use round parentheses rather than square brackets (i.e., $[0, \infty]$) because neither 0 nor infinity are included in the distribution's support set.] such as a Gamma or log-normal distribution.
+There are many distributions we can assign to $R$ depending on how we quantify it.
+For instance, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
+However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).
+Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution.
 If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
-Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
-However, since there is no commonly used distribution with that support, the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).
 
@@
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-emptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
+If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
 
@@
 We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
-Firstly, $H$ scales superlinearly with the standard diviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
+Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
 Secondly, the animal's movement was simulated to be range resident within an homogeneous environment.
@@
 With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas.
-In light of the UN's objctive to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
+In light of the UN's objective to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
 
@@
 Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
-However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less sotchastic resources.
+However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less stochastic resources.
 
@@
 The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currently need and will need in the future under different climate change scenarios.
+
+# Acknowledgements
+
+We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
+In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
 

### 2023-08-06 writing/manuscript.Rmd
reorganized application section

@@
 
-# Applying the framework: The effects of $\e(R)$ and $\var(R)$ on a lowland tapir
+# Applying the framework
 
-\noindent In this section, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011].
+\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011].
+To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
 Appendix 2 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
@@
 
-We start by providing examples and considerations on how one may model $R$, animals' spatial needs, and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
+We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
 Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
@@
 
-### Estimating resource abundance using NDVI
+### Estimating $R$ using NDVI
 
@@
 
-### Modeling the effects of $\e(R)$ and $\var(R)$
+### Modeling the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs
 
@@
 
-### Empirical results
+### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's spatial needs
 
@@
 The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
-knitr::include_graphics('../figures/tapir-example.png')
+knitr::include_graphics('../figures/tapir-example-with-data.png')
 ```

### 2023-08-06 analysis/tapirus-terrestris/moving-window-figure-anna-betals.R

@@ -151,11 +151,13 @@ preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
                              type = 'response'))
 
 p_d <- ggplot() +
+  # geom_point(aes(mu, hr_est), tapir, color = pal[3], alpha = 0.5) +
   geom_line(aes(mu, hr_mu), preds, color = pal[1], linewidth = 2) +
   xlab(e_r) +
   scale_y_continuous(hr_lab)
 
 p_e <- ggplot() +
+  # geom_point(aes(sigma2, hr_est), tapir, color = pal[3], alpha = 0.5) +
   geom_line(aes(sigma2, hr_sigma2), preds, color = pal[2], linewidth = 2) +
   xlab(v_r) +
   scale_y_continuous(hr_lab)
@@ -180,5 +182,5 @@ p_right <- plot_grid(plotlist = r_grobs, ncol = 1, labels = c('d.', 'e.', 'f.'),
 # add space to avoid cutting off x axis for (e.)
 p <- plot_grid(p_left, p_right, NULL, nrow = 1, rel_widths = c(1, 1, 0.01))
 
-ggsave('figures/tapir-example.png', plot = p, height = 11, width = 14,
-       units = 'in', dpi = 600, bg = 'white')
+ggsave('figures/tapir-example-without-data.png', plot = p, height = 11,
+       width = 14, units = 'in', dpi = 600, bg = 'white')

### 2023-08-24 analysis/aepyceros-melampus-1757/download-ndvi-aepyceros-melampus-1757.R
switched raster to terra

@@ -69,7 +69,7 @@ names(rasters) <-
   as.Date(format = '%Y_%j')
 
 if(FALSE) {
-  raster::plot(rasters[[1]])
+  terra::plot(rasters[[1]])
   plot(ud_poly, add = TRUE, color = 'transparent')
   tel %>%
     SpatialPoints.telemetry() %>%
@@ -80,7 +80,7 @@ if(FALSE) {
   rasters %>%
     mask(ud_poly) %>%
     crop(ud_poly) %>%
-    raster::plot()
+    terra::plot()
 }
 
 # save NDVI data as an rds file of a tibble

### 2023-08-24 analysis/figures/barga-italy-2023/simulation-example-barga-2023.R

@@ -2,6 +2,7 @@ library('ctmm')    # for movement modeling
 library('dplyr')   # for data wrangling (%>%, mutate())
 library('purrr')   # for functional programming (e.g., map*())
 library('tidyr')   # for data wrangling (e.g., unnest())
+library('terra')   # for working with raster data
 library('ggplot2') # for plotting
 library('cowplot') # for plots in grids
 source('functions/label_visits.R') # label food encounters based on a raster
@@ -16,8 +17,9 @@ DELTA_T <- 0.1 # time between measurements
 SAMPLES <- seq(0, 30, by = DELTA_T) # sampling times
 DIM <- 15 # dimensions of raster (number of cells per side)
 HABITAT <- matrix(data = 1, nrow = DIM, ncol = DIM) %>% # raster of patches
-  raster(xmx = 20, xmn = -20, ymx = 20, ymn = -20)
-HABITAT_tbl <- rasterToPoints(HABITAT) %>% as_tibble() # raster in tibble format
+  rast(extent = c(-20, 20, -20, 20))
+HABITAT_tbl <- as.data.frame(HABITAT, xy = TRUE) %>%
+  as_tibble() # raster in tibble format
 
 # resource abundance palette
 LOW <- 'white'
@@ -104,7 +106,7 @@ akde_95 <-
 
 # create the figure ----
 HABITAT_tbl <-
-  rasterToPoints(HABITAT) %>% # convert raster to a data.frame
+  as.data.frame(HABITAT, xy = TRUE) %>% # convert raster to a data.frame
   as_tibble() %>% # convert to a tibble for ease of use
   filter(x >= min(akde_95$long) - 1, x <= max(akde_95$long),
          y >= min(akde_95$lat) - 1, y <= max(akde_95$lat) + 1)

### 2023-08-24 analysis/figures/simulation-example.R

@@ -8,15 +8,15 @@ source('functions/label_visits.R') # label food encounters based on a raster
 source('functions/pull_parameter.R') # to extract parameters from ctmm and akde objects
 source('analysis/figures/default-figure-styling.R') # defaults for figures (theme, size)
 theme_set(theme_map() + theme(legend.position = 'none'))
-select <- dplyr::select # don't use raster::select as a default
 
 REQUIRED <- 8 # amount of food required for satiety
 DELTA_T <- 0.1 # time between measurements
 SAMPLES <- seq(0, 30, by = DELTA_T) # sampling times
 DIM <- 15 # dimensions of raster (number of cells per side)
 HABITAT <- matrix(data = 1, nrow = DIM, ncol = DIM) %>% # raster of patches
-  raster(xmx = 20, xmn = -20, ymx = 20, ymn = -20)
-HABITAT_tbl <- rasterToPoints(HABITAT) %>% as_tibble() # raster in tibble format
+  rast(extent = c(-20, 20, -20, 20))
+HABITAT_tbl <- as.data.frame(HABITAT, xy = TRUE) %>%
+  as_tibble() # raster in tibble format
 
 # resource abundance palette
 LOW <- 'white'
@@ -103,7 +103,7 @@ akde_95 <-
 
 # create the figure ----
 HABITAT_tbl <-
-  rasterToPoints(HABITAT) %>% # convert raster to a data.frame
+  as.data.frame(HABITAT, xy = TRUE) %>% # convert raster to a data.frame
   as_tibble() %>% # convert to a tibble for ease of use
   filter(x >= min(akde_95$long) - 1, x <= max(akde_95$long),
          y >= min(akde_95$lat) - 1, y <= max(akde_95$lat) + 1)

### 2023-08-24 analysis/simulations/delta-t-sensitivity.R

@@ -1,204 +1,0 @@
-library('terra')  # for working with rasters
-library('ctmm')    # for generating movement models
-library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
-library('tidyr')   # for data wrangling (unnest, etc.)
-library('purrr')   # for functional programming (map_*(), etc.)
-library('ggplot2') # for fancy plots
-library('cowplot') # for multi-panel fancy plots
-source('analysis/simulations/movement-model.R') # for consistency between scripts
-theme_set(theme_bw() + # change default ggplot theme
-            theme(legend.position = 'none', strip.background = element_blank(),
-                  strip.text = element_blank()))
-
-# simulate some movement
-get_tracks <- function(.seed) {
-  simulate(model, # ctmm movement model
-           t = 0:(max(SAMPLES)), # 12 hours of data, sampled every second
-           seed = .seed, # set a seed for consistent results
-           complete = TRUE, # prevents missing projection warning
-           crs = PROJECTION) # projection CRS string
-}
-tels <- tibble(seed = 1:3, tel = map(seed, get_tracks)) %>%
-  mutate(seed = factor(seed))
-
-# plot intervals between encounters
-intervals <-
-  transmute(tels,
-            seed,
-            ints = map(tel, \(.tel) {
-              .tel %>%
-                data.frame() %>%
-                mutate(cell_id = cellFromXY(object = HABITAT,
-                                            xy = SpatialPoints.telemetry(.tel) %>%
-                                              as.data.frame() %>%
-                                              select(x, y)),
-                       new_cell = c(1, diff(cell_id)) != 0) %>%
-                filter(new_cell) %>%
-                mutate(interval = c(NA, diff(t)))
-            })) %>%
-  unnest(ints)
-
-# check the main quantiles
-quantile(intervals$interval, na.rm = TRUE)
-
-# histogram of number of events as by the time interval between events
-p_hist <-
-  ggplot(intervals, aes(interval)) +
-  facet_wrap(~ seed, ncol = 1) +
-  geom_histogram(aes(fill = seed), binwidth = 30, center = 15, lwd = 0.1,
[…]

### 2023-08-24 analysis/simulations/movement-model.R

@@ -5,7 +5,7 @@ SAMPLES <- seq(0, 60 * 60 * 12, by = DELTA_T) # sample 12 hours every DELTA_T se
 # projected raster of resources
 PROJECTION <- '+proj=aeqd +lon_0=0 +lat_0=0 +datum=WGS84'
 HABITAT <- matrix(data = 1, nrow = 500, ncol = 500) %>%
-  raster(xmx = 1e3, xmn = -1e3, ymx = 1e3, ymn = -1e3, crs = PROJECTION)
+  rast(extent = c(-1e3, 1e3, -1e3, 1e3), crs = PROJECTION)
 
 # infinitely diffusive movement model
 model <- ctmm(tau = c(Inf, 1e3), sigma = 0.1, mu = c(0, 0))

### 2023-08-24 writing/appendix-1-simulations.Rmd
cleaned appendix 1

@@
     email: stefano.mezzini@ubc.ca
-    institute: ubco
-  - name: Adam T. Ford
-    institute: ubco
-  - name: Jacob R. Goheen
-    institute: wyom
+    institute: [biol, braes]
+  - name: Simon Wood
+    institute: uofe
   - name: E. Patrícia Medici
@@
     email: michael.noonan@ubc.ca
-    institute: ubco
+    institute: [biol, braes, cmps]
 institute:
-  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
-  - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
@@
 header-includes:
+    - \usepackage{setspace}\doublespacing # for double-spaced text
     - \usepackage{indentfirst} # for indenting first line of each paragraph
@@
 ```{r, echo=FALSE}
+hook_chunk <- knitr::knit_hooks$get('chunk')
+
+knitr::knit_hooks$set(chunk = function(x, options) {
+  regular_output = hook_chunk(x, options)
+  # add latex commands if chunk option singlespacing is TRUE
+  if (isTRUE(options$singlespacing)) 
+    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
+  else
+    regular_output
+})
+
 knitr::opts_chunk$set(eval = FALSE,         # do not run R code
@@
                       comment = '',         # don't add anything before outputs
-                      cache = TRUE)    # cache all chunk outputs
-knitr::opts_knit$set(root.dir = '..')
+                      fig.align = 'center', # change figure alignment
+                      cache = TRUE,         # cache all chunk outputs
+                      singlespacing = TRUE) # use single spacing for code chunks
 ```
@@
 \newcommand{\var}{\mathbb V}
-\newcommand{\vpos}{\mathbb V(\text{position})}
 
@@
 
-* fix captions (incomplete, HR not V(pos)); take captions from manuscript
+* fix captions (incomplete); take captions from manuscript
 
@@
 
-```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated spatial requirements for animals living in habitats where the mean and variance in resource abundance ($R$) are constant, linearly increasing, cyclical, drifting, or erratic over time.
-The lower green line indicates the animal's core home range (50\\% quantile), while the top line indicates the 0.95 utilization quantile.
-Note how both quantiles decrease nonlinearly as $\\e(R)$ increases, and they increase nonlinearly as $\\var(R)$ increases.
-Additionally, the variance in both quantiles is higher when $\\var(R)$ is higher, and changes in $\\var(R)$ have greater impacts when $\\e(R)$ is low.\nSimulations were run such that animals followed the same 1000 tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they returned to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's home range was then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021).
-", out.width='100%'}
+```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
+Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
+The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022).
+$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.", out.width='100%'}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/mean-variance-5-by-5-hr-sims.png',
@@
 ```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Linear Model with a Gamma family of distributions that accounted for the effects of both $\\e(R)$ and $\\var(R)$ as well as differences between the two quantiles.
-The bottom line indicates the relationships with the animal's core home range (0.5 quantile), while the top line indicates the relationship with the 0.95 utilization quantile.
-Note the nonlinear decrease in both utilization quantiles as $\\e(R)$ increases and the nonlinear increase in both utilization quantiles as $\\var(R)$ increases.", out.width='100%'}
+The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
+The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both the location (i.e., mean) of $H$ and its scale parameter.
+Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
+Additionally, note how the variance in spatial needs strongly increases with mean spatial needs.", out.width='100%'}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-regression-plots.png',
@@
 
-To reduce sampling variance between simulations, we use the same set of simulated tracks for each time point in each panel (figure \@ref(fig:sims-example)).
-In the `analysis/simulations/tracks.R` script, we generate $2^{10} = 1024$ tracks to check how many tracks are necessary to obtain stable home range estimates in the best- and worst-case scenarios.
-Most intermediate and diagnostic checks included in the `R` scripts are not included in this document for the sake of brevity and simplicity.
-In this script, we use the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for movement modeling, the `raster` package [version 3.6-3, @hijmans_raster_2022] to work with the simulated resource rasters, the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling, and the the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
+To reduce sampling variance between simulations at each time point in each panel of figure \@ref(fig:5-5-sims)), we used the same set of $2^{10} = 1024$ simulated tracks, which we generate in the `analysis/simulations/tracks.R` script.
+Most intermediate and diagnostic checks are not included in this document for the sake of brevity and simplicity, but their outputs and conclusions are listed in this section.
+In this script, we use the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for movement modeling; the `terra` package [version 1.7-39, @hijmans_terra_2023] to work with the simulated resource rasters; the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling; and the the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
 
@@
 (d.) The process was repeated $2^{10}=1024$ times (13 tracks pictured in this panel).
-(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% and 50\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
+(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-example.png',
@@
 
+\newpage
+
 ```{r simulate-tracks, eval=TRUE, warning=FALSE}
-# NOTE: change working directory to "hr-resource-stoch" or modify file paths
+# NOTE: change working directory to "hr-resource-stoch" or modify paths
 setwd('H:/GitHub/hr-resource-stoch')
@@
 library('ctmm')    # for generating movement models and movement modeling
-library('raster') # for working with raster data
+library('terra')   # for working with raster data
 library('dplyr')   # for data wrangling
@@
 
+# source custom functions
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/figures/mean-variance-trends-panel-data.R') # mu and sigma2
-source('analysis/movement-model.R') # for consistency across scripts
+source('analysis/mean-variance-trends-panel-data.R') # mu and sigma2
+source('analysis/simulations/movement-model.R') # for consistency
 source('functions/get_hr.R') # for extracting gaussian home range
@@
 
-Script: `analysis/figures/simulations/return-sensitivity.R`
+Script: `analysis/figures/return-sensitivity.R`
 
-Adding return trips to $\langle0, 0\rangle$ after an animal reached satiety doubled computational times without appreciable improvements on the home range estimates (including the 95% CIs).
+Adding return trips to $\langle0, 0\rangle$ after an animal reached satiety doubled computational times without appreciable improvements on the home range estimates (including the 95% confidence intervals of the estimates).
 
@@
 
-Script: `analysis/figures/simulations/delta-t-sensitivity.R`
+Script: `analysis/figures/delta-t-sensitivity.R`
 
-Using three tracks generated using three arbitrary seeds (1, 2, and 3), we explored the effects of sampling interval on the number of encounters with food (i.e., movements to new cells) detected.
+Using three tracks generated with three arbitrary seeds (1, 2, and 3), we explored the effects of sampling interval ($\Delta t$) on the number of encounters with food (i.e., movements to new cells) detected.
 From each of the four checks, we created an exploratory plot that we present in figure \@ref(fig:dt-sensitivity).
@@
 
-**Exploratory plot \@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks while avoiding excessive amounts of data and an inflated amount of encounters with resources while an animal was near cell boundaries.
+**Exploratory plot \@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
 
 **Exploratory plot \@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
-All three tracks have a reasonable amount of tortuous movement and directed movement.
+All three tracks have a reasonable amount of both tortuous and directed movement.
 
@@
 (a.) Histograms of the number of encounters as a function of the interval between encounters, with a binwith of 30 seconds.
-Although some encounters occur with less than 30 seconds between them, most of them occur at least 60 seconds apart.
-(b.) Number of encounters with food detected as a function of sampling interval.
+Although some encounters occur with less than 30 seconds between them, approximately 93\\% of them occur at least 30 seconds apart.
+(b.) Number of detected encounters as a function of sampling interval.
 The colored lines indicate the estimated relationship based on a Generalized Additive Model fit using the $\\tt{geom\\_smooth}$ function from the $\\tt{ggplot2}$ package.
-Although the number of encounters detected decreases as sampling interval doubles, the loss at 30 seconds is negligible.
-(c.) Beginning of the track generated with seed \"1\" (purple line) with the location of each sample for different sampling intervals.
-Red dots indicate samples where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell, and thus encountered food.
+Although the number of encounters detected decreases as sampling interval doubles, the loss at $\\Delta = 30$s is negligible.
+(c.) Beginning of the track generated with seed \"1\" (purple line) for different sampling intervals.
+Red dots indicate locations where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell and thus encountered food.
 While the number of encounters detected decreases as the sampling interval increases, most of the encounters lost at $\\Delta t = 30$s occured because the animal remained almost adjacent to the borders between cells.
-(d.) The three tracks used in these tests over the raster used to determine when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
+Additionally, the track at $\\Delta t = 30$s is still sufficiently tortuous to represent realistic animal movement.
+(d.) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/thinning-examples.png',
@@
 In this script, we check how many tracks are necessary to produce stable and accurate estimates of the spatial needs.
-We do so by estimating the home ranges of animals in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
+We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
 
@@
 set.seed(1) # for consistent results
-
 tels <- readRDS('simulations/tracks.rds') # list of telemetry tracks
@@
 days <-
-  # modify WORST and BEST to follow the syntax in 'offspring-simulations.R'
   transmute(bind_rows(WORST, BEST),
@@
   # calculate the total visits, total calories, and if animal is full
-  mutate(satiety = cumsum(food), # for diagnostics if animals don't get full
+  mutate(satiety = cumsum(food), # for diagnostics if animals aren't full
          full = satiety >= REQUIRED) %>% # did the animal reach its needs?
@@
     group_by(scenario, day) %>%
-    filter(full, ! duplicated(full)) %>% # take 1st row where animal is full
+    filter(full, ! duplicated(full)) %>% # take 1st row where full
     rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
@@
   mutate(t_start = lag(2 * t_expl), # add the return time before next "day"
-         t_start = if_else(is.na(t_start), 0, t_start), # start at 0, not NA
+         t_start = if_else(is.na(t_start), 0, t_start), # start at 0
          t_start = cumsum(t_start), # make start times comsecutive
          tel_day = map2(day, t_expl,
-                        \(i, te) tels$tel[[i]] %>% # extract tel for the day
+                        \(i, te) tels$tel[[i]] %>% # extract day's tel
                           data.frame() %>% # for filtering
@@
          individual.local.identifier = scenario, # ctmm identifier
-         timestamp = as.POSIXct(t, origin = '2000-01-01')) %>% # use new times 
+         timestamp = as.POSIXct(t, origin = '2000-01-01')) %>% # new times 
   ungroup() # remove grouping by scenario
@@
          sigma = map_dbl(m, \(.m) ctmm:::area.covm(.m$sigma)), # var(pos)
-         hr = get_hr(.sigma = sigma, quantile = 0.95)) # Gaussian home range
+         hr = get_hr(.sigma = sigma, quantile = 0.95)) # Gaussian HR
 
@@
 
+\newpage
+
 # Main scripts (to be run in the following order)

### 2023-08-24 writing/manuscript.Rmd
minor change to a figure caption

@@
 ```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions.
-The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
+The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
+The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both the location (i.e., mean) of $H$ and its scale parameter.
 Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
+Additionally, note how the variance in spatial needs strongly increases with mean spatial needs.", out.width='100%'}
 knitr::include_graphics('../figures/simulation-regression-plots.png')
@@
 \noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 50% and 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Although here we quantify animal's spatial needs using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
 Note, however, that although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's movement but not its behavior.

### 2023-08-24 writing/appendix-1-simulations.Rmd
added hanging paragraphs in refs

@@
     - \usepackage{setspace}\doublespacing # for double-spaced text
-    - \usepackage{indentfirst} # for indenting first line of each paragraph
     - \usepackage[small]{titlesec} # for smaller font for headings
@@
     - \captionsetup[figure]{font=scriptsize}
+    - \usepackage{hanging} # for hanging indents in references
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 \tableofcontents
-
-# To do {-}
-
-* fix captions (incomplete); take captions from manuscript
 
@@
 # References {-}
+
+\hangparas{1em}{1} <!-- indent all lines but the first -->

### 2023-08-24 writing/appendix-2-empirical-modeling.Rmd
finished appendix 2

--- 2022-11-03 writing/appendix-2-empirical-modeling.Rmd
@@
     email: stefano.mezzini@ubc.ca
-    institute: ubco
-  - name: Emilia Patrícia Medici
+    institute: [biol, braes]
+  - name: Simon Wood
+    institute: uofe
+  - name: E. Patrícia Medici
     institute: [tapir, iucn, escas]
@@
     email: michael.noonan@ubc.ca
-    institute: ubco
+    institute: [biol, braes, cmps]
 institute:
-  - ubco: The Irving K. Barber Faculty of Science, The University of British Columbia, Okanagan Campus, Kelowna, Canada.
-  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960‑000, Brazil.
-bibliography: 'env-var-review.bib'
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+bibliography: 'hr-resource-stoch.bib'
 csl: 'the-american-naturalist.csl'
 fontsize: 12pt
-indent: true
 header-includes:
-    - \usepackage{indentfirst} # for indenting first line of each paragraph
+    - \usepackage{setspace}\doublespacing # for double-spaced text
     - \usepackage[small]{titlesec} # for smaller font for headings
@@
     - \captionsetup[figure]{font=scriptsize}
+    -  \usepackage{hanging} # for hanging indents in references
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
 
-```{r, echo=FALSE}
-knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, fig.align = 'center')
+```{r setup, echo=FALSE}
+hook_chunk <- knitr::knit_hooks$get('chunk')
+
+knitr::knit_hooks$set(chunk = function(x, options) {
+  regular_output = hook_chunk(x, options)
+  # add latex commands if chunk option singlespacing is TRUE
+  if (isTRUE(options$singlespacing)) 
+    sprintf("\\singlespacing\n %s \n\\doublespacing", regular_output)
+  else
+    regular_output
+})
+
+knitr::opts_chunk$set(eval = FALSE,         # do not run R code
+                      echo = TRUE,          # print R code
+                      message = FALSE,      # hide messages
+                      comment = '',         # don't add anything before outputs
+                      fig.align = 'center', # change figure alignment
+                      cache = TRUE,         # cache all chunk outputs
+                      singlespacing = TRUE) # use single spacing for code chunks
+knitr::opts_knit$set(root.dir = '..')
 ```
@@
 
-# To do {-}
+\newpage
 
-* add scripts to the correct folders and check
-* add "for more info, see the tapir project" at the beginning
-* add refs for each package used
+This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript.
+The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
+For ease of reference, we also include the figure below (figure \@ref(fig:tapir-mw)).
+Estimating the effects of resource abundance and unpredictability on the tapir's spatial needs requires us to first estimate the changes in the tapir's spatial needs (section \@ref(movement), figure \@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figures \@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figures \@ref(fig:tapir-mw)d-f).
+
+```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
+(a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
+(b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir.
+(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
+(d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
+The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
+(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
+Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
+See Appendix 2 for additional information.
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
+knitr::include_graphics('../figures/tapir-example-with-data.png')
+```
+
+To minimize the computational costs of creating this appendix, we load the necessary objects through hidden `R` chunks rather than re-running all the code.
+Still, those interested in replicating the analyses can do so by using the code in the pdf document or the related `R` Markdown (`Rmd`) document (as well as the `R` scripts).
+All the packages and source scripts required to run the analyses in this document are listed in the code chunk below.
+For spatial data, we use the `MODIStsp` package [version 2.0.9, @busetto_modistsp_2016] to download the NDVI rasters, the `terra` package [version 1.7-39, @hijmans_terra_2023] to work with the NDVI rasters, and the `sf` package [version 1.0-8, @pebesma_simple_2018].
+We use the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling and the `lubridate` package [version 1.8.0, @grolemund_dates_2011] for converting calendar dates to decimal dates.
+Finally, we used the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] and the `mgcv` package [version 1.8-41, @wood_generalized_2017] for modeling and the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
 
@@
 
-This appendix illustrates the steps necessary to reproduce the tapir movement analysis and figures in the main manuscript. For ease of reference, we also include the figure here:
+We start by attaching all necessary packages and custom functions we need:
 
-```{r moving-window-examples, fig.cap="Seven-day home range size of a tapir ($Tapirus~terrestris$) in response to changes in mean and variance in resource abundance.
-(a.) GPS tracking data of the tapir.
-(b.) Mean resource abundance esimated as the mean NDVI at the locations visited by the tapir.
-(c.) Varince in resource abundance esimated as the average variance in NDVI at the locations visited by the tapir.
-(d.) Estimated home range size during each seven-day period, based on 50\\% (bold) and 95\\% (thin) utilization quantiles.
-(e.) Effect of resource abundance on home range size.
-(f.) Effect of resource unpredictability on home range size.
-The effects in panels e and f were estimated using generalized linear models with Gamma conditional distributions.
-The tapir movement data corresponds to the individual named \"Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
-knitr::include_graphics('../figures/tapir-example.png')
-```
+```{r, eval=TRUE}
+getwd() # change working directory or file paths if needed
 
-Estimating the effects of resource abundance and unpredictability on the tapir's spatial needs requires us to first estimate the changes in the tapir's spatial needs (section \@ref(movement)) and the changes in resource abundance and variance (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use to recreate the final figure (section \@ref(final-fig)).
-
-To minimize the computational costs of creating this appendix, we load the necessary objects through hidden `R` chunks rather than running all the code in the final pdf.
-Still, those interested in replicating the analysis can do so by using the code in the pdf document or the related `R` markdown (`Rmd`) document.
-Listed below are all the packages and source scripts required to run the analyses in this document:
-
-```{r, eval=TRUE, comment=FALSE}
 # attach all necessary packages
-library('raster')    # to import and save rasters
+library('terra')    # to import and save rasters
 library('dplyr')     # for data wrangling
@@
 library('ctmm')      # for movement modeling
-library('mgcv')      # for empirical Bayesian modeling
+library('mgcv')      # for empirical Bayesian GAMs
+library('scam')      # for shape-constrained splines
 library('lubridate') # for smoother date wrangling
@@
 library('MODIStsp')  # for downloading NDVI rasters
+theme_set(theme_bw()) # change default theme
 
-# resolve conflicts between packages
-select <- dplyr::select
+# attach the beta location-scale family written by Simon Wood
+source('../ndvi-stochasticity/functions/betals.r')
 
 # source all necessary scripts
-source('../analysis/figures/default-figure-styling.R') # NDVI color palette
-source('../earthdata-login-info.R') # personal login info for EarthData
-source('../analysis/figures/default-figure-styling.R') # for color palettes
-source('../functions/window_hr.R') # function to calculate HRs
+source('analysis/figures/default-figure-styling.R') # for color palettes
+source('earthdata-login-info.R') # personal login info for EarthData
+source('functions/window_hr.R') # function to calculate HRs
 ```
+
+\newpage
 
@@
 
-The script `analysis/tapir/tapirs-moving-window.R` estimates the seven-day spatial use of various tapirs from the Brazilian Cerrado.
+\noindent The script `analysis/tapir/tapirs-moving-window.R` estimates the seven-day spatial use of various tapirs from the Brazilian Cerrado.
 Here, we simplified the code so that it only estimates the spatial use of the tapir in the manuscript, Anna:
@@
 # import tapir data from https://github.com/StefanoMezzini/tapirs
-anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
+anna <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+  filter(name.short == 'ANNA')
+anna_tel <- anna$data[[1]] # telemetry data
+
+# re-project using the appropriate UTM projection for the Brazilian Cerrado
+ctmm::projection(anna_tel) <- '+proj=utm +zone=22 +datum=NAD83 +units=m'
+
+# calculate the 7-day home range estimate
+window_hr(
+  tel = anna_tel,
+  window = 7 %#% 'day', # 1 week of data for sufficient sample size
+  dt = 1 %#% 'day', # move window over by a single day each time
+  fig_path = 'figures',
+  rds_path = 'models')
+anna_mw <- readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
+anna_mw
+```
+
+```{r, echo=FALSE, eval=TRUE}
+anna <- readRDS('../tapirs/models/tapirs-final.rds') %>%
   filter(name.short == 'ANNA')
@@
 
-# calculate the 7-day home range estimate
-anna_mw <-
-  window_hr(
-    anna_tel,
-    window = 7 %#% 'day', # 1 week of data for sufficient sample size
-    dt = 1 %#% 'day', # move window over by a single day each time
-    fig_path = '../figures',
-    rds_path = '../models'
-  )
+anna_mw <- readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
+anna_mw
 ```
 
-```{r, echo=FALSE, eval=TRUE}
-anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
-  filter(name.short == 'ANNA')
-anna_tel <- anna$data[[1]] # telemetry data
+The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days (using `ctmm`'s `%#%` operator for unit conversions) that starts with the set of days from `r format(as.POSIXct(anna_mw$t_start[1]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[1]), format = '%Y-%m-%d')` (extremes included), then shifts by one day (`dt = 1 %#% 'day'`), repeats the analysis for the next seven-day set (`r format(as.POSIXct(anna_mw$t_start[2]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[2]), format = '%Y-%m-%d')`), and continues doing so until it reaches the last set of days, `r format(as.POSIXct(anna_mw$t_start[nrow(anna_mw)]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[nrow(anna_mw)]), format = '%Y-%m-%d')`.
+For each set of days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019].
+Finally, it saves an exploratory figure (figure \@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
 
-# projection for the region in the Brazilian Cerrado
-ctmm::projection(anna_tel) <- '+proj=utm +zone=22 +datum=NAD83 +units=m'
-
-anna_mw <- readRDS('../models/CE_31_ANNA-window-7-days-dt-1-days.rds')
+```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function.
+Panel $\\mathbf{a.}$ shows the tapir's GPS locations, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals.", out.width='\\textwidth'}
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png', rel_path = FALSE)
 ```
 
-The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days that starts with the first seven days of data (2017-06-24 to 2017-07-01) and then repeats the analysis for the second seven-day set (2017-06-25 to 2017-07-02) until it reaches the last set of days (2018-09-23 to 2018-09-30).
-It then saves an exploratory figure (figure \@ref(fig:mw)) to the `figures` folder and the list of movement models to the `models` folder.
-
-```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function.
-Panel $\\mathbf{a.}$ shows the tapir's movement data, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals."}
-knitr::include_graphics('../figures/CE_31_ANNA-window-7-days-dt-1-days.png')
-```
+\newpage
 
@@
 
-To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli-ref].
+To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli_using_2005; @pettorelli_normalized_2011].
 We downloaded the data using the `MODIStsp` `R` package with the following code:
@@
 ```{r}
-anna_ud <- anna$akde[[1]]
+anna_ud <- anna$akde[[1]] # extract the tapir's utilization distribution
 
 bbox <-
-  SpatialPolygonsDataFrame.UD(anna_ud,
+  SpatialPolygonsDataFrame.UD(anna_ud, # convert to a spatial object
                               level.UD = 0.9995, # utilization quantile
@@
 MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
-         out_folder = '../data/ndvi-rasters/tapir-anna',
+         out_folder = 'data/ndvi-rasters/tapir-anna',
          selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)',
@@
 # save NDVI data as an rds file of a tibble
-anna_ndvi <-
-  list.files(
-    path = '../data/ndvi-rasters/tapir-anna/VI_16Days_250m_v61/NDVI/',
+list.files(path = 'data/ndvi-rasters/tapir-anna/VI_16Days_250m_v61/NDVI/',
            pattern = '.tif', full.names = TRUE) %>%
-  stack() %>% # import all rasters as a single stack
-  rasterToPoints() %>% # convert to a matrix of points
-  data.frame() %>% # convert to a data frame
+  rast() %>% # import all rasters as a single stack
+  as.data.frame(xy = TRUE) %>% # convert to a data frame
   pivot_longer(-c(x, y)) %>% # change to long format (x, y, name, value)
@@
             ndvi = value, # rename value column
-            dec_date = decimal_date(date))
+            dec_date = decimal_date(date)) %>%
+  saveRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds')
+
+# import NDVI data
+anna_ndvi <-
+  readRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
+  mutate(dec_date = decimal_date(date))
+anna_ndvi
 ```
@@
 anna_ndvi <-
-  readRDS('../data/tapir-anna-ndvi-data.rds') %>%
+  readRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
   mutate(dec_date = decimal_date(date))
+anna_ndvi
 ```
@@
 
-```{r, echo=TRUE, eval=TRUE, fig.height=4, fig.width=5, out.width='100%'}
+```{r, echo=TRUE, eval=TRUE, fig.height=8, fig.width=10, out.width='100%', }
 anna_ndvi %>%
@@
 
-Ideally, one should model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1.
-The beta family would be appropriate after applying the linear transformation $$y^* = \frac{y + 1}{2},$$ where $y$ is the original NDVI value (between -1 and 1) while $y^*$ is the NDVI value scaled between 0 and 1.
-However, the `mgcv` package [@ref] does not allow one to model changes in variance in NDVI using a location-scale model with a beta family.
-Although the `brms` package allows one to fit beta location-scale models [@ref], the computational costs of using a fully Bayesian approach can be prohibitive for large datasets.
-Therefore, we decided to use a location-scale Gaussian model via the `mgcv` package (`family = gaulss()` in the code chunk below).
-Given the large amounts of data, the predictions are sufficiently far from -1 and 1 that any bias appears to be negligible.
+Next, we estimate the mean and variance in NDVI using a Generalized Additive Model for location and scale [GAMLS: @stasinopoulos_generalized_2007].
+Ideally, we would model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1, but no such family is available in the `mgcv` package.
+However, since the NDVI data is far from 0 (see the histogram below), we decided to keep NDVI unscaled.
+This prevents the model from predicting values that correspond to water or snow (which are not expected to occur in the study area).
+Mathematically, this approach would be comparable to predicting in a Bayesian framework with a prior with zero probability for any NDVI values below zero.
+In environments where values below zero are probable, one could use the beta family after applying the linear transformation $$y^* = \frac{y + 1}{2},$$ where $y$ is the original NDVI value (between -1 and 1) and $y^*$ is the NDVI value scaled between 0 and 1.
+
+We fit the beta GAMLS via the `mgcv` package (note `family = betals()` in the code chunk below).
+The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see figure \@ref(fig:gamls-terms)).
+The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.
+
+```{r, echo=FALSE, eval=TRUE, fig.height=4}
+ggplot(anna_ndvi, aes(ndvi)) +
+  geom_histogram(na.rm = TRUE) +
+  labs(x = 'NDVI', y = 'Count') +
+  xlim(c(-1, 1))
+```
+
+\clearpage
 
@@
     # mean predictor
-    ndvi ~
-      s(long, lat, bs = 'ds', k = 20) + # mean over space (k = 20 * 20)
+    ndvi ~ # not scaling because range is in (0, 1)
+      s(long, lat, bs = 'ds', k = 20) + # mean over space (uses k^2 df)
       s(dec_date, bs = 'tp', k = 20),   # high k to account for adaptions
-    # precision (1/standard deviation) predictor
+    # scale predictor (sigma2 = mu * (1 - mu) * scale)
     ~
-      s(long, lat, bs = 'ds', k = 10) + # precision over space
-      s(dec_date, bs = 'tp', k = 10)),  # precision over time
-    family = gaulss(b = 0.0001), # minimum standard deviation of 0.0001
+      s(long, lat, bs = 'ds', k = 10) + # scale over space (uses k^2 df)
+      s(dec_date, bs = 'tp', k = 10)),  # scale over time
+    family = betals(),
     data = anna_ndvi,
     method = 'REML') # performs better than GCV (the default)
-
-# plot smooths (predictions are on the link scales!)
-plot(m_ndvi, pages = 1, scheme = 1, n = 250)
 ```
 
-```{r, echo=FALSE, eval=TRUE}
-m_ndvi <- readRDS('../models/CE_31_ANNA-mgcv-ndvi-gaulss.rds')
-plot(m_ndvi, pages = 1, scheme = 1, n = 250)
+Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term.
+While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial.
+Excessively smooth terms for the mean can inflate the scale term (figure \@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (figure \@ref(fig:oversmooth)).
+Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
+If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.
+Additionally, note that computation time depends strongly on the complexity of the two-dimensional spatial terms, since changes in `k` alter the size of the term quadratically (since the total size is $\tt k \times \tt k = \tt k^2$).
+
+\clearpage
+
+```{r gamls-terms, echo=FALSE, eval=TRUE, fig.height=8, fig.cap="Estimated spatiotemporal trends in mean and scale parameters using the model detailed in the code chunk above.
+Estimates are provided on the logit link scale.
+The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
+Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
+m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
+plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
 ```
+
+```{r}
+gam(list(
+  # mean predictor (default ks are too small)
+  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
+  # scale predictor (excessively high k)
+  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
+  family = betals(), data = anna_ndvi, method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+
+gam(list(
+  # mean predictor (ks too large)
+  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
+    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
+  # scale predictor (ks too small)
+  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
+  family = betals(), data = anna_ndvi, method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+\clearpage
+
+```{r undersmooth, fig.cap="Estimated mean and scale parameters with excessively small basis size $\\tt{k}$ for the mean's smooth terms and excessively large for the scale's smooth terms.
+The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
+Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=8, fig.pos='!h', echo=FALSE}
+gam(list(
+  # mean predictor (default ks are too small)
+  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
+  # scale predictor (excessively high k)
+  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
+  family = betals(), data = anna_ndvi, method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+\clearpage
+
+```{r oversmooth, fig.cap="Estimated mean and scale parameters in NDVI with excessively large basis size $\\tt{k}$ for the smooth term of time ($\\tt{dec\\_date}$) for the mean in NDVI and excessively low for the variance temporal smooth term.
+The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
+Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=8, fig.pos='!h', echo=FALSE}
+gam(list(
+  # mean predictor (ks too large)
+  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
+    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
+  # scale predictor (ks too small)
+  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
+  family = betals(), data = anna_ndvi, method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
+\newpage
 
@@
 
-We start by simulating a movement track that passes near all recorded locations using the movement model estimated from the telemetry data.
-Note that this step can take hours to run, so we suggest skipping it if one is only interested in replicating the most important portions of the analysis (since the track is only used for panel $\mathbf{a.}$ in the figure).
+We start by predicting the mean and variance in NDVI experienced by the tapir in the known positions using the beta GAMLS.
 
-```{r}
-# simulate the movement track with an interval of 100 seconds
-anna_ouf <- anna$model[[1]]
-track <-
-  predict(anna_ouf, data = anna_tel, dt = 100, complete = TRUE) %>%
-  data.frame() %>%
-  select(timestamp, longitude, latitude) %>%
+```{r, eval=TRUE}
+anna_tel <-
+  data.frame(anna_tel) %>% # convert telemetry to data frame
   rename(long = longitude, lat = latitude) %>%
-  mutate(dec_date = decimal_date(timestamp))
+  mutate(dec_date = decimal_date(timestamp)) %>% # needed for predictions
+  bind_cols(., # bind telemetry to predictions
+            predict(m_ndvi, newdata = ., type = 'response',
+                    se.fit = FALSE) %>%
+              data.frame() %>% # convert list of predictions to data frame
+              # didn't scale NDVI to [0, 1], so no need to back-transform
+              transmute(mu = X1, sigma2 = X1 * (1 - X1) * X2)) %>%
+  as_tibble()
 ```
 
-```{r, eval=TRUE, echo=FALSE}
-track <- readRDS('../models/CE_31_ANNA-track-dt-100.rds')
+Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period.
+
+```{r, eval=TRUE}
+tapir <-
+  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+  mutate(est = map(dataset,
+                   \(.d) filter(anna_tel, timestamp %in% .d$timestamp)),
+         mu = map_dbl(est, \(.d) mean(.d$mu)),
+         sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
+  select(t_center, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
+  pivot_longer(c(hr_lwr_95, hr_est_95, hr_upr_95),
+               names_to = c('.value', 'quantile'),
+               names_pattern = '(.+)_(.+)') %>%
+  mutate(t_center = as.POSIXct(t_center),
+         quantile = paste0(quantile, '%'))
 ```
 
-To create panels $\mathbf{b.}$-$\mathbf{f.}$, we use the NDVI Gaussian location-scale model to estimate the mean and variance in NDVI for each known location of the tapir.
+We now have all we need to create the left side of figure \@ref(fig:tapir-mw).
 
-```{r, eval=TRUE}
-# function to make predictions from the NDVI model based on telemetry data
-ndvi_preds <- function(.data) {
-  .data <- .data %>%
-    data.frame() %>% # convert telemetry data to a data.frame
-    mutate(year = year(timestamp),
-           doy = yday(timestamp))
+```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
+theme_set(theme_bw()) # need to set the theme again for some reason
 
-  predict.gam(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE)%>%
-    data.frame() %>% # from list to data frame
-    tibble() %>% # from data frame to tibble
-    transmute(mu = X1, # rename mean column
-              sigma2 = (1/X2)^2) %>% # convert precision to variance
-    return()
+# create the figure
+date_labs <- range(anna_tel$timestamp) %>% as.Date()
+YLIMS <- c(0, 13)
+
+# axis labels (spaces in hr_lab to avoid overlap with the plot label)
+# removed unicode characters to avoid knitting errors
+e_r <- 'Resource abundance'
+v_r <- 'Resource stochasticity'
+hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
+
+l_grobs <-
+  lapply(
+    list(
+      # mean
+      ggplot(tapir, aes(t_center, mu)) +
+        geom_line(color = pal[1]) +
+        labs(x = NULL, y = e_r),
+      # variance
+      ggplot(tapir, aes(t_center, sigma2)) +
+        geom_line(color = pal[2]) +
+        labs(x = NULL, y = v_r),
+      # 95% home range
+      ggplot(tapir, aes(t_center, hr_est)) +
+        geom_line(color = pal[3]) +
+        labs(x = NULL, y = hr_lab)),
+    as_grob) # convert to grid graphical objects (grobs)
+
+# align left margins of all plots
+aligned_widths <- align_margin(map(l_grobs,
+                                   function(x) {x$widths}), 'first')
+
+# Setting the dimensions of plots to the aligned dimensions
+for (i in seq_along(l_grobs)) {
+  l_grobs[[i]]$widths <- aligned_widths[[i]]
 }
 
-anna_tel <- data.frame(anna_tel) %>%
-  tibble() %>%
-  rename(long = longitude, lat = latitude) %>%
-  mutate(dec_date = decimal_date(timestamp))
-anna_tel <- bind_cols(anna_tel, ndvi_preds(anna_tel))
+# Draw aligned plots
+plot_grid(plotlist = l_grobs, nrow = 1, labels = c('a.', 'b.', 'c.'))
 ```
 
-We also apply the function to the subset of the telemetry data belonging to each of the seven-day periods.
+To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's spatial needs.
+To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.
+However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see figure \@ref(fig:hr-gam)).
 
-```{r, eval=TRUE}
-anna_mw <-
-  mutate(
-    anna_mw,
-    # estimate mean and variance in NDVI
-    preds = map(dataset, \(.d)
-                filter(anna_tel, timestamp %in% .d$timestamp)),
-    # extract the column of mean NDVI and take the average
-    mu = map_dbl(preds, \(.d) mean(.d$mu)),
-    # extract the column of variance in NDVI and take the average
-    sigma2 = map_dbl(preds, \(.d) mean(.d$sigma2))) %>%
-  # only keep necessary coliumns
-  select(t_center, mu, sigma2, hr_lwr_50, hr_est_50, hr_upr_50, hr_lwr_95,
-         hr_est_95, hr_upr_95) %>%
-  # change to long format (one column for both utilization quantiles)
-  pivot_longer(c(hr_lwr_50, hr_est_50, hr_upr_50, hr_lwr_95, hr_est_95,
-                 hr_upr_95), names_to = c('.value', 'quantile'),
-               names_pattern = '(.+)_(.+)') %>%
-  mutate(t_center = as.POSIXct(t_center, origin = '1970-01-01'),
-         quantile = paste0(quantile, '%'),
-         quantile = factor(quantile))
-anna_mw
+```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's spatial needs (on the log link scale) using a GAM with no shape constraints.
+The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
+Dashed lines inticate the 95\\% credible intervals for the temporal terms."}
+gam(hr_est ~ s(mu,  k = 3) + s(sigma2, k = 3), family = Gamma('log'),
+    data = tapir) %>%
+  plot(pages = 1, scheme = 3)
 ```
 
-We now have all the necessary objects to create the final figure:
+\clearpage
 
-```{r}
-# min and max of the tracking dates
-date_labs <- range(anna_tel$timestamp) %>% as.Date()
+Thus, we use shape-constrained splines to inform the model about the shape we expect the terms to have while also imposing monotonicity.
 
-p_track <-
-  ggplot() +
-  coord_equal() +
-  geom_path(aes(long, lat), track, alpha = 0.3) +
-  geom_point(aes(long, lat, color = dec_date), anna_tel) +
-  scale_color_viridis_c(NULL, breaks = range(anna_tel$dec_date),
-                        labels = date_labs) +
-  labs(x = NULL, y = NULL)
-
-p_mu <-
-  ggplot(anna_mw, aes(t_center, mu)) +
-  geom_line(color = pal[1]) +
-  labs(x = NULL, y = '\U1D707(t)')
-
-p_sigma2 <-
-  ggplot(anna_mw, aes(t_center, sigma2)) +
-  geom_line(color = pal[2]) +
-  labs(x = NULL, y = '\U1D70E\U00B2(t)')
-
-p_hr <-
-  ggplot(anna_mw) +
-  geom_line(aes(t_center, hr_est, group = quantile, size = quantile),
-            color = pal[3], show.legend = FALSE) +
-  geom_ribbon(aes(t_center, ymin = `50%`, ymax = `95%`),
-              anna_mw %>%
-                select(t_center, hr_est, quantile) %>%
-                pivot_wider(values_from = hr_est, names_from = quantile),
-              fill = pal[3], color = 'transparent', alpha = 0.2,
-              inherit.aes = FALSE) +
-  labs(x = NULL, y = expression(Home~range~size~(km^2)~'   ')) +
-  scale_size_manual(values = c(1, 0.5))
+```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's spatial needs (on the log link scale) using a shape-constrained GAM.
+The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
+Dashed lines inticate the 95\\% credible intervals for the temporal terms."}
+m <- scam(hr_est ~
+            s(mu, bs = 'mpd', k = 4) + # monotone decreasing p-spline
+            s(sigma2, bs = 'mpi', k = 4), # monotone increasing p-spline
+          family = Gamma('log'),
+          data = tapir)
+plot(m, pages = 1, scheme = 3)
 ```
 
-To ensure all panels are properly aligned, we convert each plot to a grid object (a "grob") and then align each grob.
+Finally, we can create the predictions from the shape-constrained GAM.
 
-```{r}
-# convert to grid graphical objects (grobs)
-grobs <- map(list(p_track, p_mu, p_sigma2, p_hr), as_grob)
+```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
+theme_set(theme_bw()) # need to set the theme again for some reason
 
-# align left margins of all plots
-aligned_widths <- align_margin(map(grobs, function(x) {x$widths}), 'first')
+# predict marginal effects for each term
+preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
+                sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
+  bind_cols(
+    # predictions for the marginal effect of mu
+    predict(m, newdata = ., terms = 's(mu)', type = 'link',
+            se.fit = TRUE) %>%
+      as.data.frame() %>% # convert list to data frame
+      transmute(hr_mu_est = exp(fit), # transform to the response scale
+                hr_mu_lwr = exp(fit - 1.96 * se.fit), # 95% cred. intervals
+                hr_mu_upr = exp(fit + 1.96 * se.fit)),
+    # predictions for the marginal effect of sigma2
+    predict(m, newdata = ., terms = 's(sigma2)', type = 'link',
+            se.fit = TRUE) %>%
+      as.data.frame() %>%
+      transmute(hr_sigma2_est = exp(fit),
+                hr_sigma2_lwr = exp(fit - 1.96 * se.fit),
+                hr_sigma2_upr = exp(fit + 1.96 * se.fit)))
+
+p_d <- ggplot() +
+  coord_cartesian(ylim = c(0, 12.5)) +
+  geom_point(aes(mu, hr_est), tapir, color = pal[3], alpha = 0.5) +
+  geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
+              fill = pal[1], alpha = 0.3) +
+  geom_line(aes(mu, hr_mu_est), preds, color = pal[1]) +
+  xlab(e_r) +
+  scale_y_continuous(hr_lab)
+
+p_e <- ggplot() +
+  coord_cartesian(ylim = c(0, 12.5)) +
+  geom_point(aes(sigma2, hr_est), tapir, color = pal[3], alpha = 0.5) +
+  geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
+              preds, fill = pal[2], alpha = 0.3) +
+  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2]) +
+  xlab(v_r) +
+  scale_y_continuous(hr_lab)
+
+p_f <- ggplot(tapir) +
+  geom_point(aes(mu, sigma2), alpha = 0.5) +
+  labs(x = e_r, y = v_r)
+
+r_grobs <- map(list(p_d, p_e, p_f), as_grob)
+
+# align right margins of all plots
+aligned_widths <- align_margin(map(r_grobs,
+                                   function(x) {x$widths}), 'first')
 
 # Setting the dimensions of plots to the aligned dimensions
-for (i in seq_along(grobs)) {
-  grobs[[i]]$widths <- aligned_widths[[i]]
+for (i in seq_along(r_grobs)) {
+  r_grobs[[i]]$widths <- aligned_widths[[i]]
 }
-# Draw aligned plots
-p_left <- plot_grid(plotlist = grobs, ncol = 1,
-                    labels = paste0(letters, '.'))
+
+plot_grid(plotlist = r_grobs, ncol = 3, labels = c('d.', 'e.', 'f.'))
 ```
 
-## there should be a model for anna instead of `geom_smooth()!
+\clearpage
 
-```{r}
-reg_mu <-
-  ggplot(anna_mw) +
-  coord_cartesian(ylim = c(0, 13)) +
-  geom_point(aes(mu, hr_est, group = quantile, shape = quantile), alpha = 0.3,
-             color = pal[3]) +
-  geom_smooth(aes(mu, hr_est, group = quantile, size = quantile),
-              color = pal[1], se = FALSE, method = 'gam',
-              formula = y ~ x, method.args = list(family = "Gamma")) +
-  scale_x_continuous('Resource abundance') +
-  scale_y_continuous(expression(Home~range~size~(km^2)), expand = c(0, 0)) +
-  scale_size_manual(values = c(1, 0.5)) +
-  theme(legend.position = 'none')
+# References {-}
 
-# variance
-reg_s2 <-
-  ggplot(anna_mw) +
-  coord_cartesian(ylim = c(0, 13)) +
-  geom_point(aes(sigma2, hr_est, group = quantile, shape = quantile),
-             alpha = 0.3, color = pal[3]) +
-  geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
-              color = pal[2], se = FALSE, method = 'gam',
-              formula = y ~ x, method.args = list(family = "Gamma")) +
-  scale_x_continuous('Resource unpredictability') +
-  scale_y_continuous(expression(Home~range~size~(km^2)), expand = c(0, 0)) +
-  scale_size_manual('Quantile', values = c(1, 0.5)) +
-  scale_shape('Quantile') +
-  theme(legend.position = 'none')
-
-leg <-
-  get_legend(
-    ggplot(anna_mw) +
-      geom_smooth(aes(sigma2, hr_est, group = quantile, size = quantile),
-                  se = FALSE, method = 'gam', color = 'black',
-                  formula = y ~ x, method.args = list(family = "Gamma")) +
-      geom_point(aes(sigma2, hr_est, group = quantile, shape = quantile),
-                 color = pal[3]) +
-      scale_size_manual('Utilization quantile', values = c(1, 0.5)) +
-      scale_shape('Utilization quantile') +
-      theme(legend.position = 'top')
-  )
-
-# add a legend on the top 
-p_regs <- plot_grid(leg, reg_mu, reg_s2, ncol = 1,
-                    rel_heights = c(0.1, 1, 1), labels = c('', 'e.', 'f.'))
-
-# group all the plots together
-plot_grid(p_left, p_regs)
-```
-
-```{r, echo=FALSE, eval=TRUE, out.width='100%'}
-knitr::include_graphics('../figures/tapir-example.png')
-```
+\hangparas{1em}{1} <!-- indent all lines but the first -->

### 2023-08-24 analysis/tapirus-terrestris/moving-window-figure-anna-betals.R
cleaned up the code for better efficiency

@@ -18,14 +18,11 @@ hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
 # import location-scale model for predictions of mean and variance in NDVI
 m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
 
-# function to make predictions from the model
+# function to make predictions from the model for each telemetry
 ndvi_preds <- function(.data) {
-  .data <- .data %>%
-    data.frame() %>% # convert telemetry data to a data.frame
-    mutate(year = year(timestamp),
-           doy = yday(timestamp))
+  .data <- data.frame(.data) # convert telemetry data to a data.frame
   
-  predict.gam(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE) %>%
+  predict(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE) %>%
     data.frame() %>%
     tibble() %>%
     transmute(mu = X1,
@@ -33,29 +30,28 @@ ndvi_preds <- function(.data) {
     return()
 }
 
-# import tapir data (https://doi.org/10.1186/s40462-022-00313-w)
+# import tapir data (from https://doi.org/10.1186/s40462-022-00313-w)
 tapir <- readRDS('../tapirs/models/tapirs-final.rds') %>%
   filter(name.short == 'ANNA')
-tel <- tapir$data[[1]]
 
-tel <- data.frame(tel) %>%
+tel <- data.frame(tapir$data[[1]]) %>%
   rename(long = longitude, lat = latitude) %>%
-  mutate(dec_date = decimal_date(timestamp))
-tel <- bind_cols(tel, ndvi_preds(tel))
+  mutate(dec_date = decimal_date(timestamp)) %>%
+  bind_cols(., ndvi_preds(.)) %>% 
+  as_tibble()
 
 tapir <-
   readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
   # mutate(ess = map_dbl(model, \(.m) summary(.m)$DOF['area']),
   #        weight = ess / mean(ess)) %>%
-  mutate(preds = map(dataset, \(.d) filter(tel, timestamp %in% .d$timestamp)),
-         mu = map_dbl(preds, \(.d) mean(.d$mu)),
-         sigma2 = map_dbl(preds, \(.d) mean(.d$sigma2))) %>%
[…]

### 2023-08-25 analysis/simulations/modeling-R-and-hr.R
moved script to analysis folder, added 95% CIs

@@ -51,18 +51,21 @@ plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
 
 # estimate predictions ----
 # effect of mu
-newd_mu <- expand_grid(mu = seq(min(sims$mu), max(sims$mu),
-                                length.out = 400),
-                       sigma2 = mean(sims$sigma2),
-                       quantile = unique(sims$quantile))
+newd_mu <-
+  expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
+              sigma2 = mean(sims$sigma2),
+              quantile = unique(sims$quantile))
 
 preds_mu <-
-  bind_cols(
+  bind_cols( # bind new data and predictions
     newd_mu,
-    predict(m, newdata = newd_mu, type = 'response', se.fit = TRUE) %>%
-      data.frame() %>%
-      transmute(hr = fit.1,
-                var_hr = fit.1 * exp(fit.2)))
+    predict(m, newdata = newd_mu, type = 'link', se.fit = TRUE) %>%
+      data.frame() %>% # convert list to data frame
+      # 95% CIs assuming Gaussian credible intervals on the link scale
+      transmute(hr = exp(fit.1),
+                hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
+                hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
+
 
 # effect of sigma2
 newd_sigma2 <-
@@ -74,10 +77,11 @@ newd_sigma2 <-
 preds_sigma2 <-
   bind_cols(
     newd_sigma2,
-    predict(m, newdata = newd_sigma2, type = 'response', se.fit = TRUE) %>%
+    predict(m, newdata = newd_sigma2, type = 'link', se.fit = TRUE) %>%
       data.frame() %>%
-      transmute(hr = fit.1,
-                var_hr = fit.1 * exp(fit.2)))
+      transmute(hr = exp(fit.1),
+                hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
+                hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
 
 # final figure
 preds <- bind_rows(mutate(preds_mu, x = e_r) %>%
@@ -93,21 +97,24 @@ sims_l <- sims %>%
   mutate(x = if_else(name == 'mu', e_r, v_r))
 
 ggplot() +
-  facet_grid(. ~ x, scales = 'free', switch = 'x') +
[…]

### 2023-08-25 writing/appendix-1-simulations.Rmd
added GAMLS to appendix 1

@@
 library('tidyr')   # for data wrangling (e.g., nested tibbles)
+library('mgcv')    # for empirical Bayes GAMs
 library('ggplot2') # for fancy plots
+library('cowplot') # for fancy multi-panel plots
+theme_set(theme_bw()) # change default theme
 
@@
 source('functions/label_visits.R') # decides when animal encounters food
+source('analysis/figures/default-figure-styling.R') # for color palette
 
@@
 
-if(FALSE) {
-  # check if the ends of each day are correct and make sense
-  days_end <-
-    days %>%
-    group_by(scenario, day) %>%
-    filter(full, ! duplicated(full)) %>% # take 1st row where full
-    rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
-    # remove unneded columns (also avoids duplicated colnames with tracks)
-    dplyr::select(-c(x, y, vx, vy, longitude, latitude, food))
-
-  # check max fraction of time used (should be < 1)
-  max(days_end$t_expl) / MAX_T
-
-  # are animals are full only once/day? (should be == 1)
-  sum(days_end$full) / (max(days$day) * 2)
-
-  # plot of satiety over time by animal
-  ggplot(days, aes(t, satiety, group = day)) +
-    facet_wrap(~ scenario) +
-    geom_line(alpha = 0.05) +
-    geom_point(aes(t_expl), days_end, alpha = 0.1) +
-    geom_hline(yintercept = REQUIRED, color = 'red') +
-    geom_vline(xintercept = MAX_T, color = 'blue')
-
-  # check distribution of animals
-  ggplot(days_end, aes(scenario, t_expl)) +
-    geom_hline(yintercept = MAX_T, color = 'red') +
-    geom_violin(fill = 'forestgreen', alpha = 0.3) +
-    geom_boxplot(fill = NA) +
-    labs(x = '', y = 'Exploration time')
-
-  # check home ranges of animals
-  ggplot(days) +
-    facet_grid(. ~ scenario) +
-    coord_equal() +
-    geom_hex(aes(longitude, latitude)) +
-    scale_fill_distiller('Count', type = 'seq', na.value = 'transparent') +
-    theme(legend.position = 'top')
-}
-
 # single estimates that eventually converge to the asymptote ----
@@
   ungroup() # remove grouping by scenario
-
-if(FALSE) {
-  # check times are adding up correctly
-  # best case should require less time
-  # blue and red lines should have same length (within the pair)
-  # black lines should be horizontal
-  days_summarized %>%
-    filter(day <= 10) %>%
-    ggplot(aes(day, timestamp)) +
-    facet_wrap(~ scenario, scales = 'free_y') +
-    geom_line() +
-    geom_line(aes(group = day), color = 'blue', lwd = 30) +
-    geom_line(aes(day, timestamp + t_expl, group = day), color = 'red',
-              lwd = 30)
-
-  days_summarized %>%
-    filter(day <= 10) %>%
-    ggplot() +
-    facet_wrap(~ scenario) +
-    coord_equal() +
-    geom_path(aes(x, y, group = day), alpha = 0.5) +
-    geom_point(aes(0, 0)) +
-    geom_point(aes(x, y), filter(days_summarized, day <= 10, t == 0),
-               color = 'red')
-}
 
@@
 
+# Modeling the results from the simulations
+
+This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated spatial needs.
+Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity.
+The home range was caulculated using a Gaussian approximation using the simulated organism's positional variance:
+
+$$\widehat{h_q} = -2 \log(1 - q) \pi \sigma^2_{pos},$$
+
+where $widehat{h_q}$ is the estimated home range for the $q^{\text{th}}$ quantile and $\sigma^2_{pos}$ is the positional variance.
+Consequently, the estimated core home range estimate is simply a multiple of the 95% quantile:
+
+$$\widehat{h_{0.95}} / \widehat{h_{0.5}} = \frac{-2 \log(1 - 0.95) \pi \sigma^2_{pos}}{-2 \log(1 - 0.5) \pi \sigma^2_{pos}} = \frac{\log(1 - 0.95)}{\log(1 - 0.5)} = \frac{\log(0.05)}{\log(0.5)} = \log_{0.5}(0.05) \approx 4.322,$$
+
+and therefore $\e(R)$ and $\var(R)$ would have similar effects (but about 4.322 times weaker) on the core home range than they do on the 95% home range.
+Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
+When estimating spatial needs based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 2 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+
+```{r, eval=TRUE}
+# switch data to long format with a column of quantile
+sims <- readRDS('simulations/days-hrs.rds') %>%
+  tibble() %>%
+  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr')%>%
+  mutate(quantile = factor(quantile)) # necessary for GAMs
+```
+
+The sampling of the data is not uniform:
+
+```{r, eval=TRUE}
+ggplot(sims, aes(mu, sigma2)) +
+  geom_point(alpha = 0.3)
+```
+
+But it is not particularly problematic:
+
+```{r, eval=TRUE}
+ggplot(sims, aes(mu, sigma2)) +
+  geom_hex(bins = 20) +
+  scale_fill_viridis_c('Count', limits = c(0, NA), option = 'A',
+                       direction = -1) +
+  theme_bw() # Rmd only keeps theme_bw as default for a single plot
+```
+
+The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown).
+To estimate the effects of resource abundance and stochasticity on the simulated spatial needs, we fit 3 location-scale GAMs (GAMLSs) with a location-scale gamma distribution (since home ranges are strictly positive) of increasing complexity.
+We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones.
+The three models are detailed below and compared using the Akaike Information Criterion and the Bayesian Information Criterion.
+Both information criteria demonstrate that the third model, which accounts for the interaction effects of $\e(R)$ and $\var(R)$, is the best model.
+
+```{r, eval=TRUE}
+# marginal effects of mu and sigma only (linear on the log link scale)
+m_1 <- gam(
+  list(
+    # predictor for the mean
+    hr ~ quantile + mu + sigma2,
+    # predictor for the scale (variance = scale * mean)
+    ~ quantile + mu + sigma2),
+  family = gammals(),
+  data = sims)
+
+# marginal effects of mu and sigma only (nonlinear on the log link scale)
+m_2 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2),
+    ~ quantile + s(mu) + s(sigma2)),
+  family = gammals(),
+  data = sims)
+
+# marginal and interaction effects (nonlinear on the log link scale)
+m_3 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5),
+    ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5)),
+  family = gammals(),
+  data = sims)
+
+AIC(m_1, m_2, m_3) # smooth model with interactions is best
+BIC(m_1, m_2, m_3) # smooth model with interactions is best
+
+m <- m_3 # select the best model
+summary(m)
+```
+
+Figure \@ref(fig:gamls-terms) shows each of the terms from model 3.
+The estimated effect of `quantile` (see the output of `summary()` above) supports the relationship between the gaussian HR estimates detailed above, since $\exp(1.464) = 4.323218 \approx \log_{0.5}(0.05),$ but there is no differences in the scale parameter between the two quantiles.
+Additionally, the interaction term of $\mu(t)$ and $\sigma^2(t)$ on the mean HR indicates that the effect of $\sigma^2(t)$ is stronger at lower values of $\mu(t)$ (e.g., $\mu(t) = 100$), since the estimated effect is negative for low values of $\sigma^2(t)$ and positive for high values of $\sigma^2(t)$, which accentuates the effect of $\sigma^2(t)$ on the mean home range.
+The opposite is true for high values of $\mu(t)$, where the effect of $\sigma^2(t)$ becomes flatter, and thus weaker.
+Similar but opposite conclusions can be drawn from the interaction term for the scale parameter, since the effect of $\sigma^2(t)$ on the scale parameter is weaker when $\mu(t)$ is low and stronger when $\mu(t)$ is high.
+
+```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales).
+The dashed lines indicate the 95\\% Bayesian credible intervals.
+The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=16, fig.width=8, out.width='65%', echo=FALSE}
+# using custom layout to make figure easier to read
+layout(matrix(c(7, 1, 2, 3, 8, 4, 5, 6), nrow = 4, byrow = TRUE))
+plot(m, all.terms = TRUE, scheme = 3, scale = 0)
+```
+
+Finally, we can use the model to recreate figure \@ref(fig:5-5-reg) and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
+
+```{r, eval=TRUE}
+# effect of mu
+newd_mu <-
+  expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
+              sigma2 = mean(sims$sigma2),
+              quantile = unique(sims$quantile))
+
+preds_mu <- bind_cols( # bind new data and predictions
+  newd_mu,
+  predict(m, newdata = newd_mu, type = 'link', se.fit = TRUE) %>%
+    data.frame() %>% # convert list to data frame
+    # 95% CIs assuming Gaussian credible intervals on the link scale
+    transmute(hr = exp(fit.1),
+              hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
+              hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
+
+# effect of sigma2
+newd_sigma2 <-
+  expand_grid(mu = mean(sims$mu),
+              sigma2 = seq(min(sims$sigma2), max(sims$sigma2),
+                           length.out = 400),
+              quantile = unique(sims$quantile))
+
+preds_sigma2 <- bind_cols(
+  newd_sigma2,
+  predict(m, newdata = newd_sigma2, type = 'link', se.fit = TRUE) %>%
+    data.frame() %>%
+    transmute(hr = exp(fit.1),
+              hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
+              hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
+
+# final figure
+e_r <- 'Resource abundance'
+v_r <- 'Resource stochasticity'
+
+preds <- bind_rows(mutate(preds_mu, x = e_r) %>%
+                     rename(value = mu) %>%
+                     select(-sigma2),
+                   mutate(preds_sigma2, x = v_r) %>%
+                     rename(value = sigma2) %>%
+                     select(-mu))
+
+sims_l <- sims %>%
+  mutate(e = hr - predict(m, newdata = sims, type = 'response')[, 1]) %>%
+  pivot_longer(c(mu, sigma2)) %>%
+  mutate(x = if_else(name == 'mu', e_r, v_r))
+
+ggplot() +
+  facet_grid(. ~ x, scales = 'free', switch = 'x') + # label below x axis
+  geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
+             alpha = 0.1, color = pal[3]) +
+  # 95% CIs are present but hidden under the estimates
+  geom_ribbon(aes(value, ymin = hr_lwr, ymax = hr_upr, fill = x),
+              filter(preds, quantile == 'hr_95'), alpha = 0.5) +
+  geom_line(aes(value, hr, color = x),
+            filter(preds, quantile == 'hr_95')) +
+  scale_color_manual(values = pal) +
+  scale_x_continuous(NULL, breaks = NULL) + 
+  scale_y_continuous('Home range size', breaks = NULL) +
+  theme_bw() +
+  # to make facet strip look like the x axis
+  theme(legend.position = 'none',
+        strip.background = element_blank(),
+        strip.text = element_text(size = 12))
+```
+
 \clearpage

### 2023-08-25 writing/manuscript.Rmd
added 95% CIs to the simulation regression

@@
 
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% credible intervals for the mean (overlapping the line).
+Credible intervals were calculated assuming a gaussian posterior distribution on the link scale.
 The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
@@
 We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
+In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Chris Fleming, Lauren Mills, and Sarah Wyse.
 

### 2023-09-08 writing/appendix-1-concepts-and-definitions.Rmd
created appendix 1: concepts and definitions

--- 2023-08-06 writing/manuscript.Rmd
@@
 title: \large How resource abundance and stochasticity affect animals' spatial needs
+subtitle: "Appendix 1: Concepts and definitions"
 author:
@@
     institute: uofe
-  # - name: Adam T. Ford
-  #   institute: biol
-  #   institute: braes
-  # - name: Jacob R. Goheen
-  #   institute: wyom
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
   - name: E. Patrícia Medici
@@
   - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
-  # - wyom: Department of Zoology and Physiology, University of Wyoming, Laramie, Wyoming, United States of America.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
@@
     - \usepackage{setspace}\doublespacing # for double-spaced text
-    - \usepackage{indentfirst} # for indenting first line of each paragraph
     - \usepackage[small]{titlesec} # for smaller font for headings
-    - \usepackage{wrapfig} # for wrapping text around figures
     - \usepackage{caption} # for more customization of captions
-    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
-    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
+    - \captionsetup[figure]{font=scriptsize} # smaller caption text
     - \usepackage{hanging} # for hanging indents in references
 subparagraph: true # needed for \usepackage[small]{titlesec}
-linenumbers: true
 output:
   bookdown::pdf_document2:
-    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
+    number_sections: false
+    toc: false
     pandoc_args:
-      - '--lua-filter=scholarly-metadata.lua'
-      - '--lua-filter=author-info-blocks.lua'
-    toc: false
-    number_sections: false
-editor_options: 
-  chunk_output_type: console
+      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
+      - '--lua-filter=author-info-blocks.lua' # for author affiliations
 ---
 
-<!-- define custom math symbols -->
+\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
+\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->
 
-\newcommand*\e{\mathbb E} <!-- expectation (blackboard bold) E -->
-
-\newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->
-
-<!-- place table of contents on a new page -->
 \newpage
-
-\singlespacing
 
@@
 
-<!-- start main body on a new page -->
 \newpage
 
-\doublespacing
+\noindent In this appendix, we presents the foundational concepts and terms used in the main manuscript.
+We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.
 
-```{r setup, include=FALSE}
-# set chunk defaults: don't print code, images 75% of page width and centered
-knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
-```
-
-```{r wrap-figure-function, include = FALSE}
-# this chunk is needed to wrap text around or beside figures
-# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures
-
-defOut <- knitr::knit_hooks$get('plot')  # save the default plot hook
-
-knitr::knit_hooks$set(plot = function(x, options) {  # set the new plot hook ...
-
-  x <- defOut(x, options)  # first apply the default hook
-
-  # if wrapfigure options are given, create the new opening string for the wrapfigure
-  # environment using custom figure alignment and text width
-  if(!is.null(options$wrapfigure)) {
-
-    # options must be numeric, they cannot be characters
-    wf <- sprintf('\\begin{wrapfigure}{%s}{%g\\textwidth}',
-                  options$wrapfigure[[1]], # figure alignment
-                  options$wrapfigure[[2]]) # caption width relative to text width
-    x  <- gsub(pattern = '\\begin{figure}', # and replace the default one with it
-               replacement = wf,
-               x = x,
-               fixed = TRUE) # pattern is a string, so evaluate as is
-    x  <- gsub(pattern = '{figure}', # also replace the environment ending
-               replacement = '{wrapfigure}',
-               x = x,
-               fixed = TRUE)
-  }
-  return(x)
-}
-)
-
-# chuck options used to wrap figures:
-## fig.align = "right":          align figures right
-## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
-## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width
-
-#' place figures before the paragraph they are referred to so `wrapfigure` places them
-#' beside the paragraph
-```
-
-# Abstract
-
-\noindent The amount of space animals use is thought to be tightly linked to the availability of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
-This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown.
-This is not a trivial oversight.
-Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs, challenging ecological theory and applied conservation alike.
-In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
-We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use.
-Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado <!--an impala (*Aepyceros melampus*), and multiple moose (*Alces alces*)--> to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
-Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
-
-\newpage
-
-# Introduction
-
-\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction.
-Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018].
-Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
-
-The relationship between animal movement and resource abundance has been of interest to biologists for over half a century.
-@burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range.
-Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how animals decide where to live and when to reproduce.
-Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home-range size is
-
-\begin{equation}
-H = C / R,
-(\#eq:harestad-eq)
-\end{equation}
-
-\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** the animal can access.
-Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
-While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
-Instead, results are reported as independent, novel findings.
-Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs.
-In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
-
-In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
-We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use.
-For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
-In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for.
-Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
-
-Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources.
-First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on animals' spatial needs while suggesting a second unifying hypothesis.
-Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
-Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
-Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.
-
-# Concepts and definitions
-
-<!--\noindent In this section, we presents the foundational concepts and terms used in this manuscript.
-We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.-->
-
-## Resources as a random variable
+# Resources as a random variable
 
@@
 Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Resource abundance is often unpredictable (and difficult to quantify), since it depends on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
-Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
+Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
+Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time.
 Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
 
-### Probability distributions
+## Probability distributions
 
@@
 
-### Expected resource abundance, $\e(R)$
+## Expected resources, $\e(R)$
 
@@
 The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term.
-Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$.
+Here we use $\e(R)$ to indicate the expectation of the random variable $R$.
 When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$.
-We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
+We indicate the *estimated* average amount of resources over time by adding a caret on the symbol: $\hat \mu(t)$.
+Note, however, that the estimate of $\mu(t)$ and what values it can take depend on what distribution we assume $R$ follows.
+For a Gamma distribution, $\mu(t)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t)$ is necessarily within the interval $(-1, 1)$.
+See the section on modeling $R$ in the main manuscript for more information on how to choose probability distributions.
 
-### Variance in resource abundance, $\var(R)$
+## Variance in resources, $\var(R)$
 
@@
 We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$).
-Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
+<!-- Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others due to changes in $\sigma^2(t)$.
 Additionally, we distinguish between two different sources of variation.
 We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
-In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$.
-For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley.
-In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.
+In contrast, we use the term **stochasticity** for unpredictable variation (i.e., due to $\sigma^2(t)$).--> For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons.
+Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t)$ to be lower in the winter since it does not expect to find berries (i.e., $\mu(t) \approx 0$).
+Also note that, like with $\mu(t)$, the estimate of $\sigma^2(t)$ and what values it can take depend on what distribution we assume $R$ follows.
 
-### Simulating resource abundance
+Below are four examples of Gamma probability distributions with different means and variances:
 
-\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
-Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
-However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa).
-As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
-This assumption also holds biologically; in the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
-
-# Effects of resource abundance, $\e(R)$
-
-```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
-Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
-```
-
-```{r, eval=FALSE, echo=FALSE}
+```{r, eval=TRUE, echo=FALSE, fig.height=5}
 # to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
@@
 plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
+     xlab = 'r', ylab = 'Density',
+     main = '\U03BC = 1, \U03C3\U00B2 = 0.5')
 plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
+     xlab = 'r', ylab = 'Density',
+     main = '\U03BC = 5, \U03C3\U00B2 = 0.5')
 plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
+     xlab = 'r', ylab = 'Density',
+     main = '\U03BC = 1, \U03C3\U00B2 = 5')
 plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
+     xlab = 'r', ylab = 'Density',
+     main = '\U03BC = 5, \U03C3\U00B2 = 5')
 layout(1)
 ```
-
-\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
-All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
-The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
-
-When $\e(R)$ is high, we expect animals' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
-For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
-
-It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
-Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).
-
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
-(a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
-```
-
-Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
-Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
-However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
-
-This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
-Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
-Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
-Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
-
-# Effects of resource stochasticity, $\var(R)$
-
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
-For example, since berries are scarce outside the growing season, the variance in berries will also be low.
-But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location only after competitors ate all the berries.
-In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
-
-```{r s2-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource stochasticity ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low spatial needs as animals are able to depend on somewhat predictable resources.
-As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
-```
-
-Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
-While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
-
-```{r, eval=FALSE, echo=FALSE}
-# the median decreases as V(R) increases, even if E(R) is constant
-source('functions/qgamma2.R')
-plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
-```
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
-
-Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
-
-Like the scenario illustrated in figure \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
-We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)).
-When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
-Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
-Biologically, this is important because some animals may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
-An animal's perception of changes in $R$ will depend strongly on the animal's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
-Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
-
-# Interactive effects of $\e(R)$ and $\var(R)$
-
-\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' spatial needs, but we presented the two parameters as independent drivers of movement.
-However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, animals may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on animal's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
-A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
-Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
-
-<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
-However, the effects of $\var(R)$ on animals' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
--->
-
-@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
-However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality).
-Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
-A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
-Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on animals' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
-Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
-The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
-
-@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
-Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
-However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
-Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
-Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.
-
-## Simulating animal responses to $\e(R)$ and $\var(R)$
-
-To illustrate how $\e(R)$ and $\var(R)$ affect animals' spatial needs as well as the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations in figure \@ref(fig:5-5-sims) (for methodological details, see Appendix 1).
-The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the animal.
-As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
-
-```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
-Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022).
-$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
-Additional information is provided in Appendix 1.", out.width='100%'}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
-```
-
-In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
-Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
-In cases where $\mu(t)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
-
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
-Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
-This is because, similarly to resource-poor periods, times of greater stochasticity require the animal to move over larger areas for longer periods of time.
-This results in a greater uncertainty in how much time and space the animal will require to reach satiety, or indeed whether an animal living in highly stochastic environments can even reach satiety within a finite amount of time.
-
-The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
-Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
-When both $\e(R)$ and $\var(R)$ increase linearly, $H$ initially increases since the effect of $\var(R)$ is stronger, but then decreases as the effect of $\e(R)$ begins to dominate.
-However, when $\mu(t)$ and $\sigma^2(t)$ follow fairly different trends (e.g., cyclical $\e(R)$ and linear $\var(R)$), it is easy to see how an increase in $\var(R)$ amplifies the effects of $\e(R)$.
-In contrast, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
-Thus, $H$ is largest when $\e(R)$ is low and $\var(R)$ is high, and it is smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
-
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
-$\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
-Although it is impossible for $\mu(t)$ and $\sigma^2(t)$ to increase linearly forever, these examples are useful to demonstrate that linear changes in $\mu(t)$ and $\sigma^2(t)$ affect $H$ nonlinearly.
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
-Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
-However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
-
-Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
-Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
-We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
-Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
-Secondly, the animal's movement was simulated to be range resident within an homogeneous environment.
-Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood.
-Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
-Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
-While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.
-
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
-The relationships were estimated using a Generalized Additive Model for Location and Scale (Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma family of distributions.
-The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both $\\e(H)$ and $\\var(H)$.
-Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in spatial needs increases with mean spatial needs.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-plots.png')
-```
-
-# Applying the framework
-
-\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011].
-To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
-Appendix 2 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
-
-The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
-We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
-In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as new findings.
-Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models (HGAMs) with individual-level intercepts and smooth terms [see models "GS" and "GI" in @pedersen_hierarchical_2019].
-Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
-Individuals and populations that were once thought to have different spatial needs due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
-In the following sections, we illustrate how one can apply this framework to empirical data.
-
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
-Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals' spatial needs under different scenarios.
-
-### Modeling $R$
-
-\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
-its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).
-
-The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
-However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].
-
-The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->.
-The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009].
-In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
-If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
-
-Modeling the mean and variance terms of $R$ should be done carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be spatiotemporally nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms.
-An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
-Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
-Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
-We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model.
-@simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.
-
-### Estimating $R$ using NDVI
-
-\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the **linear transformation**
-
-\begin{equation}
-\nu_{[0, 1]} = \frac{\nu + 1}{2},
-(\#eq:ndvi-to-01)
-\end{equation}
-
-\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
-<!--Note that only **linear transformations** (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
-We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
-
-\begin{equation}
-\e(R) = \frac{\alpha}{\alpha + \beta}
-(\#eq:beta-mean)
-\end{equation}
-
-\noindent while the variance is
-
-\begin{equation}
-\var(R) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
-(\#eq:beta-var)
-\end{equation}
-
-\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):
-
-\begin{equation}
-\nu = 2 \nu_{[0, 1]} - 1.
-(\#eq:01-to-ndvi)
-\end{equation}
-
-\noindent Consequently, we have
-
-\begin{equation}
-\e(\nu|t) = 2 \mu(t) - 1.
-(\#eq:mu-to-e-ndvi)
-\end{equation}
-
-\noindent and
-
-\begin{equation}
-\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
-(\#eq:sigma2-to-v-ndvi)
-\end{equation}
-
-\noindent Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
-Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
-
-<!--
-## Animal movement as a stochastic process
-
-### Drivers of animal movement
-
-\noindent @nathan_movement_2008 suggested we can generalize animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
-
-\begin{equation}
-\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
-(\#eq:nathan-movement)
-\end{equation}
-
-\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
-This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on animals' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
-
-### Visualizing movement as a stochastic process
-
-\noindent To understand how movement can be viewed as a **stochastic process**, imagine an animal that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
-If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another.
-However, we can use the animal's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
-This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
-Similarly, we can define $\tau_v$ to be the amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
-$\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
-
-By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data.
-While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].
-
-Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
-However, since the time required for an animal to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
-Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
-
-Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less on *why* it moved.
-Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
-Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in animal movement while focusing on the animals' movement behaviors rather than the causes.
-
-### Quantifying spatial needs
-
-\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 50% and 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
-Note, however, that although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's movement but not its behavior.
-Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
-Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
--->
-
-<!-- ### Modeling spatial needs -->
-
-<!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
-However, most commonly used movement models assume animals' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
--->
-
-<!-- To quantify an animal's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
-The width of the window, $w$, should be a fixed value greater than two or three times the animal's expected range crossing time [@fleming_overcoming_2019].
-If the animal's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
-Alternatively, if the animal's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
--->
-
-<!-- Modeling animal movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
-In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022].
--->
-
-### Modeling the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs
-
-\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate family of distributions.
-Since $H$ is strictly positive and continuous, a Gamma family seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017], but the Tweedie family would also be appropriate if the variance does not scale with the squared mean.
-
-As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
-If there is not sufficient data to fit a location-scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
-Due to potential correlation between $\e(R)$ and $\var(R)$ in a small and autocorrelated sample, the relationship between $R$ and $H$ should be modeled carefully and with parsimony.
-To appropriately constrain the shape of the effects of $\e(R)$ and $\var(R)$, we suggest using a GLM or a shape-constrained GAM (as long as the wiggliness of each term is chosen carefully).
-Here, we model the effects of $\e(R)$ and $\var(R)$ on $H$ using a shape-constrained GAM with via the `scam` package [@pya_scam_2022].
-Additional information on the model can be found in Appendix 2.
-
-### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's spatial needs
-
-```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
-(a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
-(b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir.
-(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
-(d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
-(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
-Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
-See Appendix 2 for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
-knitr::include_graphics('../figures/tapir-example-with-data.png')
-```
-
-\noindent Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
-Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period.
-Panel **c** shows the changes in the tapir's space use over time.
-Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018).
-In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (January 2018).
-Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
-Thus, while the time series shown in panel **c** may appear complicated, the tapir's response can be well described by two simple functions.
-
-# Discussion
-
-<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as new findings.
--->
-
-## Animal adaptations to changes in $\e(R)$ and $\var(R)$
-
-\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
-Events that occur at the scale of the organism [i.e.
-approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019].
-Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them.
-Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006].
-Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-<!-- 
-with species names:
-Animals may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006].
-Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
-However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
-
-In contrast, individuals with short lifespans are less likely to experience stochastic events and develop memory about them.
-However, the short generation time and high reproductive rate of r-selected species [as opposed to k-selected species, see @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events.
-Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021].
-Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022].
-Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity.
-Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-
-A population's fitness and chances of survival in a changing environment depend strongly on its adaptability and responsiveness to change [@riotte-lambert_environmental_2020], and its ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on its members' chances of survival and reproduction.
-For example, ungulates' ability to time their yearly migrations on available forage biomass, snow fall, snow melt, and hunting pressure allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time and energy spent exploring.
-Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences in $\mu(t)$ between years [@jonzen_rapid_2006].
-However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources during unpredictable times [whooping cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
-Instead, if anomalies occur repeatedly over time, animals may adapt to a higher $\sigma^2(t)$ or shift their ranges altogether [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
-Still, while range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021], they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [e.g., @wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range, since it is determined by their food sources' and associates' ability to move and adapt, too.
-
-## Applications for conservation
-
-\noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on animal behavior.
-In these times of great change and climatic uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021], it is increasingly important to recognize how environmental stochasticity affects animal behavior and shapes ecosystems.
-Accounting for changes in both resource abundance and resource stochasticity is crucial when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
-Protecting areas with high resource abundance and low resource stochasticity will reduce animal's spatial requirements.
-
-The designation of high-quality habitats is a central concern of many conservation initiatives.
-This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat.
-With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas.
-In light of the UN's objective to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
-
-Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
-Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' spatial needs, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
-This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in spatial needs.
-
-It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels.
-While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity.
-In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants.
-Consequently, the abundance of primary consumers may correlate with productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019].
-Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay.
-Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
-However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less stochastic resources.
-
-# Conclusion
-
-\noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance and stochasticity.
-We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
-We demonstrate that animals' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving window approach that allows one to model animal movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism.
-The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currently need and will need in the future under different climate change scenarios.
-
-# Acknowledgements
-
-We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
-
-\newpage
-
-\clearpage <!-- to prevent figures from floating to the references -->
 

### 2023-09-08 writing/appendix-2-simulations.Rmd
renamed sims and modeling appendices

--- 2023-08-25 writing/appendix-1-simulations.Rmd
@@
 title: \large How resource abundance and stochasticity affect animals' spatial needs
-subtitle: "Appendix 1: Simulations"
+subtitle: "Appendix 2: Simulations"
 author:
@@
     - \usepackage{caption} # for more customization of captions
-    # smaller caption text, less space above and below caption
-    - \captionsetup[figure]{font=scriptsize}
+    - \captionsetup[figure]{font=scriptsize} # smaller caption text
     - \usepackage{hanging} # for hanging indents in references
@@
   bookdown::pdf_document2:
+    toc: false
     pandoc_args:
@@
       - '--lua-filter=author-info-blocks.lua' # for author affiliations
-    toc: false
 ---
@@
 
+# set default chunk options
 knitr::opts_chunk$set(eval = FALSE,         # do not run R code
@@
 <!-- add custom commands for formulae -->
-
-\newcommand{\e}{\mathbb E}
-\newcommand{\var}{\mathbb V}
+\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
+\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->
 
@@
 To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
-Instead, the R Markdown document (`writing/appendix-1-simulations.Rmd`) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file.
+Instead, the R Markdown document (`writing/appendix-2-simulations.Rmd`) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file.
 Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes and as a general guide.
@@
 
-```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
+```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated home range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (and homogenous over space).
 Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022).
-$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.", out.width='100%'}
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/mean-variance-5-by-5-hr-sims.png',
@@
 ```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' spatial needs were simulated.
-(a.) Animal tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU; model, an infinitely diffusive and continous-velocity movement mode), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
+(a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
 (b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$.
@@
 
-```{r simulate-tracks, eval=TRUE, warning=FALSE}
+```{r simulate-tracks, warning=FALSE}
 # NOTE: change working directory to "hr-resource-stoch" or modify paths
-setwd('H:/GitHub/hr-resource-stoch')
 
@@
 # generate simulated tracks (will be truncated at satiety later)
-tracks <- tibble(day = 1:N_DAYS, # a simulation for each day
+tels <- tibble(day = 1:N_DAYS, # a simulation for each day
                tel = map(.x = day, # set a seed for consistent results
                          .f = get_tracks)) # function to generate tracks
-tracks
+tels
+```
 
+```{r, eval=TRUE, echo=FALSE}
+# attach all necessary packages
+library('ctmm')    # for generating movement models and movement modeling
+library('terra')   # for working with raster data
+library('dplyr')   # for data wrangling
+library('purrr')   # for functional programming
+library('tidyr')   # for data wrangling (e.g., nested tibbles)
+library('mgcv')    # for empirical Bayes GAMs
+library('ggplot2') # for fancy plots
+library('cowplot') # for fancy multi-panel plots
+theme_set(theme_bw()) # change default theme
+
+# source custom functions
+# source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
+# source('analysis/mean-variance-trends-panel-data.R') # mu and sigma2
+# source('analysis/simulations/movement-model.R') # for consistency
+# source('functions/get_hr.R') # for extracting gaussian home range
+# source('functions/label_visits.R') # decides when animal encounters food
+source('../analysis/figures/default-figure-styling.R') # for color palette
+
+readRDS('H:/GitHub/hr-resource-stoch//simulations/tracks.rds')
+```
+
+```{r warning=FALSE}
 # find patch visits and calories consumed from the tracks
-tracks <-
-  transmute(tracks, # drop tel column
+tracks <- transmute(tels, # drop tel column
                     day, # keep day column
@@
                                 }))
+tracks <- tidyr::unnest(tracks, track) # make a single, large tibble
+tracks
+```
 
-# make a single, large tibble
-tracks <- tidyr::unnest(tracks, track)
-tracks
+```{r, eval=TRUE, echo=FALSE}
+readRDS('H:/GitHub/hr-resource-stoch//simulations/labelled-tracks.rds')
 ```
@@
 Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
-When estimating spatial needs based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 2 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+When estimating spatial needs based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 3 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
@@
 # switch data to long format with a column of quantile
-sims <- readRDS('simulations/days-hrs.rds') %>%
+sims <- readRDS('H:/GitHub/hr-resource-stoch//simulations/days-hrs.rds') %>%
   tibble() %>%
-  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr')%>%
+  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr') %>%
   mutate(quantile = factor(quantile)) # necessary for GAMs
@@
 
-```{r, eval=TRUE}
+```{r, eval=TRUE, fig.height=4}
 ggplot(sims, aes(mu, sigma2)) +
-  geom_point(alpha = 0.3)
+  geom_point(alpha = 0.3) +
+  theme_bw() # Rmd only keeps theme_bw as default for a single plot
 ```
@@
                        direction = -1) +
-  theme_bw() # Rmd only keeps theme_bw as default for a single plot
+  theme_bw()
 ```
@@
 AIC(m_1, m_2, m_3) # smooth model with interactions is best
+
 BIC(m_1, m_2, m_3) # smooth model with interactions is best

### 2023-09-08 writing/appendix-3-empirical-modeling.Rmd
renamed sims and modeling appendices

--- 2023-08-24 writing/appendix-2-empirical-modeling.Rmd
@@
 title: \large How resource abundance and stochasticity affect animals' spatial needs
-subtitle: "Appendix 2: Empirical modeling"
+subtitle: "Appendix 3: Empirical modeling"
 author:
@@
     - \usepackage{caption} # for more customization of captions
-  # smaller caption text, less space above and below caption
-    - \captionsetup[figure]{font=scriptsize}
+    - \captionsetup[figure]{font=scriptsize} # smaller caption text
     -  \usepackage{hanging} # for hanging indents in references
@@
 Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
-See Appendix 2 for additional information.
+See Appendix 3 for additional information.
 The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
-knitr::include_graphics('../figures/tapir-example-with-data.png')
+knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/tapir-example-with-data.png',
+                        rel_path = FALSE)
 ```
@@
 theme_set(theme_bw()) # change default theme
+getwd()
 
@@
 
-```{r, echo=FALSE, eval=TRUE, fig.height=4}
+```{r, echo=FALSE, eval=TRUE, fig.height=2}
 ggplot(anna_ndvi, aes(ndvi)) +
@@
   labs(x = 'NDVI', y = 'Count') +
-  xlim(c(-1, 1))
+  xlim(c(-1, 1)) +
+  theme_bw()
 ```

### 2023-09-08 writing/manuscript.Rmd
updated based on Chris's comments

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' spatial needs
+title: \large How resource abundance and stochasticity affect organisms' spatial needs
 author:
@@
     - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
-    - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
+    # - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
     - \usepackage{hanging} # for hanging indents in references
@@
 <!-- define custom math symbols -->
-
-\newcommand*\e{\mathbb E} <!-- expectation (blackboard bold) E -->
-
-\newcommand*\var{\mathbb V} <!-- variance (blackboard bold) V -->
+\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
+\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->
 
@@
 
-#' place figures before the paragraph they are referred to so `wrapfigure` places them
-#' beside the paragraph
+#' place figures before the paragraph they are referred to so `wrapfigure`
+#' (generally) places them beside the paragraph
 ```
@@
 
-\noindent The amount of space animals use is thought to be tightly linked to the availability of resources within their habitats, such that animals living in productive habitats generally require less space than those in resource-poor habitats.
+\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
 This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown.
 This is not a trivial oversight.
-Animals adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of animals' spatial needs, challenging ecological theory and applied conservation alike.
+Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' spatial needs, challenging ecological theory and applied conservation alike.
 In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
-We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on animals' space use.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in animal space use.
-Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado <!--an impala (*Aepyceros melampus*), and multiple moose (*Alces alces*)--> to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
+We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use.
+Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use.
+Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
 Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
@@
 
-The relationship between animal movement and resource abundance has been of interest to biologists for over half a century.
-@burt_territoriality_1943 considered the search for food as the primary driver for animal movement within an animal's home range.
-Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how animals decide where to live and when to reproduce.
-Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an animal's home-range size is
+The relationship between organisms' movement and resource abundance has been of interest to biologists for over half a century.
+@burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
+Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce.
+Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is
 
@@
 
-\noindent where $H$ is the animal's **home-range size**, $C$ is the animal's resource consumption, and $R$ is the **resources** the animal can access.
+\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption, and $R$ is the **resources** the organism can access.
 Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
-While many researchers have since demonstrated that animals adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
+While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
 Instead, results are reported as independent, novel findings.
-Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on animals' spatial needs.
-In addition, while much work has been done on estimating animals' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
+Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' spatial needs.
+In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
 
 In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
-We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an animal might use.
+We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
 For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
@@
 Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for.
-Although it is generally expected that resource unpredictability will decrease animals' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-Here, we illustrate how animals' spatial needs depend on both the abundance and unpredictability of resources.
+Here, we illustrate how organisms' spatial needs depend on both the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource *abundance* on animals' spatial needs while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on animals' spatial needs while suggesting a second unifying hypothesis.
+Next, we provide a review of the effects of resource *abundance* on organisms' spatial needs while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' spatial needs while suggesting a second unifying hypothesis.
 Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
@@
 
-# Concepts and definitions
+# Simulating resources as a random variable
 
-<!--\noindent In this section, we presents the foundational concepts and terms used in this manuscript.
-We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.-->
-
-## Resources as a random variable
-
-\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
-Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
-Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Resource abundance is often unpredictable (and difficult to quantify), since it depends on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
-Thus, we can let the random variable $R$ indicate the resource abundance at a given point in space and time.
-Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
-
-### Probability distributions
-
-\noindent Random variables are defined by specifying the **distribution** the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Bernoulli).
-Since the variable is random, it can take multiple possible values, each with different probabilities.
-The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's **support**.
-
-There are many distributions we can assign to $R$ depending on how we quantify it.
-For instance, if we measure $R$ using the Normalized Difference Vegetation Index [**NDVI**, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
-However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).
-Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution.
-If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
-
-### Expected resource abundance, $\e(R)$
-
-\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average.
-The **expectation** of a random variable is the value one can *expect* the random variable to take, on average, in the long term.
-Thus, we can let $\e(R)$ indicate the expectation of the random variable $R$.
-When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$.
-We indicate the *estimated* average resource abundance over time by adding a caret on the symbol: $\hat \mu(t)$.
-
-### Variance in resource abundance, $\var(R)$
-
-\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$.
-A random variable's **variance** is a measure of its unpredictability (i.e., variability).
-We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$).
-Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others.
-Additionally, we distinguish between two different sources of variation.
-We refer to known or expected spatiotemporal differences or changes as **heterogeneity**, which we consider as a change in $\mu(t)$.
-In contrast, we use the term **stochasticity** for unpredictable and unexpected variation (i.e., due to $\sigma^2(t)$.
-For example, the location of (high-yield) berry bushes in a valley and the time at which they produce fruit may be heterogeneous, but it becomes predictable once a bear visits all patches in the valley.
-In contrast, whether next year will be a good or bad year for berries is stochastic because it depends on factors the bear cannot predict.
-
-### Simulating resource abundance
-
-\noindent In this paper, we simulate $R$ using a Gamma distribution with time-dependent mean $\mu(t)$ and time-dependent variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
+\noindent In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
 Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
@@
 As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
-This assumption also holds biologically; in the example from above, a mean of zero kg of berries within the bear's habitat during winter implies the variance in berries is also zero.
+This assumption also holds biologically, since resources tend to be less variable when they are less abundant.
+We assume readers are familiar with the concepts of: random variables, probability distributions, the expected value of a random variable, and the variance of a random variable.
+We define and explain each of the four concepts in Appendix 1 for readers who need additional information.
 
-# Effects of resource abundance, $\e(R)$
+## Effects of resource abundance, $\e(R)$
 
-```{r mu-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high spatial needs as animals are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the animal to survive.
+```{r mu-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high spatial needs as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
 Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
@@
 
-```{r, eval=FALSE, echo=FALSE}
-# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
-dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
-    # mean = k * theta
-    # sigma^2 = k * theta^2
-    dgamma(x = x,
-           shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
-           scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
-}
-x <- seq(0, 10, length.out = 250)
-layout(matrix(1:4, ncol = 2, byrow = TRUE))
-plot(x, dgamma2(x, mu = 1, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, low \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 5, sigma2 = 0.5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, low \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 1, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'low \U03BC, high \U03C3\U00B2')
-plot(x, dgamma2(x, mu = 5, sigma2 = 5), type = 'l', ylim = c(0, 0.8),
-     xlab = NA, ylab = 'Density', main = 'high \U03BC, high \U03C3\U00B2')
-layout(1)
-```
-
-\noindent While animals' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
+\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
 All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
@@
 This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an animal is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
 
-When $\e(R)$ is high, we expect animals' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-However, as $\e(R)$ decreases, we expect animals' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force animals to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-It is unclear when animals switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on animal's spatial needs and when an animal may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an animal's lifespan.
+When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+However, as $\e(R)$ decreases, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organism's spatial needs and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
+Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an organism's lifespan.
 For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
-
-It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
-Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).
 
@@
 
+It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
+Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
+Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).
+
 Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
 We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
-Since $R$ is spatiotemporally random, an animal that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
+Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
 However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an animal may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
+peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
 
@@
 When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
-Most often, $\e(R)$ will vary over time, and animals' behaviors will change in response.
+Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response.
 
-# Effects of resource stochasticity, $\var(R)$
+## Effects of resource stochasticity, $\var(R)$
 
-\noindent Although the effect of $R$ on animals' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can fluctuate both spatially and temporally due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
+\noindent Although the effect of $R$ on organisms' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
 For example, since berries are scarce outside the growing season, the variance in berries will also be low.
-But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location only after competitors ate all the berries.
+But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location before fruiting time or after competitors ate all the berries.
 In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
 
-```{r s2-hyp, fig.cap="Hypothesized spatial needs of an animal as a function of resource stochasticity ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low spatial needs as animals are able to depend on somewhat predictable resources.
-As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident animals) or a switch to dispersal, nomadism, or migration.
+```{r s2-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource stochasticity ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low spatial needs as organisms are able to depend on relatively predictable resources.
+As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration.
 Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
@@
 
-Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases animals' fitness [@berger_climate_2018] and increases their spatial needs.
+Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [@berger_climate_2018] and increases their spatial needs.
 While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
 A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-Thus, there remains a need for developing a more complete understanding of how animals' spatial needs change with environmental stochasticity.
+Thus, there remains a need for developing a more complete understanding of how organisms' spatial needs change with environmental stochasticity.
 
@@
 ```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(a.) Although resource abundance ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
+(a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
 (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
@@
 
-Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on animals' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases animals' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), animals may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, animals may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
+Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' spatial needs.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
+If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
 
@@
 Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
-Biologically, this is important because some animals may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
-An animal's perception of changes in $R$ will depend strongly on the animal's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires animals have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
+Biologically, this is important because some organisms may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
+An organism's perception of changes in $R$ will depend strongly on the organism's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
+The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires organisms have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
 Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
 
-# Interactive effects of $\e(R)$ and $\var(R)$
+## Interactive effects of $\e(R)$ and $\var(R)$
 
 \noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' spatial needs, but we presented the two parameters as independent drivers of movement.
-However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, animals may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on animal's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
+Consequently, models estimating the effects of $R$ on organism's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
@@
 <!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
-However, the effects of $\var(R)$ on animals' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small animals, while animals with larger spatial needs may not even notice a change.
+However, the effects of $\var(R)$ on organisms' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
+Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small organisms, while organisms with larger spatial needs may not even notice a change.
 -->
@@
 However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality, heterogeneity, or continentality).
+Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
 Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
 A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
-Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on animals' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
-Thus, it is important to disentangle the effects of spatiotemporal heterogeneity (i.e., changes in $\e(R)$, such as seasonal trends in $\mu(t)$) and stochasticity ($\var(R)$).
+Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
+Thus, it is important to disentangle the effects of changes in $\e(R)$, such as seasonal trends in $\mu(t)$, and stochasticity ($\var(R)$).
 The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
 
-@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since animals in habitats with high and low seasonality had opposite responses to $\e(R)$.
+@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since organisms in habitats with high and low seasonality had opposite responses to $\e(R)$.
 Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because animals respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
-However, it does not explain why animals may respond to $\e(R)$ in opposite ways in different habitats.
+This may be because organisms respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
+However, it does not explain why organisms may respond to $\e(R)$ in opposite ways in different habitats.
 Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
@@
 
-## Simulating animal responses to $\e(R)$ and $\var(R)$
+## Simulating responses to $\e(R)$ and $\var(R)$
 
-To illustrate how $\e(R)$ and $\var(R)$ affect animals' spatial needs as well as the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations in figure \@ref(fig:5-5-sims) (for methodological details, see Appendix 1).
+To support our hypothesis of how organisms' spatial needs are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate $2^{10}=1024$ tracks from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
+The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
+However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not assume the movement process is spatially stationary, so the organism is not assumed to be range-resident.
+Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.
+
+Each of the 1024 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
+Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution.
+The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in figure \@ref(fig:5-5-sims)).
+Note that $t$ was constant within each set of 1024 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
+Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ with no correlation to the following track.
+Finally, each set of 1024 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution.
+Additional information is provided in Appendix 2.
+
+Figure \@ref(fig:5-5-sims) shows how the simulated spatial needs changed in response to $\mu(t)$ and $\sigma^2(t)$.
 The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the animal.
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
 As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
 
-```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an animal living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time.
+```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (and homogenous over space).
 Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.\\newline{}Simulations were run such that animals followed the same $2^{10}=1024$ tracks at each time point starting from the location $\\langle 0, 0\\rangle$ until they reach satiety, after which they were allowed to return to $\\langle 0, 0\\rangle$ over the same amount of time required to reach satiety.
-The animal's movement and utilization distribution were then estimated using an Ornstein-Uhlenbeck Foraging (OUF) model and Autocorrelated Kernel Density Estimation (AKDE), respectively, via the $\\tt{ctmm}$ package (Fleming and Calabrese 2021) for $\\tt{R}$ (R Core Team 2022).
-$H$ was estimated using the UD's 0.95 quantile, but the effect is true for all quantiles since the UD was estimated via AKDE.
-Additional information is provided in Appendix 1.", out.width='100%'}
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
 knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
@@
 
-In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], animals should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase animals' spatial needs suddenly and unpredictably.
+In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
+However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase organisms' spatial needs suddenly and unpredictably.
 Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
 This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
-In cases where $\mu(t)$ is highly unpredictable, animals should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
+In cases where $\mu(t)$ is highly unpredictable, organisms should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
 
@@
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
-This is because, similarly to resource-poor periods, times of greater stochasticity require the animal to move over larger areas for longer periods of time.
-This results in a greater uncertainty in how much time and space the animal will require to reach satiety, or indeed whether an animal living in highly stochastic environments can even reach satiety within a finite amount of time.
+This is because, similarly to resource-poor periods, times of greater stochasticity require the organism to move over larger areas for longer periods of time.
+This results in a greater uncertainty in how much time and space the organism will require to reach satiety, or indeed whether an organism living in highly stochastic environments can even reach satiety within a finite amount of time.
 
@@
 
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on animal space use are useful examples that can be thought of as simplified scenarios.
+Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on space use are useful examples that can be thought of as simplified scenarios.
 $\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
@@
 Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
-However, if stochastic changes are sufficiently small and frequent, animals may perceive them as continuous and smooth changes rather a series of small and sudden changes.
+However, if stochastic changes are sufficiently small and frequent, organisms may perceive them as continuous and smooth changes rather a series of small and sudden changes.
 
@@
 Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
-Secondly, the animal's movement was simulated to be range resident within an homogeneous environment.
-Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, animals may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood.
-Thus, we hypothesize that our simulations may be underestimating animals' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated animals cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
+Secondly, the organism's movement was simulated to be range resident within an homogeneous environment.
+Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, organisms may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood.
+Thus, we hypothesize that our simulations may be underestimating organisms' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated organisms cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
 Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
@@
 
-\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging animals via remote sensing data such as NDVI [@pettorelli_normalized_2011].
+\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging organisms via remote sensing data such as NDVI [@pettorelli_normalized_2011].
 To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
-Appendix 2 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
+Appendix 3 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
 
@@
 
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs.
+We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' spatial needs.
 Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating animals' spatial needs under different scenarios.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' spatial needs under different scenarios.
 
-### Modeling $R$
+### Modeling $R$ {#modeling-r}
 
@@
 An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
-Although there is no error-proof system, choosing the complexity of the terms based on the animal's ability to detect change and adapt is a reasonable starting point.
+Although there is no error-proof system, choosing the complexity of the terms based on the organism's ability to detect change and adapt is a reasonable starting point.
 Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
@@
 
+\singlespace <!-- to avoid excessive spacing -->
 \begin{equation}
@@
 
-\noindent Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
+\doublespace <!-- back to double spaced -->
+
+Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
 Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
@@
 <!--
-## Animal movement as a stochastic process
+## Movement as a stochastic process
 
-### Drivers of animal movement
+### Drivers of movement
 
-\noindent @nathan_movement_2008 suggested we can generalize animal movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
+\noindent @nathan_movement_2008 suggested we can generalize movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
 
@@
 
-\noindent such that an animal's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
+\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
 This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
 The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on animals' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
+Thus, we can instead focus on organisms' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
 
@@
 
-\noindent To understand how movement can be viewed as a **stochastic process**, imagine an animal that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
-If we do not know why the animal is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the animal may move from one time point to another.
-However, we can use the animal's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
+\noindent To understand how movement can be viewed as a **stochastic process**, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
+If we do not know why the organism is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the organism may move from one time point to another.
+However, we can use the organism's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
 This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
-Similarly, we can define $\tau_v$ to be the amount of time the animal moves with the same direction and speed, or the time required for an animal's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
-$\tau_v$ is referred to the animal's **directional persistence** or the velocity autocorrelation parameter.
+Similarly, we can define $\tau_v$ to be the amount of time the organism moves with the same direction and speed, or the time required for an organism's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
+$\tau_v$ is referred to the organism's **directional persistence** or the velocity autocorrelation parameter.
 
@@
 
-Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the animal is moving with a similar speed and direction.
-However, since the time required for an animal to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the animal's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the animal's spatial needs and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
+Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
+However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's spatial needs and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
 Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
 
-Predicting animal movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the animal moved and less on *why* it moved.
-Animals may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
-Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in animal movement while focusing on the animals' movement behaviors rather than the causes.
+Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the organism moved and less on *why* it moved.
+Organisms may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
+Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in organism movement while focusing on the organisms' movement behaviors rather than the causes.
 
@@
 
-\noindent The concept of a home range has a long history in ecological research as an indicator of the space an animal requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for animals which are not range-resident, such as migrating animals [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic animals [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify animal's spatial needs using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of animals that are not range-resident.In the case of animals that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
-Note, however, that although range-resident animals will generally have smaller positional variance ($\vpos$) than migratory and nomadic animals, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an animal's movement but not its behavior.
-Instead, the animal's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
-Animals with a $\tau_p$ shorter than a year are likely range-resident, while animals with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an animal's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
+\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
+Although here we quantify organism's spatial needs using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
+In the case of organisms that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
+Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
+Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
+In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
 -->
@@
 <!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
-However, most commonly used movement models assume animals' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
+However, most commonly used movement models assume organisms' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
 -->
 
-<!-- To quantify an animal's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
-The width of the window, $w$, should be a fixed value greater than two or three times the animal's expected range crossing time [@fleming_overcoming_2019].
-If the animal's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
-Alternatively, if the animal's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
+<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
+The width of the window, $w$, should be a fixed value greater than two or three times the organism's expected range crossing time [@fleming_overcoming_2019].
+If the organism's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
+Alternatively, if the organism's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
 -->
 
-<!-- Modeling animal movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
+<!-- Modeling movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
 In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022].
@@
 
-### Modeling the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs
+### Modeling the effects of $\e(R)$ and $\var(R)$ on spatial needs
 
@@
 Here, we model the effects of $\e(R)$ and $\var(R)$ on $H$ using a shape-constrained GAM with via the `scam` package [@pya_scam_2022].
-Additional information on the model can be found in Appendix 2.
+Additional information on the model can be found in Appendix 3.
 
 ### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's spatial needs
-
-```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
-(a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
-(b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir.
-(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
-(d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
-(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
-Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
-See Appendix 2 for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
-knitr::include_graphics('../figures/tapir-example-with-data.png')
-```
 
@@
 
+```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resources.
+(a.) Trends in resource abundance over time, $\\mu(t)$, esimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
+(b.) Varince in resources over time, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
+(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
+(d., e.) Estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
+The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
+(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
+Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
+See Appendix 3 for additional information.
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
+knitr::include_graphics('../figures/tapir-example-with-data.png')
+```
+
 # Discussion
 
-<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on animals' spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on animals' spatial needs by building upon previous work rather than stating their results as new findings.
+<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on spatial needs.
+The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on spatial needs by building upon previous work rather than stating their results as new findings.
 -->
 
-## Animal adaptations to changes in $\e(R)$ and $\var(R)$
+## Adaptations to changes in $\e(R)$ and $\var(R)$
 
-\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an animal will be able to predict the event and how it will respond to it (or fail to do so).
+<!-- spores are also an adaptation to periods of low $\e(R)$ or high $\var(R)$ -->
+
+\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so).
 Events that occur at the scale of the organism [i.e.
-approximately one or more times per generation and on the spatial scale of the animal's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019].
+approximately one or more times per generation and on the spatial scale of the organism's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019].
 Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them.
@@
 with species names:
-Animals may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006].
+Organisms may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006].
 Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
@@
 Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022].
-Of course, an animal with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an animal with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity.
+Of course, an organism with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an organism with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity.
 Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].
@@
 However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources during unpredictable times [whooping cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
-Instead, if anomalies occur repeatedly over time, animals may adapt to a higher $\sigma^2(t)$ or shift their ranges altogether [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
+Instead, if anomalies occur repeatedly over time, organisms may adapt to a higher $\sigma^2(t)$ or shift their ranges altogether [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
 Still, while range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021], they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [e.g., @wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range, since it is determined by their food sources' and associates' ability to move and adapt, too.
@@
 
-\noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on animal behavior.
-In these times of great change and climatic uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021], it is increasingly important to recognize how environmental stochasticity affects animal behavior and shapes ecosystems.
+\noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on organisms' behaviors.
+In these times of great change and climatic uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021], it is increasingly important to recognize how environmental stochasticity affects organisms' behaviors and shapes ecosystems.
 Accounting for changes in both resource abundance and resource stochasticity is crucial when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
-Protecting areas with high resource abundance and low resource stochasticity will reduce animal's spatial requirements.
+Protecting areas with high resource abundance and low resource stochasticity will reduce animals' spatial requirements.
 
@@
 
-\noindent The work presented here provides a unifying framework for viewing animal movement as a function of resource abundance and stochasticity.
-We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on animals' spatial needs and movement behavior.
-We demonstrate that animals' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving window approach that allows one to model animal movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism.
-The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space animals currently need and will need in the future under different climate change scenarios.
+\noindent The work presented here provides a unifying framework for viewing movement as a function of resource abundance and stochasticity.
+We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' spatial needs and movement behavior.
+We demonstrate that organisms' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
+We also provide a moving window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism.
+The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
 
@@
 We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Chris Fleming, Lauren Mills, and Sarah Wyse.
+In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
 

### 2023-09-09 analysis/simulations/hr-mean-variance-simulations-days.R
changed log10 to log2

@@ -7,7 +7,7 @@ source('analysis/figures/mean-variance-trends-panel-data.R') # trends in E(U) an
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
 
 tracks <- readRDS('simulations/labelled-tracks.rds') %>% # movement tracks
-  filter(day <= 100) # 100 are sufficient for stable estimates
+  filter(day <= 200) # 100 are sufficient for stable estimates
 types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic') # trend types
 
 # using for loops to reduce computation costs

### 2023-09-09 analysis/simulations/hr-simulation-extreme-scenarios.R

@@ -3,7 +3,7 @@ library('dplyr')   # for data wrangling (e.g., %>%)
 library('tidyr')   # for data wrangling (e.g., nested tibbles)
 library('ggplot2') # for fancy plots
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/figures/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/mean-variance-trends-panel-data.R') # means & variances
 source('analysis/simulations/movement-model.R') # for consistency across scripts
 source('functions/get_hr.R') # for extracting gaussian home range
 theme_set(theme_bw())
@@ -151,6 +151,9 @@ saturation_days %>%
   select(case, n_days, sigma, hr) %>%
   readr::write_csv('simulations/hr-saturation-days.csv')
 
+# to avoid having to run all code above
+saturation_days <- readRDS('simulations/hr-saturation-days.rds')
+
 p_hr_days <-
   ggplot(saturation_days, aes(n_days, hr)) +
   facet_wrap(~ case, nrow = 1) +
@@ -162,7 +165,8 @@ p_hr_days <-
   scale_x_continuous(expression(Number~of~days~sampled~(log[2]~scale)),
                      trans = 'log2', breaks = c(2, 16, 128, 1024),
                      limits = c(2, 1100)) +
-  scale_y_log10(expression(Estimated~home~range~(log[10]~scale))); p_hr_days
+  scale_y_continuous(trans = 'log2', expression(atop(Estimated~'space-use',
+                                requirements~(log[2]~scale)))); p_hr_days
 
-ggsave('figures/simulations/hr-over-days.png', plot = p_hr_days,
+ggsave('figures/hr-over-days.png', plot = p_hr_days,
        width = 6, height = 3, dpi = 'print')

### 2023-09-09 analysis/simulations/movement-model.R
requires terra

@@ -1,4 +1,5 @@
 library('ctmm') # for movement modeling
+library('terra') # for rasters
 DELTA_T <- 60
 SAMPLES <- seq(0, 60 * 60 * 12, by = DELTA_T) # sample 12 hours every DELTA_T seconds
 

### 2023-09-09 writing/appendix-1-concepts-and-definitions.Rmd
removed bold because all figures have unicode

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' spatial needs
+title: \large How resource abundance and stochasticity affect animals' space-use requirements
 subtitle: "Appendix 1: Concepts and definitions"

### 2023-09-09 writing/appendix-2-simulations.Rmd
removed bold because all figures have unicode

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' spatial needs
+title: \large How resource abundance and stochasticity affect animals' space-use requirements
 subtitle: "Appendix 2: Simulations"
@@
 Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/mean-variance-5-by-5-hr-sims.png',
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%', cache=FALSE}
+knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png',
                         rel_path = FALSE)
@@
 
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements.
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated space-use requirements.
 The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
@@
 Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in spatial needs strongly increases with mean spatial needs.", out.width='100%'}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-regression-plots.png',
+Additionally, note how the variance in space-use requirements strongly increases with mean space-use requirements.", out.width='100%', cache=FALSE}
+knitr::include_graphics('../figures/simulation-regression-plots.png',
                         rel_path = FALSE)
@@
 
-```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' spatial needs were simulated.
+```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' space-use requirements were simulated.
 (a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
@@
 (d.) The process was repeated $2^{10}=1024$ times (13 tracks pictured in this panel).
-(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h'}
+(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-example.png',
@@
 Additionally, the track at $\\Delta t = 30$s is still sufficiently tortuous to represent realistic animal movement.
-(d.) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE}
+(d.) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/thinning-examples.png',
@@
 
-In this script, we check how many tracks are necessary to produce stable and accurate estimates of the spatial needs.
+In this script, we check how many tracks are necessary to produce stable and accurate estimates of the space-use requirements.
 We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
@@
                      limits = c(2, 1100)) +
-  scale_y_log10(expression(Estimated~home~range~(log[10]~scale)))
+  scale_y_continuous(expression(atop(Estimated~'space-use',
+                                     requirements~(log[2]~scale))),
+                     trans = 'log2')
 ```
 
-```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated spatial needs as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
-In both cases, 100 days are sufficient to produce stable estimates of spatial needs."}
+```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated space-use requirements as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
+In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE}
 knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/hr-over-days.png',
@@
 
-This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated spatial needs.
+This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated space-use requirements.
 Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity.
@@
 Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
-When estimating spatial needs based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 3 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+When estimating space-use requirements based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 3 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
@@
 The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown).
-To estimate the effects of resource abundance and stochasticity on the simulated spatial needs, we fit 3 location-scale GAMs (GAMLSs) with a location-scale gamma distribution (since home ranges are strictly positive) of increasing complexity.
+To estimate the effects of resource abundance and stochasticity on the simulated space-use requirements, we fit 3 location-scale GAMs (GAMLSs) with a location-scale gamma distribution (since home ranges are strictly positive) of increasing complexity.
 We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones.

### 2023-09-09 writing/appendix-3-empirical-modeling.Rmd
removed bold because all figures have unicode

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' spatial needs
+title: \large How resource abundance and stochasticity affect animals' space-use requirements
 subtitle: "Appendix 3: Empirical modeling"
@@
 For ease of reference, we also include the figure below (figure \@ref(fig:tapir-mw)).
-Estimating the effects of resource abundance and unpredictability on the tapir's spatial needs requires us to first estimate the changes in the tapir's spatial needs (section \@ref(movement), figure \@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figures \@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figures \@ref(fig:tapir-mw)d-f).
+Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement), figure \@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figures \@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figures \@ref(fig:tapir-mw)d-f).
 
@@
 (d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
+The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them.
 (f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
@@
 
-# Modeling the effects of $\e(R)$ and $\var(R)$ on spatial needs {#final-fig}
+# Modeling the effects of $\e(R)$ and $\var(R)$ on space-use requirements {#final-fig}
 
@@
 
-To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's spatial needs.
+To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space-use requirements.
 To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.
@@
 
-```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's spatial needs (on the log link scale) using a GAM with no shape constraints.
+```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space-use requirements (on the log link scale) using a GAM with no shape constraints.
 The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
@@
 
-```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's spatial needs (on the log link scale) using a shape-constrained GAM.
+```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space-use requirements (on the log link scale) using a shape-constrained GAM.
 The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.

### 2023-09-09 writing/manuscript.Rmd
removed bold because all figures have unicode

@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' spatial needs
+title: \large How resource abundance and stochasticity affect organisms' space-use requirements
 author:
@@
     - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
-    # - \usepackage{afterpage} # for \afterpage{\<command>}, e.g. \afterpage{\clearpage}
     - \usepackage{hanging} # for hanging indents in references
@@
 This is not a trivial oversight.
-Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' spatial needs, challenging ecological theory and applied conservation alike.
-In this study, we first review the literature of studies that estimate animals' spatial needs as a function of resource abundance.
+Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
+In this study, we first review the literature of studies that estimate animals' space-use requirements as a function of resource abundance.
 We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use.
@@
 Instead, results are reported as independent, novel findings.
-Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' spatial needs.
+Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements.
 In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
@@
 
-Here, we illustrate how organisms' spatial needs depend on both the abundance and unpredictability of resources.
+Here, we illustrate how organisms' space-use requirements depend on both the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource *abundance* on organisms' spatial needs while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' spatial needs while suggesting a second unifying hypothesis.
+Next, we provide a review of the effects of resource *abundance* on organisms' space-use requirements while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space-use requirements while suggesting a second unifying hypothesis.
 Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
@@
 
-```{r mu-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high spatial needs as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, spatial needs should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
-Note that the relationship between $\\e(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
+Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
-The hypothesis that spatial needs decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
 This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between spatial needs and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
 
@@
 However, as $\e(R)$ decreases, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organism's spatial needs and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
+It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
 Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an organism's lifespan.
@@
 
-\noindent Although the effect of $R$ on organisms' spatial needs is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+\noindent Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
 However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
@@
 
-```{r s2-hyp, fig.cap="Hypothesized spatial needs of an organism as a function of resource stochasticity ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low spatial needs as organisms are able to depend on relatively predictable resources.
-As $\\var(R)$ increases, spatial needs should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and spatial needs cannot be linear because it would require spatial needs to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+```{r s2-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource stochasticity ($\\var(R)$).
+We expect low values of $\\var(R)$ to result in low space-use requirements as organisms are able to depend on relatively predictable resources.
+As $\\var(R)$ increases, space-use requirements should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration.
+Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [@berger_climate_2018] and increases their spatial needs.
+Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [@berger_climate_2018] and increases their space-use requirements.
 While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
 A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-Thus, there remains a need for developing a more complete understanding of how organisms' spatial needs change with environmental stochasticity.
+Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.
 
@@
 
-Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' spatial needs.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect spatial needs to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
+Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' space-use requirements.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
@@
 
-\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' spatial needs, but we presented the two parameters as independent drivers of movement.
+\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement.
 However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
@@
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
-Thus, we expect $\var(R)$ to have a stronger effect on spatial needs when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
+Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
 
 <!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
-However, the effects of $\var(R)$ on organisms' spatial needs likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small organisms, while organisms with larger spatial needs may not even notice a change.
+However, the effects of $\var(R)$ on organisms' space-use requirements likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
+Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small organisms, while organisms with larger space-use requirements may not even notice a change.
 -->
 
-@rizzuto_forage_2021 found that the spatial needs of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
-However, it is hard to determine how $\sigma^2(t)$ affected the hares' spatial needs since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
+@rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
+However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
 Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
@@
 A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
-Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' spatial needs, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
+Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' space-use requirements, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
 Thus, it is important to disentangle the effects of changes in $\e(R)$, such as seasonal trends in $\mu(t)$, and stochasticity ($\var(R)$).
@@
 
-To support our hypothesis of how organisms' spatial needs are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+To support our hypothesis of how organisms' space-use requirements are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate $2^{10}=1024$ tracks from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
@@
 
-Figure \@ref(fig:5-5-sims) shows how the simulated spatial needs changed in response to $\mu(t)$ and $\sigma^2(t)$.
-The top row (constant $\var(R)$) shows how spatial needs vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the spatial needs decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
+Figure \@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$.
+The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
 As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
@@
 In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase organisms' spatial needs suddenly and unpredictably.
-Thus, it is best to include a "buffer" area so the available space is larger than the estimated spatial needs.
+However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase organisms' space-use requirements suddenly and unpredictably.
+Thus, it is best to include a "buffer" area so the available space is larger than the estimated space-use requirements.
 This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated spatial needs while $\e(R)$ remains constant.
+The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
@@
 Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
-Both $\e(R)$ and $\var(R)$ affect spatial needs precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
+Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
 We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
@@
 Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in spatial needs strongly increases with mean spatial needs.", out.width='100%'}
+Additionally, note how the variance in space-use requirements strongly increases with mean space-use requirements.", out.width='100%'}
 knitr::include_graphics('../figures/simulation-regression-plots.png')
@@
 To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
-Appendix 3 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's spatial needs.
+Appendix 3 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
 
@@
 Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
-Individuals and populations that were once thought to have different spatial needs due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
+Individuals and populations that were once thought to have different space-use requirements due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
 In the following sections, we illustrate how one can apply this framework to empirical data.
 
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' spatial needs.
+We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements.
 Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' spatial needs under different scenarios.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' space-use requirements under different scenarios.
 
@@
 The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on organisms' movement and spatial needs as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
+Thus, we can instead focus on organisms' movement and space-use requirements as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
 
@@
 However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's spatial needs and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
 Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
@@
 
-### Quantifying spatial needs
+### Quantifying space-use requirements
 
 \noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify organism's spatial needs using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
-In the case of organisms that are range resident for short periods of time, we can model their movement and short-term spatial needs within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Although here we quantify organism's space-use requirements using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
+In the case of organisms that are range resident for short periods of time, we can model their movement and short-term space-use requirements within distinct and discrete periods or using a sliding window approach (see the discussion section).
 Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
@@
 
-<!-- ### Modeling spatial needs -->
+<!-- ### Modeling space-use requirements -->
 
@@
 
-### Modeling the effects of $\e(R)$ and $\var(R)$ on spatial needs
+### Modeling the effects of $\e(R)$ and $\var(R)$ on space-use requirements
 
@@
 
-### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's spatial needs
+### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's space-use requirements
 
@@
 (d., e.) Estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean spatial needs and the variance around them.
+The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them.
 (f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
@@
 
-<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on spatial needs.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on spatial needs by building upon previous work rather than stating their results as new findings.
+<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on space-use requirements.
+The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on space-use requirements by building upon previous work rather than stating their results as new findings.
 -->
@@
 Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
-Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' spatial needs, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
-This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in spatial needs.
+Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' space-use requirements, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
+This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in space-use requirements.
 
@@
 \noindent The work presented here provides a unifying framework for viewing movement as a function of resource abundance and stochasticity.
-We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' spatial needs and movement behavior.
-We demonstrate that organisms' spatial needs respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in spatial needs and shifts between range-residency and nomadism.
+We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior.
+We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
+We also provide a moving window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.
 The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.

### 2023-09-09 analysis/figures/default-figure-styling.R

@@ -5,7 +5,7 @@ library('stringi') # for working with strings
 # cannot bold text because unicode characters fail
 theme_set(theme_bw() +
             theme(legend.position = 'none',
-                  text = element_text(size = 15, face = 'bold')))
+                  text = element_text(size = 15)))
 
 # custom color-blind palette
 pal <- c('#ff8c00', '#4477AA', '#009900', '#66CCEE',

### 2023-09-09 analysis/figures/habitat-examples.R

@@ -14,11 +14,9 @@ library('ragg')      # needed for custom alpha with coord_cartesian
 source('functions/rgamma2.R') # function to generate rgamma() from mean and variance
 source('functions/qgamma2.R') # function to generate qgamma() from mean and variance
 source('analysis/figures/default-figure-styling.R')
-# remove bold because it doesn't work with unicode characters
 theme_set(theme_get() + 
             theme(legend.position = 'top',
-                  panel.grid = element_blank(),
-                  text = element_text(size = 15, face = 'plain')))
+                  panel.grid = element_blank()))
 
 # resource abundance palette
 LOW <- '#744700'

### 2023-09-09 analysis/figures/home-range-hypotheses.R

@@ -4,12 +4,12 @@ library('tidyr')   # for data wrangling
 library('ggplot2') # for fancy plots
 library('ggbrace') # for curly braces on fancy plots
 library('cowplot') # for fancy multi-panel plots
-source('analysis/figures/default-figure-styling.R') # for common theme and color palette
+source('analysis/figures/default-figure-styling.R') # ggplot theme and palettes
 
 theme_set(theme_get() + theme(legend.position = 'none',
                               text = element_text(size = 22)))
 
-hr_lab <- 'Spatial needs' # label for y axis
+hr_lab <- 'Space-use requirements' # label for y axis
 
 # hypothesis for change in H over E(R) ----
 d <-
@@ -33,7 +33,7 @@ p_mu <-
   geom_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180) +
   geom_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0) +
   geom_text(aes(x = x, y = y, label = text), notes, size = 6) +
-  scale_x_continuous('Resource abundance, \U1D53C(\U1D445)', breaks = NULL,
+  scale_x_continuous('Resource abundance, E(\U1D445)', breaks = NULL,
                      expand = c(0, 0)) +
   scale_y_continuous(hr_lab, breaks = 0, expand = c(0, 0)); p_mu
 
@@ -46,7 +46,7 @@ p_s2 <-
   geom_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180) +
   geom_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0) +
   geom_text(aes(x = x, y = y, label = text), notes, size = 6) +
-  scale_x_reverse('Resource stochasticity, \U1D54D(\U1D445)', breaks = NULL,
+  scale_x_reverse('Resource stochasticity, Var(\U1D445)', breaks = NULL,
                   expand = c(0, 0)) +
   scale_y_continuous(hr_lab, breaks = 0, expand = c(0, 0)); p_s2
 
@@ -56,113 +56,9 @@ ggsave('figures/mean-abundance-hr-hypotheses.png', plot = p_mu, width = 8,
 ggsave('figures/variance-abundance-hr-hypotheses.png', plot = p_s2, width = 8,
        height = 8, scale = 1, dpi = 300, bg = 'white')
 
+
[…]

### 2023-09-09 analysis/figures/hr-mean-variance-simulations.R

@@ -7,17 +7,13 @@ source('analysis/mean-variance-trends-panel-data.R') # tibble of mu, var
 days <- readRDS('simulations/days-hrs.rds')
 
 # main simulation figure of change in home range (5x5 plot) ----
-e_r <- 'Resource abundance, \U1D53C(\U1D445|t) = \U1D707(\U1D461)'
-v_r <- 'Resource stochasticity, \U1D54D(\U1D445|t) = \U1D70E\U00B2(\U1D461)'
-hr_lab <- 'Home range size, \U1D43B' # label for y axis
+e_r <- 'Resource abundance, E(\U1D445|t) = \U1D707(\U1D461)'
+v_r <- 'Resource stochasticity, Var(\U1D445|t) = \U1D70E\U00B2(\U1D461)'
+hr_lab <- 'Space-use requirements, \U1D43B' # label for y axis
 p_sim <-
   ggplot(days) +
   facet_grid(variance ~ mean, drop = FALSE) + # facet by trends in mu and var
-  # area between 50% and 95% HRs
-  # geom_ribbon(aes(animal, ymin = hr_50, ymax = hr_95), fill = pal[3],
-  #             alpha = 0.2) +
-  # add lines for 50% and 95% HRs
-  # geom_line(aes(animal, hr_50), color = pal[3]) +
+  # add lines for 95% HR
   geom_line(aes(animal, hr_95), color = pal[3], linewidth = 1) +
   scale_x_continuous('Time', breaks = NULL) +
   scale_y_continuous(hr_lab, breaks = NULL, limits = c(0, NA)) +
@@ -52,43 +48,3 @@ p_full
 # save the plot as a png
 ggsave('figures/mean-variance-5-by-5-hr-sims.png', p_full,
        width = 10, height = 5.625, units = 'in', dpi = 'print', bg = 'white')
-
-# estimate fraction of animals not satisfied if only mean is accounted for ----
-days_suff <-
-  left_join(days,
-            filter(days, variance == 'constant') %>%
-              select(mean, animal, hr_50, hr_95) %>%
-              rename(hr_50_x = hr_50, hr_95_x = hr_95),
-            by = c('mean', 'animal')) %>%
-  # see if the animal has enough space based on the 95% quantile
-  mutate(enough_50 = hr_50 <= hr_95_x,
-         enough_95 = hr_95 <= hr_95_x,
-         additional = if_else(enough_95, hr_95, hr_95_x))
-
-p_sim_suff <-
-  ggplot(days_suff) +
[…]

### 2023-09-09 analysis/figures/moving-window-figure-anna-betals.R

@@ -42,8 +42,6 @@ tel <- data.frame(tapir$data[[1]]) %>%
 
 tapir <-
   readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
-  # mutate(ess = map_dbl(model, \(.m) summary(.m)$DOF['area']),
-  #        weight = ess / mean(ess)) %>%
   mutate(est = map(dataset, \(.d) filter(tel, timestamp %in% .d$timestamp)),
          mu = map_dbl(est, \(.d) mean(.d$mu)),
          sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
@@ -95,48 +93,6 @@ m <- scam(hr_est ~ s(mu, bs = 'mpd', k = 4) + s(sigma2, bs = 'mpi', k = 4),
           family = Gamma('log'),
           data = tapir)
 plot(m, pages = 1, scheme = 2, trans = exp, ylim = c(log(0), log(4)))
-
-# p_d <-
-#   tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
-#          sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
-#   mutate(hr_mu = predict(m, newdata = ., terms = 's(mu)',
-#                          type = 'response'),
-#          hr_sigma2 = predict(m, newdata = ., terms = 's(sigma2)',
-#                              type = 'response')) %>%
-#   pivot_longer(c(hr_mu, hr_sigma2), names_to = 'pred',
-#                names_pattern = 'hr_(.+)', values_to = 'hr') %>%
-#   pivot_longer(c(mu, sigma2), names_to = 'pred2', values_to = 'x') %>%
-#   filter(pred == pred2) %>%
-#   mutate(param = if_else(pred == 'mu', e_r, v_r)) %>%
-#   ggplot() +
-#   facet_wrap(~ param, scales = 'free', strip.position = 'bottom',
-#              ncol = 1) +
-#   coord_cartesian(ylim = c(0, 12.5)) +
-#   geom_point(aes(value, hr_est),
-#              tapir %>%
-#                pivot_longer(c(mu, sigma2), names_to = 'param',
-#                             values_to = 'value') %>%
-#                mutate(param = if_else(param == 'mu', e_r, v_r)),
-#              alpha = 0.5, color = pal[3]) +
-#   geom_line(aes(x, hr, color = param), linewidth = 2,
[…]

### 2023-09-09 analysis/simulations/modeling-R-and-hr.R

@@ -6,8 +6,8 @@ library('cowplot') # for fancy multi-panel plots
 source('analysis/figures/default-figure-styling.R') # for color palette
 theme_set(theme_bw() + theme(legend.position = 'none'))
 
-e_r <- 'Resource abundance, \U1D53C(\U1D445)'
-v_r <- 'Resource stochasticity, \U1D54D(\U1D445)'
+e_r <- 'Resource abundance, E(\U1D445)'
+v_r <- 'Resource stochasticity, Var(\U1D445)'
 
 sims <- readRDS('simulations/days-hrs.rds') %>%
   tibble() %>%
@@ -106,16 +106,17 @@ ggplot() +
             filter(preds, quantile == 'hr_95')) +
   scale_color_manual(values = pal) +
   scale_x_continuous(NULL, breaks = NULL) + 
-  scale_y_continuous('Home range size, \U1D43B', breaks = NULL) +
+  scale_y_continuous('Space-use requirements, \U1D43B', breaks = NULL) +
   # to make facet strip look like the x axis
   theme(legend.position = 'none',
         strip.background = element_blank(),
         strip.text = element_text(size = 12))
+
 ggsave('figures/simulation-regression-plots.png',
        width = 6, height = 3, dpi = 'print', bg = 'white', scale = 1.5)
 
 # effect of sqrt(sigma2) is nonlinear since it is on the same scale as mu
-sd_r <- "Standard deviation in \U1D445, \U221A\U1D54D(\U1D445)"
+sd_r <- "Standard deviation in \U1D445, \U221A Var(\U1D445)"
 
 preds_sd <- bind_rows(preds,
                       filter(preds, x == v_r) %>%

### 2023-09-11 writing/appendix-2-simulations.Rmd
renamed sims scripts; added sensitivity analyses
also corrected the number of simulations: only 200 used since HR estimates are stable for n >~ 100

@@
 
-To reduce sampling variance between simulations at each time point in each panel of figure \@ref(fig:5-5-sims)), we used the same set of $2^{10} = 1024$ simulated tracks, which we generate in the `analysis/simulations/tracks.R` script.
+To reduce sampling variance between simulations at each time point in each panel of figure \@ref(fig:5-5-sims)), we used the same set of 200 simulated tracks, which we generate in the `analysis/simulations/tracks.R` script.
 Most intermediate and diagnostic checks are not included in this document for the sake of brevity and simplicity, but their outputs and conclusions are listed in this section.
@@
 Once the animal collected sufficient resources, the animal stopped moving (i.e., the track was truncated).
-(d.) The process was repeated $2^{10}=1024$ times (13 tracks pictured in this panel).
+(d.) The process was repeated 200 times (13 tracks pictured in this panel).
 (e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/simulation-example.png',
-                        rel_path = FALSE)
+knitr::include_graphics('../figures/simulation-example.png', rel_path = FALSE)
 ```
@@
 
-readRDS('H:/GitHub/hr-resource-stoch//simulations/tracks.rds')
+readRDS('H:/GitHub/hr-resource-stoch/simulations/tracks.rds')
 ```
@@
 ```{r, eval=TRUE, echo=FALSE}
-readRDS('H:/GitHub/hr-resource-stoch//simulations/labelled-tracks.rds')
+readRDS('H:/GitHub/hr-resource-stoch/simulations/labelled-tracks.rds')
 ```
@@
 (d.) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/thinning-examples.png',
-                        rel_path = FALSE)
+knitr::include_graphics('../figures/thinning-examples.png', rel_path = FALSE)
 ```
@@
 ## Checking how many tracks were necessary
+
+Script: `analysis/simulations/2-hr-mean-variance-simulations-days.R`
+
+The end of this script generates figures that show the mean time spent searching for resources before reaching satiety.
+This appendix includes the two figures that show the absolute and relative differences in time required to reach satiety as a function of the number of tracks.
+The estimated times for each number of replicates can be found in the `figures/5-by-5-sensitivity-analysis/` folder.
+
+```{r}
+all <-
+  tibble(d = map_dfr(
+    list.files('simulations', pattern = '*-replicates.rds',
+               full.names = TRUE),
+    function(fn) {
+      readRDS(fn) %>%
+        mutate(mean = paste(mean, 'mean'),
+               variance = paste(variance, 'var'),
+               replicates = max(day)) %>%
+        group_by(mean, variance, animal, replicates) %>%
+        summarize(est = mean(t_expl),
+                  .groups = 'drop')
+    })) %>%
+  unnest(d) %>%
+  mutate(est = est / (1 %#% 'hours')) %>%
+  pivot_wider(names_from = replicates, values_from = est) %>%
+  pivot_longer(cols = as.character(2^(4:9)), values_to = 'est',
+               names_to = 'replicates') %>%
+  mutate(diff_est = est - `1024`,
+         rel_est = est / `1024`,
+         replicates = factor(replicates, levels = 2^(4:9)))
+
+ggplot(all) +
+  facet_wrap(~ replicates) +
+  geom_line(aes(animal, diff_est, group = paste(mean, variance)),
+            alpha = 0.1) +
+  geom_hline(yintercept = 0, color = 'darkorange') +
+  labs(x = 'Time',
+       y = 'Difference in average daily exploration time (hours)')
+```
+
+```{r, echo=FALSE, eval=TRUE, fig.cap="Difference in average time required to reach satiety for each scenario in figure 1 (indicated by each line) and different amounts of tracks (indicated at the top of each facet).
+The differences are relative to simulations with $2^{10}=1024$ tracks for each time point on the x axis (and each scenario in figure 1).
+The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h'}
+knitr::include_graphics('../figures/5-by-5-sensitivity-analysis/time-to-satiety-difference.png',
+                        rel_path = FALSE)
+```
+
+```{r}
+ggplot(all) +
+  facet_wrap(~ replicates) +
+  geom_line(aes(animal, rel_est, group = paste(mean, variance)),
+            alpha = 0.1) +
+  geom_hline(yintercept = 1, color = 'darkorange') +
+  scale_y_continuous(
+    trans = 'log2', breaks = 2^seq(-1, 1, by = 0.5),
+    labels = parse(text = paste0('2^', seq(-1, 1, by = 0.5))))+
+  labs(x = 'Time',
+       y = expression(paste('Relative difference from estimates with',
+                            '1024 replicates'~(log[2]~scale))))
+```
+
+\clearpage
+
+```{r, echo=FALSE, eval=TRUE, fig.cap="Ratios of average time required to reach satiety for each scenario in figure 1 (indicated by each line) and different amounts of tracks (indicated at the top of each facet).
+The ratios are relative to simulations with $2^{10}=1024$ tracks for each time point on the x axis (and each scenario in figure 1).
+The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h'}
+knitr::include_graphics('../figures/5-by-5-sensitivity-analysis/time-to-satiety-rel-difference.png',
+                        rel_path = FALSE)
+```
 
@@
 In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/hr-over-days.png',
-                        rel_path = FALSE)
+knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
 ```
@@
 # switch data to long format with a column of quantile
-sims <- readRDS('H:/GitHub/hr-resource-stoch//simulations/days-hrs.rds') %>%
+sims <- readRDS('H:/GitHub/hr-resource-stoch/simulations/days-hrs.rds') %>%
   tibble() %>%

### 2023-09-11 writing/manuscript.Rmd
renamed sims scripts; added sensitivity analyses
also corrected the number of simulations: only 200 used since HR estimates are stable for n >~ 100

@@
 To support our hypothesis of how organisms' space-use requirements are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
-To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate $2^{10}=1024$ tracks from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
+To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate 200 tracks (see Appendices 2 for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
 The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
@@
 
-Each of the 1024 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
+Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
 Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution.
 The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in figure \@ref(fig:5-5-sims)).
-Note that $t$ was constant within each set of 1024 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
+Note that $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
 Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ with no correlation to the following track.
-Finally, each set of 1024 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution.
+Finally, each set of 200 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution.
 Additional information is provided in Appendix 2.

### 2023-09-11 analysis/figures/delta-t-sensitivity.R

@@ -1,193 +1,0 @@
-library('terra')   # for working with rasters
-library('ctmm')    # for generating movement models
-library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
-library('tidyr')   # for data wrangling (unnest, etc.)
-library('purrr')   # for functional programming (map_*(), etc.)
-library('ggplot2') # for fancy plots
-library('cowplot') # for multi-panel fancy plots
-source('analysis/simulations/movement-model.R') # for consistency between scripts
-theme_set(theme_bw() + # change default ggplot theme
-            theme(legend.position = 'none', strip.background = element_blank(),
-                  strip.text = element_blank()))
-
-# simulate some movement
-get_tracks <- function(.seed) {
-  simulate(model, # ctmm movement model
-           t = 0:(max(SAMPLES)), # 12 hours of data, sampled every second
-           seed = .seed, # set a seed for consistent results
-           complete = TRUE, # prevents missing projection warning
-           crs = PROJECTION) # projection CRS string
-}
-tels <- tibble(seed = 1:3, tel = map(seed, get_tracks)) %>%
-  mutate(seed = factor(seed))
-
-# plot intervals between encounters
-intervals <-
-  transmute(tels,
-            seed,
-            ints = map(tel, \(.tel) {
-              .tel %>%
-                data.frame() %>%
-                mutate(cell_id = cellFromXY(object = HABITAT,
-                                            xy = SpatialPoints.telemetry(.tel)),
-                       new_cell = c(1, diff(cell_id)) != 0) %>%
-                filter(new_cell) %>%
-                mutate(interval = c(NA, diff(t)))
-            })) %>%
-  unnest(ints)
-
-# check the main quantiles
-quantile(intervals$interval, na.rm = TRUE)
-
-# histogram of number of events as by the time interval between events
-p_hist <-
-  ggplot(intervals, aes(interval)) +
-  facet_wrap(~ seed, ncol = 1) +
-  geom_histogram(aes(fill = seed), binwidth = 30, center = 15, lwd = 0.1,
-                 color = 'black', na.rm = TRUE) +
-  scale_fill_viridis_d('Seed', ) +
[…]

### 2023-09-11 analysis/figures/return-sensitivity.R

@@ -1,43 +1,0 @@
-library('ctmm')    # for generating movement models
-library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
-library('purrr')   # for functional programming (map_*(), etc.)
-library('tidyr')   # for data wrangling (pivot_*(), etc.)
-library('ggplot2') # for fancy plots
-source('analysis/simulations/movement-model.R') # for consistency between scripts
-source('functions/add_returns.R') # to add return trips to (0, 0)
-source('functions/get_hr.R') # to extract gaussian HR estimate from spatial variance
-theme_set(theme_bw() + # change default ggplot theme
-            theme(legend.position = 'none', strip.background = element_blank(),
-                  strip.text = element_blank()))
-
-DELTA_T <- 30 # sample every 30 s
-SAMPLES <- seq(0, 60 * 60 * 24, by = DELTA_T) # 24 h sampled every DELTA_T
-
-newd <- tibble(hour = 1:24, # 24 hours
-               t = as.POSIXct(60 * 60 * hour, origin = '1970-01-01'),
-               longitude = 0, # always at (0, 0) at the beginning of each hour
-               latitude = 0,
-               animal = 'test animal') %>%
-  as.telemetry(projection = PROJECTION)
-head(newd)
-sims_loop <- simulate(model, data = newd, dt = DELTA_T, seed = 1, complete = TRUE)
-sims_out <- sims_loop[sims_loop$t %% (60 * 60) < (60 * 30), ] # only first half hours
-sims_out <- sims_out[sims_out$t < (60 * 60 * 24), ] # remove last data point
-
-# check if `sims_loop` is continuous while `sims_out` is intermittent
-plot(rep(1, nrow(sims_loop)) ~ sims_loop$t, pch = '.', ylim = c(-1, 2))
-points(rep(0, nrow(sims_out)) ~ sims_out$t, pch = '.')
-
-fit_ctmm <- function(.data) {
-  tictoc::tic() # start the "stopwatch"
-  theta <- ctmm.guess(.data, interactive = FALSE)
-  model <- ctmm.fit(data = .data, CTMM = theta)
-  tictoc::toc() # print the elapsed time
-  return(model)
-}
-                              # DELTA_t     : 5 min, 1 min,  30 s
-m_loop <- fit_ctmm(sims_loop) # Fitting time:  16 s,  65 s, 168 s
[…]

### 2023-09-11 analysis/simulations/1-tracks.R

@@ -1,0 +1,120 @@
+# does not require large amounts of RAM, takes < 1 minute
+library('ctmm')   # for generating movement models
+library('terra') # for working with raster data
+library('dplyr')  # for data wrangling
+library('purrr')  # for functional programming
+source('functions/label_visits.R') # function for when animal moves to a new cell & eats
+source('analysis/simulations/movement-model.R') # for consistency between scripts
+
+N_DAYS <- 2^10 # number of "days" (i.e., simulations with different seeds)
+SIMULATE_TRACKS <- TRUE # set to "TRUE" if you want to generate new tracks
+
+# extracts simulated tracks from a ctmm movement model for given sample times
+get_tracks <- function(day, times = SAMPLES) {
+  simulate(model, # ctmm movement model
+           t = times, # sampling times in seconds
+           seed = day, # set a seed for consistent results (different track each day)
+           complete = TRUE, # add lat, long, and timestamp to the telemetry object
+           crs = PROJECTION) # CRS projection string
+}
+
+# generate simulated tracks (will be truncated at satiety later)
+if(SIMULATE_TRACKS) {
+  tracks <- tibble(day = 1:N_DAYS, # a simulation for each day
+                   tel = map(.x = day, # set a seed for consistent results
+                             .f = get_tracks)) # function to generate tracks
+  saveRDS(tracks, file = 'simulations/tracks.rds')
+} else tracks <- readRDS('simulations/tracks.rds')
+
+# find patch visits and calories consumed from the tracks
+tracks <- transmute(tracks, # drop tel column
+                    day, # keep day column
+                    track = map(.x = tel, # add a column of full tracks
+                                .f = \(x) label_visits(.tel = x, .habitat = HABITAT)))
+
+# make a single, large tibble
+tracks <- tidyr::unnest(tracks, track)
+saveRDS(tracks, file = 'simulations/labelled-tracks.rds')
+
+# check if grid any animals moved out of the habitat raster
+if(any(is.na(tracks$new_cell))) {
+  warning(paste('CAUTION:', sum(tracks$new_cell), 'animals left the habitat raster!'))
+}
+
+# check scaling is consistent and sampling frequency is sufficiently high
+if(FALSE) {
+  library('ggplot2')
+  theme_set(theme_bw())
+  
[…]

### 2023-09-11 analysis/simulations/2-hr-mean-variance-simulations-days.R

@@ -1,0 +1,134 @@
+# runtime on 1 core with 187 GB of memory: < 3 minutes for 100 tracks, < 20 min for 2^10
+library('ctmm')  # for continuous-time movement modeling
+library('dplyr') # for data wrangling (e.g., %>%)
+library('purrr') # for functional programming (e.g., map(), map_dbl())
+library('tidyr') # for data wrangling (e.g., nested tibbles)
+library('ggplot2') # for fancy plots
+theme_set(theme_bw())
+source('analysis/mean-variance-trends-panel-data.R') # trends in E(R) and V(R)
+source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
+
+tracks <- readRDS('simulations/labelled-tracks.rds') %>% # movement tracks
+  filter(day <= 200) # ~100 are sufficient for stable estimates
+types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic') # trend types
+
+# using for loops to reduce computation costs
+for(i in 1:(length(types))) {
+  mean_fcn <- get(paste0('mean_', types[i]))
+  cat('Running', types[i], 'mean\n')
+  for(j in 1:(length(types))) {
+    cat('  Running', types[j], 'variance\n')
+    var_fcn <- get(paste0('variance_', types[j]))
+    # calculate the number of visits necessary for each animal in each day
+    days <-
+      # tibble of animals and movement dataset
+      d55 %>%
+      filter(mean == types[i], variance == types[j]) %>%
+      mutate(d = list(tracks)) %>%
+      unnest(d) %>% # unnest the datasets so we have a single, large tibble
+      # generate the food for each row from a gamma distribution
+      mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
+             # if the animal visits a new cell, it finds food, otherwise it doesn't
+             food = if_else(new_cell, food, 0)) %>%
+      # end the movement once the animal has reached satiety
+      group_by(day, animal) %>%
+      # calculate the total number of visits, total calories, and if animal is full
+      mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
+             full = satiety >= REQUIRED) %>% # did the animal reach the required needs?
+      filter(full, ! duplicated(full)) %>% # take the 1st row where the animal is full
+      rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
[…]

### 2023-09-11 analysis/simulations/3-hr-mean-variance-simulations-days-summarized.R

@@ -1,0 +1,82 @@
+# runtime on 1 core with 187 GB of memory: ~ 35 min w 100 tracks, ~ 8 h w 1024
+library('ctmm')    # for continuous-time movement modeling
+library('dplyr')   # for data wrangling (e.g., %>%)
+library('purrr')   # for functional programming (e.g., map(), map_dbl())
+library('tidyr')   # for data wrangling (e.g., nested tibbles)
+source('analysis/simulations/movement-model.R') # for consistency btwn scripts
+
+tracks <- readRDS('simulations/tracks.rds') # simulated tracks
+types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic')# trend types
+
+days <- readRDS('simulations/days.rds')
+
+days_summarized <-
+  days %>% # times at which animals reached satiety on each day in each panel
+  # add labelled telemetry tracks for each day
+  left_join(tracks, by = 'day') %>%
+  # filter tracks by when animal became full
+  mutate(tel = map2(tel, t_expl,
+                    \(.t, .max) data.frame(.t) %>% filter(t <= .max))) %>%
+  # group by panel, animal, and day
+  group_by(mean, variance, animal, day) %>%
+  # group by panel and animal (not by day)
+  group_by(mean, variance, animal) %>%
+  mutate(t_start = lag(t_expl * 2), # start at end of previous explor. + return
+         t_start = if_else(is.na(t_start), 0, t_start), # start 1st day at t = 0
+         t_start = cumsum(t_start)) %>%
+  # drop groups
+  ungroup()
+
+#' splitting into two sections bc `unnest()` can't work with very long vectors
+N <- nrow(days_summarized)
+
+days_summarized <-
+  # expand nested datasets into a single big dataset
+  bind_rows(unnest(days_summarized[1:floor(N / 2), ], tel),
+            unnest(days_summarized[(floor(N / 2) + 1):N, ], tel)) %>%
+  mutate(
+    # change t to be the date and time of consecutive days, not "time of day"
+    t = t + t_start,
+    # make timestamp reflect the new values of t
+    timestamp = as.POSIXct(t, tz = 'UTC',
+                           origin = as.POSIXct('1970-01-01 00:00 UTC')),
+    # id for each animal (needed to create a telemetry object)
+    individual.local.identifier = paste(mean, ' mean,',
+                                        variance, 'variance,',
[…]

### 2023-09-11 analysis/simulations/4-hr-mean-variance-simulations-modeling.R

@@ -1,0 +1,58 @@
+# runtime: ~15 minutes
+
+# import trends in means and variances
+source('analysis/mean-variance-trends-panel-data.R')
+
+N <- nrow(d55) # to decrease computation time substantially (>~ 3 times)
+
+# use a for loop to save the models each time
+for(i in 1:N) { # use 1:2 for an initial test
+  # create an R script that fits a movement model to the data
+  cat("setwd('/scratch/st-mnoonan-1/stefano/hr-environ-stoch-phd')",
+      "library('ctmm')    # for continuous-time movement modeling",
+      "library('dplyr')   # for data wrangling (e.g., %>%)",
+      "library('purrr')   # for functional programming (e.g., map())",
+      "ROW <- readRDS('simulations/days-summarized.rds') %>% # import data",
+      paste0("  slice(", i, ") # take the ith row"),
+      "",
+      "# convert data to telemetry and add variogram",
+      "ROW <- mutate(ROW,",
+      "              tel = map(tel, \\(x) suppressMessages(as.telemetry(x))),",
+      "              theta = map(tel, \\(x) ctmm.guess(data = x, interactive = FALSE)))",
+      "",
+      "# fit movement model (OUF)",
+      "ROW[[1, 'model']] <-",
+      paste0("  ctmm.fit(data = ROW[[1, 'tel']][[1]],"),
+      paste0("           CTMM = ROW[[1, 'theta']][[1]],"),
+      "           # cores != 1 doesn't work in Windows",
+      "           control = list(cores = 1)) %>% list() # threading seems to be slower",
+      paste0("saveRDS(ROW, 'analysis/sockeye-scripts/temp-files/days-row-", i, ".rds')"),
+      "",
+      file = paste0('analysis/sockeye-scripts/temp-files/model-', i, '.R'),
+      sep = '\n')
+  
+  # create a PBS script to run the R script
+  cat('#!/bin/bash',
+      '# ------------Sockeye Parameters----------------- #',
+      '#PBS -l walltime=03:00:00,select=1:ncpus=1:mem=187gb', # > 1 h w 100 tracks
+      paste0('#PBS -N m', i, '-', N), # job name
+      '#PBS -A st-mnoonan-1', # allocation
+      # output file if the script runs fully
+      paste0('#PBS -o /scratch/st-mnoonan-1/stefano/hr-environ-stoch-phd/',
[…]

### 2023-09-11 analysis/simulations/5-hr-mean-variance-simulations-hrs.R

@@ -1,0 +1,39 @@
+library('ctmm')  # for continuous-time movement modeling
+library('dplyr') # for data wrangling (e.g., %>%)
+library('purrr') # for functional programming (e.g., map(), map_dbl())
+library('tidyr') # for data wrangling (e.g., nested tibbles)
+source('analysis/mean-variance-trends-panel-data.R') # trends mu and s2
+source('functions/get_hr.R') # for gaussian HR estimate from spatial variance
+
+N <- nrow(d55)
+files <- list.files('analysis/sockeye-scripts/temp-files/',
+                    pattern = 'days-row*')
+
+# check how many models have fit
+length(files)
+
+# ensure all models fit
+length(files) == N
+
+# check which models didn't fit
+which(! (substr(files,
+                nchar('days-row-x'),
+                nchar(files) - nchar('.rds')) %in% 1:N))
+
+days_models <-
+  map_dfr(list.files('analysis/sockeye-scripts/temp-files/',
+                     pattern = 'days-row-*'),
+          \(x) readRDS(paste0('analysis/sockeye-scripts/temp-files/', x)))
+
+days_hrs <-
+  mutate(days_models,
+         # mean exploration length
+         t_expl = map_dbl(tel, \(x) as.numeric(max(x$timestamp))),
+         # position variance
+         pos_var = map_dbl(model, \(x) ctmm:::area.covm(x$sigma)),
+         # calculate the HR as a function of the positional variance
+         hr_50 = get_hr(pos_var, quantile = 0.50),
+         hr_95 = get_hr(pos_var, quantile = 0.95)) %>%
+  select(-c(tel, theta, model))
+
+saveRDS(days_hrs, 'simulations/days-hrs.rds')

### 2023-09-11 analysis/simulations/6-modeling-R-and-hr.R

@@ -1,0 +1,144 @@
+library('dplyr')   # for data wrangling
+library('tidyr')   # for data wrangling
+library('mgcv')    # for empirical Bayes GAMs
+library('ggplot2') # for fancy plots
+library('cowplot') # for fancy multi-panel plots
+source('analysis/figures/default-figure-styling.R') # for color palette
+theme_set(theme_bw() + theme(legend.position = 'none'))
+
+e_r <- 'Resource abundance, E(\U1D445)'
+v_r <- 'Resource stochasticity, Var(\U1D445)'
+
+sims <- readRDS('simulations/days-hrs.rds') %>%
+  tibble() %>%
+  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr') %>%
+  mutate(quantile = factor(quantile))
+
+# sampling is not uniform, but not highly problematic
+# removing the data with constant mu or sigma2 has no effect on the model
+ggplot(sims, aes(mu, sigma2)) +
+  geom_point(alpha = 0.3)
+
+m_1 <- gam(
+  list(
+    # predictor for the mean
+    hr ~ quantile + mu + sigma2,
+    # predictor for the scale (variance = scale * mean)
+    ~ quantile + mu + sigma2),
+  family = gammals(),
+  data = sims)
+
+m_2 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2),
+    ~ quantile + s(mu) + s(sigma2)),
+  family = gammals(),
+  data = sims)
+
+m_3 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5),
+    ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5)),
+  family = gammals(),
+  data = sims)
+
+AIC(m_1, m_2, m_3) # smooth model with interactions is best
+BIC(m_1, m_2, m_3) # smooth model with interactions is best
+
+m <- m_3
+
+plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+
+# estimate predictions ----
+# effect of mu
+newd_mu <-
+  expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
+              sigma2 = mean(sims$sigma2),
+              quantile = unique(sims$quantile))
+
+preds_mu <-
+  bind_cols( # bind new data and predictions
+    newd_mu,
[…]

### 2023-09-11 analysis/simulations/hr-mean-variance-simulations-days-summarized.R

@@ -1,77 +1,0 @@
-# runtime on 1 core with 187 GB of memory: ~ 35 min w 100 tracks, ~ 8 h w 1000 
-library('ctmm')    # for continuous-time movement modeling
-library('dplyr')   # for data wrangling (e.g., %>%)
-library('purrr')   # for functional programming (e.g., map(), map_dbl())
-library('tidyr')   # for data wrangling (e.g., nested tibbles)
-source('analysis/simulations/movement-model.R') # for consistency btwn scripts
-
-tracks <- readRDS('simulations/tracks.rds') # simulated tracks
-types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic')# trend types
-
-days <- readRDS('simulations/days.rds')
-
-cat('Making trips consecutive...\n',
-    file = 'analysis/sockeye-scripts/progress.txt', append = FALSE)
-
-days_summarized <-
-  days %>% # times at which animals reached satiety on each day in each panel
-  # add labelled telemetry tracks for each day
-  left_join(tracks, by = 'day') %>%
-  # filter tracks by when animal became full
-  mutate(tel = map2(tel, t_expl,
-                    \(.t, .max) data.frame(.t) %>% filter(t <= .max))) %>%
-  # group by panel, animal, and day
-  group_by(mean, variance, animal, day) %>%
-  # group by panel and animal (not by day)
-  group_by(mean, variance, animal) %>%
-  mutate(t_start = lag(t_expl * 2), # start at end of previous explor. + return
-         t_start = if_else(is.na(t_start), 0, t_start), # start 1st day at t = 0
-         t_start = cumsum(t_start)) %>%
-  # drop groups
-  ungroup()
-
-#' splitting into two sections bc `unnest()` can't work with very long vectors
-N <- nrow(days_summarized)
-
-days_summarized <-
-  # expand nested datasets into a single big dataset
-  bind_rows(unnest(days_summarized[1:floor(N / 2), ], tel),
-            unnest(days_summarized[(floor(N / 2) + 1):N, ], tel)) %>%
-  mutate(
-    # change t to be the date and time of consecutive days, not "time of day"
-    t = t + t_start,
-    # make timestamp reflect the new values of t
-    timestamp = as.POSIXct(t, tz = 'UTC',
-                           origin = as.POSIXct('1970-01-01 00:00 UTC')),
-    # id for each animal (needed to create a telemetry object)
-    individual.local.identifier = paste(mean, ' mean,',
[…]

### 2023-09-11 analysis/simulations/hr-mean-variance-simulations-days.R

@@ -1,83 +1,0 @@
-# runtime on 1 core with 187 GB of memory: < 3 minutes for 100 tracks, < 20 min for 1e3
-library('ctmm')  # for continuous-time movement modeling
-library('dplyr') # for data wrangling (e.g., %>%)
-library('purrr') # for functional programming (e.g., map(), map_dbl())
-library('tidyr') # for data wrangling (e.g., nested tibbles)
-source('analysis/figures/mean-variance-trends-panel-data.R') # trends in E(U) and V(U)
-source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-
-tracks <- readRDS('simulations/labelled-tracks.rds') %>% # movement tracks
-  filter(day <= 200) # 100 are sufficient for stable estimates
-types <- c('constant', 'linear', 'cyclical', 'drifting', 'erratic') # trend types
-
-# using for loops to reduce computation costs
-for(i in 1:(length(types))) {
-  mean_fcn <- get(paste0('mean_', types[i]))
-  for(j in 1:(length(types))) {
-    var_fcn <- get(paste0('variance_', types[j]))
-    # calculate the number of visits necessary for each animal in each day
-    days <-
-      # tibble of animals and movement dataset
-      d55 %>%
-      filter(mean == types[i], variance == types[j]) %>%
-      mutate(d = list(tracks)) %>%
-      unnest(d) %>% # unnest the datasets so we have a single, large tibble
-      # generate the food for each row from a gamma distribution
-      mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
-             # if the animal visits a new cell, it finds food, otherwise it doesn't
-             food = if_else(new_cell, food, 0)) %>%
-      # end the movement once the animal has reached satiety
-      group_by(day, animal) %>%
-      # calculate the total number of visits, total calories, and if animal is full
-      mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
-             full = satiety >= REQUIRED) %>% # did the animal reach the required needs?
-      filter(full, ! duplicated(full)) %>% # take the 1st row where the animal is full
-      rename(t_expl = t) %>% # to avoid duplicated colnames with tracks
-      # remove unneeded columns (also avoids duplicated colnames with tracks)
-      dplyr::select(-c(x, y, vx, vy, timestamp, longitude, latitude, food))
-    
-    # save the simulation
-    saveRDS(days,
[…]

### 2023-09-11 analysis/simulations/hr-mean-variance-simulations-hrs.R

@@ -1,39 +1,0 @@
-library('ctmm')  # for continuous-time movement modeling
-library('dplyr') # for data wrangling (e.g., %>%)
-library('purrr') # for functional programming (e.g., map(), map_dbl())
-library('tidyr') # for data wrangling (e.g., nested tibbles)
-source('analysis/figures/mean-variance-trends-panel-data.R') # trends mu and s2
-source('functions/get_hr.R') # for gaussian HR estimate from spatial variance
-
-N <- nrow(d55)
-files <- list.files('analysis/sockeye-scripts/temp-files/',
-                    pattern = 'days-row*')
-
-# check how many models have fit
-length(files)
-
-# ensure all models fit
-length(files) == N
-
-# check which models didn't fit
-which(! (substr(files,
-                nchar('days-row-x'),
-                nchar(files) - nchar('.rds')) %in% 1:N))
-
-days_models <-
-  map_dfr(list.files('analysis/sockeye-scripts/temp-files/',
-                     pattern = 'days-row-*'),
-          \(x) readRDS(paste0('analysis/sockeye-scripts/temp-files/', x)))
-
-days_hrs <-
-  mutate(days_models,
-         # mean exploration length
-         t_expl = map_dbl(tel, \(x) as.numeric(max(x$timestamp))),
-         # position variance
-         pos_var = map_dbl(model, \(x) ctmm:::area.covm(x$sigma)),
-         # calculate the HR as a function of the positional variance
-         hr_50 = get_hr(pos_var, quantile = 0.50),
-         hr_95 = get_hr(pos_var, quantile = 0.95)) %>%
-  select(-c(tel, theta, model))
-
-saveRDS(days_hrs, 'simulations/days-hrs.rds')

### 2023-09-11 analysis/simulations/hr-mean-variance-simulations-modeling.R

@@ -1,58 +1,0 @@
-# runtime: ~15 minutes
-
-# import trends in means and variances
-source('analysis/figures/mean-variance-trends-panel-data.R')
-
-N <- nrow(d55) # to decrease computation time substantially (>~ 3 times)
-
-# use a for loop to save the models each time
-for(i in 1:N) { # use 1:2 for an initial test
-  # create an R script that fits a movement model to the data
-  cat("setwd('/scratch/st-mnoonan-1/stefano/hr-environ-stoch-phd')",
-      "library('ctmm')    # for continuous-time movement modeling",
-      "library('dplyr')   # for data wrangling (e.g., %>%)",
-      "library('purrr')   # for functional programming (e.g., map())",
-      "ROW <- readRDS('simulations/days-summarized.rds') %>% # import data",
-      paste0("  slice(", i, ") # take the ith row"),
-      "",
-      "# convert data to telemetry and add variogram",
-      "ROW <- mutate(ROW,",
-      "              tel = map(tel, \\(x) suppressMessages(as.telemetry(x))),",
-      "              theta = map(tel, \\(x) ctmm.guess(data = x, interactive = FALSE)))",
-      "",
-      "# fit movement model (OUF)",
-      "ROW[[1, 'model']] <-",
-      paste0("  ctmm.fit(data = ROW[[1, 'tel']][[1]],"),
-      paste0("           CTMM = ROW[[1, 'theta']][[1]],"),
-      "           # cores != 1 doesn't work in Windows",
-      "           control = list(cores = 1)) %>% list() # threading seems to be slower",
-      paste0("saveRDS(ROW, 'analysis/sockeye-scripts/temp-files/days-row-", i, ".rds')"),
-      "",
-      file = paste0('analysis/sockeye-scripts/temp-files/model-', i, '.R'),
-      sep = '\n')
-  
-  # create a PBS script to run the R script
-  cat('#!/bin/bash',
-      '# ------------Sockeye Parameters----------------- #',
-      '#PBS -l walltime=03:00:00,select=1:ncpus=1:mem=187gb', # > 1 h w 100 tracks
-      paste0('#PBS -N m', i, '-', N), # job name
-      '#PBS -A st-mnoonan-1', # allocation
-      # output file if the script runs fully
-      paste0('#PBS -o /scratch/st-mnoonan-1/stefano/hr-environ-stoch-phd/',
[…]

### 2023-09-11 analysis/simulations/hr-simulation-extreme-scenarios.R

@@ -1,172 +1,0 @@
-library('ctmm')    # for continuous-time movement modeling
-library('dplyr')   # for data wrangling (e.g., %>%)
-library('tidyr')   # for data wrangling (e.g., nested tibbles)
-library('ggplot2') # for fancy plots
-source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/mean-variance-trends-panel-data.R') # means & variances
-source('analysis/simulations/movement-model.R') # for consistency across scripts
-source('functions/get_hr.R') # for extracting gaussian home range
-theme_set(theme_bw())
-set.seed(1) # for consistent results
-
-tels <- readRDS('simulations/tracks.rds') # list of telemetry tracks
-tracks <- readRDS('simulations/labelled-tracks.rds') # tibble of tracks
-MAX_T <- max(tracks$t) # maximum amount of exploration time
-
-WORST <- filter(d55, mu == min(mu)) %>% # lowest mean resources
-  filter(sigma2 == max(sigma2)) %>% # with highest variance
-  slice(1) # take the first row only
-BEST <- filter(d55, mu == max(mu)) %>% # highest mean resources
-  filter(sigma2 == min(sigma2)) %>% # with lowest variance
-  slice(1) # take the first row only
-
-days <-
-  # modify WORST and BEST to follow the syntax used in 'offspring-simulations.R'
-  transmute(bind_rows(WORST, BEST),
-            animal,
-            mu,
-            sigma2,
-            d = list(tracks),
-            scenario = c('Worst case', 'Best case')) %>%
-  unnest(d) %>% # unnest the datasets so we have a single, large tibble
-  select(-timestamp) %>%
-  # generate the food for each row from a gamma distribution
-  mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
-         # if the animal visits a new cell, it finds food, otherwise it doesn't
-         food = if_else(new_cell, food, 0)) %>%
-  # end the movement once the animal has reached satiety
-  group_by(day, animal, scenario) %>%
-  # calculate the total number of visits, total calories, and if animal is full
-  mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
-         full = satiety >= REQUIRED) %>% # did the animal reach the its needs?
-  filter(cumsum(full) <= 1) %>% # full only once
-  ungroup()
-
[…]

### 2023-09-11 analysis/simulations/modeling-R-and-hr.R

@@ -1,144 +1,0 @@
-library('dplyr')   # for data wrangling
-library('tidyr')   # for data wrangling
-library('mgcv')    # for empirical Bayes GAMs
-library('ggplot2') # for fancy plots
-library('cowplot') # for fancy multi-panel plots
-source('analysis/figures/default-figure-styling.R') # for color palette
-theme_set(theme_bw() + theme(legend.position = 'none'))
-
-e_r <- 'Resource abundance, E(\U1D445)'
-v_r <- 'Resource stochasticity, Var(\U1D445)'
-
-sims <- readRDS('simulations/days-hrs.rds') %>%
-  tibble() %>%
-  pivot_longer(c(hr_50, hr_95), names_to = 'quantile', values_to = 'hr') %>%
-  mutate(quantile = factor(quantile))
-
-# sampling is not uniform, but not highly problematic
-# removing the data with constant mu or sigma2 has no effect on the model
-ggplot(sims, aes(mu, sigma2)) +
-  geom_point(alpha = 0.3)
-
-m_1 <- gam(
-  list(
-    # predictor for the mean
-    hr ~ quantile + mu + sigma2,
-    # predictor for the scale (variance = scale * mean)
-    ~ quantile + mu + sigma2),
-  family = gammals(),
-  data = sims)
-
-m_2 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(sigma2),
-    ~ quantile + s(mu) + s(sigma2)),
-  family = gammals(),
-  data = sims)
-
-m_3 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5),
-    ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5)),
-  family = gammals(),
-  data = sims)
-
-AIC(m_1, m_2, m_3) # smooth model with interactions is best
-BIC(m_1, m_2, m_3) # smooth model with interactions is best
-
-m <- m_3
-
-plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
-
-# estimate predictions ----
-# effect of mu
-newd_mu <-
-  expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
-              sigma2 = mean(sims$sigma2),
-              quantile = unique(sims$quantile))
-
-preds_mu <-
-  bind_cols( # bind new data and predictions
-    newd_mu,
[…]

### 2023-09-11 analysis/simulations/sensitivity-analyses/1a-return-sensitivity.R

@@ -1,0 +1,43 @@
+library('ctmm')    # for generating movement models
+library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
+library('purrr')   # for functional programming (map_*(), etc.)
+library('tidyr')   # for data wrangling (pivot_*(), etc.)
+library('ggplot2') # for fancy plots
+source('analysis/simulations/movement-model.R') # for consistency between scripts
+source('functions/add_returns.R') # to add return trips to (0, 0)
+source('functions/get_hr.R') # to extract gaussian HR estimate from spatial variance
+theme_set(theme_bw() + # change default ggplot theme
+            theme(legend.position = 'none', strip.background = element_blank(),
+                  strip.text = element_blank()))
+
+DELTA_T <- 30 # sample every 30 s
+SAMPLES <- seq(0, 60 * 60 * 24, by = DELTA_T) # 24 h sampled every DELTA_T
+
+newd <- tibble(hour = 1:24, # 24 hours
+               t = as.POSIXct(60 * 60 * hour, origin = '1970-01-01'),
+               longitude = 0, # always at (0, 0) at the beginning of each hour
+               latitude = 0,
+               animal = 'test animal') %>%
+  as.telemetry(projection = PROJECTION)
+head(newd)
+sims_loop <- simulate(model, data = newd, dt = DELTA_T, seed = 1, complete = TRUE)
+sims_out <- sims_loop[sims_loop$t %% (60 * 60) < (60 * 30), ] # only first half hours
+sims_out <- sims_out[sims_out$t < (60 * 60 * 24), ] # remove last data point
+
+# check if `sims_loop` is continuous while `sims_out` is intermittent
+plot(rep(1, nrow(sims_loop)) ~ sims_loop$t, pch = '.', ylim = c(-1, 2))
+points(rep(0, nrow(sims_out)) ~ sims_out$t, pch = '.')
+
+fit_ctmm <- function(.data) {
+  tictoc::tic() # start the "stopwatch"
+  theta <- ctmm.guess(.data, interactive = FALSE)
+  model <- ctmm.fit(data = .data, CTMM = theta)
+  tictoc::toc() # print the elapsed time
+  return(model)
+}
+                              # DELTA_t     : 5 min, 1 min,  30 s
+m_loop <- fit_ctmm(sims_loop) # Fitting time:  16 s,  65 s, 168 s
[…]

### 2023-09-11 analysis/simulations/sensitivity-analyses/1b-delta-t-sensitivity.R

@@ -1,0 +1,193 @@
+library('terra')   # for working with rasters
+library('ctmm')    # for generating movement models
+library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
+library('tidyr')   # for data wrangling (unnest, etc.)
+library('purrr')   # for functional programming (map_*(), etc.)
+library('ggplot2') # for fancy plots
+library('cowplot') # for multi-panel fancy plots
+source('analysis/simulations/movement-model.R') # for consistency between scripts
+theme_set(theme_bw() + # change default ggplot theme
+            theme(legend.position = 'none', strip.background = element_blank(),
+                  strip.text = element_blank()))
+
+# simulate some movement
+get_tracks <- function(.seed) {
+  simulate(model, # ctmm movement model
+           t = 0:(max(SAMPLES)), # 12 hours of data, sampled every second
+           seed = .seed, # set a seed for consistent results
+           complete = TRUE, # prevents missing projection warning
+           crs = PROJECTION) # projection CRS string
+}
+tels <- tibble(seed = 1:3, tel = map(seed, get_tracks)) %>%
+  mutate(seed = factor(seed))
+
+# plot intervals between encounters
+intervals <-
+  transmute(tels,
+            seed,
+            ints = map(tel, \(.tel) {
+              .tel %>%
+                data.frame() %>%
+                mutate(cell_id = cellFromXY(object = HABITAT,
+                                            xy = SpatialPoints.telemetry(.tel)),
+                       new_cell = c(1, diff(cell_id)) != 0) %>%
+                filter(new_cell) %>%
+                mutate(interval = c(NA, diff(t)))
+            })) %>%
+  unnest(ints)
+
+# check the main quantiles
+quantile(intervals$interval, na.rm = TRUE)
+
+# histogram of number of events as by the time interval between events
+p_hist <-
+  ggplot(intervals, aes(interval)) +
+  facet_wrap(~ seed, ncol = 1) +
+  geom_histogram(aes(fill = seed), binwidth = 30, center = 15, lwd = 0.1,
+                 color = 'black', na.rm = TRUE) +
+  scale_fill_viridis_d('Seed', ) +
[…]

### 2023-09-11 analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R

@@ -1,0 +1,172 @@
+library('ctmm')    # for continuous-time movement modeling
+library('dplyr')   # for data wrangling (e.g., %>%)
+library('tidyr')   # for data wrangling (e.g., nested tibbles)
+library('ggplot2') # for fancy plots
+source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
+source('analysis/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/simulations/movement-model.R') # for consistency across scripts
+source('functions/get_hr.R') # for extracting gaussian home range
+theme_set(theme_bw())
+set.seed(1) # for consistent results
+
+tels <- readRDS('simulations/tracks.rds') # list of telemetry tracks
+tracks <- readRDS('simulations/labelled-tracks.rds') # tibble of tracks
+MAX_T <- max(tracks$t) # maximum amount of exploration time
+
+WORST <- filter(d55, mu == min(mu)) %>% # lowest mean resources
+  filter(sigma2 == max(sigma2)) %>% # with highest variance
+  slice(1) # take the first row only
+BEST <- filter(d55, mu == max(mu)) %>% # highest mean resources
+  filter(sigma2 == min(sigma2)) %>% # with lowest variance
+  slice(1) # take the first row only
+
+days <-
+  # modify WORST and BEST to follow the syntax used in 'offspring-simulations.R'
+  transmute(bind_rows(WORST, BEST),
+            animal,
+            mu,
+            sigma2,
+            d = list(tracks),
+            scenario = c('Worst case', 'Best case')) %>%
+  unnest(d) %>% # unnest the datasets so we have a single, large tibble
+  select(-timestamp) %>%
+  # generate the food for each row from a gamma distribution
+  mutate(food = rgamma2(mu = mu, sigma2 = sigma2, N = n()),
+         # if the animal visits a new cell, it finds food, otherwise it doesn't
+         food = if_else(new_cell, food, 0)) %>%
+  # end the movement once the animal has reached satiety
+  group_by(day, animal, scenario) %>%
+  # calculate the total number of visits, total calories, and if animal is full
+  mutate(satiety = cumsum(food), # use for diagnostics if animals don't get full
+         full = satiety >= REQUIRED) %>% # did the animal reach the its needs?
+  filter(cumsum(full) <= 1) %>% # full only once
+  ungroup()
+
[…]

### 2023-09-11 analysis/simulations/tracks.R

@@ -1,120 +1,0 @@
-# does not require large amounts of RAM, takes < 1 minute
-library('ctmm')   # for generating movement models
-library('terra') # for working with raster data
-library('dplyr')  # for data wrangling
-library('purrr')  # for functional programming
-source('functions/label_visits.R') # function for when animal moves to a new cell & eats
-source('analysis/simulations/movement-model.R') # for consistency between scripts
-
-N_DAYS <- 2^10 # number of "days" (i.e., simulations with different seeds)
-SIMULATE_TRACKS <- TRUE # set to "TRUE" if you want to generate new tracks
-
-# extracts simulated tracks from a ctmm movement model for given sample times
-get_tracks <- function(day, times = SAMPLES) {
-  simulate(model, # ctmm movement model
-           t = times, # sampling times in seconds
-           seed = day, # set a seed for consistent results (different track each day)
-           complete = TRUE, # add lat, long, and timestamp to the telemetry object
-           crs = PROJECTION) # CRS projection string
-}
-
-# generate simulated tracks (will be truncated at satiety later)
-if(SIMULATE_TRACKS) {
-  tracks <- tibble(day = 1:N_DAYS, # a simulation for each day
-                   tel = map(.x = day, # set a seed for consistent results
-                             .f = get_tracks)) # function to generate tracks
-  saveRDS(tracks, file = 'simulations/tracks.rds')
-} else tracks <- readRDS('simulations/tracks.rds')
-
-# find patch visits and calories consumed from the tracks
-tracks <- transmute(tracks, # drop tel column
-                    day, # keep day column
-                    track = map(.x = tel, # add a column of full tracks
-                                .f = \(x) label_visits(.tel = x, .habitat = HABITAT)))
-
-# make a single, large tibble (will need lots of RAM)
-tracks <- tidyr::unnest(tracks, track)
-saveRDS(tracks, file = 'simulations/labelled-tracks.rds')
-
-# check if grid any animals moved out of the habitat raster
-if(any(is.na(tracks$new_cell))) {
-  warning(paste('CAUTION:', sum(tracks$new_cell), 'animals left the habitat raster!'))
-}
-
-# check scaling is consistent and sampling frequency is sufficiently high
-if(FALSE) {
-  library('ggplot2')
-  theme_set(theme_bw())
-  
[…]

### 2023-09-12 writing/appendix-2-simulations.Rmd
abbreviated figure(s) to fig(s).

@@
 This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript.
-For ease of reference, we also include the figures here (figures \@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
+For ease of reference, we also include the figures here (figs.
+\@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
 To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
@@
 ```{r, echo=FALSE, eval=TRUE, fig.cap="Ratios of average time required to reach satiety for each scenario in figure 1 (indicated by each line) and different amounts of tracks (indicated at the top of each facet).
-The ratios are relative to simulations with $2^{10}=1024$ tracks for each time point on the x axis (and each scenario in figure 1).
+The ratios are relative to simulations with $2^{10}=1024$ tracks for each time point on the x axis (and each scenario in fig.
+1).
 The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h'}
@@
 
-Figure \@ref(fig:gamls-terms) shows each of the terms from model 3.
+Fig.
+\@ref(fig:gamls-terms) shows each of the terms from model 3.
 The estimated effect of `quantile` (see the output of `summary()` above) supports the relationship between the gaussian HR estimates detailed above, since $\exp(1.464) = 4.323218 \approx \log_{0.5}(0.05),$ but there is no differences in the scale parameter between the two quantiles.
@@
 
-Finally, we can use the model to recreate figure \@ref(fig:5-5-reg) and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
+Finally, we can use the model to recreate fig. \@ref(fig:5-5-reg) and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
 

### 2023-09-12 writing/appendix-3-empirical-modeling.Rmd
abbreviated figure(s) to fig(s).

@@
 The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-For ease of reference, we also include the figure below (figure \@ref(fig:tapir-mw)).
-Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement), figure \@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figures \@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figures \@ref(fig:tapir-mw)d-f).
+For ease of reference, we also include the figure below (fig.
+\@ref(fig:tapir-mw)).
+Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement), fig.
+\@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figs.
+\@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figs.
+\@ref(fig:tapir-mw)d-f).
 
@@
 For each set of days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019].
-Finally, it saves an exploratory figure (figure \@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
+Finally, it saves an exploratory figure (fig.
+\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
 
@@
 We fit the beta GAMLS via the `mgcv` package (note `family = betals()` in the code chunk below).
-The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see figure \@ref(fig:gamls-terms)).
+The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see fig.
+\@ref(fig:gamls-terms)).
 The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.
@@
 While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial.
-Excessively smooth terms for the mean can inflate the scale term (figure \@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (figure \@ref(fig:oversmooth)).
+Excessively smooth terms for the mean can inflate the scale term (fig.
+\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig.
+\@ref(fig:oversmooth)).
 Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
@@
 To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.
-However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see figure \@ref(fig:hr-gam)).
+However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig.
+\@ref(fig:hr-gam)).
 

### 2023-09-12 writing/manuscript.Rmd
addressed Mike's comments

@@
   - name: Stefano Mezzini
-    email: stefano.mezzini@ubc.ca
     institute: [braes, biol]
@@
     institute: [braes, biol, cmps]
+    correspondence: "yes"
 institute:
@@
     - \usepackage{hanging} # for hanging indents in references
+    - \usepackage[left]{lineno} # for line numbers
 subparagraph: true # needed for \usepackage[small]{titlesec}
@@
     pandoc_args:
+# https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
       - '--lua-filter=scholarly-metadata.lua'
@@
 \newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->
+
+<!-- place authors' emails on a new page -->
+
+\newpage
+
+\noindent \textbf{Authors' emails:}
+
+\noindent Stefano Mezzini: stefano.mezzini@ubc.ca
+
+\noindent Simon Wood: simon.wood@bath.edu
+
+\noindent Chris H. Fleming: chris.h.fleming@gmail.com
+
+\noindent E. Patrícia Medici: medici@ipe.org.br
+
+\noindent Michael J. Noonan: michael.noonan@ubc.ca
 
@@
 
-knitr::knit_hooks$set(plot = function(x, options) {  # set the new plot hook ...
+knitr::knit_hooks$set(plot = function(x, options) { # set the new plot hook...
 
@@
 
-  # if wrapfigure options are given, create the new opening string for the wrapfigure
-  # environment using custom figure alignment and text width
+  # if wrapfigure options are given, create the new opening string for the
+  # wrapfigure environment using custom figure alignment and text width
   if(!is.null(options$wrapfigure)) {
@@
 
+\linenumbers
+
 # Abstract
@@
 This is not a trivial oversight.
-Organisms adjust to variable environmental conditions, so failing to account for the effects of resource unpredictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
-In this study, we first review the literature of studies that estimate animals' space-use requirements as a function of resource abundance.
+Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
+In this study, we first review the literature of studies that estimate organisms' space-use requirements as a function of resource abundance.
 We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use.
@@
 Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
+
+\vspace*{\fill}
+
+\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`.
 
@@
 
-The relationship between organisms' movement and resource abundance has been of interest to biologists for over half a century.
+The relationship between organisms' movement and resource abundance has been of interest to biologists for nearly a century.
 @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
@@
 
-\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption, and $R$ is the **resources** the organism can access.
-Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but a species' spatial requirements likely depend on numerous other factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
+\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the **resources** the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
+Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
 While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
-Instead, results are reported as independent, novel findings.
+Instead, results are typically reported as independent, novel findings.
 Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements.
@@
 
-In this paper, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
+Here, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
@@
 Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for.
-Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is little empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-Here, we illustrate how organisms' space-use requirements depend on both the abundance and unpredictability of resources.
+Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource *abundance* on organisms' space-use requirements while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space-use requirements while suggesting a second unifying hypothesis.
+Next, we provide a review of the effects of resource *abundance* on organisms' space use while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space use while suggesting a second unifying hypothesis.
 Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
@@
 
-# Simulating resources as a random variable
+# Resources as a random variable
 
-\noindent In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
+\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
+Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
+Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
+Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time.
+Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
+
+In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
 Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
@@
 This assumption also holds biologically, since resources tend to be less variable when they are less abundant.
-We assume readers are familiar with the concepts of: random variables, probability distributions, the expected value of a random variable, and the variance of a random variable.
-We define and explain each of the four concepts in Appendix 1 for readers who need additional information.
+We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix 1.
 
 ## Effects of resource abundance, $\e(R)$
-
-```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
-Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
-```
 
@@
 This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see figure \@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see fig.
+\@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+
+```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$).
+We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
+As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
+Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
+knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
+```
 
@@
 
+It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
+Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
+Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
+\@ref(fig:mu-hyp).
+
 ```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
@@
 (b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
 knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
@@
 
-It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
-Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in figure \@ref(fig:mu-hyp).
-
-Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see figure \@ref(fig:multipanel-constant-var)a).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (figure \@ref(fig:multipanel-constant-var)b).
+Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig.
+\@ref(fig:multipanel-constant-var)a).
+We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig.
+\@ref(fig:multipanel-constant-var)b).
 Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
@@
 As $\\var(R)$ increases, space-use requirements should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
+Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth'}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.align = 'right', wrapfigure = list('R', 0.5)}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
-
 Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' space-use requirements.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (figure \@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig.
+\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
@@
 
-Like the scenario illustrated in figure \@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
-We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (figure \@ref(fig:multipanel-changing-var)).
+Like the scenario illustrated in fig.
+\@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
+We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig.
+\@ref(fig:multipanel-changing-var)).
 When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
@@
 Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
+
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
+(a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
+(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
+The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
+knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
+```
 
@@
 However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and be more likely to be nomadic, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
-Likewise, @herfindal_prey_2005 found that the home ranges Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
+Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
+Likewise, @herfindal_prey_2005 found that the home ranges of Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
 A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
@@
 
-To support our hypothesis of how organisms' space-use requirements are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+\noindent To support our hypothesis of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate 200 tracks (see Appendices 2 for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
@@
 Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution.
-The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in figure \@ref(fig:5-5-sims)).
+The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig.
+\@ref(fig:5-5-sims)).
 Note that $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
@@
 Additional information is provided in Appendix 2.
-
-Figure \@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$.
-The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in figure \@ref(fig:multipanel-constant-var)).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
-As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
 
@@
 
+Fig.
+\@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$.
+The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
+\@ref(fig:multipanel-constant-var)).
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
+As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
+
 In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
@@
 
-The leftmost column of figure \@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
+The leftmost column of fig.
+\@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
@@
 
-The remaining panels in figure \@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationships can be.
+The remaining panels in fig.
+\@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationship can be.
 Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
@@
 
-Not all 25 scenarios depicted in figure \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on space use are useful examples that can be thought of as simplified scenarios.
+Not all 25 scenarios depicted in fig.
+\@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on space use are useful examples that can be thought of as simplified scenarios.
 $\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
@@
 
-Although the temporal trends in figure \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
-Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (figure \@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
-We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of figure \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
+Although the temporal trends in fig.
+\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
+Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (fig.
+\@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
+We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of fig.
+\@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
 Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
@@
 
-The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., figure \@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
-We expect the two hypotheses we present here (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
+The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., fig.
+\@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
+We expect the two hypotheses we present here (figs.
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
 In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as new findings.
@@
 Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
-Individuals and populations that were once thought to have different space-use requirements due to differences in behavior may instead be simply responding to different local conditions along the functions in figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
+Individuals and populations that were once thought to have different space-use requirements due to differences in behavior may instead be simply responding to different local conditions along the functions in figs.
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
 In the following sections, we illustrate how one can apply this framework to empirical data.
@@
 `mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families (see the help on `?mgcv::family.mgcv` and `?mgcv::twlss` for more information).
+Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
 
@@
 
-As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (figure \@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
+As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (fig.
+\@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
 If there is not sufficient data to fit a location-scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
@@
 
-\noindent Figure \@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
+\noindent Fig.
+\@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
 Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period.
@@
 Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figures \@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (figure \@ref(fig:5-5-reg)).
+Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figs.
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (fig.
+\@ref(fig:5-5-reg)).
 Thus, while the time series shown in panel **c** may appear complicated, the tapir's response can be well described by two simple functions.
+However, note that since $\mu(t)$ and $\sigma^2(t)$ are highly corrleated, modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect.
+By including the effect of $\sigma^2(t)$ in the model, we show that the tapir responds to both resource abundance and stochasticity.
+When resources are relatively low (but still high: $\mu(t) \in (0.65, 0.75)$), the tapir does not respond to changes in resources abundance.
+However, duing periods of peak greenness ($\mu \approx 0.85$), its weekly home range is approximately halved.
+In contrast, the tapir adapts much more strongly to changes in $\sigma^2(t)$ by increasing its home range sixfold.
+This example demonstrates that (1) even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in periods of high stochasticity and that (2) the effects of stochasticity in $R$ should also be considered when estimating the effects of average $R$ on animals' space use.
 
@@
 
-We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript despite not being authors, namely: Aimee Chhen, Lauren Mills, and Sarah Wyse.
+\noindent We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
+In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely: Aimee Chhen, Kim Hinz, Lauren Mills, and Sarah Wyse.
+This work was supported by the NSERC Discovery Grant RGPIN-2021-02758 to MJN and funds from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, Mitacs, and BC Parks.
+
+# Code and data availability
+
+All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of the tapir data, which is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
+
+# Conflict of interest
+
+\noindent The authors declare there are no conflicts of interest.
 

### 2023-09-13 writing/appendix-3-empirical-modeling.Rmd
fixed k typo

@@
     ndvi ~ # not scaling because range is in (0, 1)
-      s(long, lat, bs = 'ds', k = 20) + # mean over space (uses k^2 df)
+      s(long, lat, bs = 'ds', k = 20) + # mean over space
       s(dec_date, bs = 'tp', k = 20),   # high k to account for adaptions
@@
     ~
-      s(long, lat, bs = 'ds', k = 10) + # scale over space (uses k^2 df)
+      s(long, lat, bs = 'ds', k = 10) + # scale over space
       s(dec_date, bs = 'tp', k = 10)),  # scale over time
@@
 If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.
-Additionally, note that computation time depends strongly on the complexity of the two-dimensional spatial terms, since changes in `k` alter the size of the term quadratically (since the total size is $\tt k \times \tt k = \tt k^2$).
 

### 2023-09-13 writing/appendix-3-empirical-modeling.Rmd
corrected comment on reml and gcv

@@
     data = anna_ndvi,
-    method = 'REML') # performs better than GCV (the default)
+    method = 'REML') # REstricted Maximum Likelihood 
 ```

### 2023-09-14 analysis/tapirus-terrestris/anna-betals.R
added script for betals family

@@ -4,7 +4,7 @@ library('ggplot2')   # for fancy plots
 library('mgcv')      # for GAMs
 library('lubridate') # for smoother date wrangling
 
-source('../ndvi-stochasticity/functions/betals.r') # beta location-scale fam 
+source('functions/betals.r') # beta location-scale fam 
 source('analysis/figures/default-figure-styling.R') # for NDVI color palette
 
 # import 95% home range estimates with no CIs (optional)

### 2023-09-17 analysis/figures/moving-window-figure-tapirs.R
tested if adding tapirs made the story clearer

@@ -16,7 +16,7 @@ v_r <- 'Resource stochasticity, \U1D70E\U00B2(t)'
 hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
 
 # import location-scale model for predictions of mean and variance in NDVI
-m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
+m_ndvi <- readRDS('models/tapirs/tapirs-ndvi-betals.rds')
 
 # function to make predictions from the model for each telemetry
 ndvi_preds <- function(.data) {
@@ -31,52 +31,49 @@ ndvi_preds <- function(.data) {
 }
 
 # import tapir data (from https://doi.org/10.1186/s40462-022-00313-w)
-tapir <- readRDS('../tapirs/models/tapirs-final.rds') %>%
-  filter(name.short == 'ANNA')
-
-tel <- data.frame(tapir$data[[1]]) %>%
+tel <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+  filter(region == 'cerrado') %>%
+  pull(data) %>%
+  map_dfr(data.frame) %>%
   rename(long = longitude, lat = latitude) %>%
   mutate(dec_date = decimal_date(timestamp)) %>%
   bind_cols(., ndvi_preds(.)) %>% 
   as_tibble()
 
-if(! file.exists('data/anna-hr-ndvi-data.rds')) {
-  tapir <-
-    readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+if(! file.exists('data/tapirs-hr-ndvi-data.rds')) {
+  tapirs <-
+    readRDS('data/cerrado-hrs.rds') %>%
     mutate(est = map(dataset, \(.d) filter(tel, timestamp %in% .d$timestamp)),
            mu = map_dbl(est, \(.d) mean(.d$mu)),
            sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
-    select(t_center, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
-    pivot_longer(c(hr_lwr_95, hr_est_95, hr_upr_95), names_to = c('.value', 'quantile'),
-                 names_pattern = '(.+)_(.+)') %>%
-    mutate(t_center = as.POSIXct(t_center),
-           quantile = paste0(quantile, '%'))
[…]

### 2023-09-17 analysis/tapirus-terrestris/tapirs-ndvi-betals.R

@@ -43,26 +43,26 @@ n_distinct(d$dec_date)
 # run a test model
 m_test <-
   bam(
-    (ndvi + 1) / 2 ~ # not scaling because range is in (0, 1)
-      s(long, lat, bs = 'ds', k = 20) + # mean over space
+    (ndvi + 1) / 2 ~ # scale from [-1, 1] to [0, 1]
+      s(long, lat, bs = 'ds', k = 300) + # mean over space
       s(dec_date, bs = 'tp', k = 10), # need high k to account for animal adapting
     family = betar(link = 'logit'),
     data = d,
     method = 'fREML',
     discrete = TRUE,
     control = gam.control(trace = TRUE))
-plot(m_test, pages = 1, n = 400, scheme = 3)
+plot(m_test, pages = 1, n = 400, n2 = 200, scheme = 3)
 summary(m_test)
 
 m_ndvi <-
   gam(list(
     # mean predictor
-    ndvi ~ # not scaling because range is in (0, 1)
-      s(long, lat, bs = 'ds', k = 50) + # mean over space
-      s(dec_date, bs = 'tp', k = 10), # need high k to account for animal adapting
+    (ndvi + 1) / 2 ~ # scale from [-1, 1] to [0, 1]
+      s(long, lat, bs = 'ds', k = 300) + # mean over space
+      s(dec_date, bs = 'tp', k = 30), # need high k to account for animal adapting
     # scale predictor (sigma2 = mu * (1 - mu) * scale)
     ~
-      s(long, lat, bs = 'ds', k = 25) +
+      s(long, lat, bs = 'ds', k = 150) +
       s(dec_date, bs = 'tp', k = 10)),
     family = betals(),
     data = d,
@@ -72,5 +72,5 @@ m_ndvi <-
 saveRDS(m_ndvi, file = 'models/tapirs/tapirs-ndvi-betals.rds')
 
 if(FALSE) {
-  plot(m_ndvi, pages = 1, scheme = 3) # plot smooths
+  plot(m_ndvi, pages = 1, n = 400, n2 = 200, scheme = 3) # plot smooths
 }

### 2023-09-17 writing/appendix-3-empirical-modeling.Rmd
added script for betals family
one tapir is sufficient; added weights to GAMs

@@
 
-\newcommand{\e}{\mathbb E}
-\newcommand{\var}{\mathbb V}
+\newcommand{\e}{\text{E}}
+\newcommand{\var}{\text{Var}}
 
@@
 theme_set(theme_bw()) # change default theme
-getwd()
 
 # attach the beta location-scale family written by Simon Wood
-source('../ndvi-stochasticity/functions/betals.r')
+source('functions/betals.r')
 
@@
   plot(pages = 1, scheme = 3, n = 250, scale = 0)
-
-gam(list(
-  # mean predictor (ks too large)
-  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
-    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
-  # scale predictor (ks too small)
-  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
-  family = betals(), data = anna_ndvi, method = 'REML') %>%
-  plot(pages = 1, scheme = 3, n = 250, scale = 0)
 ```
-
-\clearpage
 
@@
 The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=8, fig.pos='!h', echo=FALSE}
+Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
 gam(list(
@@
 
+```{r}
+gam(list(
+  # mean predictor (ks too large)
+  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
+    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
+  # scale predictor (ks too small)
+  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
+  family = betals(), data = anna_ndvi, method = 'REML') %>%
+  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+```
+
 ```{r oversmooth, fig.cap="Estimated mean and scale parameters in NDVI with excessively large basis size $\\tt{k}$ for the smooth term of time ($\\tt{dec\\_date}$) for the mean in NDVI and excessively low for the variance temporal smooth term.
 The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=8, fig.pos='!h', echo=FALSE}
+Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
 gam(list(
@@
 
-# Modeling the effects of $\e(R)$ and $\var(R)$ on space-use requirements {#final-fig}
+# Modeling the effects of $\e(R)$ and $\var(R)$ on space use {#final-fig}
 
@@
 Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period.
+We also add a column of weights that are based on the home range confidence estimates' 95% intervals.
 
-```{r, eval=TRUE}
+```{r}
 tapir <-
@@
          sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
-  select(t_center, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
-  pivot_longer(c(hr_lwr_95, hr_est_95, hr_upr_95),
-               names_to = c('.value', 'quantile'),
-               names_pattern = '(.+)_(.+)') %>%
-  mutate(t_center = as.POSIXct(t_center),
-         quantile = paste0(quantile, '%'))
+    select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
+    mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # small CI => large weight
+           weight = weight / mean(weight)) # standardize weights
+sum(tapir$weight) == nrow(tapir) # ensure weights are right
+```
+
+```{r, eval=TRUE, echo=FALSE}
+tapir <- readRDS('data/anna-hr-ndvi-data.rds')
+sum(tapir$weight) == nrow(tapir) # ensure weights are right
 ```
@@
 
-# create the figure
-date_labs <- range(anna_tel$timestamp) %>% as.Date()
-YLIMS <- c(0, 13)
-
 # axis labels (spaces in hr_lab to avoid overlap with the plot label)
-# removed unicode characters to avoid knitting errors
 e_r <- 'Resource abundance'
@@
       # mean
-      ggplot(tapir, aes(t_center, mu)) +
+      ggplot(tapir, aes(date, mu)) +
         geom_line(color = pal[1]) +
@@
       # variance
-      ggplot(tapir, aes(t_center, sigma2)) +
+      ggplot(tapir, aes(date, sigma2)) +
         geom_line(color = pal[2]) +
@@
       # 95% home range
-      ggplot(tapir, aes(t_center, hr_est)) +
+      ggplot(tapir, aes(date, hr_est_95)) +
         geom_line(color = pal[3]) +
@@
 
-To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space-use requirements.
+To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use.
 To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.
@@
 
-```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space-use requirements (on the log link scale) using a GAM with no shape constraints.
-The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms."}
-gam(hr_est ~ s(mu,  k = 3) + s(sigma2, k = 3), family = Gamma('log'),
-    data = tapir) %>%
-  plot(pages = 1, scheme = 3)
-```
+We can now fit a GAM similar to the one used for the simulated data.
+However, since this dataset is much smaller and autocorrelated, we should be more parsimonious with the number of terms and their complexity.
+A model with a Gamma family of distributions and relatively smooth (i.e., not wiggly) marginal terms of $\mu(t)$ and $\sigma^2(t)$ is a reasonable start.
+This dataset is too small (and autocorrelated) to estimate changes in scale parameters with reasonable accuracy and precision.
 
@@
 
-Thus, we use shape-constrained splines to inform the model about the shape we expect the terms to have while also imposing monotonicity.
+```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a GAM with no shape constraints.
+The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
+Shaded areas inticate the 95\\% credible intervals."}
+gam(hr_est_95 ~ s(mu, bs = 'ts', k = 5) + s(sigma2, bs = 'ts', k = 5),
+         family = Gamma('log'),
+         weights = weight, # weights based on CI size
+         data = tapir,
+         method = 'REML') %>%
+  plot(pages = 1, scheme = 1) # trends should be monotonic
+```
 
-```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space-use requirements (on the log link scale) using a shape-constrained GAM.
+Due to the small sample size, the spatiotemporal autocorrelation, and the correlation between $\mu(t)$ and $\sigma^2(t)$, it is hard to separate the effects of the two predictors.
+However, since it is reasonable to expect the effects of $\mu(t)$ and $\sigma^2(t)$ to be monotonic decreaseing and increasing, respectively, we can use a Shape-Constrained Additive Model (SCAM) via the `scam` package [@pya_scam_2022].
+
+```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a shape-constrained GAM.
 The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms."}
-m <- scam(hr_est ~
-            s(mu, bs = 'mpd', k = 4) + # monotone decreasing p-spline
-            s(sigma2, bs = 'mpi', k = 4), # monotone increasing p-spline
+Shaded areas inticate the 95\\% credible intervals."}
+m <- scam(hr_est_95 ~
+            s(mu, bs = 'mpd', k = 5) +
+            s(sigma2, bs = 'mpi', k = 5),
           family = Gamma('log'),
+          weights = weight,
           data = tapir)
-plot(m, pages = 1, scheme = 3)
+plot(m, pages = 1, scheme = 1)
 ```
@@
   coord_cartesian(ylim = c(0, 12.5)) +
-  geom_point(aes(mu, hr_est), tapir, color = pal[3], alpha = 0.5) +
+  geom_point(aes(mu, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
   geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
               fill = pal[1], alpha = 0.3) +
-  geom_line(aes(mu, hr_mu_est), preds, color = pal[1]) +
+  geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
   xlab(e_r) +
@@
   coord_cartesian(ylim = c(0, 12.5)) +
-  geom_point(aes(sigma2, hr_est), tapir, color = pal[3], alpha = 0.5) +
+  geom_point(aes(sigma2, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
   geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
               preds, fill = pal[2], alpha = 0.3) +
-  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2]) +
+  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2],
+            linewidth = 2) +
   xlab(v_r) +

### 2023-09-17 analysis/figures/moving-window-figure-anna-betals.R

@@ -46,11 +46,10 @@ if(! file.exists('data/anna-hr-ndvi-data.rds')) {
     mutate(est = map(dataset, \(.d) filter(tel, timestamp %in% .d$timestamp)),
            mu = map_dbl(est, \(.d) mean(.d$mu)),
            sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
-    select(t_center, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
-    pivot_longer(c(hr_lwr_95, hr_est_95, hr_upr_95), names_to = c('.value', 'quantile'),
-                 names_pattern = '(.+)_(.+)') %>%
-    mutate(t_center = as.POSIXct(t_center),
-           quantile = paste0(quantile, '%'))
+    select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
+    mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # weights based on CI width
+           weight = weight / mean(weight)) # standardize weights
+  sum(tapir$weight) == nrow(tapir) # ensure weights are right
   saveRDS(tapir, 'data/anna-hr-ndvi-data.rds')
 } else {
   tapir <- readRDS('data/anna-hr-ndvi-data.rds')
@@ -58,25 +57,24 @@ if(! file.exists('data/anna-hr-ndvi-data.rds')) {
 
 # create the figure
 date_labs <- range(tel$timestamp) %>% as.Date()
-YLIMS <- c(0, 13)
 
 # E(R) and V(R) are highly correlated
 ggplot(tapir) +
-  geom_point(aes(mu, sigma2), alpha = 0.1) +
+  geom_point(aes(mu, sigma2, alpha = weight)) +
   labs(x = '\U1D707(t)', y = '\U1D70E\U00B2(t)')
 
 l_grobs <-
   lapply(
     list(
       # mean, variance and HR
-      ggplot(tapir, aes(t_center, mu)) +
+      ggplot(tapir, aes(date, mu)) +
         geom_line(color = pal[1], linewidth = 2) +
         labs(x = NULL, y = e_r),
-      ggplot(tapir, aes(t_center, sigma2)) +
+      ggplot(tapir, aes(date, sigma2)) +
         geom_line(color = pal[2], linewidth = 2) +
         labs(x = NULL, y = v_r),
       ggplot(tapir) +
[…]

### 2023-09-17 analysis/figures/moving-window-figure-tapirs.R

@@ -1,164 +1,0 @@
-library('dplyr')     # for data wrangling
-library('purrr')     # for functional programming
-library('tidyr')     # for data wrangling
-library('ctmm')      # for movement modeling
-library('mgcv')      # for empirical Bayesian modeling
-library('scam')      # for shape-constrained splines
-library('lubridate') # for smoother date wrangling
-library('ggplot2')   # for fancy figures
-library('cowplot')   # for fancy multi-panel plots
-
-source('analysis/figures/default-figure-styling.R') # for theme & palettes
-
-# axis lables
-e_r <- 'Resource abundance, \U1D707(t)'
-v_r <- 'Resource stochasticity, \U1D70E\U00B2(t)'
-hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
-
-# import location-scale model for predictions of mean and variance in NDVI
-m_ndvi <- readRDS('models/tapirs/tapirs-ndvi-betals.rds')
-
-# function to make predictions from the model for each telemetry
-ndvi_preds <- function(.data) {
-  .data <- data.frame(.data) # convert telemetry data to a data.frame
-  
-  predict(m_ndvi, newdata = .data, type = 'response', se.fit = FALSE) %>%
-    data.frame() %>%
-    tibble() %>%
-    transmute(mu = X1,
-              sigma2 = X1 * (1 - X1) * X2) %>% # didn't scale NDVI to [0, 1]
-    return()
-}
-
-# import tapir data (from https://doi.org/10.1186/s40462-022-00313-w)
-tel <- readRDS('../tapirs/models/tapirs-final.rds') %>%
-  filter(region == 'cerrado') %>%
-  pull(data) %>%
-  map_dfr(data.frame) %>%
-  rename(long = longitude, lat = latitude) %>%
-  mutate(dec_date = decimal_date(timestamp)) %>%
-  bind_cols(., ndvi_preds(.)) %>% 
-  as_tibble()
-
-if(! file.exists('data/tapirs-hr-ndvi-data.rds')) {
-  tapirs <-
-    readRDS('data/cerrado-hrs.rds') %>%
-    mutate(est = map(dataset, \(.d) filter(tel, timestamp %in% .d$timestamp)),
-           mu = map_dbl(est, \(.d) mean(.d$mu)),
[…]

### 2023-09-17 analysis/tapirus-terrestris/tapirs-moving-window.R

@@ -10,7 +10,7 @@ theme_set(theme_bw()) # change default theme
 # import tapir data from https://github.com/StefanoMezzini/tapirs
 tapirs <- readRDS('../tapirs/models/tapirs-final.rds') %>%
   select(region, name, name.short, data) %>%
-  filter(tapirs, region == 'cerrado')
+  filter(region != 'atlantica') # VHF data is too irregular
 
 # check ranges of tracking data
 tapirs <-
@@ -48,60 +48,12 @@ cerrado <- filter(tapirs, region == 'cerrado')
 select(cerrado, name, days, clicks, daily_clicks)
 
 cerrado %>%
-  filter(name != 'CE_08_COLOMBINA') %>% # remove tapir that fails 
+  filter(name == 'CE_31_ANNA') %>% # remove tapir that fails 
   # filter(name.short %in% # filter to only a few good examples
   #          c('ANNA', 'DONALINA', 'KURUKA', 'LOU', 'ZEFA', 'ZACA', 'TITI', 'ZIGGY')) %>%
   pull(data) %>% # extract data column
   map(function(.d) {
     projection(.d) <- ce_proj # add projection based on mean long and lat
-    cat('Running', .d@info$identity, '\n')
-    .d <- window_hr(.d,
-                    window = 15 %#% 'day', # 1 week of data for sufficient sample size
-                    dt = 15 %#% 'day', # move window over by a single day each time
-                    fig_path = 'figures/tapirs',
-                    rds_path = 'models/tapirs',
-                    cores = 1) # cannot parallelize on Windows
-  })
-
-# join all the Cerrado data into a single tibble
-cerrado <-
-  list.files(path = 'models/tapirs',
-             pattern = '*-window-15-days-dt-15-days.rds',
-             full.names = TRUE) %>%
-  tibble(name = .,
-         d = map(., \(x) readRDS(x))) %>%
-  mutate(name = substr(name,
-                       nchar('models/tapirs/CE_xx_x'),
-                       nchar(name) -
-                         nchar('-window-15-days-dt-15-days.rds'))) %>%
-  unnest(cols = d) %>%
-  select(name, t_center, hr_est_95, hr_lwr_95, hr_upr_95) %>%
-  filter(! is.na(hr_est_95)) %>%
-  mutate(t_center = as.POSIXct(t_center)) %>%
[…]

### 2023-09-17 analysis/tapirus-terrestris/tapirs-ndvi-betals.R

@@ -1,76 +1,0 @@
-library('ctmm')      # for tracking data and movement modeling
-library('dplyr')     # for data wrangling
-library('ggplot2')   # for fancy plots
-library('mgcv')      # for GAMs
-library('lubridate') # for smoother date wrangling
-library('sf')        # for spatial objects
-
-source('functions/betals.r') # beta location-scale fam 
-source('analysis/figures/default-figure-styling.R') # for NDVI color palette
-
-# import 95% home range estimates with no CIs (optional)
-uds <- readRDS('../tapirs/models/tapirs-akdes.rds') %>%
-  filter(region == 'cerrado') %>%
-  pull(akde) %>%
-  purrr::map_dfr(\(x) {
-    SpatialPolygonsDataFrame.UD(x, level.UD = 0.95, level = 0) %>%
-      sf::st_as_sf() %>%
-      sf::st_transform(crs = '+proj=longlat') %>%
-      st_union() %>%
-      st_as_sf()
-  })
-uds_merged <- uds %>%
-  st_union() %>%
-  st_buffer(3e3)
-plot(uds_merged, col = 'grey')
-plot(uds, add = TRUE, col = 'forestgreen')
-
-# import NDVI data
-d <- readRDS('data/ndvi-rasters/tapirs/tapirs-data.rds') %>%
-  mutate(dec_date = decimal_date(date))
-
-d %>%
-  filter(dec_date %in% unique(dec_date)[1:6]) %>%
-  ggplot(aes(long, lat, fill = ndvi)) +
-  facet_wrap(~ round(dec_date, 2)) +
-  geom_raster() +
-  geom_sf(dat = uds, inherit.aes = FALSE, alpha = 0.3) +
-  scale_fill_gradientn('NDVI', colours = ndvi_pal, limits = c(-1, 1))
-
-range(d$ndvi)
-n_distinct(d$dec_date)
-
-# run a test model
-m_test <-
-  bam(
-    (ndvi + 1) / 2 ~ # scale from [-1, 1] to [0, 1]
-      s(long, lat, bs = 'ds', k = 300) + # mean over space
-      s(dec_date, bs = 'tp', k = 10), # need high k to account for animal adapting
-    family = betar(link = 'logit'),
-    data = d,
-    method = 'fREML',
-    discrete = TRUE,
-    control = gam.control(trace = TRUE))
[…]

### 2023-09-17 analysis/tapirus-terrestris/tapirs-ndvi-data-to-local.R

@@ -1,89 +1,0 @@
-# see https://rspatialdata.github.io/vegetation.html for example code and more info
-#' install `rgeoboundaries` with `remotes::install_github('wmgeolab/rgeoboundaries')`
-#' install `MODIStsp` with `remotes::install_github('ropensci/MODIStsp')`
-#' `MODIStsp` version >= 2.0.7 from github fixes login issue due to 
-library('ctmm') # for movement data and models
-library('dplyr')    # for data wrangling
-library('sf')       # for spatial features
-library('MODIStsp') # for downloading NDVI rasters
-library('purrr')    # for functional programming
-library('tidyr')    # for data wrangling
-library('terra')    # to import and save rasters
-source('earthdata-login-info.R') # import personal login credentials for EarthData
-
-# use a bounding box from tracking data ----
-tapirs <- filter(readRDS('../tapirs/models/tapirs-final.rds'),
-                 region == 'cerrado')
-
-# extract all telemetries and convert them to a single, large data frame
-tels <- map_dfr(tapirs$data, data.frame)
-
-# import 95% home range estimates with no CIs (optional)
-uds <-
-  map_dfr(tapirs$akde, \(x) {
-    SpatialPolygonsDataFrame.UD(x, level.UD = 0.95, level = 0) %>%
-      sf::st_as_sf() %>%
-      sf::st_transform(crs = '+proj=longlat') %>%
-      st_union() %>%
-      st_as_sf()
-  })
-uds_merged <- uds %>%
-  st_union() %>%
-  st_buffer(3e3)
-plot(uds_merged, col = 'grey')
-plot(uds, col = 'forestgreen', add = TRUE)
-
-bbox <- st_bbox(uds_merged)
-
-plot(st_as_sf(st_as_sfc(bbox)))
-plot(uds_merged, col = 'grey', add = TRUE)
-plot(uds, col = 'forestgreen', add = TRUE)
-
-# download NDVI
-MODIStsp(gui = FALSE, # do not use the browser GUI, only run in R
-         reprocess =TRUE,
-         out_folder = 'data/ndvi-rasters/tapirs', # '<folder>/VI_16Days_250m_v6/NDVI'
-         selprod = 'Vegetation Indexes_16Days_250m (M*D13Q1)', # can't specify Terra here
-         prod_version = '061', # 2022 raster version
-         bandsel = 'NDVI', # Normalized Difference Vegetation Index layer only
-         sensor = 'Terra', # only terrestrial values, ignore main bodies of water
[…]

### 2023-09-17 writing/appendix-3-empirical-modeling.Rmd
added AIC comparison with mu-only model

@@
 
-Finally, we can create the predictions from the shape-constrained GAM.
+This SCAM fits substantially better than a SCAM that only accounts for $\mu(t)$:
+
+```{r, eval=TRUE}
+m_mu <- scam(hr_est_95 ~ s(mu, bs = 'mpd', k = 5),
+             family = Gamma('log'),
+             weights = weight,
+             data = tapir)
+round(AIC(m, m_mu), 3)
+```
+
+\newpage
+
+We can now create the predictions from the shape-constrained GAM.
 

### 2023-09-17 analysis/figures/moving-window-figure-anna-betals.R

@@ -108,6 +108,12 @@ m <- scam(hr_est_95 ~
           data = tapir)
 plot(m, pages = 1, scheme = 1)
 
+m0 <- scam(hr_est_95 ~ s(mu, bs = 'mpd', k = 5),
+           family = Gamma('log'),
+           weights = weight,
+           data = tapir)
+AIC(m, m0) # model that accounts for both mu and sigma2 fits much better
+
 # removing data from d and splitting into two panels
 preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
                 sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%

### 2023-09-17 writing/manuscript.Rmd
improved section on empirical example

@@
 Due to potential correlation between $\e(R)$ and $\var(R)$ in a small and autocorrelated sample, the relationship between $R$ and $H$ should be modeled carefully and with parsimony.
-To appropriately constrain the shape of the effects of $\e(R)$ and $\var(R)$, we suggest using a GLM or a shape-constrained GAM (as long as the wiggliness of each term is chosen carefully).
-Here, we model the effects of $\e(R)$ and $\var(R)$ on $H$ using a shape-constrained GAM with via the `scam` package [@pya_scam_2022].
-Additional information on the model can be found in Appendix 3.
+Below, we present an empirical example using GPS tracking data from a lowland tapir (*Tapirus terrestris*) and satellite-derived NDVI.
+Additional information can be found in Appendix 3.
 
@@
 \noindent Fig.
-\@ref(fig:tapir-mw) illustrates  the tapir's response to changes in $\e(R)$ and $\var(R)$.
-Panels **a** and **b** show the changes in mean and variance in NDVI, respectively, during the tracking period.
-Panel **c** shows the changes in the tapir's space use over time.
-Note how the home-range size is larger during periods of lower NDVI (e.g., August 2017) and smallest during periods with high NDVI (January 2018).
-In contrast, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (January 2018).
-Finally, panels **d** and **e** show the estimated (partial) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Again, the estimated effects of $\e(R)$ and $\var(R)$ follow the simple relationships we hypothesized (figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) and supported with the simulated responses (fig.
-\@ref(fig:5-5-reg)).
-Thus, while the time series shown in panel **c** may appear complicated, the tapir's response can be well described by two simple functions.
-However, note that since $\mu(t)$ and $\sigma^2(t)$ are highly corrleated, modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect.
-By including the effect of $\sigma^2(t)$ in the model, we show that the tapir responds to both resource abundance and stochasticity.
-When resources are relatively low (but still high: $\mu(t) \in (0.65, 0.75)$), the tapir does not respond to changes in resources abundance.
-However, duing periods of peak greenness ($\mu \approx 0.85$), its weekly home range is approximately halved.
-In contrast, the tapir adapts much more strongly to changes in $\sigma^2(t)$ by increasing its home range sixfold.
-This example demonstrates that (1) even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in periods of high stochasticity and that (2) the effects of stochasticity in $R$ should also be considered when estimating the effects of average $R$ on animals' space use.
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
+Panels **a** and **b** show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
+Panel **c** shows the changes in the tapir's 7-day home range over time.
+Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018).
+Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018).
+Finally, panels **d** and **e** show the estimated (marginal) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
+Simce $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel **f**) and (spatio)temporally autocorrelated (panels **a** and **b**), the effects of $R$ on $H$ should be modeled carefully.
+To avoid over-fitting the model, we constrain the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using a shape-constrained additive models (SCAM) with monotone deceasing and increasing P-splines for $\mu(t)$ and $\sigma^2(t)$, respectively.
+
+Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and an worse model fit ($\Delta\text{AIC} \approx 87$).
+By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity.
+While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
+\@ref(fig:tapir-mw)a), and thus the tapir does not experience resource scarcity during the tracking period.
+Consequently, this example demonstrates two points central to this paper.
+Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity.
+Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.
 

### 2023-09-18 writing/appendix-1-concepts-and-definitions.Rmd
removed Simon as an author
Simon asked to be removed because of lack of experience with movement modeling data

--- 2023-09-08 writing/appendix-1-concepts-and-definitions.Rmd
@@
     institute: [braes, biol]
-  - name: Simon Wood
-    institute: uofe
   - name: Chris H. Fleming
@@
   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.

### 2023-09-18 writing/appendix-2-simulations.Rmd
removed Simon as an author
Simon asked to be removed because of lack of experience with movement modeling data

@@
     institute: [biol, braes]
-  - name: Simon Wood
-    institute: uofe
   - name: E. Patrícia Medici
@@
   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - uofe: School of Mathematics, James Clerk Maxwell Building, University of Edinburgh, Edinburgh, Unted Kingdom.
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.

### 2023-09-20 analysis/figures/default-figure-styling.R
made text in figures bold; changed color scheme

@@ -1,11 +1,9 @@
 library('ggplot2') # for fancy figures
-library('stringi') # for working with strings
 
 # set default ggplot theme
-# cannot bold text because unicode characters fail
 theme_set(theme_bw() +
             theme(legend.position = 'none',
-                  text = element_text(size = 15)))
+                  text = element_text(size = 15, face = 'bold')))
 
 # custom color-blind palette
 pal <- c('#ff8c00', '#4477AA', '#009900', '#66CCEE',
@@ -17,6 +15,6 @@ create_ndvi_pal <- colorRampPalette(c('darkblue', 'dodgerblue', '#744700', '#d9b
 ndvi_pal <- create_ndvi_pal(100)
 
 # mean and variance notations
-e_r <- expression(E(italic(R)))
-var_r <- expression(Var(italic(R)))
-h <- 'Space-use requirements'
+e_r <- bquote(paste(bold('E('), bolditalic('R'), bold(')')))
+var_r <- bquote(paste(bold('Var('), bolditalic('R'), bold(')')))
+h <- bquote(paste(bold('Space-use requirements, '), bolditalic('H')))

### 2023-09-20 analysis/figures/habitat-examples.R

@@ -16,7 +16,11 @@ source('functions/qgamma2.R') # function to generate qgamma() from mean and vari
 source('analysis/figures/default-figure-styling.R')
 theme_set(theme_get() + 
             theme(legend.position = 'top',
-                  panel.grid = element_blank()))
+                  panel.grid = element_blank(),
+                  legend.text.align = 0,
+                  legend.background = element_rect(fill = 'transparent'),
+                  legend.margin = margin(l = 10, r = 10)))
+
 
 # resource abundance palette
 LOW <- '#744700'
@@ -66,13 +70,13 @@ p_1a <-
   geom_point(aes(color = sigma2), size = 2) +
   scale_x_continuous('Time', expand = c(0, 0), breaks = NULL) +
   scale_y_discrete('Patch', expand = c(0, 0)) +
-  scale_fill_gradient2(e_r, low = LOW, mid = MID, high = HIGH,
-                       limits = lim_food, breaks = lim_food, labels = c('Low', 'High'),
-                       midpoint = quantile(lim_food, 0.5)) +
-  scale_color_viridis_c(var_r, direction = -1, option = 'D', end = 1,
-                        begin = 0.3, limits = lim_sigma2, breaks = lim_sigma2,
-                        labels = c('Low', 'High')) +
-  guides(color = guide_colorbar(order = 0), fill = guide_colorbar(order = 1))
+  scale_fill_gradient(e_r, low = 'white', high = pal[1], limits = lim_food,
+                      breaks = lim_food, labels = c('Low', 'High')) +
+  scale_color_gradient(var_r, low = 'white', high = pal[2],
+                       limits = lim_sigma2, breaks = lim_sigma2,
+                       labels = c('Low', 'High')) +
+  guides(color = guide_colorbar(order = 0),
+         fill = guide_colorbar(order = 1))
 
 # b. variable definition
 # b-1: Gamma distribution
@@ -86,9 +90,11 @@ plot_expr <- function(string) {
 # probability density of R
 r_pdf <-
   ggplot(r_density, aes(r, dens)) +
-  geom_area(fill = pal[3], alpha = 0.5, color = pal[3]) +
-  scale_x_continuous('\U1D445', breaks = NULL) +
-  scale_y_continuous('P(\U1D445 = \U1D45F)', breaks = NULL)
[…]

### 2023-09-20 analysis/figures/home-range-hypotheses.R

@@ -20,7 +20,7 @@ d <-
          h_2 = if_else(split, 2 * mu^2 - 7 * mu + 3.916873, h_1),
          h_3 = if_else(split, h_2 + 30 * (0.35 - mu)^2, h_2))
 
-notes <- tibble(x = c(0.13, 0.3, 0.175, 0.68),
+notes <- tibble(x = c(0.13, 0.32, 0.19, 0.68),
                 y = c(2.05, 3.2, 1.35, 2.05),
                 text = c('range-resident', 'nomadic or\ndispersing',
                          'evolutionary timescale', 'ecological timescale'))
@@ -32,9 +32,11 @@ p_mu <-
               linewidth = 2, fill = pal[1]) +
   geom_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180) +
   geom_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0) +
-  geom_text(aes(x = x, y = y, label = text), notes, size = 6) +
-  scale_x_continuous('Resource abundance, E(\U1D445)', breaks = NULL,
-                     expand = c(0, 0)) +
+  geom_text(aes(x = x, y = y, label = text), notes, size = 6,
+            fontface = 'bold') +
+  scale_x_continuous(bquote(paste(bold('Resource abundance, E('),
+                                       bolditalic(R), bold(')'))),
+                     breaks = NULL, expand = c(0, 0)) +
   scale_y_continuous(hr_lab, breaks = 0, expand = c(0, 0)); p_mu
 
 # hypothesis for change in H over V(R) ----
@@ -45,8 +47,11 @@ p_s2 <-
               linewidth = 2, fill = pal[2]) +
   geom_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180) +
   geom_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0) +
-  geom_text(aes(x = x, y = y, label = text), notes, size = 6) +
-  scale_x_reverse('Resource stochasticity, Var(\U1D445)', breaks = NULL,
+  geom_text(aes(x = x, y = y, label = text), notes, size = 6,
+            fontface = 'bold') +
+  scale_x_reverse(bquote(paste(bold(Resource~'stochasticity, Var('),
+                                    bolditalic(R),bold(')'))),
+                  breaks = NULL,
[…]

### 2023-09-20 analysis/figures/hr-mean-variance-simulations.R

@@ -7,9 +7,13 @@ source('analysis/mean-variance-trends-panel-data.R') # tibble of mu, var
 days <- readRDS('simulations/days-hrs.rds')
 
 # main simulation figure of change in home range (5x5 plot) ----
-e_r <- 'Resource abundance, E(\U1D445|t) = \U1D707(\U1D461)'
-v_r <- 'Resource stochasticity, Var(\U1D445|t) = \U1D70E\U00B2(\U1D461)'
-hr_lab <- 'Space-use requirements, \U1D43B' # label for y axis
+e_r <- bquote(paste(bold('Resource abundance, E('), bolditalic('R'),
+                    bold(') = '), '\U1D6CD', bold('('), bolditalic('t'), bold(')')))
+v_r <- bquote(paste(bold('Resource stochasticity, Var('), bolditalic('R'),
+                    bold(') = '), '\U1D6D4\U00B2', bold('('),
+                    bolditalic('t'), bold(')')))
+hr_lab <- bquote(paste(bold('Space-use requirements, '), bolditalic('H')))
+
 p_sim <-
   ggplot(days) +
   facet_grid(variance ~ mean, drop = FALSE) + # facet by trends in mu and var

### 2023-09-20 analysis/figures/moving-window-figure-anna-betals.R

@@ -11,9 +11,12 @@ library('cowplot')   # for fancy multi-panel plots
 source('analysis/figures/default-figure-styling.R') # for theme & palettes
 
 # axis lables
-e_r <- 'Resource abundance, \U1D707(t)'
-v_r <- 'Resource stochasticity, \U1D70E\U00B2(t)'
-hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
+e_r <- bquote(paste(bold('Resource abundance, '), '\U1D6CD', bold('('),
+                    bolditalic('t'), bold(')')))
+v_r <- bquote(paste(bold('Resource stochasticity, '), '\U1D6D4\U00B2',
+                    bold('('), bolditalic('t'), bold(')')))
+hr_lab <- bquote(paste(bold('7-day home range size (km'), '\U00B2',
+                       bold(')')))
 
 # import location-scale model for predictions of mean and variance in NDVI
 m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
@@ -61,7 +64,8 @@ date_labs <- range(tel$timestamp) %>% as.Date()
 # E(R) and V(R) are highly correlated
 ggplot(tapir) +
   geom_point(aes(mu, sigma2, alpha = weight)) +
-  labs(x = '\U1D707(t)', y = '\U1D70E\U00B2(t)')
+  labs(x = '\U1D707(t)', y = '\U1D70E\U00B2(t)') +
+  theme(text = element_text(face = 'plain'))
 
 l_grobs <-
   lapply(
@@ -171,5 +175,5 @@ p_right <- plot_grid(plotlist = r_grobs, ncol = 1, labels = c('d.', 'e.', 'f.'),
 p <- plot_grid(p_left, p_right, NULL, nrow = 1, rel_widths = c(1, 1, 0.01))
 p
 
-ggsave('figures/tapir-example-with-data.png', plot = p, height = 11,
+ggsave('figures/tapir-example-with-data.png', plot = p, height = 12,
        width = 14, units = 'in', dpi = 600, bg = 'white')

### 2023-09-20 analysis/simulations/2-hr-mean-variance-simulations-days.R

@@ -4,7 +4,7 @@ library('dplyr') # for data wrangling (e.g., %>%)
 library('purrr') # for functional programming (e.g., map(), map_dbl())
 library('tidyr') # for data wrangling (e.g., nested tibbles)
 library('ggplot2') # for fancy plots
-theme_set(theme_bw())
+source('analysis/figures/default-figure-styling.R')
 source('analysis/mean-variance-trends-panel-data.R') # trends in E(R) and V(R)
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
 
@@ -115,7 +115,7 @@ all <-
 ggplot(all) +
   facet_wrap(~ replicates) +
   geom_line(aes(animal, diff_est, group = paste(mean, variance)), alpha = 0.1) +
-  geom_hline(yintercept = 0, color = 'darkorange') +
+  geom_hline(yintercept = 0, color = 'red') +
   labs(x = 'Time', y = 'Difference in average daily time required (hours)')
 
 ggsave('figures/5-by-5-sensitivity-analysis/time-to-satiety-difference.png',
@@ -124,7 +124,7 @@ ggsave('figures/5-by-5-sensitivity-analysis/time-to-satiety-difference.png',
 ggplot(all) +
   facet_wrap(~ replicates) +
   geom_line(aes(animal, rel_est, group = paste(mean, variance)), alpha = 0.1) +
-  geom_hline(yintercept = 1, color = 'darkorange') +
+  geom_hline(yintercept = 1, color = 'red') +
   scale_y_continuous(trans = 'log2', breaks = 2^c(-0.5, 0, 0.5),
                      labels = parse(text = c('2^-0.5', '0', '2^0.5')))+
   labs(x = 'Time',

### 2023-09-20 analysis/simulations/6-modeling-R-and-hr.R

@@ -4,10 +4,9 @@ library('mgcv')    # for empirical Bayes GAMs
 library('ggplot2') # for fancy plots
 library('cowplot') # for fancy multi-panel plots
 source('analysis/figures/default-figure-styling.R') # for color palette
-theme_set(theme_bw() + theme(legend.position = 'none'))
 
-e_r <- 'Resource abundance, E(\U1D445)'
-v_r <- 'Resource stochasticity, Var(\U1D445)'
+e_r <- 'paste(bold("Resource abundance, E("), bolditalic("R"), bold(")"))'
+v_r <- 'paste(bold("Resource stochasticity, Var("), bolditalic("R"), bold(")"))'
 
 sims <- readRDS('simulations/days-hrs.rds') %>%
   tibble() %>%
@@ -42,9 +41,24 @@ m_3 <- gam(
   family = gammals(),
   data = sims)
 
-AIC(m_1, m_2, m_3) # smooth model with interactions is best
-BIC(m_1, m_2, m_3) # smooth model with interactions is best
+# check if CV = sigma/mu is better than sigma
+m_4 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
+    ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
+  family = gammals(),
+  data = mutate(sims, cv = sqrt(sigma2) / mu))
 
+# ti term of the scale parameter looks over-fit and too uncertain
+plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+
+AIC(m_1, m_2, m_3, m_4) # smooth model with interactions is best
+BIC(m_1, m_2, m_3, m_4) # smooth model with interactions is best
+
+summary(m_3)
+summary(m_4)
+
+# choose the best model
 m <- m_3
 
 plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
@@ -97,7 +111,8 @@ sims_l <- sims %>%
   mutate(x = if_else(name == 'mu', e_r, v_r))
 
 ggplot() +
-  facet_grid(. ~ x, scales = 'free', switch = 'x') + # label below x axis
+  facet_grid(. ~ x, scales = 'free', switch = 'x',
+             labeller = label_parsed) + # label below x axis
   geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
[…]

### 2023-09-20 analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R

@@ -2,11 +2,11 @@ library('ctmm')    # for continuous-time movement modeling
 library('dplyr')   # for data wrangling (e.g., %>%)
 library('tidyr')   # for data wrangling (e.g., nested tibbles)
 library('ggplot2') # for fancy plots
+source('analysis/figures/default-figure-styling.R')
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
 source('analysis/mean-variance-trends-panel-data.R') # means & variances
 source('analysis/simulations/movement-model.R') # for consistency across scripts
 source('functions/get_hr.R') # for extracting gaussian home range
-theme_set(theme_bw())
 set.seed(1) # for consistent results
 
 tels <- readRDS('simulations/tracks.rds') # list of telemetry tracks
@@ -141,7 +141,7 @@ saturation_days <-
          m = map(1:n(), \(i) {
            cat('Fitting model', i, '\n')
            ctmm.fit(tel[[i]], theta[[i]])
-           }), # fit movement model
+         }), # fit movement model
          sigma = map_dbl(m, \(.m) ctmm:::area.covm(.m$sigma)), # var(position)
          hr = get_hr(.sigma = sigma, quantile = 0.95)) # Gaussian home range
 
@@ -157,16 +157,17 @@ saturation_days <- readRDS('simulations/hr-saturation-days.rds')
 p_hr_days <-
   ggplot(saturation_days, aes(n_days, hr)) +
   facet_wrap(~ case, nrow = 1) +
-  geom_vline(xintercept = 100, color = 'darkorange') +
+  geom_vline(xintercept = 100, color = 'red') +
   geom_smooth(method = 'gam', color = 'black',
               formula = y ~ s(x, bs = 'cs', k = 10),
               method.args = list(family = Gamma(link = 'log'))) +
   geom_point(alpha = 0.3) +
-  scale_x_continuous(expression(Number~of~days~sampled~(log[2]~scale)),
+  scale_x_continuous(expression(bold(Number~of~days~sampled~(log[2]~scale))),
                      trans = 'log2', breaks = c(2, 16, 128, 1024),
                      limits = c(2, 1100)) +
-  scale_y_continuous(trans = 'log2', expression(atop(Estimated~'space-use',
-                                requirements~(log[2]~scale)))); p_hr_days
+  scale_y_continuous(trans = 'log2',
+                     expression(bold(atop(Estimated~'space-use',
[…]

### 2023-09-20 writing/appendix-2-simulations.Rmd
updated based on Chrirs's comments

@@
     institute: [biol, braes]
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
   - name: E. Patrícia Medici
@@
   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
@@
             alpha = 0.1) +
-  geom_hline(yintercept = 0, color = 'darkorange') +
+  geom_hline(yintercept = 0, color = 'red') +
   labs(x = 'Time',
@@
 The differences are relative to simulations with $2^{10}=1024$ tracks for each time point on the x axis (and each scenario in figure 1).
-The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h'}
+The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h', out.width='\\textwidth'}
 knitr::include_graphics('../figures/5-by-5-sensitivity-analysis/time-to-satiety-difference.png',
@@
 1).
-The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h'}
+The amount of time stabilizes for $\\gtrapprox 100$ tracks, which suggests that the home range estimates should also be stable with more than 100 tracks.", fig.pos='!h', out.width='\\textwidth'}
 knitr::include_graphics('../figures/5-by-5-sensitivity-analysis/time-to-satiety-rel-difference.png',
@@
 
-saveRDS(saturation_days, 'simulations/hr-saturation-days.rds')
-
 saturation_days %>%
@@
 ```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated space-use requirements as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
-In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE}
+In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
 knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
@@
 
-```{r, eval=TRUE}
+```{r, eval=TRUE, fig.height=8, fig.width=8, out.width='65%'}
 # marginal effects of mu and sigma only (linear on the log link scale)
@@
 
-AIC(m_1, m_2, m_3) # smooth model with interactions is best
+# check if CV = sigma/mu is better than sigma
+m_4 <- gam(
+  list(
+    hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
+    ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
+  family = gammals(),
+  data = mutate(sims, cv = sqrt(sigma2) / mu))
 
-BIC(m_1, m_2, m_3) # smooth model with interactions is best
+# ti term of the scale parameter looks over-fit and too uncertain
+plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
 
+AIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
+BIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
+```
+
+\clearpage
+
+```{r, eval=TRUE}
+# the difference in df (and significance) is in the ti.1() term
+summary(m_3)
+```
+
+\newpage
+
+```{r, eval=TRUE}
+summary(m_4)
+```
+
+\newpage
+
+```{r, eval=TRUE, fig.height=8, fig.width=8, out.width='65%'}
 m <- m_3 # select the best model
-summary(m)
 ```
@@
               hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
+```
 
+\newpage
+
+```{r, eval=TRUE}
 # effect of sigma2
@@
 
-# final figure
-e_r <- 'Resource abundance'
-v_r <- 'Resource stochasticity'
+# final figure (Rmd throws and error with label_bquote)
+e_r <- 'paste(bold("Resource abundance, E("), bolditalic("R"), bold(")"))'
+v_r <-
+  'paste(bold("Resource stochasticity, Var("), bolditalic("R"), bold(")"))'
 
@@
   mutate(x = if_else(name == 'mu', e_r, v_r))
+```
 
+\newpage
+
+```{r, eval=TRUE}
 ggplot() +
-  facet_grid(. ~ x, scales = 'free', switch = 'x') + # label below x axis
+  facet_grid(. ~ x, scales = 'free', switch = 'x',
+             labeller = label_parsed) + # label below x axis
   geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
@@
   geom_line(aes(value, hr, color = x),
-            filter(preds, quantile == 'hr_95')) +
+            filter(preds, quantile == 'hr_95'), linewidth = 2) +
   scale_color_manual(values = pal) +
   scale_x_continuous(NULL, breaks = NULL) + 
-  scale_y_continuous('Home range size', breaks = NULL) +
+  scale_y_continuous(bquote(paste(bold('Space-use requirements, '),
+                                  bolditalic('H'))), breaks = NULL) +
   theme_bw() +

### 2023-09-20 writing/appendix-3-empirical-modeling.Rmd
updated based on Chrirs's comments

@@
                       singlespacing = TRUE) # use single spacing for code chunks
-knitr::opts_knit$set(root.dir = '..')
 ```
@@
 The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/tapir-example-with-data.png',
-                        rel_path = FALSE)
+knitr::include_graphics('../figures/tapir-example-with-data.png', rel_path = FALSE)
 ```
@@
 # attach the beta location-scale family written by Simon Wood
-source('functions/betals.r')
+source('../functions/betals.r')
 
 # source all necessary scripts
-source('analysis/figures/default-figure-styling.R') # for color palettes
-source('earthdata-login-info.R') # personal login info for EarthData
-source('functions/window_hr.R') # function to calculate HRs
+source('../analysis/figures/default-figure-styling.R') # for color palettes
+source('../earthdata-login-info.R') # personal login info for EarthData
+source('../functions/window_hr.R') # function to calculate HRs
 ```
@@
 # import tapir data from https://github.com/StefanoMezzini/tapirs
-anna <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
   filter(name.short == 'ANNA')
@@
   rds_path = 'models')
-anna_mw <- readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
+anna_mw <-readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
 anna_mw
@@
 ```{r, echo=FALSE, eval=TRUE}
-anna <- readRDS('../tapirs/models/tapirs-final.rds') %>%
+anna <- readRDS('../../tapirs/models/tapirs-final.rds') %>%
   filter(name.short == 'ANNA')
@@
 
-anna_mw <- readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
+anna_mw <- readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
 anna_mw
@@
 Panel $\\mathbf{a.}$ shows the tapir's GPS locations, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals.", out.width='\\textwidth'}
-knitr::include_graphics('H:/GitHub/hr-resource-stoch/figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png', rel_path = FALSE)
+knitr::include_graphics('../figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png')
 ```
@@
 theme_set(theme_bw()) # need to set the theme again for some reason
+# axis labels (spaces in hr_lab to avoid overlap with the plot label)
+# unicode characters removed because they knitting errors
+e_r <- bquote(paste(bold('Resource abundance, E('), bolditalic(R),
+                    bold(')')))
+v_r <- bquote(paste(bold('Resource stochasticity, Var('), bolditalic(R),
+                    bold(')')))
+hr_lab <- bquote(bold(paste('7-day home range size (', km^2, ')   ')))
 
-# axis labels (spaces in hr_lab to avoid overlap with the plot label)
-e_r <- 'Resource abundance'
-v_r <- 'Resource stochasticity'
-hr_lab <- expression('7-day'~home~range~size~(km^2)~'   ')
-
-l_grobs <-
-  lapply(
-    list(
-      # mean
-      ggplot(tapir, aes(date, mu)) +
+l_grobs <- lapply(
+  list(ggplot(tapir, aes(date, mu)) + # mean
          geom_line(color = pal[1]) +
          labs(x = NULL, y = e_r),
-      # variance
-      ggplot(tapir, aes(date, sigma2)) +
+       ggplot(tapir, aes(date, sigma2)) + # variance
          geom_line(color = pal[2]) +
          labs(x = NULL, y = v_r),
-      # 95% home range
-      ggplot(tapir, aes(date, hr_est_95)) +
+       ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
          geom_line(color = pal[3]) +
@@
 # align left margins of all plots
-aligned_widths <- align_margin(map(l_grobs,
-                                   function(x) {x$widths}), 'first')
+aligned_widths <- align_margin(map(l_grobs, \(x) {x$widths}), 'first')
 
 # Setting the dimensions of plots to the aligned dimensions
-for (i in seq_along(l_grobs)) {
-  l_grobs[[i]]$widths <- aligned_widths[[i]]
-}
+for(i in seq_along(l_grobs)) l_grobs[[i]]$widths <- aligned_widths[[i]]
 
@@
 ```
+
+\clearpage
 
@@
 
-\clearpage
-
 ```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a GAM with no shape constraints.
@@
 
-\newpage
-
 We can now create the predictions from the shape-constrained GAM.
@@
 theme_set(theme_bw()) # need to set the theme again for some reason
-
 # predict marginal effects for each term
@@
 # align right margins of all plots
-aligned_widths <- align_margin(map(r_grobs,
-                                   function(x) {x$widths}), 'first')
+aligned_widths <- align_margin(map(r_grobs, \(x) {x$widths}), 'first')
 
 # Setting the dimensions of plots to the aligned dimensions
-for (i in seq_along(r_grobs)) {
-  r_grobs[[i]]$widths <- aligned_widths[[i]]
-}
+for(i in seq_along(r_grobs)) r_grobs[[i]]$widths <- aligned_widths[[i]]
 

### 2023-09-20 writing/manuscript.Rmd
updated based on Chrirs's comments

@@
 
-\noindent Simon Wood: simon.wood@bath.edu
-
-\noindent Chris H. Fleming: chris.h.fleming@gmail.com
+\noindent Chris H. Fleming: christen.fleming@ucf.edu
 
@@
 
-@rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation over the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
+@rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation divided by the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
 However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.

### 2023-09-28 writing/manuscript.Rmd
added Simon in the acknowledgements

@@
 
-\noindent We would like to thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely: Aimee Chhen, Kim Hinz, Lauren Mills, and Sarah Wyse.
+\noindent We would like to thank Dr.
+Simon Wood for providing code to fit a Beta location-scale GAM despite not being involved directly with the project.
+Additionally, we thank all those who provided feedback on all posters, presentation, and writings related to this project.
+In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely: Aimee Chhen, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr.
+Simon Wood.
 This work was supported by the NSERC Discovery Grant RGPIN-2021-02758 to MJN and funds from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, Mitacs, and BC Parks.

### 2023-10-04 writing/manuscript.Rmd
added a ref of E(R) and H

@@
 Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
+However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
 

### 2023-11-23 analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R
improved figure scaling

@@ -169,5 +169,5 @@ p_hr_days <-
                      expression(bold(atop(Estimated~'space-use',
                                           requirements~(log[2]~scale)))))
 
-ggsave('figures/hr-over-days.png', plot = p_hr_days,
+ggsave('figures/hr-over-days.png', plot = p_hr_days, scale = 2,
        width = 6, height = 3, dpi = 'print')

### 2023-11-23 analysis/figures/moving-window-figure-anna-betals.R
switched to capital panel labels

@@ -91,7 +91,7 @@ for (i in seq_along(l_grobs)) {
 }
 
 # Draw aligned plots
-p_left <- plot_grid(plotlist = l_grobs, ncol = 1, labels = c('a.', 'b.', 'c.'),
+p_left <- plot_grid(plotlist = l_grobs, ncol = 1, labels = 'AUTO',
                     label_size = 22, label_x = -0.01, label_y = 1.025)
 
 # regression plots ----
@@ -141,7 +141,11 @@ p_d <- ggplot() +
               fill = pal[1], alpha = 0.3) +
   geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
   labs(x = e_r, y = hr_lab) +
-  scale_alpha_continuous(range = c(0.3, 1))
+  scale_alpha_continuous('Weight', range = c(0.3, 1)) +
+  theme(legend.position = c(0.93, 0.86),
+        legend.box.background = element_rect(color = '#00000080'),
+        legend.background = element_blank()) +
+  guides(alpha = guide_legend(override.aes = list(color = 'black')))
 
 p_e <- ggplot() +
   coord_cartesian(ylim = c(0, 12.5)) +
@@ -155,8 +159,7 @@ p_e <- ggplot() +
 p_f <- ggplot(tapir) +
   geom_point(aes(mu, sigma2, alpha = weight)) +
   labs(x = e_r, y = v_r) +
-  scale_alpha_continuous('Weight', range = c(0.3, 1)) +
-  theme(legend.position = 'right')
+  scale_alpha_continuous(range = c(0.3, 1))
 
 r_grobs <- map(list(p_d, p_e, p_f), as_grob)
 
@@ -168,8 +171,9 @@ for (i in seq_along(r_grobs)) {
   r_grobs[[i]]$widths <- aligned_widths[[i]]
 }
 
-p_right <- plot_grid(plotlist = r_grobs, ncol = 1, labels = c('d.', 'e.', 'f.'),
-                     label_size = 22, label_x = -0.01, label_y = 1.025)
+p_right <-
+  plot_grid(plotlist = r_grobs, ncol = 1, labels = c('D', 'E', 'F'),
+            label_size = 22, label_x = -0.01, label_y = 1.025)
 
 # add space to avoid cutting off x axis for (e.)
 p <- plot_grid(p_left, p_right, NULL, nrow = 1, rel_widths = c(1, 1, 0.01))

### 2023-11-23 analysis/figures/simulation-example.R

@@ -176,8 +176,9 @@ p <-
               theme(legend.key.height = unit(0.5, 'cm'),
                     text = element_text(face = 'bold', size = 18),
                     legend.key = element_rect(color = 'black'))),
-          nrow = 1, labels = c('a.', 'b.', 'c.', 'd.', 'e.', ''),
-          rel_widths = c(rep(1, 5), 0.4), label_size = 22)
+          nrow = 1, labels = c('A', 'B', 'C', 'D', 'E', ''),
+          rel_widths = c(rep(1, 5), 0.4),
+          label_size = 22)
 
 ggsave('figures/simulation-example.png', plot = p, width = 20,
        height = 3, scale = 1, bg = 'white', dpi = 600)

### 2023-11-23 analysis/simulations/sensitivity-analyses/1b-delta-t-sensitivity.R

@@ -1,5 +1,5 @@
-library('terra')   # for working with rasters
 library('ctmm')    # for generating movement models
+library('terra')   # for working with rasters
 library('dplyr')   # for data wrangling (%>%, mutate(), slice(), filter(), etc.)
 library('tidyr')   # for data wrangling (unnest, etc.)
 library('purrr')   # for functional programming (map_*(), etc.)
@@ -113,7 +113,8 @@ tel_short <- tel[tel$t <= max(consecutive_encounters$t) + 500, ]
 
 track <- tel_short %>%
   data.frame() %>%
-  mutate(cell_id = cellFromXY(HABITAT, SpatialPoints.telemetry(tel_short)),
+  mutate(cell_id = cellFromXY(HABITAT,
+                              SpatialPoints.telemetry(tel_short)),
          new_cell = c(1, diff(cell_id)) != 0)
 
 thin <- function(dt) {
@@ -186,7 +187,7 @@ panels <-
   plot_grid(plot_grid(get_legend(p_hist + theme(legend.position = 'top')),
                       get_legend(p_thinning + theme(legend.position = 'top'))),
             plot_grid(p_hist, p_prop, p_thinning, p_tracks,
-                      labels = c('a.', 'b.', 'c.', 'd.'), ncol = 2),
+                      labels = 'AUTO', ncol = 2),
             ncol = 1, rel_heights = c(1, 10))
 
 ggsave('figures/thinning-examples.png', plot = panels,

### 2023-11-23 functions/window_hr.R

@@ -131,7 +131,7 @@ window_hr <- function(tel, window, dt, projection, full_ud = NULL,
       geom_line(aes(date, hr_est_50, color = posixct))
   }
   
-  plt <- cowplot::plot_grid(plt_a, plt_b, labels = c('a.', 'b.'), nrow = 1,
+  plt <- cowplot::plot_grid(plt_a, plt_b, labels = c('A', 'B'), nrow = 1,
                             align = 'hv')
   
   if (! is.null(fig_path)) {

### 2023-11-23 writing/appendix-A-concepts-and-definitions.Rmd
prepared manuscript and appendices for submission

---
title: \large How resource abundance and stochasticity affect animals' space-use requirements
subtitle: "Appendix A: Concepts and definitions"
# author:
#   - name: Stefano Mezzini
#     email: stefano.mezzini@ubc.ca
#     institute: [braes, biol]
#   - name: Chris H. Fleming
#     institute: [ucf, smith]
#   - name: E. Patrícia Medici
#     institute: [tapir, iucn, escas]
#   - name: Michael J. Noonan
#     email: michael.noonan@ubc.ca
#     institute: [braes, biol, cmps]
# institute:
#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
bibliography: 'hr-resource-stoch.bib'
csl: 'the-american-naturalist.csl'
fontsize: 12pt
indent: true
header-includes:
    - \renewcommand{\figurename}{Figure A\!\!} # for "Figure Ax".
    - \usepackage{setspace}\doublespacing # for double-spaced text
    - \usepackage[small]{titlesec} # for smaller font for headings
    - \usepackage{caption} # for more customization of captions
    - \captionsetup[figure]{font=scriptsize} # smaller caption text
    - \usepackage{hanging} # for hanging indents in references
subparagraph: true # needed for \usepackage[small]{titlesec}
output:
  bookdown::pdf_document2:
    number_sections: false
    toc: false
    # latex_engine: xelatex # for sigma characters in figure
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua' # for author affiliations
      - '--lua-filter=author-info-blocks.lua' # for author affiliations
---

\newcommand*\e{\text{E}} <!-- expectation; non-italic E -->
\newcommand*\var{\text{Var}} <!-- variance; non-italic Var -->

\newpage

```{r setup, include=FALSE}
# set chunk defaults: don't print code, images 75% of page width and centered
knitr::opts_chunk$set(echo = FALSE, fig.align = 'center', out.width = '100%')
```

\noindent In this appendix, we presents the foundational concepts and terms used in the main manuscript. We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.

## Probability distributions

\noindent Random variables are defined by specifying the distribution the variable is assumed to follow (e.g., Gaussian, Gamma, Poisson, Bernoulli). Since the variable is random, it can take multiple possible values, each with different probabilities. The set or range of values which have non-zero probabilities in a distribution is referred to as the distribution's support.

There are many distributions we can assign to $R$ depending on how we quantify it. For instance, if we measure $R$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included). However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework). Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution. If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.

## Expected resources, $\e(R)$

\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average. The expectation of a random variable is the value one can expect the random variable to take, on average, in the long term. Here we use $\e(R)$ to indicate the expectation of the random variable $R$. When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$ (Fig. A\@ref(fig:multipanel-constant-var)). We indicate the estimated average amount of resources over time by adding a caret on the symbol: $\hat \mu(t)$. Note, however, that the estimate of $\mu(t)$ and what values it can take depend on what distribution we assume $R$ follows. For a Gamma distribution, $\mu(t)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t)$ is necessarily within the interval $(-1, 1)$. See the section on modeling $R$ in the main manuscript for more information on how to choose probability distributions.

```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean). (A) Resources ($R$, orange fill) vary over time and space with a constant the variance (dot color). (B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
```

## Variance in resources, $\var(R)$

\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$. A random variable's variance is a measure of its unpredictability (i.e., variability). We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$; see Fig. A\@ref(fig:multipanel-changing-var)). <!-- Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others due to changes in $\sigma^2(t)$. Additionally, we distinguish between two different sources of variation. We refer to known or expected spatiotemporal differences or changes as heterogeneity, which we consider as a change in $\mu(t)$. In contrast, we use the term stochasticity for unpredictable variation (i.e., due to $\sigma^2(t)$).--> For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons. Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t)$ to be lower in the winter since it does not expect to find berries (i.e., $\mu(t) \approx 0$). Also note that, like with $\mu(t)$, the estimate of $\sigma^2(t)$ and what values it can take depend on what distribution we assume $R$ follows. Figure A\@ref(fig:ex) shows four examples of Gamma distributions with different means and variances.

```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean). (A) Although resources ($R$, orange fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks. (B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$. The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
```

```{r ex, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Gamma distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
library('dplyr') # for data wrangling
library('tidyr') # for data wrangling
library('ggplot2') # for fancy plots

# to help visualize the consequences of different E(R) and V(R) if R ~ Gamma
dgamma2 <- function(x, mu, sigma2, lower.tail = FALSE) {
  # mean = k * theta
  # sigma^2 = k * theta^2
  dgamma(x = x,
         shape = mu^2 / sigma2, # (k * theta)^2 / (k * theta^2)
         scale = sigma2 / mu)   # (k * theta^2) / (k * theta)
}

expand_grid(x = tibble(x = seq(0, 10, length.out = 1000)),
            mean = c(1, 5),
            var = c(0.5, 5)) %>%
  unnest(x) %>%
  mutate(dens = dgamma2(x = x, mu = mean, sigma2 = var)) %>%
  ggplot(aes(x, dens)) +
  coord_cartesian(xlim = c(-0.5, 10.5), ylim = c(0, 0.8)) +
  # using '\U03BC' because mu results in \propto, for some reason
  facet_wrap(. ~ mean + var,
             labeller = label_bquote('\U03BC' == .(mean)~','~sigma^2 == .(var)))+
  geom_line(na.rm = TRUE) +
  geom_area(alpha = 0.3, na.rm = TRUE) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_continuous(expand = c(0, 0)) +
  labs(x = 'r', y = 'Density') +
  theme_bw()
```

\clearpage <!-- reference on a new page without figures -->

# References

\hangparas{1em}{1} <!-- indent all lines but the first -->

### 2023-11-23 writing/appendix-B-simulations.Rmd
prepared manuscript and appendices for submission

--- 2023-09-20 writing/appendix-2-simulations.Rmd
@@
 title: \large How resource abundance and stochasticity affect animals' space-use requirements
-subtitle: "Appendix 2: Simulations"
-author:
-  - name: Stefano Mezzini
-    email: stefano.mezzini@ubc.ca
-    institute: [biol, braes]
-  - name: Chris H. Fleming
-    institute: [ucf, smith]
-  - name: E. Patrícia Medici
-    institute: [tapir, iucn, escas]
-  - name: Michael J. Noonan
-    email: michael.noonan@ubc.ca
-    institute: [biol, braes, cmps]
-institute:
-  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+subtitle: "Appendix B: Simulations"
+# author:
+#   - name: Stefano Mezzini
+#     email: stefano.mezzini@ubc.ca
+#     institute: [biol, braes]
+#   - name: Chris H. Fleming
+#     institute: [ucf, smith]
+#   - name: E. Patrícia Medici
+#     institute: [tapir, iucn, escas]
+#   - name: Michael J. Noonan
+#     email: michael.noonan@ubc.ca
+#     institute: [biol, braes, cmps]
+# institute:
+#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'
@@
 header-includes:
+    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Sx.
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 
-<!-- have table of contents on pages of its own -->
-
-\newpage
-
-\tableofcontents
-
 \newpage
@@
 
-This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript.
-For ease of reference, we also include the figures here (figs.
-\@ref(fig:5-5-sims) and \@ref(fig:5-5-reg)).
+This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript (figs.
+3 and 4).
 To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
@@
 
-```{r 5-5-sims, eval=TRUE, echo=FALSE, fig.cap="Simulated home range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (and homogenous over space).
-Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%', cache=FALSE}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png',
-                        rel_path = FALSE)
-```
-
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated space-use requirements.
-The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
-The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both the location (i.e., mean) of $H$ and its scale parameter.
-Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in space-use requirements strongly increases with mean space-use requirements.", out.width='100%', cache=FALSE}
-knitr::include_graphics('../figures/simulation-regression-plots.png',
-                        rel_path = FALSE)
-```
-
 \newpage
@@
 
-To reduce sampling variance between simulations at each time point in each panel of figure \@ref(fig:5-5-sims)), we used the same set of 200 simulated tracks, which we generate in the `analysis/simulations/tracks.R` script.
+To reduce sampling variance between simulations at each time point in each panel of fig.
+3 in the main text, we used the same set of 200 simulated tracks, which we generate in the `analysis/simulations/tracks.R` script.
 Most intermediate and diagnostic checks are not included in this document for the sake of brevity and simplicity, but their outputs and conclusions are listed in this section.
@@
 (c.) Each time the animal collected more resources, its satiety (purple) increased.
-Once the animal collected sufficient resources, the animal stopped moving (i.e., the track was truncated).
-(d.) The process was repeated 200 times (13 tracks pictured in this panel).
-(e.) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
+Once the animal collected sufficient resources, the animal stopped exploring (i.e., the track was truncated) and was allowed to return to $\\langle0, 0 \\rangle$ over the same amount of time needed to reach satiety.
+(D) The process was repeated 200 times (13 tracks pictured in this panel).
+(E) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
 knitr::include_graphics('../figures/simulation-example.png', rel_path = FALSE)
@@
 Using three tracks generated with three arbitrary seeds (1, 2, and 3), we explored the effects of sampling interval ($\Delta t$) on the number of encounters with food (i.e., movements to new cells) detected.
-From each of the four checks, we created an exploratory plot that we present in figure \@ref(fig:dt-sensitivity).
+From each of the four checks, we created an exploratory plot that we present in figure B\@ref(fig:dt-sensitivity).
 
-**Exploratory plot \@ref(fig:dt-sensitivity)a.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
+**Exploratory plot B\@ref(fig:dt-sensitivity)a.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
 Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.
 
-**Exploratory plot \@ref(fig:dt-sensitivity)b.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$
+**Exploratory plot B\@ref(fig:dt-sensitivity)b.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$
 
-**Exploratory plot \@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
+**Exploratory plot B\@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
 
-**Exploratory plot \@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
+**Exploratory plot B\@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
 All three tracks have a reasonable amount of both tortuous and directed movement.
@@
 ```{r dt-sensitivity, fig.cap="Exploratory plots used to decide an appropriate sampling interval.
-(a.) Histograms of the number of encounters as a function of the interval between encounters, with a binwith of 30 seconds.
+(A) Histograms of the number of encounters as a function of the interval between encounters, with a binwith of 30 seconds.
 Although some encounters occur with less than 30 seconds between them, approximately 93\\% of them occur at least 30 seconds apart.
-(b.) Number of detected encounters as a function of sampling interval.
+(B) Number of detected encounters as a function of sampling interval.
 The colored lines indicate the estimated relationship based on a Generalized Additive Model fit using the $\\tt{geom\\_smooth}$ function from the $\\tt{ggplot2}$ package.
 Although the number of encounters detected decreases as sampling interval doubles, the loss at $\\Delta = 30$s is negligible.
-(c.) Beginning of the track generated with seed \"1\" (purple line) for different sampling intervals.
+(C) Beginning of the track generated with seed \"1\" (purple line) for different sampling intervals.
 Red dots indicate locations where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell and thus encountered food.
@@
 Additionally, the track at $\\Delta t = 30$s is still sufficiently tortuous to represent realistic animal movement.
-(d.) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
+(D) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
 knitr::include_graphics('../figures/thinning-examples.png', rel_path = FALSE)
@@
 Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
-When estimating space-use requirements based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix 3 on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+When estimating space-use requirements based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
@@
 Fig.
-\@ref(fig:gamls-terms) shows each of the terms from model 3.
+B\@ref(fig:gamls-terms) shows each of the terms from model 3.
 The estimated effect of `quantile` (see the output of `summary()` above) supports the relationship between the gaussian HR estimates detailed above, since $\exp(1.464) = 4.323218 \approx \log_{0.5}(0.05),$ but there is no differences in the scale parameter between the two quantiles.
@@
 
-Finally, we can use the model to recreate fig. \@ref(fig:5-5-reg) and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
+Finally, we can use the model to recreate fig. 4 from the main text and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
 

### 2023-11-23 writing/appendix-C-empirical-modeling.Rmd
prepared manuscript and appendices for submission

--- 2023-09-20 writing/appendix-3-empirical-modeling.Rmd
@@
 title: \large How resource abundance and stochasticity affect animals' space-use requirements
-subtitle: "Appendix 3: Empirical modeling"
-author:
-  - name: Stefano Mezzini
-    email: stefano.mezzini@ubc.ca
-    institute: [biol, braes]
-  - name: Chris H. Fleming
-    institute: [ucf, smith]
-  - name: E. Patrícia Medici
-    institute: [tapir, iucn, escas]
-  - name: Michael J. Noonan
-    email: michael.noonan@ubc.ca
-    institute: [biol, braes, cmps]
-institute:
-  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+subtitle: "Appendix C: Empirical modeling"
+# author:
+#   - name: Stefano Mezzini
+#     email: stefano.mezzini@ubc.ca
+#     institute: [biol, braes]
+#   - name: Chris H. Fleming
+#     institute: [ucf, smith]
+#   - name: E. Patrícia Medici
+#     institute: [tapir, iucn, escas]
+#   - name: Michael J. Noonan
+#     email: michael.noonan@ubc.ca
+#     institute: [biol, braes, cmps]
+# institute:
+#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'
@@
 header-includes:
+    - \renewcommand{\figurename}{Figure C\!\!} # for "Figure Sx.
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 
-This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript.
+This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig.
+5).
 The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-For ease of reference, we also include the figure below (fig.
-\@ref(fig:tapir-mw)).
-Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement), fig.
-\@ref(fig:tapir-mw)c) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi), figs.
-\@ref(fig:tapir-mw)a-b) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig), figs.
-\@ref(fig:tapir-mw)d-f).
-
-```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resource abundance.
-(a.) Mean resource abundance, $\\mu(t)$, esimated as the mean NDVI at the locations visited by the tapir.
-(b.) Varince in resource abundance, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir.
-(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
-(d., e.) estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them.
-(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
-Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
-See Appendix 3 for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%', eval=TRUE, echo=FALSE}
-knitr::include_graphics('../figures/tapir-example-with-data.png', rel_path = FALSE)
-```
+Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig)).
 
@@
 
+```{r, eval=FALSE}
+getwd() # change working directory or file paths if needed
+```
+
+```{r, eval=TRUE, echo=FALSE}
+print('H:/GitHub/hr-resource-stoch/writing') # for anonymity in peer review
+```
+
 ```{r, eval=TRUE}
-getwd() # change working directory or file paths if needed
-
 # attach all necessary packages
@@
 Finally, it saves an exploratory figure (fig.
-\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
+C\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
 
 ```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function.
-Panel $\\mathbf{a.}$ shows the tapir's GPS locations, while panel $\\mathbf{b.}$ shows the seven-day home range estimates (95\\% and 50\\% utilization quantiles) with 95\\% confidence intervals.", out.width='\\textwidth'}
+Panel $\\mathbf{A}$ shows the tapir's GPS locations, while panel $\\mathbf{B}$ shows the seven-day home range estimates (95\\% utilization quantile) with 95\\% confidence intervals.", out.width='\\textwidth'}
 knitr::include_graphics('../figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png')
@@
 anna_ndvi <-
-  readRDS('data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
+  readRDS('../data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
   mutate(dec_date = decimal_date(date))
@@
 The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see fig.
-\@ref(fig:gamls-terms)).
+C\@ref(fig:gamls-terms)).
 The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.
@@
 Excessively smooth terms for the mean can inflate the scale term (fig.
-\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig.
-\@ref(fig:oversmooth)).
+C\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig.
+C\@ref(fig:oversmooth)).
 Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
@@
 Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
-m_ndvi <- readRDS('models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
+m_ndvi <- readRDS('../models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
 plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
@@
 tapir <-
-  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+  readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
   mutate(est = map(dataset,
@@
 ```{r, eval=TRUE, echo=FALSE}
-tapir <- readRDS('data/anna-hr-ndvi-data.rds')
+tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
 sum(tapir$weight) == nrow(tapir) # ensure weights are right
@@
 
-We now have all we need to create the left side of figure \@ref(fig:tapir-mw).
+We now have all we need to create the left side of figure 5 from the main manuscript.
 
@@
   list(ggplot(tapir, aes(date, mu)) + # mean
-         geom_line(color = pal[1]) +
+         geom_line(color = pal[1], size = 2) +
          labs(x = NULL, y = e_r),
        ggplot(tapir, aes(date, sigma2)) + # variance
-         geom_line(color = pal[2]) +
+         geom_line(color = pal[2], size = 2) +
          labs(x = NULL, y = v_r),
        ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
-         geom_line(color = pal[3]) +
+         geom_line(color = pal[3], size = 2) +
          labs(x = NULL, y = hr_lab)),
@@
 # Draw aligned plots
-plot_grid(plotlist = l_grobs, nrow = 1, labels = c('a.', 'b.', 'c.'))
+plot_grid(plotlist = l_grobs, nrow = 1, labels = 'AUTO')
 ```
@@
 However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig.
-\@ref(fig:hr-gam)).
+C\@ref(fig:hr-gam)).
 
@@
 
-plot_grid(plotlist = r_grobs, ncol = 3, labels = c('d.', 'e.', 'f.'))
+plot_grid(plotlist = r_grobs, ncol = 3, labels = c('D', 'E', 'F'))
 ```

### 2023-11-23 writing/manuscript.Rmd
prepared manuscript and appendices for submission

@@
 title: \large How resource abundance and stochasticity affect organisms' space-use requirements
-author:
-  - name: Stefano Mezzini
-    institute: [braes, biol]
-  - name: Chris H. Fleming
-    institute: [ucf, smith]
-  - name: E. Patrícia Medici
-    institute: [tapir, iucn, escas]
-  - name: Michael J. Noonan
-    email: michael.noonan@ubc.ca
-    institute: [braes, biol, cmps]
-    correspondence: "yes"
-institute:
-  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+# author:
+#   - name: Stefano Mezzini
+#     institute: [braes, biol]
+#   - name: Chris H. Fleming
+#     institute: [ucf, smith]
+#   - name: E. Patrícia Medici
+#     institute: [tapir, iucn, escas]
+#   - name: Michael J. Noonan
+#     email: michael.noonan@ubc.ca
+#     institute: [braes, biol, cmps]
+#     correspondence: "yes"
+# institute:
+#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl'
+csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
+# biblio-style: 'amnatnat'
 fontsize: 12pt
@@
     - \usepackage{caption} # for more customization of captions
-    - \captionsetup[figure]{font=scriptsize, aboveskip=4pt, belowskip=-15pt}
+    - \captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
     - \usepackage{hanging} # for hanging indents in references
@@
   bookdown::pdf_document2:
-    latex_engine: xelatex # @reed_frequency_2003 has unicode character U+0308
+    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
     pandoc_args:
@@
 
+\bibliographystyle{amnatnat}
+
 <!-- define custom math symbols -->
@@
 
-<!-- place authors' emails on a new page -->
+```{r, echo=FALSE}
+# needs to be <= 200 words for AmNat
+abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
+This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
+This is not a trivial oversight.
+Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
+In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on organisms' space use.
+Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use.
+Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
+```
 
-\newpage
+<!-- elements of the title page -->
 
-\noindent \textbf{Authors' emails:}
+\noindent \textbf{Article type}: Major article
 
-\noindent Stefano Mezzini: stefano.mezzini@ubc.ca
+\noindent \textbf{Words in abstract}: `r stringi::stri_count_words(abstract)`
 
-\noindent Chris H. Fleming: christen.fleming@ucf.edu
+\noindent \textbf{Words in main text}: `r suppressMessages(wordcountaddin::word_count())`
 
-\noindent E. Patrícia Medici: medici@ipe.org.br
+\noindent \textbf{Figures}: 5
 
-\noindent Michael J. Noonan: michael.noonan@ubc.ca
+\noindent \textbf{Tables}: 0
 
-<!-- place table of contents on a new page -->
-\newpage
+\noindent \textbf{References}: 141 <!-- counted by using nature csl -->
 
-\singlespacing
+\noindent \textbf{Appendices}: 3
 
-\tableofcontents
+\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`
+
+<!-- Authors' emails:
+Stefano Mezzini: stefano.mezzini@ubc.ca
+Chris H. Fleming: christen.fleming@ucf.edu
+E. Patrícia Medici: medici@ipe.org.br
+Michael J. Noonan: michael.noonan@ubc.ca
+-->
 
@@
 
-\noindent The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
-This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the *mean* amount of resources, while responses to the variance around the mean are still largely unknown.
-This is not a trivial oversight.
-Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
-In this study, we first review the literature of studies that estimate organisms' space-use requirements as a function of resource abundance.
-We then leverage this information to provide a unifying framework and hypotheses for the effect of mean *and* variance in resources on organisms' space use.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use.
-Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals.
-Results from the simulations and empirical examples are presented using a fully transparent approach that allows researchers to apply the framework to their own data and inform area-based conservation efforts.
-
-\vspace*{\fill}
-
-\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`.
+\noindent `r abstract`
 
@@
 
-\noindent where $H$ is the organism's **home-range size**, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the **resources** the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
+\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
 Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
@@
 
-Here, we refer to a location's average amount of resources as **resource abundance**, while we use the phrase **resource stochasticity** to indicate the variability in resource abundance due to unpredictable causes.
+Here, we refer to a location's average amount of resources as resource abundance, while we use the phrase resource stochasticity to indicate the variability in resource abundance due to unpredictable causes.
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
@@
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource *abundance* on organisms' space use while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource *stochasticity* on organisms' space use while suggesting a second unifying hypothesis.
+Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis.
+Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis.
 Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
@@
 
-\noindent In statistics, **random variables** indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
 Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
@@
 This assumption also holds biologically, since resources tend to be less variable when they are less abundant.
-We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix 1.
+We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.
 
@@
 
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
-(a.) Resources ($R$, brown-green fill) vary over time and space with a constant the variance (dot color).
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
-knitr::include_graphics('../figures/habitat-examples-constant-variance.png')
-```
-
 Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig.
-\@ref(fig:multipanel-constant-var)a).
+A1).
 We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig.
-\@ref(fig:multipanel-constant-var)b).
+A2).
 Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
 However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict *where* $R$ is high if $\e(R)$ does not vary predictably over space.
+peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict where $R$ is high if $\e(R)$ does not vary predictably over space.
 
@@
 Like the scenario illustrated in fig.
-\@ref(fig:multipanel-constant-var), we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
+A1, we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
 We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig.
-\@ref(fig:multipanel-changing-var)).
+A2).
 When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
@@
 Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
-
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(a.) Although resources ($R$, brown-green fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(b.) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
-The points in the central panel indicate the ralizations of $R$ in panel a.", out.width = '0.5\\textwidth', fig.pos='!t'}
-knitr::include_graphics('../figures/habitat-examples-changing-variance.png')
-```
 
@@
 Finally, each set of 200 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution.
-Additional information is provided in Appendix 2.
+Additional information is provided in Appendix B.
 
@@
 The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
-\@ref(fig:multipanel-constant-var)).
+A1).
 As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
@@
 \@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
-Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases *sub*linearly with $\var(R)$.
+Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases sublinearly with $\var(R)$.
 Secondly, the organism's movement was simulated to be range resident within an homogeneous environment.
@@
 To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
-Appendix 3 contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
+Appendix C contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
 
@@
 
-\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the **linear transformation**
+\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation
 
@@
 \noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
-<!--Note that only **linear transformations** (i.e.
+<!--Note that only linear transformations (i.e.
 only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
@@
 
-\noindent To understand how movement can be viewed as a **stochastic process**, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
+\noindent To understand how movement can be viewed as a stochastic process, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
 If we do not know why the organism is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the organism may move from one time point to another.
 However, we can use the organism's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
-This period is often referred to as the **range crossing time** or the positional autocorrelation parameter, $\tau_p$.
+This period is often referred to as the range crossing time or the positional autocorrelation parameter, $\tau_p$.
 Similarly, we can define $\tau_v$ to be the amount of time the organism moves with the same direction and speed, or the time required for an organism's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
-$\tau_v$ is referred to the organism's **directional persistence** or the velocity autocorrelation parameter.
+$\tau_v$ is referred to the organism's directional persistence or the velocity autocorrelation parameter.
 
@@
 
-Accurate speed estimates require the **sampling interval**, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
+Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
 However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total **sampling duration**, $T$, is sufficiently long.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
 Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
 
-Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on *how* the organism moved and less on *why* it moved.
+Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on how the organism moved and less on why it moved.
 Organisms may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
@@
 
-<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a **sliding window** approach.
+<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a sliding window approach.
 The width of the window, $w$, should be a fixed value greater than two or three times the organism's expected range crossing time [@fleming_overcoming_2019].
@@
 Below, we present an empirical example using GPS tracking data from a lowland tapir (*Tapirus terrestris*) and satellite-derived NDVI.
-Additional information can be found in Appendix 3.
+Additional information can be found in Appendix C.
 
@@
 \@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
-Panels **a** and **b** show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
-Panel **c** shows the changes in the tapir's 7-day home range over time.
+Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
+Panel C shows the changes in the tapir's 7-day home range over time.
 Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018).
 Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018).
-Finally, panels **d** and **e** show the estimated (marginal) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Simce $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel **f**) and (spatio)temporally autocorrelated (panels **a** and **b**), the effects of $R$ on $H$ should be modeled carefully.
+Finally, panels D and E show the estimated (marginal) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
+Simce $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel F) and (spatio)temporally autocorrelated (panels A and B), the effects of $R$ on $H$ should be modeled carefully.
 To avoid over-fitting the model, we constrain the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using a shape-constrained additive models (SCAM) with monotone deceasing and increasing P-splines for $\mu(t)$ and $\sigma^2(t)$, respectively.
 
-Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and an worse model fit ($\Delta\text{AIC} \approx 87$).
+Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and a worse model fit ($\Delta\text{AIC} \approx 87$, see Appendix C).
 By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity.
 While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
-\@ref(fig:tapir-mw)a), and thus the tapir does not experience resource scarcity during the tracking period.
+\@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period.
 Consequently, this example demonstrates two points central to this paper.
@@
 ```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resources.
-(a.) Trends in resource abundance over time, $\\mu(t)$, esimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
-(b.) Varince in resources over time, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
-(c.) Estimated seven-day home range based on the 95\\% utilization quantiles.
-(d., e.) Estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
+(A) Trends in resource abundance over time, $\\mu(t)$, esimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
+(B) Varince in resources over time, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
+(C) Estimated seven-day home range based on the 95\\% utilization quantiles.
+(D, E) Estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
 The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them.
-(f.) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
+(F) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
 Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
-See Appendix 3 for additional information.
+See Appendix C for additional information.
 The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
@@
 
+<!--
 # Acknowledgements
@@
 All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of the tapir data, which is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
+
+# Author contribution statement
+After acceptance, all multi-author papers must provide an Author Contribution Statement in the final files.
+It appears after the Acknowledgments.
+The goal of the Author Contribution statement is to clearly assign responsibility for parts of the paper.
+This statement should be carefully written to indicate the role that every author played in generating the paper.
+However, it is essential that authors do not get credited for steps in which they were not deeply involved.
+Sample roles are provided below but be sure to assemble a Contribution Statement that fits the specifics of your paper:
+
+* Conceptualization
+* Funding acquisition
+* Methods development/experimental design
+* Data collection  
+* Data analysis
+* Data validation
+* Data visualization
+* Provided resources [including specimens, reagents, equipment]
+* Software development
+* Model analysis
+* Coding simulation
+* Supervision
+* Writing - original draft
+* Writing - review & editing
+
+-->
 

### 2023-11-23 analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R
moved panel data to `analysis/simulations` folder

@@ -4,7 +4,7 @@ library('tidyr')   # for data wrangling (e.g., nested tibbles)
 library('ggplot2') # for fancy plots
 source('analysis/figures/default-figure-styling.R')
 source('functions/rgamma2.R') # rgamma() parameterized by mean and variance
-source('analysis/mean-variance-trends-panel-data.R') # means & variances
+source('analysis/simulations/mean-variance-trends-panel-data.R')
 source('analysis/simulations/movement-model.R') # for consistency across scripts
 source('functions/get_hr.R') # for extracting gaussian home range
 set.seed(1) # for consistent results

### 2024-02-02 analysis/simulations/6-modeling-R-and-hr.R
added ti() panel

@@ -4,6 +4,10 @@ library('mgcv')    # for empirical Bayes GAMs
 library('ggplot2') # for fancy plots
 library('cowplot') # for fancy multi-panel plots
 source('analysis/figures/default-figure-styling.R') # for color palette
+theme_set(theme_get() +
+            theme(legend.position = 'none',
+                  strip.background = element_blank(),
+                  strip.text = element_text(size = 14, face = 'bold')))
 
 e_r <- 'paste(bold("Resource abundance, E("), bolditalic("R"), bold(")"))'
 v_r <- 'paste(bold("Resource stochasticity, Var("), bolditalic("R"), bold(")"))'
@@ -68,7 +72,7 @@ plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
 newd_mu <-
   expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
               sigma2 = mean(sims$sigma2),
-              quantile = unique(sims$quantile))
+              quantile = 'hr_95')
 
 preds_mu <-
   bind_cols( # bind new data and predictions
@@ -80,13 +84,25 @@ preds_mu <-
                 hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
                 hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
 
+p_mu <-
+  ggplot() +
+  geom_point(aes(mu, hr), filter(sims, quantile == 'hr_95'),
+             alpha = 0.1, color = pal[3]) +
+  geom_ribbon(aes(mu, ymin = hr_lwr, ymax = hr_upr), fill = pal[1],
+              preds_mu, alpha = 0.5) +
+  geom_line(aes(mu, hr), color = pal[1], linewidth = 1, preds_mu) +
+  scale_x_continuous(
+    bquote(paste(bold('Resource abundance, E('),
+                 bolditalic('R'), bold(')'))), breaks = NULL) +
+  scale_y_continuous(bquote(paste(bold('Space-use requirements, '),
+                                  bolditalic('H'))), breaks = NULL)
 
 # effect of sigma2
 newd_sigma2 <-
   expand_grid(mu = mean(sims$mu),
               sigma2 = seq(min(sims$sigma2), max(sims$sigma2),
                            length.out = 400),
-              quantile = unique(sims$quantile))
+              quantile = 'hr_95')
 
 preds_sigma2 <-
   bind_cols(
@@ -97,39 +113,61 @@ preds_sigma2 <-
[…]

### 2024-02-02 writing/appendix-A-concepts-and-definitions.Rmd
started major edits

@@
 We introduce each concept without assuming a statistical or quantitative background to maximize accessibility and minimize misunderstandings and misinterpretations.
+
+# Resources as a random variable
+
+\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
+Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
+Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
+
+In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
+Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
+However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa).
+As the mean approaches zero from the positive side, the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
+This assumption also holds biologically, since resources tend to be less variable when they are less abundant.
 

### 2024-02-02 writing/manuscript.Rmd
started major edits

@@
 csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
-# biblio-style: 'amnatnat'
 fontsize: 12pt
@@
 <!-- start main body on a new page -->
+# To do:
+
+* make sure appendices flow (added stuff from main text)
+
 \newpage
@@
 Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018].
-Thus, many organisms will adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.
-Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, movement represents one of the most readily available behaviors that species can adjust, whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].
+Thus, many organisms adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.<!-- Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
+However, --> Movement represents one of the most readily available behaviors that species can adjust to optimize their resource use [<!--@krivan_ideal_2008; @brown_ecology_1999; -->@kacelnik_ideal_1992; @charnov_optimal_1976].<!-- whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].-->
 
@@
 \noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
-Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].
-While many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, few studies build upon others' previous work.
+Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., Figure 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work.
 Instead, results are typically reported as independent, novel findings.
-Consequently, we currently lack a unifying framework for quantifying the effects of resource abundance on organisms' space-use requirements.
-In addition, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources.
+Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements.
+Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].
 
-Here, we refer to a location's average amount of resources as resource abundance, while we use the phrase resource stochasticity to indicate the variability in resource abundance due to unpredictable causes.
+Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean.
 We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
-For instance, a grassland with relatively low but constant forage availability will require drastically different behaviors and adaptations from a desert location with equally scarce forage but rare, sudden, and strong pulses of resources.
-In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [*sensu* @teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal as it searches for high-resource patches and short-term range residency until the local resources are depleted.
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in substantially different movement and life history strategies being selected for.
-Although it is generally expected that resource unpredictability will decrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity).
+In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976].
+Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies.
+Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
@@
 
+\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
+Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time.
+Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
+We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.
+
+<!--
 \noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
@@
 We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.
+-->
 
@@
 Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources, but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
+Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
 The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
 This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
-Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime as well as the minimum amount of space it requires to survive [which is necessarily greater than zero -- see fig.
-\@ref(fig:mu-hyp) as well as: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020; and contrast them to estimates based on linear models, e.g.: @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022].
+Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
 
@@
 
-When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-However, as $\e(R)$ decreases, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+Fig.
+\@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, large-scale changes in movement behavior (such as species-wide shifts to dispersal, migration, or nomadism) are more likely to occur over evolutionary timescales than over an organism's lifespan.
-For instance, larger home ranges requires greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
+Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
 
-It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
-Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
-\@ref(fig:mu-hyp).
+Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (Fig.
+A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
+However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
+peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.
 
-Consider the scenario where the average in resource abundance, $\e(R)$, changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (see fig.
-A1).
-We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant (fig.
-A2).
-Since $R$ is spatiotemporally random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
-However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $R$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017], but it will not be able to predict where $R$ is high if $\e(R)$ does not vary predictably over space.
-
-This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
+<!-- This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
 Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
@@
 Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response.
+-->
 
@@
 
-\noindent Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@burt_territoriality_1943; @southwood_habitat_1977;  @lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns as well as more stochastic ones.
+\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when data is scarce, but it likely not realistic, since $\var(R)$ often also changes over time.
+Periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$.
+In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$ because $R$ tends to vary across wider ranges (e.g., due to high spatial variability, or high $R$ followed by low $R$ after grazing).
+When both $\e(R)$ and $\var(R)$ change over time (fig.
+A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
+Statistically, this is because the more change one attributes to $\mu(t)$ (i.e., the wigglier it is), the smaller $\sigma^2(t)$ becomes.
+Conversely, the smoother $\mu(t)$ is, the larger $\sigma^2(t)$ becomes.
+Biologically, this is important because an organism's perception scale [@levin_problem_1992] determines whether it sees a change in $R$ as a trend in $\e(R)$ or as a stochastic event (i.e., due to $\var(R)$).
+An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
+Whether, an organism is able to predict trends in $\sigma^2(t)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their space-use requirements.
+While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still limited.
+Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.
+
+<!-- Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@stephens_optimal_1982; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
+However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns.
 For example, since berries are scarce outside the growing season, the variance in berries will also be low.
@@
 In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
+-->
+
+```{r, eval=FALSE, echo=FALSE}
+# the median decreases as V(R) increases, even if E(R) is constant
+source('functions/qgamma2.R')
+plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
+```
+
+Similarly to $\e(R)$, we hypothesize $\var(R)$ to have a nonlinear effect on organisms' space-use requirements.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect.
+But as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig.
+\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974], and (4) the added movement required to search for food increases organisms' energetic requirements.
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
+If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food, migrating, or relying on alternative food sources during stochastic events (e.g., the seasonal arrival of competitors).
 
@@
 
-Environmental variability can reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [@berger_climate_2018] and increases their space-use requirements.
-While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect.
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still very limited.
-Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.
-
-```{r, eval=FALSE, echo=FALSE}
-# the median decreases as V(R) increases, even if E(R) is constant
-source('functions/qgamma2.R')
-plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
-```
-
-Similar to the effect of $\e(R)$, we expect $\var(R)$ to have a nonlinear effect on organisms' space-use requirements.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect, but as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig.
-\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], and (3) the added movement required to search for food increases organisms' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food or migrating, or relying on alternative food sources during stochastic events (e.g., after the arrival of competitors).
-
-Like the scenario illustrated in fig.
-A1, we can imagine an environment where both $\e(R)$ and $\var(R)$ change over time.
-We can then define $\var(R)$ as a function of time using the same notation as we did for $\e(R)$, i.e., $\var(R | t) = \sigma^2(t)$ (fig.
-A2).
-When both $\e(R)$ and $\var(R)$ change over time, predicting $R$ becomes more complex.
-Statistically, this is because $\sigma^2(t)$ decreases as we allow $\mu(t)$ to change more over time and explain a greater proportion of $\var(R)$.
-Biologically, this is important because some organisms may perceive changes in $R$ as a change in $\e(R)$, while others may perceive the change as a stochastic event.
-An organism's perception of changes in $R$ will depend strongly on the organism's cognitive capacities, including its memory and lifespan [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-The ability to predict trends in $\mu(t)$ and $\sigma^2(t)$ requires organisms have high cognitive capacity, since distinguishing between changes in $\e(R)$ and $\var (R)$ is not easy [@steixner-kumar_strategies_2020], especially if $\e(R)$ and $\var(R)$ are not independent, as in the case of $R \sim \Gamma(\mu(t), \sigma^2(t))$.
-Regardless of whether an organism is capable of predicting changes in $\sigma^2(t)$, the effect of $\var(R)$ is arguably as important as the effect of $\e(R)$ in determining how said organism will respond to changes in $R$ and consequently adapt its space use.
+***HERE***
 
@@
 
-@rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation divided by the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
-However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
-Likewise, @herfindal_prey_2005 found that the home ranges of Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
-A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
-Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' space-use requirements, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
-Thus, it is important to disentangle the effects of changes in $\e(R)$, such as seasonal trends in $\mu(t)$, and stochasticity ($\var(R)$).
-The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
-
-@nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since organisms in habitats with high and low seasonality had opposite responses to $\e(R)$.
-Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because organisms respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
-However, it does not explain why organisms may respond to $\e(R)$ in opposite ways in different habitats.
-Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
-Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.
-
 ## Simulating responses to $\e(R)$ and $\var(R)$
 
-\noindent To support our hypothesis of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
-To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022] to generate 200 tracks (see Appendices 2 for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model; @gurarie_correlated_2017].
+\noindent To support our hypotheses of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendices 2 for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
 The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
@@
 As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
+
+***HERE***
 
@@
 Although the temporal trends in fig.
-\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two surprisingly simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
-Both $\e(R)$ and $\var(R)$ affect space-use requirements precisely as we hypothesized (fig.
-\@ref(fig:5-5-reg)).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
+\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
+These relationships follow the hypotheses we presented in figs.
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) and are represented in panels A and B of fig.
+\@ref(fig:5-5-reg), while panel C shows the amount by which the effect of $\mu(t)$ on $H$ changes as a function of $\sigma^2(t)$ (and vice-versa; panel C).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
 We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of fig.
@@
 
-# Applying the framework
+***to do***:
+
+* add a paragraph and figure on effects of $\mu(t)$ and $\sigma^2(t)$ on $\var(R)$.
+
+# A lowland tapir in the Brazilian Cerrado: a case study
+
+***to do***:
+
+* open with limitations of sims
+* then highlight shortcomings of empirical data
+* move trasnformation stuff to the appendix
+* move modeling details to appendix C
 
@@
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2022] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
+`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
 Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
@@
 <!-- Modeling movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
-In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2022].
+In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023].
 -->
@@
 
-<!-- \noindent This paper presents a unifying framework for describing the effects of resource abundance and stochasticity on space-use requirements.
-The hypotheses and methods we present allow researchers to generalize the effects $\e(R)$ and $\var(R)$ have on space-use requirements by building upon previous work rather than stating their results as new findings.
--->
+***to do***:
+
+* add another intro paragraph to tie back to general idea and connecting to previous theory
+* provided hyp, sims, empirical examples, framework to test ideas
+
+\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on space-use requirements.
 
 ## Adaptations to changes in $\e(R)$ and $\var(R)$
+
+Paragraphs to tie in:
+
+* It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
+Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
+Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
+\@ref(fig:mu-hyp).
+* @rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation divided by the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
+However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
+Similarly, @mueller_how_2011 suggested that ecosystem variability [measured as spatial semivariance in Normalized Difference Vegetation Index, NDVI; see @pettorelli_normalized_2011] causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
+Likewise, @herfindal_prey_2005 found that the home ranges of Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
+A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
+Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' space-use requirements, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
+Thus, it is important to disentangle the effects of changes in $\e(R)$, such as seasonal trends in $\mu(t)$, and stochasticity ($\var(R)$).
+The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
+* @nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since organisms in habitats with high and low seasonality had opposite responses to $\e(R)$.
+Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
+This may be because organisms respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
+However, it does not explain why organisms may respond to $\e(R)$ in opposite ways in different habitats.
+Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
+Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.
 

### 2024-03-06 writing/appendix-B-simulations.Rmd
moved stuff to appendices; working on discussion

@@
 library('cowplot') # for fancy multi-panel plots
+library('dagitty') # for directed acyclical graphs
+library('ggdag')   # for directed acyclical graphs
 theme_set(theme_bw()) # change default theme
@@
 Both information criteria demonstrate that the third model, which accounts for the interaction effects of $\e(R)$ and $\var(R)$, is the best model.
+Each of the three models are based on the directed acyclical graph (DAG) shown in figure B\@ref(fig:dag).
+Within a Bayesian framework, DAGs can be used to infer causality within statistical models [as the latter alone does not allow one to infer causality, see @mcelreath_statistical_2016].
+The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's space-use requirements, $H$.
+Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).
+
+```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=4}
+dag <- dagitty('dag{
+"E(R)" -> R <- "Var(R)"
+"E(R)" -> "Var(R)"
+R -> H
+}')
+coordinates(dag) <- list(x = c(R = 0, 'E(R)' = -1, 'Var(R)' = 1, H = 0),
+                         y = c(R = 1, 'E(R)' = 0, 'Var(R)' = 0, H = 2))
+ggdag(dag) +
+  theme_dag()
+```
 
@@
   family = gammals(),
-  data = sims)
+  data = sims,
+  method = 'REML')
 
@@
   list(
-    hr ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5),
-    ~ quantile + s(mu) + s(sigma2) + ti(mu, sigma2, k = 5)),
+    hr ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5),
+    ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5)),
   family = gammals(),
-  data = sims)
+  data = sims,
+  method = 'REML')
 
@@
   family = gammals(),
-  data = mutate(sims, cv = sqrt(sigma2) / mu))
+  data = mutate(sims, cv = sqrt(sigma2) / mu),
+  method = 'REML')
 
@@
 plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+```
 
+\clearpage
+
+```{r, eval=TRUE}
 AIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4

### 2024-03-06 writing/appendix-C-empirical-modeling.Rmd
moved stuff to appendices; working on discussion

@@
 
-This appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig.
+We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements.
+Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' space-use requirements under different scenarios.
+
+# Modeling $R$ {#modeling-r}
+
+\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
+its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
+`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
+Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
+
+The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
+However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
+available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].
+
+The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->.
+The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009].
+In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
+If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
+
+Modeling the mean and variance terms of $R$ should be done carefully.
+Since trends in both $\e(R)$ and $\var(R)$ can be spatiotemporally nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
+However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms.
+An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
+Although there is no error-proof system, choosing the complexity of the terms based on the organism's ability to detect change and adapt is a reasonable starting point.
+Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
+We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
+Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model.
+@simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.
+
+# Estimating $R$ using NDVI
+
+\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation
+
+\begin{equation}
+\nu_{[0, 1]} = \frac{\nu + 1}{2},
+(\#eq:ndvi-to-01)
+\end{equation}
+
+\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
+<!--Note that only linear transformations (i.e.
+only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
+
+\begin{equation}
+\e(R) = \frac{\alpha}{\alpha + \beta}
+(\#eq:beta-mean)
+\end{equation}
+
+\noindent while the variance is
+
+\begin{equation}
+\var(R) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
+(\#eq:beta-var)
+\end{equation}
+
+\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):
+
+\singlespace <!-- to avoid excessive spacing -->
+\begin{equation}
+\nu = 2 \nu_{[0, 1]} - 1.
+(\#eq:01-to-ndvi)
+\end{equation}
+
+\noindent Consequently, we have
+
+\begin{equation}
+\e(\nu|t) = 2 \mu(t) - 1.
+(\#eq:mu-to-e-ndvi)
+\end{equation}
+
+\noindent and
+
+\begin{equation}
+\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
+(\#eq:sigma2-to-v-ndvi)
+\end{equation}
+
+\doublespace <!-- back to double spaced -->
+
+Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
+Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
+
+<!--
+## Movement as a stochastic process
+
+### Drivers of movement
+
+\noindent @nathan_movement_2008 suggested we can generalize movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
+
+\begin{equation}
+\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
+(\#eq:nathan-movement)
+\end{equation}
+
+\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
+This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
+The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
+Thus, we can instead focus on organisms' movement and space-use requirements as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
+
+### Visualizing movement as a stochastic process
+
+\noindent To understand how movement can be viewed as a stochastic process, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
+If we do not know why the organism is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the organism may move from one time point to another.
+However, we can use the organism's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
+This period is often referred to as the range crossing time or the positional autocorrelation parameter, $\tau_p$.
+Similarly, we can define $\tau_v$ to be the amount of time the organism moves with the same direction and speed, or the time required for an organism's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
+$\tau_v$ is referred to the organism's directional persistence or the velocity autocorrelation parameter.
+
+By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data.
+While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].
+
+Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
+However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
+Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
+
+Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on how the organism moved and less on why it moved.
+Organisms may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
+Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in organism movement while focusing on the organisms' movement behaviors rather than the causes.
+
+### Quantifying space-use requirements
+
+\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
+Although here we quantify organism's space-use requirements using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
+In the case of organisms that are range resident for short periods of time, we can model their movement and short-term space-use requirements within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
+Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
+Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
+In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
+-->
+
+<!-- ### Modeling space-use requirements -->
+
+<!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
+However, most commonly used movement models assume organisms' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
+-->
+
+<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a sliding window approach.
+The width of the window, $w$, should be a fixed value greater than two or three times the organism's expected range crossing time [@fleming_overcoming_2019].
+If the organism's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
+Alternatively, if the organism's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
+-->
+
+<!-- Modeling movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
+In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023].
+-->
+
+# Reproducing the analyses
+
+This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig.
 5).
@@
 library('MODIStsp')  # for downloading NDVI rasters
+library('dagitty')   # for directed acyclical graphs
+library('ggdag')     # for directed acyclical graphs
 theme_set(theme_bw()) # change default theme
@@
 ggplot(anna_ndvi, aes(ndvi)) +
-  geom_histogram(na.rm = TRUE) +
+  geom_histogram(na.rm = TRUE, color = 'black', fill = 'grey') +
   labs(x = 'NDVI', y = 'Count') +
@@
 
-\clearpage
+<!-- \clearpage -->
 
 To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use.
-To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.
-However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig.
-C\@ref(fig:hr-gam)).
+To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.<!-- However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig.
+C\@ref(fig:hr-gam)).-->
 
@@
 This dataset is too small (and autocorrelated) to estimate changes in scale parameters with reasonable accuracy and precision.
+As in Appendix B, we provide the causal Directed Acyclical Graph (DAG) in figure C\@ref(fig:dag).
+See the section on strengths and limitations of the empirical approach in the main text for a discussion of the DAG.
+
+```{r dag, eval=TRUE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
+$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=4}
+dag <- dagitty('dag{
+Z -> R -> "NDVI" <- S <- R
+R -> H <- Z
+}')
+coordinates(dag) <- list(x = c(R = 0, 'NDVI' = 0, Z = 1, S = 1, H = 0),
+                         y = c(R = 0, 'NDVI' = 1, Z = 0, S = 1, H = -1))
+plot(dag)
+```
+
+\clearpage
 

### 2024-03-06 writing/manuscript.Rmd
moved stuff to appendices; working on discussion

@@
 
-\noindent \textbf{Figures}: 5
+\noindent \textbf{Figures}: 6
 
@@
 * make sure appendices flow (added stuff from main text)
+* check number of references
 
@@
 \noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
-Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., Figure 2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work.
+Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig.
+2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work.
 Instead, results are typically reported as independent, novel findings.
@@
 The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., figure 2 in the work of @bista_effect_2022].
+This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., fig.
+2 in the work of @bista_effect_2022].
 Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
@@
 
-Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (Fig.
+Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig.
 A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
@@
 
-***HERE***
-
 ## Interactive effects of $\e(R)$ and $\var(R)$
@@
 \noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement.
-However, in the more realistic scenario where both $\e(R)$ and $\var(R)$ fluctuate over time, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on organism's movement should account for not only the effects of $\e(R)$ and $\var(R)$ separately, but also their interactive effects.
+However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
+Consequently, models estimating the effects of $R$ on organism's movement should account for not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
 Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
+We explore this more in the following section.
 
@@
 
-## Simulating responses to $\e(R)$ and $\var(R)$
+## Simulating responses to $\e(R)$ and $\var(R)$ {#sims}
 
 \noindent To support our hypotheses of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
-To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendices 2 for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
+To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
 The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
-However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not assume the movement process is spatially stationary, so the organism is not assumed to be range-resident.
+However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
 Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.
@@
 \@ref(fig:5-5-sims)).
-Note that $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
-Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ with no correlation to the following track.
-Finally, each set of 200 truncated tracks was modeled using an OUF model and Autocorrelated Kernel Density Estimate to estimate the 95% utilization distribution.
+The value of $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
+Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track.
+Finally, as each set of 200 truncated was now spatially restricted and centered around $\langle 0, 0\rangle$, the set of tracks was modeled to calculate the 95% utilization distribution from an OUF model and Autocorrelated Kernel Density Estimate.
+We designed the simulations presented here to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments.
+Consequently, we made the following four assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$).
+(2) Organisms move randomly in the environment with no energetic costs.
+(3) Organisms have no perceptive range or memory, so they are unable to detect, learn, or predict where resources are abundant (high $\e(R)$) or reliable (low $\var(R)$) over time.
+(4) Resources do not deplete, and there is no competition or predator avoidance.
 Additional information is provided in Appendix B.
@@
 Fig.
-\@ref(fig:5-5-sims) shows how the simulated space-use requirements changed in response to $\mu(t)$ and $\sigma^2(t)$.
-The top row (constant $\var(R)$) shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
+\@ref(fig:5-5-sims) shows how simulated space-use requirements responded to changes in $\mu(t)$ and $\sigma^2(t)$ in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically.
+The top row <!--(constant $\var(R)$) -->shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
 A1).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) the space-use requirements decrease nonlinearly, with larger changes when $\e(R)$ is low, until home-range size approaches the minimum size required by the organism.
-As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$) $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism.
+Although less evident, note how the noise in the lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->
 
-***HERE***
-
-In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
+<!-- In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
 However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase organisms' space-use requirements suddenly and unpredictably.
@@
 In cases where $\mu(t)$ is highly unpredictable, organisms should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
+-->
 
 The leftmost column of fig.
-\@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
+\@ref(fig:5-5-sims) <!--(constant $\e(R)$) -->illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
@@
 The remaining panels in fig.
-\@ref(fig:5-5-sims) illustrate how the effect of $\var(R)$ depends on $\e(R)$ (and vice-versa) as well as how complex the relationship can be.
+\@ref(fig:5-5-sims) illustrate $\e(R)$ and $\var(R)$ jointly affect $H$ and how complex the effects can be.
 Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
 When both $\e(R)$ and $\var(R)$ increase linearly, $H$ initially increases since the effect of $\var(R)$ is stronger, but then decreases as the effect of $\e(R)$ begins to dominate.
-However, when $\mu(t)$ and $\sigma^2(t)$ follow fairly different trends (e.g., cyclical $\e(R)$ and linear $\var(R)$), it is easy to see how an increase in $\var(R)$ amplifies the effects of $\e(R)$.
-In contrast, the effect of $\var(R)$ is stronger when $\e(R)$ is low (e.g., linear $\e(R)$ with cyclical $\var(R)$).
-Thus, $H$ is largest when $\e(R)$ is low and $\var(R)$ is high, and it is smallest when $\e(R)$ is high and $\var(R)$ is low [see the panel with cyclical $\e(R)$ and $\var(R)$ as well as @lai_movement_2017].
+Difficulties in disentangling the two effects are explored in greater depth in the case study in the following section.
 
-Not all 25 scenarios depicted in fig.
+<!-- Not all 25 scenarios depicted in fig.
 \@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on space use are useful examples that can be thought of as simplified scenarios.
@@
 However, if stochastic changes are sufficiently small and frequent, organisms may perceive them as continuous and smooth changes rather a series of small and sudden changes.
+-->
 
 Although the temporal trends in fig.
-\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when home-range size is shown as a function of either $\e(R)$ or $\var(R)$, rather than time.
-These relationships follow the hypotheses we presented in figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) and are represented in panels A and B of fig.
-\@ref(fig:5-5-reg), while panel C shows the amount by which the effect of $\mu(t)$ on $H$ changes as a function of $\sigma^2(t)$ (and vice-versa; panel C).<!--, although we found that the effect of $\var(R)$ is closer linear than we expected (it is mostly sub-linear as long as $H$ is sufficiently far from zero).
-We ensured this was not due to a low maximum $\var(R)$ (as in the left-hand side of fig.
-\@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> We believe the approximately linear effect of $\var(R)$ on $H$ is mainly due to two factors.
-Firstly, $H$ scales superlinearly with the standard deviation in $R$ (as $\sqrt{\var(R)}$ is on the same scale as $\e(R)$; not shown), which would necessarily imply that $H$ increases sublinearly with $\var(R)$.
-Secondly, the organism's movement was simulated to be range resident within an homogeneous environment.
-Since real-world landscapes are spatially heterogeneous in both $\e(R)$ and $\var(R)$, organisms may choose to shift their home range to a new area following a stochastic event that caused $\mu(t)$ to decrease, such as a fire or a flood.
-Thus, we hypothesize that our simulations may be underestimating organisms' responses to spatiotemporal changes in $\e(R)$ and $\var(R)$, since the simulated organisms cannot select for areas of higher $\e(R)$ or lower $\var(R)$.
-Additionally, the simulations also ignore how competition, movement costs, predation, and other factors may affect the effects of $\e(R)$ and $\var(R)$ on $H$.
-While these are definitely important effects that should be accounted for [@jetz_scaling_2004], doing so was beyond the scope of this paper.
+\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time (panels A and B of fig.
+\@ref(fig:5-5-reg-mean)).
+The estimated relationships follow the hypotheses we presented in figs.
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation as $\var(R)$ increased.<!--We ensured this was not due to a high minimum or low maximum $\var(R)$ (fig.
+\@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C).
+The increase in $\e(H)$ over $\var(R)$ is strongly nonlinear at low values of $\e(R)$, but it is not affected by $\var(R)$ much when $\e(R)$ is very low (and $\e(H)$ is consequently very high).
 
-```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% credible intervals for the mean (overlapping the line).
-Credible intervals were calculated assuming a gaussian posterior distribution on the link scale.
-The relationships were estimated using a Generalized Additive Model for Location and Scale (GAMLS; Rigby and Stasinopoulos 2005; Stasinopoulos and Rigby 2007) with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
-The model accounted for the effects of $\\e(R)$ and $\\var(R)$ and the interaction effects between the two on both the location (i.e., mean) of $H$ and its scale parameter.
-Note the nonlinear decrease in $H$ as $\\e(R)$ increases and the increase in $H$ as $\\var(R)$ increases.
-Additionally, note how the variance in space-use requirements strongly increases with mean space-use requirements.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-plots.png')
+<!-- While we modeled the effects of $\e(R)$ and $\var(R)$ on both the 95% and the 50% space-use requirements, we only show results for the 95^th^ percentile because the results the 50^th^ quantile followed the same trends, other than the multiplicative rescaling required to account for the difference in magnitude between the two quantiles.
+-->
+
+```{r 5-5-reg-mean, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% Bayesian credible intervals for the mean.
+While the effect of $\\var(R)$ is sublinear in panel B, the effect of $\\var(R)$ depends strongly on $\\e(R)$ (see panel C).\nCredible intervals were calculated using 10,000 samples from the posterior distribution.
+The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
+Details on the model structure are provided in Appendix B.", out.width='100%'}
+knitr::include_graphics('../figures/simulation-regression-mean.png')
 ```
 
-***to do***:
+As expected by the changes in the spread of the points in panels A and B of figure \@ref(fig:5-5-reg-mean), the variance in $H$ also depends on $\e(R)$ and $\var(R)$ (figure \@ref(fig:5-5-reg-var)A-B).
+Here, the effects of the two variables are more pronounced, even for low values of $\e(R)$ or high values of $\var(R)$.
+Consequences of these effects are explored in the discussion section.
 
-* add a paragraph and figure on effects of $\mu(t)$ and $\sigma^2(t)$ on $\var(R)$.
+```{r 5-5-reg-var, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on the variance in simulated spatial requirements, with 95\\% Bayesian credible intervals for the estimated variance.
+Credible intervals were calculated using 10,000 samples from the posterior distribution.
+Additional details on the model structure are provided in Appendix B.", out.width='100%'}
+knitr::include_graphics('../figures/simulation-regression-var.png')
+```
 
-# A lowland tapir in the Brazilian Cerrado: a case study
+# A case study on a lowland tapir in the Brazilian Cerrado
 
-***to do***:
+\noindent The simulations in the section above mostly follow the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural enviroments.
+Organisms live in spatiotemporally heterogeneous and dynamic environments, which the development of perceptual ranges, navigation, and memory.
+Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].
 
-* open with limitations of sims
-* then highlight shortcomings of empirical data
-* move trasnformation stuff to the appendix
-* move modeling details to appendix C
+In this section, we test the hypotheses illustrated in the introduction using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with estimates of $\e(R)$ and $\var(R)$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], a remote-sensed measure of lanscape greeness, as a proxy of forage abundance.
+Appendix C contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
+Through this example, we aim to provide support for the two hypotheses using empirical data that do not depend on as strong assumptions as those required by the simulations in the section above.
+Concurrently, we also hope to show the potential complications that arise with empirical data.
 
-\noindent In this section, we show how this framework can be applied to better understand the movement ecology of free-ranging organisms via remote sensing data such as NDVI [@pettorelli_normalized_2011].
-To illustrate the methods, we use empirical tracking data on a lowland tapir from the Brazilian Cerrado.
-Appendix C contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
-
-The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., fig.
+<!-- The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., fig.
 \@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
@@
 In the following sections, we illustrate how one can apply this framework to empirical data.
-
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements.
-Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' space-use requirements under different scenarios.
-
-### Modeling $R$ {#modeling-r}
-
-\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
-its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
-
-The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
-However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].
-
-The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->.
-The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009].
-In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
-If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
-
-Modeling the mean and variance terms of $R$ should be done carefully.
-Since trends in both $\e(R)$ and $\var(R)$ can be spatiotemporally nonlinear and non-monotonic, we suggest using a GAM rather than a GLM.
-However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms.
-An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
-Although there is no error-proof system, choosing the complexity of the terms based on the organism's ability to detect change and adapt is a reasonable starting point.
-Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
-We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model.
-@simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.
-
-### Estimating $R$ using NDVI
-
-\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation
-
-\begin{equation}
-\nu_{[0, 1]} = \frac{\nu + 1}{2},
-(\#eq:ndvi-to-01)
-\end{equation}
-
-\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
-<!--Note that only linear transformations (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
-We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
-
-\begin{equation}
-\e(R) = \frac{\alpha}{\alpha + \beta}
-(\#eq:beta-mean)
-\end{equation}
-
-\noindent while the variance is
-
-\begin{equation}
-\var(R) = \frac{\alpha\beta}{(\alpha+\beta)^2(\alpha+\beta+1)}.
-(\#eq:beta-var)
-\end{equation}
-
-\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):
-
-\singlespace <!-- to avoid excessive spacing -->
-\begin{equation}
-\nu = 2 \nu_{[0, 1]} - 1.
-(\#eq:01-to-ndvi)
-\end{equation}
-
-\noindent Consequently, we have
-
-\begin{equation}
-\e(\nu|t) = 2 \mu(t) - 1.
-(\#eq:mu-to-e-ndvi)
-\end{equation}
-
-\noindent and
-
-\begin{equation}
-\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
-(\#eq:sigma2-to-v-ndvi)
-\end{equation}
-
-\doublespace <!-- back to double spaced -->
-
-Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
-Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
-
-<!--
-## Movement as a stochastic process
-
-### Drivers of movement
-
-\noindent @nathan_movement_2008 suggested we can generalize movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
-
-\begin{equation}
-\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
-(\#eq:nathan-movement)
-\end{equation}
-
-\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
-This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on organisms' movement and space-use requirements as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
-
-### Visualizing movement as a stochastic process
-
-\noindent To understand how movement can be viewed as a stochastic process, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
-If we do not know why the organism is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the organism may move from one time point to another.
-However, we can use the organism's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
-This period is often referred to as the range crossing time or the positional autocorrelation parameter, $\tau_p$.
-Similarly, we can define $\tau_v$ to be the amount of time the organism moves with the same direction and speed, or the time required for an organism's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
-$\tau_v$ is referred to the organism's directional persistence or the velocity autocorrelation parameter.
-
-By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data.
-While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].
-
-Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
-However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
-Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
-
-Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on how the organism moved and less on why it moved.
-Organisms may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
-Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in organism movement while focusing on the organisms' movement behaviors rather than the causes.
-
-### Quantifying space-use requirements
-
-\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify organism's space-use requirements using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
-In the case of organisms that are range resident for short periods of time, we can model their movement and short-term space-use requirements within distinct and discrete periods or using a sliding window approach (see the discussion section).
-Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
-Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
-Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
 -->
 
-<!-- ### Modeling space-use requirements -->
-
-<!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
-However, most commonly used movement models assume organisms' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
--->
-
-<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a sliding window approach.
-The width of the window, $w$, should be a fixed value greater than two or three times the organism's expected range crossing time [@fleming_overcoming_2019].
-If the organism's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
-Alternatively, if the organism's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
--->
-
-<!-- Modeling movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
-In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023].
--->
-
-### Modeling the effects of $\e(R)$ and $\var(R)$ on space-use requirements
-
-\noindent As discussed in the previous section, the first step to modeling $H$ is to choose an appropriate family of distributions.
-Since $H$ is strictly positive and continuous, a Gamma family seems the most appropriate among the distributions currently offered by the `mgcv` package [@wood_generalized_2017], but the Tweedie family would also be appropriate if the variance does not scale with the squared mean.
-
-As with the simulated effects of $\mu(t)$ and $\sigma^2(t)$ on $H$ (fig.
-\@ref(fig:5-5-reg)), a location-scale Gamma model is likely required.
-If there is not sufficient data to fit a location-scale model (which is not the case here), a simple Gamma model may be sufficient, but we suggest selecting the best family of distributions using an information criterion such as the Akaike Information Criterion [@akaike_new_1974].
-Due to potential correlation between $\e(R)$ and $\var(R)$ in a small and autocorrelated sample, the relationship between $R$ and $H$ should be modeled carefully and with parsimony.
-Below, we present an empirical example using GPS tracking data from a lowland tapir (*Tapirus terrestris*) and satellite-derived NDVI.
-Additional information can be found in Appendix C.
-
-### The effects of $\e(R)$ and $\var(R)$ on a lowland tapir's space-use requirements
-
 \noindent Fig.
-\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from the individual labelled as "Anna" in the dataset from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
+The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007, examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
 Panel C shows the changes in the tapir's 7-day home range over time.
@@
 Finally, panels D and E show the estimated (marginal) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Simce $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel F) and (spatio)temporally autocorrelated (panels A and B), the effects of $R$ on $H$ should be modeled carefully.
-To avoid over-fitting the model, we constrain the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using a shape-constrained additive models (SCAM) with monotone deceasing and increasing P-splines for $\mu(t)$ and $\sigma^2(t)$, respectively.
-
-Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and a worse model fit ($\Delta\text{AIC} \approx 87$, see Appendix C).
-By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity.
-While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
-\@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period.
-Consequently, this example demonstrates two points central to this paper.
-Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity.
-Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.
+Since $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel F) and (spatio)temporally autocorrelated (panels A and B), the effects of $R$ on $H$ should be modeled carefully.
+To avoid over-fitting the model, we constrained the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using a shape-constrained additive model [@pya_scam_2022] with monotone deceasing and increasing P-splines for $\mu(t)$ and $\sigma^2(t)$, respectively.
+Additional information is provided in appendix C.
+Overall, the results presented in fig.
+\@ref(fig:tapir-mw) support our hypotheses, as the taprir's 7-day home range decreases nonlinearly with $\mu(t)$ and increases nonlinearly with $\sigma^2(t)$.
+However, due to the strong nonlinear correlation between the two predictors, the marginal effects do not follow the data closely, particularly in the case of $\mu(t)$.
+We provide comments about these results in the related section in the discussion below.
 
@@
 The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
-knitr::include_graphics('../figures/tapir-example-with-data.png')
+knitr::include_graphics('../figures/tapir-example.png')
 ```
@@
 
-***to do***:
+\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on organisms' space-use requirements.
+We support the hypotheses using quantitative simulations and an empirical case study.
+Each of the two methods allows us to highlight the strengths and limitations of simulation-based and data-driven approaches, which we discuss below.
 
-* add another intro paragraph to tie back to general idea and connecting to previous theory
-* provided hyp, sims, empirical examples, framework to test ideas
+## Strengths and limitations of the simulation-based approach
 
-\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on space-use requirements.
+\noindent The simulations we present here provide a simple and direct estimate of the effects of resource abundance and stochasticity on organisms' space-use requirements.
+The combination of simplicity (few drivers of movement) and realism (continuous-time movement models) allowed us to estimate the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements using simple causal models of inference without the confounding variables that are often present in empirical data (e.g., predator avoidance, territoriality, competition, landscape connectivity).
+Additionally, since our sample size was virtually unlimited, we were able to simulate sufficient data to ensure the simulations were stable and the resulting estimated coefficients were well estimated (see Appendix B).
+
+Avoiding confounding variables rather than attempting to account for them (as is required with empirical data) allowed us to develop theory easily while providing a baseline model against which one can contrast empirical examples.
+In figures B6, we present the Directed Acyclical Graph [DAG, see @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct, partial effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with figure C5 and the section below).
+
+By quantifying the deviations from the simple theoretical models, one can then assess the importance of other drivers movement and behavior.
+Still, the simplicity of the underlying model and its (often invalid) assumptions often limit the applicability of simulated models in realistic scenarios.
+Responses to $\e(R)$ and $\var(R)$ in real ecosystems likely depend on many other drivers of movement, including competition, movement costs, and predation, which may affect the effects of $\e(R)$ and $\var(R)$ on both $\e(H)$ and $\var(H)$ [@jetz_scaling_2004; @noonan_search_2023].
+And while it is possible to account for all of these factors through simulations, empirical data often can provide evidence for these effects without the need for complex model structures.
+
+## Strengths and limitations of the empirical approach
+
+\noindent In the absence of direct measurements of $R$, empirical models estimating the effects of $R$ on $H$ must depend on proxies of $R$, such as NDVI.
+However, using proxies of $R$ rather than direct measurements introduces sources of bias.
+While $R$ and NDVI are correlated [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019] and nonlinear [@ref].
+This is because remote-sensed proxies of $R$ and they do not measure resource availability directly, so they introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the DAG in figure C5) and satellite-level confounding variables ($S$ in figure C5).
+
+Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and $NDVI$ [@jetz_scaling_2004; @prox_framework_2020].
+Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999].
+While such pressures may increase avoidance behaviors, the behaviors may result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020]
+
+***here***
+
+, but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
+Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
+\@ref(fig:mu-hyp).
+
+Due to data limitations, disentangling the effects of such confounds is often complex or impossible [e.g., @wang_stochastic_2019].
+
+Satellite-level confounds include coarse spatiotemporal resolution [and the consequent aggregation of signals, see @ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing, including the inability to quantify the abundance of specific resources, whether these be preferred forage or the distribution of prey.
+In the case of preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @], nonlinear models such as Generalized Additive Models can help account for a nonmonotonic relationship between $H$ and $E(NDVI)$, as long as data are sufficiently abundant to avoid any artifacts due to data paucity.
+In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$, which prevented us from estimating any nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$).
+A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this paper.
+
+Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and a quantifiably worse model fit ($\Delta\text{AIC} \approx 87$, see Appendix C).
+By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity.
+While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
+\@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period.
+Consequently, this example demonstrates two points central to this paper.
+Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity.
+Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.
 
@@
 
-* It is also worth noting that high $\e(R)$ does not does not necessarily imply that organisms have access to high amounts of resources, as predator avoidance and competition within an organism's habitat will decrease the proportion of $R$ it is able to access.
-Thus, organisms in habitats with strong competitive pressure and predation may require larger home ranges [@jetz_scaling_2004; @prox_framework_2020], but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
-\@ref(fig:mu-hyp).
 * @rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation divided by the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
 However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability [measured as spatial semivariance in Normalized Difference Vegetation Index, NDVI; see @pettorelli_normalized_2011] causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
+Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
 Likewise, @herfindal_prey_2005 found that the home ranges of Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
@@
 
-## Applications for conservation
+<!-- ## Applications for conservation -->
 
-\noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on organisms' behaviors.
-In these times of great change and climatic uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021], it is increasingly important to recognize how environmental stochasticity affects organisms' behaviors and shapes ecosystems.
-Accounting for changes in both resource abundance and resource stochasticity is crucial when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
-Protecting areas with high resource abundance and low resource stochasticity will reduce animals' spatial requirements.
+<!-- \noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on organisms' behaviors.
+In these times of great change and uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021; @sih_evolution_2011], it is increasingly important to recognize how environmental stochasticity affects organisms' behaviors and shapes ecosystems, particularly when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
+Protecting areas with high resource abundance and low resource stochasticity is an important first step towards reducing animals' physiological stress, inter- and intra-specific competition, community instability, and human-wildlife conflict.
+-->
 
-The designation of high-quality habitats is a central concern of many conservation initiatives.
+<!-- The designation of high-quality habitats is a central concern of many conservation initiatives.
 This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat.
@@
 In light of the UN's objective to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
+-->
 
-Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
+<!-- Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
 Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' space-use requirements, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
 This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in space-use requirements.
+-->
 
-It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels.
+<!-- It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels.
 While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity.
@@
 However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less stochastic resources.
+-->
 

### 2024-03-19 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R
numbered scripts

@@ -100,18 +100,19 @@ m <- gam(hr_est_95 ~ s(mu, bs = 'ts', k = 5) + s(sigma2, bs = 'ts', k = 5),
          weights = weight, # weights based on CI size
          data = tapir,
          method = 'REML')
-# trends should be monotonic...
+#' non monotonicity is likely an artifact of autocorrelation in `mu` and
+#' `sigma2` and poor data availability rather than an actual effect
 plot(m, pages = 1, scheme = 1)
 
 # using shape-constrained splines to avoid artifacts due to correlation
-m_mu <- gam(hr_est_95 ~ s(mu, k = 3),
+m_mu <- gam(hr_est_95 ~ s(mu, k = 4),
             family = Gamma('log'),
             weights = weight,
             data = tapir,
             method = 'REML')
 plot(m_mu, trans = exp, scheme = 1)
 
-m_sigma2 <- gam(hr_est_95 ~ s(sigma2, k = 3),
+m_sigma2 <- gam(hr_est_95 ~ s(sigma2, k = 4),
                 family = Gamma('log'),
                 weights = weight,
                 data = tapir,
@@ -123,9 +124,9 @@ m <- gam(hr_est_95 ~ mu + s(sigma2, k = 4), # linear mu bc of autocorrelation
          weights = weight,
          data = tapir,
          method = 'REML')
-plot(m, scheme = 3, too.far = 1, pages = 1, all.terms = TRUE)
-AIC(m_mu, m_sigma2, m) # model that accounts for both fits best
-BIC(m_mu, m_sigma2, m) # model that accounts for both fits best
+plot(m, scheme = 1, pages = 1, all.terms = TRUE)
+arrange(AIC(m_mu, m_sigma2, m), AIC) # m is best
+arrange(BIC(m_mu, m_sigma2, m), BIC) # m is best
 
 # marginals of mu and sigma2, and a plot to show the lack of independence
 preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
@@ -169,6 +170,18 @@ p_f <- ggplot(tapir) +
   geom_point(aes(mu, sigma2, alpha = weight)) +
   labs(x = e_r, y = v_r) +
   scale_alpha_continuous(range = c(0.3, 1))
+
+expand_grid(mu = gratia:::seq_min_max(tapir$mu, n = 250),
[…]

### 2024-03-19 analysis/simulations/6-modeling-R-and-hr.R
improved models, predictions, and color palettes

@@ -31,21 +31,27 @@ m_1 <- gam(
     # predictor for the scale (variance = scale * mean)
     ~ quantile + mu + sigma2),
   family = gammals(),
-  data = sims)
+  data = sims,
+  method = 'REML')
+plot(m_1, pages = 1, all.terms = TRUE) # acceptable but likely too rigid
 
 m_2 <- gam(
   list(
     hr ~ quantile + s(mu) + s(sigma2),
     ~ quantile + s(mu) + s(sigma2)),
   family = gammals(),
-  data = sims)
+  data = sims,
+  method = 'REML')
+plot(m_2, pages = 1, all.terms = TRUE) # scale terms are inappropriate
 
 m_3 <- gam(
   list(
     hr ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5),
     ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5)),
   family = gammals(),
-  data = sims)
+  data = sims,
+  method = 'REML')
+plot(m_3, pages = 1, all.terms = TRUE, scheme = 3) # much better
 
 # check if CV = sigma/mu is better than sigma
 m_4 <- gam(
@@ -53,9 +59,9 @@ m_4 <- gam(
     hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
     ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
   family = gammals(),
-  data = mutate(sims, cv = sqrt(sigma2) / mu))
-
-# ti term of the scale parameter looks over-fit and too uncertain
+  data = mutate(sims, cv = sqrt(sigma2) / mu),
+  method = 'REML')
+# ti term of the scale parameter looks over-fit
 plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
 
 AIC(m_1, m_2, m_3, m_4) # smooth model with interactions is best
@@ -72,7 +78,7 @@ plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
 # filter to 95% HR for simplicity
 sims <- filter(sims, quantile == 'hr_95')
 
-# predictions
+# predictions for calculating (squared) residuals
 sims$hr_est_95 <- predict(m, newdata = sims, type = 'response',
                           se.fit = FALSE)[, 1]
 
@@ -118,8 +124,7 @@ p_mu_var <-
[…]

### 2024-03-19 writing/manuscript.Rmd
finished editing discussion

@@
 
-\noindent \textbf{References}: 141 <!-- counted by using nature csl -->
+\noindent \textbf{References}: 95 <!-- counted by using nature csl -->
 
@@
 * make sure appendices flow (added stuff from main text)
-* check number of references
+* check number of references (updated on 2024-03-18)
 
@@
 Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999].
-While such pressures may increase avoidance behaviors, the behaviors may result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020]
+While such pressures may increase avoidance behaviors, avoidance may have a cmplex effect on $H$, since it can result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020] or a contaction of an animal's movement, since larger home ranges can be harder to defend and also result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [theoretical framework: @despotic_ref; examples: @rich_anthropogenic_2012; @bengsen_feral_2016; @torrez-herrera_monkeys_2020].
 
-***here***
+Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources, competitive pressure, and predator avoidance).
+However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @ref].
+In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$ that prevented us from estimating nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$), since estimating the divers of movement and disentangling them from potential confounds in a stochastic environemtn is often complex or impossible [e.g., @wang_stochastic_2019].
+A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this empirical example.
 
-, but larger home ranges can be harder to defend and also result in higher rates of competition and movement costs [@grant_whether_1993; @jetz_scaling_2004; but also see @dickie_resource_2022].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [wolves, *Canis lupus*: @rich_anthropogenic_2012; feral cats, *Felis catus*: @bengsen_feral_2016; Capuchin monkeys, *Cebus capucinus*: @torrez-herrera_monkeys_2020], so the effect of $\e(R)$ on space use may vary between individuals, species, and locations.
-Still, these pressures simply alter the intensity with which organisms respond to $\e(R)$ but not the shape of the relationship, since each of these cases fall along different parts of the continuum in $\e(R)$ shown in fig.
-\@ref(fig:mu-hyp).
+**here:** need a few lines to tie this back to the general
 
-Due to data limitations, disentangling the effects of such confounds is often complex or impossible [e.g., @wang_stochastic_2019].
-
-Satellite-level confounds include coarse spatiotemporal resolution [and the consequent aggregation of signals, see @ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing, including the inability to quantify the abundance of specific resources, whether these be preferred forage or the distribution of prey.
-In the case of preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @], nonlinear models such as Generalized Additive Models can help account for a nonmonotonic relationship between $H$ and $E(NDVI)$, as long as data are sufficiently abundant to avoid any artifacts due to data paucity.
-In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$, which prevented us from estimating any nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$).
-A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this paper.
-
-Modeling the tapir's space use as a function of resource abundance alone would result in an excessively strong estimated effect and a quantifiably worse model fit ($\Delta\text{AIC} \approx 87$, see Appendix C).
-By including the effect of $\sigma^2(t)$ in the model along with that of $\mu(t)$, we show that the tapir responds to both resource abundance and stochasticity.
-While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
+<!-- Feel free to delete this: -->
+<!-- While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
 \@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period.
@@
 Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.
+-->
 
+<!-- 
 ## Adaptations to changes in $\e(R)$ and $\var(R)$
@@
 
-<!-- spores are also an adaptation to periods of low $\e(R)$ or high $\var(R)$ -->
-
-\noindent The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so).
+The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so).
 Events that occur at the scale of the organism [i.e.
@@
 Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
+-->
+
 <!-- 
@@
 Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
-However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
+<!-- However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
 
@@
 
-\noindent The work presented here provides a unifying framework for viewing movement as a function of resource abundance and stochasticity.
+**needs a few more lines at the end**
+
+\noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
 We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior.
 We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.
-The methods presented here can be applied to assess the current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
+We also provide a moving-window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.
+The methods presented here can be applied to assess both current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
 

### 2024-03-28 writing/appendix-B-simulations.Rmd
updated appendices and methods for empirical model

@@
 library('ggdag')   # for directed acyclical graphs
+library('gratia')  # for ggplot-based GAM figures
 theme_set(theme_bw()) # change default theme
@@
 
-```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=4}
+```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=6, out.width='50%'}
 dag <- dagitty('dag{
@@
 
-```{r, eval=TRUE, fig.height=8, fig.width=8, out.width='65%'}
+```{r, eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth'}
 # marginal effects of mu and sigma only (linear on the log link scale)
@@
 # ti term of the scale parameter looks over-fit and too uncertain
-plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+draw(m_4) & theme_bw() + theme(panel.grid = element_blank())
 ```
@@
 The dashed lines indicate the 95\\% Bayesian credible intervals.
-The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=16, fig.width=8, out.width='65%', echo=FALSE}
+The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
 # using custom layout to make figure easier to read
 layout(matrix(c(7, 1, 2, 3, 8, 4, 5, 6), nrow = 4, byrow = TRUE))
-plot(m, all.terms = TRUE, scheme = 3, scale = 0)
+draw(m) & theme_bw() + theme(panel.grid = element_blank())
 ```

### 2024-03-28 writing/appendix-C-empirical-modeling.Rmd
updated appendices and methods for empirical model

@@
 
-\newpage
-
 \tableofcontents
@@
 
-\noindent Location-scale models [@stasinopoulos_generalized_2007; @rigby_generalized_2005] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
+\noindent Location-scale models [theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] are a class of statistical models that allow us to estimate changes in a random variable's mean (i.e.
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
@@
 
-\noindent Since there is no commonly-used distribution with a support over the interval $[-1, 1]$, we use beta distribution after scaling NDVI to the interval $[0, 1]$ by applying the linear transformation
+\noindent Although there is no commonly-used distribution with a support over the full range of NDVI (i.e., $[-1, 1]$), we use beta distribution (support (0, 1)) since the all NDVI values were sufficiently greater than 0 (range: 0.3534 to 0.9475).
+Thus, here we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
 
-\begin{equation}
-\nu_{[0, 1]} = \frac{\nu + 1}{2},
-(\#eq:ndvi-to-01)
-\end{equation}
-
-\noindent where $\nu$ indicates the NDVI value in $[-1, 1]$ and $\nu_{[0, 1]}$ is the scaled NDVI value [see @denny_fallacy_2017 for more information on transforming response data].
-<!--Note that only linear transformations (i.e.
-only addition, subtraction, multiplication, and division) should be applied to random variables to avoid biases while estimating the distribution and its parameters [@denny_fallacy_2017; @jensen_sur_1906; @ohara_not_2010].--> In this case, we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
-We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
+\singlespace
 
@@
 
-\noindent We can easily convert $\mu(t)$ and $\sigma^2(t)$ back to mean and variance in NDVI using the inverse of equation \@ref(eq:ndvi-to-01):
-
-\singlespace <!-- to avoid excessive spacing -->
-\begin{equation}
-\nu = 2 \nu_{[0, 1]} - 1.
-(\#eq:01-to-ndvi)
-\end{equation}
-
-\noindent Consequently, we have
-
-\begin{equation}
-\e(\nu|t) = 2 \mu(t) - 1.
-(\#eq:mu-to-e-ndvi)
-\end{equation}
-
-\noindent and
-
-\begin{equation}
-\var(\nu|t) = 2^2 \sigma^2(t) + 0 = 4 \sigma^2(t).
-(\#eq:sigma2-to-v-ndvi)
-\end{equation}
-
 \doublespace <!-- back to double spaced -->
@@
 Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
+The following section addresses these concerns.
 
@@
 The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and spatial use (section \@ref(final-fig)).
+Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).
 
@@
 
-\newpage
-
 We start by attaching all necessary packages and custom functions we need:
 
-```{r, eval=FALSE}
-getwd() # change working directory or file paths if needed
-```
-
-```{r, eval=TRUE, echo=FALSE}
-print('H:/GitHub/hr-resource-stoch/writing') # for anonymity in peer review
-```
-
-```{r, eval=TRUE}
+```{r, eval=TRUE, warning=FALSE}
 # attach all necessary packages
@@
 library('mgcv')      # for empirical Bayesian GAMs
-library('scam')      # for shape-constrained splines
 library('lubridate') # for smoother date wrangling
@@
 library('ggdag')     # for directed acyclical graphs
+library('gratia')    # for ggplot-based GAM figures
 theme_set(theme_bw()) # change default theme
-
-# attach the beta location-scale family written by Simon Wood
-source('../functions/betals.r')
-
-# source all necessary scripts
+source('../functions/betals.r') # betals family written by Simon Wood
 source('../analysis/figures/default-figure-styling.R') # for color palettes
@@
          verbose = TRUE) # print processing messages
+```
 
+\newpage
+
+```{r}
 # save NDVI data as an rds file of a tibble
@@
 We removed the raster for 2017-12-19 because the values were unusually low for the region.
-We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood (which is common for the Cerrado):
+We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood.
+While sudden floods are common for the Cerrado, we believe NDVI was not representative of the available forage availability.
 
@@
     ndvi ~ # not scaling because range is in (0, 1)
-      s(long, lat, bs = 'ds', k = 20) + # mean over space
-      s(dec_date, bs = 'tp', k = 20),   # high k to account for adaptions
+      s(long, lat, bs = 'ds', k = 50) + # mean over space
+      s(dec_date, bs = 'tp', k = 10), # mean over time
     # scale predictor (sigma2 = mu * (1 - mu) * scale)
     ~
-      s(long, lat, bs = 'ds', k = 10) + # scale over space
+      s(long, lat, bs = 'ds', k = 30) + # scale over space
       s(dec_date, bs = 'tp', k = 10)),  # scale over time
@@
 Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term.
-While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, finding the right balance between each of the `k` values is crucial.
-Excessively smooth terms for the mean can inflate the scale term (fig.
-C\@ref(fig:undersmooth)), while excessively wiggly terms for the mean can cause the scale to be under-estimated (fig.
-C\@ref(fig:oversmooth)).
+While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, it does not prevent over-fitting, so finding the right balance between each of the `k` values is crucial.
+Excessively smooth terms for the mean can inflate the scale term, while excessively wiggly terms for the mean can cause the scale to be under-estimated.
 Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
@@
 The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms.", fig.pos='!h'}
+Shaded ribbons inticate the 95\\% credible intervals for the temporal terms.", fig.height=6, fig.pos='!h'}
 m_ndvi <- readRDS('../models/tapirs/CE_31_ANNA-mgcv-ndvi-betals.rds')
-plot(m_ndvi, pages = 1, scheme = 3, n = 250, scale = 0)
-```
-
-```{r}
-gam(list(
-  # mean predictor (default ks are too small)
-  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
-  # scale predictor (excessively high k)
-  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
-  family = betals(), data = anna_ndvi, method = 'REML') %>%
-  plot(pages = 1, scheme = 3, n = 250, scale = 0)
-```
-
-```{r undersmooth, fig.cap="Estimated mean and scale parameters with excessively small basis size $\\tt{k}$ for the mean's smooth terms and excessively large for the scale's smooth terms.
-The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
-gam(list(
-  # mean predictor (default ks are too small)
-  ndvi ~ s(long, lat, bs = 'ds', k = 10) + s(dec_date, bs = 'tp', k = 10),
-  # scale predictor (excessively high k)
-  ~ s(long, lat, bs = 'ds', k = 20) + s(dec_date, bs = 'tp', k = 20)),
-  family = betals(), data = anna_ndvi, method = 'REML') %>%
-  plot(pages = 1, scheme = 3, n = 250, scale = 0)
-```
-
-\clearpage
-
-```{r}
-gam(list(
-  # mean predictor (ks too large)
-  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
-    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
-  # scale predictor (ks too small)
-  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
-  family = betals(), data = anna_ndvi, method = 'REML') %>%
-  plot(pages = 1, scheme = 3, n = 250, scale = 0)
-```
-
-```{r oversmooth, fig.cap="Estimated mean and scale parameters in NDVI with excessively large basis size $\\tt{k}$ for the smooth term of time ($\\tt{dec\\_date}$) for the mean in NDVI and excessively low for the variance temporal smooth term.
-The estimated degrees of freedom for each term can be seen in parentheses in the title of the spatial terms and the y-axis labels of the temporal terms.
-Dashed lines inticate the 95\\% credible intervals for the temporal terms.", eval=TRUE, fig.height=6, fig.pos='!h', echo=FALSE}
-gam(list(
-  # mean predictor (ks too large)
-  ndvi ~ s(long, lat, bs = 'ds', k = 50) +
-    s(dec_date, bs = 'tp', k = n_distinct(anna_ndvi$dec_date)),
-  # scale predictor (ks too small)
-  ~ s(long, lat, bs = 'ds', k = 5) + s(dec_date, bs = 'tp', k = 5)),
-  family = betals(), data = anna_ndvi, method = 'REML') %>%
-  plot(pages = 1, scheme = 3, n = 250, scale = 0)
+draw(m_ndvi, rug = FALSE) &
+  theme_bw() + theme(panel.grid = element_blank())
 ```
@@
 Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period.
-We also add a column of weights that are based on the home range confidence estimates' 95% intervals.
 
@@
 tapir <-
-  readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
-  mutate(est = map(dataset,
-                   \(.d) filter(anna_tel, timestamp %in% .d$timestamp)),
-         mu = map_dbl(est, \(.d) mean(.d$mu)),
-         sigma2 = map_dbl(est, \(.d) mean(.d$sigma2))) %>%
-  select(date, mu, sigma2, hr_lwr_95, hr_est_95, hr_upr_95) %>%
-  mutate(weight = 1 / (hr_upr_95 - hr_lwr_95), # small CI => large weight
-         weight = weight / mean(weight)) # standardize weights
-sum(tapir$weight) == nrow(tapir) # ensure weights are right
-```
-
-```{r, eval=TRUE, echo=FALSE}
-tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
-sum(tapir$weight) == nrow(tapir) # ensure weights are right
+  readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+  mutate(sub_tel = map(dataset,
+                       \(.d) filter(tel, timestamp %in% .d$timestamp)),
+         mu = map_dbl(sub_tel, \(.d) mean(.d$mu)),
+         sigma2 = map_dbl(sub_tel, \(.d) mean(.d$sigma2))) %>%
+  select(date, mu, sigma2, hr_est_95)
 ```
@@
 
+```{r, eval=TRUE, echo=FALSE}
+tapir <- readRDS('../data/anna-hr-ndvi-data.rds')
+```
+
+\clearpage
+
 ```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
-theme_set(theme_bw()) # need to set the theme again for some reason
-# axis labels (spaces in hr_lab to avoid overlap with the plot label)
-# unicode characters removed because they knitting errors
-e_r <- bquote(paste(bold('Resource abundance, E('), bolditalic(R),
-                    bold(')')))
-v_r <- bquote(paste(bold('Resource stochasticity, Var('), bolditalic(R),
-                    bold(')')))
-hr_lab <- bquote(bold(paste('7-day home range size (', km^2, ')   ')))
+# need to set the theme again for some reason
+theme_set(theme_bw() + theme(panel.grid = element_blank()))
+
+# axis labels (unicode characters removed because they knitting errors)
+e_r <- bquote(paste(bold('Resource abundance')))
+v_r <- bquote(paste(bold('Resource stochasticity')))
+hr_lab <- bquote(bold(paste('7-day space-use requirement (', km^2, ')')))
 
@@
   list(ggplot(tapir, aes(date, mu)) + # mean
-         geom_line(color = pal[1], size = 2) +
+         geom_line(color = pal[1], linewidth = 2) +
          labs(x = NULL, y = e_r),
        ggplot(tapir, aes(date, sigma2)) + # variance
-         geom_line(color = pal[2], size = 2) +
+         geom_line(color = pal[2], linewidth = 2) +
          labs(x = NULL, y = v_r),
        ggplot(tapir, aes(date, hr_est_95)) + # 95% home range
-         geom_line(color = pal[3], size = 2) +
+         geom_line(color = pal[3], linewidth = 2) +
          labs(x = NULL, y = hr_lab)),
@@
 
-<!-- \clearpage -->
+\clearpage
 
@@
 ```{r dag, eval=TRUE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
-$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=4}
+$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=6, out.width='50%'}
 dag <- dagitty('dag{
@@
                          y = c(R = 0, 'NDVI' = 1, Z = 0, S = 1, H = -1))
-plot(dag)
+ggdag(dag) + theme_dag()
 ```
@@
 
-```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a GAM with no shape constraints.
+```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale).
 The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
-Shaded areas inticate the 95\\% credible intervals."}
-gam(hr_est_95 ~ s(mu, bs = 'ts', k = 5) + s(sigma2, bs = 'ts', k = 5),
-    family = Gamma('log'),
-    weights = weight, # weights based on CI size
-    data = tapir,
-    method = 'REML') %>%
-  plot(pages = 1, scheme = 1) # trends should be monotonic
+Shaded areas inticate the 95\\% credible intervals.", fig.height=3}
+m <- gam(hr_est_95 ~ s(mu, k = 4) + s(sigma2, k = 4),
+         family = Gamma('log'), data = tapir, method = 'REML')
+draw(m) & theme_bw() + theme(panel.grid = element_blank())
 ```
 
-\clearpage
+We can now create the predictions from the GAM.
 
-Due to the small sample size, the spatiotemporal autocorrelation, and the correlation between $\mu(t)$ and $\sigma^2(t)$, it is hard to separate the effects of the two predictors.
-However, since it is reasonable to expect the effects of $\mu(t)$ and $\sigma^2(t)$ to be monotonic decreaseing and increasing, respectively, we can use a Shape-Constrained Additive Model (SCAM) via the `scam` package [@pya_scam_2022].
-
-```{r hr-scam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale) using a shape-constrained GAM.
-The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
-Shaded areas inticate the 95\\% credible intervals."}
-m <- scam(hr_est_95 ~
-            s(mu, bs = 'mpd', k = 5) +
-            s(sigma2, bs = 'mpi', k = 5),
-          family = Gamma('log'),
-          weights = weight,
-          data = tapir)
-plot(m, pages = 1, scheme = 1)
-```
-
-This SCAM fits substantially better than a SCAM that only accounts for $\mu(t)$:
-
-```{r, eval=TRUE}
-m_mu <- scam(hr_est_95 ~ s(mu, bs = 'mpd', k = 5),
-             family = Gamma('log'),
-             weights = weight,
-             data = tapir)
-round(AIC(m, m_mu), 3)
-```
-
-We can now create the predictions from the shape-constrained GAM.
-
-```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
-theme_set(theme_bw()) # need to set the theme again for some reason
+```{r, eval=TRUE, out.width='\\linewidth'}
 # predict marginal effects for each term
 preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
-                sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
+                sigma2 = seq(5e-4, 25e-4, length.out = 250)) %>%
   bind_cols(
     # predictions for the marginal effect of mu
-    predict(m, newdata = ., terms = 's(mu)', type = 'link',
-            se.fit = TRUE) %>%
-      as.data.frame() %>% # convert list to data frame
-      transmute(hr_mu_est = exp(fit), # transform to the response scale
-                hr_mu_lwr = exp(fit - 1.96 * se.fit), # 95% cred. intervals
+    predict(m, newdata = ., exclude = 's(sigma2)',
+            type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
+      as.data.frame() %>%
+      transmute(hr_mu_est = exp(fit),
+                hr_mu_lwr = exp(fit - 1.96 * se.fit),
                 hr_mu_upr = exp(fit + 1.96 * se.fit)),
     # predictions for the marginal effect of sigma2
-    predict(m, newdata = ., terms = 's(sigma2)', type = 'link',
-            se.fit = TRUE) %>%
+    predict(m, newdata = ., exclude = 's(mu)',
+            type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
       as.data.frame() %>%
@@
                 hr_sigma2_upr = exp(fit + 1.96 * se.fit)))
+```
+
+```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
+theme_set(theme_bw() + theme(panel.grid = element_blank()))
 
@@
   coord_cartesian(ylim = c(0, 12.5)) +
-  geom_point(aes(mu, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
+  geom_point(aes(mu, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
   geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
@@
   geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
-  xlab(e_r) +
-  scale_y_continuous(hr_lab)
+  labs(x = e_r, y = hr_lab)
 
 p_e <- ggplot() +
-  coord_cartesian(ylim = c(0, 12.5)) +
-  geom_point(aes(sigma2, hr_est_95), tapir, color = pal[3], alpha = 0.5) +
+  geom_point(aes(sigma2, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
   geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
@@
             linewidth = 2) +
-  xlab(v_r) +
-  scale_y_continuous(hr_lab)
+  labs(x = v_r, y = hr_lab)
 
-p_f <- ggplot(tapir) +
-  geom_point(aes(mu, sigma2), alpha = 0.5) +
-  labs(x = e_r, y = v_r)
-
-r_grobs <- map(list(p_d, p_e, p_f), as_grob)
+p_f <-
+  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
+                       to = ceiling(max(tapir$mu) * 100) / 100,
+                       length.out = 250),
+              sigma2 = seq(from = 0, to = 0.0035, length.out = 250)) %>%
+  mutate(hr_full_est = predict(m, newdata = ., type = 'response')) %>%
+  ggplot() +
+  geom_raster(aes(mu, sigma2, fill = hr_full_est)) +
+  geom_contour(aes(mu, sigma2, z = hr_full_est), color = 'black') +
+  geom_point(aes(mu, sigma2), tapir, alpha = 0.3, show.legend = FALSE) +
+  scale_x_continuous(e_r, expand = c(0, 0)) +
+  scale_y_continuous(v_r, expand = c(0, 0)) +
+  scale_fill_gradient(bquote(atop(bold('7-day space-use'),
+                                  paste(bold('requirement (km'), '\U00B2',
+                                        bold(')')))),
+                      low = 'grey90', high = pal[3], limits = c(0, NA)) +
+  theme(legend.position = c(1, 1),
+        legend.justification = c('right', 'top'),
+        legend.box.background = element_rect(),
+        legend.background = element_rect(),
+        legend.key.width = unit(0.35, 'in')) +
+  guides(fill = guide_colorbar(
+    title.position = 'top',
+    theme = theme(legend.title = element_text(hjust = 1)),
+    direction = 'horizontal'))
 
 # align right margins of all plots
+r_grobs <- map(list(p_d, p_e, p_f), as_grob)
 aligned_widths <- align_margin(map(r_grobs, \(x) {x$widths}), 'first')
-
-# Setting the dimensions of plots to the aligned dimensions
 for(i in seq_along(r_grobs)) r_grobs[[i]]$widths <- aligned_widths[[i]]

### 2024-03-28 writing/manuscript.Rmd
updated appendices and methods for empirical model

@@
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
-The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007, examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
+The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
 Panel C shows the changes in the tapir's 7-day home range over time.
@@
 Since $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel F) and (spatio)temporally autocorrelated (panels A and B), the effects of $R$ on $H$ should be modeled carefully.
-To avoid over-fitting the model, we constrained the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using a shape-constrained additive model [@pya_scam_2022] with monotone deceasing and increasing P-splines for $\mu(t)$ and $\sigma^2(t)$, respectively.
+To avoid over-fitting the model, we constrained the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using low number of knots ($\tt{k = 3}$).
 Additional information is provided in appendix C.

### 2024-04-12 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R
removed alpha in panel d

@@ -141,17 +141,15 @@ preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
                         hr_sigma2_upr = exp(fit + 1.96 * se.fit)))
 
 p_d <- ggplot() +
-  coord_cartesian(ylim = c(0, 12.5)) +
+  coord_cartesian(ylim = c(0, 12.5)) + # CIs go higher
   geom_point(aes(mu, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
   geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
               fill = pal[1], alpha = 0.3) +
   geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
-  labs(x = e_r, y = hr_lab) +
-  scale_alpha_continuous(expression(bold(sqrt(Range~crossings))),
-                         range = c(0.3, 1), breaks = c(4, 6, 8))
+  labs(x = e_r, y = hr_lab)
 
 p_e <- ggplot() +
-  coord_cartesian(ylim = c(0, 12.5)) +
+  coord_cartesian(ylim = c(0, 12.5)) + # to have same y breaks as p_d
   geom_point(aes(sigma2, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
   geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr), preds,
               fill = pal[2], alpha = 0.3) +

### 2024-04-12 writing/manuscript.Rmd
addressed mike's comments; see ***changes***

@@
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
+csl: 'the-american-naturalist.csl'
+# csl: 'nature.csl' # to count refs
 fontsize: 12pt
@@
 
-\noindent \textbf{References}: 95 <!-- counted by using nature csl -->
+\noindent \textbf{References}: 102 (updated on 2024-04-12) <!-- counted by using nature csl -->
 
@@
 <!-- start main body on a new page -->
-# To do:
 
-* make sure appendices flow (added stuff from main text)
-* check number of references (updated on 2024-03-18)
+\newpage
 
@@
 Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig.
-2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work.
-Instead, results are typically reported as independent, novel findings.
+2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, ***but, unlike other areas of ecological research, findings from these studies are not structured around a set of theoretical expectations.*** Instead, results are typically reported as independent, novel findings.
 Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements.
@@
 For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity).
-In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976].
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies.
-Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+In the grassland, an animal may require a large but constant home range as it moves between patches in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency ***within*** patches [@charnov_optimal_1976].
+Although the two habitats may have the same long-term average resource abundance, the ephemerality of resource in the second environment would promote a different selection of movement and life history strategies.
+Overall, resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [@chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources.
+Here, we illustrate how organisms' space use ***should be expected to*** depend on both the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
@@
 \noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
-Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time.
-Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
+***Thus, we can treat the amount of resources at a given point in space ($u$) and time ($t$) as a random variable, denoted $R(u, t)$.
+The benefit of treating*** resources as a numerical random variable, ***is that it*** provides us with the capacity to leverage techniques from probability theory and statistics.
 We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.
@@
 All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
-The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., fig.
-2 in the work of @bista_effect_2022].
-Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
+***As the abundance of essential resources fluctuates,*** motile organisms can move to new locations ***or 'patches', to meet their requirements*** [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
 
@@
 Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
+***moved from above*** The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+This is problematic because, ***conceptually,*** the relationship between space-use requirements and $\e(R)$ must be nonlinear, since: (1) there ***is an upper limit*** to how much space an organism is able to explore in its lifetime and ***(2)*** the minimum amount of space it requires to survive ***is necessarily greater than zero*** [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
 
 Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig.
-A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
+A1).
+Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
 However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
@@
 
-\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when data is scarce, but it likely not realistic, since $\var(R)$ often also changes over time.
-Periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$.
-In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$ because $R$ tends to vary across wider ranges (e.g., due to high spatial variability, or high $R$ followed by low $R$ after grazing).
-When both $\e(R)$ and $\var(R)$ change over time (fig.
+\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when ***information on how $\e(R)$ changes is limited,*** but it likely not realistic, since ***the variance in $R$ around $\e(R)$, which we indicate as $\var(R)$, can also differ between locations or time periods***.
+Generally, periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$***, which implies a smaller $\var(R)$***.
+In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$, ***as*** $R$ tends to vary ***more (e.g., due differences in phenology, environmental conditions, and patch depletion by competitors).
+Recognizing differences in $\var(R)$ between locations or periods helps account for the small-scale stochasticity in $R$, after accounting for the large-scale average $R$ [@levin_problem_1992].
+However,*** when both $\e(R)$ and $\var(R)$ change over time (fig.
 A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
@@
 While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still limited.
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is ***unknown***.
 Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.
@@
 
-\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement.
+\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ ***should be expected to*** affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement.
 However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on organism's movement should account for not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
+Consequently, ***an organism’s movement behavior is likely to be a function of*** not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
@@
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
-The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
-However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
+The IOU model's correlated velocity produces realistic tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
 Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.
 
-Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
+Each of the 200 tracks were placed on a raster ***(this was a raster.
+is ``grid" more understandable?)*** with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
 Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution.
@@
 We designed the simulations presented here to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments.
-Consequently, we made the following four assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$).
+Consequently, we made the following ***five*** assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$).
 (2) Organisms move randomly in the environment with no energetic costs.
@@
 (4) Resources do not deplete, and there is no competition or predator avoidance.
-Additional information is provided in Appendix B.
+***(5) Since $\mu(t)$ and $\sigma^2(t)$ are spatially homogeneous for a given $t$, organisms respond to $\e(R)$ and $\var(R)$ by adapting the extent of the spatial needs (i.e., their positional variance), rather than dispersing or abandoning a range (i.e., their mean location).
+Additional information is provided in Appendix B, including the directed acyclical graph [see figure B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct, partial effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with figure C5 and the empirical case study below).***
 
@@
 As $\e(R)$ increases at a constant slope (linear $\mu(t)$) $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism.
-Although less evident, note how the noise in the lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->
+Although less evident, note how the noise in the ***green*** lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->
 
@@
 
+***changed the paragraph below***
+
 Although the temporal trends in fig.
-\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time (panels A and B of fig.
-\@ref(fig:5-5-reg-mean)).
+\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two ***surprisingly (I don't think they're surprising, especially after our rationale)*** simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time (panels A and B of fig.
+\@ref(fig:5-5-reg)A-C).
 The estimated relationships follow the hypotheses we presented in figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation as $\var(R)$ increased.<!--We ensured this was not due to a high minimum or low maximum $\var(R)$ (fig.
-\@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C).
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation as $\var(R)$ increased.
+However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C).
 The increase in $\e(H)$ over $\var(R)$ is strongly nonlinear at low values of $\e(R)$, but it is not affected by $\var(R)$ much when $\e(R)$ is very low (and $\e(H)$ is consequently very high).
+As expected by the changes in the spread of the points in panels A and B of figure \@ref(fig:5-5-reg), the variance in $H$ also depends on $\e(R)$ and $\var(R)$ (figure \@ref(fig:5-5-reg)D-F).
+Since we modeled $H$ using a Gamma family of distributions, we expected $\var(H)$ to increase with $\e(H)$, but the location-scale model removes the assumption of a constant mean-variance relationship.
+Flexibility in the mean-variance relationship to change over $\e(R)$ and $\var(R)$, allowed us to show the strong exponential effect of $\e(R)$ and $\var(R)$ on $\var(H)$.
+Consequences of these effects are explored in the discussion section.
 
@@
 
-```{r 5-5-reg-mean, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% Bayesian credible intervals for the mean.
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% Bayesian credible intervals for the mean.
 While the effect of $\\var(R)$ is sublinear in panel B, the effect of $\\var(R)$ depends strongly on $\\e(R)$ (see panel C).\nCredible intervals were calculated using 10,000 samples from the posterior distribution.
@@
 Details on the model structure are provided in Appendix B.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-mean.png')
-```
-
-As expected by the changes in the spread of the points in panels A and B of figure \@ref(fig:5-5-reg-mean), the variance in $H$ also depends on $\e(R)$ and $\var(R)$ (figure \@ref(fig:5-5-reg-var)A-B).
-Here, the effects of the two variables are more pronounced, even for low values of $\e(R)$ or high values of $\var(R)$.
-Consequences of these effects are explored in the discussion section.
-
-```{r 5-5-reg-var, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on the variance in simulated spatial requirements, with 95\\% Bayesian credible intervals for the estimated variance.
-Credible intervals were calculated using 10,000 samples from the posterior distribution.
-Additional details on the model structure are provided in Appendix B.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-var.png')
+knitr::include_graphics('../figures/simulation-regression-mean-and-var.png')
 ```
@@
 
-\noindent Fig.
+Fig.
 \@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from the individual labelled as "Anna" in the dataset from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
@@
 
-\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on organisms' space-use requirements.
-We support the hypotheses using quantitative simulations and an empirical case study.
-Each of the two methods allows us to highlight the strengths and limitations of simulation-based and data-driven approaches, which we discuss below.
+\noindent This paper presents two hypotheses ***that describe*** the effects of resource abundance and stochasticity on organisms' space-use requirements.
+***We show that, together, these hypotheses provide a framework for understanding how organisms adapt their spatial needs to both average conditions and the variance around the average conditions.
+By moving beyond considering only average conditions, we are able to create a more realistic view of how animals respond to changes in both predictable and unpredictable environmental conditions.
+We support the hypotheses using quantitative simulations and test it using a simple case study with methods that can implemented easily by animal movement ecologists (see Appendix C).
+In this section, we discuss the strengths and limitations of both the simulation-based and empirically driven approaches along with comments on the consequences for free-moving animals in a changing world.***
 
@@
 
-Avoiding confounding variables rather than attempting to account for them (as is required with empirical data) allowed us to develop theory easily while providing a baseline model against which one can contrast empirical examples.
-In figures B6, we present the Directed Acyclical Graph [DAG, see @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct, partial effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with figure C5 and the section below).
+Avoiding confounding variables rather than attempting to account for them (as is ***often*** required with empirical data) allowed us to develop theory easily while providing a baseline model against which one can contrast empirical examples.
 
@@
 However, using proxies of $R$ rather than direct measurements introduces sources of bias.
-While $R$ and NDVI are correlated [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019] and nonlinear [@ref].
-This is because remote-sensed proxies of $R$ and they do not measure resource availability directly, so they introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the DAG in figure C5) and satellite-level confounding variables ($S$ in figure C5).
+While $R$ and NDVI are correlated [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent, and nonlinear [@huang_commentary_2021; @fan_global_2016].
+This is because remote-sensed proxies of $R$ and they do not measure resource availability directly, so they introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in figure C5) and satellite-level confounding variables ($S$ in figure C5).
 
-Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and $NDVI$ [@jetz_scaling_2004; @prox_framework_2020].
-Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999].
+Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020].
+Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition and predation [@charnov_optimal_1976; @brown_ecology_1999].
 While such pressures may increase avoidance behaviors, avoidance may have a cmplex effect on $H$, since it can result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020] or a contaction of an animal's movement, since larger home ranges can be harder to defend and also result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [theoretical framework: @despotic_ref; examples: @rich_anthropogenic_2012; @bengsen_feral_2016; @torrez-herrera_monkeys_2020].
+Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary ***under the Ideal Free Distribution*** [theoretical framework: @fretwell_territorial_1969; examples: @rich_anthropogenic_2012; @bengsen_feral_2016; @torrez-herrera_monkeys_2020].
 
-Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources, competitive pressure, and predator avoidance).
-However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @ref].
+Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@huang_commentary_2021; @fan_global_2016], satellite-level error [@huang_commentary_2021; @tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources, competitive pressure, and predator avoidance).
+However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @huang_commentary_2021].
 In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$ that prevented us from estimating nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$), since estimating the divers of movement and disentangling them from potential confounds in a stochastic environemtn is often complex or impossible [e.g., @wang_stochastic_2019].
 A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this empirical example.
-
-**here:** need a few lines to tie this back to the general
 
@@
 
-**needs a few more lines at the end**
-
 \noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
 We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior.
-We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving-window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.
-The methods presented here can be applied to assess both current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
+We demonstrate that organisms' space-use requirements respond ***(nonlinearly)*** to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.<!-- We also provide a moving-window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.-->
+
+***The hypotheses we presented here provide a framework of how one should expect organisms to adapt their space use in response to changes in $\e(R)$ and $\var(R)$.
+Similarly to the introduction of the Ideal Free Distribution by @fretwell_territorial_1969, the framework we present here allows one to detect substantial deviations from the expected behavior and study the causes of such deviations [e.g., @fretwell_populations_2020].
+Recent advances in computation power have greatly increased analysts' ability to fit computationally demanding models [@nathan_big-data_2022; @wood_generalized_2017-1], which allows (need a different verb) biologists to move beyond only considering changes in mean conitions and instead start thinking more about changes in the stochasticity of environments organisms live in, which is crucial in ecosystems of high and changing stochsticity [@rypkema_modeling_2021; @yao_emergence_2022; @grant_evolution_2017; @berger_climate_2018; @schmidt_interplay_2020; @intergovernmental_panel_on_climate_change_climate_2023].***
 

### 2024-04-12 analysis/simulations/6-modeling-R-and-hr.R

@@ -247,7 +247,7 @@ p_full_var <-
     bquote(paste(bold('Resource stochasticity, Var('),
                  bolditalic('R'), bold(')'))), breaks = NULL,
     expand = c(0, 0)) +
-  scale_fill_gradient(fill_var_lab, low = 'grey90', high = pal[3], 
+  scale_fill_gradient(fill_var_lab, low = 'grey90', high = pal[7], 
                       breaks = range(preds_full$hr_var),
                       labels = c('low', 'high')) +
   theme(legend.position = 'top')
@@ -266,6 +266,14 @@ cowplot::plot_grid(p_mu_var, p_sigma2_var, p_full_var, labels = 'AUTO',
 
 ggsave('figures/simulation-regression-var.png',
        width = 9, height = 3, dpi = 'print', bg = 'white', scale = 1.5)
+
+# mean and variance together
+cowplot::plot_grid(p_mu_mean, p_sigma2_mean, p_full_mean,
+                   p_mu_var, p_sigma2_var, p_full_var,
+                   labels = 'AUTO', nrow = 2)
+
+ggsave('figures/simulation-regression-mean-and-var.png',
+       width = 9, height = 6, dpi = 'print', bg = 'white', scale = 1.5)
 
 # effect of sigma2 for a few values of mu ----
 newd_mu_2 <-

### 2024-05-30 analysis/figures/default-figure-styling.R
made color palettes colorblind-friendly

@@ -1,4 +1,7 @@
 library('ggplot2') # for fancy figures
+library('khroma')  # for color palettes
+
+.check_colors <- FALSE
 
 # set default ggplot theme
 theme_set(theme_bw() +
@@ -6,16 +9,18 @@ theme_set(theme_bw() +
                   text = element_text(size = 15, face = 'bold'),
                   panel.grid = element_blank()))
 
-# custom color-blind palette
-pal <- c('#ff8c00', '#4477AA', '#009900', '#66CCEE',
-         '#CCBB44', '#EE6677', '#AA3377', '#BBBBBB')
+# custom color-blind-friendly palette
+pal <- c('#F09D00', '#3388E1', '#1C7734', '#992266')
+
+if(.check_colors) plot_scheme_colorblind(pal)
 
 # custom NDVI color palette
-create_ndvi_pal <- colorRampPalette(c('darkblue', 'dodgerblue', '#744700', '#d9bb94',
-                                      'darkgreen'))
+create_ndvi_pal <- colorRampPalette(c('darkblue', 'dodgerblue', '#744700',
+                                      '#d9bb94', 'darkgreen'))
 ndvi_pal <- create_ndvi_pal(100)
+if(.check_colors) plot_scheme_colorblind(ndvi_pal)
 
 # mean and variance notations
 e_r <- bquote(paste(bold('E('), bolditalic('R'), bold(')')))
 var_r <- bquote(paste(bold('Var('), bolditalic('R'), bold(')')))
-h <- bquote(paste(bold('Space-use requirements, '), bolditalic('H')))
+h <- bquote(paste(bold('Home-range size, '), bolditalic('H')))

### 2024-05-30 analysis/figures/habitat-examples.R
removed unnecessary package

@@ -6,7 +6,6 @@ library('purrr')     # for functional programming
 library('ctmm')      # for movement modeling
 library('lubridate') # for working with data more smoothly
 library('ggplot2')   # for fancy plots
-library('latex2exp') # for LaTeX in plots
 library('cowplot')   # for fancy plots in grids
 library('ragg')      # needed for custom alpha with coord_cartesian
 # see https://github.com/tidyverse/ggplot2/issues/4029

### 2024-05-30 analysis/figures/home-range-hypotheses.R
changed ylab to "(home) range size"

@@ -1,6 +1,4 @@
-library('ctmm')    # for movement modeling
 library('dplyr')   # for data wrangling
-library('tidyr')   # for data wrangling
 library('ggplot2') # for fancy plots
 library('ggbrace') # for curly braces on fancy plots
 library('cowplot') # for fancy multi-panel plots
@@ -9,7 +7,7 @@ source('analysis/figures/default-figure-styling.R') # theme & color palette
 theme_set(theme_get() + theme(legend.position = 'none',
                               text = element_text(size = 22)))
 
-hr_lab <- 'Space-use requirements' # label for y axis
+hr_lab <- 'Range size' # label for y axis
 
 # hypothesis for change in H over E(R) ----
 d <-

### 2024-05-30 analysis/figures/hr-mean-variance-simulations.R

@@ -12,7 +12,7 @@ e_r <- bquote(paste(bold('Resource abundance, E('), bolditalic('R'),
 v_r <- bquote(paste(bold('Resource stochasticity, Var('), bolditalic('R'),
                     bold(') = '), '\U1D6D4\U00B2', bold('('),
                     bolditalic('t'), bold(')')))
-hr_lab <- bquote(paste(bold('Space-use requirements, '), bolditalic('H')))
+hr_lab <- bquote(paste(bold('Home-range size, '), bolditalic('H')))
 
 p_sim <-
   ggplot(days) +
@@ -45,8 +45,8 @@ p_variance <-
         axis.title.x = element_text(color = 'transparent'))
 
 # create a single plot
-p_full <- plot_grid(NULL, p_mean, p_variance, p_sim, rel_widths = c(1, 4.5),
-                    rel_heights = c(1, 3.5), nrow = 2)
+p_full <- plot_grid(NULL, p_mean, p_variance, p_sim, rel_widths = c(1, 4),
+                    rel_heights = c(1, 3.1), nrow = 2)
 p_full
 
 # save the plot as a png

### 2024-05-30 analysis/simulations/6-modeling-R-and-hr.R
cleaned up scripts, improved models

@@ -24,56 +24,14 @@ sims <- readRDS('simulations/days-hrs.rds') %>%
 ggplot(sims, aes(mu, sigma2)) +
   geom_point(alpha = 0.3)
 
-m_1 <- gam(
-  list(
-    # predictor for the mean
-    hr ~ quantile + mu + sigma2,
-    # predictor for the scale (variance = scale * mean)
-    ~ quantile + mu + sigma2),
-  family = gammals(),
-  data = sims,
-  method = 'REML')
-plot(m_1, pages = 1, all.terms = TRUE) # acceptable but likely too rigid
-
-m_2 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(sigma2),
-    ~ quantile + s(mu) + s(sigma2)),
-  family = gammals(),
-  data = sims,
-  method = 'REML')
-plot(m_2, pages = 1, all.terms = TRUE) # scale terms are inappropriate
-
-m_3 <- gam(
+m <- gam(
   list(
     hr ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5),
     ~ quantile + s(mu, k = 5) + s(sigma2, k = 5) + ti(mu, sigma2, k = 5)),
   family = gammals(),
   data = sims,
   method = 'REML')
-plot(m_3, pages = 1, all.terms = TRUE, scheme = 3) # much better
-
-# check if CV = sigma/mu is better than sigma
-m_4 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
-    ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
-  family = gammals(),
-  data = mutate(sims, cv = sqrt(sigma2) / mu),
-  method = 'REML')
-# ti term of the scale parameter looks over-fit
-plot(m_4, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
-
-AIC(m_1, m_2, m_3, m_4) # smooth model with interactions is best
-BIC(m_1, m_2, m_3, m_4) # smooth model with interactions is best
-
-summary(m_3)
-summary(m_4)
-
-# choose the best model
-m <- m_3
-
-plot(m, all.terms = TRUE, pages = 1, scheme = 3, scale = 0)
+draw(m, rug = FALSE)
 
 # filter to 95% HR for simplicity
[…]

### 2024-05-30 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R

@@ -16,7 +16,7 @@ e_r <- bquote(paste(bold('Resource abundance, '), '\U1D6CD', bold('('),
                     bolditalic('t'), bold(')')))
 v_r <- bquote(paste(bold('Resource stochasticity, '), '\U1D6D4\U00B2',
                     bold('('), bolditalic('t'), bold(')')))
-hr_lab <- bquote(paste(bold('7-day home range size (km'), '\U00B2',
+hr_lab <- bquote(paste(bold('7-day home-range size (km'), '\U00B2',
                        bold(')')))
 
 # import location-scale model for predictions of mean and variance in NDVI
@@ -101,12 +101,14 @@ p_left <- plot_grid(plotlist = l_grobs, ncol = 1, labels = 'AUTO',
 # adding weights biases towards smaller HRs and gives bad q-q plots
 m <- gam(hr_est_95 ~
            s(mu, k = 4) +
-           s(sigma2, k = 4),
+           s(sigma2, k = 4) +
+           ti(mu, sigma2, k = 3),
          family = Gamma('log'),
          data = tapir,
          method = 'REML')
+draw(m, rug = FALSE, scales = 'fixed', dist = 1)
 appraise(m, method = 'simulate', n_simulate = 1e3)
-draw(m, rug = FALSE, scales = 'fixed')
+summary(m)
 
 # diagnostics aren't bad given the autocorrelation
 ggplot(mapping = aes(tapir$mu, resid(m))) +
@@ -125,14 +127,14 @@ ggplot(mapping = aes(tapir$mu, tapir$sigma2)) +
 
 # marginals of mu and sigma2, and a plot to show the lack of independence
 preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
-                sigma2 = seq(5e-4, 25e-4, length.out = 250)) %>%
-  bind_cols(predict(m, newdata = ., exclude = 's(sigma2)',
+                sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
+  bind_cols(predict(m, newdata = ., terms = c('(Intercept)', 's(mu)'),
                     type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
               as.data.frame() %>%
               transmute(hr_mu_est = exp(fit),
                         hr_mu_lwr = exp(fit - 1.96 * se.fit),
                         hr_mu_upr = exp(fit + 1.96 * se.fit)),
-            predict(m, newdata = ., exclude = 's(mu)',
[…]

### 2024-05-30 functions/rbeta2.R
wrote rbeta() function for mean and variance

@@ -1,0 +1,22 @@
+rbeta2 <- function(mu, sigma2, N = 1e4) {
+  
+  phi <- sigma2 / (mu * (1 - mu)) # scale parameter
+  a <- mu*(1 / phi - 1) # shape_1
+  b <- (1 - mu) * (1 / phi - 1) # shape 2
+  
+  cat(paste0('a = ', a, ', b = ', b, ', phi = ', phi, '\n'))
+  rbeta(n = N, shape1 = a, shape2 = b)
+}
+
+
+if(FALSE) {
+  eps <- .Machine$double.eps
+  layout(matrix(1:4, ncol = 2, byrow = TRUE))
+  for(.mu in c(0.5, 0.9)) {
+    for(.sigma2 in c(1e-4, 0.05)) {
+      hist(rbeta2(mu = .mu, sigma2 = .sigma2 - eps), xlab = 'R',
+           main = paste0('\U03BC = ', .mu, ', \U03C3\U00B2 = ', .sigma2))
+    }
+  }
+  layout(1)
+}

### 2024-05-30 writing/appendix-A-concepts-and-definitions.Rmd
updated writing

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' space-use requirements
+title: \large How resource abundance and stochasticity affect organisms' range sizes
 subtitle: "Appendix A: Concepts and definitions"
@@
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl'
+# csl: 'the-american-naturalist.csl'
 fontsize: 12pt
@@
 
-```{r ex, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Gamma distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
+```{r ex-gamma, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Gamma distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
 library('dplyr') # for data wrangling
@@
   ggplot(aes(x, dens)) +
-  coord_cartesian(xlim = c(-0.5, 10.5), ylim = c(0, 0.8)) +
+  coord_cartesian(xlim = c(-0, 10), ylim = c(0, 0.8)) +
   # using '\U03BC' because mu results in \propto, for some reason
@@
   geom_area(alpha = 0.3, na.rm = TRUE) +
+  scale_x_continuous(expand = c(0, 0), breaks = c(0, 2.5, 5, 7.5)) +
   scale_y_continuous(expand = c(0, 0)) +
-  scale_x_continuous(expand = c(0, 0)) +
-  labs(x = 'r', y = 'Density') +
-  theme_bw()
+  labs(x = 'R', y = 'Density') +
+  theme_bw() +
+  theme(panel.grid = element_blank())
+```
+
+```{r ex-beta, eval=TRUE, echo=FALSE, fig.height=5, fig.cap="Examples of Beta distributions with different means ($\\mu$) and variances ($\\sigma^2$).", fig.pos='!h', message=FALSE, warning=FALSE, out.width = '75%'}
+# to help visualize the consequences of different E(R) and V(R) if R ~ Beta
+dbeta2 <- function(x, mu, sigma2, lower.tail = FALSE,
+                   eps = .Machine$double.eps) {
+  phi <- sigma2 / (mu * (1 - mu)) # scale parameter
+
+  dbeta(x = x,
+        shape1 = mu * (1 / phi - 1),
+        shape2 = (1 - mu) * (1 / phi - 1))
+}
+
+expand_grid(x = tibble(x = seq(0, 1, length.out = 1000)),
+            mean = c(0.5, 0.75),
+            var = c(0.002, 0.05)) %>%
+  unnest(x) %>%
+  mutate(dens = dbeta2(x = x, mu = mean, sigma2 = var)) %>%
+  ggplot(aes(x, dens)) +
+  coord_cartesian(ylim = c(0, NA)) +
+  # using '\U03BC' because mu results in \propto, for some reason
+  facet_wrap(. ~ mean + var,
+             labeller = label_bquote('\U03BC' == .(mean)~','~sigma^2 == .(var)))+
+  geom_line(na.rm = TRUE) +
+  geom_area(alpha = 0.3, na.rm = TRUE) +
+  scale_x_continuous(expand = c(0, 0), breaks = c(0, 0.25, 0.5, 0.75, 1),
+                     labels = c('0', '0.25', '0.5', '0.75', '1')) +
+  scale_y_continuous(expand = c(0, 0)) +
+  labs(x = 'R', y = 'Density') +
+  theme_bw() +
+  theme(panel.grid = element_blank())
 ```

### 2024-05-30 writing/appendix-B-simulations.Rmd
updated writing

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' space-use requirements
+title: \large How resource abundance and stochasticity affect organisms' range sizes
 subtitle: "Appendix B: Simulations"
@@
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl'
+# csl: 'the-american-naturalist.csl'
 fontsize: 12pt
 header-includes:
-    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Sx.
+    - \renewcommand{\figurename}{Figure B\!\!} # for "Figure Bx."
     - \usepackage{setspace}\doublespacing # for double-spaced text
@@
 
-```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' space-use requirements were simulated.
+```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' home-range sizes were simulated.
 (a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
@@
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)a.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
+**Exploratory plot B\@ref(fig:dt-sensitivity)A.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
 Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)b.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$
+**Exploratory plot B\@ref(fig:dt-sensitivity)B.** Halving the sampling interval had little to no effect on the total number of encounters for $\Delta t \lessapprox 60 \,s.$
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)c.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
+**Exploratory plot B\@ref(fig:dt-sensitivity)C.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)d.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
+**Exploratory plot B\@ref(fig:dt-sensitivity)D.** The three tracks used for these exploratory plots are sufficiently different that we considered them to be a representative sample of movement tracks simulated by the OUF model.
 All three tracks have a reasonable amount of both tortuous and directed movement.
@@
 ## Checking how many tracks were necessary
+
+As the organism searched for resources, it encountered $R$ resources sampled from a family of Gamma distributions.
+The figure below shows the range in Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\,500, 126\,250, 250\,000$).
+
+\renewcommand{\figurename}{Figure B\!\!\!\!}
+
+```{r, eval=TRUE, echo=FALSE, fig.cap="Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\\,500, 126\\,250, 250\\,000$) with the corresponding coefficients of variation.
+The median and mode of the distributions increase with the mean but decrease with the variance.
+Greater coefficients of variation result in wider distributions with greater density near zero."}
+setwd('..')
+source('analysis/simulations/mean-variance-trends-panel-data.R')
+source('functions/rgamma2.R')
+setwd('writing')
+
+expand_grid(mu = quantile(d55$mu, c(0, 0.5, 1)),
+            sigma2 = quantile(d55$sigma2, c(0, 0.5, 1))) %>%
+  mutate(sample = map2(mu, sigma2,
+                       \(.m, .s) tibble(R = rgamma2(.m, .s, 1e5))),
+         parameters = paste('list(', '\U03BC~"="~', mu, ', ',
+                            'sigma^2~"="~', round(sigma2), ', ',
+                            'sigma/\U03BC~"="~', round(sqrt(sigma2) / mu, 2),
+                            ')') %>%
+           factor(., levels = unique(.))) %>%
+  unnest(sample) %>%
+  ggplot(aes(R)) +
+  facet_wrap(~ parameters, labeller = label_parsed, scales = 'fixed') +
+  geom_density(fill = 'grey', bw = 5, na.rm = TRUE) +
+  scale_x_continuous(expand = c(0, 0), limits = c(0, 399)) +
+  scale_y_continuous('Density', expand = c(0, 0)) +
+  theme_bw() +
+  theme(panel.grid = element_blank())
+```
+
+\renewcommand{\figurename}{Figure B\!\!}
+
+\clearpage
 
@@
 
-In this script, we check how many tracks are necessary to produce stable and accurate estimates of the space-use requirements.
+In this script, we check how many tracks are necessary to produce stable and accurate home-range size estimates.
 We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
@@
                      limits = c(2, 1100)) +
-  scale_y_continuous(expression(atop(Estimated~'space-use',
-                                     requirements~(log[2]~scale))),
+  scale_y_continuous(expression(atop(Estimated~'home-range',
+                                     size~(log[2]~scale))),
                      trans = 'log2')
@@
 
-```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated space-use requirements as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
-In both cases, 100 days are sufficient to produce stable estimates of space-use requirements.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
+```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated home-range sizes as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
+In both cases, 100 days are sufficient to produce stable estimates of home-range size.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
 knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
@@
 
-This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated space-use requirements.
+This final section illustrates the location-scale GAM used to estimate the effects of resource abundance and stochasticity on the simulated home-range sizes.
 Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity.
@@
 
-$$\widehat{h_q} = -2 \log(1 - q) \pi \sigma^2_{pos},$$
+$$\widehat{h_q} = -2 \log(1 - q) \pi \varsigma^2,$$
 
-where $widehat{h_q}$ is the estimated home range for the $q^{\text{th}}$ quantile and $\sigma^2_{pos}$ is the positional variance.
+where $\widehat{h_q}$ is the estimated home range for the $q^{\text{th}}$ quantile and $\varsigma^2$ is the positional variance.
 Consequently, the estimated core home range estimate is simply a multiple of the 95% quantile:
@@
 Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
-When estimating space-use requirements based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+When estimating home-range sizes based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
@@
 
-The sampling of the data is not uniform:
+\clearpage
 
-```{r, eval=TRUE, fig.height=4}
+The sampling of the data is not uniform, but it is not particularly problematic.
+
+```{r, eval=TRUE, fig.height=4, echo=FALSE}
 ggplot(sims, aes(mu, sigma2)) +
@@
   theme_bw() # Rmd only keeps theme_bw as default for a single plot
-```
 
-But it is not particularly problematic:
-
-```{r, eval=TRUE}
 ggplot(sims, aes(mu, sigma2)) +
-  geom_hex(bins = 20) +
+  geom_hex(bins = 10) +
   scale_fill_viridis_c('Count', limits = c(0, NA), option = 'A',
@@
 
+\clearpage
+
 The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown).
-To estimate the effects of resource abundance and stochasticity on the simulated space-use requirements, we fit 3 location-scale GAMs (GAMLSs) with a location-scale gamma distribution (since home ranges are strictly positive) of increasing complexity.
-We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones.
-The three models are detailed below and compared using the Akaike Information Criterion and the Bayesian Information Criterion.
-Both information criteria demonstrate that the third model, which accounts for the interaction effects of $\e(R)$ and $\var(R)$, is the best model.
-Each of the three models are based on the directed acyclical graph (DAG) shown in figure B\@ref(fig:dag).
+To estimate the effects of resource abundance and stochasticity on the simulated home-range sizes, we fit a location-scale GAMs (GAMLS) with a location-scale gamma distribution (since home ranges are strictly positive).
+We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones (fig.
+B\@ref(fig:dag)).
 Within a Bayesian framework, DAGs can be used to infer causality within statistical models [as the latter alone does not allow one to infer causality, see @mcelreath_statistical_2016].
-The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's space-use requirements, $H$.
+The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's home-range sizes, $H$.
 Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).
@@
 ```{r, eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth'}
-# marginal effects of mu and sigma only (linear on the log link scale)
-m_1 <- gam(
-  list(
-    # predictor for the mean
-    hr ~ quantile + mu + sigma2,
-    # predictor for the scale (variance = scale * mean)
-    ~ quantile + mu + sigma2),
-  family = gammals(),
-  data = sims,
-  method = 'REML')
-
-# marginal effects of mu and sigma only (nonlinear on the log link scale)
-m_2 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(sigma2),
-    ~ quantile + s(mu) + s(sigma2)),
-  family = gammals(),
-  data = sims,
-  method = 'REML')
-
-# marginal and interaction effects (nonlinear on the log link scale)
-m_3 <- gam(
+m <- gam(
   list(
@@
   method = 'REML')
+```
 
-# check if CV = sigma/mu is better than sigma
-m_4 <- gam(
-  list(
-    hr ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5),
-    ~ quantile + s(mu) + s(cv) + ti(mu, cv, k = 5)),
-  family = gammals(),
-  data = mutate(sims, cv = sqrt(sigma2) / mu),
-  method = 'REML')
-
-# ti term of the scale parameter looks over-fit and too uncertain
-draw(m_4) & theme_bw() + theme(panel.grid = element_blank())
+```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales).
+The dashed lines indicate the 95\\% Bayesian credible intervals.
+The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
+draw(m) & theme_bw() + theme(panel.grid = element_blank())
 ```
@@
 ```{r, eval=TRUE}
-AIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
-BIC(m_1, m_2, m_3, m_4) # not much of a difference between m_3 and m_4
-```
-
-\clearpage
-
-```{r, eval=TRUE}
-# the difference in df (and significance) is in the ti.1() term
-summary(m_3)
-```
-
-\newpage
-
-```{r, eval=TRUE}
-summary(m_4)
-```
-
-\newpage
-
-```{r, eval=TRUE, fig.height=8, fig.width=8, out.width='65%'}
-m <- m_3 # select the best model
+summary(m)
 ```
@@
 Similar but opposite conclusions can be drawn from the interaction term for the scale parameter, since the effect of $\sigma^2(t)$ on the scale parameter is weaker when $\mu(t)$ is low and stronger when $\mu(t)$ is high.
-
-```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales).
-The dashed lines indicate the 95\\% Bayesian credible intervals.
-The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
-# using custom layout to make figure easier to read
-layout(matrix(c(7, 1, 2, 3, 8, 4, 5, 6), nrow = 4, byrow = TRUE))
-draw(m) & theme_bw() + theme(panel.grid = element_blank())
-```
 
@@
   scale_x_continuous(NULL, breaks = NULL) + 
-  scale_y_continuous(bquote(paste(bold('Space-use requirements, '),
+  scale_y_continuous(bquote(paste(bold('Home-range size, '),
                                   bolditalic('H'))), breaks = NULL) +

### 2024-05-30 writing/appendix-C-empirical-modeling.Rmd
updated writing

@@
 ---
-title: \large How resource abundance and stochasticity affect animals' space-use requirements
+title: \large How resource abundance and stochasticity affect organisms' range sizes
 subtitle: "Appendix C: Empirical modeling"
@@
 
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements.
+We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' home-range sizes.
 Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' space-use requirements under different scenarios.
+Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' range sizes under different scenarios.
 
@@
 The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on organisms' movement and space-use requirements as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
+Thus, we can instead focus on organisms' movement and range sizes as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
 
@@
 However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's space-use requirements and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
+Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's home-range sizes and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
 Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
@@
 
-### Quantifying space-use requirements
+### Quantifying home-range sizes
 
 \noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify organism's space-use requirements using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
-In the case of organisms that are range resident for short periods of time, we can model their movement and short-term space-use requirements within distinct and discrete periods or using a sliding window approach (see the discussion section).
+Although here we quantify organism's home-range sizes using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
+In the case of organisms that are range resident for short periods of time, we can model their movement and short-term home-range sizes within distinct and discrete periods or using a sliding window approach (see the discussion section).
 Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
@@
 
-<!-- ### Modeling space-use requirements -->
+<!-- ### Modeling home-range sizes -->
 
@@
 The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-Estimating the effects of resource abundance and unpredictability on the tapir's space-use requirements requires us to first estimate the changes in the tapir's space-use requirements (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).
+Estimating the effects of resource abundance and unpredictability on the tapir's home-range size requires us to first estimate the changes in the tapir's home-range sizes (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).
 
@@
 
-```{r, eval=TRUE, warning=FALSE}
+```{r, eval=TRUE, warning=FALSE, echo=FALSE}
 # attach all necessary packages
@@
 v_r <- bquote(paste(bold('Resource stochasticity')))
-hr_lab <- bquote(bold(paste('7-day space-use requirement (', km^2, ')')))
+hr_lab <- bquote(bold(paste('7-day home-range size (', km^2, ')')))
 
@@
   scale_y_continuous(v_r, expand = c(0, 0)) +
-  scale_fill_gradient(bquote(atop(bold('7-day space-use'),
-                                  paste(bold('requirement (km'), '\U00B2',
+  scale_fill_gradient(bquote(atop(bold('7-day home-range'),
+                                  paste(bold('size (km'), '\U00B2',
                                         bold(')')))),

### 2024-05-30 writing/manuscript.Rmd
updated writing

--- 2024-03-28 writing/manuscript.Rmd
@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' space-use requirements
+title: \large How resource abundance and stochasticity affect organisms' range sizes
 # author:
@@
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl' # use 'nature.csl' to count refs
+# csl: 'the-american-naturalist-2024-03.csl'
+# csl: 'nature.csl' # to count refs
 fontsize: 12pt
@@
 This is not a trivial oversight.
-Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of organisms' space-use requirements, challenging ecological theory and applied conservation alike.
-In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on organisms' space use.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in space use.
+Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of an organism's range size, which challenges ecological theory and applied conservation alike.
+In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on range sizes.
+Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in range size.
 Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
@@
 
-\noindent \textbf{Figures}: 6
+\noindent \textbf{Figures}: 5
 
@@
 
-\noindent \textbf{References}: 95 <!-- counted by using nature csl -->
+\noindent \textbf{References}: 109 (updated on 2024-05-29) <!-- counted by using nature csl -->
 
@@
 
-\noindent \textbf{Key words:} energetics, energetic landscape, environmental stochasticity, space use, spatial needs, `ctmm`
+\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations
 
@@
 knitr::opts_chunk$set(echo = FALSE, out.width = '75%', fig.align = 'center')
-```
-
-```{r wrap-figure-function, include = FALSE}
-# this chunk is needed to wrap text around or beside figures
-# code modified from https://stackoverflow.com/questions/54100806/r-markdown-how-do-i-make-text-float-around-figures
-
-defOut <- knitr::knit_hooks$get('plot')  # save the default plot hook
-
-knitr::knit_hooks$set(plot = function(x, options) { # set the new plot hook...
-
-  x <- defOut(x, options)  # first apply the default hook
-
-  # if wrapfigure options are given, create the new opening string for the
-  # wrapfigure environment using custom figure alignment and text width
-  if(!is.null(options$wrapfigure)) {
-
-    # options must be numeric, they cannot be characters
-    wf <- sprintf('\\begin{wrapfigure}{%s}{%g\\textwidth}',
-                  options$wrapfigure[[1]], # figure alignment
-                  options$wrapfigure[[2]]) # caption width relative to text width
-    x  <- gsub(pattern = '\\begin{figure}', # and replace the default one with it
-               replacement = wf,
-               x = x,
-               fixed = TRUE) # pattern is a string, so evaluate as is
-    x  <- gsub(pattern = '{figure}', # also replace the environment ending
-               replacement = '{wrapfigure}',
-               x = x,
-               fixed = TRUE)
-  }
-  return(x)
-}
-)
-
-# chuck options used to wrap figures:
-## fig.align = "right":          align figures right
-## out.width = "0.5\\textwidth": output width = 0.5 * the width of the text
-## wrapfigure = list("R", 0.5)   wrap caption to right, caption width = 0.5 text width
-
-#' place figures before the paragraph they are referred to so `wrapfigure`
-#' (generally) places them beside the paragraph
 ```
@@
 \noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction.
-Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018], along with changes to community structure [@haney_tipping_2018; @burson_competition_2018; @riotte-lambert_environmental_2020; @ghislandi_resource_2018].
-Thus, many organisms adapt their behaviors in response to changes in local resource abundance to ensure their needs are met.<!-- Some species may respond to fluctuations in resource abundance by, if possible, switching to other food sources [@steinmetz_effects_2021; @le_bot_fishery_2019], reducing energetic costs by lowering body heat [@schmidt_interplay_2020], or entering hibernation or torpor [@fjelldal_nightly_2021; @mohr_cellular_2020; @boyles_optimal_2020].
-However, --> Movement represents one of the most readily available behaviors that species can adjust to optimize their resource use [<!--@krivan_ideal_2008; @brown_ecology_1999; -->@kacelnik_ideal_1992; @charnov_optimal_1976].<!-- whether this be by modifying their home range [@hirt_environmental_2021; @bista_effect_2022; @bradsworth_using_2022; @lucherini_habitat_1996; @relyea_home_2000; @yu_artificial_2022; @arechavala-lopez_common_2019], migrating [@middleton_green-wave_2018; @geremia_migrating_2019], moving nomadically [@teitelbaum_beyond_2019; @nandintsetseg_variability_2019], or dispersing [framework: @southwood_habitat_1977; amphibians: @cayuela_determinants_2020; birds: @pretorius_movement_2020; @wheat_migrate_2017; mammals: @singh_migration_2012].-->
+Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018].
+Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met.
 
-The relationship between organisms' movement and resource abundance has been of interest to biologists for nearly a century.
-@burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
-Three decades after, @southwood_habitat_1977 suggested that change in resource abundance is a strong determinant of how organisms decide where to live and when to reproduce.
+While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976].
+The relationship between organisms' movement and resource abundance has long been of interest to biologists.
+In his seminal paper, @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
+Three decades after, @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce.
 Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is
@@
 
+```{r, echo=FALSE, eval=FALSE}
+# visualizing the relationship
+r <- seq(0.5, 5, by = 0.01)
+her <- function(R, C = 1) return(C / R)
+plot(r, her(R = r, C = 5), col = 'white', xlab = 'E(R)', ylab = 'E(H)',
+     ylim = c(0, max(her(R = r, C = 4))))
+for(i in 1:4) {
+  lines(r, her(R = r, C = c(1, 2, 3, 4)[i]), col = i, lwd = 2)
+}
+```
+
 \noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
-Harestad and Bunnel's model is simple to conceptualize and allows for testable predictions, but it fails to account for the nonlinear relationship required to avoid negative home range estimates [e.g., fig.
-2 in the work of @bista_effect_2022].<!-- to an organism's values of $C$ and $R$ likely depend on numerous factors, such as competition, metabolic rate, diet, and body weight [@harvey_primate_1981; @lindstedt_home_1986; @reiss_scaling_1988; @harvey_primate_1981; @gittleman_carnivore_1982; @jetz_scaling_2004; @boratynski_energetic_2020; @noonan_effects_2020].--> Many researchers have since demonstrated that organisms adapt their home ranges in response to resources abundance, but studies often fail to build upon previous work.
-Instead, results are typically reported as independent, novel findings.
-Thus, there remains a need for a clear hypothesis for the effects of resource abundance on organisms' space-use requirements.
-Additionally, while much work has been done on estimating organisms' responses to average resource abundance, there is little information on how they respond to unpredictable changes in resources [but see: @stephens_optimal_1982; @rizzuto_forage_2021].
+Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis.
+Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings.
+Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021].
+Thus, there remains a need for a clear hypothesis for the effects of both resource abundance and stochasticity on organisms’ range sizes.
 
 Here, we refer to a location's average amount of resources as "resource abundance", while we use the phrase "resource stochasticity" to indicate the variability in resources after accounting for changes in the mean.
-We argue that, on its own, a habitat's long-term resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
-For instance, herbivores in a grassland with relatively low but constant forage availability (low mean and variance) will adopt different behaviors and adaptations from herbivores in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (low mean and high stochasticity).
-In the grassland, an animal may require a large but constant home range as it moves between locations in search of food [@teitelbaum_beyond_2019], while an animal in the desert may switch between dispersal in search for high-resource patches and short-term range residency whinith patches [@charnov_optimal_1976].
-Although it may be possible for both habitats to have the same long-term average resource abundance, the differences in resource unpredictability result in different selection of movement and life history strategies.
-Although resource unpredictability (and seasonality) may ecrease organisms' fitness and a landscape's energetic balance [@chevin_adaptation_2010], there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
+To see this, consider,, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance).
+This individual will adopt a different behavior and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity).
+Although these two habitats may have the same long-term average resource abundance, the ephemerality of resources in the stochastic desert environment would promote a different selection of movement and life history strategies.
+In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
+Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
 
-Here, we illustrate how organisms' space use depends on both the abundance and unpredictability of resources.
+In this paper, we illustrate how an organism's range size can be expected to depend on both the abundance and unpredictability of resources.
 First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation.
-Next, we provide a review of the effects of resource abundance on organisms' space use while suggesting a unifying hypothesis.
-Afterwards, we provide a review of the effects of resource stochasticity on organisms' space use while suggesting a second unifying hypothesis.
-Subsequently, we demonstrate the power of these two hypotheses using quantitative, simulated responses to changes in resource abundance and unpredictability.
+Next, we provide a review of the effects of resource abundance on range sizes while suggesting a simple and unifying hypothesis.
+Afterwards, we provide a review of the effects of resource stochasticity on organisms' range sizes while suggesting a second simple and unifying hypothesis.
+Subsequently, we support the two hypotheses using quantitative, simulated responses in range size to changes in resource abundance and stochasticity.
 Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
@@
 
-\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
-Thus, let the random variable $R$ indicate the amount of resources at a given point in space and time.
-Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
-We assume readers are familiar with fundamental statistical concepts (e.g., random variables, probability distributions, expected value and variance of a random variable), but we define and explain the necessary concepts in Appendix A.
-
-<!--
-\noindent In statistics, random variables indicate random (i.e., unknown) quantities and are indicated with capital letters (e.g., $R$).
-Known values, such as realizations of random variables (i.e., known observations or instances), are indicated with lower-case letters (e.g., $r$).
-Using this notation, we can write the statement "the probability of random variable $R$ taking the value $r$" as $P(R = r)$.
-Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at the individual [@schmidt_interplay_2020] and species level [@jetz_scaling_2004].
-Thus, we can let the random variable $R$ indicate the amount of resources at a given point in space and time.
-Quantifying resources as a numerical random variable, as opposed to using *ad hoc* qualitative descriptions, provides us with the capacity to leverage techniques from probability theory and statistics.
-
-In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
-Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
-However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa).
-As the mean approaches zero (from the positive side, which we write as $\e(R) \rightarrow0^+$) the variance also does: $\e(R) \rightarrow 0^+ \iff k \theta \rightarrow 0^+ \iff k \theta^2 = 0^+ \iff \var(R) = 0^+$.
-This assumption also holds biologically, since resources tend to be less variable when they are less abundant.
-We assume readers are familiar with the concepts of: random variables, probability distributions, expected value of a random variable, and variance of a random variable, but we define and explain each of the four concepts in Appendix A.
--->
+\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004].
+Thus, we can treat the amount of resources $R$ at a given point in space ($u$) and time ($t$) as a random variable, denoted as $R(u, t)$.
+Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean.
+We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, u)$ and $\sigma^2(t, u)$ to indicate them as functions of time ($t$) and space ($u$).
+Additional definitions and explanations are available in Appendix A.
 
@@
 
-\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for most species for survival and reproduction.
-All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients to survive, grow, and reproduce [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-Failing to acquire sufficient resources can result in lower fitness, behavioral changes, physiological changes or damage, inability to reproduce, and death.
-Motile organisms can move to new locations in search of resources [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for a higher metabolism and movement costs [@taylor_energetics_1982].
-The hypothesis that space-use requirements decrease with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @harestad_home_1979; @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because a linear relationship can lead to questionable estimates and problematic biases, such as negative or excessively small home ranges [e.g., fig.
-2 in the work of @bista_effect_2022].
-Conceptually, the relationship between space-use requirements and $\e(R)$ must be nonlinear, since there are limits to how much space an organism is able to explore in its lifetime and a minimum amount of space it requires to survive [which is necessarily greater than zero -- see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
+\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for the growth, survival, and reproduction of most organisms.
+All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
+As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for costs of movement [@taylor_energetics_1982].
 
-```{r mu-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource abundance ($\\e(R)$).
-We expect low values of $\\e(R)$ to result in high space-use requirements as organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident or nomadic.
-As $\\e(R)$ increases, space-use requirements should decrease nonlinearly until they reach the minimum amount of space required by the organism to survive.
-Note that the relationship between $\\e(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
+Fig.
+\@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979.
+When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+Like @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
+These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
+Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
+
+```{r mu-hyp, fig.cap="Hypothesized range size of an organism as a function of resource abundance, $\\e(R)$.
+We expect low values of $\\e(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory.
+As $\\e(R)$ increases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.
+Note that the relationship between $\\e(R)$ and range size cannot be linear because it would require range size to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
 knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
@@
 
-Fig.
-\@ref(fig:mu-hyp) illustrates our first of two hypotheses: When $\e(R)$ is high, we expect organisms' spatial requirements to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-At lower values of $\e(R)$, we expect organisms' spatial requirements to increase nonlinearly, since low values of $\e(R)$ force organisms to expand their home ranges [@lucherini_habitat_1996; @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005; @bista_effect_2022], migrate to better locations [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], or move nomadically [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the connection between these types of movement is important for quantifying the effect of resource abundance on organisms' space-use requirements and when an organism may choose to migrate or disperse rather than remaining range-resident [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
-
-Consider the scenario where $\e(R) = \mu(t)$ changes over time but $\var(R) = \sigma^2 > 0$ is constant over time and space (fig.
-A1).<!-- We can use the notation $R \sim \Gamma\big(\mu(t), \sigma^2\big)$ to indicate that $R$ follows a Gamma distribution with a mean $\mu(t)$ that changes over time (i.e., it is a function of time) while the variance is constant.--> Since $R$ is random, an organism that moves in the landscape will not find the same $R$ at different time points or throughout the habitat.
-However, if $\mu(t)$ changes repetitively and regularly over time (e.g.
-peaks each spring and is lowest in winter), an organism may learn to predict times of high or low $\e(R)$ [e.g., @falcon-cortes_hierarchical_2021; @geremia_migrating_2019; @abrahms_memory_2019; @samarra_movements_2017] but will still not be able to predict $R$ exactly.
-
-<!-- This model is somewhat simplistic, but its simplicity makes it easy to fit and conceptualize.
-Such a model may be appropriate in regions where variability does not change noticeably spatiotemporally, or when data availability is too low to produce appreciable measures of changes in variance.
-Examples of temporally homogeneous habitats include regions where productivity remains fairly predictable throughout the year (e.g., equatorial rain forests or highly homogeneous deserts).
-When productivity is approximately stable over long periods of time, it may be possible to further simplify the model by assuming a constant mean, but this is likely rarely the case.
-Most often, $\e(R)$ will vary over time, and organisms' behaviors will change in response.
--->
+Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
+Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.
 
@@
 
-\noindent The example illustrated in the section above can occasionally be a useful simplification of relatively stable environments or when data is scarce, but it likely not realistic, since $\var(R)$ often also changes over time.
-Periods of scarcity (e.g., winter, droughts) not only have lower $\e(R)$ but also have narrower ranges of $R$.
-In contrast, periods of high $\e(R)$ (e.g., spring green-up) also often tend to have higher $\var(R)$ because $R$ tends to vary across wider ranges (e.g., due to high spatial variability, or high $R$ followed by low $R$ after grazing).
-When both $\e(R)$ and $\var(R)$ change over time (fig.
-A2), predicting $R$ becomes more complex, since disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
-Statistically, this is because the more change one attributes to $\mu(t)$ (i.e., the wigglier it is), the smaller $\sigma^2(t)$ becomes.
-Conversely, the smoother $\mu(t)$ is, the larger $\sigma^2(t)$ becomes.
-Biologically, this is important because an organism's perception scale [@levin_problem_1992] determines whether it sees a change in $R$ as a trend in $\e(R)$ or as a stochastic event (i.e., due to $\var(R)$).
+\noindent Assuming resource stochasticity is constant over time and space can be a useful simplification of relatively stable environments or when information on how $\e(R)$ changes is limited and estimating changes in $\var(R)$ is unreasonable.
+However, such an assumption is likely not realistic, since $\var(R)$ often differ across space and over time.
+Generally, strictly positive quantities tend to have correlated mean and variances.
+For example, the number of berries in a patch will be low during periods of scarcity (e.g., winter, droughts) due to both a low mean (e.g., 15 berries) and a narrow range of values (e.g., 0 to 50 berries), which implies a small variance.
+In contrast, periods of high productivity (e.g., spring green-up) will have a higher mean (e.g., 500 berries) and wider range of $R$ (e.g., 300 to 700 berries), which implies a higher variance (e.g., due differences in phenology, environmental conditions, and patch depletion).
+This relationship between a random variable's mean and variance is present in many statistical distributions where values are strictly positive, such the Gamma and Poisson distributions.
+See the section on probability distributions in Appendix A for more information.
+
+```{r, echo=FALSE, eval=FALSE}
+layout(t(1:2))
+hist(rgamma(n = 1e4, shape = 5, scale = 3))
+hist(rgamma(n = 1e4, shape = 500/3, scale = 3))
+layout(1)
+```
+
+Recognizing changes in $\var(R)$ helps account for the residual, fine-scale variation in $R$ after accounting for trends in the large-scale average $R$ [e.g., variations in plant phenology between years after accounting for mean seasonal trends, see @levin_problem_1992].
+However, when both $\e(R)$ and $\var(R)$ change over time (fig.
+A2), disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
+Statistically, this is because the more change one attributes to $\mu(t, u)$ (i.e., the wigglier it is), the smaller $\sigma^2(t, u)$ becomes.
+Conversely, the smoother $\mu(t, u)$ is, the larger $\sigma^2(t, u)$ becomes.
+Biologically, this is important because an organism's perception scale determines whether it attributes a change in $R$ to a trend in $\e(R)$ or as a stochastic event [i.e., due to $\var(R)$; see @levin_problem_1992].
 An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-Whether, an organism is able to predict trends in $\sigma^2(t)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their space-use requirements.
+Whether an organism is able to predict trends in $\sigma^2(t, u)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their range size.
 While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is still limited.
-Thus, there remains a need for developing a more complete understanding of how organisms' space-use requirements change with environmental stochasticity.
-
-<!-- Although the effect of $R$ on organisms' space-use requirements is often recognized and accounted for in ecology [@stephens_optimal_1982; @williams-guillen_resource_2006; @rickbeil_plasticity_2019], most of the focus has been on the effects of $\e(R)$, while $\var(R)$ has received far less attention [but see @nilsen_can_2005; @di_stefano_resource_2011; @rizzuto_forage_2021; @seigle-ferrand_systematic_2021].
-However, $\var(R)$ can fluctuate temporally (and spatially) due to many important drivers, including repetitive and predictable patterns.
-For example, since berries are scarce outside the growing season, the variance in berries will also be low.
-But during the growing season, both the average number of berries and the variance in berries are higher, since a bear may explore a location before fruiting time or after competitors ate all the berries.
-In contrast, other sources of stochasticity can be much less predictable, such as the arrival of new competitors [@alexander_novel_2015], the emergence of new diseases [@hollings_trophic_2014], droughts [@foley_severe_2008; @haig_assessment_2013], fires [@jolly_animal_2022], floods [@ramos_pereira_structure_2013], changes in climate and phenology [@berger_climate_2018; @grant_evolution_2017; @jonzen_rapid_2006; @inouye_climate_2000; @severson_spring_2021; @woolway_global_2020], and other extreme events [@logares_black_2012; @anderson_black-swan_2017].
--->
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is currently unknown.
+Thus, there remains a need for developing a more complete understanding of how organisms' range sizes changes with environmental stochasticity.
 
@@
 source('functions/qgamma2.R')
-plot(1:10, qgamma2(p = 0.5, mu = 1, sigma2 = 1:10))
+plot(1:100, qgamma2(p = 0.5, mu = 3, sigma2 = 1:100), type = 'l',
+     xlab = expression(sigma^2~('u,'~t)), ylab = 'Median R')
 ```
 
-Similarly to $\e(R)$, we hypothesize $\var(R)$ to have a nonlinear effect on organisms' space-use requirements.
-When $\var(R)$ is low enough that $R$ is relatively predictable, we do not expect changes in $\var(R)$ to have a noticeable effect.
-But as resources become increasingly unpredictable, we expect space-use requirements to increase progressively faster (fig.
-\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) stochastic environments tend to be less productive [@chevin_adaptation_2010], (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974], and (4) the added movement required to search for food increases organisms' energetic requirements.
-If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new behaviors (such as nomadism, dispersal, and migration) or adaptations (such as increased fat storage or food caching).
-If changes in $\sigma^2(t)$ are sufficiently predictable, organisms may learn to anticipate and prepare for periods of greater stochasticity by pre-preemptively caching food, migrating, or relying on alternative food sources during stochastic events (e.g., the seasonal arrival of competitors).
+Similarly to $\e(R)$, we hypothesize $\var(R)$ has a nonlinear effect on an organism's range size.
+When $\var(R)$ is low enough that $R$ is relatively predictable, we expect organisms to be range-resident with small home ranges, and we do not expect small changes in $\var(R)$ to have a noticeable effect.
+As resources become increasingly unpredictable, we expect home range size to increase progressively faster (fig.
+\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) the added movement required to search for food increases organisms' energetic requirements, and (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974].
+If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new and consistent behaviors (e.g, nomadism) or adaptations (e.g., increased fat storage or food caching) to buffer themselves against times of unpredictably low $R$.
+Conversely, if changes in $\sigma^2(t, u)$ are sufficiently predictable, organisms may learn to anticipate and prepare for times of greater stochasticity by pre-preemptively caching food, reducing energetic needs, migrating, or relying on alternative food sources [e.g., @van_baalen_alternative_2001].
 
-```{r s2-hyp, fig.cap="Hypothesized space-use requirements of an organism as a function of resource stochasticity ($\\var(R)$).
-We expect low values of $\\var(R)$ to result in low space-use requirements as organisms are able to depend on relatively predictable resources.
-As $\\var(R)$ increases, space-use requirements should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and space-use requirements cannot be linear because it would require space-use requirements to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth'}
+```{r s2-hyp, fig.cap="Hypothesized range size of an organism as a function of resource stochasticity, $\\var(R)$.
+We expect low values of $\\var(R)$ to result in small home-rages, since organisms are able to depend on relatively predictable resources.
+As $\\var(R)$ increases, range size should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to a larger range via dispersal, nomadism, or migration.
+Note that the relationship between $\\var(R)$ and range size cannot be linear because it would require range size to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth'}
 knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
@@
 
-\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ affect organisms' space-use requirements, but we presented the two parameters as independent drivers of movement.
-However, organisms may respond to changes in $\var(R)$ more when resources are scarce than when they are abundant.
-Consequently, models estimating the effects of $R$ on organism's movement should account for not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
+\noindent We have provided the case for why both $\e(R)$ and $\var(R)$ should be expected to affect organisms' range size, but we presented the two parameters as independent drivers of movement.
+However, organisms may respond to changes in $\sigma^2(t, u)$ more when resources are scarce than when they are abundant.
+Consequently, an organism’s movement behavior is likely to be a function of not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
 A highly unpredictable habitat may be very inhospitable if resources are poor, but $\var(R)$ may have little effect if resources are stochastic but always abundant.
-Thus, we expect $\var(R)$ to have a stronger effect on space-use requirements when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
+Thus, we expect $\var(R)$ to have a stronger effect on range size when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
 We explore this more in the following section.
-
-<!-- Most habitats with appreciable seasonal climatic changes will require models of $R$ to account for changes in both $\e(R)$ and $\var(R)$, as $\sigma^2(t)$ will often change within and between years, and periods of high $\mu(t)$ will often be more variable than times of lower productivity or dormancy.
-However, the effects of $\var(R)$ on organisms' space-use requirements likely vary between species, since movement capacity, and energetic and nutritional needs also vary between species [@boratynski_energetic_2020; @noonan_effects_2020].
-Small-scale variations in $R$ may be sufficient to causes changes in behavior for some small organisms, while organisms with larger space-use requirements may not even notice a change.
--->
 
@@
 
-\noindent To support our hypotheses of how organisms' space use is affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+\noindent To support our hypotheses of how organisms' range sizes are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
-The IOU model's correlated velocity produces realistic tracks with directional persistence despite the tracks being discrete samples of continuous-time movement processes.
-However, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
-Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory movement.
+The IOU model's correlated velocity produced realistic tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
+Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory or memoryless movement.
 
-Each of the 200 tracks were placed on a raster with common starting point $\langle 0, 0\rangle$ and sufficient time between tracks to be independent of one another (other than the starting point).
-Each time the track moved to a new cell, the organism collected resources $R$ sampled from a Gamma distribution.
-The mean and variance of the distribution were defined by deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig.
+Each of the 200 tracks were placed on a grid with common starting point $\langle 0, 0\rangle$ (fig.
+B1).
+Each time the simulated individual moved to a new cell, it collected $R$ resources sampled from a Gamma distribution.
+The mean and variance of the distribution were defined by a series of deterministic functions $\mu(t)$ and $\sigma^2(t)$ (orange and blue lines in fig.
 \@ref(fig:5-5-sims)).
-The value of $t$ was constant within each set of 200 tracks, so the time spent moving by the organism in each track did not affect the distribution $R$ was sampled from.
-Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track.
-Finally, as each set of 200 truncated was now spatially restricted and centered around $\langle 0, 0\rangle$, the set of tracks was modeled to calculate the 95% utilization distribution from an OUF model and Autocorrelated Kernel Density Estimate.
-We designed the simulations presented here to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments.
-Consequently, we made the following four assumptions: (1) $\e(R)$ and $\var(R)$ are constant over space and short periods of time (i.e., within each set of 200 tracks, but not over $t$).
-(2) Organisms move randomly in the environment with no energetic costs.
-(3) Organisms have no perceptive range or memory, so they are unable to detect, learn, or predict where resources are abundant (high $\e(R)$) or reliable (low $\var(R)$) over time.
-(4) Resources do not deplete, and there is no competition or predator avoidance.
-Additional information is provided in Appendix B.
+The value of $t$ was constant within each set of 200 tracks, so the distribution $R$ was sampled from was independent of both the organism's location and its time spent moving.
+Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track (section 2.1 of Appendix B).
+Finally, we fit an OUF movement model [@fleming_fine-scale_2014] to the set of tracks to calculate the 95% Gaussian home-range size using the formula $$H_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,$$ where $\hat \varsigma^2$ is the positional variance estimated by the movement model.
 
-```{r 5-5-sims, fig.cap="Simulated home range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (and homogenous over space).
+We designed the simulations to estimate the effects of $\e(R)$ and $\var(R)$ in simplistic environments where organisms could only respond by searching for longer periods of time.
+Consequently, we made the following assumptions:
+
+1) Environments are homogeneous for a given $t$.
+Given $t$, $\e(R) = \mu(t)$ and $\var(R) = \sigma^2(t)$ are constant over space and within each set of 200 tracks, but $R$ is random and follows a $\Gamma\large(\mu(t), \sigma^2(t)\large)$ distribution.
+2) The are no external pressures on the simulated organism. Resources do not deplete, and there is no competition nor predator avoidance.
+3) The organism has a fixed daily energetic requirement that is independent of movement rates, and it cannot alter its metabolism or physiology.
+Additionally, the organism does not have energetic reserves, so excess resources cannot be carried over to the next track or $t$.
+4) The organism is range-resident and can only respond to changes in $\e(R)$ and $\var(R)$ by altering its home-range size. The organism does not disperse or abandon a range.
+5) The organism's movement is simplistic. The organism's movement speed and direction are stochastic and independent of $\e(R)$ and $\var(R)$.
+6) The organism has no perceptive range or memory. It is unable to detect, learn, or predict where resources are abundant (high $\e(R)$) or reliable (low $\var(R)$) over time or space.
+7) Animals only move to search for food or return to the center of their home-range after reaching satiety.
+
+Additional information is provided in Appendix B, including the directed acyclical graph [see fig.
+B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with fig.
+C5 and the empirical case study below).
+
+```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogenous over space for a given $t$).
 Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
@@
 Fig.
-\@ref(fig:5-5-sims) shows how simulated space-use requirements responded to changes in $\mu(t)$ and $\sigma^2(t)$ in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically.
-The top row <!--(constant $\var(R)$) -->shows how space-use requirements vary for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
+\@ref(fig:5-5-sims) shows how simulated home-range size, $H$, responded to changes in $\mu(t)$ and $\sigma^2(t)$ in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically.
+The top row (constant $\var(R)$) shows how $H$ varies for different trends in $\mu(t)$ while $\var(R)$ remains constant (like in fig.
 A1).
-As $\e(R)$ increases at a constant slope (linear $\mu(t)$) $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism.
-Although less evident, note how the noise in the lines also decreases as $\e(R)$ increases.<!-- As described in the section on the effects of resource abundance, the nonlinear decrease is because changes in $\mu(t)$ have a larger effect when $\e(R)$ is low [e.g., @lucherini_habitat_1996; @nilsen_can_2005; @herfindal_prey_2005].-->
-
-<!-- In regions where the $\mu(t)$ changes over time [e.g., seasonal changes -- see cyclical $\e(R)$ and @lai_movement_2017], organisms should have access to sufficient space to fulfill their needs during periods of scarcity, whether the space available is sufficiently large year-round or it changes seasonally with the expected changes in $\mu(t)$ (e.g., winter park closures).
-However, estimates of spatial requirements based on estimated changes in $\mu(t)$ should be interpreted carefully, since model error and unforeseeable decreases in $\e(R)$ (such as following fires or floods) may increase organisms' space-use requirements suddenly and unpredictably.
-Thus, it is best to include a "buffer" area so the available space is larger than the estimated space-use requirements.
-This is particularly the case in environments where resource abundance changes unpredictably (drifting $\mu(t)$), since accurate long-term estimates and predictions of $\mu(t)$ may be hard to produce, if not impossible.
-In cases where $\mu(t)$ is highly unpredictable, organisms should have enough space to collect resources during times of greatest predicted scarcity, particularly if the changes in resource abundance occur rapidly, often, or for long periods of time (erratic $\mu(t)$).
--->
+As $\e(R)$ increases at a constant slope (linear $\mu(t)$), $H$ decreases nonlinearly, with larger changes when $\e(R)$ is low, until it approaches the minimum size required by the organism.
+Also note how the noise in the green lines also decreases as $\e(R)$ increases.
 
 The leftmost column of fig.
-\@ref(fig:5-5-sims) <!--(constant $\e(R)$) -->illustrates the effects of $\var(R)$ on the simulated space-use requirements while $\e(R)$ remains constant.
+\@ref(fig:5-5-sims) (constant $\e(R)$) illustrates the effects of $\var(R)$ on $H$ while $\e(R)$ remains constant.
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
@@
 The remaining panels in fig.
-\@ref(fig:5-5-sims) illustrate $\e(R)$ and $\var(R)$ jointly affect $H$ and how complex the effects can be.
+\@ref(fig:5-5-sims) illustrate how $\e(R)$ and $\var(R)$ jointly affect $H$ and how confusing the effects can be.
 Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
@@
 
-<!-- Not all 25 scenarios depicted in fig.
-\@ref(fig:5-5-sims) may be realistic, but the trends in $\e(R)$ and $\var(R)$, and their impacts on space use are useful examples that can be thought of as simplified scenarios.
-$\e(R)$ and $\var(R)$ may be assumed to be (approximately) constant in highly homogeneous environments.
-Although it is impossible for $\mu(t)$ and $\sigma^2(t)$ to increase linearly forever, these examples are useful to demonstrate that linear changes in $\mu(t)$ and $\sigma^2(t)$ affect $H$ nonlinearly.
-Cyclical oscillations in $\e(R)$ and $\var(R)$ may occur in urban environments as human activity changes within and between days [@peron_periodic_2017; @ikeda_home_2022] and as temperatures fluctuate daily and seasonally [@geremia_migrating_2019; @alston_temperature_2020], while $\e(R)$ and $\var(R)$ may drift randomly in highly complex environments which are too hard to predict.
-Finally, erratic changes in $\e(R)$ and $\var(R)$ may occur in environments where changes are very sudden, such as areas prone to fires or floods, as well as habitats with drastic human alteration (e.g., a forest which is clear-cut with a subsequent artificial re-forestation).
-However, if stochastic changes are sufficiently small and frequent, organisms may perceive them as continuous and smooth changes rather a series of small and sudden changes.
--->
-
 Although the temporal trends in fig.
 \@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time (panels A and B of fig.
-\@ref(fig:5-5-reg-mean)).
+\@ref(fig:5-5-reg)).
 The estimated relationships follow the hypotheses we presented in figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation as $\var(R)$ increased.<!--We ensured this was not due to a high minimum or low maximum $\var(R)$ (fig.
-\@ref(fig:s2-hyp)) by simulating within the range $\sigma^2(t) \in [2\,500, 250\,000] \implies \sigma \in [50, 500]$, since the range of $\e(R)$ was $\mu(t) \in [100, 200]$.--> However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C).
-The increase in $\e(H)$ over $\var(R)$ is strongly nonlinear at low values of $\e(R)$, but it is not affected by $\var(R)$ much when $\e(R)$ is very low (and $\e(H)$ is consequently very high).
+\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation at high values of $\var(R)$.
+However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C): When $\e(R)$ is low, $\e(H)$ is high and $\var(R)$ does not have a strong effect, but when $\e(R)$ is high the effect of $\var(R)$ on $\e(H)$ is exponential.
+Similarly, $\e(H)$ decreases exponentially with $\e(R)$ except when $\var(R)$ is very high.
 
-<!-- While we modeled the effects of $\e(R)$ and $\var(R)$ on both the 95% and the 50% space-use requirements, we only show results for the 95^th^ percentile because the results the 50^th^ quantile followed the same trends, other than the multiplicative rescaling required to account for the difference in magnitude between the two quantiles.
--->
-
-```{r 5-5-reg-mean, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on simulated spatial requirements with 95\\% Bayesian credible intervals for the mean.
-While the effect of $\\var(R)$ is sublinear in panel B, the effect of $\\var(R)$ depends strongly on $\\e(R)$ (see panel C).\nCredible intervals were calculated using 10,000 samples from the posterior distribution.
-The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
-Details on the model structure are provided in Appendix B.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-mean.png')
-```
-
-As expected by the changes in the spread of the points in panels A and B of figure \@ref(fig:5-5-reg-mean), the variance in $H$ also depends on $\e(R)$ and $\var(R)$ (figure \@ref(fig:5-5-reg-var)A-B).
-Here, the effects of the two variables are more pronounced, even for low values of $\e(R)$ or high values of $\var(R)$.
+As expected by the changes in the spread of the points in panels A and B of fig.
+\@ref(fig:5-5-reg), the variance in $H$, $\var(H)$, also depends on $\e(R)$ and $\var(R)$ (fig.
+\@ref(fig:5-5-reg)D-F).
+Since we modeled $H$ using a Gamma family of distributions, we expected $\var(H)$ to increase with $\e(H)$, but the location-scale model removes the assumption of a constant mean-variance relationship (i.e., constant coefficient of variation, $\frac{\mu(t)}{\sigma^2(t)}$.
+This allowed us to show that the effect of $R$ on $\var(H)$ is much stronger than the effect of $R$ on $\e(H)$.
 Consequences of these effects are explored in the discussion section.
 
-```{r 5-5-reg-var, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on the variance in simulated spatial requirements, with 95\\% Bayesian credible intervals for the estimated variance.
-Credible intervals were calculated using 10,000 samples from the posterior distribution.
+```{r 5-5-reg, eval=TRUE, echo=FALSE, fig.cap="Effects of $\\e(R)$ and $\\var(R)$ on on the mean (A-C) and variance (D-F) in simulated home-range size with 95\\% Bayesian credible intervals.
+While the estimated marginal effect of $\\var(R)$ on $\\e(H)$ is sublinear (panel B), the effect of $\\var(R)$ is superlinear for high values of $\\e(R)$ (panel C).
+The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
+Credible intervals were calculated using 10,000 samples from the posterior distribution while assuming multivariate Gaussian coefficients.
 Additional details on the model structure are provided in Appendix B.", out.width='100%'}
-knitr::include_graphics('../figures/simulation-regression-var.png')
+knitr::include_graphics('../figures/simulation-regression-mean-and-var.png')
 ```
@@
 
-\noindent The simulations in the section above mostly follow the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural enviroments.
-Organisms live in spatiotemporally heterogeneous and dynamic environments, which the development of perceptual ranges, navigation, and memory.
+\noindent The simulations in the section above support the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural environments.
+Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory.
 Together, these abilities result in selective space use that depends on resource availability [@kacelnik_ideal_1992] and resource depletion [@charnov_optimal_1976].
 
-In this section, we test the hypotheses illustrated in the introduction using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with estimates of $\e(R)$ and $\var(R)$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], a remote-sensed measure of lanscape greeness, as a proxy of forage abundance.
-Appendix C contains additional information on how NDVI was modeled along with details on the continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022] used to quantify the tapir's space-use requirements.
-Through this example, we aim to provide support for the two hypotheses using empirical data that do not depend on as strong assumptions as those required by the simulations in the section above.
-Concurrently, we also hope to show the potential complications that arise with empirical data.
+In this section, we test the hypotheses using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with empirical estimates of $\e(R)$ and $\var(R)$.
+We measure $R$ using Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], a remote-sensed measure of landscape greenness, as a proxy for forage abundance.
+Appendix C contains additional information on how we modeled NDVI and the tapir's movement using continuous-time movement models [@fleming_ctmm_2021; @noonan_scale-insensitive_2019] and autocorrelated kernel density estimation [@noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
-<!-- The hypotheses we present here allow researchers to combine otherwise complicated findings (e.g., fig.
-\@ref(fig:5-5-sims)) and generalize results to a single, common set of functions.
-We expect the two hypotheses we present here (figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp)) to be applicable to all motile organisms, once differences in size, metabolic needs, and diet are accounted for.
-In this subsection, we illustrate how researchers can leverage these hypotheses using their own data to build on previous work rather than listing the results as new findings.
-Additionally, the analyses can be expanded to the population or species level with the use of Hierarchical Generalized Additive Models (HGAMs) with individual-level intercepts and smooth terms [see models "GS" and "GI" in @pedersen_hierarchical_2019].
-Such models would allow one to test whether members of the same population or species respond similarly to $\e(R)$ and $\var(R)$ while also quantifying any deviation from the hypothesized behavior.
-Individuals and populations that were once thought to have different space-use requirements due to differences in behavior may instead be simply responding to different local conditions along the functions in figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp) [e.g., @singh_migration_2012].
-In the following sections, we illustrate how one can apply this framework to empirical data.
--->
-
-\noindent Fig.
-\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado [data from the individual labelled as "Anna" in the dataset from @medici_movement_2022] adapts its spatial needs to changes in $\e(R)$ and $\var(R)$.
+Fig.
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in $\mu(t, u)$ and $\sigma^2(t,u)$ [telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022].
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
-The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; examples: @bjorndahl_abrupt_2022; @marien_timing_2022; @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
+The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; example: @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
 Panel C shows the changes in the tapir's 7-day home range over time.
@@
 Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018).
-Finally, panels D and E show the estimated (marginal) effects of $\mu(t)$ and $\sigma^2(t)$ on the tapir's space use.
-Since $\mu(t)$ and $\sigma^2(t)$ are strongly correlated (panel F) and (spatio)temporally autocorrelated (panels A and B), the effects of $R$ on $H$ should be modeled carefully.
-To avoid over-fitting the model, we constrained the smooth effects of $\mu(t)$ and $\sigma^2(t)$ using low number of knots ($\tt{k = 3}$).
+Finally, panels D and E show the estimated (marginal) effects of $\mu(t, u)$ and $\sigma^2(t, u)$ on the tapir's 7-day home-range size.
+Since $\mu(t, u)$ and $\sigma^2(t, u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully.
+To avoid over-fitting the model, we constrained the smooth effects of $\mu(t, u)$ and $\sigma^2(t, u)$ and their interaction effect to a small basis size ($\tt{k = 3}$).
 Additional information is provided in appendix C.
-Overall, the results presented in fig.
-\@ref(fig:tapir-mw) support our hypotheses, as the taprir's 7-day home range decreases nonlinearly with $\mu(t)$ and increases nonlinearly with $\sigma^2(t)$.
-However, due to the strong nonlinear correlation between the two predictors, the marginal effects do not follow the data closely, particularly in the case of $\mu(t)$.
-We provide comments about these results in the related section in the discussion below.
+The results presented in panels D-F of fig.
+\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\mu(t, u)$ and increases with $\sigma^2(t, u)$, and the effect of $\mu(t, u)$ depends on $\sigma^2(t, u)$, and vice-versa.
+Alone, $\mu(t, u)$ and $\sigma^2(t, u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
+Additionally, note how high NDVI values (> 0.8) cause $\sigma^2(t, u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
 
-```{r tapir-mw, fig.cap="Seven-day home-range size of a lowland tapir (\\emph{Tapirus terrestris}) in response to changes in mean and variance in resources.
-(A) Trends in resource abundance over time, $\\mu(t)$, esimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
-(B) Varince in resources over time, $\\sigma^2(t)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
-(C) Estimated seven-day home range based on the 95\\% utilization quantiles.
-(D, E) Estimated marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ and their interaction on mean space-use requirements and the variance around them.
-(F) The effect of $\\mu(t)$ does not follow the data closely because $\\e(R)$ and $\\var(R)$ are highly correlated.
-Consequently, while estimating the effects of $\\e(R)$ and $\\var(R)$ via separate models would allow result in a closer fit, the estimated effects would be inappropriate because they do not disentangle the effects of $\\e(R)$ and $\\var(R)$.
+```{r tapir-mw, fig.cap="Effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}).
+(A) Trends in resource abundance over time, $\\mu(t, u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
+(B) Varince in resources over time, $\\sigma^2(t, u)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
+(C) Seven-day 95% home range estimated using Autocorrelated Kernel Density Estimation.
+(D, E) Estimated marginal effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ on home-range size.
+The model accounted for the marginal effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ and their interaction effect.
+(F) Estimated home-range size in response to changes in both $\\mu(t, u)$ and $\\sigma^2(t, u)$.
+Note how the effect of $\\sigma^2(t, u)$ is more pronounced when $\\mu(t, u)$ is low.
 See Appendix C for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici $et~al.$ (2022).", out.width='100%'}
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} (2022).", out.width='100%'}
 knitr::include_graphics('../figures/tapir-example.png')
@@
 
-\noindent This paper presents two hypotheses for describing the effects of resource abundance and stochasticity on organisms' space-use requirements.
-We support the hypotheses using quantitative simulations and an empirical case study.
-Each of the two methods allows us to highlight the strengths and limitations of simulation-based and data-driven approaches, which we discuss below.
+\noindent The amount of space organisms is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
+This paper presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
+We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments.
+
+Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$.
+Together, the degree to which $\e(R)$ affects $H$ depends on$ $\var(R)$, and vice-versa, so organisms' responses to resource dynamics can be complex.
+The simulated and empirical results suggest qualitatively similar marginal effects of $\e(R)$ and $\var(R)$, but there are differences in the estimated interactive effects.
+In the simulated data, $\var(R)$ has little effect when $\e(R)$ is low and a strong effect when $\e(R)$ is high, while the opposite is true for the empirical data.
+This is due to two reasons.
+Firstly, the shape and symmetry of bounded distributions such as Gamma ($R > 0$) and Beta ($0 < R < 1$) distributions depend on both $\e(R)$ and $\var(R)$ (figs.
+A3, A4), but $\var(R)$ does not affect the shape of a Gamma distribution as much if $\e(R)$ is low (fig.
+B3).
+Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to $\e(R)$ and $\var(R)$ such as selective space use, which we account for in the empirical approach.
+Below we discuss the strengths and limitations of each approach.
 
@@
 
-\noindent The simulations we present here provide a simple and direct estimate of the effects of resource abundance and stochasticity on organisms' space-use requirements.
-The combination of simplicity (few drivers of movement) and realism (continuous-time movement models) allowed us to estimate the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements using simple causal models of inference without the confounding variables that are often present in empirical data (e.g., predator avoidance, territoriality, competition, landscape connectivity).
-Additionally, since our sample size was virtually unlimited, we were able to simulate sufficient data to ensure the simulations were stable and the resulting estimated coefficients were well estimated (see Appendix B).
+\noindent Our simulations are based on a simplistic environment with many assumptions that allowed us to estimate how resource abundance and stochasticity affect organisms' home-range sizes if organisms can only respond to changes by adapting the amount of time spent searching for food (with no energetic cost to movement).
+The use of continuous-time movement models coupled with few drivers of movement supported realistic data that could be explained by straightforward causal models.
+The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).
 
-Avoiding confounding variables rather than attempting to account for them (as is required with empirical data) allowed us to develop theory easily while providing a baseline model against which one can contrast empirical examples.
-In figures B6, we present the Directed Acyclical Graph [DAG, see @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct, partial effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with figure C5 and the section below).
-
-By quantifying the deviations from the simple theoretical models, one can then assess the importance of other drivers movement and behavior.
-Still, the simplicity of the underlying model and its (often invalid) assumptions often limit the applicability of simulated models in realistic scenarios.
-Responses to $\e(R)$ and $\var(R)$ in real ecosystems likely depend on many other drivers of movement, including competition, movement costs, and predation, which may affect the effects of $\e(R)$ and $\var(R)$ on both $\e(H)$ and $\var(H)$ [@jetz_scaling_2004; @noonan_search_2023].
-And while it is possible to account for all of these factors through simulations, empirical data often can provide evidence for these effects without the need for complex model structures.
+Deviations from the simulations offer a means of detecting when the underlying assumptions are inappropriate and how additional factors may affect organisms' responses to changes in $\e(R)$ and $\var(R)$.
+For example, energetic costs of movement are often non-negligible and depend on organism size [@taylor_energetics_1982], movement speed [@taylor_energetics_1982], and ambient temperature [@brown_toward_2004; @hou_cold_2020].
+In addition, an organism may alter its movement behavior, physiology, and energetic needs to buffer itself against changes in $\e(R)$ and $\var(R)$ by using space selectively [@johnson_comparison_1980; @rickbeil_plasticity_2019; @riotte-lambert_environmental_2020; @ranc_memory_2022] and adapting their behavior and physiology over time [@southwood_habitat_1977; @levins_evolution_1974].
+Before or during periods of scarcity, organisms may cache resources [@nespolo_why_2022], enter states of dormancy [@goldberg_hibernation_2021; @reher_short_2018], or even pause fetal growth [@schmidt_interplay_2020].
+However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], and avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969].
+Failure to respond optimally may thus force organisms to relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], and may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.
 
@@
 
-\noindent In the absence of direct measurements of $R$, empirical models estimating the effects of $R$ on $H$ must depend on proxies of $R$, such as NDVI.
-However, using proxies of $R$ rather than direct measurements introduces sources of bias.
-While $R$ and NDVI are correlated [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019] and nonlinear [@ref].
-This is because remote-sensed proxies of $R$ and they do not measure resource availability directly, so they introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the DAG in figure C5) and satellite-level confounding variables ($S$ in figure C5).
+\noindent There are two main advantages of using and empirical approach.
+Firstly, using real-world animal movement data can produce scale-appropriate and easily interpretable estimates.
+Secondly, empirical models directly quantify the effects of $\e(R)$, $\var(R)$, and confounding variables without having to design complex and time-consuming simulations.
+However, it is not always possible to quantify confounding variables, such as competition.
+While there may be some appropriate proxies, such as density competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (the presence of competitors may not reflect competitive pressure).
+This is problematic if one is interested in the direct causal effect of $\e(R)$ and $\var(R)$, which requires removing any non-negligible confounding effects [@mcelreath_statistical_2016].<!-- However, it may not be an issue if researchers are interested in estimating total causal effects, whether they be direct causal paths or not [@mcelreath_statistical_2016].
+For example, one may choose not to account for competition if $\e(R)$ influences competitor density but competitors do not have a measurable effect on $R$.
+In contrast, if competitor abundance determines $R$ (and not vice-versa), then failing to account for competitors will produce incorrect estimates of the causal effects.-->
 
-Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and $NDVI$ [@jetz_scaling_2004; @prox_framework_2020].
-Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators, which may increase pressures from competition [@ref] and predation [@brown_ecology_1999].
-While such pressures may increase avoidance behaviors, avoidance may have a cmplex effect on $H$, since it can result in both an expansion of the home range [@jetz_scaling_2004; @prox_framework_2020] or a contaction of an animal's movement, since larger home ranges can be harder to defend and also result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004].
-Additionally, persistent territorial defense from competitors may prevent organisms from using space freely and as necessary [theoretical framework: @despotic_ref; examples: @rich_anthropogenic_2012; @bengsen_feral_2016; @torrez-herrera_monkeys_2020].
+Similarly, if $R$ is often non-measurable.
+Proxies of $R$, such as NDVI [@pettorelli_normalized_2011], which may introduce complexities.
+While $R$ and NDVI are correlated for many species [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016].
+This can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig.
+C3) and satellite-level confounding variables ($S$ in fig.
+C3).
 
-Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@ref], satellite-level error [@tian_evaluating_2015; @huang_commentary_2021; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources, competitive pressure, and predator avoidance).
-However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of NDVI [e.g., young grass rather than mature grasslands, see @ref].
-In the case study we present here, sampling limitations resulted in a strong nonlinear correlation between $\e(R)$ and $\var(R)$ that prevented us from estimating nonlinear preferences in mean NDVI (i.e., the proxy for $\e(R)$), since estimating the divers of movement and disentangling them from potential confounds in a stochastic environemtn is often complex or impossible [e.g., @wang_stochastic_2019].
-A tracking time series of two or more years or additional individuals would have provided better estimates, but that was beyond the scope of this empirical example.
+***HERE***
 
-**here:** need a few lines to tie this back to the general
+Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020].
+Resource-rich patches can attract larger amounts of competitors [@kacelnik_ideal_1992] and predators [@duncan_lifehistory_2015], which may, in turn, increase pressures from competition and predation [@charnov_optimal_1976; @brown_ecology_1999].
+However, such pressures may result in both an expansion of the range [@jetz_scaling_2004; @prox_framework_2020] or a contraction, since larger ranges can be harder to defend and result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004] and encounter rates [@martinez-garcia_how_2020].
 
-<!-- Feel free to delete this: -->
-<!-- While the tracking period may be too short to see the effects of $\e(R)$ clearly, it should also be noted that the Cerrado is relatively rich in vegetation throughout the entire year (fig.
-\@ref(fig:tapir-mw)A), and thus the tapir does not experience resource scarcity during the tracking period.
-Consequently, this example demonstrates two points central to this paper.
-Firstly, even animals in resource-rich regions such as the Brazilian Cerrado adapt their behavior in response to reosurce stochasticity.
-Secondly, trends in mean resources alone are not sufficient to explain the tapir's space use, and the stochasticity in $R$ is an important driver of the tapir's movement.
--->
-
-<!-- 
-## Adaptations to changes in $\e(R)$ and $\var(R)$
-
-Paragraphs to tie in:
-
-* @rizzuto_forage_2021 found that the space-use requirements of snowshoe hares (*Lepus americanus*) increased with average carbon to nitrogen ratio (C:N, a measure of N scarcity) in lowbush blueberry as well as the coefficient of variation (the standard deviation divided by the mean, $\frac{\sqrt{\sigma^2(t)}}{\mu(t)}$) in C:N.
-However, it is hard to determine how $\sigma^2(t)$ affected the hares' space-use requirements since the coefficient of variation is a function of both $\mu(t)$ and $\sigma^2(t)$, so the effects of the two variables are confounded.
-Similarly, @mueller_how_2011 suggested that ecosystem variability (measured as spatial semivariance in NDVI) causes ungulates to move more and adopt more nomadic behaviors, but it is hard to determine whether the higher semivariances in NDVI are due to an increase in $\var(R)$ or simply spatial changes in $\e(R)$ (e.g., increased seasonality or continentality).
-Likewise, @herfindal_prey_2005 found that the home ranges of Canadian lynx (*Lynx canadensis*) decreased with seasonality, but this may be an artifact of a lower $\mu(t)$ in highly seasonal environments, rather than stochasticity in $R$.
-A subsequent analysis by @nilsen_can_2005 of the lynx data of @herfindal_prey_2005 along with data on wolverines (*Gulo gulo*), bobcats (*Lynx rufus*), and coyotes (*Canis latrans*) found that carnivores respond differently to seasonality, which suggests that different species may have different perceptions of the changes in $R$ over the seasons.
-Collectively, these studies provide support towards the effects of both $\e(R)$ and $\var(R)$ on organisms' space-use requirements, but the lack of a unifying framework across these studies makes it challenging to synthesize their findings.
-Thus, it is important to disentangle the effects of changes in $\e(R)$, such as seasonal trends in $\mu(t)$, and stochasticity ($\var(R)$).
-The need to account for changes in both $\e(R)$ and $\var(R)$ is compounded by recent changes in climate, which expose species to novel situations and increasingly common and extreme stochastic events [@intergovernmental_panel_on_climate_change_climate_2023; @noonan_situ_2018; @yao_emergence_2022].
-* @nilsen_can_2005 also show that the effect of $\e(R)$ on the home-range size of wolves and fishers (*Martes pennanti*) depends on seasonality (and vice-versa), since organisms in habitats with high and low seasonality had opposite responses to $\e(R)$.
-Therefore, it may not be sufficient to account for the effects of $\e(R)$ and $\var(R)$; models may also need a term for interaction between the two.
-This may be because organisms respond strongly to $\sigma^2(t)$ when $\mu(t)$ is low, but they may not respond if $\mu(t)$ is sufficiently high.
-However, it does not explain why organisms may respond to $\e(R)$ in opposite ways in different habitats.
-Instead, we suspect this inconsistency is because @nilsen_can_2005 do not distinguish between changes in $\mu(t)$ and $\sigma^2(t)$ between seasons, so the effects of the two are confounded in the seasonality term.
-Distinguishing between changes in $\e(R)$ and $\var(R)$ allows one to separate the two seasonal cycles and produce more consistent results.
-
-The spatiotemporal scale over which an event occurs is a main determinant of whether an organism will be able to predict the event and how it will respond to it (or fail to do so).
-Events that occur at the scale of the organism [i.e.
-approximately one or more times per generation and on the spatial scale of the organism's spatial range, see @frankham_importance_2004] are more likely to be perceived as a threat or boon worth preparing for [@foley_severe_2008; @berger_climate_2018; @geremia_migrating_2019].
-Consequently, organisms with long generations (e.g., elephants, *Loxodonta sp.*) or large spatial ranges (e.g., migratory birds) are more likely to experience stochastic events and thus are also more likely to respond to and prepare for them.
-Animals may adapt by adjusting the timing of reproduction [@southwood_habitat_1977; @boersma_wildfire_2021] and torpor or hibernation [@inouye_climate_2000; @goldberg_hibernation_2021], or storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [@armitage_energetics_2003; @nespolo_why_2022] or resource caching [@post_caching_2006].
-Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory [@fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; @geremia_migrating_2019; @abrahms_memory_2019; @cavedon_genomic_2022; @foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; @ranc_memory_2022], which is required for migration [@mueller_social_2013; @geremia_migrating_2019; @merkle_spatial_2019; @abrahms_memory_2019; @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].
--->
-
-<!-- 
-with species names:
-Organisms may adapt by adjusting the timing of reproduction [framework: @southwood_habitat_1977; red-backed fairywren, *Malurus melanocephalus*: @boersma_wildfire_2021] and torpor or hibernation [marmots, *Marmota flaviventris*: @inouye_climate_2000; ground squirrels, *Urocitellus brunneus*: @goldberg_hibernation_2021], storing large amounts of energy [@lindstedt_seasonality_1985] through fat storage [marmots: @armitage_energetics_2003; bears, *Ursus sp.*: @nespolo_why_2022] or resource caching [@post_caching_2006].
-Alternatively, animals may alter their movement and travel long distances over short periods of time [@jetz_scaling_2004; @hirt_general_2017], whether this be for migration [@geremia_migrating_2019; @mueller_social_2013] or nomadism [@teitelbaum_how_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
-Over the years, animals may also develop more complex cognitive abilities [@brown_toward_2004], including the development of memory of events [framework: @fagan_spatial_2013; @riotte-lambert_environmental_2020; @steixner-kumar_strategies_2020; bison and mule deer: @geremia_migrating_2019; blue whales, *Balaenoptera musculus*: @abrahms_memory_2019; caribou, *Rangifer tarandus*: @cavedon_genomic_2022; elephants: @foley_severe_2008; @polansky_elucidating_2015; elk, *Cervus canadensis*: @falcon-cortes_hierarchical_2021; @rickbeil_plasticity_2019; roe deer, *Capreolus capreolus*: @ranc_memory_2022], which is required for migration [whooping cranes, *Grus americana*: @mueller_social_2013; bison and mule deer: @geremia_migrating_2019; mule deer: @merkle_spatial_2019; blue whales: @abrahms_memory_2019; elk: @rickbeil_plasticity_2019; @middleton_green-wave_2018; but see: @cuadrado_assessing_2021; @stefanescu_timing_2021].-->
-<!-- However, adapting to changes in environmental cues rather than relying on memory alone [@severson_spring_2021; @rickbeil_plasticity_2019; @buntgen_elevational_2017; @jonzen_rapid_2006], can prevent animals from becoming trapped in sub-optimal conditions [@abrahms_memory_2019].
-
-In contrast, individuals with short lifespans are less likely to experience stochastic events and develop memory about them.
-However, the short generation time and high reproductive rate of r-selected species [as opposed to k-selected species, see @pianka_r-_1970; @brown_toward_2004] may promote other traits that increase survival following extreme events.
-Examples include higher population size, dispersal, higher genetic diversity, and complex group dynamics [e.g., @cuadrado_assessing_2021; @stefanescu_timing_2021].
-Events that occur over large spatiotemporal scales can thus promote slow, large-scale changes through the adaptation, natural selection, and evolution of the species [@grant_evolution_2017; @logares_black_2012; @anderson_black-swan_2017; @gienapp_climate_2008], which is more likely to adapt if starting populations are large with high genetic diversity and short generations, as in the case of r-selected species [but see @leung_smaller_2022].
-Of course, an organism with a lifespan shorter than a year will be unable to experience the cyclical nature of the seasons, and an organism with a lifespan of a few weeks or days may not even experience substantial changes in weather other than daily cycles and some weather stochasticity.
-Still, a population may respond to spatiotemporal cycles in $\e(R)$ and $\var(R)$ by adapting the group's reproduction timing and migration [e.g., painted lady butterflies: @cuadrado_assessing_2021; @stefanescu_timing_2021].
-
-A population's fitness and chances of survival in a changing environment depend strongly on its adaptability and responsiveness to change [@riotte-lambert_environmental_2020], and its ability to recognize cues that precede extreme events or periods of scarcity or abundance can have appreciable effects on its members' chances of survival and reproduction.
-For example, ungulates' ability to time their yearly migrations on available forage biomass, snow fall, snow melt, and hunting pressure allow them to adapt to changes in both $\e(R)$ and $\var(R)$, while their spatial memory [@falcon-cortes_hierarchical_2021; @ranc_memory_2022] allows them to optimize their large-scale movement by reducing the amount of time and energy spent exploring.
-Similarly, the ability of trans-Saharan migrant birds to adapt their migration timing based on environmental cues allows them to account for differences in $\mu(t)$ between years [@jonzen_rapid_2006].
-However, relying too strongly on short-term changes rather than depending on long-term memory can also cause animals to fail to find crucial resources during unpredictable times [whooping cranes: @mueller_social_2013; elephants: @polansky_elucidating_2015; @foley_severe_2008; whales: @abrahms_memory_2019].
-Instead, if anomalies occur repeatedly over time, organisms may adapt to a higher $\sigma^2(t)$ or shift their ranges altogether [terrestrial animals: @chen_rapid_2011; ungulates: @severson_spring_2021; @buntgen_elevational_2017], including migratory animals [@jonzen_rapid_2006].
-Still, while range shifts can help species overcome changes in the long-term trends in $\mu(t)$ and $\sigma^2(t)$, including changes in phenology [@severson_spring_2021], they can also result in phenological mismatches [@lameris_arctic_2018] and add pressure to the species that already live in the habitat [e.g., @wallingford_adjusting_2020; but see @shepard_predators_2022], including specialists and obligate symbionts that may be unable to shift their range, since it is determined by their food sources' and associates' ability to move and adapt, too.
-
-<!-- ## Applications for conservation -->
-
-<!-- \noindent Technical advancements from the last few decades have allowed scientists to collect increasingly larger amounts of data at finer and finer scales [@nathan_big-data_2022], but little attention has been paid to estimating the effects of environmental stochasticity on organisms' behaviors.
-In these times of great change and uncertainty [@intergovernmental_panel_on_climate_change_climate_2023; @abrahms_human-wildlife_2021; @sih_evolution_2011], it is increasingly important to recognize how environmental stochasticity affects organisms' behaviors and shapes ecosystems, particularly when assessing whether the quality of protected areas is sufficiently high to sustain biodiversity long term.
-Protecting areas with high resource abundance and low resource stochasticity is an important first step towards reducing animals' physiological stress, inter- and intra-specific competition, community instability, and human-wildlife conflict.
--->
-
-<!-- The designation of high-quality habitats is a central concern of many conservation initiatives.
-This framework allows consultants and conservation managers to quantify the quality of a habitat based on its resource abundance and stochasticity as well as the amount of space animals would need in the habitat.
-With the methods presented in this paper, one can (1) assess the quality of current protected areas, (2) predict whether or not the areas will be able to sustain current populations in future decades, and (3) evaluate the quality of new potential conservation areas.
-In light of the UN's objective to conserve 30% of global landmasses and waters by 2030 [@ipbes_global_2019], the framework and methods presented here provide useful metrics for evaluating whether protected areas will have sufficient quality and long-term stability to support species under various possible climate change scenarios [@intergovernmental_panel_on_climate_change_climate_2023].
--->
-
-<!-- Ignoring the effects of recent and predicted increases in $\var(R)$, including an increase in the intensity and frequency of extreme events [@yao_emergence_2022; @intergovernmental_panel_on_climate_change_climate_2023], has the potential to greatly hinder future conservation efforts.
-Since not accounting for an increase in $\sigma^2(t)$ will result in an under-estimation of animals' space-use requirements, protected areas which were designated based on the area's $\mu(t)$ but not $\sigma^2(t)$ will likely be insufficient in both size and resources in the future, which may lead to decreases in the fitness and size of a population while also increasing the risk of population collapse and human-wildlife conflict [@mukeka_human-wildlife_2019].
-This is particularly the case in resource-poor, stochastic regions, where animals are forced to endure longer and more unpredictable searches, which results in a larger variance in space-use requirements.
--->
-
-<!-- It is also worth noting that carnivores may be more susceptible to changes in $\e(R)$ and $\var(R)$, and that this susceptibility is likely amplified further at higher trophic levels.
-While primary consumers can rely on relatively predictable and static resources that can be estimated easily via remote sensing measures such as NDVI, the  fine-scale location of secondary consumers' resources may not correlate as strongly with measures of habitat productivity.
-In fact, primary consumers' ability to move makes their location and abundance more stochastic than those of plants.
-Consequently, the abundance of primary consumers may correlate with productivity weakly or at a temporal lag [e.g., @geremia_migrating_2019].
-Thus, primary consumers' delayed responses to fluctuations in $\e(R)$ and $\var(R)$ may result in an increase in secondary consumers' resource stochasticity, which would also be perceived with a delay.
-Similar hypotheses can be formulated for higher trophic levels but with amplified temporal lags, since each trophic level adds another layer of uncertainty and delayed response, which together increase the unpredictability on how carnivores indirectly respond to changes in primary productivity.
-However, the effect is likely less pronounced for omnivores and other diet generalists, as they may have the ability to switch to more abundant and less stochastic resources.
--->
+Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@huang_commentary_2021; @fan_global_2016], satellite-level error [@huang_commentary_2021; @tian_evaluating_2015; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources or small-scale resource depletion).
+However, nonlinear models such as Generalized Additive Models [@wood_generalized_2017] can help account for preferences for intermediate values of remotely-sensed  $R$ [e.g., young grass rather than mature grasslands, see @huang_commentary_2021].
 
@@
 
-**needs a few more lines at the end**
+\noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
+We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' range sizes and movement behavior.
+We demonstrate that organisms' range sizes decrease with resource abundance, increase with resource stochasticity, and that the effects of $\var(R)$ can depend strongly on $\e(R)$.
 
-\noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
-We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' space-use requirements and movement behavior.
-We demonstrate that organisms' space-use requirements respond nonlinearly to both resource abundance and stochasticity, and we demonstrate the importance of accounting for $\var(R)$ besides $\e(R)$.
-We also provide a moving-window approach that allows one to model movement continuously rather than imposing arbitrary thresholds between different periods or types of behaviors, including temporal changes in space-use requirements and shifts between range-residency and nomadism.
-The methods presented here can be applied to assess both current and future quality of protected areas while also estimating the amount of space organisms currently need and will need in the future under different climate change scenarios.
+Recent advances in computational power have greatly increased analysts' ability to fit computationally demanding models [@nathan_big-data_2022; @wood_generalized_2017-1] that allow biologists to move beyond only considering changes in mean conditions.
+By accounting for changes in stochasticity, we can start developing a more comprehensive understanding of how organisms adapt to the dynamic environments organisms live in, including recent changes in climate [@intergovernmental_panel_on_climate_change_climate_2023] and increases in the frequenct and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017; @grant_evolution_2017; @rypkema_modeling_2021; @yao_emergence_2022].
 
@@
 Additionally, we thank all those who provided feedback on all posters, presentation, and writings related to this project.
-In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely: Aimee Chhen, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr.
+In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely, in alphabetical order by first name: Aimee Chhen, Jessa Marley, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr.
 Simon Wood.

### 2024-06-06 writing/appendix-A-concepts-and-definitions.Rmd
cleaned up appendix A

@@
 
-In this section, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$, which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
-Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations.
+In our simulations, we simulate $R$ using a Gamma distribution with time-dependent (but spatially homogeneous) mean $\mu(t)$ and variance $\sigma^2(t)$ (see the sections below), which we write as $R \sim \Gamma\big(\mu(t), \sigma^2(t)\big)$.
+Although Gamma distributions are more often parameterized using parameters shape and scale ($k > 0$ and $\theta > 0$) or shape and rate ($\alpha > 0$ and $\beta = 1/\theta > 0$), we use $\e(R) = k \theta$ and $\var(R) = k \theta^2$ to facilitate visualizing the simulations within the context of resource abundance and stochasticity.
 However, note that $\e(R)$ and $\var(R)$ are not independent because the variance depends strongly on the mean (and vice-versa).
@@
 There are many distributions we can assign to $R$ depending on how we quantify it.
-For instance, if we measure $R$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
-However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the section below on applying this framework).
-Alternatively, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution.
+For instance, if $R$ is the number of calories an animal is able to access from food in a given location, we can let $R$ follow a distribution with support positive real numbers, such as a Gamma or log-normal distribution.
+Alternatively, if we measure $R$ using the Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], we should use a distribution with support over the interval $[-1, 1]$, since NDVI can only take on values between -1 and 1 (extremes included).
+However, there is no commonly used distribution with that support, so the best option is to rescale NDVI to (0, 1) and use a beta distribution (see the main text and Appendix C).
 If $R$ is a discrete integer variable, such as the number of prey in a location during a period of time, we can use a Poisson or negative binomial distribution.
@@
 
-\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what value we believe $R$ will take, on average.
-The expectation of a random variable is the value one can expect the random variable to take, on average, in the long term.
+\noindent Since the exact value of $R$ at a given time and location is unknown, comparing the magnitude in $R$ between two locations or time periods requires a quantitative measure of what values we believe $R$ will take, on average, in each of the two locations or time periods.
+The expectation of a random variable is the value one can expect the random variable to take, on average, over the long term.
 Here we use $\e(R)$ to indicate the expectation of the random variable $R$.
-When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that $\e(R)$ changes over time by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$ (Fig.
+When the mean changes over time, such as in strongly seasonal regions, we explicitly indicate that by writing $\e(R)$ as a function of time, $t$: $\e(R|t) = \mu(t)$ (Fig.
 A\@ref(fig:multipanel-constant-var)).
-We indicate the estimated average amount of resources over time by adding a caret on the symbol: $\hat \mu(t)$.
-Note, however, that the estimate of $\mu(t)$ and what values it can take depend on what distribution we assume $R$ follows.
-For a Gamma distribution, $\mu(t)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t)$ is necessarily within the interval $(-1, 1)$.
-See the section on modeling $R$ in the main manuscript for more information on how to choose probability distributions.
+If $\e(R)$ changes over time and space, we write $\e(R|t, u) = \mu(t, u)$.
+We indicate the estimated $\mu(t, u)$ by adding a caret on the symbol: $\hat \mu(t, u)$.
+Note, however, that the estimate of $\mu(t, u)$ and what values it can take depend on what distribution we assume $R$ follows.
+For a Gamma distribution, $\mu(t, u)$ can take any positive value, but if we use NDVI as a proxy for $R$, then $\mu(t, u)$ is necessarily within the interval $(-1, 1)$.
 
-```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance (after accounting for changes in the mean).
-(A) Resources ($R$, orange fill) vary over time and space with a constant the variance (dot color).
-(B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
+```{r multipanel-constant-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with constant variance around the mean.
+(A) Resources ($R$, orange fill) vary over time and space with a time-varying but spatially homogeneous mean ($\\mu(t)$) and constant variance (dot color).
+(B) $R$ follows a Gamma distribution with mean $\\mu(t)$ and constant variance $\\var(R) = \\sigma^2$.
 The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}
@@
 
-\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$.
-A random variable's variance is a measure of its unpredictability (i.e., variability).
-We use the notation $\var(R)$ to indicate the variance in $R$ after accounting for changes in $\mu(t)$, and we use $\sigma^2(t)$ to indicate its function over time (with estimate $\widehat{\sigma^2}(t)$; see Fig.
+\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$, which we quantify with its variance.
+We use the notation $\var(R)$ to indicate the variance in $R$ around the mean, i.e., after accounting for changes in $\mu(t, u)$, and we use $\sigma^2(t, u)$ to indicate its function over time and time (with estimate $\widehat{\sigma^2}(t, u)$; see Fig.
 A\@ref(fig:multipanel-changing-var)).
-<!-- Consequently, while $R$ may vary within years, we define $\var(R)$ as the variation in $R$ after accounting for the fact that $R$ is (predictably) lower during some seasons and higher in others due to changes in $\sigma^2(t)$.
-Additionally, we distinguish between two different sources of variation.
-We refer to known or expected spatiotemporal differences or changes as heterogeneity, which we consider as a change in $\mu(t)$.
-In contrast, we use the term stochasticity for unpredictable variation (i.e., due to $\sigma^2(t)$).--> For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons.
-Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t)$ to be lower in the winter since it does not expect to find berries (i.e., $\mu(t) \approx 0$).
-Also note that, like with $\mu(t)$, the estimate of $\sigma^2(t)$ and what values it can take depend on what distribution we assume $R$ follows.
-Figure A\@ref(fig:ex) shows four examples of Gamma distributions with different means and variances.
+For example, the time at which berry bushes produce fruit may seem highly stochastic (i.e., unpredictable) to a young bear, but it becomes predictable as it learns to understand and predict how the mean amount of berry changes with the seasons.
+Still, whether next year will be a good or bad year for berries remains stochastic because it depends on factors the bear cannot predict (e.g., droughts, fires), but the bear can still expect $\sigma^2(t, u)$ to be lower in the winter since berries will be hard to find (i.e., $\mu(t, u) \approx 0$).
+Also note that, like with $\mu(t, u)$, what values $\sigma^2(t, u)$ can take depends on what distribution we assume $R$ follows.
+Figures A\@ref(fig:ex-gamma) and A\@ref(fig:ex-beta) show four examples of Gamma and Beta distributions, respectively, with different means and variances.
 
-```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance (even after accounting for changes in the mean).
-(A) Although resources ($R$, orange fill) varies over time and space, the variance (dot color) is lowest at the beginning and end of the observational period and highest when $\\e(R)$ peaks.
-(B) Arbitrary definition of $R$ as following a Gamma distribution with time-varying mean $\\mu(t)$ and variance $\\sigma^2(t)$.
+```{r multipanel-changing-var, fig.cap="Fictitious example of variation in resources in a heterogeneous environment with changing variance around the mean.
+(A) Although resources ($R$, orange fill) vary over time and space, the variance (dot color) is lowest at the beginning and end of the observational period, and it is highest when $\\mu(t)$ peaks.
+(B) $R$ follows a Gamma distribution with time-varying but spatially homogenous mean $\\mu(t)$ and variance $\\sigma^2(t)$.
 The points in the central panel indicate the ralizations of $R$ in panel a.", fig.pos='!h'}

### 2024-06-07 writing/appendix-B-simulations.Rmd
cleaned up appendices B and C

@@
       - '--lua-filter=author-info-blocks.lua' # for author affiliations
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-This appendix illustrates all the steps necessary to produce the simulation figures in the main manuscript (figs.
+This appendix illustrates all the steps necessary to produce the simulation used in the main manuscript (figs.
 3 and 4).
-To achieve full transparency while minimizing computational times, the code illustrated in this pdf is not executed during the knitting of the document.
+To achieve full transparency while minimizing computational times, the code illustrated in this pdf was not executed during the knitting of the document.
 Instead, the R Markdown document (`writing/appendix-2-simulations.Rmd`) contains code chunks that import the `RDS` files saved by the scripts used during the analysis via `R` code that is not printed in the pdf file.
-Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes and as a general guide.
+Although one can replicate the analyses by running the code in this pdf, we suggest only using this document for illustrative purposes.
 We suggest using the `R` scripts to replicate the simulations, instead.
@@
 Most intermediate and diagnostic checks are not included in this document for the sake of brevity and simplicity, but their outputs and conclusions are listed in this section.
-In this script, we use the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] for movement modeling; the `terra` package [version 1.7-39, @hijmans_terra_2023] to work with the simulated resource rasters; the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling; and the the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
+In this appendix, we use the `ctmm` package [version 1.2.0, @fleming_ctmm_2021] for movement modeling; the `terra` package [version 1.7-71, @hijmans_terra_2024] to work with the rasters of simulated resources; the `dplyr` [version 1.1.4, @wickham_dplyr_2023], `purrr` [version 1.0.2, @wickham_purrr_2023], and `tidyr` [version 1.3.1, @wickham_tidyr_2024] packages for data wrangling; and the the `ggplot2` [version 3.5.1, @wickham_ggplot2_2016], `cowplot` [version 1.1.3, @wilke_cowplot_2024], and `gratia` [version 0.9.0, @simpson_gratia_2024] packages for plotting.
 
-```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how animals' home-range sizes were simulated.
-(a.) Tracks were simulated using an Integrated Ornstein-Uhlenbeck model (IOU model, an infinitely diffusive and continous-velocity movement model), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
-(b.) Each time the track crossed into a new cell (green dots), the animal collected a random amount of resources that followed a Gamma distribution with common mean $\\mu (t)$ and variance $\\sigma^2 (t)$.
-(c.) Each time the animal collected more resources, its satiety (purple) increased.
-Once the animal collected sufficient resources, the animal stopped exploring (i.e., the track was truncated) and was allowed to return to $\\langle0, 0 \\rangle$ over the same amount of time needed to reach satiety.
-(D) The process was repeated 200 times (13 tracks pictured in this panel).
-(E) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range estimates using Autocorrelated Kernel Density Estimates.", out.width='100%', fig.pos='!h', cache=FALSE}
+```{r sims-example, echo=FALSE, eval=TRUE, fig.cap="Overview of how organisms' home-range sizes were simulated.
+(a.) A track was simulated using an infinitely diffusive and continous-velocity movement model (Integrated Ornstein-Uhlenbeck, IOU), starting from the point $\\langle 0, 0 \\rangle$ (black and yellow square).
+(b.) Each time the track crossed into a new cell (green dots), the organism collected a random amount of resources that followed a Gamma distribution with mean $\\mu (t)$ and variance $\\sigma^2 (t)$, with $t$ being fixed for the entirety of the track.
+(c.) Each time the organism collected more resources, its satiety (purple) increased.
+Once the organism collected sufficient resources, the organism stopped exploring (i.e., the track was truncated) and was allowed to return to $\\langle0, 0 \\rangle$ over the same amount of time needed to reach satiety.
+(D) The process was repeated 200 times (13 tracks pictured in this panel) with the same value for $t$.
+(E) The final set of (truncated) tracks was then modeled using Ornstein-Uhlenbeck Foraging models to estimate the 95\\% home range size (see section 4).", out.width='100%', fig.pos='!h', cache=FALSE}
 knitr::include_graphics('../figures/simulation-example.png', rel_path = FALSE)
@@
 ```{r simulate-tracks, warning=FALSE}
-# NOTE: change working directory to "hr-resource-stoch" or modify paths
-
+# NOTE: assuming the working directory is "hr-resource-stoch"
 # attach all necessary packages
@@
 library('cowplot') # for fancy multi-panel plots
-library('dagitty') # for directed acyclical graphs
 library('ggdag')   # for directed acyclical graphs
@@
 source('functions/get_hr.R') # for extracting gaussian home range
-source('functions/label_visits.R') # decides when animal encounters food
+source('functions/label_visits.R') # decides when organism encounters food
 source('analysis/figures/default-figure-styling.R') # for color palette
@@
 # source('functions/get_hr.R') # for extracting gaussian home range
-# source('functions/label_visits.R') # decides when animal encounters food
+# source('functions/label_visits.R') # decides when organism encounters food
 source('../analysis/figures/default-figure-styling.R') # for color palette
@@
 ```{r warning=FALSE}
-# find patch visits and calories consumed from the tracks
+# find patch visits
 tracks <- transmute(tels, # drop tel column
@@
 tracks <- tidyr::unnest(tracks, track) # make a single, large tibble
-tracks
+head(tracks)
 ```
@@
 ```{r, eval=TRUE, echo=FALSE}
-readRDS('H:/GitHub/hr-resource-stoch/simulations/labelled-tracks.rds')
+head(readRDS('H:/GitHub/hr-resource-stoch/simulations/labelled-tracks.rds'))
 ```
+
+\newpage
 
@@
 
-Script: `analysis/figures/return-sensitivity.R`
+Script: `analysis/simulations/sensitivity-analyses/1a-return-sensitivity.R`
 
-Adding return trips to $\langle0, 0\rangle$ after an animal reached satiety doubled computational times without appreciable improvements on the home range estimates (including the 95% confidence intervals of the estimates).
+Adding return trips to $\langle0, 0\rangle$ after an organism reached satiety doubled computational times without appreciable improvements on the Gaussian home range estimates (including their 95% confidence intervals).
 
@@
 
-Script: `analysis/figures/delta-t-sensitivity.R`
+Script: `analysis/simulations/sensitivity-analyses/1b-delta-t-sensitivity.R`
 
@@
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)A.** The amount of time between encounters ranged from 1 second (the minimum sampling interval) to 24 minutes and 10 seconds.
+**Exploratory plot B\@ref(fig:dt-sensitivity)A.** The amount of time between encounters ranged from 1 second (the minimum $\Delta t$) to 24 minutes and 10 seconds.
 Approximately 93% of the encounters (500/536, excluding the first 3 events of each track) occurred with 30 or more seconds between events.
@@
 
-**Exploratory plot B\@ref(fig:dt-sensitivity)C.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an animal was near cell boundaries.
+**Exploratory plot B\@ref(fig:dt-sensitivity)C.** A sampling interval of $\Delta t = 30$ seconds was small enough to capture fine-scale movement in the tracks but large enough to avoid excessive amounts of data and an inflated amount of encounters when an organism was near cell boundaries.
 
@@
 The colored lines indicate the estimated relationship based on a Generalized Additive Model fit using the $\\tt{geom\\_smooth}$ function from the $\\tt{ggplot2}$ package.
-Although the number of encounters detected decreases as sampling interval doubles, the loss at $\\Delta = 30$s is negligible.
+Although the number of encounters detected decreases as sampling interval doubles, the loss at $\\Delta = 30$s is negligible (note that all sampling times included > 60\\% of the encounters).
 (C) Beginning of the track generated with seed \"1\" (purple line) for different sampling intervals.
-Red dots indicate locations where the animal remained in the same cell, while the blue squares indicate when an animal was in a new cell and thus encountered food.
-While the number of encounters detected decreases as the sampling interval increases, most of the encounters lost at $\\Delta t = 30$s occured because the animal remained almost adjacent to the borders between cells.
-Additionally, the track at $\\Delta t = 30$s is still sufficiently tortuous to represent realistic animal movement.
-(D) The three tracks used in these tests over the raster used for determining when the animals encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
+Red dots indicate locations where the organism remained in the same cell, while the blue squares indicate when an organism was in a new cell and thus encountered food.
+While the number of encounters detected decreases as the sampling interval increases, most of the encounters lost at $\\Delta t = 30$s occured because the organism remained almost adjacent to the borders between cells.
+Additionally, the track at $\\Delta t = 30$s is still sufficiently tortuous to represent realistic movement.
+(D) The three tracks used in these tests over the raster used for determining when the organism encountered food.", out.width='100%', echo=FALSE, eval=TRUE, cache=FALSE}
 knitr::include_graphics('../figures/thinning-examples.png', rel_path = FALSE)
@@
 
-As the organism searched for resources, it encountered $R$ resources sampled from a family of Gamma distributions.
-The figure below shows the range in Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\,500, 126\,250, 250\,000$).
+As the organism moved, it encountered $R$ resources sampled from a family of Gamma distributions with mean $\mu(t)$ and variance $\sigma^2(t)$.
+The figure below shows the range in Gamma distributions for the minimum, median, and maximum $\mu(t)$ and $\sigma^2(t)$.
 
@@
 ```{r, eval=TRUE, echo=FALSE, fig.cap="Gamma distributions for the minimum, median, and maximum mean (100, 150, 200) and variance ($2\\,500, 126\\,250, 250\\,000$) with the corresponding coefficients of variation.
-The median and mode of the distributions increase with the mean but decrease with the variance.
+The medians and modes of the distributions increase with the mean but decrease with the variance.
 Greater coefficients of variation result in wider distributions with greater density near zero."}
@@
 The end of this script generates figures that show the mean time spent searching for resources before reaching satiety.
-This appendix includes the two figures that show the absolute and relative differences in time required to reach satiety as a function of the number of tracks.
-The estimated times for each number of replicates can be found in the `figures/5-by-5-sensitivity-analysis/` folder.
+This appendix includes the two figures that show the change in time required to reach satiety as a function of the number of tracks.
+The estimated times for each number of replicates can be found in the `figures/5-by-5-sensitivity-analysis` folder.
 
@@
   geom_hline(yintercept = 0, color = 'red') +
-  labs(x = 'Time',
-       y = 'Difference in average daily exploration time (hours)')
+  labs(x = 'Time', y = 'Difference in average daily time required (hours)')
 ```
+
+\clearpage
 
@@
   labs(x = 'Time',
-       y = expression(paste('Relative difference from estimates with',
-                            '1024 replicates'~(log[2]~scale))))
+       y = expression(bold(paste('Relative difference in time to satiety'~
+                                   (log[2]~scale)))))
 ```
@@
 
-Script: `analysis/hr-simulation-extreme-scenarios.R`
+Script: `analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R`
 
 In this script, we check how many tracks are necessary to produce stable and accurate home-range size estimates.
-We do so by estimating the home ranges of animals for varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
+We do so by estimating the home-range sizes from varying numbers of track in the best-case scenario (highest $\e(R)$ and lowest $\var(R)$) and worst-case scenario (lowest $\e(R)$ and highest $\var(R)$).
 
@@
   facet_wrap(~ case, nrow = 1) +
-  geom_vline(xintercept = 100, color = 'darkorange') +
+  geom_vline(xintercept = 200, color = 'darkorange') +
   geom_smooth(method = 'gam', color = 'black',
@@
 
-```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated home-range sizes as a function of the number of days sampled for an animal in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an animal with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
-In both cases, 100 days are sufficient to produce stable estimates of home-range size.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
+```{r, eval=TRUE, echo=FALSE, fig.cap="Estimated home-range sizes as a function of the number of days sampled for an organism in a habitat with the highest $\\e(R)$ and lowest $\\var(R)$ (left) and an organism with the lowest $\\e(R)$ and highest $\\var(R)$ (right).
+In both cases, 200 days are sufficient to produce stable estimates of home-range size.", cache=FALSE, out.width='75%', fig.height=8, fig.width=12}
 knitr::include_graphics('../figures/hr-over-days.png', rel_path = FALSE)
@@
 
-1. $\tt{analysis/simulations/hr-mean-variance-simulations-days.R}$
-2. $\tt{analysis/simulations/hr-mean-variance-simulations-days-summarized.R}$
-3. $\tt{analysis/simulations/hr-mean-variance-simulations-modeling.R}$
-4. $\tt{analysis/simulations/hr-mean-variance-simulations-hrs.R}$
-6. $\tt{analysis/simulations/modeling-R-and-hr.R}$
+1. `analysis/simulations/hr-mean-variance-simulations-days.R`
+2. `analysis/simulations/hr-mean-variance-simulations-days-summarized.R`
+3. `analysis/simulations/hr-mean-variance-simulations-modeling.R`
+4. `analysis/simulations/hr-mean-variance-simulations-hrs.R`
+6. `analysis/simulations/modeling-R-and-hr.R`
 
@@
 Although the model was fit to both the 95% and core (50%) home range estimates, the results were presented only for the 95% quantile for simplicity.
-The home range was caulculated using a Gaussian approximation using the simulated organism's positional variance:
+The home range was caulculated using a Gaussian approximation and the simulated organism's positional variance via the formula:
 
@@
 
-$$\widehat{h_{0.95}} / \widehat{h_{0.5}} = \frac{-2 \log(1 - 0.95) \pi \sigma^2_{pos}}{-2 \log(1 - 0.5) \pi \sigma^2_{pos}} = \frac{\log(1 - 0.95)}{\log(1 - 0.5)} = \frac{\log(0.05)}{\log(0.5)} = \log_{0.5}(0.05) \approx 4.322,$$
+$$\widehat{h_{0.95}} / \widehat{h_{0.5}} = \frac{-2 \log(1 - 0.95) \pi \sigma^2_{pos}}{-2 \log(1 - 0.5) \pi \sigma^2_{pos}} = \frac{\log(1 - 0.95)}{\log(1 - 0.5)} = \frac{\log(0.05)}{\log(0.5)} \approx 4.322,$$
 
-and therefore $\e(R)$ and $\var(R)$ would have similar effects (but about 4.322 times weaker) on the core home range than they do on the 95% home range.
+and therefore $\e(R)$ and $\var(R)$ would have similar effects on the core home range as on the 95% home range (but about 4.322 times weaker).
 Note that we used this approximation because, unlike real organisms, the simulated organism did not use the available space selectively (e.g., spending more time near cell boundaries), so the Gaussian approximation is appropriate.
-When estimating home-range sizes based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C on empirical modeling and @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
+When estimating home-range sizes based on empirical data, we suggest estimating utilization distributions using methods such as Autocorrelated Kernel Density Estimation [see Appendix C and: @noonan_comprehensive_2019; @alston_mitigating_2022; @silva_autocorrelationinformed_2022].
 
@@
   mutate(quantile = factor(quantile)) # necessary for GAMs
+sims
 ```
@@
 
-The abundance of data for the average $\e(R)$ and $\var(R)$ did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown).
-To estimate the effects of resource abundance and stochasticity on the simulated home-range sizes, we fit a location-scale GAMs (GAMLS) with a location-scale gamma distribution (since home ranges are strictly positive).
-We hypothesized that the effect of $\e(R)$ would depend on $\var(R)$ (and the effect of $\var(R)$ would depend on $\e(R)$), since we expected organisms to respond to changes in $\mu(t)$ less in predictable environments than in stochastic ones (fig.
+The abundance of data for the average $\mu(t)$ and $\sigma^2(t)$ (see figures above) did not create biases in the models, since removing such values had no appreciable effect on the model estimates (not shown).
+We estimated the effects of $\e(R)$ and $\var(R)$ on the simulated home-range sizes using a location-scale GAMs (GAMLS) with a gamma family of distributions (since home-range sizes are strictly positive).
+We hypothesized that the effect of $\var(R)$ would depend on $\e(R)$, since we expected organisms to respond to changes in $\sigma^2(t)$ less when resources were abundant than when they were scarce (fig.
 B\@ref(fig:dag)).
-Within a Bayesian framework, DAGs can be used to infer causality within statistical models [as the latter alone does not allow one to infer causality, see @mcelreath_statistical_2016].
-The DAG we present in this appendix can be interpreted as follows: $\e(R)$ and $\var(R)$ are the controlled variables that determine both the amount of resources in a given encounter with resources, $R$, and, consequently, an organism's home-range sizes, $H$.
+
+Within a Bayesian framework, Directed Acyclical Graphs (DAGs) can be used to infer causality [@mcelreath_statistical_2016].
+The DAG below can be interpreted as follows: $\e(R)$ and $\var(R)$ are the (controlled) variables that determine resources, $R$, and, consequently, an organism's home-range size, $H$.
 Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).
 
-```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=6, out.width='50%'}
-dag <- dagitty('dag{
-"E(R)" -> R <- "Var(R)"
-"E(R)" -> "Var(R)"
-R -> H
-}')
-coordinates(dag) <- list(x = c(R = 0, 'E(R)' = -1, 'Var(R)' = 1, H = 0),
-                         y = c(R = 1, 'E(R)' = 0, 'Var(R)' = 0, H = 2))
-ggdag(dag) +
-  theme_dag()
+```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=6, out.width='100%'}
+dagify(H ~ R,
+       R ~ E + V,
+       outcome = 'H',
+       labels = c('H' = 'H', 'R' = 'R', 'E' = 'E(R)', 'V' = 'Var(R)'),
+       coords = list(x = c(R = 0, E = -1, V = 1, H = 0),
+                     y = c(R = 1, E = 2, V = 2, H = 0))) %>%
+  tidy_dagitty() %>%
+  mutate(col = case_when(name == 'E' ~ '1',
+                         name == 'EN' ~ '1',
+                         name == 'V' ~ '2',
+                         name == 'VN' ~ '2',
+                         name == 'H' ~ '3',
+                         TRUE ~ '4')) %>%
+  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_point(aes(color = col), size = 20, alpha = 0.3) +
+  geom_dag_edges() +
+  geom_dag_text(aes(label = label), color = 'black', size = 4.5) +
+  scale_color_manual(values = c(pal[1:3], 'black')) +
+  theme_dag() +
+  theme(legend.position = 'none')
 ```
+
+\clearpage
 
@@
 
-```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scales).
-The dashed lines indicate the 95\\% Bayesian credible intervals.
-The data points are included as rug plots in all marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
+```{r gamls-terms, fig.cap = "Marginal and interaction effects of each of the terms present in the final GAMLS (on the log link scale).
+The shaded ribbons indicate the 95\\% Bayesian credible intervals.
+The data are included as rug plots in the marginal terms and as points in the interaction terms.", eval=TRUE, fig.height=7.5, fig.width=12, out.width='\\linewidth', echo=FALSE}
 draw(m) & theme_bw() + theme(panel.grid = element_blank())
@@
 Fig.
-B\@ref(fig:gamls-terms) shows each of the terms from model 3.
-The estimated effect of `quantile` (see the output of `summary()` above) supports the relationship between the gaussian HR estimates detailed above, since $\exp(1.464) = 4.323218 \approx \log_{0.5}(0.05),$ but there is no differences in the scale parameter between the two quantiles.
-Additionally, the interaction term of $\mu(t)$ and $\sigma^2(t)$ on the mean HR indicates that the effect of $\sigma^2(t)$ is stronger at lower values of $\mu(t)$ (e.g., $\mu(t) = 100$), since the estimated effect is negative for low values of $\sigma^2(t)$ and positive for high values of $\sigma^2(t)$, which accentuates the effect of $\sigma^2(t)$ on the mean home range.
-The opposite is true for high values of $\mu(t)$, where the effect of $\sigma^2(t)$ becomes flatter, and thus weaker.
-Similar but opposite conclusions can be drawn from the interaction term for the scale parameter, since the effect of $\sigma^2(t)$ on the scale parameter is weaker when $\mu(t)$ is low and stronger when $\mu(t)$ is high.
-
-Finally, we can use the model to recreate fig. 4 from the main text and visualize the marginal effects of $\mu(t)$ and $\sigma^2(t)$ on the 95% home range.
-
-```{r, eval=TRUE}
-# effect of mu
-newd_mu <-
-  expand_grid(mu = seq(min(sims$mu), max(sims$mu), length.out = 400),
-              sigma2 = mean(sims$sigma2),
-              quantile = unique(sims$quantile))
-
-preds_mu <- bind_cols( # bind new data and predictions
-  newd_mu,
-  predict(m, newdata = newd_mu, type = 'link', se.fit = TRUE) %>%
-    data.frame() %>% # convert list to data frame
-    # 95% CIs assuming Gaussian credible intervals on the link scale
-    transmute(hr = exp(fit.1),
-              hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
-              hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
-```
-
-\newpage
-
-```{r, eval=TRUE}
-# effect of sigma2
-newd_sigma2 <-
-  expand_grid(mu = mean(sims$mu),
-              sigma2 = seq(min(sims$sigma2), max(sims$sigma2),
-                           length.out = 400),
-              quantile = unique(sims$quantile))
-
-preds_sigma2 <- bind_cols(
-  newd_sigma2,
-  predict(m, newdata = newd_sigma2, type = 'link', se.fit = TRUE) %>%
-    data.frame() %>%
-    transmute(hr = exp(fit.1),
-              hr_lwr = exp(fit.1 - se.fit.1 * 1.96),
-              hr_upr = exp(fit.1 + se.fit.1 * 1.96)))
-
-# final figure (Rmd throws and error with label_bquote)
-e_r <- 'paste(bold("Resource abundance, E("), bolditalic("R"), bold(")"))'
-v_r <-
-  'paste(bold("Resource stochasticity, Var("), bolditalic("R"), bold(")"))'
-
-preds <- bind_rows(mutate(preds_mu, x = e_r) %>%
-                     rename(value = mu) %>%
-                     select(-sigma2),
-                   mutate(preds_sigma2, x = v_r) %>%
-                     rename(value = sigma2) %>%
-                     select(-mu))
-
-sims_l <- sims %>%
-  mutate(e = hr - predict(m, newdata = sims, type = 'response')[, 1]) %>%
-  pivot_longer(c(mu, sigma2)) %>%
-  mutate(x = if_else(name == 'mu', e_r, v_r))
-```
-
-\newpage
-
-```{r, eval=TRUE}
-ggplot() +
-  facet_grid(. ~ x, scales = 'free', switch = 'x',
-             labeller = label_parsed) + # label below x axis
-  geom_point(aes(value, hr), filter(sims_l, quantile == 'hr_95'),
-             alpha = 0.1, color = pal[3]) +
-  # 95% CIs are present but hidden under the estimates
-  geom_ribbon(aes(value, ymin = hr_lwr, ymax = hr_upr, fill = x),
-              filter(preds, quantile == 'hr_95'), alpha = 0.5) +
-  geom_line(aes(value, hr, color = x),
-            filter(preds, quantile == 'hr_95'), linewidth = 2) +
-  scale_color_manual(values = pal) +
-  scale_x_continuous(NULL, breaks = NULL) + 
-  scale_y_continuous(bquote(paste(bold('Home-range size, '),
-                                  bolditalic('H'))), breaks = NULL) +
-  theme_bw() +
-  # to make facet strip look like the x axis
-  theme(legend.position = 'none',
-        strip.background = element_blank(),
-        strip.text = element_text(size = 12))
-```
+B\@ref(fig:gamls-terms) shows each of the terms from the model.
+The estimated effect of `quantile` (see the output of `summary()` above) supports the relationship between the gaussian HR estimates detailed above, since $\exp(1.464) = 4.323 \approx \frac{\log 0.5}{\log 0.05},$ but there is no differences in the scale parameter between the two quantiles.
 

### 2024-06-07 writing/appendix-C-empirical-modeling.Rmd
cleaned up appendices B and C

@@
 bibliography: 'hr-resource-stoch.bib'
-# csl: 'the-american-naturalist.csl'
+csl: 'the-american-naturalist.csl'
 fontsize: 12pt
@@
   bookdown::pdf_document2:
+    latex_engine: xelatex # for unicode characters
     pandoc_args:
@@
     toc: false
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-We start by providing examples and considerations on how one may model $R$ and the effects of $\e(R)$ and $\var(R)$ on organisms' home-range sizes.
-Next, we apply the methods to the tapir's GPS tracking data and use NDVI as a proxy for $R$.
-Finally, we offer suggestions on how this approach can be used to inform conservation-related decisions, including assessing habitat quality and estimating organisms' range sizes under different scenarios.
-
 # Modeling $R$ {#modeling-r}
@@
 its location) and variance (which depends on its scale) while allowing the mean-variance relationship to vary.
-`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit Generalized Linear Models [GLMs, see @zuur_mixed_2009] and Generalized Additive Models [GAMs, see @wood_generalized_2017], including hierarchical and location-scale GLMs and GAMs.
-Currently, the `mgcv` package allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
-
-The Gaussian location-scale family of distributions is very flexible, since the mean and variance parameters are assumed to be independent, and the response can be either positive or negative.
-However, the distribution's flexibility can also result in unreliable estimates for non-Gaussian responses, such as strictly positive data (e.g.
-available biomass), count data (e.g., number of prey), proportions (e.g., percentage of forested habitat), and bounded ratios [e.g., NDVI, see @pettorelli_normalized_2011].
-
-The Gamma location-scale family is best for strictly positive responses, such as areas (including home ranges), elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass<!-- (e.g., kg of carbon per m$^2$)-->, or energetic intake<!-- (e.g., kcal/day)-->.
+`mgcv` [@wood_generalized_2017] is a commonly used package for `R` [@r_core_team_r_2023] that allows one to fit location-scale models with various families of distributions, including Gaussian (i.e., normal), gamma, and Tweedie location-scale families.
+The Gaussian location-scale family is very flexible, since the mean and variance parameters are assumed to be independent, but it is inappropriate for strictly positive data (e.g.
+available biomass) and other bounded data [e.g., proportions and Normalized Difference Vegetation Index, i.e.
+NDVI, see @pettorelli_using_2005; @pettorelli_normalized_2011].
+The Gamma location-scale family is best for strictly positive responses, such as elemental compositions [e.g., carbon to nitrogen ratio, see @rizzuto_forage_2021], total biomass, or energetic intake.
 The Tweedie location-scale family is similar to the Gamma family, but it allows for zero data, so it is appropriate for data with a non-trivial amount of zeros, such as daily precipitation or prey density [but see zero-inflated distributions: @zuur_zero-truncated_2009].
-In this paper, we estimate $R$ by modeling NDVI using `mgcv` and a beta location-scale family (not available in `mgcv` at the time of publication).
-If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which allows full control over all of a family's parameters via a fully Bayesian approach [as opposed to `mgcv`'s Empirical Bayes method -- see @burkner_advanced_2018].
+For this paper, we estimated $R$ by modeling NDVI using `mgcv` and a beta location-scale family.
+While the family is not available in `mgcv` at the time of publication, the code for it is avialable on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch/blob/main/functions/betals.r](https://github.com/QuantitativeEcologyLab/hr-resource-stoch/blob/main/functions/betals.r).
+If one is interested in families of distributions which are not available in `mgcv`, we suggest using the `brms` package [@burkner_brms_2017], which supports fully distributional, Bayesian models [@burkner_advanced_2018].
 
@@
 However, the complexity of the spatiotemporal terms should be chosen carefully, particularly for the mean's terms.
-An excessively wiggly $\hat \mu(t)$ will cause $\sigma^2(t)$ to be under-estimated, while an excessively smooth $\hat \mu(t)$ will cause $\sigma^2(t)$ to be over-estimated.
+An excessively wiggly $\hat \mu(t, u)$ will cause $\sigma^2(t, u)$ to be under-estimated, while an excessively smooth $\hat \mu(t, u)$ will cause $\sigma^2(t, u)$ to be over-estimated.
 Although there is no error-proof system, choosing the complexity of the terms based on the organism's ability to detect change and adapt is a reasonable starting point.
-Additionally, setting the basis dimension (`k`) of the scale terms to be half or less than that of the mean terms and using restricted marginal likelihood [@wood_fast_2011] should provide reasonably accurate results.
-We suggest starting with low values of `k` and adjusting `k` based on the trends in the residuals.
-Note that since $R$ is likely spatiotemporally autocorrelated, it may be easy to overfit the model.
+Additionally, using restricted marginal likelihood [`method = 'REML'`, see @wood_fast_2011] should help constrain the complexity of the smooths.
 @simpson_modelling_2018 provides a useful introduction to GAMs for biological time series.
@@
 
-\noindent Although there is no commonly-used distribution with a support over the full range of NDVI (i.e., $[-1, 1]$), we use beta distribution (support (0, 1)) since the all NDVI values were sufficiently greater than 0 (range: 0.3534 to 0.9475).
-Thus, here we can define $R$ as following a beta distribution with mean and variance that depend on time using the notation $B (\mu(t), \sigma^2(t))$.
+\noindent Since all NDVI values in our dataset were sufficiently greater than 0 (fig.
+C\@ref(fig:NDVI-hist)), we defined $R$ as following a spatiotemporally-varying beta distribution with mean $\mu(t, u)$ and variance $\sigma^2(t, u)$: $R \sim B (\mu(t, u), \sigma^2(t, u))$.
 We use this parameterization here for ease of explanation, but note that beta distributions are generally parameterized using the shape parameters $\alpha$ and $\beta$ such that the mean is
@@
 
-Particular attention should be be given when deciding what distribution to use and how to estimate means and variances in $R$.
-Improper models and simulations of resource abundance can fail to produce robust, sensible, and accurate estimates of $R$.
-The following section addresses these concerns.
+If NDVI values are near or below zero (e.g., in barren or snowy ecosystems), we suggest using the equation
 
-<!--
-## Movement as a stochastic process
-
-### Drivers of movement
-
-\noindent @nathan_movement_2008 suggested we can generalize movement from location $\vec l_t$ to location $\vec l_{t+1}$ using the model
+\singlespace
 
 \begin{equation}
-\vec l_{t+1} = F(\Omega, \Phi, \vec w_t, \vec l_t, \vec \xi_t),
-(\#eq:nathan-movement)
+\nu^* = \frac{\nu + 1}{2},
+(\#eq:ndvi-scaling)
 \end{equation}
 
-\noindent such that an organism's position as time $t+1$ depends on its motion capacity ($\Omega$), navigation capacity ($\Phi$), internal state ($\vec w_t$) and current location ($\vec l_t$), as well as how each of the parameters may change in response to the current environmental factors ($\vec \xi_t$).
-This model is conceptually useful, as it can form the basis for many models, experiments, and hypotheses, but fitting such a model would be very difficult in practice (if not impossible).
-The parameters $\Omega, \Phi, \vec w_t$ and $\vec \xi_t$ cannot be quantified directly, and although it may be possible to monitor proxies for the parameters (e.g., limb length, body size, site fidelity, cortisol levels, reproductive stage), doing so with sufficient frequency and accuracy would not be simple.
-Thus, we can instead focus on organisms' movement and range sizes as a response to overall resource abundance and unpredictability while accepting other drivers of movement as causes of stochastic behaviors.
+\doublespace
 
-### Visualizing movement as a stochastic process
+where $\nu$ is the NDVI values in the [-1, 1] scale and $\nu^*$ is the NDVI values scaled to [0, 1].
+Since the transformation is linear (i.e., it only involves addition and division), estimates of $\e(\nu^*)$ and $\var(\nu^*)$ can be back-transformed to the [-1, 1] scale with no bias, unlike with nonlinear transformations such as $\arcsin \sqrt \nu$ and $\log(\nu+1)$ [@jensen_sur_1906; @denny_fallacy_2017].
 
-\noindent To understand how movement can be viewed as a stochastic process, imagine an organism that leaves its den or nest to collect resources and returns to it once it has collected enough resources for the day.
-If we do not know why the organism is choosing to move in one direction or another, nor do we know what the environment is like, we cannot predict where the organism may move from one time point to another.
-However, we can use the organism's positional data to estimate the amount of time it spends collecting resources before returning "home" (or the time required for the position to "decay" back to the average position).
-This period is often referred to as the range crossing time or the positional autocorrelation parameter, $\tau_p$.
-Similarly, we can define $\tau_v$ to be the amount of time the organism moves with the same direction and speed, or the time required for an organism's velocity to "decay" back to the average velocity, $\e(v)$, estimated by $\widehat{\e(v)} = \bar v$.
-$\tau_v$ is referred to the organism's directional persistence or the velocity autocorrelation parameter.
+```{r NDVI-hist, fig.cap="Histogram of the NDVI values used for the beta location-scale model (after removing the problematic raster for 2017-12-19; see section 5).
+Note that all values are far from zero (range: 0.3534 to 0.9475).", echo=FALSE, eval=TRUE, fig.height=2}
+library('dplyr')
+library('lubridate')
+library('ggplot2')
 
-By modeling movement via stochastic models which account for autocorrelation in position [e.g., Ornstein–Uhlenbeck or OU models, see @uhlenbeck_theory_1930] and autocorrelation in velocity [Ornstein–Uhlenbeck Foraging or OUF models, see @fleming_fine-scale_2014], one is able to account for and leverage autocorrelation in tracking data.
-While autocorrelation is problematic for models that assume locations to be independent, modeling movement as a stochastic, autocorrelated process avoids systematic biases in the estimation of home ranges [@noonan_comprehensive_2019] as well as travelling distances and speeds [@noonan_scale-insensitive_2019].
+anna_ndvi <-
+  readRDS('../data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
+  mutate(dec_date = decimal_date(date))
 
-Accurate speed estimates require the sampling interval, $\Delta t$, to be sufficiently small [$\Delta t < \tau_v$, see @noonan_scale-insensitive_2019], since the data must contain multiple consecutive samples where the organism is moving with a similar speed and direction.
-However, since the time required for an organism to cross its home range is orders of magnitude larger than the time it moves with a consistent behavior (i.e., $\tau_p > \tau_v$), it is still possible to produce good estimates of range crossing time even if there is no information on the organism's speed.
-Similarly, if the data is too sparse to estimate $\tau_p$, one can still obtain good estimates of the organism's home-range sizes and home-range size, provided that the total sampling duration, $T$, is sufficiently long.
-Model selection procedures with criteria such as the Akaike Information Criterion [@akaike_new_1974] can help choose the most appropriate model for the available data [@silva_autocorrelationinformed_2022].
-
-Predicting movement using stochastic models may seem less straightforward and more complex than using deterministic models, but ignoring the exact causes and decisions that determine the movement allows us to focus on how the organism moved and less on why it moved.
-Organisms may temporarily change their behavior in response to many causes, including predators [mule deer: @ciuti_effects_2012; shovelnose sturgeon: @hintz_shovelnose_2013; rodents: @randler_predator_2020; macaques: @liu_predator_2022; but also see beaked whales: @siegal_beaked_2022], competitors [Capuchin monkeys: @torrez-herrera_monkeys_2020], weather [framework: @gibert_crossing_2016; giant anteaters: @giroux_role_2022], and mates or mating periods [feral cats: @zhang_home_2022; yellow mongoose: @cronk_home_2021].
-Accounting for these causes of variation as a sum of stochastic processes allows us to account for the unpredictability in organism movement while focusing on the organisms' movement behaviors rather than the causes.
-
-### Quantifying home-range sizes
-
-\noindent The concept of a home range has a long history in ecological research as an indicator of the space an organism requires to satisfy its essential requirements during a period of time [@burt_territoriality_1943], but home range analysis is not appropriate for organisms which are not range-resident, such as migrating organisms [e.g., @jonzen_rapid_2006; @geremia_migrating_2019; @abrahms_memory_2019] and nomadic organisms [e.g., @morato_space_2016; @nandintsetseg_variability_2019].
-Although here we quantify organism's home-range sizes using 95% home-range estimates, we expect that similar conclusions can be drawn about the positional variance, $\vpos$, of organisms that are not range-resident.
-In the case of organisms that are range resident for short periods of time, we can model their movement and short-term home-range sizes within distinct and discrete periods or using a sliding window approach (see the discussion section).
-Note, however, that although range-resident organisms will generally have smaller positional variance ($\vpos$) than migratory and nomadic organisms, $\vpos$ is not sufficient to distinguish between range residency, migration, and nomadism, since $\vpos$ determines the scale and size of an organism's movement but not its behavior.
-Instead, the organism's $\tau_p$, $\tau_v$, and $\bar v$ can help discriminate between different behaviors.
-Organisms with a $\tau_p$ shorter than a year are likely range-resident, while organisms with a $\tau_p$ on the order of a year are likely to be migratory (as long as their life expectancy is greater than a year).
-In contrast, when an organism's $\tau_p$ is on the order of its expected lifespan, it is most often nomadic, since it does not have an average position it returns to regularly.
--->
-
-<!-- ### Modeling home-range sizes -->
-
-<!-- \noindent Recent improvements in animal tracking technologies (e.g., increasing sampling duration and frequency of GPS units) have allowed scientists to demonstrate that changes in animals movement and behavior may be more common than previously thought [leatherback turtles: @hays_flexible_2006; arctic foxes: @lai_movement_2017; elephants: @foley_severe_2008; bison and mule deer: @geremia_migrating_2019; bald eagles: @wheat_migrate_2017].
-However, most commonly used movement models assume organisms' behaviors to remain consistent during the entirety of the tracking period [but see behavior-switching state-space models, e.g., @michelot_movehmm_2016].
--->
-
-<!-- To quantify an organism's changes in behavior in response to $\e(R)$ and $\var(R)$, we suggest using a sliding window approach.
-The width of the window, $w$, should be a fixed value greater than two or three times the organism's expected range crossing time [@fleming_overcoming_2019].
-If the organism's spatial use is relatively consistent over time, one may determine $w$ using the value of $\tau_p$ estimated using the whole dataset.
-Alternatively, if the organism's behavior changes repeatedly (e.g., uses a small area for a short period of time before moving to a new one), then one may want to test multiple values of $w$ to find a value sufficiently large for an effective sample size of at least 2 or 3 range crossings (i.e., $w \gtrapprox 2.5 \,\tau_p$).
--->
-
-<!-- Modeling movement as a continuous-time stochastic process is particularly useful in this approach, since it accounts for the uncertainty and limited amount of information contained in short periods of (autocorrelated) tracking data by differentiating between the sample size (i.e., the number of locations) and effective sample size [i.e., the number of range crossings, see @noonan_scale-insensitive_2019; @noonan_comprehensive_2019].
-In this paper, we model tracking data using continuous-time movement models via the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023].
--->
+ggplot(anna_ndvi, aes(ndvi)) +
+  coord_cartesian(xlim = c(-1, 1)) +
+  geom_histogram(na.rm = TRUE, color = 'black', fill = 'grey',
+                 binwidth = 0.1, center = 0.05) +
+  labs(x = 'NDVI', y = 'Count') +
+  theme_bw()
+```
 
@@
 
-This section of the appendix illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig.
+This section illustrates the steps necessary to reproduce the tapir movement analysis and the related figure in the manuscript (fig.
 5).
 The tapir data used here is from the work of @medici_movement_2022 and can be found at the GitHub repository located at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-Estimating the effects of resource abundance and unpredictability on the tapir's home-range size requires us to first estimate the changes in the tapir's home-range sizes (section \@ref(movement)) and the changes in resource abundance and variance experienced by the tapir (section \@ref(ndvi)) before we can estimate the relationship between resource dynamics and space use (section \@ref(final-fig)).
-
 To minimize the computational costs of creating this appendix, we load the necessary objects through hidden `R` chunks rather than re-running all the code.
@@
 All the packages and source scripts required to run the analyses in this document are listed in the code chunk below.
-For spatial data, we use the `MODIStsp` package [version 2.0.9, @busetto_modistsp_2016] to download the NDVI rasters, the `terra` package [version 1.7-39, @hijmans_terra_2023] to work with the NDVI rasters, and the `sf` package [version 1.0-8, @pebesma_simple_2018].
-We use the `dplyr` [version 1.0.10, @wickham_dplyr_2022], `purrr` [version 0.3.5, @henry_purrr_2022], and `tidyr` [version 1.2.1, @wickham_tidyr_2022] packages for data wrangling and the `lubridate` package [version 1.8.0, @grolemund_dates_2011] for converting calendar dates to decimal dates.
-Finally, we used the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] and the `mgcv` package [version 1.8-41, @wood_generalized_2017] for modeling and the `ggplot2` [version 3.4.0, @wickham_ggplot2_2016] and `cowplot` [version 1.1.1, @wilke_cowplot_2020] packages for plotting.
+For spatial data, we use the `MODIStsp` package [version 2.1.0, @busetto_modistsp_2016] to download the NDVI rasters, the `terra` package [version 1.7-71, @hijmans_terra_2024] to work with the NDVI rasters, and the `sf` package [version 1.0-16, @pebesma_simple_2018; @pebesma_spatial_2023] to work with simple features (e.g., telemetry data and shapefiles).
+We use the `dplyr` [version 1.1.4, @wickham_dplyr_2023], `purrr` [version 1.0.2, @wickham_purrr_2023], and `tidyr` [version 1.3.1, @wickham_tidyr_2024] packages for data wrangling, and the `lubridate` package [version 1.9.3, @grolemund_dates_2011] for converting calendar dates to decimal dates.
+Finally, we used the `ctmm` package [version 1.1.0, @fleming_ctmm_2021] and the `mgcv` package [version 1.9-1, @wood_generalized_2017] for modeling, and the `ggplot2` [version 3.5.1, @wickham_ggplot2_2016] and `cowplot` [version 1.1.3, @wilke_cowplot_2024] packages for plotting.
+We start by attaching all the packages and custom functions we need for the following sections.
 
-We start by attaching all necessary packages and custom functions we need:
-
-```{r, eval=TRUE, warning=FALSE, echo=FALSE}
-# attach all necessary packages
+```{r, eval=TRUE, warning=FALSE, echo=TRUE}
+# NOTE: assuming the working directory is "hr-resource-stoch/writing"
 library('terra')     # to import and save rasters
@@
 
-\noindent The script `analysis/tapir/tapirs-moving-window.R` estimates the seven-day spatial use of various tapirs from the Brazilian Cerrado.
-Here, we simplified the code so that it only estimates the spatial use of the tapir in the manuscript, Anna:
+\noindent The script `analysis/tapir/tapirs-moving-window.R` estimates the seven-day home-range size of various tapirs from the Brazilian Cerrado.
+Here, we simplified the code so that it only estimates the spatial use of the tapir in the manuscript, Anna, which we chose because of the large sample size and high variation in home-range size.
 
@@
 
-# calculate the 7-day home range estimate
+# calculate the 7-day home-range estimate
 window_hr(
@@
 anna_mw <- readRDS('../models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds')
-anna_mw
+anna_mw %>%
+  select(- t_start, - t_end) %>%
+  relocate(date, .before = 1)
 ```
 
-The `window_hr()` function estimates the tapir's home range using a sliding window approach with a window size of 7 days (using `ctmm`'s `%#%` operator for unit conversions) that starts with the set of days from `r format(as.POSIXct(anna_mw$t_start[1]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[1]), format = '%Y-%m-%d')` (extremes included), then shifts by one day (`dt = 1 %#% 'day'`), repeats the analysis for the next seven-day set (`r format(as.POSIXct(anna_mw$t_start[2]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[2]), format = '%Y-%m-%d')`), and continues doing so until it reaches the last set of days, `r format(as.POSIXct(anna_mw$t_start[nrow(anna_mw)]), format = '%Y-%m-%d')` to `r format(as.POSIXct(anna_mw$t_end[nrow(anna_mw)]), format = '%Y-%m-%d')`.
-For each set of days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019].
+The `window_hr()` function estimates the tapir's home range using a sliding window approach with a 7-day window (`window = 7 %#% 'day'`) and a one-day slide (`dt = 1 %#% 'day'`).
+For each set of 7 days, it fits a positional variogram, a continuous-time movement model [@fleming_ctmm_2021], and a utilization distribution via autocorrelated kernel density estimation [@silva_autocorrelationinformed_2022; @noonan_comprehensive_2019].
 Finally, it saves an exploratory figure (fig.
-C\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home range estimates (with 95% confidence intervals) to the `models` folder.
+C\@ref(fig:mw)) to the `figures` folder and the tibble of times, telemetries, movement models, utilization distributions, and home-range estimates (with 95% confidence intervals) to the `models` folder.
 
 ```{r mw, eval=TRUE, echo=FALSE, fig.cap="Exploratory figure created by the $\\tt{window\\_hr()}$ function.
-Panel $\\mathbf{A}$ shows the tapir's GPS locations, while panel $\\mathbf{B}$ shows the seven-day home range estimates (95\\% utilization quantile) with 95\\% confidence intervals.", out.width='\\textwidth'}
+Panel $\\mathbf{A}$ shows the tapir's GPS locations, while panel $\\mathbf{B}$ shows the seven-day home-range estimates (95\\% utilization quantile) with 95\\% confidence intervals.", out.width='\\textwidth'}
 knitr::include_graphics('../figures/tapirs/CE_31_ANNA-window-7-days-dt-1-days.png')
@@
 
-To estimate the resources in the tapir's habitat, we used satellite-measured Normalized Difference Vegetation Index [NDVI, see @pettorelli_using_2005; @pettorelli_normalized_2011].
-We downloaded the data using the `MODIStsp` `R` package with the following code:
+We estimated the resources in the tapir's habitat using NDVI that we downloaded using the `MODIStsp` package for `R` using the code below.
 
 ```{r}
-anna_ud <- anna$akde[[1]] # extract the tapir's utilization distribution
-
+# find the extent of tapir's range 
 bbox <-
-  SpatialPolygonsDataFrame.UD(anna_ud, # convert to a spatial object
+  SpatialPolygonsDataFrame.UD(anna$akde[[1]], # convert to a spatial object
                               level.UD = 0.9995, # utilization quantile
@@
 ```{r, echo=FALSE, eval=TRUE}
-# import NDVI data
-anna_ndvi <-
-  readRDS('../data/ndvi-rasters/tapir-anna/tapir-anna-data.rds') %>%
-  mutate(dec_date = decimal_date(date))
 anna_ndvi
@@
 
-We removed the raster for 2017-12-19 because the values were unusually low for the region.
+We removed the raster for 2017-12-19 because a large portion of the values were unusually low for the region (fig.
+C\@ref(fig:ndvi-rasters)).
 We hypothesize the change in NDVI was drastic, temporary, and widespread because of a sudden flood.
@@
 
-```{r, echo=TRUE, eval=TRUE, fig.height=8, fig.width=10, out.width='100%', }
+```{r ndvi-rasters, echo=TRUE, eval=TRUE, fig.height=8, fig.width=10, out.width='100%', fig.cap="Subset of the NDVI rasters used to estimate the mean and variance in NDVI experienced by the tapir.
+Notice how many of the values for 2017-12-19 are near zero (brown) but values for the two adjacent rasters are closer to 1 (more green)."}
 anna_ndvi %>%
@@
 
-anna_ndvi <- filter(anna_ndvi, date != '2017-12-19') # remove bad values
+anna_ndvi <- filter(anna_ndvi, date != '2017-12-19') # remove biased values
 ```
 
-Next, we estimate the mean and variance in NDVI using a Generalized Additive Model for location and scale [GAMLS: @stasinopoulos_generalized_2007].
-Ideally, we would model NDVI using a family of distributions that accounts for the fact that NDVI cannot be less than -1 or greater than 1, but no such family is available in the `mgcv` package.
-However, since the NDVI data is far from 0 (see the histogram below), we decided to keep NDVI unscaled.
-This prevents the model from predicting values that correspond to water or snow (which are not expected to occur in the study area).
-Mathematically, this approach would be comparable to predicting in a Bayesian framework with a prior with zero probability for any NDVI values below zero.
-In environments where values below zero are probable, one could use the beta family after applying the linear transformation $$y^* = \frac{y + 1}{2},$$ where $y$ is the original NDVI value (between -1 and 1) and $y^*$ is the NDVI value scaled between 0 and 1.
+\newpage
 
-We fit the beta GAMLS via the `mgcv` package (note `family = betals()` in the code chunk below).
+Next, we estimate the mean and variance in NDVI using a Generalized Additive Model for location and scale [GAMLS: @stasinopoulos_generalized_2007] via the `mgcv` package (`family = betals()` in the code chunk below).
 The `betals` family accepts a list of two predictors: one for the mean parameter, $\mu$, and one for the scale parameter, $\phi$, and it uses logit link functions for both parameters (see fig.
 C\@ref(fig:gamls-terms)).
-The variance of the distribution is a function of both parameters: $\sigma^2 = \mu (1 - \mu) \phi$.
+The variance of the distribution is a function of both parameters:
 
-```{r, echo=FALSE, eval=TRUE, fig.height=2}
-ggplot(anna_ndvi, aes(ndvi)) +
-  geom_histogram(na.rm = TRUE, color = 'black', fill = 'grey') +
-  labs(x = 'NDVI', y = 'Count') +
-  xlim(c(-1, 1)) +
-  theme_bw()
-```
+\singlespace
 
-\clearpage
+\begin{equation}
+\sigma^2 = \mu (1 - \mu) \phi.
+(\#eq:betals-var)
+\end{equation}
 
@@
 
-Note that when fitting location-scale GAMs, one should pay particular attention to the number of knots used for each smooth term.
-While using a penalized maximum likelihood method such as REML (`method = 'REML'`) helps avoid over-fitting the model, it does not prevent over-fitting, so finding the right balance between each of the `k` values is crucial.
-Excessively smooth terms for the mean can inflate the scale term, while excessively wiggly terms for the mean can cause the scale to be under-estimated.
-Ultimately, each of the `k` values should be decided in such a way as to mimic the animal's responsiveness, adaptability, motility, and memory (or ability to predict cycles or events).
-If one is unsure where to start from, keeping the `k` for the scale terms below half the `k` for the mean terms is a good starting point.
-
-\clearpage
+\doublespace
 
@@
 
-\newpage
+\clearpage
 
@@
 
-We start by predicting the mean and variance in NDVI experienced by the tapir in the known positions using the beta GAMLS.
+We start by predicting the mean and variance in NDVI experienced by the tapir at its GPS locations using the beta GAMLS.
 
@@
 
-Next, we can estimate the mean and variance in NDVI for each 7-day period by taking the average for the GPS locations within each period.
+Next, we can estimate the mean and variance in NDVI for each 7-day period using the GPS locations within each period to create the left side of figure 5 from the main manuscript.
 
@@
 
-We now have all we need to create the left side of figure 5 from the main manuscript.
-
 ```{r, eval=TRUE, echo=FALSE}
@@
 
-\clearpage
+```{r, eval=TRUE, echo=FALSE, fig.height=4, fig.width=12, out.width='\\linewidth'}
+# need to set the theme again for some reason
+source('../analysis/figures/default-figure-styling.R')
 
-```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
-# need to set the theme again for some reason
-theme_set(theme_bw() + theme(panel.grid = element_blank()))
-
-# axis labels (unicode characters removed because they knitting errors)
-e_r <- bquote(paste(bold('Resource abundance')))
-v_r <- bquote(paste(bold('Resource stochasticity')))
-hr_lab <- bquote(bold(paste('7-day home-range size (', km^2, ')')))
+# axis labels (sigma unicode is knitted as "s"; cannot fix the issue)
+e_r <- 'Resource abundance, \U03BC(t, u)'
+v_r <- bquote(bold(paste('Resource stochasticity,'~sigma, '\U00B2(t, u)')))
+hr_lab <- '7-day home-range size (km\U00B2)'
 
@@
 
-\clearpage
-
-To create the right side of the figure, we will need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use.
-To do this, we fit a GAM to the the tapir's 7-day home range estimates using the mean and variance in NDVI as predictors.<!-- However, since the two predictors are highly correlated, using a regular GAM causes the effects to be non-monotonic (after strongly restricting `k` to avoid over-fitting; see fig.
-C\@ref(fig:hr-gam)).-->
-
-We can now fit a GAM similar to the one used for the simulated data.
-However, since this dataset is much smaller and autocorrelated, we should be more parsimonious with the number of terms and their complexity.
-A model with a Gamma family of distributions and relatively smooth (i.e., not wiggly) marginal terms of $\mu(t)$ and $\sigma^2(t)$ is a reasonable start.
-This dataset is too small (and autocorrelated) to estimate changes in scale parameters with reasonable accuracy and precision.
+To create the right side of the figure, we need to estimate the effects of $\e(R)$ and $\var(R)$ on the tapir's space use.
+To do this, we fit a GAM to the the tapir's 7-day home-range estimates using the mean and variance in NDVI as predictors.
 As in Appendix B, we provide the causal Directed Acyclical Graph (DAG) in figure C\@ref(fig:dag).
@@
 
-```{r dag, eval=TRUE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
-$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=4, fig.width=6, out.width='50%'}
-dag <- dagitty('dag{
-Z -> R -> "NDVI" <- S <- R
-R -> H <- Z
-}')
-coordinates(dag) <- list(x = c(R = 0, 'NDVI' = 0, Z = 1, S = 1, H = 0),
-                         y = c(R = 0, 'NDVI' = 1, Z = 0, S = 1, H = -1))
-ggdag(dag) + theme_dag()
+```{r dag, eval=TRUE, echo=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
+$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=6, fig.width=7, out.width='\\linewidth'}
+dagify(NDVI ~ S,
+       EN ~ NDVI + E,
+       VN ~ NDVI + V,
+       E ~ R,
+       V ~ R,
+       Z ~ E + V,
+       S ~ R,
+       H ~ E + V + Z,
+       outcome = 'H',
+       labels = c(H = 'H', R = 'R', E = 'E(R)', V = 'Var(R)', S = 'S',
+                  Z = 'Z', NDVI = 'NDVI', EN = 'E(NDVI)', VN = 'Var(NDVI)'),
+       coords = list(x = c(R = 0, NDVI = 0, Z = 0, S = 0, H = 0,
+                           E = -1, V = 1, EN = -1, VN = 1),
+                     y = c(R = 3,
+                           Z = 1.5, H = 0, E = 1.5, V = 1.5,
+                           S = 4.5, NDVI = 6, EN = 4.5, VN = 4.5))) %>%
+  tidy_dagitty() %>%
+  mutate(col = case_when(name == 'E' ~ '1',
+                         name == 'EN' ~ '1',
+                         name == 'V' ~ '2',
+                         name == 'VN' ~ '2',
+                         name == 'H' ~ '3',
+                         TRUE ~ '4')) %>%
+  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_point(aes(color = col), size = 20, alpha = 0.3) +
+  geom_dag_edges() +
+  geom_dag_text(aes(label = label), color = 'black', size = 3) +
+  scale_color_manual(values = c(pal[1:3], 'black')) +
+  theme_dag() +
+  theme(legend.position = 'none')
 ```
@@
 
-```{r hr-gam, eval=TRUE, fig.pos='!h', fig.height=3, fig.cap="Marginal effects of $\\mu(t)$ and $\\sigma^2(t)$ on the tapir's space use (on the log link scale).
+```{r hr-gam, eval=TRUE, fig.pos='!h', fig.width=8, fig.height=8, fig.cap="Effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ on the tapir's space use (on the log link scale).
 The estimated degrees of freedom for each term can be seen in parentheses in the y-axis labels.
-Shaded areas inticate the 95\\% credible intervals.", fig.height=3}
-m <- gam(hr_est_95 ~ s(mu, k = 4) + s(sigma2, k = 4),
+Shaded areas inticate the 95\\% credible intervals."}
+m <- gam(hr_est_95 ~ s(mu, k = 4) + s(sigma2, k = 4) + ti(mu, sigma2, k =3),
          family = Gamma('log'), data = tapir, method = 'REML')
@@
 
-We can now create the predictions from the GAM.
+\clearpage
 
-```{r, eval=TRUE, out.width='\\linewidth'}
-# predict marginal effects for each term
-preds <- tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
-                sigma2 = seq(5e-4, 25e-4, length.out = 250)) %>%
+We can now predict from the GAM to create the right side of the figure.
+
+```{r, eval=TRUE}
+marginal_preds <-
+  tibble(mu = gratia:::seq_min_max(tapir$mu, n = 250),
+         sigma2 = gratia:::seq_min_max(tapir$sigma2, n = 250)) %>%
   bind_cols(
     # predictions for the marginal effect of mu
-    predict(m, newdata = ., exclude = 's(sigma2)',
+    predict(m, newdata = ., terms = c('(Intercept)', 's(mu)'),
             type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
@@
     # predictions for the marginal effect of sigma2
-    predict(m, newdata = ., exclude = 's(mu)',
+    predict(m, newdata = ., terms = c('(Intercept)', 's(sigma2)'),
             type = 'link', se.fit = TRUE, unconditional = TRUE) %>%
@@
                 hr_sigma2_upr = exp(fit + 1.96 * se.fit)))
+full_preds <-
+  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
+                       to = ceiling(max(tapir$mu) * 100) / 100,
+                       length.out = 250),
+              sigma2 = seq(from = 0.5e-3, to = 2.2e-3, length.out = 250))%>%
+  mutate(hr_full_est = predict(m, newdata = ., type = 'response') %>%
+           # to avoid excessively large predictions
+           if_else(. < 20, ., NA_real_))
 ```
 
-```{r, eval=TRUE, fig.height=4, fig.width=12, out.width='\\linewidth'}
-theme_set(theme_bw() + theme(panel.grid = element_blank()))
+```{r, eval=TRUE, echo=FALSE, fig.height=4, fig.width=12, out.width='\\linewidth', warning=FALSE}
+source('../analysis/figures/default-figure-styling.R')
 
@@
   geom_point(aes(mu, hr_est_95), tapir, alpha = 0.3, color = pal[3]) +
-  geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), preds,
+  geom_ribbon(aes(mu, ymin = hr_mu_lwr, ymax = hr_mu_upr), marginal_preds,
               fill = pal[1], alpha = 0.3) +
-  geom_line(aes(mu, hr_mu_est), preds, color = pal[1], linewidth = 2) +
-  labs(x = e_r, y = hr_lab)
+  geom_line(aes(mu, hr_mu_est), marginal_preds, color = pal[1], linewidth = 2) +
+  labs(x = 'Resource abundance, \U03BC(t, u)', y = hr_lab)
 
@@
   geom_ribbon(aes(sigma2, ymin = hr_sigma2_lwr, ymax = hr_sigma2_upr),
-              preds, fill = pal[2], alpha = 0.3) +
-  geom_line(aes(sigma2, hr_sigma2_est), preds, color = pal[2],
+              marginal_preds, fill = pal[2], alpha = 0.3) +
+  geom_line(aes(sigma2, hr_sigma2_est), marginal_preds, color = pal[2],
             linewidth = 2) +
@@
 p_f <-
-  expand_grid(mu = seq(from = floor(min(tapir$mu) * 100) / 100,
-                       to = ceiling(max(tapir$mu) * 100) / 100,
-                       length.out = 250),
-              sigma2 = seq(from = 0, to = 0.0035, length.out = 250)) %>%
-  mutate(hr_full_est = predict(m, newdata = ., type = 'response')) %>%
-  ggplot() +
+  ggplot(full_preds) +
   geom_raster(aes(mu, sigma2, fill = hr_full_est)) +
@@
   geom_point(aes(mu, sigma2), tapir, alpha = 0.3, show.legend = FALSE) +
-  scale_x_continuous(e_r, expand = c(0, 0)) +
+  scale_x_continuous('Resource abundance, \U03BC(t, u)', expand = c(0, 0)) +
   scale_y_continuous(v_r, expand = c(0, 0)) +
@@
                                         bold(')')))),
-                      low = 'grey90', high = pal[3], limits = c(0, NA)) +
-  theme(legend.position = c(1, 1),
-        legend.justification = c('right', 'top'),
-        legend.box.background = element_rect(),
-        legend.background = element_rect(),
-        legend.key.width = unit(0.35, 'in')) +
-  guides(fill = guide_colorbar(
-    title.position = 'top',
-    theme = theme(legend.title = element_text(hjust = 1)),
-    direction = 'horizontal'))
+                      low = 'grey90', high = pal[3], limits = c(0, NA))
 

### 2024-06-07 analysis/simulations/sensitivity-analyses/1c-hr-simulation-extreme-scenarios.R

@@ -157,7 +157,7 @@ saturation_days <- readRDS('simulations/hr-saturation-days.rds')
 p_hr_days <-
   ggplot(saturation_days, aes(n_days, hr)) +
   facet_wrap(~ case, nrow = 1) +
-  geom_vline(xintercept = 100, color = 'red') +
+  geom_vline(xintercept = 200, color = 'red') +
   geom_smooth(method = 'gam', color = 'black',
               formula = y ~ s(x, bs = 'cs', k = 10),
               method.args = list(family = Gamma(link = 'log'))) +
@@ -166,8 +166,8 @@ p_hr_days <-
                      trans = 'log2', breaks = c(2, 16, 128, 1024),
                      limits = c(2, 1100)) +
   scale_y_continuous(trans = 'log2',
-                     expression(bold(atop(Estimated~'space-use',
-                                          requirements~(log[2]~scale)))))
+                     expression(bold(Estimated~'home-range size'~
+                                            (log[2]~scale))))
 
 ggsave('figures/hr-over-days.png', plot = p_hr_days, scale = 2,
        width = 6, height = 3, dpi = 'print')

### 2024-06-07 writing/manuscript.Rmd
fixed ylabs of fig 5, ref to 2nd DAG, formula typo

@@
 B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with fig.
-C5 and the empirical case study below).
+C3 and the empirical case study below).
 
@@
 Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$.
-Together, the degree to which $\e(R)$ affects $H$ depends on$ $\var(R)$, and vice-versa, so organisms' responses to resource dynamics can be complex.
+Together, the degree to which $\e(R)$ affects $H$ depends on $\var(R)$, and vice-versa, so organisms' responses to resource dynamics can be complex.
 The simulated and empirical results suggest qualitatively similar marginal effects of $\e(R)$ and $\var(R)$, but there are differences in the estimated interactive effects.
@@
 C3).
-
-***HERE***
 

### 2024-06-07 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R

@@ -13,9 +13,10 @@ theme_set(theme_get() + theme(legend.title = element_text(size = 13)))
 
 # axis lables
 e_r <- bquote(paste(bold('Resource abundance, '), '\U1D6CD', bold('('),
-                    bolditalic('t'), bold(')')))
+                    bolditalic('t'), bold(', '), bolditalic('u'), bold(')')))
 v_r <- bquote(paste(bold('Resource stochasticity, '), '\U1D6D4\U00B2',
-                    bold('('), bolditalic('t'), bold(')')))
+                    bold('('), bolditalic('t'), bold(', '), bolditalic('u'),
+                    bold(')')))
 hr_lab <- bquote(paste(bold('7-day home-range size (km'), '\U00B2',
                        bold(')')))
 
@@ -67,7 +68,7 @@ date_labs <- range(tel$timestamp) %>% as.Date()
 # E(R) and V(R) are highly correlated
 ggplot(tapir) +
   geom_point(aes(mu, sigma2)) +
-  labs(x = '\U1D707(t)', y = '\U1D70E\U00B2(t)') +
+  labs(x = e_r, y = v_r) +
   theme(text = element_text(face = 'plain'))
 
 l_grobs <-

### 2024-06-21 writing/manuscript.Rmd
addressed Chris's comments

@@
 bibliography: 'hr-resource-stoch.bib'
-# csl: 'the-american-naturalist-2024-03.csl'
+csl: 'the-american-naturalist-2024-03.csl'
 # csl: 'nature.csl' # to count refs
@@
 
-\noindent \textbf{References}: 109 (updated on 2024-05-29) <!-- counted by using nature csl -->
+\noindent \textbf{References}: 109 (updated on 2024-06-21) <!-- counted by using nature csl -->
 
@@
 \noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004].
-Thus, we can treat the amount of resources $R$ at a given point in space ($u$) and time ($t$) as a random variable, denoted as $R(u, t)$.
+Thus, we can treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$.
 Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean.
-We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, u)$ and $\sigma^2(t, u)$ to indicate them as functions of time ($t$) and space ($u$).
+We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ to indicate them as functions of time ($t$) and space ($\vec u$).
 Additional definitions and explanations are available in Appendix A.
@@
 A2), disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
-Statistically, this is because the more change one attributes to $\mu(t, u)$ (i.e., the wigglier it is), the smaller $\sigma^2(t, u)$ becomes.
-Conversely, the smoother $\mu(t, u)$ is, the larger $\sigma^2(t, u)$ becomes.
+Statistically, this is because the more change one attributes to $\mu(t, \vec u)$ (i.e., the wigglier it is), the smaller $\sigma^2(t, \vec u)$ becomes.
+Conversely, the smoother $\mu(t, \vec u)$ is, the larger $\sigma^2(t, \vec u)$ becomes.
 Biologically, this is important because an organism's perception scale determines whether it attributes a change in $R$ to a trend in $\e(R)$ or as a stochastic event [i.e., due to $\var(R)$; see @levin_problem_1992].
 An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
-Whether an organism is able to predict trends in $\sigma^2(t, u)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their range size.
+Whether an organism is able to predict trends in $\sigma^2(t, \vec u)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their range size.
 While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
@@
 plot(1:100, qgamma2(p = 0.5, mu = 3, sigma2 = 1:100), type = 'l',
-     xlab = expression(sigma^2~('u,'~t)), ylab = 'Median R')
+     xlab = expression(sigma^2~('t,'~u)), ylab = 'Median R')
 ```
@@
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new and consistent behaviors (e.g, nomadism) or adaptations (e.g., increased fat storage or food caching) to buffer themselves against times of unpredictably low $R$.
-Conversely, if changes in $\sigma^2(t, u)$ are sufficiently predictable, organisms may learn to anticipate and prepare for times of greater stochasticity by pre-preemptively caching food, reducing energetic needs, migrating, or relying on alternative food sources [e.g., @van_baalen_alternative_2001].
+Conversely, if changes in $\sigma^2(t, \vec u)$ are sufficiently predictable, organisms may learn to anticipate and prepare for times of greater stochasticity by pre-preemptively caching food, reducing energetic needs, migrating, or relying on alternative food sources [e.g., @van_baalen_alternative_2001].
 
@@
 \noindent We have provided the case for why both $\e(R)$ and $\var(R)$ should be expected to affect organisms' range size, but we presented the two parameters as independent drivers of movement.
-However, organisms may respond to changes in $\sigma^2(t, u)$ more when resources are scarce than when they are abundant.
+However, organisms may respond to changes in $\sigma^2(t, \vec u)$ more when resources are scarce than when they are abundant.
 Consequently, an organism’s movement behavior is likely to be a function of not only the marginal effects of $\e(R)$ and $\var(R)$ but also their interactive effects.
@@
 Fig.
-\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in $\mu(t, u)$ and $\sigma^2(t,u)$ [telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022].
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ [telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022].
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
@@
 Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018).
-Finally, panels D and E show the estimated (marginal) effects of $\mu(t, u)$ and $\sigma^2(t, u)$ on the tapir's 7-day home-range size.
-Since $\mu(t, u)$ and $\sigma^2(t, u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully.
-To avoid over-fitting the model, we constrained the smooth effects of $\mu(t, u)$ and $\sigma^2(t, u)$ and their interaction effect to a small basis size ($\tt{k = 3}$).
+Finally, panels D and E show the estimated (marginal) effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ on the tapir's 7-day home-range size.
+Since $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully.
+To avoid over-fitting the model, we constrained the smooth effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ and their interaction effect to a small basis size ($\tt{k = 3}$).
 Additional information is provided in appendix C.
 The results presented in panels D-F of fig.
-\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\mu(t, u)$ and increases with $\sigma^2(t, u)$, and the effect of $\mu(t, u)$ depends on $\sigma^2(t, u)$, and vice-versa.
-Alone, $\mu(t, u)$ and $\sigma^2(t, u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
-Additionally, note how high NDVI values (> 0.8) cause $\sigma^2(t, u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
+\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\mu(t, \vec u)$ and increases with $\sigma^2(t, \vec u)$, and the effect of $\mu(t, \vec u)$ depends on $\sigma^2(t, \vec u)$, and vice-versa.
+Alone, $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
+Additionally, note how high NDVI values (> 0.8) cause $\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
 
-```{r tapir-mw, fig.cap="Effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}).
-(A) Trends in resource abundance over time, $\\mu(t, u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
-(B) Varince in resources over time, $\\sigma^2(t, u)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
+```{r tapir-mw, fig.cap="Effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}).
+(A) Trends in resource abundance over time, $\\mu(t, \\vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
+(B) Varince in resources over time, $\\sigma^2(t, \\vec u)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
 (C) Seven-day 95% home range estimated using Autocorrelated Kernel Density Estimation.
-(D, E) Estimated marginal effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t, u)$ and $\\sigma^2(t, u)$ and their interaction effect.
-(F) Estimated home-range size in response to changes in both $\\mu(t, u)$ and $\\sigma^2(t, u)$.
-Note how the effect of $\\sigma^2(t, u)$ is more pronounced when $\\mu(t, u)$ is low.
+(D, E) Estimated marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on home-range size.
+The model accounted for the marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ and their interaction effect.
+(F) Estimated home-range size in response to changes in both $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$.
+Note how the effect of $\\sigma^2(t, \\vec u)$ is more pronounced when $\\mu(t, \\vec u)$ is low.
 See Appendix C for additional information.
@@
 
-\noindent The amount of space organisms is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
+\noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
 This paper presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.

### 2024-07-26 writing/appendix-A-concepts-and-definitions.Rmd
small change

@@
 
-\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$, which we quantify with its variance.
+\noindent In viewing resources as a random variable, we not only obtain a formal framework for describing $\e(R)$, but also the spread around $\e(R)$, which we quantify with its variance (i.e., stochasticity).
 We use the notation $\var(R)$ to indicate the variance in $R$ around the mean, i.e., after accounting for changes in $\mu(t, u)$, and we use $\sigma^2(t, u)$ to indicate its function over time and time (with estimate $\widehat{\sigma^2}(t, u)$; see Fig.

### 2024-07-26 writing/manuscript.Rmd
formatted for preprint

@@
 title: \large How resource abundance and stochasticity affect organisms' range sizes
-# author:
-#   - name: Stefano Mezzini
-#     institute: [braes, biol]
-#   - name: Chris H. Fleming
-#     institute: [ucf, smith]
-#   - name: E. Patrícia Medici
-#     institute: [tapir, iucn, escas]
-#   - name: Michael J. Noonan
-#     email: michael.noonan@ubc.ca
-#     institute: [braes, biol, cmps]
-#     correspondence: "yes"
-# institute:
-#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+author:
+  - name: Stefano Mezzini
+    institute: [braes, biol]
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
+  - name: E. Patrícia Medici
+    institute: [tapir, iucn, escas]
+  - name: Michael J. Noonan
+    email: michael.noonan@ubc.ca
+    institute: [braes, biol, cmps]
+    correspondence: "yes"
+institute:
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist-2024-03.csl'
-# csl: 'nature.csl' # to count refs
+csl: 'movement-ecology.csl'
 fontsize: 12pt
@@
   bookdown::pdf_document2:
-    latex_engine: xelatex # @reed_frequency_2003 has character U+0308
+    latex_engine: xelatex # @reed_frequency_2003 has character \U0308
     pandoc_args:
@@
 
-\bibliographystyle{amnatnat}
-
 <!-- define custom math symbols -->
@@
 ```{r, echo=FALSE}
-# needs to be <= 200 words for AmNat
 abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
@@
 
+\clearpage
+
 \noindent \textbf{Article type}: Major article
@@
 
-\noindent \textbf{References}: 109 (updated on 2024-06-21) <!-- counted by using nature csl -->
+\noindent \textbf{References}: 107 (updated on 2024-07-22)
 
@@
 
-While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976].
+While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008].
 The relationship between organisms' movement and resource abundance has long been of interest to biologists.
@@
 r <- seq(0.5, 5, by = 0.01)
-her <- function(R, C = 1) return(C / R)
-plot(r, her(R = r, C = 5), col = 'white', xlab = 'E(R)', ylab = 'E(H)',
-     ylim = c(0, max(her(R = r, C = 4))))
+har <- function(R, C = 1) return(C / R)
+plot(r, har(R = r, C = 5), col = 'white', xlab = 'E(R)', ylab = 'E(H)',
+     ylim = c(0, max(har(R = r, C = 4))))
 for(i in 1:4) {
-  lines(r, her(R = r, C = c(1, 2, 3, 4)[i]), col = i, lwd = 2)
+  lines(r, har(R = r, C = c(1, 2, 3, 4)[i]), col = i, lwd = 2)
 }
@@
 Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021].
-Thus, there remains a need for a clear hypothesis for the effects of both resource abundance and stochasticity on organisms’ range sizes.
+Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.
 
@@
 We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use.
-To see this, consider,, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance).
-This individual will adopt a different behavior and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity).
-Although these two habitats may have the same long-term average resource abundance, the ephemerality of resources in the stochastic desert environment would promote a different selection of movement and life history strategies.
+To see this, consider, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance).
+This individual will adopt different behaviors and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity).
 In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
@@
 Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
-Results from the simulations and empirical example are presented using a fully transparent approach that allows researchers to replicate the work and apply the methods to their own tracking data.
 
@@
 We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ to indicate them as functions of time ($t$) and space ($\vec u$).
-Additional definitions and explanations are available in Appendix A.
+Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for gamma and beta distributions.
 
@@
 It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, species-level changes in movement behavior are more likely to occur over evolutionary timescales than over an organism's lifespan, since larger ranges require greater vagility, which, in turn, is facilitated by morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
+Still, switches from range residency to nomadism (or vice-versa) will occur over evolutionary timescales rather than over an organism's lifespan (Fig.
+\@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
 
@@
 However, such an assumption is likely not realistic, since $\var(R)$ often differ across space and over time.
-Generally, strictly positive quantities tend to have correlated mean and variances.
-For example, the number of berries in a patch will be low during periods of scarcity (e.g., winter, droughts) due to both a low mean (e.g., 15 berries) and a narrow range of values (e.g., 0 to 50 berries), which implies a small variance.
-In contrast, periods of high productivity (e.g., spring green-up) will have a higher mean (e.g., 500 berries) and wider range of $R$ (e.g., 300 to 700 berries), which implies a higher variance (e.g., due differences in phenology, environmental conditions, and patch depletion).
-This relationship between a random variable's mean and variance is present in many statistical distributions where values are strictly positive, such the Gamma and Poisson distributions.
-See the section on probability distributions in Appendix A for more information.
-
-```{r, echo=FALSE, eval=FALSE}
-layout(t(1:2))
-hist(rgamma(n = 1e4, shape = 5, scale = 3))
-hist(rgamma(n = 1e4, shape = 500/3, scale = 3))
-layout(1)
-```
+Generally, bounded qualities quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., beta).
+See Appendix A for more information.
 
@@
 A2), disentangling changes in $\e(R)$ and $\var(R)$ is not simple [@steixner-kumar_strategies_2020].
-Statistically, this is because the more change one attributes to $\mu(t, \vec u)$ (i.e., the wigglier it is), the smaller $\sigma^2(t, \vec u)$ becomes.
+Statistically, this confound occurs because the more change one attributes to $\mu(t, \vec u)$ (i.e., the wigglier it is), the smaller $\sigma^2(t, \vec u)$ becomes.
 Conversely, the smoother $\mu(t, \vec u)$ is, the larger $\sigma^2(t, \vec u)$ becomes.
-Biologically, this is important because an organism's perception scale determines whether it attributes a change in $R$ to a trend in $\e(R)$ or as a stochastic event [i.e., due to $\var(R)$; see @levin_problem_1992].
+Biologically, it is important because an organism's perception scale determines whether it attributes a change in $R$ to a trend in $\e(R)$ or as a stochastic event [i.e., due to $\var(R)$; see @levin_problem_1992].
 An organism's perception of changes in $R$ will also depend strongly on the its cognitive capacities and memory [@mueller_social_2013; @abrahms_memory_2019; @foley_severe_2008; @falcon-cortes_hierarchical_2021; @fagan_spatial_2013].
 Whether an organism is able to predict trends in $\sigma^2(t, \vec u)$ or not, environmental variability is thought to reduce a landscape's energetic balance [@chevin_adaptation_2010], which, in turn, decreases organisms' fitness [e.g., @berger_climate_2018] and increases their range size.
-While this is true for both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
+While this behavioral response occurs with both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
 A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is currently unknown.
@@
 Thus, we expect $\var(R)$ to have a stronger effect on range size when $\e(R)$ is low, and less of an effect when $\e(R)$ is high.
-We explore this more in the following section.
+We explore this interaction effect more in the following section.
 
@@
 Additional information is provided in Appendix B, including the directed acyclical graph [see fig.
-B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct effects of $\e(R)$ and $\var(R)$ on $H$ (contrast this with fig.
+B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct effects of $\e(R)$ and $\var(R)$ on $H$ (contrast the graph with fig.
 C3 and the empirical case study below).
@@
 Overall, both mean $H$ and the variance around it increase with $\sigma^2(t)$ (most visible with constant $\e(R)$ and linear $\var(R)$).
-This is because, similarly to resource-poor periods, times of greater stochasticity require the organism to move over larger areas for longer periods of time.
-This results in a greater uncertainty in how much time and space the organism will require to reach satiety, or indeed whether an organism living in highly stochastic environments can even reach satiety within a finite amount of time.
+Similarly to resource-poor periods, times of greater stochasticity require the organism to move over larger areas for longer periods of time.
+Additionally, the greater in uncertainty in how much time and space the organism will require to reach satiety, or indeed whether an organism living in highly stochastic environments can even reach satiety within a finite amount of time.
 
@@
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
-The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, theory: @rigby_generalized_2005; @stasinopoulos_generalized_2007; example: @gushulak_impacts_2024] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
+The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale [GAMLS, @wood_smoothing_2016] with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475).
 Panel C shows the changes in the tapir's 7-day home range over time.
@@
 \noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
-This paper presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
+This paper builds on earlier theoretical work [e.g., @harestad_home_1979; @stephens_optimal_1982; @southwood_habitat_1977] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
 We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments.
-
 Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$.
@@
 In the simulated data, $\var(R)$ has little effect when $\e(R)$ is low and a strong effect when $\e(R)$ is high, while the opposite is true for the empirical data.
-This is due to two reasons.
+This difference is due to two reasons.
 Firstly, the shape and symmetry of bounded distributions such as Gamma ($R > 0$) and Beta ($0 < R < 1$) distributions depend on both $\e(R)$ and $\var(R)$ (figs.
@@
 In addition, an organism may alter its movement behavior, physiology, and energetic needs to buffer itself against changes in $\e(R)$ and $\var(R)$ by using space selectively [@johnson_comparison_1980; @rickbeil_plasticity_2019; @riotte-lambert_environmental_2020; @ranc_memory_2022] and adapting their behavior and physiology over time [@southwood_habitat_1977; @levins_evolution_1974].
-Before or during periods of scarcity, organisms may cache resources [@nespolo_why_2022], enter states of dormancy [@goldberg_hibernation_2021; @reher_short_2018], or even pause fetal growth [@schmidt_interplay_2020].
-However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], and avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969].
-Failure to respond optimally may thus force organisms to relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], and may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.
+Before or during periods of scarcity, organisms may cache resources [@nespolo_why_2022], build up fat reserves [@middleton_green-wave_2018], enter states of dormancy [@goldberg_hibernation_2021; @reher_short_2018; @mohr_cellular_2020], or even pause fetal growth [@schmidt_interplay_2020].
+However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969], or a physiology that is not amenable to things like hibernation or fat storage.
+Thus organisms may relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], which may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.
 
@@
 
-\noindent There are two main advantages of using and empirical approach.
-Firstly, using real-world animal movement data can produce scale-appropriate and easily interpretable estimates.
+\noindent There are two main advantages of taking an empirical approach.
+Firstly, modeling real-world animal movement data can produce scale-appropriate and easily interpretable estimates.
 Secondly, empirical models directly quantify the effects of $\e(R)$, $\var(R)$, and confounding variables without having to design complex and time-consuming simulations.
-However, it is not always possible to quantify confounding variables, such as competition.
-While there may be some appropriate proxies, such as density competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (the presence of competitors may not reflect competitive pressure).
-This is problematic if one is interested in the direct causal effect of $\e(R)$ and $\var(R)$, which requires removing any non-negligible confounding effects [@mcelreath_statistical_2016].<!-- However, it may not be an issue if researchers are interested in estimating total causal effects, whether they be direct causal paths or not [@mcelreath_statistical_2016].
-For example, one may choose not to account for competition if $\e(R)$ influences competitor density but competitors do not have a measurable effect on $R$.
-In contrast, if competitor abundance determines $R$ (and not vice-versa), then failing to account for competitors will produce incorrect estimates of the causal effects.-->
+However, it is not always possible to quantify confounding variables.
+For exaple, while there may be some appropriate proxies of competition, such as density of competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (i.e., the presence of competitors may not reflect competitive pressure).
+This is problematic if one is interested in estimating the direct causal effect of $\e(R)$ and $\var(R)$, which requires removing any non-negligible confounding effects [@mcelreath_statistical_2016].
 
@@
 While $R$ and NDVI are correlated for many species [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016].
-This can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig.
+This complexity can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig.
 C3) and satellite-level confounding variables ($S$ in fig.
 C3).
-
 Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020].
@@
 However, such pressures may result in both an expansion of the range [@jetz_scaling_2004; @prox_framework_2020] or a contraction, since larger ranges can be harder to defend and result in higher movement costs [@grant_whether_1993; @jetz_scaling_2004] and encounter rates [@martinez-garcia_how_2020].
-
 Satellite-level confounds include information loss due to coarse spatiotemporal resolution [@huang_commentary_2021; @fan_global_2016], satellite-level error [@huang_commentary_2021; @tian_evaluating_2015; @fan_global_2016], and other limitations of remote sensing (e.g., inability to quantify specific resources or small-scale resource depletion).
@@
 
-<!--
 # Acknowledgements
@@
 All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of the tapir data, which is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
-
-# Author contribution statement
-After acceptance, all multi-author papers must provide an Author Contribution Statement in the final files.
-It appears after the Acknowledgments.
-The goal of the Author Contribution statement is to clearly assign responsibility for parts of the paper.
-This statement should be carefully written to indicate the role that every author played in generating the paper.
-However, it is essential that authors do not get credited for steps in which they were not deeply involved.
-Sample roles are provided below but be sure to assemble a Contribution Statement that fits the specifics of your paper:
-
-* Conceptualization
-* Funding acquisition
-* Methods development/experimental design
-* Data collection  
-* Data analysis
-* Data validation
-* Data visualization
-* Provided resources [including specimens, reagents, equipment]
-* Software development
-* Model analysis
-* Coding simulation
-* Supervision
-* Writing - original draft
-* Writing - review & editing
-
--->
 

### 2024-07-30 analysis/figures/default-figure-styling.R
fixed image sizes

@@ -24,3 +24,8 @@ if(.check_colors) plot_scheme_colorblind(ndvi_pal)
 e_r <- bquote(paste(bold('E('), bolditalic('R'), bold(')')))
 var_r <- bquote(paste(bold('Var('), bolditalic('R'), bold(')')))
 h <- bquote(paste(bold('Home-range size, '), bolditalic('H')))
+
+# figure dimensions
+HALF_WIDTH <- 85
+FULL_WIDTH <- 170
+MAX_HEIGHT <- 225

### 2024-07-30 analysis/figures/home-range-hypotheses.R

@@ -59,9 +59,11 @@ p_s2 <-
 
 # save the plots ----
 ggsave('figures/mean-abundance-hr-hypotheses.png', plot = p_mu,
-       width = 8, height = 8, scale = 1, dpi = 300, bg = 'white')
+       width = HALF_WIDTH, height = HALF_WIDTH, scale = 2.5, units = 'mm',
+       dpi = 600, bg = 'white')
 ggsave('figures/variance-abundance-hr-hypotheses.png', plot = p_s2,
-       width = 8, height = 8, scale = 1, dpi = 300, bg = 'white')
+       width = HALF_WIDTH, height = HALF_WIDTH, scale = 2.5, units = 'mm',
+       dpi = 600, bg = 'white')
 
 # figure the GRS and GRC poster
 p_grid <- plot_grid(p_mu, NULL, p_s2, labels = c('a.', '', 'b.'),

### 2024-07-30 analysis/figures/hr-mean-variance-simulations.R

@@ -51,4 +51,5 @@ p_full
 
 # save the plot as a png
 ggsave('figures/mean-variance-5-by-5-hr-sims.png', p_full,
-       width = 10, height = 5.625, units = 'in', dpi = 'print', bg = 'white')
+       width = FULL_WIDTH, height = 100, units = 'mm', scale = 1.5,
+       dpi = 600, bg = 'white')

### 2024-07-30 analysis/simulations/6-modeling-R-and-hr.R

@@ -220,14 +220,16 @@ cowplot::plot_grid(p_mu_mean, p_sigma2_mean, p_full_mean, labels = 'AUTO',
                    nrow = 1)
 
 ggsave('figures/simulation-regression-mean.png',
-       width = 9, height = 3, dpi = 'print', bg = 'white', scale = 1.5)
+       width = FULL_WIDTH, height = FULL_WIDTH / 3, units = 'mm',
+       dpi = 600, bg = 'white', scale = 2)
 
 # variance
 cowplot::plot_grid(p_mu_var, p_sigma2_var, p_full_var, labels = 'AUTO',
                    nrow = 1)
 
 ggsave('figures/simulation-regression-var.png',
-       width = 9, height = 3, dpi = 'print', bg = 'white', scale = 1.5)
+       width = FULL_WIDTH, height = FULL_WIDTH / 3, units = 'mm',
+       dpi = 600, bg = 'white', scale = 2)
 
 # mean and variance together
 cowplot::plot_grid(p_mu_mean, p_sigma2_mean, p_full_mean,
@@ -235,7 +237,8 @@ cowplot::plot_grid(p_mu_mean, p_sigma2_mean, p_full_mean,
                    labels = 'AUTO', nrow = 2)
 
 ggsave('figures/simulation-regression-mean-and-var.png',
-       width = 9, height = 6, dpi = 'print', bg = 'white', scale = 1.5)
+       width = FULL_WIDTH, height = FULL_WIDTH * 2 / 3, units = 'mm',
+       dpi = 600, bg = 'white', scale = 2)
 
 # effect of SD(R) on E(H) and Var(H) ----
 ggplot() +

### 2024-07-30 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R

@@ -211,5 +211,6 @@ p_right <-
 p <- plot_grid(p_left, p_right, NULL, nrow = 1, rel_widths = c(1, 1, 0.01))
 p
 
-ggsave('figures/tapir-example.png', plot = p, height = 14, width = 14,
-       units = 'in', dpi = 600, bg = 'white')
+ggsave('figures/tapir-example.png', plot = p,
+       width = FULL_WIDTH, height = FULL_WIDTH, units = 'mm',
+       dpi = 600, bg = 'white', scale = 2)

### 2024-07-30 writing/manuscript.Rmd
cleaned up abstract, fixed some typos

@@
 This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
-This is not a trivial oversight.
-Organisms adjust to variable environmental conditions, so failing to consider the effects of resource (un)predictability can result in a limited understanding of an organism's range size, which challenges ecological theory and applied conservation alike.
-In this study, we leverage the available literature to provide a unifying framework and hypotheses for the effect of mean and variance in resources on range sizes.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resource abundance interact to shape predictable patterns in range size.
+Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation.
+In this study, we leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes.
+Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size.
 Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
@@
 
-\noindent \textbf{Article type}: Major article
+\noindent \textbf{Article type}: Research article
 
@@
 
-\noindent \textbf{References}: 107 (updated on 2024-07-22)
+\noindent \textbf{References}: 109
 
@@
 
-\noindent \textbf{Key words:} energetics, environmental stochasticity, home range, range size, `ctmm`, simulations
+\noindent \textbf{Key words:} energetics, resource abundance, environmental stochasticity, home range, range size, `ctmm`
 
@@
 Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings.
-Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021].
+Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024].
 Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.
@@
 While this behavioral response occurs with both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude [@logares_black_2012; @anderson_black-swan_2017].
-A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020], but many of them are limited in geographic and taxonomic scales, so the extent to which these preliminary findings can be generalized is currently unknown.
+A few recent studies support these hypotheses [@chevin_adaptation_2010; @morellet_seasonality_2013; @nandintsetseg_variability_2019; @riotte-lambert_environmental_2020; @broekman_environmental_2024], but many of them are limited in geographic and taxonomic scales or fail to account for nonlinear relationships, so the extent to which these preliminary findings can be generalized is currently unknown.
 Thus, there remains a need for developing a more complete understanding of how organisms' range sizes changes with environmental stochasticity.
@@
 However, it is not always possible to quantify confounding variables.
-For exaple, while there may be some appropriate proxies of competition, such as density of competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (i.e., the presence of competitors may not reflect competitive pressure).
+For example, while there may be some appropriate proxies of competition, such as density of competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (i.e., the presence of competitors may not reflect competitive pressure).
 This is problematic if one is interested in estimating the direct causal effect of $\e(R)$ and $\var(R)$, which requires removing any non-negligible confounding effects [@mcelreath_statistical_2016].
@@
 Recent advances in computational power have greatly increased analysts' ability to fit computationally demanding models [@nathan_big-data_2022; @wood_generalized_2017-1] that allow biologists to move beyond only considering changes in mean conditions.
-By accounting for changes in stochasticity, we can start developing a more comprehensive understanding of how organisms adapt to the dynamic environments organisms live in, including recent changes in climate [@intergovernmental_panel_on_climate_change_climate_2023] and increases in the frequenct and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017; @grant_evolution_2017; @rypkema_modeling_2021; @yao_emergence_2022].
+By accounting for changes in stochasticity, we can start developing a more comprehensive understanding of how organisms adapt to the dynamic environments organisms live in, including recent changes in climate [@intergovernmental_panel_on_climate_change_climate_2023] and increases in the frequency and intensity of extreme events [@logares_black_2012; @anderson_black-swan_2017; @grant_evolution_2017; @rypkema_modeling_2021; @yao_emergence_2022].
 

### 2024-07-31 analysis/tapirus-terrestris/4-moving-window-figure-anna-betals.R
added hats on estimates of E(R) and Var(R)

@@ -12,9 +12,9 @@ source('analysis/figures/default-figure-styling.R') # for theme & palettes
 theme_set(theme_get() + theme(legend.title = element_text(size = 13)))
 
 # axis lables
-e_r <- bquote(paste(bold('Resource abundance, '), '\U1D6CD', bold('('),
+e_r <- bquote(paste(bold('Resource abundance, '), hat('\U1D6CD'), bold('('),
                     bolditalic('t'), bold(', '), bolditalic('u'), bold(')')))
-v_r <- bquote(paste(bold('Resource stochasticity, '), '\U1D6D4\U00B2',
+v_r <- bquote(paste(bold('Resource stochasticity, '), hat('\U1D6D4'), '\U00B2',
                     bold('('), bolditalic('t'), bold(', '), bolditalic('u'),
                     bold(')')))
 hr_lab <- bquote(paste(bold('7-day home-range size (km'), '\U00B2',

### 2024-07-31 writing/manuscript.Rmd
formatted for ME, added abbreviations, fixed typos

@@
 ```{r, echo=FALSE}
-abstract <- "The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
+abstract <- "\\textbf{Background:} The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
 This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
 Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation.
-In this study, we leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes.
-Next, we use simulated movement data to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size.
-Finally, we use real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado to show how this framework can be applied to better understand the movement ecology of free-ranging animals."
+\\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes.
+We then use simulated movement data using continuous-time movement models to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size.
+Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
+\\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity.
+Additionally, the effects of resource stochasticity depend strongly on resource abundance, and organisms' responses to resource abundance and stochasticity can be complex, since the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation.
+\\textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms.
+Failing to account for resource stochasticity can lead to an incomplete and incorrect understanding of how and why organisms move, particularly during periods of rapid change."
+# stringi::stri_count_words(abstract)
 ```
@@
 
-\noindent \textbf{Key words:} energetics, resource abundance, environmental stochasticity, home range, range size, `ctmm`
-
-<!-- Authors' emails:
-Stefano Mezzini: stefano.mezzini@ubc.ca
-Chris H. Fleming: christen.fleming@ucf.edu
-E. Patrícia Medici: medici@ipe.org.br
-Michael J. Noonan: michael.noonan@ubc.ca
--->
+\noindent \textbf{Key words:} energetics, resource abundance, resource stochasticity, environmental stochasticity, home range, range size, movement behavior, `ctmm`
 
@@
 
-# Introduction
+# Background
 
@@
 The relationship between organisms' movement and resource abundance has long been of interest to biologists.
-In his seminal paper, @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
-Three decades after, @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce.
-Two years later, @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is
+In his seminal paper, Burt @burt_territoriality_1943 considered the search for food as the primary driver for movement within an organism's home range.
+Three decades after, Southwood @southwood_habitat_1977 suggested that change in resource abundance drives how organisms decide where to live and when to reproduce.
+Two years later, Harestad and Bunnel @harestad_home_1979 proposed that the simplest relationship between resource abundance and an organism's home-range size is
 
@@
 
-\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
+\noindent where $H$ is the organism's home-range size, $C$ is the organism's resource consumption rate (kcal day$^{-1}$), and $R$ is the resources the organism can access (kcal day$^{-1}$ unit area$^{-1}$).
 Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis.
@@
 When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
-Like @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
+Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
 These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
@@
 Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to $\langle 0, 0\rangle$ independently from the following track (section 2.1 of Appendix B).
-Finally, we fit an OUF movement model [@fleming_fine-scale_2014] to the set of tracks to calculate the 95% Gaussian home-range size using the formula $$H_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,$$ where $\hat \varsigma^2$ is the positional variance estimated by the movement model.
+Finally, we fit an OUF movement model [@fleming_fine-scale_2014] to the set of tracks to calculate the 95% Gaussian home-range size using the formula $$\hat{H}_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,$$ where $\hat \varsigma^2$ is the positional variance estimated by the movement model.
 
@@
 
-```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogenous over space for a given $t$).
+```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogeneous over space for a given $t$).
 Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
@@
 
-\noindent The simulations in the section above support the hypotheses we presented in the introduction, but they are based on assumptions that are often not met in real natural environments.
+\noindent The simulations in the section above support the hypotheses we presented in the background section, but they are based on assumptions that are often not met in real natural environments.
 Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory.
@@
 Fig.
-\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ [telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022].
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in estimated $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ (telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022).
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
@@
 Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018).
-Finally, panels D and E show the estimated (marginal) effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ on the tapir's 7-day home-range size.
-Since $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully.
-To avoid over-fitting the model, we constrained the smooth effects of $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ and their interaction effect to a small basis size ($\tt{k = 3}$).
+Finally, panels D and E show the estimated (marginal) effects of $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ on the tapir's 7-day home-range size.
+Since $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of $R$ on $H$ should be modeled carefully.
+To avoid over-fitting the model, we constrained the smooth effects of $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ and their interaction effect to a small basis size ($\tt{k = 3}$).
 Additional information is provided in appendix C.
 The results presented in panels D-F of fig.
-\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\mu(t, \vec u)$ and increases with $\sigma^2(t, \vec u)$, and the effect of $\mu(t, \vec u)$ depends on $\sigma^2(t, \vec u)$, and vice-versa.
-Alone, $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
-Additionally, note how high NDVI values (> 0.8) cause $\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
+\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\hat\mu(t, \vec u)$ and increases with $\hat\sigma^2(t, \vec u)$, and the effect of $\hat\mu(t, \vec u)$ depends on $\hat\sigma^2(t, \vec u)$, and vice-versa.
+Alone, $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
+Additionally, note how high NDVI values ($\hat \mu(t, \vec u) > 0.8$) cause $\hat\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
 
-```{r tapir-mw, fig.cap="Effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}).
-(A) Trends in resource abundance over time, $\\mu(t, \\vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
-(B) Varince in resources over time, $\\sigma^2(t, \\vec u)$, esimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
+```{r tapir-mw, fig.cap="Effects of estimated $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on the home-range size of a lowland tapir (\\emph{Tapirus terrestris}).
+(A) Trends in resource abundance over time, $\\hat\\mu(t, \\vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period.
+(B) Variance in resources over time, $\\hat\\sigma^2(t, \\vec u)$, estimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period.
 (C) Seven-day 95% home range estimated using Autocorrelated Kernel Density Estimation.
-(D, E) Estimated marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ on home-range size.
-The model accounted for the marginal effects of $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$ and their interaction effect.
-(F) Estimated home-range size in response to changes in both $\\mu(t, \\vec u)$ and $\\sigma^2(t, \\vec u)$.
-Note how the effect of $\\sigma^2(t, \\vec u)$ is more pronounced when $\\mu(t, \\vec u)$ is low.
+(D, E) Estimated marginal effects of $\\hat\\mu(t, \\vec u)$ and $\\hat\\sigma^2(t, \\vec u)$ on home-range size.
+The model accounted for the marginal effects of $\\hat\\mu(t, \\vec u)$, $\\hat\\sigma^2(t, \\vec u)$, and their interaction effect.
+(F) Estimated home-range size in response to changes in both $\\hat\\mu(t, \\vec u)$ and $\\hat\\sigma^2(t, \\vec u)$.
+Note how the effect of $\\hat\\sigma^2(t, \\vec u)$ is more pronounced when $\\hat\\mu(t, \\vec u)$ is low.
 See Appendix C for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} (2022).", out.width='100%'}
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} [27].", out.width='100%'}
 knitr::include_graphics('../figures/tapir-example.png')
@@
 However, organisms may be unable to respond to changes in $\e(R)$ and $\var(R)$ optimally due to various reasons, including limited perceptive range [@steixner-kumar_strategies_2020], lack of experience [@foley_severe_2008; @polansky_elucidating_2015; @falcon-cortes_hierarchical_2021; @merkle_spatial_2019; @fagan_spatial_2013; @abrahms_memory_2019], avoidance of competitors and predators [@kacelnik_ideal_1992; @fretwell_territorial_1969], or a physiology that is not amenable to things like hibernation or fat storage.
-Thus organisms may relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], which may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.
+Thus, organisms may relocate their range to a sub-optimal location [@ciuti_effects_2012; @rich_anthropogenic_2012; @burson_competition_2018; @torrez-herrera_monkeys_2020], which may exacerbate the effects of $\e(R)$ and $\var(R)$ on both mean range size and the variance around it.
 
@@
 
-# Conclusion
+# Conclusions
 
@@
 
-# Acknowledgements
+# List of abbreviations
+
+```{r abbreviations, message=FALSE, warning=FALSE, eval=TRUE, echo=FALSE}
+library(dplyr)
+read.csv('../writing/abbreviations.csv') %>%
+  knitr::kable(format = 'latex', escape = FALSE) %>%
+  kableExtra::row_spec(row = 0, bold = TRUE)
+```
+
+# Declarations
+
+## Ethics approval and consent to participate
+
+\noindent Not applicable.
+
+## Consent for publication
+
+\noindent Not applicable.
+
+## Availability of data and materials
+
+\noindent All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of two simulated datasets that were greater than 100 MB and the tapir data.
+The simulated data can be produced by running the scripts in the repository, while the tapir data is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
+
+## Competing interests
+
+\noindent The authors declare that they have no competing interests.
+
+## Funding
+
+\noindent SM was supported by funding from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, BC Parks Living Labs, and MITACS.
+MJN was supported by NSERC Discovery Grant RGPIN-2021-02758 and the Canadian Foundation for Innovation.
+CHF was supported by NSF IIBR 1915347.
+
+## Authors' contributions
+
+\noindent SM performed the literature review, designed the simulations, analyzed the data, and wrote the manuscript.
+CHF contributed to the analyses.
+EPM provided the tapir telemetry data.
+MJN conceived the project idea and provided support throughout the analyses.
+All authors contributed to the writing and read and approved the final manuscript.
+
+## Acknowledgements
 
@@
 Simon Wood.
-This work was supported by the NSERC Discovery Grant RGPIN-2021-02758 to MJN and funds from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, Mitacs, and BC Parks.
 
-# Code and data availability
+## Authors' information
 
-All code and data used for this manuscript is available on GitHub at [https://github.com/QuantitativeEcologyLab/hr-resource-stoch](https://github.com/QuantitativeEcologyLab/hr-resource-stoch), with the exception of the tapir data, which is available at [https://github.com/StefanoMezzini/tapirs](https://github.com/StefanoMezzini/tapirs).
+\noindent See page 1 for affiliations.
 
-# Conflict of interest
+\noindent Authors' emails:
 
-\noindent The authors declare there are no conflicts of interest.
+\noindent SM: $\text{stefano.mezzini@ubc.ca}$
 
-\newpage
+\noindent CHF: $\text{christen.fleming@ucf.edu}$
+
+\noindent EPM: $\text{medici@ipe.org.br}$
+
+\noindent MJN: $\text{michael.noonan@ubc.ca}$
 

### 2025-02-07 writing/manuscript.Rmd
started responding to comments from reviewers

@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' range sizes
+title: \large How resource abundance and resource stochasticity affect organisms' range sizes
 author:
@@
 output:
+#   bookdown::word_document2:
+#     pandoc_args:
+# # https://github.com/pandoc/lua-filters/blob/master/author-info-blocks/README.md
+#       - '--lua-filter=scholarly-metadata.lua'
+#       - '--lua-filter=author-info-blocks.lua'
+#     toc: false
+#     number_sections: false
   bookdown::pdf_document2:
@@
 ```{r, echo=FALSE}
-abstract <- "\\textbf{Background:} The amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
-This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to the mean amount of resources, while responses to the variance around the mean are still largely unknown.
-Since organisms adjust to variable environmental conditions, failing to consider the effects of resource (un)predictability can result in an insufficient understanding of an organism's range size, which results in challenges for both ecological theory and applied conservation.
-\\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on range sizes.
-We then use simulated movement data using continuous-time movement models to demonstrate how the combined effects of mean and variance in resources interact to shape predictable patterns in range size.
+abstract <- "\\textbf{Background:} From megafauna to amoebas, the amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
+This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to spatiotemporal changes in mean resources, while responses to unpredictable changes in resources (i.e., variance in resources or resource stochasticity) are still largely unknown.
+Since organisms adjust to variable environmental conditions, failing to consider the effects of resource unpredictability can result in an insufficient understanding of an organism's range size.
+\\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on organisms' range sizes.
+We then use simulated movement data to demonstrate how the combined effects of resource abundance and stochasticity interact to shape predictable patterns in range size.
 Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
-\\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity.
-Additionally, the effects of resource stochasticity depend strongly on resource abundance, and organisms' responses to resource abundance and stochasticity can be complex, since the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation.
+\\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity, and the effects of resource stochasticity depend strongly on resource abundance.
+Additionally, the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation.
 \\textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms.
@@
 
-\noindent The amount of resources an organism is able to access is a strong determinant of its odds of survival and reproduction.
+\noindent The amount of resources an organism is able to access is a strong determinant of its fitness.
 Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness [@hou_cold_2020; @le_bot_fishery_2019], altered physiology [@dai_pra_ground_2022; @rocha_life_2021; @le_bot_fishery_2019; @wessling_seasonal_2018], lower chance of reproduction [@stefanescu_timing_2021; @schmidt_interplay_2020; @le_bot_fishery_2019; @douglas_relative_2014], and even death [@foley_severe_2008; @berger_climate_2018].
-Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met.
+Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met [e.g., soil amoebae *Dictyostelium spp.*: @van_haastert_food_2009; plants: @taub_root_1996; and animals: @harestad_home_1979].
 
-While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008].
+While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that motile species can adjust [@kacelnik_ideal_1992; @charnov_optimal_1976; @nathan_movement_2008].
 The relationship between organisms' movement and resource abundance has long been of interest to biologists.
@@
 Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis.
-Many researchers have since demonstrated that organisms adapt their range sizes in response to resources abundance, but results are typically reported as independent, novel findings.
-Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to mean resource abundance, there is little information on how organisms respond to variance around the mean [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024].
-Thus, there remains a need for clear hypotheses of the effects of both resource abundance and stochasticity on organisms’ range sizes.
+Many researchers have since demonstrated that organisms adapt their range sizes in response to resource abundance, but results are typically reported as independent, novel findings.
+Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to changes in mean resource abundance, there is little information on how organisms respond to unpredictable changes in resources [i.e., resource stochasticity, but see: @stephens_optimal_1982; @duncan_lifehistory_2015; @rizzuto_forage_2021; @broekman_environmental_2024].
+Thus, there remains a need for a clear, unifying hypothesis of the effects of both resource abundance and stochasticity on organisms’ range sizes.
 
@@
 To see this, consider, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance).
+The animal may require a large but constant home range size as it moves between patches in search of food.
+If, instead, it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low long-term mean and high stochasticity), it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; $ $ see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
+Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].
+
 This individual will adopt different behaviors and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity).
-In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
-Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @herfindal_prey_2005; @nilsen_can_2005; @rizzuto_forage_2021].
+In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches
 
@@
 
-\noindent Resources are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004].
+\noindent Resources (e.g., food, water, shelter, heat) are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004].
 Thus, we can treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$.
@@
 We indicate the expected value and variance of random variable $R$ using $\e(R)$ and $\var(R)$, respectively, and we use $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ to indicate them as functions of time ($t$) and space ($\vec u$).
-Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for gamma and beta distributions.
+Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for Gamma and Beta distributions.
 
@@
 All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients [@harvey_primate_1981; @baldwin_nutritional_1984; @reich_body_2001].
-As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but they must also account for costs of movement [@taylor_energetics_1982].
+As the abundance of essential resources fluctuates, motile organisms can move to new locations or 'patches' to meet their requirements [@charnov_optimal_1976; @brown_ecology_1999], but movement also increases energetic needs [@taylor_energetics_1982].
 
-Fig.
-\@ref(fig:mu-hyp) illustrates our first of two hypotheses, which is similar to that presented by @harestad_home_1979.
-When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [e.g., @relyea_home_2000; @nilsen_can_2005; @herfindal_prey_2005].
+When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [see Fig.
+\@ref(fig:mu-hyp) as well as: @herfindal_prey_2005; @nilsen_can_2005; @relyea_home_2000].
 Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
@@
 It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, switches from range residency to nomadism (or vice-versa) will occur over evolutionary timescales rather than over an organism's lifespan (Fig.
+Still, large-scale switches from range residency to nomadism (or vice-versa) are more likely to occur over evolutionary timescales rather than over an organism's lifespan (Fig.
 \@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
@@
 
-Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @relyea_home_2000; @rizzuto_forage_2021; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
-This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see: @lucherini_habitat_1996; @herfindal_prey_2005; @nilsen_can_2005; @simcharoen_female_2014; @watson_ferruginous_2020, and contrast them to the estimates based on linear models listed above].
+Overall, the hypothesis that range size decreases with resource abundance, $\e(R)$, is commonly accepted and well supported, but many studies assume a linear relationship [e.g., @rizzuto_forage_2021; @relyea_home_2000; @bista_effect_2022; @bradsworth_using_2022; @mcclintic_effects_2014].
+This is problematic because, conceptually, the relationship between range size and $\e(R)$ must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero [see @herfindal_prey_2005; @nilsen_can_2005; @lucherini_habitat_1996; @watson_ferruginous_2020; @simcharoen_female_2014, and contrast them to the earlier references that assume a linear relationship between $H$ and $R$].
 Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.
@@
 However, such an assumption is likely not realistic, since $\var(R)$ often differ across space and over time.
-Generally, bounded qualities quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., beta).
+Generally, bounded quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., Beta).
+For example, prey abundance in a given area over time may approximately follow a Poisson distribution, which implies that the mean and variance will be approximately equal.
+When prey are scarce (e.g., a mean of 10), the variance will also be low, and when prey are abundant (e.g., a mean of 300) the variance will also be high.
+This occurs because the behavior, fitness, and predator-prey dynamics of ~300 prey are more stochastic than those of 10 prey [@campillo_effect_2012].
+Similarly, in the case of fully bounded random variables, the variance is generally lowest when the mean is near either boundary.
+For example, successful predation events are predictably scarce if the probability of capture is near 0, predictably common if the probability is near 1, and most stochastic if the probability is near 0.5 [i.e., as far as possible from both 0 and 1; see @lee_effects_2011].
 See Appendix A for more information.
@@
 While the estimated marginal effect of $\\var(R)$ on $\\e(H)$ is sublinear (panel B), the effect of $\\var(R)$ is superlinear for high values of $\\e(R)$ (panel C).
-The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals})$.
+The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\\tt{mgcv\\!::\\!gammals}$).
 Credible intervals were calculated using 10,000 samples from the posterior distribution while assuming multivariate Gaussian coefficients.
@@
 \noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
-This paper builds on earlier theoretical work [e.g., @harestad_home_1979; @stephens_optimal_1982; @southwood_habitat_1977] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
+This paper builds on earlier theoretical work [e.g., @southwood_habitat_1977; @harestad_home_1979; @stephens_optimal_1982] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
 We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments.
@@
 Proxies of $R$, such as NDVI [@pettorelli_normalized_2011], which may introduce complexities.
-While $R$ and NDVI are correlated for many species [e.g., @phillips_evaluating_2008; @middleton_green-wave_2018; @seigle-ferrand_systematic_2021; @merkle_large_2016; @geremia_migrating_2019; @merkle_spatial_2019], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016].
+While $R$ and NDVI are correlated for many species [e.g., @middleton_green-wave_2018; @geremia_migrating_2019; @merkle_spatial_2019; @phillips_evaluating_2008; @seigle-ferrand_systematic_2021; @merkle_large_2016], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016].
 This complexity can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig.

### 2025-02-10 writing/manuscript.Rmd
addressed most of reviewers' comments

@@
 The animal may require a large but constant home range size as it moves between patches in search of food.
-If, instead, it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low long-term mean and high stochasticity), it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; $ $ see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
+If, instead, it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low long-term mean and high stochasticity), it may switch between dispersal in search for high-resource patches and short-term range residency within patches [*sensu* @charnov_optimal_1976; see @singh_migration_2012; @wheat_migrate_2017; @teitelbaum_beyond_2019].
 Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances [e.g., @chevin_adaptation_2010], but there is still limited empirical evidence to support this hypothesis [but see: @rizzuto_forage_2021; @herfindal_prey_2005; @nilsen_can_2005].
-
-This individual will adopt different behaviors and adaptations if it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low mean and high stochasticity).
-In the grassland, the grazer may require a large but constant home range size as it moves between patches in search of food, while in the desert it may switch between dispersal in search for high-resource patches and short-term range residency within patches
 
@@
 When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [see Fig.
-\@ref(fig:mu-hyp) as well as: @herfindal_prey_2005; @nilsen_can_2005; @relyea_home_2000].
+\@ref(fig:hr-hyp)A as well as: @herfindal_prey_2005; @nilsen_can_2005; @relyea_home_2000].
 Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
@@
 It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
-Still, large-scale switches from range residency to nomadism (or vice-versa) are more likely to occur over evolutionary timescales rather than over an organism's lifespan (Fig.
-\@ref(fig:mu-hyp)), since larger ranges require greater vagility, which, in turn, is facilitated by the development of morphological features such as hinged joints and elongated limbs [@jetz_scaling_2004; @hirt_general_2017; @andersson_elbow-joint_2004; @samuels_postcranial_2013].
 
-```{r mu-hyp, fig.cap="Hypothesized range size of an organism as a function of resource abundance, $\\e(R)$.
-We expect low values of $\\e(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory.
-As $\\e(R)$ increases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.
-Note that the relationship between $\\e(R)$ and range size cannot be linear because it would require range size to be negative for high values of $\\e(R)$.", out.width = '0.5\\textwidth'}
-knitr::include_graphics('../figures/mean-abundance-hr-hypotheses.png')
+```{r hr-hyp, fig.cap="Hypothesized range size of an organism as a function of (A) resource abundance and (B) resource stochasticity.
+We expect low values of $\\e(R)$ and large values of $\\var(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory.
+As $\\e(R)$ increases or $\\var(R)$ decreases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.
+While our hypotheses imply that small-scale changes in range size can occur over ecological timescales (i.e., within an individual's lifetime), we suggest that large-scale (i.e., population-level and species-level) shifts to range residency and nomadism or frequent dispersal (or vice-versa) are more likely to occur within evolotionary timescales.\\\\Note that the relationship between range size and both $\\e(R)$ and $\\var(R)$ cannot be linear because it would require range size to be negative for high values of $\\e(R)$ or low values of $\\var(R)$.", out.width = '\\textwidth'}
+knitr::include_graphics('../figures/hr-hypotheses.png')
 ```
@@
 Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size.
+While the relationship may be approximately linear for some range of $\e(R)$, this assumption often does not hold for low or high values of $\e(R)$ [e.g., @bista_effect_2022].
+Additionally, identifying inflection points in nonlinear relationships can help understand the pressures and limitations of increasing range size.
 
@@
 As resources become increasingly unpredictable, we expect home range size to increase progressively faster (fig.
-\@ref(fig:s2-hyp)) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) the added movement required to search for food increases organisms' energetic requirements, and (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974].
+\@ref(fig:hr-hyp)B) because: (1) as $\var(R)$ increases, the chances of finding low $R$ increase superlinearly, (2) the added movement required to search for food increases organisms' energetic requirements, and (3) stochasticity reduces an organism's ability to specialize and reduce competition for $R$ [@levins_evolution_1974].
 If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new and consistent behaviors (e.g, nomadism) or adaptations (e.g., increased fat storage or food caching) to buffer themselves against times of unpredictably low $R$.
 Conversely, if changes in $\sigma^2(t, \vec u)$ are sufficiently predictable, organisms may learn to anticipate and prepare for times of greater stochasticity by pre-preemptively caching food, reducing energetic needs, migrating, or relying on alternative food sources [e.g., @van_baalen_alternative_2001].
-
-```{r s2-hyp, fig.cap="Hypothesized range size of an organism as a function of resource stochasticity, $\\var(R)$.
-We expect low values of $\\var(R)$ to result in small home-rages, since organisms are able to depend on relatively predictable resources.
-As $\\var(R)$ increases, range size should increase nonlinearly, whether this results in an expansion of the home range (in the case of range-resident organisms) or a switch to a larger range via dispersal, nomadism, or migration.
-Note that the relationship between $\\var(R)$ and range size cannot be linear because it would require range size to be negative for low values of $\\var(R)$.", out.width = '0.5\\textwidth'}
-knitr::include_graphics('../figures/variance-abundance-hr-hypotheses.png')
-```
 
@@
 
-\noindent To support our hypotheses of how organisms' range sizes are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+\noindent To evaluate our hypotheses of how organisms' range sizes are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
-The IOU model's correlated velocity produced realistic tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not to range-resident.
+The IOU model's correlated velocity produced tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not range-resident.
 Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory or memoryless movement.
@@
 1) Environments are homogeneous for a given $t$.
-Given $t$, $\e(R) = \mu(t)$ and $\var(R) = \sigma^2(t)$ are constant over space and within each set of 200 tracks, but $R$ is random and follows a $\Gamma\large(\mu(t), \sigma^2(t)\large)$ distribution.
+Given $t$, $\e(R) = \mu(t)$ and $\var(R) = \sigma^2(t)$ are constant over space and within each set of 200 tracks, but $R$ is random and follows a $\text{Gamma}\large(\mu(t), \sigma^2(t)\large)$ distribution.
 2) The are no external pressures on the simulated organism. Resources do not deplete, and there is no competition nor predator avoidance.
@@
 The remaining panels in fig.
-\@ref(fig:5-5-sims) illustrate how $\e(R)$ and $\var(R)$ jointly affect $H$ and how confusing the effects can be.
+\@ref(fig:5-5-sims) illustrate how $\e(R)$ and $\var(R)$ jointly affect $H$ and how unintuitive the effects can be.
 Since $\e(R)$ and $\var(R)$ have opposite effects on $H$, disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear $\e(R)$ and $\var(R)$).
@@
 Although the temporal trends in fig.
-\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time (panels A and B of fig.
+\@ref(fig:5-5-sims) are complex and the effects of $\e(R)$ and $\var(R)$ can be hard to disentangle, two simple relationships emerge when $H$ is shown as a function of either $\e(R)$ or $\var(R)$, rather than time: $H$ decreases nonlinearly with $\e(R)$ and increases with $\var(R)$ (panels A and B of fig.
 \@ref(fig:5-5-reg)).
-The estimated relationships follow the hypotheses we presented in figs.
-\@ref(fig:mu-hyp) and \@ref(fig:s2-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation at high values of $\var(R)$.
+The estimated relationships thus follow the hypotheses we presented in fig.
+\@ref(fig:hr-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation at high values of $\var(R)$.
 However, notice that the effect of $\var(R)$ on $E(H)$ depends strongly on $\e(R)$ (panel C): When $\e(R)$ is low, $\e(H)$ is high and $\var(R)$ does not have a strong effect, but when $\e(R)$ is high the effect of $\var(R)$ on $\e(H)$ is exponential.
@@
 The results presented in panels D-F of fig.
-\@ref(fig:tapir-mw) match our findings from the simulations: The tapir's 7-day home range decreases with $\hat\mu(t, \vec u)$ and increases with $\hat\sigma^2(t, \vec u)$, and the effect of $\hat\mu(t, \vec u)$ depends on $\hat\sigma^2(t, \vec u)$, and vice-versa.
+\@ref(fig:tapir-mw) match our findings from the simulations (Fig.
+\@ref(fig:5-5-reg)A-C): The tapir's 7-day home range decreases with $\hat\mu(t, \vec u)$ and increases with $\hat\sigma^2(t, \vec u)$, and the effect of $\hat\mu(t, \vec u)$ depends on $\hat\sigma^2(t, \vec u)$, and vice-versa.
 Alone, $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km^2^; see panel F).
@@
 
-\noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of animal how much and where organisms move.
+\noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of how much and where organisms move.
 This paper builds on earlier theoretical work [e.g., @southwood_habitat_1977; @harestad_home_1979; @stephens_optimal_1982] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
@@
 B3).
-Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to $\e(R)$ and $\var(R)$ such as selective space use, which we account for in the empirical approach.
+Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to $\e(R)$ and $\var(R)$ such as selective space use, which are included (but not explicitly accounted for) in the empirical approach.
 Below we discuss the strengths and limitations of each approach.
@@
 Firstly, modeling real-world animal movement data can produce scale-appropriate and easily interpretable estimates.
-Secondly, empirical models directly quantify the effects of $\e(R)$, $\var(R)$, and confounding variables without having to design complex and time-consuming simulations.
+Secondly, empirical data contain information on the effects of $\e(R)$, $\var(R)$, and confounding variables without having to design complex and time-consuming simulations.
 However, it is not always possible to quantify confounding variables.
@@
 
-Similarly, if $R$ is often non-measurable.
-Proxies of $R$, such as NDVI [@pettorelli_normalized_2011], which may introduce complexities.
+Similarly, if $R$ non-measurable (as is often the case), $R$ must be estimated with proxies such as NDVI [@pettorelli_normalized_2011], which may introduce complexities.
 While $R$ and NDVI are correlated for many species [e.g., @middleton_green-wave_2018; @geremia_migrating_2019; @merkle_spatial_2019; @phillips_evaluating_2008; @seigle-ferrand_systematic_2021; @merkle_large_2016], the relationship between the two can be weak [@gautam_ndvi_2019], satellite-dependent [@huang_commentary_2021], and nonlinear [@huang_commentary_2021; @fan_global_2016].
@@
 \noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
-We provide realistic and flexible hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' range sizes and movement behavior.
+We provide sensible and unifying hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' range sizes and movement behavior.
 We demonstrate that organisms' range sizes decrease with resource abundance, increase with resource stochasticity, and that the effects of $\var(R)$ can depend strongly on $\e(R)$.

### 2025-02-10 analysis/figures/home-range-hypotheses.R

@@ -18,20 +18,15 @@ d <-
          h_2 = if_else(split, 2 * mu^2 - 7 * mu + 3.916873, h_1),
          h_3 = if_else(split, h_2 + 30 * (0.35 - mu)^2, h_2))
 
-notes <- tibble(x = c(0.13, 0.32, 0.19, 0.68),
-                y = c(2.05, 3.2, 1.35, 2.05),
-                text = c('range-resident', 'nomadic or\ndispersing',
-                         'evolutionary timescale', 'ecological timescale'))
+notes <- tibble(x = c(0.13, 0.32),
+                y = c(2.05, 3.2),
+                text = c('range-resident', 'nomadic or\ndispersing'))
 
 p_mu <-
   ggplot() +
   coord_cartesian(ylim = c(0, 4)) +
   geom_ribbon(aes(mu, ymin = h_1, ymax = h_3), d, color = pal[1],
               alpha = 0.2, linewidth = 2, fill = pal[1]) +
-  stat_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180,
-             outside = FALSE) +
-  stat_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0,
-             outside = FALSE) +
   geom_text(aes(x = x, y = y, label = text), notes, size = 6,
             fontface = 'bold') +
   scale_x_continuous(bquote(paste(bold('Resource abundance, E('),
@@ -45,10 +40,6 @@ p_s2 <-
   coord_cartesian(ylim = c(0, 4)) +
   geom_ribbon(aes(mu, ymin = h_1, ymax = h_3), d, color = pal[2],
               alpha = 0.2, linewidth = 2, fill = pal[2]) +
-  stat_brace(aes(x = c(0, 0.345), y = c(1.45, 1.7)), rotate = 180,
-             outside = FALSE) +
-  stat_brace(aes(x = c(0.36, 1), y = c(1.95, 1.7)), rotate = 0,
-             outside = FALSE) +
   geom_text(aes(x = x, y = y, label = text), notes, size = 6,
             fontface = 'bold') +
   scale_x_reverse(bquote(paste(bold(Resource~'stochasticity, Var('),
@@ -58,6 +49,7 @@ p_s2 <-
   scale_y_continuous(hr_lab, breaks = 0, expand = c(0, 0)); p_s2
 
 # save the plots ----
+# individual plots
[…]

### 2025-02-17 writing/appendix-A-concepts-and-definitions.Rmd
remove AmNat csl line

@@
 bibliography: 'hr-resource-stoch.bib'
-# csl: 'the-american-naturalist.csl'
 fontsize: 12pt

### 2025-02-17 writing/appendix-B-simulations.Rmd
"resources" to "distribution of R"; decr fig size

@@
 Within a Bayesian framework, Directed Acyclical Graphs (DAGs) can be used to infer causality [@mcelreath_statistical_2016].
-The DAG below can be interpreted as follows: $\e(R)$ and $\var(R)$ are the (controlled) variables that determine resources, $R$, and, consequently, an organism's home-range size, $H$.
-Additionally, as indicated by the arrow going from $\e(R)$ to $\var(R)$, the effect of $\var(R)$ depends of the value of $\e(R)$ (and vice-versa, but arrows in DAGs have to be unidirectional).
+The DAG below can be interpreted as follows: $\e(R)$ and $\var(R)$ are the (controlled) variables that determine the distribution of resources, $R$, which, in turn determines an organism's home-range size, $H$.
 
-```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$.", fig.height=4, fig.width=6, out.width='100%'}
+```{r dag, eval=TRUE, echo = FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on the distributions of $R$ and $H$ in the simulations.", fig.height=4, fig.width=4, out.width='50%'}
 dagify(H ~ R,

### 2025-02-17 writing/appendix-C-empirical-modeling.Rmd
rm "unaccounted"; reduced fig size

@@
 bibliography: 'hr-resource-stoch.bib'
-csl: 'the-american-naturalist.csl'
 fontsize: 12pt
@@
 ```{r dag, eval=TRUE, echo=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
-$Z$ and $S$ indicate unaccounted confounding factors that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=6, fig.width=7, out.width='\\linewidth'}
+$Z$ and $S$ indicate confounds that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=6, fig.width=6, out.width='75%'}
 dagify(NDVI ~ S,

### 2025-02-17 writing/manuscript.Rmd
finished addressing reviewer comments

@@
 ```{r, echo=FALSE}
-abstract <- "\\textbf{Background:} From megafauna to amoebas, the amount of space organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
+abstract <- "\\textbf{Background:} From megafauna to amoebas, the amount of space heterotrophic organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats.
 This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to spatiotemporal changes in mean resources, while responses to unpredictable changes in resources (i.e., variance in resources or resource stochasticity) are still largely unknown.
 Since organisms adjust to variable environmental conditions, failing to consider the effects of resource unpredictability can result in an insufficient understanding of an organism's range size.
-\\textbf{Methods:} We leverage the available literature to provide a unifying framework and two hypotheses for the effects of resource abundance and stochasticity on organisms' range sizes.
+\\textbf{Methods:} We leverage the available literature to provide a unifying framework and hypothesis for the effects of resource abundance and stochasticity on organisms' range sizes.
 We then use simulated movement data to demonstrate how the combined effects of resource abundance and stochasticity interact to shape predictable patterns in range size.
-Finally, we test the two hypotheses using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
+Finally, we test the hypothesis using real-world tracking data on a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado.
 \\textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity, and the effects of resource stochasticity depend strongly on resource abundance.
@@
 
-\noindent \textbf{References}: 109
+\noindent \textbf{References}: 112
 
@@
 Afterwards, we provide a review of the effects of resource stochasticity on organisms' range sizes while suggesting a second simple and unifying hypothesis.
-Subsequently, we support the two hypotheses using quantitative, simulated responses in range size to changes in resource abundance and stochasticity.
+Subsequently, we support the hypothesis using quantitative, simulated responses in range size to changes in resource abundance and stochasticity.
 Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (*Tapirus terrestris*) from the Brazilian Cerrado [@medici_movement_2022].
@@
 \noindent Resources (e.g., food, water, shelter, heat) are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate [@lindstedt_seasonality_1985; @schmidt_interplay_2020; @morellet_seasonality_2013], weather [@fjelldal_nightly_2021; @morellet_seasonality_2013], competitive pressure [@torrez-herrera_monkeys_2020; @rich_anthropogenic_2012], and differences in energetics at among individuals [@schmidt_interplay_2020] and species [@jetz_scaling_2004].
-Thus, we can treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$.
+Thus, it is possible to treat the amount of resources $R$ at a given point in time ($t$) and space (location vector $\vec u$) as a random variable, denoted as $R(t, \vec u)$.
 Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean.
@@
 
-When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [see Fig.
+When $\e(R)$ is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive [see fig.
 \@ref(fig:hr-hyp)A as well as: @herfindal_prey_2005; @nilsen_can_2005; @relyea_home_2000].
 Like Harestad and Bunnel @harestad_home_1979, we also expect organisms' range sizes to increase nonlinearly as $\e(R)$ decreases, but we highlight that organisms may adopt different behaviors at low values of $\e(R)$.
-These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019; @torrez-herrera_monkeys_2020], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
+These behaviors include maximal home range expansion [home range size is limited by vagility, habitat structure, competition, and predation, e.g., @torrez-herrera_monkeys_2020; @rich_anthropogenic_2012; @dawe_influence_2014; @berger-tal_invisible_2019], migration [@samarra_movements_2017; @middleton_green-wave_2018; @geremia_migrating_2019], and nomadism [@singh_migration_2012; @polansky_elucidating_2015; @teitelbaum_beyond_2019; @nandintsetseg_variability_2019].
 It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior [mammals: @teitelbaum_how_2015; moose, *Alces alces*: @singh_migration_2012; eagles, *Haliaeetus leucocephalus*: @wheat_migrate_2017; @poessel_interpreting_2022; lesser flamingos, *Phoeniconaias minor*: @pretorius_movement_2020].
@@
 We expect low values of $\\e(R)$ and large values of $\\var(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory.
-As $\\e(R)$ increases or $\\var(R)$ decreases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.
-While our hypotheses imply that small-scale changes in range size can occur over ecological timescales (i.e., within an individual's lifetime), we suggest that large-scale (i.e., population-level and species-level) shifts to range residency and nomadism or frequent dispersal (or vice-versa) are more likely to occur within evolotionary timescales.\\\\Note that the relationship between range size and both $\\e(R)$ and $\\var(R)$ cannot be linear because it would require range size to be negative for high values of $\\e(R)$ or low values of $\\var(R)$.", out.width = '\\textwidth'}
+As $\\e(R)$ increases or $\\var(R)$ decreases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.\\\\Note that the relationship between range size and both $\\e(R)$ and $\\var(R)$ cannot be of the form $H = \\beta_0 + \\beta_1 \\e(R) + \\beta_2 \\var(R)$ because it would require range size to be negative for high values of $\\e(R)$ or low values of $\\var(R)$.", out.width = '\\textwidth'}
 knitr::include_graphics('../figures/hr-hypotheses.png')
@@
 For example, prey abundance in a given area over time may approximately follow a Poisson distribution, which implies that the mean and variance will be approximately equal.
-When prey are scarce (e.g., a mean of 10), the variance will also be low, and when prey are abundant (e.g., a mean of 300) the variance will also be high.
-This occurs because the behavior, fitness, and predator-prey dynamics of ~300 prey are more stochastic than those of 10 prey [@campillo_effect_2012].
+When prey are scarce, the variance will also be low, and when prey are abundant the variance will also be high.
+This occurs because the behavior, fitness, and predator-prey dynamics of many prey are more stochastic than those of few prey [@campillo_effect_2012].
 Similarly, in the case of fully bounded random variables, the variance is generally lowest when the mean is near either boundary.
@@
 
-\noindent To evaluate our hypotheses of how organisms' range sizes are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
+\noindent To evaluate our hypothesis of how organisms' range sizes are affected by $\e(R)$, $\var(R)$, and the interaction effect of $\e(R)$ and $\var(R)$, we present the results from a series of quantitative simulations.
 To start, we used the `ctmm` package [@fleming_ctmm_2021] for `R` [@r_core_team_r_2023] to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model [IOU model, see @gurarie_correlated_2017].
@@
 
-Additional information is provided in Appendix B, including the directed acyclical graph [see fig.
-B6 and @mcelreath_statistical_2016] we used to infer causal the mechanisms of changes in $H$ and estimate the direct effects of $\e(R)$ and $\var(R)$ on $H$ (contrast the graph with fig.
-C3 and the empirical case study below).
+Based on the assumptions above, we constructed the following causal model for the simulated effects of $\e(R)$ and $\var(R)$ on $H$ [see fig.
+\@ref(fig:dag-sims) and @mcelreath_statistical_2016]: $\e(R)$ and $\var(R)$ were determined independently of each other, but they jointly determined the distribution of $R$, which, in turn, determined the distribution of $H$.
+Additional information is provided in Appendix B.
 
-```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogeneous over space for a given $t$).
-Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
-Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
-knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+```{r dag-sims, eval=TRUE,  message=FALSE, warning=FALSE, fig.cap="Directed acyclical graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on the distributions of $R$ and $H$ in the simulations.", fig.height=4, fig.width=4, out.width='50%'}
+library('dplyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+library('ggdag')   # for directed acyclical graphs
+source('../analysis/figures/default-figure-styling.R') # for color palette
+
+dagify(H ~ R,
+       R ~ E + V,
+       outcome = 'H',
+       labels = c('H' = 'H', 'R' = 'R', 'E' = 'E(R)', 'V' = 'Var(R)'),
+       coords = list(x = c(R = 0, E = -1, V = 1, H = 0),
+                     y = c(R = 1, E = 2, V = 2, H = 0))) %>%
+  tidy_dagitty() %>%
+  mutate(col = case_when(name == 'E' ~ '1',
+                         name == 'EN' ~ '1',
+                         name == 'V' ~ '2',
+                         name == 'VN' ~ '2',
+                         name == 'H' ~ '3',
+                         TRUE ~ '4')) %>%
+  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_point(aes(color = col), size = 20, alpha = 0.3) +
+  geom_dag_edges() +
+  geom_dag_text(aes(label = label), color = 'black', size = 4.5) +
+  scale_color_manual(values = c(pal[1:3], 'black')) +
+  theme_dag() +
+  theme(legend.position = 'none')
 ```
@@
 Also note how the noise in the green lines also decreases as $\e(R)$ increases.
+
+```{r 5-5-sims, fig.cap="Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogeneous over space for a given $t$).
+Note how $H$ decreases nonlinearly as $\\mu(t)$ increases and increases nonlinearly as $\\sigma^2(t)$ increases.
+Additionally, the variance in $H$ is higher when $\\mu(t)$ is lower or $\\sigma^2(t)$ is higher, and changes in $\\sigma^2(t)$ have greater impacts when $\\mu(t)$ is low.", out.width='100%'}
+knitr::include_graphics('../figures/mean-variance-5-by-5-hr-sims.png')
+```
 
@@
 \@ref(fig:5-5-reg)).
-The estimated relationships thus follow the hypotheses we presented in fig.
+The estimated relationships thus follow the hypothesis we presented in fig.
 \@ref(fig:hr-hyp), although we found that the effect of $\var(R)$ at average $\e(R)$ was linear with a slight sublinear saturation at high values of $\var(R)$.
@@
 
-\noindent The simulations in the section above support the hypotheses we presented in the background section, but they are based on assumptions that are often not met in real natural environments.
+\noindent The simulations in the section above support the hypothesis we presented in the background section, but they are based on assumptions that are often not met in real natural environments.
 Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory.
@@
 
-In this section, we test the hypotheses using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with empirical estimates of $\e(R)$ and $\var(R)$.
+In this section, we test the hypothesis using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with empirical estimates of $\e(R)$ and $\var(R)$.
 We measure $R$ using Normalized Difference Vegetation Index [NDVI, see @pettorelli_normalized_2011], a remote-sensed measure of landscape greenness, as a proxy for forage abundance.
@@
 
+```{r, eval=FALSE}
+library('dplyr')
+library('purrr')
+library('ctmm')
+anna <- readRDS('models/tapirs/CE_31_ANNA-window-7-days-dt-1-days.rds') %>%
+  mutate(tau_p_days = map_dbl(model, \(.m) {
+    ci <- summary(.m, units = FALSE)$CI # values in seconds
+    if(any(rownames(ci) == 'τ[position] (seconds)')) {
+      return('days' %#% ci['τ[position] (seconds)', 'est'])
+    } else {
+      return(NA_real_)
+    }
+  }),
+  dof_area = map_dbl(model, \(.m) {
+    unname(summary(.m, units = FALSE)$DOF['area'])
+  }))
+range(anna$tau_p_days, na.rm = TRUE)
+mean(! is.na(anna$tau_p_days))
+range(anna$dof_area, na.rm = TRUE)
+```
+
 Fig.
-\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapts its 7-day home-range size to spatiotemporal changes in estimated $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ (telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022).
+\@ref(fig:tapir-mw) illustrates how a tapir in the Brazilian Cerrado adapted its 7-day home-range size to spatiotemporal changes in estimated $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ (telemetry data from the individual labelled as "Anna" in the dataset from @medici_movement_2022).
 Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period.
@@
 Panel C shows the changes in the tapir's 7-day home range over time.
+All 457 of the 7-day windows had a minimum effective sample size of 7 range crossings [range: 7.7 -- 69.6, see @calabrese_ctmm_2016], and 92% had resolvable (i.e., non-NA) home range crossing times, all of which were below 17 hours.
 Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018).
@@
 The results presented in panels D-F of fig.
-\@ref(fig:tapir-mw) match our findings from the simulations (Fig.
+\@ref(fig:tapir-mw) match our findings from the simulations (fig.
 \@ref(fig:5-5-reg)A-C): The tapir's 7-day home range decreases with $\hat\mu(t, \vec u)$ and increases with $\hat\sigma^2(t, \vec u)$, and the effect of $\hat\mu(t, \vec u)$ depends on $\hat\sigma^2(t, \vec u)$, and vice-versa.
@@
 Additionally, note how high NDVI values ($\hat \mu(t, \vec u) > 0.8$) cause $\hat\sigma^2(t, \vec u)$ to have little to no effect on home-range size, as indicated by the vertical contour line in panel F.
+Similar conclusions can be drawn for the animal's diffusion (i.e., area covered per unit time), which is a more appropriate measure of space use when animals are not range resident [@calabrese_ctmm_2016].
 
@@
 See Appendix C for additional information.
-The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} [27].", out.width='100%'}
+The tapir movement data corresponds to the individual named ``Anna\" from the Cerrado sample of Medici \\emph{et al.} [29].", out.width='100%'}
 knitr::include_graphics('../figures/tapir-example.png')
+```
+
+Quantifying the direct effects of $\e(R)$ and $\var(R)$ on $H$ using empirical data is more complex than with simulated data, and it requires a different causal framework, particularly in the case of observational studies (as opposed to experimentally-controlled studies; see fig.
+\@ref(fig:dag-empirical)).
+Unlike with the simulations, $\e(R)$ and $\var(R)$ are not controlled variables and instead depend on the distribution of $R$, which depends on a variety of other factors (that we exclude from the figure for simplicity).
+Both $\e(R)$ and $\var(R)$ then impact $H$ as well as habitat-level variables (e.g., competition, predation, etc.; indicated as $Z$) that also affect $H$.
+Additionally, estimating $R$ via a proxy (NDVI) adds satellite-level noise and confounds [e.g., saturation, cloud cover, spatiotemporal averaging -- indicated as $S$, see @fan_global_2016; @gautam_ndvi_2019; @huang_commentary_2021].
+However, $\e(R)$ and $\var(R)$ can be correlated to $\e(\text{NDVI})$ and $\var(\text{NDVI})$, respectively, provided that analysts use models that are sufficiently smooth and flexible at the relevant spatiotemporal scale [@pease_ecological_2024].
+We discuss this in further detail in the section below on the strengths and limitations of the empirical approach.
+
+```{r dag-empirical, eval=TRUE, echo=FALSE, fig.cap="Directed Acyclical Graph assumed for inferring the causal effects of $\\e(R)$ and $\\var(R)$ on $H$, where NDVI was used as a proxy for $R$.
+$Z$ and $S$ indicate confounds that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).", fig.height=6, fig.width=7, out.width='75%'}
+library('dplyr')   # for data wrangling
+library('ggplot2') # for fancy plots
+library('ggdag')   # for directed acyclical graphs
+source('../analysis/figures/default-figure-styling.R') # for color palette
+
+dagify(NDVI ~ S,
+       EN ~ NDVI + E,
+       VN ~ NDVI + V,
+       E ~ R,
+       V ~ R,
+       Z ~ E + V,
+       S ~ R,
+       H ~ E + V + Z,
+       outcome = 'H',
+       labels = c(H = 'H', R = 'R', E = 'E(R)', V = 'Var(R)', S = 'S',
+                  Z = 'Z', NDVI = 'NDVI', EN = 'E(NDVI)', VN = 'Var(NDVI)'),
+       coords = list(x = c(R = 0, NDVI = 0, Z = 0, S = 0, H = 0,
+                           E = -1, V = 1, EN = -1, VN = 1),
+                     y = c(R = 3,
+                           Z = 1.5, H = 0, E = 1.5, V = 1.5,
+                           S = 4.5, NDVI = 6, EN = 4.5, VN = 4.5))) %>%
+  tidy_dagitty() %>%
+  mutate(col = case_when(name == 'E' ~ '1',
+                         name == 'EN' ~ '1',
+                         name == 'V' ~ '2',
+                         name == 'VN' ~ '2',
+                         name == 'H' ~ '3',
+                         TRUE ~ '4')) %>%
+  ggplot(aes(x = x, y = y, xend = xend, yend = yend)) +
+  geom_dag_point(aes(color = col), size = 20, alpha = 0.3) +
+  geom_dag_edges() +
+  geom_dag_text(aes(label = label), color = 'black', size = 3) +
+  scale_color_manual(values = c(pal[1:3], 'black')) +
+  theme_dag() +
+  theme(legend.position = 'none')
 ```
@@
 \noindent The amount of space organisms use is determined by a multitude of factors [@nathan_movement_2008], but the search for resources is often a main driver of how much and where organisms move.
-This paper builds on earlier theoretical work [e.g., @southwood_habitat_1977; @harestad_home_1979; @stephens_optimal_1982] and presents two hypotheses that describe the effects of resource abundance and stochasticity on organisms' range sizes.
-We use quantitative simulations and an empirical case study to support the hypotheses and show that, together, they provide a simple framework for understanding how motile organisms adapt their movement in dynamic environments.
+This paper builds on earlier theoretical work [e.g., @southwood_habitat_1977; @harestad_home_1979; @stephens_optimal_1982] and presents a unifying hypothesis that describes the effects of resource abundance and stochasticity on organisms' range sizes.
+We use quantitative simulations and an empirical case study to support the hypothesis and show that it provides a simple framework for understanding how motile organisms adapt their movement in dynamic environments.
 Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: $H$ decreases with $\e(R)$ and increases with $\var(R)$.
@@
 The use of continuous-time movement models coupled with few drivers of movement supported realistic data that could be explained by straightforward causal models.
-The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).
+The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity; see Figure \@ref(fig:dag-sims)) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).
 
@@
 This complexity can introduce two sources of bias: ecosystem-level biases (indicated as $Z$ in the directed acyclical graph in fig.
-C3) and satellite-level confounding variables ($S$ in fig.
-C3).
+\@ref(fig:dag-empirical)) and satellite-level confounding variables ($S$ in fig.
+\@ref(fig:dag-empirical)).
 Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to $R$ and NDVI [@jetz_scaling_2004; @prox_framework_2020].
@@
 \noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity.
-We provide sensible and unifying hypotheses of the effects of $\e(R)$ and $\var(R)$ on organisms' range sizes and movement behavior.
+We provide a sensible and unifying hypothesis of the effects of $\e(R)$ and $\var(R)$ on organisms' range sizes and movement behavior.
 We demonstrate that organisms' range sizes decrease with resource abundance, increase with resource stochasticity, and that the effects of $\var(R)$ can depend strongly on $\e(R)$.

### 2025-02-17 writing/manuscript.Rmd
changed "Figure" to "fig."; checked diffs

@@
 The use of continuous-time movement models coupled with few drivers of movement supported realistic data that could be explained by straightforward causal models.
-The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity; see Figure \@ref(fig:dag-sims)) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).
+The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity; see fig.
+\@ref(fig:dag-sims)) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).
 

### 2025-02-18 writing/appendix-A-concepts-and-definitions.Rmd
submitting response to reviews

@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' range sizes
+title: \large How resource abundance and resource stochasticity affect organisms' range sizes
 subtitle: "Appendix A: Concepts and definitions"
-# author:
-#   - name: Stefano Mezzini
-#     email: stefano.mezzini@ubc.ca
-#     institute: [braes, biol]
-#   - name: Chris H. Fleming
-#     institute: [ucf, smith]
-#   - name: E. Patrícia Medici
-#     institute: [tapir, iucn, escas]
-#   - name: Michael J. Noonan
-#     email: michael.noonan@ubc.ca
-#     institute: [braes, biol, cmps]
-# institute:
-#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+author:
+  - name: Stefano Mezzini
+    email: stefano.mezzini@ubc.ca
+    institute: [braes, biol]
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
+  - name: E. Patrícia Medici
+    institute: [tapir, iucn, escas]
+  - name: Michael J. Noonan
+    email: michael.noonan@ubc.ca
+    institute: [braes, biol, cmps]
+institute:
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'

### 2025-02-18 writing/appendix-B-simulations.Rmd
submitting response to reviews

@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' range sizes
+title: \large How resource abundance and resource stochasticity affect organisms' range sizes
 subtitle: "Appendix B: Simulations"
-# author:
-#   - name: Stefano Mezzini
-#     email: stefano.mezzini@ubc.ca
-#     institute: [biol, braes]
-#   - name: Chris H. Fleming
-#     institute: [ucf, smith]
-#   - name: E. Patrícia Medici
-#     institute: [tapir, iucn, escas]
-#   - name: Michael J. Noonan
-#     email: michael.noonan@ubc.ca
-#     institute: [biol, braes, cmps]
-# institute:
-#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+author:
+  - name: Stefano Mezzini
+    email: stefano.mezzini@ubc.ca
+    institute: [biol, braes]
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
+  - name: E. Patrícia Medici
+    institute: [tapir, iucn, escas]
+  - name: Michael J. Noonan
+    email: michael.noonan@ubc.ca
+    institute: [biol, braes, cmps]
+institute:
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'

### 2025-02-18 writing/appendix-C-empirical-modeling.Rmd
submitting response to reviews

@@
 ---
-title: \large How resource abundance and stochasticity affect organisms' range sizes
+title: \large How resource abundance and resource stochasticity affect organisms' range sizes
 subtitle: "Appendix C: Empirical modeling"
-# author:
-#   - name: Stefano Mezzini
-#     email: stefano.mezzini@ubc.ca
-#     institute: [biol, braes]
-#   - name: Chris H. Fleming
-#     institute: [ucf, smith]
-#   - name: E. Patrícia Medici
-#     institute: [tapir, iucn, escas]
-#   - name: Michael J. Noonan
-#     email: michael.noonan@ubc.ca
-#     institute: [biol, braes, cmps]
-# institute:
-#   - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
-#   - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
-#   - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
-#   - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
-#   - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
-#   - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+author:
+  - name: Stefano Mezzini
+    email: stefano.mezzini@ubc.ca
+    institute: [biol, braes]
+  - name: Chris H. Fleming
+    institute: [ucf, smith]
+  - name: E. Patrícia Medici
+    institute: [tapir, iucn, escas]
+  - name: Michael J. Noonan
+    email: michael.noonan@ubc.ca
+    institute: [biol, braes, cmps]
+institute:
+  - biol: Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - braes: Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
+  - ucf: Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.
+  - smith: Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.
+  - tapir: Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - iucn: IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.
+  - escas: Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.
+  - cmps: Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.
 bibliography: 'hr-resource-stoch.bib'

### 2025-02-18 writing/manuscript.Rmd
submitting response to reviews

@@
     number_sections: false
+    keep_tex: true
 editor_options: 
@@
 <!-- start main body on a new page -->
-
-\newpage
 

### 2025-02-18 writing/manuscript.tex
submitting response to reviews

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace}\doublespacing
\usepackage{indentfirst}
\usepackage[small]{titlesec}
\usepackage{wrapfig}
\usepackage{caption}
\captionsetup[figure]{font={scriptsize, doublespacing}, labelfont=bf, aboveskip=4pt, belowskip=-15pt}
\usepackage{hanging}
\usepackage[left]{lineno}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={How resource abundance and resource stochasticity affect organisms' range sizes},
  pdfauthor={Stefano Mezzini1,2; Chris H. Fleming3,4; E. Patrícia Medici5,6,7; Michael J. Noonan1,2,8,},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{\large How resource abundance and resource stochasticity affect organisms' range sizes}
\author{Stefano Mezzini\textsuperscript{1,2} \and Chris H. Fleming\textsuperscript{3,4} \and E. Patrícia Medici\textsuperscript{5,6,7} \and Michael J. Noonan\textsuperscript{1,2,8,*}}
\date{}

\begin{document}
\maketitle

\textsuperscript{1} Okanagan Institute for Biodiversity, Resilience, and Ecosystem Services, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.\\
\textsuperscript{2} Department of Biology, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.\\
\textsuperscript{3} Department of Biology, University of Central Florida, Orlando, Florida 32816, United States.\\
\textsuperscript{4} Smithsonian Conservation Biology Institute, National Zoological Park, 1500 Remount Rd., Front Royal, VA 22630, United States.\\
\textsuperscript{5} Lowland Tapir Conservation Initiative (LTCI), Instituto de Pesquisas Ecológicas (IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.\\
\textsuperscript{6} IUCN SSC Tapir Specialist Group (TSG), Campo Grande, Brazil.\\
\textsuperscript{7} Escola Superior de Conservação Ambiental E Sustentabilidade (ESCAS/IPÊ), Rodovia Dom Pedro I, km 47, Nazaré Paulista, São Paulo 12960-000, Brazil.\\
\textsuperscript{8} Department of Computer Science, Math, Physics, and Statistics, The University of British Columbia Okanagan, Kelowna, British Columbia, Canada.

\textsuperscript{*} Correspondence: \href{mailto:michael.noonan@ubc.ca}{Michael J. Noonan \textless{}\href{mailto:michael.noonan@ubc.ca}{\nolinkurl{michael.noonan@ubc.ca}}\textgreater{}}

\newcommand*\e{\text{E}}

\newcommand*\var{\text{Var}}

\clearpage

\noindent \textbf{Article type}: Research article

\noindent \textbf{Words in abstract}: 274

\noindent \textbf{Words in main text}: 6722

\noindent \textbf{Figures}: 6

\noindent \textbf{Tables}: 0

\noindent \textbf{References}: 112

\noindent \textbf{Appendices}: 3

\noindent \textbf{Key words:} energetics, resource abundance, resource stochasticity, environmental stochasticity, home range, range size, movement behavior, \texttt{ctmm}

\newpage

\doublespacing

\linenumbers

\section{Abstract}\label{abstract}

\noindent \textbf{Background:} From megafauna to amoebas, the amount of space heterotrophic organisms use is thought to be tightly linked to the availability of resources within their habitats, such that organisms living in productive habitats generally require less space than those in resource-poor habitats. This hypothesis has widespread empirical support, but existing studies have focused primarily on responses to spatiotemporal changes in mean resources, while responses to unpredictable changes in resources (i.e., variance in resources or resource stochasticity) are still largely unknown. Since organisms adjust to variable environmental conditions, failing to consider the effects of resource unpredictability can result in an insufficient understanding of an organism's range size. \textbf{Methods:} We leverage the available literature to provide a unifying framework and hypothesis for the effects of resource abundance and stochasticity on organisms' range sizes. We then use simulated movement data to demonstrate how the combined effects of resource abundance and stochasticity interact to shape predictable patterns in range size. Finally, we test the hypothesis using real-world tracking data on a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado. \textbf{Results:} Organisms' range sizes decrease nonlinearly with resource abundance and increase nonlinearly with resource stochasticity, and the effects of resource stochasticity depend strongly on resource abundance. Additionally, the distribution and predictability of resources can exacerbate the effects of other drivers of movement, such as resource depletion, competition, and predation. \textbf{Conclusions:} Accounting for resource abundance and stochasticity is crucial for understanding the movement behavior of free-ranging organisms. Failing to account for resource stochasticity can lead to an incomplete and incorrect understanding of how and why organisms move, particularly during periods of rapid change.

\newpage

\section{Background}\label{background}

\noindent The amount of resources an organism is able to access is a strong determinant of its fitness. Resource limitations can cause individuals to experience a negative energetic balance, which can then result in lower fitness {[}1,2{]}, altered physiology {[}2--5{]}, lower chance of reproduction {[}2,6--8{]}, and even death {[}9,10{]}. Thus, many organisms adapt their behaviors and/or physiology in response to changes in local resource abundance to ensure their needs are met {[}e.g., soil amoebae \emph{Dictyostelium spp.}: 11,plants: 12,and animals: 13{]}.

While there are many ways that individuals can respond to resource availability, movement represents one of the most readily available traits that motile species can adjust {[}14--16{]}. The relationship between organisms' movement and resource abundance has long been of interest to biologists. In his seminal paper, Burt {[}17{]} considered the search for food as the primary driver for movement within an organism's home range. Three decades after, Southwood {[}18{]} suggested that change in resource abundance drives how organisms decide where to live and when to reproduce. Two years later, Harestad and Bunnel {[}13{]} proposed that the simplest relationship between resource abundance and an organism's home-range size is

\begin{equation}
H = C / R,
\label{eq:harestad-eq}
\end{equation}

\noindent where \(H\) is the organism's home-range size, \(C\) is the organism's resource consumption rate (kcal day\(^{-1}\)), and \(R\) is the resources the organism can access (kcal day\(^{-1}\) unit area\(^{-1}\)). Harestad and Bunnel's model is simple to conceptualize, and it allows for testable predictions, but few studies are structured around a set of theoretical expectations such as Harestad and Bunnel's hypothesis. Many researchers have since demonstrated that organisms adapt their range sizes in response to resource abundance, but results are typically reported as independent, novel findings. Perhaps more problematic is the fact that, while much work has been done on estimating organisms' responses to changes in mean resource abundance, there is little information on how organisms respond to unpredictable changes in resources {[}i.e., resource stochasticity, but see: 19,20--22{]}. Thus, there remains a need for a clear, unifying hypothesis of the effects of both resource abundance and stochasticity on organisms' range sizes.

Here, we refer to a location's average amount of resources as ``resource abundance'', while we use the phrase ``resource stochasticity'' to indicate the variability in resources after accounting for changes in the mean. We argue that, on its own, a habitat's resource abundance is not sufficient to assess the habitat's quality, nor make predictions about how much space an organism might use. To see this, consider, for instance, a herbivore grazing in a grassland with relatively low but constant forage availability (i.e., low mean and variance). The animal may require a large but constant home range size as it moves between patches in search of food. If, instead, it lived in a desert with equally scarce forage but rare, sudden, and strong pulses of resources (i.e., low long-term mean and high stochasticity), it may switch between dispersal in search for high-resource patches and short-term range residency within patches {[}\emph{sensu} 15,see 23,24,25{]}. Previous studies suggest that resource stochasticity may decrease organisms' fitness and landscapes' energetic balances {[}e.g., 26{]}, but there is still limited empirical evidence to support this hypothesis {[}but see: 21,27,28{]}.

In this paper, we illustrate how an organism's range size can be expected to depend on both the abundance and unpredictability of resources. First, we set the theoretical background necessary for the successive sections by introducing key concepts and notation. Next, we provide a review of the effects of resource abundance on range sizes while suggesting a simple and unifying hypothesis. Afterwards, we provide a review of the effects of resource stochasticity on organisms' range sizes while suggesting a second simple and unifying hypothesis. Subsequently, we support the hypothesis using quantitative, simulated responses in range size to changes in resource abundance and stochasticity. Finally, we demonstrate how this framework can be used in practice to describe the movement ecology of a lowland tapir (\emph{Tapirus terrestris}) from the Brazilian Cerrado {[}29{]}.

\section{Resources as a random variable}\label{resources-as-a-random-variable}

\noindent Resources (e.g., food, water, shelter, heat) are often unpredictable (and difficult to quantify), since they depend on various factors which cannot be accounted for easily, including climate {[}7,30,31{]}, weather {[}31,32{]}, competitive pressure {[}33,34{]}, and differences in energetics at among individuals {[}7{]} and species {[}35{]}. Thus, it is possible to treat the amount of resources \(R\) at a given point in time (\(t\)) and space (location vector \(\vec u\)) as a random variable, denoted as \(R(t, \vec u)\). Treating resources as a random variable allows us to leverage techniques from probability theory and statistics, such as the expectation of a random variable (i.e., its mean) and its variance around the mean. We indicate the expected value and variance of random variable \(R\) using \(\text{E}(R)\) and \(\text{Var}(R)\), respectively, and we use \(\mu(t, \vec u)\) and \(\sigma^2(t, \vec u)\) to indicate them as functions of time (\(t\)) and space (\(\vec u\)). Appendix A defines and expands on the concepts of probability distributions, expected value, variance, and provides examples of them for Gamma and Beta distributions.

\subsection{\texorpdfstring{Effects of resource abundance, \(\text{E}(R)\)}{Effects of resource abundance, \textbackslash text\{E\}(R)}}\label{effects-of-resource-abundance-texter}

\noindent While organisms' needs vary greatly between taxonomic groups, some needs are essential for the growth, survival, and reproduction of most organisms. All heterotrophic organisms require sources of chemical energy (i.e., food), water, and various limiting nutrients {[}36--38{]}. As the abundance of essential resources fluctuates, motile organisms can move to new locations or `patches' to meet their requirements {[}15,39{]}, but movement also increases energetic needs {[}40{]}.

When \(\text{E}(R)\) is high, we expect organisms' ranges to be relatively small and near the smallest amount of space required to survive {[}see fig.~\ref{fig:hr-hyp}A as well as: 27,28,41{]}. Like Harestad and Bunnel {[}13{]}, we also expect organisms' range sizes to increase nonlinearly as \(\text{E}(R)\) decreases, but we highlight that organisms may adopt different behaviors at low values of \(\text{E}(R)\). These behaviors include maximal home range expansion {[}home range size is limited by vagility, habitat structure, competition, and predation, e.g., 33,34,42,43{]}, migration {[}44--46{]}, and nomadism {[}23,25,47,48{]}. It is unclear when organisms switch from range residency to migration or nomadism (or vice-versa), but understanding the gradient among these types of movement is necessary for quantifying the effect of resource abundance on organisms' range size and movement behavior {[}mammals: 49,moose, \emph{Alces alces}: 23,eagles, \emph{Haliaeetus leucocephalus}: 24,50,lesser flamingos, \emph{Phoeniconaias minor}: 51{]}.

\begin{figure}

{\centering \includegraphics[width=\textwidth]{../figures/hr-hypotheses}

}

\caption{Hypothesized range size of an organism as a function of (A) resource abundance and (B) resource stochasticity. We expect low values of $\text{E}(R)$ and large values of $\text{Var}(R)$ to result in a large range, since organisms are forced to explore large areas to collect the resources they require to survive, whether they be range-resident, nomadic, or migratory. As $\text{E}(R)$ increases or $\text{Var}(R)$ decreases, range size should decrease nonlinearly until it reaches the minimum amount of space required by the organism to survive.\\Note that the relationship between range size and both $\text{E}(R)$ and $\text{Var}(R)$ cannot be of the form $H = \beta_0 + \beta_1 \text{E}(R) + \beta_2 \text{Var}(R)$ because it would require range size to be negative for high values of $\text{E}(R)$ or low values of $\text{Var}(R)$.}\label{fig:hr-hyp}
\end{figure}

Overall, the hypothesis that range size decreases with resource abundance, \(\text{E}(R)\), is commonly accepted and well supported, but many studies assume a linear relationship {[}e.g., 21,41,52--54{]}. This is problematic because, conceptually, the relationship between range size and \(\text{E}(R)\) must be nonlinear, since: (1) there is an upper limit to how much space an organism is able to explore in its finite lifetime and (2) the minimum amount of space it requires to survive is necessarily greater than zero {[}see 27,28,55,56,57, and contrast them to the earlier references that assume a linear relationship between \(H\) and \(R\){]}. Consequently, we suggest analysts use models that account for this nonlinearity when estimating the effects of resource abundance on range size. While the relationship may be approximately linear for some range of \(\text{E}(R)\), this assumption often does not hold for low or high values of \(\text{E}(R)\) {[}e.g., 52{]}. Additionally, identifying inflection points in nonlinear relationships can help understand the pressures and limitations of increasing range size.

\subsection{\texorpdfstring{Effects of resource stochasticity, \(\text{Var}(R)\)}{Effects of resource stochasticity, \textbackslash text\{Var\}(R)}}\label{effects-of-resource-stochasticity-textvarr}

\noindent Assuming resource stochasticity is constant over time and space can be a useful simplification of relatively stable environments or when information on how \(\text{E}(R)\) changes is limited and estimating changes in \(\text{Var}(R)\) is unreasonable. However, such an assumption is likely not realistic, since \(\text{Var}(R)\) often differ across space and over time. Generally, bounded quantities have correlated means and variances, as in the case of random variables that are strictly positive (e.g., Gamma and Poisson) or fully bounded (e.g., Beta). For example, prey abundance in a given area over time may approximately follow a Poisson distribution, which implies that the mean and variance will be approximately equal. When prey are scarce, the variance will also be low, and when prey are abundant the variance will also be high. This occurs because the behavior, fitness, and predator-prey dynamics of many prey are more stochastic than those of few prey {[}58{]}. Similarly, in the case of fully bounded random variables, the variance is generally lowest when the mean is near either boundary. For example, successful predation events are predictably scarce if the probability of capture is near 0, predictably common if the probability is near 1, and most stochastic if the probability is near 0.5 {[}i.e., as far as possible from both 0 and 1; see {[}59{]}{]}. See Appendix A for more information.

Recognizing changes in \(\text{Var}(R)\) helps account for the residual, fine-scale variation in \(R\) after accounting for trends in the large-scale average \(R\) {[}e.g., variations in plant phenology between years after accounting for mean seasonal trends, see 60{]}. However, when both \(\text{E}(R)\) and \(\text{Var}(R)\) change over time (fig.~A2), disentangling changes in \(\text{E}(R)\) and \(\text{Var}(R)\) is not simple {[}61{]}. Statistically, this confound occurs because the more change one attributes to \(\mu(t, \vec u)\) (i.e., the wigglier it is), the smaller \(\sigma^2(t, \vec u)\) becomes. Conversely, the smoother \(\mu(t, \vec u)\) is, the larger \(\sigma^2(t, \vec u)\) becomes. Biologically, it is important because an organism's perception scale determines whether it attributes a change in \(R\) to a trend in \(\text{E}(R)\) or as a stochastic event {[}i.e., due to \(\text{Var}(R)\); see {[}60{]}{]}. An organism's perception of changes in \(R\) will also depend strongly on the its cognitive capacities and memory {[}9,62--65{]}. Whether an organism is able to predict trends in \(\sigma^2(t, \vec u)\) or not, environmental variability is thought to reduce a landscape's energetic balance {[}26{]}, which, in turn, decreases organisms' fitness {[}e.g., 10{]} and increases their range size. While this behavioral response occurs with both predictable and unpredictable stochasticity, extreme and rare events are more likely to have a stronger effect due to their unpredictability and magnitude {[}66,67{]}. A few recent studies support these hypotheses {[}22,26,31,48,68{]}, but many of them are limited in geographic and taxonomic scales or fail to account for nonlinear relationships, so the extent to which these preliminary findings can be generalized is currently unknown. Thus, there remains a need for developing a more complete understanding of how organisms' range sizes changes with environmental stochasticity.

Similarly to \(\text{E}(R)\), we hypothesize \(\text{Var}(R)\) has a nonlinear effect on an organism's range size. When \(\text{Var}(R)\) is low enough that \(R\) is relatively predictable, we expect organisms to be range-resident with small home ranges, and we do not expect small changes in \(\text{Var}(R)\) to have a noticeable effect. As resources become increasingly unpredictable, we expect home range size to increase progressively faster (fig.~\ref{fig:hr-hyp}B) because: (1) as \(\text{Var}(R)\) increases, the chances of finding low \(R\) increase superlinearly, (2) the added movement required to search for food increases organisms' energetic requirements, and (3) stochasticity reduces an organism's ability to specialize and reduce competition for \(R\) {[}69{]}. If resources remain highly unpredictable over long periods of time (e.g., multiple lifespans), organisms may evolve or develop new and consistent behaviors (e.g, nomadism) or adaptations (e.g., increased fat storage or food caching) to buffer themselves against times of unpredictably low \(R\). Conversely, if changes in \(\sigma^2(t, \vec u)\) are sufficiently predictable, organisms may learn to anticipate and prepare for times of greater stochasticity by pre-preemptively caching food, reducing energetic needs, migrating, or relying on alternative food sources {[}e.g., 70{]}.

\subsection{\texorpdfstring{Interactive effects of \(\text{E}(R)\) and \(\text{Var}(R)\)}{Interactive effects of \textbackslash text\{E\}(R) and \textbackslash text\{Var\}(R)}}\label{interactive-effects-of-texter-and-textvarr}

\noindent We have provided the case for why both \(\text{E}(R)\) and \(\text{Var}(R)\) should be expected to affect organisms' range size, but we presented the two parameters as independent drivers of movement. However, organisms may respond to changes in \(\sigma^2(t, \vec u)\) more when resources are scarce than when they are abundant. Consequently, an organism's movement behavior is likely to be a function of not only the marginal effects of \(\text{E}(R)\) and \(\text{Var}(R)\) but also their interactive effects. A highly unpredictable habitat may be very inhospitable if resources are poor, but \(\text{Var}(R)\) may have little effect if resources are stochastic but always abundant. Thus, we expect \(\text{Var}(R)\) to have a stronger effect on range size when \(\text{E}(R)\) is low, and less of an effect when \(\text{E}(R)\) is high. We explore this interaction effect more in the following section.

\subsection{\texorpdfstring{Simulating responses to \(\text{E}(R)\) and \(\text{Var}(R)\)}{Simulating responses to \textbackslash text\{E\}(R) and \textbackslash text\{Var\}(R)}}\label{sims}

\noindent To evaluate our hypothesis of how organisms' range sizes are affected by \(\text{E}(R)\), \(\text{Var}(R)\), and the interaction effect of \(\text{E}(R)\) and \(\text{Var}(R)\), we present the results from a series of quantitative simulations. To start, we used the \texttt{ctmm} package {[}71{]} for \texttt{R} {[}72{]} to generate 200 tracks (see Appendix B for sensitivity analyses) from an Integrated Ornstein-Uhlenbeck movement model {[}IOU model, see 73{]}. The IOU model's correlated velocity produced tracks with directional persistence, but, unlike Ornstein-Uhlenbeck (OU) and Ornstein-Uhlenbeck Foraging (OUF) models, IOU models do not produce spatially stationary movement, so the organism is not range-resident. Consequently, each track is spatially unrestricted and can be interpreted as purely exploratory or memoryless movement.

Each of the 200 tracks were placed on a grid with common starting point \(\langle 0, 0\rangle\) (fig.~B1). Each time the simulated individual moved to a new cell, it collected \(R\) resources sampled from a Gamma distribution. The mean and variance of the distribution were defined by a series of deterministic functions \(\mu(t)\) and \(\sigma^2(t)\) (orange and blue lines in fig.~\ref{fig:5-5-sims}). The value of \(t\) was constant within each set of 200 tracks, so the distribution \(R\) was sampled from was independent of both the organism's location and its time spent moving. Tracks were truncated once the organism reached satiety, and the organism was given enough time to return to \(\langle 0, 0\rangle\) independently from the following track (section 2.1 of Appendix B). Finally, we fit an OUF movement model {[}74{]} to the set of tracks to calculate the 95\% Gaussian home-range size using the formula \[\hat{H}_{95\%} = -2 \log(1 - 0.95) \pi \hat \varsigma^2,\] where \(\hat \varsigma^2\) is the positional variance estimated by the movement model.

We designed the simulations to estimate the effects of \(\text{E}(R)\) and \(\text{Var}(R)\) in simplistic environments where organisms could only respond by searching for longer periods of time. Consequently, we made the following assumptions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  Environments are homogeneous for a given \(t\). Given \(t\), \(\text{E}(R) = \mu(t)\) and \(\text{Var}(R) = \sigma^2(t)\) are constant over space and within each set of 200 tracks, but \(R\) is random and follows a \(\text{Gamma}\large(\mu(t), \sigma^2(t)\large)\) distribution.
\item
  The are no external pressures on the simulated organism. Resources do not deplete, and there is no competition nor predator avoidance.
\item
  The organism has a fixed daily energetic requirement that is independent of movement rates, and it cannot alter its metabolism or physiology. Additionally, the organism does not have energetic reserves, so excess resources cannot be carried over to the next track or \(t\).
\item
  The organism is range-resident and can only respond to changes in \(\text{E}(R)\) and \(\text{Var}(R)\) by altering its home-range size. The organism does not disperse or abandon a range.
\item
  The organism's movement is simplistic. The organism's movement speed and direction are stochastic and independent of \(\text{E}(R)\) and \(\text{Var}(R)\).
\item
  The organism has no perceptive range or memory. It is unable to detect, learn, or predict where resources are abundant (high \(\text{E}(R)\)) or reliable (low \(\text{Var}(R)\)) over time or space.
\item
  Animals only move to search for food or return to the center of their home-range after reaching satiety.
\end{enumerate}

Based on the assumptions above, we constructed the following causal model for the simulated effects of \(\text{E}(R)\) and \(\text{Var}(R)\) on \(H\) {[}see fig.~\ref{fig:dag-sims} and 75{]}: \(\text{E}(R)\) and \(\text{Var}(R)\) were determined independently of each other, but they jointly determined the distribution of \(R\), which, in turn, determined the distribution of \(H\). Additional information is provided in Appendix B.

\begin{figure}

{\centering \includegraphics[width=0.5\linewidth]{manuscript_files/figure-latex/dag-sims-1}

}

\caption{Directed acyclical graph assumed for inferring the causal effects of $\text{E}(R)$ and $\text{Var}(R)$ on the distributions of $R$ and $H$ in the simulations.}\label{fig:dag-sims}
\end{figure}

Fig. \ref{fig:5-5-sims} shows how simulated home-range size, \(H\), responded to changes in \(\mu(t)\) and \(\sigma^2(t)\) in scenarios where both functions can remain constant, increase linearly, oscillate cyclically, drift stochastically, or change erratically. The top row (constant \(\text{Var}(R)\)) shows how \(H\) varies for different trends in \(\mu(t)\) while \(\text{Var}(R)\) remains constant (like in fig.~A1). As \(\text{E}(R)\) increases at a constant slope (linear \(\mu(t)\)), \(H\) decreases nonlinearly, with larger changes when \(\text{E}(R)\) is low, until it approaches the minimum size required by the organism. Also note how the noise in the green lines also decreases as \(\text{E}(R)\) increases.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figures/mean-variance-5-by-5-hr-sims}

}

\caption{Simulated home-range sizes, $H$, of an organism living in habitats where the mean and variance in resources are constant, linearly increasing, cyclical, drifting, or erratic over time (but homogeneous over space for a given $t$). Note how $H$ decreases nonlinearly as $\mu(t)$ increases and increases nonlinearly as $\sigma^2(t)$ increases. Additionally, the variance in $H$ is higher when $\mu(t)$ is lower or $\sigma^2(t)$ is higher, and changes in $\sigma^2(t)$ have greater impacts when $\mu(t)$ is low.}\label{fig:5-5-sims}
\end{figure}

The leftmost column of fig.~\ref{fig:5-5-sims} (constant \(\text{E}(R)\)) illustrates the effects of \(\text{Var}(R)\) on \(H\) while \(\text{E}(R)\) remains constant. Overall, both mean \(H\) and the variance around it increase with \(\sigma^2(t)\) (most visible with constant \(\text{E}(R)\) and linear \(\text{Var}(R)\)). Similarly to resource-poor periods, times of greater stochasticity require the organism to move over larger areas for longer periods of time. Additionally, the greater in uncertainty in how much time and space the organism will require to reach satiety, or indeed whether an organism living in highly stochastic environments can even reach satiety within a finite amount of time.

The remaining panels in fig.~\ref{fig:5-5-sims} illustrate how \(\text{E}(R)\) and \(\text{Var}(R)\) jointly affect \(H\) and how unintuitive the effects can be. Since \(\text{E}(R)\) and \(\text{Var}(R)\) have opposite effects on \(H\), disentangling the effects can be particularly difficult when both parameters change in a correlated manner (e.g., linear \(\text{E}(R)\) and \(\text{Var}(R)\)). When both \(\text{E}(R)\) and \(\text{Var}(R)\) increase linearly, \(H\) initially increases since the effect of \(\text{Var}(R)\) is stronger, but then decreases as the effect of \(\text{E}(R)\) begins to dominate. Difficulties in disentangling the two effects are explored in greater depth in the case study in the following section.

Although the temporal trends in fig.~\ref{fig:5-5-sims} are complex and the effects of \(\text{E}(R)\) and \(\text{Var}(R)\) can be hard to disentangle, two simple relationships emerge when \(H\) is shown as a function of either \(\text{E}(R)\) or \(\text{Var}(R)\), rather than time: \(H\) decreases nonlinearly with \(\text{E}(R)\) and increases with \(\text{Var}(R)\) (panels A and B of fig.~\ref{fig:5-5-reg}). The estimated relationships thus follow the hypothesis we presented in fig.~\ref{fig:hr-hyp}, although we found that the effect of \(\text{Var}(R)\) at average \(\text{E}(R)\) was linear with a slight sublinear saturation at high values of \(\text{Var}(R)\). However, notice that the effect of \(\text{Var}(R)\) on \(E(H)\) depends strongly on \(\text{E}(R)\) (panel C): When \(\text{E}(R)\) is low, \(\text{E}(H)\) is high and \(\text{Var}(R)\) does not have a strong effect, but when \(\text{E}(R)\) is high the effect of \(\text{Var}(R)\) on \(\text{E}(H)\) is exponential. Similarly, \(\text{E}(H)\) decreases exponentially with \(\text{E}(R)\) except when \(\text{Var}(R)\) is very high.

As expected by the changes in the spread of the points in panels A and B of fig.~\ref{fig:5-5-reg}, the variance in \(H\), \(\text{Var}(H)\), also depends on \(\text{E}(R)\) and \(\text{Var}(R)\) (fig.~\ref{fig:5-5-reg}D-F). Since we modeled \(H\) using a Gamma family of distributions, we expected \(\text{Var}(H)\) to increase with \(\text{E}(H)\), but the location-scale model removes the assumption of a constant mean-variance relationship (i.e., constant coefficient of variation, \(\frac{\mu(t)}{\sigma^2(t)}\). This allowed us to show that the effect of \(R\) on \(\text{Var}(H)\) is much stronger than the effect of \(R\) on \(\text{E}(H)\). Consequences of these effects are explored in the discussion section.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figures/simulation-regression-mean-and-var}

}

\caption{Effects of $\text{E}(R)$ and $\text{Var}(R)$ on on the mean (A-C) and variance (D-F) in simulated home-range size with 95\% Bayesian credible intervals. While the estimated marginal effect of $\text{Var}(R)$ on $\text{E}(H)$ is sublinear (panel B), the effect of $\text{Var}(R)$ is superlinear for high values of $\text{E}(R)$ (panel C). The relationships were estimated using a Generalized Additive Model for Location and Scale with a Gamma location-scale family of distributions ($\tt{mgcv\!::\!gammals}$). Credible intervals were calculated using 10,000 samples from the posterior distribution while assuming multivariate Gaussian coefficients. Additional details on the model structure are provided in Appendix B.}\label{fig:5-5-reg}
\end{figure}

\section{A case study on a lowland tapir in the Brazilian Cerrado}\label{a-case-study-on-a-lowland-tapir-in-the-brazilian-cerrado}

\noindent The simulations in the section above support the hypothesis we presented in the background section, but they are based on assumptions that are often not met in real natural environments. Organisms live in spatiotemporally heterogeneous and dynamic environments that promote the use of perceptual ranges, navigation, and memory. Together, these abilities result in selective space use that depends on resource availability {[}14{]} and resource depletion {[}15{]}.

In this section, we test the hypothesis using empirical tracking data on a lowland tapir from the Brazilian Cerrado along with empirical estimates of \(\text{E}(R)\) and \(\text{Var}(R)\). We measure \(R\) using Normalized Difference Vegetation Index {[}NDVI, see 76{]}, a remote-sensed measure of landscape greenness, as a proxy for forage abundance. Appendix C contains additional information on how we modeled NDVI and the tapir's movement using continuous-time movement models {[}71,77{]} and autocorrelated kernel density estimation {[}78--80{]}.

Fig. \ref{fig:tapir-mw} illustrates how a tapir in the Brazilian Cerrado adapted its 7-day home-range size to spatiotemporal changes in estimated \(\mu(t, \vec u)\) and \(\sigma^2(t, \vec u)\) (telemetry data from the individual labelled as ``Anna'' in the dataset from {[}29{]}). Panels A and B show the changes in seven-day average mean and variance in NDVI, respectively, experienced by the tapir during the tracking period. The mean and variance in NDVI were estimated using a Generalized Additive Model for Location and Scale {[}GAMLS, 81{]} with a Beta family of distributions (NDVI values ranged from 0.3534 to 0.9475). Panel C shows the changes in the tapir's 7-day home range over time. All 457 of the 7-day windows had a minimum effective sample size of 7 range crossings {[}range: 7.7 -- 69.6, see 82{]}, and 92\% had resolvable (i.e., non-NA) home range crossing times, all of which were below 17 hours. Note how the tapir uses more space during periods of lower NDVI (e.g., August 2017) and less space during periods with high NDVI (January 2018). Additionally, when resources are scarce and highly unpredictable (August 2018), the tapir uses up to 5 times more space than when resources are abundant and predictable (e.g., January 2018). Finally, panels D and E show the estimated (marginal) effects of \(\hat\mu(t, \vec u)\) and \(\hat\sigma^2(t, \vec u)\) on the tapir's 7-day home-range size. Since \(\hat\mu(t, \vec u)\) and \(\hat\sigma^2(t, \vec u)\) are correlated (panel F) and spatiotemporally autocorrelated (panels A, B, and F), the effects of \(R\) on \(H\) should be modeled carefully. To avoid over-fitting the model, we constrained the smooth effects of \(\hat\mu(t, \vec u)\) and \(\hat\sigma^2(t, \vec u)\) and their interaction effect to a small basis size (\(\tt{k = 3}\)). Additional information is provided in appendix C. The results presented in panels D-F of fig.~\ref{fig:tapir-mw} match our findings from the simulations (fig.~\ref{fig:5-5-reg}A-C): The tapir's 7-day home range decreases with \(\hat\mu(t, \vec u)\) and increases with \(\hat\sigma^2(t, \vec u)\), and the effect of \(\hat\mu(t, \vec u)\) depends on \(\hat\sigma^2(t, \vec u)\), and vice-versa. Alone, \(\hat\mu(t, \vec u)\) and \(\hat\sigma^2(t, \vec u)\) cause the tapir to double her home range (panels D and E), but together they result in an approximate 15-fold change in home-range size (observed range: 0.8 to 12.4 km\textsuperscript{2}; see panel F). Additionally, note how high NDVI values (\(\hat \mu(t, \vec u) > 0.8\)) cause \(\hat\sigma^2(t, \vec u)\) to have little to no effect on home-range size, as indicated by the vertical contour line in panel F. Similar conclusions can be drawn for the animal's diffusion (i.e., area covered per unit time), which is a more appropriate measure of space use when animals are not range resident {[}82{]}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{../figures/tapir-example}

}

\caption{Effects of estimated $\mu(t, \vec u)$ and $\sigma^2(t, \vec u)$ on the home-range size of a lowland tapir (\emph{Tapirus terrestris}). (A) Trends in resource abundance over time, $\hat\mu(t, \vec u)$, estimated as the average mean NDVI at the locations visited by the tapir during a seven-day period. (B) Variance in resources over time, $\hat\sigma^2(t, \vec u)$, estimated as the average variance in NDVI at the locations visited by the tapir during a seven-day period. (C) Seven-day 95\% home range estimated using Autocorrelated Kernel Density Estimation. (D, E) Estimated marginal effects of $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$ on home-range size. The model accounted for the marginal effects of $\hat\mu(t, \vec u)$, $\hat\sigma^2(t, \vec u)$, and their interaction effect. (F) Estimated home-range size in response to changes in both $\hat\mu(t, \vec u)$ and $\hat\sigma^2(t, \vec u)$. Note how the effect of $\hat\sigma^2(t, \vec u)$ is more pronounced when $\hat\mu(t, \vec u)$ is low. See Appendix C for additional information. The tapir movement data corresponds to the individual named ``Anna" from the Cerrado sample of Medici \emph{et al.} [29].}\label{fig:tapir-mw}
\end{figure}

Quantifying the direct effects of \(\text{E}(R)\) and \(\text{Var}(R)\) on \(H\) using empirical data is more complex than with simulated data, and it requires a different causal framework, particularly in the case of observational studies (as opposed to experimentally-controlled studies; see fig.~\ref{fig:dag-empirical}). Unlike with the simulations, \(\text{E}(R)\) and \(\text{Var}(R)\) are not controlled variables and instead depend on the distribution of \(R\), which depends on a variety of other factors (that we exclude from the figure for simplicity). Both \(\text{E}(R)\) and \(\text{Var}(R)\) then impact \(H\) as well as habitat-level variables (e.g., competition, predation, etc.; indicated as \(Z\)) that also affect \(H\). Additionally, estimating \(R\) via a proxy (NDVI) adds satellite-level noise and confounds {[}e.g., saturation, cloud cover, spatiotemporal averaging -- indicated as \(S\), see 83,84,85{]}. However, \(\text{E}(R)\) and \(\text{Var}(R)\) can be correlated to \(\text{E}(\text{NDVI})\) and \(\text{Var}(\text{NDVI})\), respectively, provided that analysts use models that are sufficiently smooth and flexible at the relevant spatiotemporal scale {[}86{]}. We discuss this in further detail in the section below on the strengths and limitations of the empirical approach.

\begin{figure}

{\centering \includegraphics[width=0.75\linewidth]{manuscript_files/figure-latex/dag-empirical-1}

}

\caption{Directed Acyclical Graph assumed for inferring the causal effects of $\text{E}(R)$ and $\text{Var}(R)$ on $H$, where NDVI was used as a proxy for $R$. $Z$ and $S$ indicate confounds that result from habitat-level variables (e.g., competition, predation, etc.) and satellite-level variables (e.g., noise, cloud cover).}\label{fig:dag-empirical}
\end{figure}

\section{Discussion}\label{discussion}

\noindent The amount of space organisms use is determined by a multitude of factors {[}16{]}, but the search for resources is often a main driver of how much and where organisms move. This paper builds on earlier theoretical work {[}13,e.g., 18,19{]} and presents a unifying hypothesis that describes the effects of resource abundance and stochasticity on organisms' range sizes. We use quantitative simulations and an empirical case study to support the hypothesis and show that it provides a simple framework for understanding how motile organisms adapt their movement in dynamic environments. Separately, resource abundance and stochasticity have simple but opposing effects on organisms' range sizes: \(H\) decreases with \(\text{E}(R)\) and increases with \(\text{Var}(R)\). Together, the degree to which \(\text{E}(R)\) affects \(H\) depends on \(\text{Var}(R)\), and vice-versa, so organisms' responses to resource dynamics can be complex. The simulated and empirical results suggest qualitatively similar marginal effects of \(\text{E}(R)\) and \(\text{Var}(R)\), but there are differences in the estimated interactive effects. In the simulated data, \(\text{Var}(R)\) has little effect when \(\text{E}(R)\) is low and a strong effect when \(\text{E}(R)\) is high, while the opposite is true for the empirical data. This difference is due to two reasons. Firstly, the shape and symmetry of bounded distributions such as Gamma (\(R > 0\)) and Beta (\(0 < R < 1\)) distributions depend on both \(\text{E}(R)\) and \(\text{Var}(R)\) (figs. A3, A4), but \(\text{Var}(R)\) does not affect the shape of a Gamma distribution as much if \(\text{E}(R)\) is low (fig.~B3). Secondly, and perhaps more interestingly, the simulation approach does not account for real-world adaptations to \(\text{E}(R)\) and \(\text{Var}(R)\) such as selective space use, which are included (but not explicitly accounted for) in the empirical approach. Below we discuss the strengths and limitations of each approach.

\subsection{Strengths and limitations of the simulation-based approach}\label{strengths-and-limitations-of-the-simulation-based-approach}

\noindent Our simulations are based on a simplistic environment with many assumptions that allowed us to estimate how resource abundance and stochasticity affect organisms' home-range sizes if organisms can only respond to changes by adapting the amount of time spent searching for food (with no energetic cost to movement). The use of continuous-time movement models coupled with few drivers of movement supported realistic data that could be explained by straightforward causal models. The absence of confounding variables (e.g., predator avoidance, territoriality, competition, landscape connectivity; see fig.~\ref{fig:dag-sims}) or sample size limitation allowed us to ensure estimates were accurate and robust (sensitivity analysis available in Appendix B).

Deviations from the simulations offer a means of detecting when the underlying assumptions are inappropriate and how additional factors may affect organisms' responses to changes in \(\text{E}(R)\) and \(\text{Var}(R)\). For example, energetic costs of movement are often non-negligible and depend on organism size {[}40{]}, movement speed {[}40{]}, and ambient temperature {[}1,87{]}. In addition, an organism may alter its movement behavior, physiology, and energetic needs to buffer itself against changes in \(\text{E}(R)\) and \(\text{Var}(R)\) by using space selectively {[}68,88--90{]} and adapting their behavior and physiology over time {[}18,69{]}. Before or during periods of scarcity, organisms may cache resources {[}91{]}, build up fat reserves {[}45{]}, enter states of dormancy {[}92--94{]}, or even pause fetal growth {[}7{]}. However, organisms may be unable to respond to changes in \(\text{E}(R)\) and \(\text{Var}(R)\) optimally due to various reasons, including limited perceptive range {[}61{]}, lack of experience {[}9,47,63--65,95{]}, avoidance of competitors and predators {[}14,96{]}, or a physiology that is not amenable to things like hibernation or fat storage. Thus, organisms may relocate their range to a sub-optimal location {[}33,34,97,98{]}, which may exacerbate the effects of \(\text{E}(R)\) and \(\text{Var}(R)\) on both mean range size and the variance around it.

\subsection{Strengths and limitations of the empirical approach}\label{strengths-and-limitations-of-the-empirical-approach}

\noindent There are two main advantages of taking an empirical approach. Firstly, modeling real-world animal movement data can produce scale-appropriate and easily interpretable estimates. Secondly, empirical data contain information on the effects of \(\text{E}(R)\), \(\text{Var}(R)\), and confounding variables without having to design complex and time-consuming simulations. However, it is not always possible to quantify confounding variables. For example, while there may be some appropriate proxies of competition, such as density of competitors, these variables may be hard to quantify, and they may not account for the confounding effects appropriately (i.e., the presence of competitors may not reflect competitive pressure). This is problematic if one is interested in estimating the direct causal effect of \(\text{E}(R)\) and \(\text{Var}(R)\), which requires removing any non-negligible confounding effects {[}75{]}.

Similarly, if \(R\) non-measurable (as is often the case), \(R\) must be estimated with proxies such as NDVI {[}76{]}, which may introduce complexities. While \(R\) and NDVI are correlated for many species {[}e.g., 45,46,95,99--101{]}, the relationship between the two can be weak {[}84{]}, satellite-dependent {[}85{]}, and nonlinear {[}83,85{]}. This complexity can introduce two sources of bias: ecosystem-level biases (indicated as \(Z\) in the directed acyclical graph in fig.~\ref{fig:dag-empirical}) and satellite-level confounding variables (\(S\) in fig.~\ref{fig:dag-empirical}). Examples of ecosystem-level biases are the effects of competition, predation, habitat connectivity, and movement costs, all of which can depend on habitat quality, and, consequently, be correlated nonlinearly to \(R\) and NDVI {[}35,102{]}. Resource-rich patches can attract larger amounts of competitors {[}14{]} and predators {[}20{]}, which may, in turn, increase pressures from competition and predation {[}15,39{]}. However, such pressures may result in both an expansion of the range {[}35,102{]} or a contraction, since larger ranges can be harder to defend and result in higher movement costs {[}35,103{]} and encounter rates {[}104{]}. Satellite-level confounds include information loss due to coarse spatiotemporal resolution {[}83,85{]}, satellite-level error {[}83,85,105{]}, and other limitations of remote sensing (e.g., inability to quantify specific resources or small-scale resource depletion). However, nonlinear models such as Generalized Additive Models {[}106{]} can help account for preferences for intermediate values of remotely-sensed \(R\) {[}e.g., young grass rather than mature grasslands, see 85{]}.

\section{Conclusions}\label{conclusions}

\noindent The work presented here provides a unifying framework for viewing movement as a response to resource abundance and stochasticity. We provide a sensible and unifying hypothesis of the effects of \(\text{E}(R)\) and \(\text{Var}(R)\) on organisms' range sizes and movement behavior. We demonstrate that organisms' range sizes decrease with resource abundance, increase with resource stochasticity, and that the effects of \(\text{Var}(R)\) can depend strongly on \(\text{E}(R)\).

Recent advances in computational power have greatly increased analysts' ability to fit computationally demanding models {[}107,108{]} that allow biologists to move beyond only considering changes in mean conditions. By accounting for changes in stochasticity, we can start developing a more comprehensive understanding of how organisms adapt to the dynamic environments organisms live in, including recent changes in climate {[}109{]} and increases in the frequency and intensity of extreme events {[}66,67,110--112{]}.

\section{List of abbreviations}\label{list-of-abbreviations}

\begin{tabular}{l|l}
\hline
\textbf{Abbreviation} & \textbf{Definition}\\
\hline
$H$ & Range size\\
\hline
$\hat H_{95\%}$ & Estimated $95\%$ home range size\\
\hline
$C$ & Resource consumption rate\\
\hline
$R$ & Resources\\
\hline
$t$ & Moment in time\\
\hline
$\vec u$ & Location in space (vector of coordinates)\\
\hline
$\text{E}(R)$ & Resource abundance\\
\hline
$\mu(t)$ & Resource abundance as a function of time\\
\hline
$\mu(t, \vec u)$ & Resource abundance as a function of time and space\\
\hline
$\text{Var}(R)$ & Resource stochasticity\\
\hline
$\sigma^2(t)$ & Resource stochasticity as a function of time\\
\hline
$\sigma^2(t, \vec u)$ & Resource stochasticity as a function of time and space\\
\hline
$\hat{\varsigma^2}$ & Estimated positional variance\\
\hline
$\Gamma(\mu, \sigma^2)$ & Gamma distribution with mean $\mu$ and variance $\sigma^2$\\
\hline
NDVI & Normalized Difference Vegetation Index\\
\hline
GAMLS & Generalized Additive Model for Location and Scale\\
\hline
\end{tabular}

\section{Declarations}\label{declarations}

\subsection{Ethics approval and consent to participate}\label{ethics-approval-and-consent-to-participate}

\noindent Not applicable.

\subsection{Consent for publication}\label{consent-for-publication}

\noindent Not applicable.

\subsection{Availability of data and materials}\label{availability-of-data-and-materials}

\noindent All code and data used for this manuscript is available on GitHub at \url{https://github.com/QuantitativeEcologyLab/hr-resource-stoch}, with the exception of two simulated datasets that were greater than 100 MB and the tapir data. The simulated data can be produced by running the scripts in the repository, while the tapir data is available at \url{https://github.com/StefanoMezzini/tapirs}.

\subsection{Competing interests}\label{competing-interests}

\noindent The authors declare that they have no competing interests.

\subsection{Funding}\label{funding}

\noindent SM was supported by funding from the University of British Columbia Okanagan, the Canadian Foundation for Innovation, BC Parks Living Labs, and MITACS. MJN was supported by NSERC Discovery Grant RGPIN-2021-02758 and the Canadian Foundation for Innovation. CHF was supported by NSF IIBR 1915347.

\subsection{Authors' contributions}\label{authors-contributions}

\noindent SM performed the literature review, designed the simulations, analyzed the data, and wrote the manuscript. CHF contributed to the analyses. EPM provided the tapir telemetry data. MJN conceived the project idea and provided support throughout the analyses. All authors contributed to the writing and read and approved the final manuscript.

\subsection{Acknowledgements}\label{acknowledgements}

\noindent We would like to thank Dr.~Simon Wood for providing code to fit a Beta location-scale GAM despite not being involved directly with the project. Additionally, we thank all those who provided feedback on all posters, presentation, and writings related to this project. In particular, we thank all those who provided feedback on the manuscript and appendices despite not being authors, namely, in alphabetical order by first name: Aimee Chhen, Jessa Marley, Kim Hinz, Lauren Mills, Sarah Wyse, and Dr.~Simon Wood.

\subsection{Authors' information}\label{authors-information}

\noindent See page 1 for affiliations.

\noindent Authors' emails:

\noindent SM: \(\text{stefano.mezzini@ubc.ca}\)

\noindent CHF: \(\text{christen.fleming@ucf.edu}\)

\noindent EPM: \(\text{medici@ipe.org.br}\)

\noindent MJN: \(\text{michael.noonan@ubc.ca}\)

\clearpage

\newpage

\section{References}\label{references}

\hangparas{1em}{1}

\phantomsection\label{refs}
\begin{CSLReferences}{0}{1}
\bibitem[\citeproctext]{ref-hou_cold_2020}
1. Hou R, Chapman CA, Jay O, Guo S, Li B, Raubenheimer D. Cold and hungry: Combined effects of low temperature and resource scarcity on an edge‐of‐range temperate primate, the golden snub‐nose monkey. Ecography {[}Internet{]}. 2020 {[}cited 2022 Oct 3{]};43:1672--82. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/ecog.05295}

\bibitem[\citeproctext]{ref-le_bot_fishery_2019}
2. Le Bot T, Lescroël A, Fort J, Péron C, Gimenez O, Provost P, et al. Fishery discards do not compensate natural prey shortage in northern gannets from the english channel. Biological Conservation {[}Internet{]}. 2019 {[}cited 2022 Oct 3{]};236:375--84. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0006320718310930}

\bibitem[\citeproctext]{ref-dai_pra_ground_2022}
3. Dai Pra R, Mohr SM, Merriman DK, Bagriantsev SN, Gracheva EO. Ground squirrels initiate sexual maturation during hibernation. Current Biology {[}Internet{]}. 2022 {[}cited 2022 Sep 2{]};32:1822--1828.e4. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0960982222002548}

\bibitem[\citeproctext]{ref-rocha_life_2021}
4. Rocha JL, Godinho R, Brito JC, Nielsen R. Life in deserts: The genetic basis of mammalian desert adaptation. Trends in Ecology \& Evolution {[}Internet{]}. 2021 {[}cited 2022 Sep 2{]};36:637--50. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0169534721000744}

\bibitem[\citeproctext]{ref-wessling_seasonal_2018}
5. Wessling EG, Deschner T, Mundry R, Pruetz JD, Wittig RM, Kühl HS. Seasonal variation in physiology challenges the notion of chimpanzees (pan troglodytes verus) as a forest-adapted species. Front Ecol Evol {[}Internet{]}. 2018 {[}cited 2022 Sep 2{]};6:60. Available from: \url{http://journal.frontiersin.org/article/10.3389/fevo.2018.00060/full}

\bibitem[\citeproctext]{ref-stefanescu_timing_2021}
6. Stefanescu C, Ubach A, Wiklund C. Timing of mating, reproductive status and resource availability in relation to migration in the painted lady butterfly. Animal Behaviour {[}Internet{]}. 2021 {[}cited 2022 Sep 2{]};172:145--53. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0003347220303742}

\bibitem[\citeproctext]{ref-schmidt_interplay_2020}
7. Schmidt NM, Grøndahl C, Evans AL, Desforges J-P, Blake J, Hansen LH, et al. On the interplay between hypothermia and reproduction in a high arctic ungulate. Sci Rep {[}Internet{]}. 2020 {[}cited 2022 Sep 2{]};10:1514. Available from: \url{http://www.nature.com/articles/s41598-020-58298-8}

\bibitem[\citeproctext]{ref-douglas_relative_2014}
8. Douglas DJT, Pearce-Higgins JW. Relative importance of prey abundance and habitat structure as drivers of shorebird breeding success and abundance: Drivers of shorebird breeding success and abundance. Anim Conserv {[}Internet{]}. 2014 {[}cited 2022 Nov 8{]};17:535--43. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/acv.12119}

\bibitem[\citeproctext]{ref-foley_severe_2008}
9. Foley C, Pettorelli N, Foley L. Severe drought and calf survival in elephants. Biology Letters {[}Internet{]}. 2008 {[}cited 2020 Feb 12{]};4:541--4. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rsbl.2008.0370}

\bibitem[\citeproctext]{ref-berger_climate_2018}
10. Berger J, Hartway C, Gruzdev A, Johnson M. Climate degradation and extreme icing events constrain life in cold-adapted mammals. Scientific Reports {[}Internet{]}. 2018 {[}cited 2020 Jan 24{]};8:1156. Available from: \url{http://www.nature.com/articles/s41598-018-19416-9}

\bibitem[\citeproctext]{ref-van_haastert_food_2009}
11. Van Haastert PJM, Bosgraaf L. Food searching strategy of amoeboid cells by starvation induced run length extension. Dornhaus A, editor. {PLoS} {ONE} {[}Internet{]}. 2009 {[}cited 2025 Feb 6{]};4:e6814. Available from: \url{https://dx.plos.org/10.1371/journal.pone.0006814}

\bibitem[\citeproctext]{ref-taub_root_1996}
12. Taub DR, Goldberg D. Root system topology of plants from habitats differing in soil resource availability. Functional Ecology {[}Internet{]}. 1996 {[}cited 2025 Feb 6{]};10:258. Available from: \url{https://www.jstor.org/stable/2389851?origin=crossref}

\bibitem[\citeproctext]{ref-harestad_home_1979}
13. Harestad AS, Bunnel FL. Home range and body weight--a reevaluation. Ecology {[}Internet{]}. 1979 {[}cited 2022 Sep 5{]};60:389--402. Available from: \url{http://doi.wiley.com/10.2307/1937667}

\bibitem[\citeproctext]{ref-kacelnik_ideal_1992}
14. Kacelnik A, Krebs JR, Bernstein C. The ideal free distribution and predator-prey populations. Trends in Ecology \& Evolution {[}Internet{]}. 1992 {[}cited 2024 Jan 31{]};7:50--5. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/016953479290106L}

\bibitem[\citeproctext]{ref-charnov_optimal_1976}
15. Charnov EL. Optimal foraging, the marginal value theorem. Theoretical Population Biology {[}Internet{]}. 1976 {[}cited 2024 Jan 31{]};9:129--36. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/004058097690040X}

\bibitem[\citeproctext]{ref-nathan_movement_2008}
16. Nathan R, Getz WM, Revilla E, Holyoak M, Kadmon R, Saltz D, et al. A movement ecology paradigm for unifying organismal movement research. Proc Natl Acad Sci USA {[}Internet{]}. 2008 {[}cited 2022 Mar 9{]};105:19052--9. Available from: \url{https://pnas.org/doi/full/10.1073/pnas.0800375105}

\bibitem[\citeproctext]{ref-burt_territoriality_1943}
17. Burt WH. Territoriality and home range concepts as applied to mammals. Journal of Mammalogy {[}Internet{]}. 1943 {[}cited 2022 Jan 31{]};24:346. Available from: \url{https://academic.oup.com/jmammal/article-lookup/doi/10.2307/1374834}

\bibitem[\citeproctext]{ref-southwood_habitat_1977}
18. Southwood TRE. Habitat, the templet for ecological strategies? The Journal of Animal Ecology {[}Internet{]}. 1977 {[}cited 2022 Feb 4{]};46:336. Available from: \url{https://www.jstor.org/stable/3817?origin=crossref}

\bibitem[\citeproctext]{ref-stephens_optimal_1982}
19. Stephens DW, Charnov EL. Optimal foraging: Some simple stochastic models. Behav Ecol Sociobiol {[}Internet{]}. 1982 {[}cited 2024 Jan 31{]};10:251--63. Available from: \url{http://link.springer.com/10.1007/BF00302814}

\bibitem[\citeproctext]{ref-duncan_lifehistory_2015}
20. Duncan C, Nilsen EB, Linnell JDC, Pettorelli N. Life‐history attributes and resource dynamics determine intraspecific home‐range sizes in carnivora. Remote Sens Ecol Conserv {[}Internet{]}. 2015 {[}cited 2024 May 29{]};1:39--50. Available from: \url{https://zslpublications.onlinelibrary.wiley.com/doi/10.1002/rse2.6}

\bibitem[\citeproctext]{ref-rizzuto_forage_2021}
21. Rizzuto M, Leroux SJ, Vander Wal E, Richmond IC, Heckford TR, Balluffi-Fry J, et al. Forage stoichiometry predicts the home range size of a small terrestrial herbivore. Oecologia {[}Internet{]}. 2021 {[}cited 2022 Mar 2{]};197:327--38. Available from: \url{https://link.springer.com/10.1007/s00442-021-04965-0}

\bibitem[\citeproctext]{ref-broekman_environmental_2024}
22. Broekman MJE, Hilbers JP, Hoeks S, Huijbregts MAJ, Schipper AM, Tucker MA. Environmental drivers of global variation in home range size of terrestrial and marine mammals. Journal of Animal Ecology {[}Internet{]}. 2024 {[}cited 2024 Jul 29{]};93:488--500. Available from: \url{https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2656.14073}

\bibitem[\citeproctext]{ref-singh_migration_2012}
23. Singh NJ, Börger L, Dettki H, Bunnefeld N, Ericsson G. From migration to nomadism: Movement variability in a northern ungulate across its latitudinal range. Ecological Applications {[}Internet{]}. 2012 {[}cited 2022 Nov 17{]};22:2007--20. Available from: \url{http://doi.wiley.com/10.1890/12-0245.1}

\bibitem[\citeproctext]{ref-wheat_migrate_2017}
24. Wheat RE, Lewis SB, Wang Y, Levi T, Wilmers CC. To migrate, stay put, or wander? Varied movement strategies in bald eagles (haliaeetus leucocephalus). Mov Ecol {[}Internet{]}. 2017 {[}cited 2022 Oct 17{]};5:9. Available from: \url{http://movementecologyjournal.biomedcentral.com/articles/10.1186/s40462-017-0102-4}

\bibitem[\citeproctext]{ref-teitelbaum_beyond_2019}
25. Teitelbaum CS, Mueller T. Beyond migration: Causes and consequences of nomadic animal movements. Trends in Ecology \& Evolution {[}Internet{]}. 2019 {[}cited 2022 Feb 1{]};34:569--81. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0169534719300527}

\bibitem[\citeproctext]{ref-chevin_adaptation_2010}
26. Chevin L-M, Lande R, Mace GM. Adaptation, plasticity, and extinction in a changing environment: Towards a predictive theory. Kingsolver JG, editor. {PLoS} Biology {[}Internet{]}. 2010 {[}cited 2020 Nov 11{]};8:e1000357. Available from: \url{https://dx.plos.org/10.1371/journal.pbio.1000357}

\bibitem[\citeproctext]{ref-herfindal_prey_2005}
27. Herfindal I, Linnell JDC, Odden J, Nilsen EB, Andersen R. Prey density, environmental productivity and home‐range size in the eurasian lynx ( \emph{lynx lynx} ). Journal of Zoology {[}Internet{]}. 2005 {[}cited 2022 Sep 23{]};265:63--71. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1017/S0952836904006053}

\bibitem[\citeproctext]{ref-nilsen_can_2005}
28. Nilsen EB, Herfindal I, Linnell JDC. Can intra-specific variation in carnivore home-range size be explained using remote-sensing estimates of environmental productivity? Écoscience {[}Internet{]}. 2005 {[}cited 2021 Nov 29{]};12:68--75. Available from: \url{https://www.tandfonline.com/doi/full/10.2980/i1195-6860-12-1-68.1}

\bibitem[\citeproctext]{ref-medici_movement_2022}
29. Medici EP, Mezzini S, Fleming CH, Calabrese JM, Noonan MJ. Movement ecology of vulnerable lowland tapirs between areas of varying human disturbance. Mov Ecol {[}Internet{]}. 2022 {[}cited 2022 Mar 14{]};10:14. Available from: \url{https://movementecologyjournal.biomedcentral.com/articles/10.1186/s40462-022-00313-w}

\bibitem[\citeproctext]{ref-lindstedt_seasonality_1985}
30. Lindstedt SL, Boyce MS. Seasonality, fasting endurance, and body size in mammals. The American Naturalist {[}Internet{]}. 1985 {[}cited 2022 Feb 4{]};125:873--8. Available from: \url{https://www.journals.uchicago.edu/doi/10.1086/284385}

\bibitem[\citeproctext]{ref-morellet_seasonality_2013}
31. Morellet N, Bonenfant C, Börger L, Ossi F, Cagnacci F, Heurich M, et al. Seasonality, weather and climate affect home range size in roe deer across a wide latitudinal gradient within europe. Coulson T, editor. Journal of Animal Ecology {[}Internet{]}. 2013 {[}cited 2020 Nov 11{]};82:1326--39. Available from: \url{http://doi.wiley.com/10.1111/1365-2656.12105}

\bibitem[\citeproctext]{ref-fjelldal_nightly_2021}
32. Fjelldal MA, Wright J, Stawski C. Nightly torpor use in response to weather conditions and individual state in an insectivorous bat. Oecologia {[}Internet{]}. 2021 {[}cited 2022 Oct 3{]};197:129--42. Available from: \url{https://link.springer.com/10.1007/s00442-021-05022-6}

\bibitem[\citeproctext]{ref-torrez-herrera_monkeys_2020}
33. Tórrez-Herrera LL, Davis GH, Crofoot MC. Do monkeys avoid areas of home range overlap because they are dangerous? A test of the risk hypothesis in white-faced capuchin monkeys (cebus capucinus). Int J Primatol {[}Internet{]}. 2020 {[}cited 2022 Mar 9{]};41:246--64. Available from: \url{http://link.springer.com/10.1007/s10764-019-00110-0}

\bibitem[\citeproctext]{ref-rich_anthropogenic_2012}
34. Rich LN, Mitchell MS, Gude JA, Sime CA. Anthropogenic mortality, intraspecific competition, and prey availability influence territory sizes of wolves in montana. J Mammal {[}Internet{]}. 2012 {[}cited 2022 Mar 10{]};93:722--31. Available from: \url{https://academic.oup.com/jmammal/article-lookup/doi/10.1644/11-MAMM-A-079.2}

\bibitem[\citeproctext]{ref-jetz_scaling_2004}
35. Jetz W, Carbone C, Fulford J, Brown JH. The scaling of animal space use. Science {[}Internet{]}. 2004 {[}cited 2022 Mar 3{]};306:266--8. Available from: \url{https://www.science.org/doi/10.1126/science.1102138}

\bibitem[\citeproctext]{ref-harvey_primate_1981}
36. Harvey PH, Clutton-Brock TH. Primate home-range size and metabolic needs. Behav Ecol Sociobiol {[}Internet{]}. 1981 {[}cited 2022 Nov 12{]};8:151--5. Available from: \url{http://link.springer.com/10.1007/BF00300828}

\bibitem[\citeproctext]{ref-baldwin_nutritional_1984}
37. Baldwin R, Bywater A. \href{https://doi.org/10.1146/annurev.nu.04.070184.000533}{Nutritional energetics of animals}. Annual review of nutrition. 1984;4:101--14.

\bibitem[\citeproctext]{ref-reich_body_2001}
38. Reich PB. Body size, geometry, longevity and metabolism: Do plant leaves behave like animal bodies? Trends in Ecology \& Evolution {[}Internet{]}. 2001 {[}cited 2022 Oct 17{]};16:674--80. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0169534701023060}

\bibitem[\citeproctext]{ref-brown_ecology_1999}
39. Brown JS, Laundre JW, Gurung M. The ecology of fear: Optimal foraging, game theory, and trophic interactions. Journal of Mammalogy {[}Internet{]}. 1999 {[}cited 2024 Jan 31{]};80:385--99. Available from: \url{https://academic.oup.com/jmammal/article-lookup/doi/10.2307/1383287}

\bibitem[\citeproctext]{ref-taylor_energetics_1982}
40. Taylor CR, Heglund NC, Maloiy GM. Energetics and mechanics of terrestrial locomotion. I. Metabolic energy consumption as a function of speed and body size in birds and mammals. Journal of Experimental Biology {[}Internet{]}. 1982 {[}cited 2022 Dec 12{]};97:1--21. Available from: \url{https://journals.biologists.com/jeb/article/97/1/1/34642/Energetics-and-mechanics-of-terrestrial-locomotion}

\bibitem[\citeproctext]{ref-relyea_home_2000}
41. Relyea RA, Lawrence RK, Demarais S. Home range of desert mule deer: Testing the body-size and habitat-productivity hypotheses. The Journal of Wildlife Management {[}Internet{]}. 2000 {[}cited 2021 Nov 29{]};64:146. Available from: \url{https://www.jstor.org/stable/3802984?origin=crossref}

\bibitem[\citeproctext]{ref-dawe_influence_2014}
42. Dawe KL, Bayne EM, Boutin S. Influence of climate and human land use on the distribution of white-tailed deer ( \emph{odocoileus} \emph{virginianus} ) in the western boreal forest. Canadian Journal of Zoology {[}Internet{]}. 2014 {[}cited 2020 Oct 23{]};92:353--63. Available from: \url{http://www.nrcresearchpress.com/doi/10.1139/cjz-2013-0262}

\bibitem[\citeproctext]{ref-berger-tal_invisible_2019}
43. Berger-Tal O, Saltz D. Invisible barriers: Anthropogenic impacts on inter- and intra-specific interactions as drivers of landscape-independent fragmentation. Phil Trans R Soc B {[}Internet{]}. 2019 {[}cited 2022 Aug 11{]};374:20180049. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rstb.2018.0049}

\bibitem[\citeproctext]{ref-samarra_movements_2017}
44. Samarra FIP, Tavares SB, Béesau J, Deecke VB, Fennell A, Miller PJO, et al. Movements and site fidelity of killer whales (orcinus orca) relative to seasonal and long-term shifts in herring (clupea harengus) distribution. Mar Biol {[}Internet{]}. 2017 {[}cited 2022 Nov 7{]};164:159. Available from: \url{http://link.springer.com/10.1007/s00227-017-3187-9}

\bibitem[\citeproctext]{ref-middleton_green-wave_2018}
45. Middleton AD, Merkle JA, McWhirter DE, Cook JG, Cook RC, White PJ, et al. Green-wave surfing increases fat gain in a migratory ungulate. Oikos {[}Internet{]}. 2018 {[}cited 2022 Sep 2{]};127:1060--8. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/oik.05227}

\bibitem[\citeproctext]{ref-geremia_migrating_2019}
46. Geremia C, Merkle JA, Eacker DR, Wallen RL, White PJ, Hebblewhite M, et al. Migrating bison engineer the green wave. Proc Natl Acad Sci {USA} {[}Internet{]}. 2019 {[}cited 2022 Mar 2{]};116:25707--13. Available from: \url{http://www.pnas.org/lookup/doi/10.1073/pnas.1913783116}

\bibitem[\citeproctext]{ref-polansky_elucidating_2015}
47. Polansky L, Kilian W, Wittemyer G. Elucidating the significance of spatial memory on movement decisions by african savannah elephants using state--space models. Proc R Soc B {[}Internet{]}. 2015 {[}cited 2022 Feb 10{]};282:20143042. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rspb.2014.3042}

\bibitem[\citeproctext]{ref-nandintsetseg_variability_2019}
48. Nandintsetseg D, Bracis C, Leimgruber P, Kaczensky P, Buuveibaatar B, Lkhagvasuren B, et al. Variability in nomadism: Environmental gradients modulate the movement behaviors of dryland ungulates. Ecosphere {[}Internet{]}. 2019 {[}cited 2020 Nov 11{]};10. Available from: \url{https://onlinelibrary.wiley.com/doi/abs/10.1002/ecs2.2924}

\bibitem[\citeproctext]{ref-teitelbaum_how_2015}
49. Teitelbaum CS, Fagan WF, Fleming CH, Dressler G, Calabrese JM, Leimgruber P, et al. How far to go? Determinants of migration distance in land mammals. Festa-Bianchet M, editor. Ecol Lett {[}Internet{]}. 2015 {[}cited 2022 Sep 23{]};18:545--52. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/ele.12435}

\bibitem[\citeproctext]{ref-poessel_interpreting_2022}
50. Poessel SA, Woodbridge B, Smith BW, Murphy RK, Bedrosian BE, Bell DA, et al. Interpreting long‐distance movements of non‐migratory golden eagles: Prospecting and nomadism? Ecosphere {[}Internet{]}. 2022 {[}cited 2022 Nov 18{]};13. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1002/ecs2.4072}

\bibitem[\citeproctext]{ref-pretorius_movement_2020}
51. Pretorius MD, Leeuwner L, Tate GJ, Botha A, Michael MD, Durgapersad K, et al. Movement patterns of lesser flamingos \emph{phoeniconaias minor} : Nomadism or partial migration? Wildlife Biology {[}Internet{]}. 2020 {[}cited 2022 Nov 17{]};2020:1--11. Available from: \url{https://onlinelibrary.wiley.com/doi/10.2981/wlb.00728}

\bibitem[\citeproctext]{ref-bista_effect_2022}
52. Bista D, Baxter GS, Hudson NJ, Lama ST, Murray PJ. Effect of disturbances and habitat fragmentation on an arboreal habitat specialist mammal using {GPS} telemetry: A case of the red panda. Landsc Ecol {[}Internet{]}. 2022 {[}cited 2022 Aug 24{]};37:795--809. Available from: \url{https://link.springer.com/10.1007/s10980-021-01357-w}

\bibitem[\citeproctext]{ref-bradsworth_using_2022}
53. Bradsworth N, White JG, Rendall AR, Carter N, Whisson DA, Cooke R. Using thresholds to determine priorities for apex predator conservation in an urban landscape. Landscape and Urban Planning {[}Internet{]}. 2022 {[}cited 2022 Oct 18{]};228:104559. Available from: \url{https://github.com/ropensci/MODIStsp}

\bibitem[\citeproctext]{ref-mcclintic_effects_2014}
54. McClintic LF, Taylor JD, Jones JC, Singleton RD, Wang G. Effects of spatiotemporal resource heterogeneity on home range size of american beaver. J Zool {[}Internet{]}. 2014 {[}cited 2022 Nov 12{]};293:134--41. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/jzo.12128}

\bibitem[\citeproctext]{ref-lucherini_habitat_1996}
55. Lucherini M, Lovari S. Habitat richness affects home range size in the red fox vulpes vulpes. Behavioural Processes {[}Internet{]}. 1996 {[}cited 2021 Nov 29{]};36:103--5. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/0376635795000186}

\bibitem[\citeproctext]{ref-watson_ferruginous_2020}
56. Watson J. Ferruginous hawk (buteo regalis) home range and resource use on northern grasslands in canada. 2020 {[}cited 2022 Nov 8{]}; Available from: \url{http://rgdoi.net/10.13140/RG.2.2.32404.32648}

\bibitem[\citeproctext]{ref-simcharoen_female_2014}
57. Simcharoen A, Savini T, Gale GA, Simcharoen S, Duangchantrasiri S, Pakpien S, et al. Female tiger \emph{panthera tigris} home range size and prey abundance: Important metrics for management. Oryx {[}Internet{]}. 2014 {[}cited 2022 Nov 8{]};48:370--7. Available from: \url{https://www.cambridge.org/core/product/identifier/S0030605312001408/type/journal_article}

\bibitem[\citeproctext]{ref-campillo_effect_2012}
58. Campillo F, Lobry C. Effect of population size in a predator--prey model. Ecological Modelling {[}Internet{]}. 2012 {[}cited 2025 Feb 6{]};246:1--10. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0304380012003432}

\bibitem[\citeproctext]{ref-lee_effects_2011}
59. Lee S-H. Effects of the probability of a predator catching prey on predator--prey system stability. Journal of Asia-Pacific Entomology {[}Internet{]}. 2011 {[}cited 2025 Feb 7{]};14:159--62. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S1226861510001329}

\bibitem[\citeproctext]{ref-levin_problem_1992}
60. Levin SA. The problem of pattern and scale in ecology: The robert h. {MacArthur} award lecture. Ecology {[}Internet{]}. 1992 {[}cited 2024 Jan 31{]};73:1943--67. Available from: \url{https://esajournals.onlinelibrary.wiley.com/doi/10.2307/1941447}

\bibitem[\citeproctext]{ref-steixner-kumar_strategies_2020}
61. Steixner-Kumar S, Gläscher J. Strategies for navigating a dynamic world. Science {[}Internet{]}. 2020 {[}cited 2022 Mar 9{]};369:1056--7. Available from: \url{https://www.science.org/doi/10.1126/science.abd7258}

\bibitem[\citeproctext]{ref-mueller_social_2013}
62. Mueller T, O'Hara RB, Converse SJ, Urbanek RP, Fagan WF. Social learning of migratory performance. Science {[}Internet{]}. 2013 {[}cited 2022 Nov 24{]};341:999--1002. Available from: \url{https://www.science.org/doi/10.1126/science.1237139}

\bibitem[\citeproctext]{ref-abrahms_memory_2019}
63. Abrahms B, Hazen EL, Aikens EO, Savoca MS, Goldbogen JA, Bograd SJ, et al. Memory and resource tracking drive blue whale migrations. Proc Natl Acad Sci {USA} {[}Internet{]}. 2019 {[}cited 2022 Mar 3{]};116:5582--7. Available from: \url{http://www.pnas.org/lookup/doi/10.1073/pnas.1819031116}

\bibitem[\citeproctext]{ref-falcon-cortes_hierarchical_2021}
64. Falcón-Cortés A, Boyer D, Merrill E, Frair JL, Morales JM. Hierarchical, memory-based movement models for translocated elk (cervus canadensis). Front Ecol Evol {[}Internet{]}. 2021 {[}cited 2022 Feb 4{]};9:702925. Available from: \url{https://www.frontiersin.org/articles/10.3389/fevo.2021.702925/full}

\bibitem[\citeproctext]{ref-fagan_spatial_2013}
65. Fagan WF, Lewis MA, Auger-Méthé M, Avgar T, Benhamou S, Breed G, et al. Spatial memory and animal movement. Clobert J, editor. Ecol Lett {[}Internet{]}. 2013 {[}cited 2022 Mar 9{]};16:1316--29. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/ele.12165}

\bibitem[\citeproctext]{ref-logares_black_2012}
66. Logares R, Nuñez M. Black swans in ecology and evolution: The importance of improbable but highly influential events. Ideas in Ecology and Evolution {[}Internet{]}. 2012 {[}cited 2020 Feb 12{]}; Available from: \url{https://ojs.library.queensu.ca/index.php/IEE/article/view/4311}

\bibitem[\citeproctext]{ref-anderson_black-swan_2017}
67. Anderson SC, Branch TA, Cooper AB, Dulvy NK. Black-swan events in animal populations. Proceedings of the National Academy of Sciences {[}Internet{]}. 2017 {[}cited 2020 Jan 24{]};114:3252--7. Available from: \url{http://www.pnas.org/lookup/doi/10.1073/pnas.1611525114}

\bibitem[\citeproctext]{ref-riotte-lambert_environmental_2020}
68. Riotte-Lambert L, Matthiopoulos J. Environmental predictability as a cause and consequence of animal movement. Trends in Ecology \& Evolution {[}Internet{]}. 2020 {[}cited 2020 Nov 11{]};35:163--74. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0169534719302885}

\bibitem[\citeproctext]{ref-levins_evolution_1974}
69. Levins RA. Evolution in changing environments: Some theoretical explorations. 3. printing. Princeton, {NJ}: Princeton Univ. Press; 1974.

\bibitem[\citeproctext]{ref-van_baalen_alternative_2001}
70. Van Baalen M, Křivan V, Van Rijn PCJ, Sabelis MW. Alternative food, switching predators, and the persistence of predator‐prey systems. The American Naturalist {[}Internet{]}. 2001 {[}cited 2024 May 20{]};157:512--24. Available from: \url{https://www.journals.uchicago.edu/doi/10.1086/319933}

\bibitem[\citeproctext]{ref-fleming_ctmm_2021}
71. Fleming CH, Calabrese JM. Ctmm: Continuous-time movement modeling {[}Internet{]}. 2021. Available from: \href{https://github.com/ctmm-initiative/ctmm,\%20https://groups.google.com/g/ctmm-user}{https://github.com/ctmm-initiative/ctmm, https://groups.google.com/g/ctmm-user}

\bibitem[\citeproctext]{ref-r_core_team_r_2023}
72. R Core Team. R: A language and environment for statistical computing {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing; 2023. Available from: \url{https://www.R-project.org/}

\bibitem[\citeproctext]{ref-gurarie_correlated_2017}
73. Gurarie E, Fleming CH, Fagan WF, Laidre KL, Hernández-Pliego J, Ovaskainen O. Correlated velocity models as a fundamental unit of animal movement: Synthesis and applications. Mov Ecol {[}Internet{]}. 2017 {[}cited 2023 Sep 6{]};5:13. Available from: \url{http://movementecologyjournal.biomedcentral.com/articles/10.1186/s40462-017-0103-3}

\bibitem[\citeproctext]{ref-fleming_fine-scale_2014}
74. Fleming CH, Calabrese JM, Mueller T, Olson KA, Leimgruber P, Fagan WF. From fine-scale foraging to home ranges: A semivariance approach to identifying movement modes across spatiotemporal scales. The American Naturalist {[}Internet{]}. 2014 {[}cited 2022 Jul 26{]};183:E154--67. Available from: \url{https://www.journals.uchicago.edu/doi/10.1086/675504}

\bibitem[\citeproctext]{ref-mcelreath_statistical_2016}
75. McElreath R. Statistical rethinking: A bayesian course with examples in r and stan. Boca Raton: {CRC} Press/Taylor \& Francis Group; 2016.

\bibitem[\citeproctext]{ref-pettorelli_normalized_2011}
76. Pettorelli N, Ryan S, Mueller T, Bunnefeld N, Jedrzejewska B, Lima M, et al. The normalized difference vegetation index ({NDVI}): Unforeseen successes in animal ecology. Clim Res {[}Internet{]}. 2011 {[}cited 2022 Mar 8{]};46:15--27. Available from: \url{http://www.int-res.com/abstracts/cr/v46/n1/p15-27/}

\bibitem[\citeproctext]{ref-noonan_scale-insensitive_2019}
77. Noonan MJ, Fleming CH, Akre TS, Drescher-Lehman J, Gurarie E, Harrison A-L, et al. Scale-insensitive estimation of speed and distance traveled from animal tracking data. Mov Ecol {[}Internet{]}. 2019 {[}cited 2021 Jun 23{]};7:35. Available from: \url{https://movementecologyjournal.biomedcentral.com/articles/10.1186/s40462-019-0177-1}

\bibitem[\citeproctext]{ref-noonan_comprehensive_2019}
78. Noonan MJ, Tucker MA, Fleming CH, Akre TS, Alberts SC, Ali AH, et al. A comprehensive analysis of autocorrelation and bias in home range estimation. Ecological Monographs {[}Internet{]}. 2019 {[}cited 2020 Oct 23{]};89:e01344. Available from: \url{https://onlinelibrary.wiley.com/doi/abs/10.1002/ecm.1344}

\bibitem[\citeproctext]{ref-alston_mitigating_2022}
79. Alston JM, Fleming CH, Kays R, Streicher JP, Downs CT, Ramesh T, et al. Mitigating pseudoreplication and bias in resource selection functions with autocorrelation‐informed weighting. Methods Ecol Evol {[}Internet{]}. 2022 {[}cited 2022 Dec 12{]};2041--210X.14025. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/2041-210X.14025}

\bibitem[\citeproctext]{ref-silva_autocorrelationinformed_2022}
80. Silva I, Fleming CH, Noonan MJ, Alston J, Folta C, Fagan WF, et al. Autocorrelation‐informed home range estimation: A review and practical guide. Methods Ecol Evol {[}Internet{]}. 2022 {[}cited 2022 Jul 26{]};13:534--44. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/2041-210X.13786}

\bibitem[\citeproctext]{ref-wood_smoothing_2016}
81. Wood SN, Pya N, Säfken B. Smoothing parameter and model selection for general smooth models. Journal of the American Statistical Association {[}Internet{]}. 2016 {[}cited 2020 Apr 3{]};111:1548--63. Available from: \url{https://www.tandfonline.com/doi/full/10.1080/01621459.2016.1180986}

\bibitem[\citeproctext]{ref-calabrese_ctmm_2016}
82. Calabrese JM, Fleming CH, Gurarie E. Ctmm: An {\textless{}}span style="font-variant:small-caps;"{\textgreater{}}r{\textless{}}/span{\textgreater{}} package for analyzing animal relocation data as a continuous‐time stochastic process. Freckleton R, editor. Methods Ecol Evol {[}Internet{]}. 2016 {[}cited 2025 Feb 14{]};7:1124--32. Available from: \url{https://besjournals.onlinelibrary.wiley.com/doi/10.1111/2041-210X.12559}

\bibitem[\citeproctext]{ref-fan_global_2016}
83. Fan X, Liu Y. A global study of {NDVI} difference among moderate-resolution satellite sensors. {ISPRS} Journal of Photogrammetry and Remote Sensing {[}Internet{]}. 2016 {[}cited 2024 Feb 14{]};121:177--91. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0924271616303975}

\bibitem[\citeproctext]{ref-gautam_ndvi_2019}
84. Gautam H, Arulmalar E, Kulkarni MR, Vidya TNC. {NDVI} is not reliable as a surrogate of forage abundance for a large herbivore in tropical forest habitat. Biotropica {[}Internet{]}. 2019 {[}cited 2024 Feb 14{]};51:443--56. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/btp.12651}

\bibitem[\citeproctext]{ref-huang_commentary_2021}
85. Huang S, Tang L, Hupy JP, Wang Y, Shao G. A commentary review on the use of normalized difference vegetation index ({NDVI}) in the era of popular remote sensing. J For Res {[}Internet{]}. 2021 {[}cited 2024 Feb 14{]};32:1--6. Available from: \url{https://link.springer.com/10.1007/s11676-020-01155-1}

\bibitem[\citeproctext]{ref-pease_ecological_2024}
86. Pease BS. Ecological scales of effect vary across space and time. Ecography {[}Internet{]}. 2024 {[}cited 2024 Aug 8{]};2024:e07163. Available from: \url{https://nsojournals.onlinelibrary.wiley.com/doi/10.1111/ecog.07163}

\bibitem[\citeproctext]{ref-brown_toward_2004}
87. Brown JH, Gillooly JF, Allen AP, Savage VM, West GB. Toward a metabolic theory of ecology. Ecology {[}Internet{]}. 2004 {[}cited 2022 Mar 3{]};85:1771--89. Available from: \url{http://doi.wiley.com/10.1890/03-9000}

\bibitem[\citeproctext]{ref-johnson_comparison_1980}
88. Johnson DH. The comparison of usage and availability measurements for evaluating resource preference. Ecology {[}Internet{]}. 1980 {[}cited 2024 May 17{]};61:65--71. Available from: \url{https://esajournals.onlinelibrary.wiley.com/doi/10.2307/1937156}

\bibitem[\citeproctext]{ref-rickbeil_plasticity_2019}
89. Rickbeil GJM, Merkle JA, Anderson G, Atwood MP, Beckmann JP, Cole EK, et al. Plasticity in elk migration timing is a response to changing environmental conditions. Glob Change Biol {[}Internet{]}. 2019 {[}cited 2022 Jan 20{]};25:2368--81. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/gcb.14629}

\bibitem[\citeproctext]{ref-ranc_memory_2022}
90. Ranc N, Cagnacci F, Moorcroft PR. Memory drives the formation of animal home ranges: Evidence from a reintroduction. Coulson T, editor. Ecology Letters {[}Internet{]}. 2022 {[}cited 2022 Nov 16{]};25:716--28. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/ele.13869}

\bibitem[\citeproctext]{ref-nespolo_why_2022}
91. Nespolo RF, Mejias C, Bozinovic F. Why bears hibernate? Redefining the scaling energetics of hibernation. Proc R Soc B {[}Internet{]}. 2022 {[}cited 2022 Nov 21{]};289:20220456. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rspb.2022.0456}

\bibitem[\citeproctext]{ref-goldberg_hibernation_2021}
92. Goldberg AR, Conway CJ. Hibernation behavior of a federally threatened ground squirrel: Climate change and habitat selection implications. Hayes L, editor. Journal of Mammalogy {[}Internet{]}. 2021 {[}cited 2022 Nov 21{]};102:574--87. Available from: \url{https://academic.oup.com/jmammal/article/102/2/574/6224539}

\bibitem[\citeproctext]{ref-reher_short_2018}
93. Reher S, Ehlers J, Rabarison H, Dausmann KH. Short and hyperthermic torpor responses in the malagasy bat macronycteris commersoni reveal a broader hypometabolic scope in heterotherms. J Comp Physiol B {[}Internet{]}. 2018 {[}cited 2022 Oct 3{]};188:1015--27. Available from: \url{http://link.springer.com/10.1007/s00360-018-1171-4}

\bibitem[\citeproctext]{ref-mohr_cellular_2020}
94. Mohr SM, Bagriantsev SN, Gracheva EO. Cellular, molecular, and physiological adaptations of hibernation: The solution to environmental challenges. Annu Rev Cell Dev Biol {[}Internet{]}. 2020 {[}cited 2022 Oct 3{]};36:315--38. Available from: \url{https://www.annualreviews.org/doi/10.1146/annurev-cellbio-012820-095945}

\bibitem[\citeproctext]{ref-merkle_spatial_2019}
95. Merkle JA, Sawyer H, Monteith KL, Dwinnell SPH, Fralick GL, Kauffman MJ. Spatial memory shapes migration and its benefits: Evidence from a large herbivore. Gaillard J, editor. Ecol Lett {[}Internet{]}. 2019 {[}cited 2022 Sep 5{]};22:1797--805. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1111/ele.13362}

\bibitem[\citeproctext]{ref-fretwell_territorial_1969}
96. Fretwell SD, Lucas HL. On territorial behavior and other factors influencing habitat distribution in birds: I. Theoretical development. Acta Biotheor {[}Internet{]}. 1969 {[}cited 2024 Apr 8{]};19:16--36. Available from: \url{http://link.springer.com/10.1007/BF01601953}

\bibitem[\citeproctext]{ref-ciuti_effects_2012}
97. Ciuti S, Northrup JM, Muhly TB, Simi S, Musiani M, Pitt JA, et al. Effects of humans on behaviour of wildlife exceed those of natural predators in a landscape of fear. Moreira N, editor. {PLoS} {ONE} {[}Internet{]}. 2012 {[}cited 2022 Aug 11{]};7:e50611. Available from: \url{https://dx.plos.org/10.1371/journal.pone.0050611}

\bibitem[\citeproctext]{ref-burson_competition_2018}
98. Burson A, Stomp M, Greenwell E, Grosse J, Huisman J. Competition for nutrients and light: Testing advances in resource competition with a natural phytoplankton community. Ecology {[}Internet{]}. 2018 {[}cited 2022 Sep 2{]};99:1108--18. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1002/ecy.2187}

\bibitem[\citeproctext]{ref-phillips_evaluating_2008}
99. Phillips LB, Hansen AJ, Flather CH. Evaluating the species energy relationship with the newest measures of ecosystem energy: {NDVI} versus {MODIS} primary production. Remote Sensing of Environment {[}Internet{]}. 2008 {[}cited 2024 Feb 14{]};112:4381--92. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0034425708002460}

\bibitem[\citeproctext]{ref-seigle-ferrand_systematic_2021}
100. Seigle-Ferrand J, Atmeh K, Gaillard J-M, Ronget V, Morellet N, Garel M, et al. A systematic review of within-population variation in the size of home range across ungulates: What do we know after 50 years of telemetry studies? Front Ecol Evol {[}Internet{]}. 2021 {[}cited 2022 Aug 16{]};8:555429. Available from: \url{https://www.frontiersin.org/articles/10.3389/fevo.2020.555429/full}

\bibitem[\citeproctext]{ref-merkle_large_2016}
101. Merkle JA, Monteith KL, Aikens EO, Hayes MM, Hersey KR, Middleton AD, et al. Large herbivores surf waves of green-up during spring. Proc R Soc B {[}Internet{]}. 2016 {[}cited 2024 Feb 14{]};283:20160456. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rspb.2016.0456}

\bibitem[\citeproctext]{ref-prox_framework_2020}
102. Prox L, Farine D. A framework for conceptualizing dimensions of social organization in mammals. Ecol Evol {[}Internet{]}. 2020 {[}cited 2022 Aug 23{]};10:791--807. Available from: \url{https://onlinelibrary.wiley.com/doi/10.1002/ece3.5936}

\bibitem[\citeproctext]{ref-grant_whether_1993}
103. Grant JWA. Whether or not to defend? The influence of resource distribution. Marine Behaviour and Physiology {[}Internet{]}. 1993 {[}cited 2022 Feb 3{]};23:137--53. Available from: \url{http://www.tandfonline.com/doi/abs/10.1080/10236249309378862}

\bibitem[\citeproctext]{ref-martinez-garcia_how_2020}
104. Martinez-Garcia R, Fleming CH, Seppelt R, Fagan WF, Calabrese JM. How range residency and long-range perception change encounter rates. Journal of Theoretical Biology {[}Internet{]}. 2020 {[}cited 2022 Mar 11{]};498:110267. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0022519320301223}

\bibitem[\citeproctext]{ref-tian_evaluating_2015}
105. Tian F, Fensholt R, Verbesselt J, Grogan K, Horion S, Wang Y. Evaluating temporal consistency of long-term global {NDVI} datasets for trend analysis. Remote Sensing of Environment {[}Internet{]}. 2015 {[}cited 2024 Feb 14{]};163:326--40. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0034425715001285}

\bibitem[\citeproctext]{ref-wood_generalized_2017}
106. Wood SN. Generalized additive models: An introduction with r. Second edition. Boca Raton: {CRC} Press/Taylor \& Francis Group; 2017.

\bibitem[\citeproctext]{ref-nathan_big-data_2022}
107. Nathan R, Monk CT, Arlinghaus R, Adam T, Alós J, Assaf M, et al. Big-data approaches lead to an increased understanding of the ecology of animal movement. Science {[}Internet{]}. 2022 {[}cited 2022 Mar 9{]};375:eabg1780. Available from: \url{https://www.science.org/doi/10.1126/science.abg1780}

\bibitem[\citeproctext]{ref-wood_generalized_2017-1}
108. Wood SN, Li Z, Shaddick G, Augustin NH. Generalized additive models for gigadata: Modeling the u.k. Black smoke network daily data. Journal of the American Statistical Association {[}Internet{]}. 2017 {[}cited 2022 Mar 11{]};112:1199--210. Available from: \url{https://www.tandfonline.com/doi/full/10.1080/01621459.2016.1195744}

\bibitem[\citeproctext]{ref-intergovernmental_panel_on_climate_change_climate_2023}
109. Intergovernmental Panel On Climate Change. Climate change 2021 -- the physical science basis: Working group i contribution to the sixth assessment report of the intergovernmental panel on climate change {[}Internet{]}. 1st ed. Cambridge University Press; 2023 {[}cited 2023 Jun 30{]}. Available from: \url{https://www.cambridge.org/core/product/identifier/9781009157896/type/book}

\bibitem[\citeproctext]{ref-grant_evolution_2017}
110. Grant PR, Grant BR, Huey RB, Johnson MTJ, Knoll AH, Schmitt J. Evolution caused by extreme events. Phil Trans R Soc B {[}Internet{]}. 2017 {[}cited 2022 Nov 18{]};372:20160146. Available from: \url{https://royalsocietypublishing.org/doi/10.1098/rstb.2016.0146}

\bibitem[\citeproctext]{ref-rypkema_modeling_2021}
111. Rypkema D, Tuljapurkar S. Modeling extreme climatic events using the generalized extreme value ({GEV}) distribution. Handbook of statistics {[}Internet{]}. Elsevier; 2021 {[}cited 2023 Oct 18{]}. p. 39--71. Available from: \url{https://linkinghub.elsevier.com/retrieve/pii/S0169716120300511}

\bibitem[\citeproctext]{ref-yao_emergence_2022}
112. Yao Q, Fan J, Meng J, Lucarini V, Jensen HJ, Christensen K, et al. Emergence of universal scaling in weather extreme events. 2022 {[}cited 2022 Nov 20{]}; Available from: \url{https://arxiv.org/abs/2209.02292}

\end{CSLReferences}

\end{document}
